### 2011-08-03 doc/paper_ambient/paper_ambient.tex
clean up

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN PACKAGE
\documentclass[11pt]{amsart}       % AMS template just for the draft
\usepackage{geometry}                 % See geometry.pdf to learn the layout options. There are lots.
\geometry{a4paper}                        % .letter or a4paper or a5paper or ... 
%\geometry{landscape}                 % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}     % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}                   % FIgure
\usepackage{amssymb}                 % Math symbol
\usepackage{epstopdf}                   % for working in .pdf and  not .dvi (convert .eps, .jpg,  etc to .pdf)
\usepackage{psfrag}
\usepackage[T1]{fontenc}
\usepackage{aeguill}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% END PACKAGE

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN AUTHORS AND INSTITUTIONS
\title{AMBIENT; an  execution and an MPI-CUDA scheduler  }
\author[A. Kosenkov, T. Ewart, B. Bauerb, A. Kantian,  M. Troyer, T. Giarmarchy]{Alexandr Kosenkov, Timoth\'ee Ewart, Bela Bauerb, Adrian Kantian, Matthias Troyer and Thierry Giamarchy }
%\author[UNIGE, ETHZ, CSCS]{UNIGE, ETHZ, CSCS}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN AUTHORS AND INSTITUTIONS

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN MACRO
\makeatletter
\newcommand\etc{\textit{etc}\@ifnextchar.{}{.\@}}
\newcommand\eg{\textit{e.g. }}
\newcommand\ie{\textit{i.e. }}
\newcommand\cf{\textit{c.f. }}

\makeatother
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% END MACRO
\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN SYMBOL
\def\cloud{\includegraphics[width=0.2cm]{FIGURES/cloud}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% END SYMBOL
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN INTRODUCTION
\section{Introduction}

%Nowadays, the numerical simulations get an essential position, as experiments,  for studying physical 
%properties of many problems. The numerical simulations are presented inside every fields of sciences 
%and engineering, as classical and quantum mechanic, quantum chemistry,  \etc. As usual, the users always
%want increase the size of their problems, as billion number of particles for a kinetic simulation, billion 
%of meshes for a fluid dynamic computation, or else, matrixes of million entries for a quantum mechanic,
%or the linear algebra problems. Thus, the high performance computing is the solution for solving very large problem,
%with the peta flops  and the future generation of exa flops computers, the engineers and the scientists will
%have a quasi unlimited set of resources for modeling, nevertheless, they will be confronted to a challenge, have compatible
%programs for these machines.

%A large set of programs; solvers, and software solutions are now availabled to solve large problem in a lot of fields, \eg we can 
%cite as popular in a few scientist field :  {\sc Quantum ESPRESSO} \cite{QE-2009} in quantum chemistry, Abinit \cite{ABINIT-2005, ABINIT-2009} in quantum mechanic,  ScaLAPACK \cite{SCALAPACK-1997} in linear 
%algebra,  or else SMILE in hypersonic rarefied flow \cite{SMILE-2005}. Of course, this short list is not exhaustive,
%and it can be fill up easily in every fields of science and engineering.

%If the scientists or the  engineers have the right tools, it will be necessarily  limited by the scalability of the code, in the case of the famous mathematical library ScaLAPACK, the
%efficiency  and thus the scalability, with the appropriated setting will reach up to 512 processors ($\sim  \mathcal{O}(100)$) \cite{SCALAPACK-1992}, beyond the interest is useless
%due to the degradation of the performance. In the worst case, the programmers will start  from a white page, and must design and efficient code.
%Due to the incredible number of cores, and optional accelerators as GPU, FGPA or "CELL like\footnote{Although the CELL processor is now depreciated, it herebies  
%a way of integrated processor (CPU+GPU), as the prove the last processor products of Intel (core i series) and AMD (fusion).}", the programmers 
%will have to introduce a mix programming model usually named mix-mode, based on a main MPI layer coupled with one or several additional layers \cite{EXASCALE-2010}.

%The most popular layer for mix-mode is OpenMP, and it is based on the preprocessor directives \texttt{\#pragma}, it demonstrated its performance. It is included successfully inside
%the  linear algebra libraries like the MKL from Intel or the PESSLSMP from IBM.  This ideology of the utilization of the pre-directives becomes more and more 
%important, thus it is used for vectorization of loops using the intel compiler and even for GPU with the PGI compiler. Although this ideology gives results, there is 
%a limitation for the main following reason: the creation/destruction of the threads will be extremely time consuming,
%when the  grain problem size becomes too fine \cite{SMITH-2001}. To achieve the best performance, the number of thread should be  equal to the number of cores on a socket, 
%thus the scalability will be increased at maximum by an order of magnitude.  The maximum number of processors now reach, becomes ($\sim \mathcal{O}(1000)$). 
%This ideology of implementation of the mix mode is more an evolution of the programming model than a revolution. It is an fast hand-on to implement that prolongates the life and the performance on a existing code. But,
%it will not prepare at all the code for the future, that will be automatically depreciated  by the new generation of cluster, in term of scalability.

%For getting several order of magnitude higher, we need a new programming model. The introduction of the GPU computing by NVidia a few years ago was
%a real innovation in HPC \cite{NICKOLLS-2010}. On a GPU, it is usual to program  thousand of threads. Thus, in this paper, we propose a innovative programming model to program on very large distributed cluster. We develop
%a framework named Ambient (C++) where we expend the CUDA programming ideology to  the a full cluster, thus we keep away the MPI programming model from the users.  In a way, we  extended
%the philosophy of OpenCL, OpenCL extents  the GPU programming model to a processor, assimilate a socket as a GPU, where every cores of the socket are assimilated to a streaming processor.
%We also developed inside a dynamic stack manager, that stocks and manages the execution of numerical "operations-solver" in functions of the resources of the distributed cluster.

%In this paper, we are going to describe our object oriented framework, and promote a first application based on linear algebra. Our main objective, is to 
%promote a new technology to develop  parallel application, without any interaction with the MPI and thread layer, and allow, a real access in term of performance to the next generation
%of clusters. The first possible  targets are applications that require large buffer of contiguous memory: quantum mechanic, linear algebra,  computational fluid dynamics, molecular dynamic \ldots 
%\, Until the end of this paper, we will focus on the linear algebra problems and their applications.

- Stress Performance of Ambient 
  - in terms of flexible and dynamical resource allocation
  - Stress performance boost derived from this (we need the benchmarks for that!)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% END INTRODUCTION

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN DESCRIPTION OF THE FRAMEWORK

\section{Design of the software layer}
In this section, we describe our framework in simple terms, without  a deep technical description.
We will focus on the main functionalities  and show the essential features.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN CUDA IDEOLOGY SECTION
\subsection{Extending the CUDA approach} \label{CUDAIDEOLOGYSECTION}

Ambient employs an extension on the CUDA approach to a distributed cluster. This is based on the observation that modern clusters employ many thousands of nodes, just like a GPU consists of thousands of small, simple cores, on each of which an runs an independent thread. In such a comparison however, the technical specifics of the basic cluster unit in terms of processor power, memory bandwidth, network speed, \etc must be considered. A basic comparison between CUDA and Ambient is shown in the table \ref{ANALOGY}.

First, given the entirely different capabilities of a CPU core over a single GPU core, the scope of data for a single CPU (on which a single Ambient "thread" is typically running) in Ambient is rather comparable to that of a CUDA block or an OpenCL work group, the typical  blocks of size is $ 128 \times 128$.
Of course, the user has full control of the size and can adapt it to the treated problem. Like in CUDA, blocks (i.e. groups of "threads") and a grid of blocks can then be formed in Ambient.

\begin{figure}[t]
	\begin{center}
	                  %psfrag for changing the layout
	                  \psfrag{New data layout}{\tiny{new layout}}
			\includegraphics[scale=0.44]{FIGURES/MemMatrix.eps} 
	\end{center}
          \caption{Representation of the data layout inside the memory for linear algebra. We move from a single contiguous column data layout (left ) to several tile data layouts (right).}
          \label{MEMMATRIX}
\end{figure}

The work group approach imposes the adaptation of the data layout to the framework. In linear algebra,  BLAS \cite{LAWSON-1979} is \textit{de facto}  the standard for high performance dense Linear Algebra computations. It imposes column order on matrix storage, as well as a contiguous storage in memory (figure \ref{MEMMATRIX}).

For the sake of backward compatibility with the standard BLAS-based mathematical libraries (BLAS, LaPACK ScaLAPACK), we keep, the column order, but do not conserve a full contiguous buffer. As in the plasma library \cite{LTAIEF-2011}, we select
a tile data layout. It confers the ability to fit perfectly with the work group representation of Ambient, and further allows an excellent MPI/Thread control (section \ref{MPIABSTRACTION}). One of the main features of Ambient is to map this new data layout  to our work group approach, as shown in Fig. \ref{CUDAANALOGY}.

\begin{table}[h]
	\begin{center}
	 	\begin{tabular}{c  c  c}
		 	                                                      & CUDA-OpenCL                                & Ambient            \\  \hline 
			 Device                                        & Graphic card/Processor (socket)   & Cluster               \\  
			  Entity  of Calculation               & GPU-Thread                                      & CPU-Thread     \\  
			 \begin{tabular}{c}
			 Scope of the \\ 
			 entity calculation  
                             \end{tabular}
                             &  Buffer entry                            & Full buffer \\
			 Execution model                       & Kernel                                                &  \begin{tabular}{c}  Logistic and \\ computational kernel   \end{tabular} \\
			 work group-ID                           &  \texttt{blockIdx}                                 & \texttt{get\_block\_id(A)}   \\  
			 work group size                        &  \texttt{blockDim}                               & \texttt{get\_dim\_id(A)}      \\  
			 grid size                                     &  \texttt{gridDim}                                  & \texttt{get\_grid\_dim(A)}  \\ \hline

		\end{tabular}
		\caption{Analogy between our framework and  the CUDA and OpenCL ideology.  \texttt{A} is the associated matrix to the  logistic and computational kernel.  
		The access of the cartesian values are done by the classical operator C/C++ \guillemotleft.\guillemotright     \, \eg \texttt{get\_dim\_id(A).x}  	\label{ANALOGY}}
	\end{center}
\end{table}

In CUDA and OpenCL, the execution model imposes kernel initialization with the appropriate resources (\eg numthread, numblock and grid dimension),
after which the user does not have to manage the memory access inside GPU memory; further, all threads can readily access any part of the GPU. Due to memory distribution in clusters we need to adapt this method of execution. Thus, we split this execution into two parts.

First, we prepare the memory, and then execute our "numerical" operation \eg a matrix multiplication in the following sequence \footnote{From here onwards, we will illustrate the execution model through parallel matrix addition and multiplication.}:
\begin{enumerate}
	\item A logistic kernel; where we select the number of processors, and map the memory with the datas.
	\item A computational kernel; where we execute the associated functions (the kernel in CUDA).
\end{enumerate}

\begin{figure}[!t]
	\begin{center}
           	        \psfrag{Ambient representation}{\tiny{\hspace{-0.4cm}Ambient representation}}
           	        \psfrag{CUDA/OpenCL representation}{\tiny{\hspace{-0.9cm}CUDA/OpenCL representation}}
	                 \psfrag{Associated mapping of the data with Ambient}{\tiny{\hspace{+0.15cm}Associated data mapping}}
	                 \psfrag{1x1 workgroup}{\tiny{\hspace{-0.5cm} \vspace{-0.1cm}1x1 workgroup}}
	                  \psfrag{of Ambient}{\tiny{\hspace{-0.4cm}\vspace{-0.9cm} of Ambient}}
	                  \psfrag{Workgroup of 4X4 threads}{\tiny{\hspace{-0.5cm} \vspace{-0.1cm}4x4 workgroup}}
	                  \psfrag{of CUDA}{\tiny{\hspace{-0.25cm}of CUDA}}
	                  \psfrag{Scope of GPU thread (left)}{\tiny{\hspace{-0.25cm}GPU scope left}}
	                  \psfrag{CPU (right)}{\tiny{\hspace{-0.7cm}CPU scope right}}       
		        \includegraphics[scale=0.44]{FIGURES/CUDAanalogy.eps} 
	\end{center}
          \caption{Corresponding of the CUDA representation, to the ambient representation, in terms of the scope of the CPU with the associated data.}
          \label{CUDAANALOGY}
\end{figure}

Thus, every operations will have an associated logistics and computational kernel \eg multiplication, addition, or more sophisticated linear algebra operations

\subsubsection{The logistics kernel} The role of the logistics kernel is to initialize both memory and data for execution. It consists of creating a MPI-group and mapping the data to the group with the appropriate pattern (1D or 2D block-cyclic distribution for a matrix, or alternatively
a row/column-wise block distribution for two-dimensional Finite Difference Methods in fluid dynamics).  The creation of the group is done by an SQL-like request, and the data is mapped into this newly created group using a modular assignment
function. Both these functionalities are described in section \ref{SQLSECTION}.

In linear algebra, the optimal distribution in terms of load balancing is the 2D block-cyclic distributions, esp. in ScaLAPACK, but this will generally depend of the solver used. For example, in a QR factorization \cite{DEMMEL-2008} it can be demonstrated that a row-wise distribution is superior to a 2D block-cyclic decomposition for matrices with more rows than columns. Switching from one distribution scheme to another, which is a challenge in ScaLAPACK, is easily accomplished with Ambient.
 \begin{table}[t]  
	\begin{center}
	 	\begin{tabular}{l l }
                           \hline 
                           \textbf{Logistic kernel }  Algorithm of the addition logistics kernel  &    \\ \hline
                           \textbf{Require}                 A class Matrix  that fits with Ambient framework  &  \\
                           \begin{tabular}{c l} 
                           \tiny{1:} &  Addition\_logistic(Matrix $A$,  Matrix $B$)  \\
                           \tiny{2:} &  SQL request for selecting $N$ number of processors   \\
                           \tiny{3:} &  Perform 2/1-D cyclic distribution of the Matrix $A$  $\vartriangleright$ Note: the distribution \\
                                         & is performed with the number of processors selected inside the SQL request. \\
                           \tiny{4:} &  Perform 2/1-D cyclic distribution of the Matrix $B$ \\
%                           \tiny{5:} &  Perform 2/1-D cyclic distribution of the Matrix $C$ \\ 
                           \end{tabular} & \\ \hline
		 \end{tabular}  
		 \caption{Logistic kernel associated to the matrix multiplication or  the addition.  \label{LKERNEL}}
	\end{center}
\end{table}

To illustrate, we describe the example of a simplified\footnote{without any C++ language annotations} logistics kernel of the matrix addition in table {\ref{LKERNEL}}. 
\begin{itemize}
\item Line 1:  the signature of the logistics kernel, showing the input/ouput by the keyword\texttt{const}. In this example there are onlz three matrices, but additional arguments like float number, integer \etc can be added.
\item Line 2: The SQL-like request selecting the desired number of processors for the operation and performing the required MPI calls. 
\item Line 3-5: Performing a block-cyclic distribution of the matrixes \texttt{A} and  \texttt{B}. A detailed description of the SQL requests and the distribution functions
is given in section \ref{SQLSECTION}. Note that the present kernel can be reused for any subsequent operation that uses the same data distribution.
\end{itemize}

\subsubsection{The computational kernel}  After the logistics kernel has been executed, we are ready for the computational kernel. The mode of execution is similar to the CUDA API with a few specificities. In the CUDA parallel computing model, the programmer writes a single thread program, using the CUDA-API,  after which the GPU runs multiple instances of this thread in parallel. We transpose this approach into Ambient. Nevertheless, we should not forget the data scope of a single CPU "thread" in Ambient roughly corresponds to the scope of a full OpenCL work group.

We have introduced a keyword (as the \texttt{const} keyword) named \texttt{pinned}. This keyword indicates to the framework than the kernel execution must be executed over all Ambient work groups. Inside a kernel, we employ several Ambient API, analogous to CUDA, to obtain key information about the work groups 
(positions, dimensions, \ldots).These are summarized in table \ref{ANALOGY}.  For illustration, the computational kernel (table \ref{CKERNEL}) of an addition of the two matrices associated with the previous example of the logistics kernel is presented in the table \ref{LKERNEL}.
\begin{itemize}
\item Line 1: the signature of the computational kernel must be the same than that of the associated logistics kernel. The only difference is the use of the keyword \texttt{pinned}, indicating an execution over all Ambient work groups.
\item Line 2: the cartesian coordinates of the work group are obtained from the Ambient instruction \texttt{get\_block\_id(A).x}  for $i$ and \texttt{get\_block\_id(A).y}  for $j$. \label{CURRENT}
\item Line 3 to 5 : we refer the work group $i,j$ to a temporary variable $A_{ref}$ using the Ambient instruction \texttt{current(A)(i,j)}. Note that if the processor requires data outside its memory, it will perform a remote access in the background (see section ... for details).
\item Line 6 to 8 : the addition is performed by looping over the total size of the work group, in parallel for each workgroup (because of the use of the keyword \texttt{pinned}.
\end{itemize}

 \begin{table}[t] 
	\begin{center}
	 	\begin{tabular}{l l }
                           \hline 
                           \textbf{Computatiomal kernel }  Algorithm of the addition computational kernel  &    \\ \hline
                           \textbf{Require}                 A class Matrix  that fits with Ambient framework  &  \\
                           \begin{tabular}{c l} 
                               \tiny{1:} &  Addition\_computational(Matrix $A$,  \texttt{pinned const}  Matrix $B$) \\
                               \tiny{2:} &  Get the cartesian coordinate $i,j$ of the work group \\ 
                               \tiny{3:} &  Refer  the  work group$(i,j)$  of A into $A_{ref}$ \\
                               \tiny{4:} &  Refer  the  work group$(i,j)$  of B into $B_{ref}$ \\
%                               \tiny{5:} &  Refer  the  work group$(i,j)$  of C into $C_{ref}$  \\
                               \tiny{5:} &  \textbf{for} $k$ from $0$ to total size of the work group \textbf{do} \\
                               \tiny{6:} & \hspace{0.2 cm}   $A_{ref}[k] \, + \!\! = B_{ref}[k]$ \\
                               \tiny{7:} &  \textbf{end for} \\                               
                          \end{tabular} & \\ \hline
		 \end{tabular} 
		 \caption{Algorithm of the computational kernel associated to the matrix addition.   \label{CKERNEL}}
	\end{center}
\end{table}

As just shown for the matrix addition, we can recode other linear algebra solvers with the Ambient approach. For the presentation of our framework, we will focus on the MPI version of BLAS named pBLAS \cite{SCALAPACK-1992}, as well as a possible alternative to the pDGEMM solver, called ...., using a recursive algorithm.

%\begin{table}[h]
%	\begin{center}
%	 	\begin{tabular}{c  c  c}		 	                                                             & CUDA/OpenCL                                 & Ambient                               \\  \hline 
%			 work group-ID                                  &  \texttt{blockIdx}                                 & \texttt{get\_block\_id(A)}   \\  
%			 work group size                               &  \texttt{blockDim}                               & \texttt{get\_dim\_id(A)}      \\  
%			 grid size                                            &  \texttt{gridDim}                                  & \texttt{get\_grid\_dim(A)}  \\ \hline
%				\end{tabular}
%		\caption{Analog¤y between Ambient and  the CUDA/OpenCL ideology. \texttt{A} is the associated matrix to the present logistic and computational kernel.  The access of the cartesian values are done by the classical operator C/C++ \guillemotleft.\guillemotright     \, \eg \texttt{get\_dim\_id(A).x}  } 	
%		\label{ANALOGY2}
%	\end{center}
%\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% END CUDA IDEOLOGY SECTION

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN SQL SECTION

\subsection{SQL resource management}  \label{SQLSECTION}
 In the previous section, we introduced the logistics kernels and their role in initializing resources for the execution of the computational kernels. 
 One of Ambients' main qualities is the ability to perform a dynamic management of the CPU resources of the cluster. To facilitate this, we integrated a light SQL framework [cite something] inside Ambient, which selects the desired number of processors inside the logistics kernel, in order to execute the computational kernel.

At the beginning of the framework execution, a main MPI group is created named \texttt{ambient}. To create groups inside this main group we used the abilities of MPI to build nested independent groups of processors  (chapter 6 of  \cite{MPI-2002}).  We encapsulated these features in the framework, and we control them through the SQL requests. This allows for a simple yet flexible management  of the MPI groups.  
These requests (done by the \texttt{scope\_select} Ambient-function) are applied to the logistics/computational addition kernels of our example by the following SQL syntax: \\

\begin{eqnarray}
       & &\texttt{scope\_select("\textit{N} from ambient as addition\_group }   \label{SQL1} \\
       & &\texttt{	where master is 0")}, \nonumber
\end{eqnarray}
This command selects \texttt{\textit{N}} processors from the original MPI group  \texttt{ambient}. The newly created group is 
named \texttt{addition\_group}, and its master processor locally has MPI-rank 0 (see Fig. \ref{MPIGROUP}).

\begin{figure}[h]
	\begin{center}
           	         \psfrag{Datas distributed over work group}{\tiny{\hspace{-0.5cm}Datas distributed over work group}}
	                  \psfrag{Local MPI rank}{\tiny{Local MPI rank}}                
	                  \psfrag{Ambient master proc}{\hspace{-0.8cm} \tiny{Ambient master proc}}        	                  
	                   \psfrag{Ambient MPI group, N process}{ \tiny Ambient MPI group, cardinal $N$ process}
	                   \psfrag{Local rank}{\tiny Local rank}
	         	         \psfrag{0}{\tiny{0}}
	                  \psfrag{1}{\hspace{-0.05cm}\tiny{1}}
	                  \psfrag{2}{\hspace{-0.15cm} \tiny{2}}
	                  \psfrag{MPI group creation}{ \hspace{-0.5cm} \tiny  Group creation}
	                  \psfrag{from the SQL request}{ \hspace{-0.6cm} \tiny  from SQL request}
	                  \psfrag{Mapping of the work group throw}{ \hspace{-0.6cm} \tiny  Mapping of the work group throw}
	                  \psfrag{the mpi GROUP created}{\tiny the new MPI group}
	                 \psfrag{(0,0)}{\hspace{-0.25cm} \tiny (0,0)}
	                 \psfrag{(0,1)}{\hspace{-0.25cm} \tiny (0,1)}
		        \psfrag{(0,2)}{\hspace{-0.25cm} \tiny (0,2)}
		        \psfrag{(1,0)}{\hspace{-0.25cm} \tiny (1,0)}
		        \psfrag{(1,1)}{\hspace{-0.25cm} \tiny (1,1)}
		        \psfrag{(1,2)}{\hspace{-0.25cm} \tiny (1,2)}
	                 \psfrag{(2,0)}{\hspace{-0.25cm} \tiny (2,0)}
		        \psfrag{(2,1)}{\hspace{-0.25cm} \tiny (2,1)}
		         \psfrag{(2,2)}{\hspace{-0.25cm} \tiny(2,2)}
                           \psfrag{n}{\hspace{-0.25cm} \tiny $n$}
	                  \includegraphics[scale=0.35]{FIGURES/MpiGroupData.eps} 
	\end{center}
          \caption[]{Creation of a MPI group by the SQL request, with \hbox{\cloud \hspace{0.02cm}} representing a MPI group with its own communicator;  $\bullet$ represents the master 
          process inside the group whereas the $\circ$ are the associated slave processors inside the corresponding group. Also shown is the assignment of the data blocks (each block corresponding to an Ambient work group) inside the newly created MPI group. In this example, the matrix is mapped, block by block, on the processors using a 1D column-cyclic distribution.}
          \label{MPIGROUP}
\end{figure}

After the SQL request, the data (workgroup) is mapped to the processes inside the newly created MPI group by a distribution function (\eg 1D or 2D block-cyclic distribution), the algorithm of which is described in table \ref{2DCYCLIC} (\cf Fig. \ref{MPIGROUP}). After this transfer, the processes of the MPI group are the primary owners in terms of data access and execution. In the Ambient approach, every operation is associated with
a logistics and a computational kernel, therefore the SQL request will be called several times. To create new groups, processes from the existing group are selected in ascending numerical rank, until all processes of the \texttt{ambient} group 
have been selected. If further groups are required, the selection is restarted from the beginning, thus a processor can belong to multiple groups simultaneously.

Every new group will have its own intra-communicator to allow intra-group communication, as well as an inter-communicator to allow for communication between groups, using the master process of the \texttt{ambient} group.

The construction of nested MPI-groups is related to the ability of MPI to create virtual and cartesian topologies  (chapter 7 of  \cite{MPI-2002}). However, Ambient allows for the simultaneous construction of the MPI topology and the mapping of the  data (work group).

 \begin{table}[t] 
	\begin{center}
	 	\begin{tabular}{l l }
                           \hline 
                           \textbf{Assign workgroup}  Algorithm of the 1/2-D assign function  &    \\ \hline
                           \textbf{Require}        An new MPI group obtained by an SQL request  &  \\
                           \begin{tabular}{c l} 
                               \tiny{1:} &  Initialize $i$ in function of the new MPI rank  of the new group \\
                               \tiny{2:} &  Initialize $j$ in function of the new MPI rank  of the new group  $\vartriangleright$ Note:   \\
                                             &  the initialization of $i,j$ will fix   the distribution (1D or 2D) \\
                               \tiny{3:} &  \textbf{for} from $i$  to the number of the work group on $x$ \textbf{do} \\
                               \tiny{4:} &  \hspace{0.2 cm}  \textbf{for} from $j$  to the number of the work group on $y$ \textbf{do} \\
                               \tiny{5:} & \hspace{0.4 cm}    Assign the work group $(i,j)$ to the corresponding processor\\
                               \tiny{6:} &   \hspace{0.2 cm}    \textbf{end for} \\
                               \tiny{7:} &  \textbf{end for} \\                               
                          \end{tabular} & \\ \hline
		 \end{tabular} 
		 \caption{Algorithm of the distribution of the work group into the new MPI group.   \label{2DCYCLIC}}
	\end{center}
\end{table}

During the execution, many group will be created in practice, and data will have to be moved between groups.
To obtain better performance and to reduce the number of data transfers between groups, there are additional possible settings for the SQL request. For example, consider the previous illustration for the SQL request in case of the matrix addition operation ($ A \, +\!\!\!= B$).
If matrix $A$ has already been created and mapped on specific processors, the logistics kernel can create a MPI group that includes its previous owner group (created in the previous operation), thereby avoiding superfluous data transfer. The concrete SQL request would be as follows:

\begin{eqnarray}  \label{SQL2}
         & & \texttt{scope\_select("\textit{N} from ambient as addition\_group }\\ 
         & & \texttt{	where master is 0 where breakdown contains" + get\_id(\textit{A})}, \nonumber
\end{eqnarray} 
The function \texttt{get\_id(\textit{A})} will get a tag  of \texttt{\textit{A}} to know is present binding.

%\begin{figure}[h]
%	\begin{center}
%			\includegraphics[scale=0.35]{FIGURES/ComGroup.eps} 
%	\end{center}
%          \caption{Data exchange inside the group.}
%          \label{COMGROUP}
%\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% END SQL SECTION

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGINNING EXECUTION AND OPERATIONS SCHEDULING SECTION
\subsection{Execution model and Operations scheduling}
In a typical HPC application, the programming model imposes a parallel execution of every function of the code, step by step, where the data granularity
of the problem  has imposed the parallelization. In ambient, we developed an scheduler, an  \guillemotleft engine\guillemotright \, that privileged a queue execution model  (promoted  and popularized by the CELL processor), where operations (\eg linear algebra operations \ldots)
will be stacked up into a list  and then executing according the model of logistics/computation kernel. This model is inspired from computer architecture design;
the queue data flow execution in a processor \cite{DENNIS-1974}, nevertheless, we are working now on a larger scale where basic instructions ($+,-,*,/$) are  replaced 
by the pair logistics/computation kernels.

An HPC application can be always decomposed as a set of functions that perform numerical operations.  Start by a basic example, we want to perform the three following algebra operations :

\begin{eqnarray} 
\text{Operation 1:} & & A + \!\! =  B, \label{3OPERATIONS}  \\
\text{Operation 2:} & & A \times \!\!= 3,   \nonumber  \\ 
\text{Operation 3:} & & C \times \!\!= D. \nonumber
\end{eqnarray}

In the usual way, the three operations will be executed in parallel one after one. Transposing this example in Ambient's conception means; each operation is now  associated to a logistics and a computational kernel. In Practice, considering our addition example(\ref{3OPERATIONS}),  the users will use the following \texttt{push} function:

\begin{eqnarray}
     \texttt{push(Addition\_logistics, Addition\_computional, \textit{A}, \textit{B})},
\end{eqnarray}

where we have the appropriated logistics and computational kernel named \texttt{Addition\_logistics} and \texttt{Addition\_computational}  presented in the table \ref{LKERNEL} and \ref{CKERNEL} previously, followed by the two arguments \texttt{\textit{A}} and \texttt{\textit{B}}. During the execution of the application, operations will be stack up, and in an ideal case, the content of the stack should be executed at the end of the application. Nevertheless, the users have the possibility to execute the content of the stack during the code,
by the basic following instruction :

\begin{eqnarray}
      \texttt{playout().}
\end{eqnarray}

This command triggered on the engine of ambient. The engine will scan through the stack of operation and identifies the independent operations, and also mark the different dependencies by incrementing a local dedicated counter. Then, it adds a  dedicated MPI packet manager\footnote{A packet manager is a tool that manages the communication between processors.} (see section \ref{MPIABSTRACTION})  for every operations and performed all logistics kernels. Due to the dependencies, the data will be map in a first MPI group and marked as need in others, with a trace of the previous owner.
Finally, the independent computational kernels are executed followed by  the depending computational kernels in the appropriated order. In case of dependency of the data, the migration of the datas will be done by the packet manager. 
When all computational kernels are over, the stack is cleaned. Thus, this method combined to the execution of the computational kernels creates two layers of parallelization : the simultaneous execution of  kernels executed in parallel.  
A schematic of the stack execution is presented on Fig. \ref{EXECUTION}

\begin{figure}[h]
	\begin{center}
	      	        \psfrag{Operation 1}{\tiny{Operation 1}}
		        \psfrag{Operation 2}{\tiny{Operation 2}}
		        \psfrag{Operation 3}{\tiny{Operation 3}}
		        \psfrag{l/c kernel 1}{\tiny{l/c kernel 1}}
                           \psfrag{l/c kernel 2}{\tiny{l/c kernel 2}}
                           \psfrag{l/c kernel 3}{\tiny{l/c kernel 3}}
                           \psfrag{Abiemt}{\tiny{Ambient}}
                           \psfrag{schedule}{\tiny{Schedule}}
                           \psfrag{Dependency}{\tiny{Dependency}}
                           \psfrag{Time}{\tiny{Time}}
                           \psfrag{evolution}{\tiny{evolution}}
                           \psfrag{Mapping and execution}{\tiny{Mapping and execution}}                          
                           \psfrag{Ambient execution}{\tiny{Ambient execution}}                          
                           \psfrag{Usual execution}{\tiny{Usual execution}}                          
	                  \includegraphics[scale=0.35]{FIGURES/execution.eps} 
	\end{center}
          \caption{Model of execution of the ambient scheduler. On the classical approach (left) the operations will be executed in parallel step by step whatever the dependencies (marked by the symbol \guillemotleft [\guillemotright \, in the schematic). In ambient (right), operations will be associated to the pair logistics/computation (abbreviate l/c) kernel, as operations 1 and 3  (l/c kernels 1 and 3) are independents, the execution of computational kernel will be done, simultaneously.  Due to the dependency between the kernel 1 and 2, the computational kernel of operation 2 (l/c kernel 2) will be executed later. It should be performed in a new MPI group but  with an appropriate SQL request (\ref{SQL2}) in logistics kernel, we can execute the computational kernel where the data are. }   \label{EXECUTION}
\end{figure}

Now, we  generalized the example \ref{3OPERATIONS} to a vectors of matrices. We  add  conditional statements to know which matrices should add and multiply (Table \ref{GRAPH}). First, from line 1 to 11, as a dry run, the application execution will determine which operations should be performed, 
and it will pushed up needed kernels on the top of stack. When the \texttt{playout()} function is called line 12, the engine will execute the stack, by solving dependency, it will  create the directed  acyclic graph of the application, naturally. Moreover this example shows the strength of Ambient, we have integrated a double parallelization without any, hard coding of MPI instruction in the original code. Thus, a full parallel linear algebra  application can be developed without taking into account the geometry of the problem, which usually done.

\begin{table}[t] 
	\begin{center}
	 	\begin{tabular}{l l }
                           \hline 
                           Determination of the directed  acyclic graph determination   &    \\ \hline
                           \textbf{Require}    A class Matrix  that fits with Ambient framework   &  \\  
                           and  vector of matrices  of size $n$, $A[n]$, $B[n]$, $C[n]$ and $D[n]$ & \\
                           \begin{tabular}{c l} 
                               \tiny{1:}   & \textbf{for} from $i$  to the size of the vector \textbf{do} \\
                               \tiny{2:}   & \hspace{0.2 cm} \textbf{for} from $j$  to the size of the vector \textbf{do} \\                               
                               \tiny{3:}   & \hspace{0.4 cm}  \textbf{if}  the conditions on $A[i]$ and  $B[j]$ are verified \textbf{do} \\                               
                               \tiny{4:}   & \hspace{0.8 cm}  $A[i]+ \!\! =B[j]$, $\vartriangleright$  push l/c kernel 1, \\
                               \tiny{5:}   & \hspace{0.4 cm} \textbf{end if} \\   
                               \tiny{6:}   & \hspace{0.4 cm} $A[i]\times \!\! =3$,  $\vartriangleright$  push l/c kernel 2, \\                   
                               \tiny{7:}   & \hspace{0.4 cm} \textbf{if}  the conditions on $C[i]$ and  $D[j]$ are verified \textbf{do} \\                               
                               \tiny{8:}   & \hspace{0.8 cm} $C[i]\times \!\! =D[j]$,  $\vartriangleright$ push l/c kernel 3, \\
                               \tiny{9:}   & \hspace{0.4 cm}  \textbf{end if} \\       
                               \tiny{10:} & \hspace{0.2 cm}  \textbf{end do} \\   
                               \tiny{11:} & \textbf{end do} \\                                 
                               \tiny{12:} & \texttt{playout()} $\vartriangleright$  execute the stack \\
                               \tiny{13:} & $\vartriangleright$ Note:  $+\!\!=$ and $\times \!\! =$ operators have wrapper functions   \\
                                                & that push the good logistics and computational kernel. \\                               
                           \end{tabular}   &  \\ \hline
        	     \end{tabular}
	\end{center}
	\caption{Generalization of the example \ref{3OPERATIONS} with conditional statements. The association of a cumulative stack and  delate stack execution will determine the  directed  acyclic graph. \label{GRAPH}}
\end{table}

To conclude this section, unlike a typical parallel execution, this model have several decisive advantages. It allows a transparent parallelization of two layers of parallel execution, and most important, it determined the directed  acyclic graph on the full application in order to get the best performance.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% END EXECUTION AND OPERATIONS SCHEDULING SECTION

\subsection{Packet manager and MPI abstraction} \label{MPIABSTRACTION}
In a normal HPC application, the  data could be prepared when communications are needed, if the data have more than one type especially. In this case  
we should prepare a MPI structure and then perform the exchange. These preliminary conversions are painful for the performance. In  Ambient,
 the structure of data layout by work groups necessitates additional metadata to coordinate the good executions of logistics and 
 computational kernels. Particularly, when a remote access is performed in 
a computational kernel. Thus,  before the block of data (literally the numbers 
inside the matrix) of the any workgroups,  contiguous data  are included. We called this ensemble \guillemotleft block of data+metadata\guillemotright \, as a netweork packet. It is built
on a MPI structure. Thus the work groups  are  ready to send or receive without any additional movements or copying. 
To assure the good execution of the framework, several additional type of packets are  available, they all derived from a standard packet. 
%The control of the packet and MPI exchange is done by a packet manager.

When we execute a computational kernel, we get the data by the Ambient command  \texttt{current(\textit{A})(i,j)} (ref. \ref{CURRENT}), where \textit{A} is a matrix 
compatible with Ambient, and the indices $i$ and $j$ are the coordinates of the work group (Fig. \ref{MPIGROUP}). Three possibilities are possibles for getting the data when a process execute this functions
in a computational kernel.
\begin{enumerate}
\item The work group is already in the memory of the process. The access is direct.
\item The work group is in the memory space of the present MPI-group (previously created by the associated logistics kernel). The applicant process sends a query request (a specific packet) to its  master process of its MPI group.
The master process of the group gets the request (the master process knows the repartition of the workgroup inside the MPI-group), it sends a new request to
the actual owner, when the actual owner gets the request, it send the workgroup needed to the first applicant. 
\item The work group is inside an other MPI-group. As previously, the applicant process sends to the master process a query  request, and the master process of 
the actual MPI group  sends an other request to the older owner group. The older master process gets the message and sends a message to actual owner of the data. Then the owner 
of the data sends the data to the first applicant.
\end{enumerate}
We illustrate the two last previous situations on the figure \ref{MPIEXCHANGE}.
\begin{figure}[h]
	\begin{center}
	    \psfrag{case 2}{\tiny case 2}
	    \psfrag{case 3}{\tiny case 3}
             \includegraphics[scale=0.35]{FIGURES/ComGroup.eps} 
	\end{center}
          \caption{On the left, we illustrate communications inside the same group. On the right, we illustrate the communications between  two MPI-groups. The dashed circle indicated the previous MPI group. In both cases, the masters process have the key 
          role to distribute informations for the exchange. The requests (\texttt{mpi\_isend} and \texttt{mpi\_irecv}) are arrows, and the orders of operations are marked by numbers.}  \label{MPIEXCHANGE}
\end{figure}

In the previous section, we reveled the importance of the data exchange.   the scheduler, its perfect execution is relevant to  the packet manager.
The packet manager  manage the intra communication inside the same MPI group \eg to perform an multiplication, and the inter communication
between the MPI groups. The packet manager is unique for every MPI-group, it is based on a singleton pattern.

The role of the p

\subsection{Abstract parallel datatypes} % <- to change ?

\subsection{Validation}
As  validation test, we code a new implementation of PDGEMM using our framework, and we compare to the classical version of PDGEMM of ScaLAPACK (MKL-Version).
The algorithm of our implementation is based on the classical method using a columns order distribution for the matrices nevertheless, we give a description of the algorithm 
into the table \ref{AMBIENTDGEMM} to show the modification due to Ambient.

 \begin{table}[t] 
	\begin{center}
	 	\begin{tabular}{l l }
                           \hline 
                           \textbf{ Algorithm:}  pDGEMM using Ambient  &    \\ \hline
                           \textbf{Require}                 A pinned computational kernel  &  \\
                           \begin{tabular}{c l} 
                                 \tiny{1:} &  Get the cartesian coordinate $i,j$ of the work group \\ 
                                 \tiny{2:} &  Refer  the  work group$(i,j)$  of B into $B_{ref}$ \\
                                 \tiny{3:} &  \textbf{for} from $k(0)$  to the max number of rows \textbf{do} \\
                                 \tiny{4:} &  \hspace{0.2 cm} Refer  the  work group$(k,j)$  of A into $A_{ref}$ \\    
                                 \tiny{5:} &  \hspace{0.2 cm}  reduction of the blocks $(k,i)$ into $C_{ref}$ , $\vartriangleright$  The reduction, \\
                                                & \hspace{0.2 cm}  is done by a proxy pattern \\
                                 \tiny{6:} &  \hspace{0.2 cm}  classical DGEMM($A_{ref}$, $B_{ref}$, $C_{ref}$)\\
                                 \tiny{7:} &  \textbf{end do} \\    
                             \end{tabular} & \\ \hline
		 \end{tabular} 
		 \caption{Algorithm of the pDGEMM using Ambient framework. We remember a proxy pattern is an interface the something else (object in memory). There, it allows the reduction between the independent workgroups.   \label{AMBIENTDGEMM}}
	\end{center}
\end{table}

The main advantages in our implementation is the size of coding, it is limited at 7 lines  of code using our framework.

----------------------------------------------------------------------------------------------------- \\
TO DO EXPLAIN THE ADVANTAGE AGAINST   SCALAPACK \\
IN TERM OF NETWORK, DATA EXCHANGE \\
----------------------------------------------------------------------------------------------------- \\

The results will be compare with the ScaLAPACK implementation, in the section \ref{CUDAIDEOLOGYSECTION}, we announced a compatibility with the ScaLAPACK.
The representation of the memory of ScaLAPACK and Ambient have significant differences, although matrixes are subdivided into small block, the layout into
the memory is different. In Ambient all blocks named work group are independent even if they belong to the same process. In ScaLAPACK block will be regrouped
by process to get a contiguous memory, the multiplication between blocks will be assure by using a correct offset inside the DEGMM call. Therefore, in ambient if we execute a solver from ScaLAPACK,
one of the first operation to perform will be the rearrangement of the work groups of a same owner into a contiguous block. This operation add a certain overload.

The validation test consists of multiplying matrices and saving the results into an output matrix, as follow :
\begin{eqnarray}
C = A \times B
\end{eqnarray}
This test will be executed with the ambient computational kernel and the Scalapack PDGEMM solver inside Ambient (with the additional workload of rearrangement of the memory)
As we are using C++ language, the operations $\times$ and $=$ have been overloaded correctly by pushing the good logistics and computational kernels.
When the calculation is over for both methods results are compared element by element respecting the following test  \cite{Dongarra:1990:SLB:77626.79170} :

\begin{eqnarray}
\frac{\left \lvert C^S_{ij} - C^A_{ij} \right \rvert}{\left \lvert  \epsilon C^S_{ij} \right \rvert } < n,
\end{eqnarray}

Where  the precision factor $\epsilon$  is equal to the \texttt{double} precision $10^{-16}$, the indices  $S$ and $A$ represent the ScaLAPACK and the Ambient version,
indices $i$ and $j$ are the entries of the matrices and $n$ is equal to 16 (decimal digits), the exponent of the numerical precision.

For the validation the size of work groups have been fixed to $128 \times 128$, and the matrices are squares from 256 up to 8196 (by step of 128 elements). The numbers of MPI-Process varied from 2 to 8.
Matrices were filled up with pseudo random number generator  \texttt{drand48()}\footnote{This function returns 
pseudo-random numbers non-negative, double-precision,  uniformly distributed over the interval $[0.0 , 1.0]$ using a linear congruential algorithm and 
48-bit integer arithmetic. }  from \texttt{stdlib}.  The validation tests have always been verified successfully . The benchmarks are performed and evaluated in the next sections.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% END DESCRIPTION OF THE FRAMEWORK

\section{Results and discutions}
\subsection{First benchmark case}
The first benchmark consists of evaluating the performance of our parallel matrix multiplication. The initial condition are the same than the validation test. The benchmarks 
were performed on a AMD-2427 2.2 Ghz ..........

\subsection{Second benchmark case}
\subsection{Third benchmark case}

\section{Conclusions}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN BIBLIO
\bibliographystyle{abbrv}
\bibliography{biblio}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% END BIBLIO

\end{document}

### 2011-08-05 doc/paper_ambient/paper_ambient.tex
paper ambient, packet manager, figure, plot, + first modif from Alex

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%% END PACKAGE
+\graphicspath{{PLOT/SinglePDGEMM/}}
 
@@
 \title{AMBIENT; an  execution and an MPI-CUDA scheduler  }
-\author[A. Kosenkov, T. Ewart, B. Bauerb, A. Kantian,  M. Troyer, T. Giarmarchy]{Alexandr Kosenkov, Timoth\'ee Ewart, Bela Bauerb, Adrian Kantian, Matthias Troyer and Thierry Giamarchy }
+\author[A. Kosenkov, T. Ewart, B. Bauerb, A. Kantian,  M. Troyer, T. Giarmarchy]{Alexandr Kosenkov, Timoth\'ee Ewart, Bela Bauer, Adrian Kantian, Matthias Troyer and Thierry Giamarchy }
 %\author[UNIGE, ETHZ, CSCS]{UNIGE, ETHZ, CSCS}
@@
 
-\section{Design of the software layer}
-In this section, we describe our framework in simple terms, without  a deep technical description.
-We will focus on the main functionalities  and show the essential features.
+\section{Design of the framework}
+%In this section, we describe our framework in simple terms, without  a deep technical description.
+%We will focus on the main functionalities  and show the essential features.
 
@@
 This is based on the observation that modern clusters employ many thousands of nodes, just like a GPU consists of thousands of small, simple cores, on each of which an runs an independent thread.
-In such a comparison however, the technical specifics of the basic cluster unit in terms of processor power, memory bandwidth, network speed, \etc must be considered.
+In such a comparison however, the technical specifics of the basic cluster unit in terms of processor power, memory bandwidth, network speed, \etc  \, must be considered.
+
+The first thing coming in mind when we compare a  cluster and a GPU is the number of available cores.
+Assimilate a single  CPU core  to  GPU core is reducer  because a CPU has additional capacities than a GPU core.
+Thus, given the entirely different capabilities of a CPU core over a single GPU core, the scope of data for a single CPU (on which a single Ambient "thread" is typically running) in Ambient is rather comparable to that of a CUDA block or an OpenCL work group, the typical  blocks of size is $ 128 \times 128$.
+Of course, the user has full control of the size and can adapt it to the treated problem.
+Like in CUDA, blocks (i.e.
+groups of "threads") and a grid of blocks can then be formed in Ambient.
 A basic comparison between CUDA and Ambient is shown in the table \ref{ANALOGY}.
-
-First, given the entirely different capabilities of a CPU core over a single GPU core, the scope of data for a single CPU (on which a single Ambient "thread" is typically running) in Ambient is rather comparable to that of a CUDA block or an OpenCL work group, the typical  blocks of size is $ 128 \times 128$.
-Of course, the user has full control of the size and can adapt it to the treated problem. Like in CUDA, blocks (i.e. groups of "threads") and a grid of blocks can then be formed in Ambient.
 
@@
                            \hline 
-                           Determination of the directed  acyclic graph determination   &    \\ \hline
+                           Determination of the directed  acyclic graph   &    \\ \hline
                            \textbf{Require}    A class Matrix  that fits with Ambient framework   &  \\  
@@
 In both cases, the masters process have the key
-          role to distribute informations for the exchange.
-The requests (\texttt{mpi\_isend} and \texttt{mpi\_irecv}) are arrows, and the orders of operations are marked by numbers.}  \label{MPIEXCHANGE}
+          role to distribute informations for the exchange. The requests  are arrows, and the orders of operations are marked by numbers.}  \label{MPIEXCHANGE}
 \end{figure}
 
-In the previous section, we reveled the importance of the data exchange.   the scheduler, its perfect execution is relevant to  the packet manager.
-The packet manager  manage the intra communication inside the same MPI group \eg to perform an multiplication, and the inter communication
+In the previous section, we reveled the importance of the communications, theirs perfect execution is relevant to  the packet manager.
+The packet manager  manages the intra communication inside the same MPI group \eg to perform an multiplication, and the inter communication
 between the MPI groups. The packet manager is unique for every MPI-group, it is based on a singleton pattern.
 
-The role of the p
+ In the previous paragraph, we presented different communication blueprint when datas are requested by a process.
+As communication path between the applicant  and the owner of the data can be more and less complex,  to solve every communication scheme,  we designed the packet manager on two main things :
+the delegate pattern and a finite state machine (FSM).
+
+To reach the best performance in HPC,  synchronization operations should be avoided  \cite{EXASCALE-2010}, because it is  a bottleneck.
+Thus, we privileged  MPI asynchronous communication into Ambient (\texttt{mpi\_isend} and \texttt{mpi\_irecv}). Nevertheless, if a message 
+is sent by a asynchronous instruction, an MPI test is necessary to check if the packet can be received.  As we refused synchronization instructions,
+we repeat the process until receiving the message by an infinite loop. Thus we integrated the delegate pattern and the FSM inside the loop.
+
+\begin{table}[t] 
+	\begin{center}
+	 	\begin{tabular}{l l }
+                           \hline 
+                           Algorithm of the packet manager   &    \\ \hline
+                           \textbf{Require}  a MPI process require a block of data &  \\                           
+                            \begin{tabular}{c l}                                
+                               \tiny{1:}   & \textbf{for}  ;;  \textbf{do} $\vartriangleright$  infinite loop \\
+                               \tiny{2:}   & \hspace{0.2 cm}  \texttt{Delegate()}  $\vartriangleright$ check if  packet has been, \\
+                                                & and received and re-send if needed \\
+                               \tiny{3:}   & \hspace{0.2 cm}  FSM $\vartriangleright$ break the loop, \\
+                                               & if all packets arrived at destination. \\
+                               \tiny{4:}   & \textbf{end do} \\
+                           \end{tabular}   &  \\ \hline
+        	     \end{tabular}
+	\end{center}
+	\caption{Description of the packet manager}
+\end{table}
+
+\begin{figure}[h]
+	\begin{center}
+	    \psfrag{1}{\tiny \hspace{-0.1 cm}  1}
+	    \psfrag{2}{\tiny \hspace{-0.1 cm}  2}
+	    \psfrag{Loose}{\tiny \hspace{-0.2 cm}   Loose}
+	    \psfrag{Closure}{\tiny \hspace{-0.2 cm}  Closure}
+	    \psfrag{Open}{\tiny \hspace{-0.4 cm}  Opening}
+             \includegraphics[scale=0.45]{FIGURES/FSM.eps} 
+	\end{center}
+          \caption{Schematic of the finite sate machine for the closing procedure.
+The states 1 and 2 correspond to open or close.
+The transitions are loose, closure or opening, they  depend of the situation of all packets (if they all reach theirs targets or not),
+                          the loose transition means the packet has not been received, thus we iterate one more time on the infinite loop.}  \label{FSM}
+\end{figure}
+
+A delegate pattern (C\#) encapsulated methods,    It is similar to pointer of function under C++  but is allows a safer execution, we ported this pattern into Ambient.
+The packet manager class contains a delegate object. The delegate class contains a series of operations pill-up  in a stack. These operations consists of forwarding requests, 
+ block of data,  layout of the MPI group (block distribution) and theirs updates. As this delegate is included, into an infinite loop, the packets
+will propagate process by process after every iterations (as shown on Fig. \ref{EXECUTION}). At the end of the delegate execution, a finite state machine checks the states of the packets. 
+If all packets received (data packet, layout packet, \ldots), the FSM (Fig. \ref{FSM}) breaks the infinite loop.
 
@@
 into the table \ref{AMBIENTDGEMM} to show the modification due to Ambient.
+
+The main advantages in our implementation is the size of coding, it is limited at 7 lines  of code using our framework.
+
+----------------------------------------------------------------------------------------------------- \\
+TO DO EXPLAIN THE ADVANTAGE AGAINST   SCALAPACK \\
+IN TERM OF NETWORK, DATA EXCHANGE \\
+----------------------------------------------------------------------------------------------------- \\
+
+The results will be compare with the ScaLAPACK implementation, in the section \ref{CUDAIDEOLOGYSECTION}, we announced a compatibility with the ScaLAPACK.
+The representation of the memory of ScaLAPACK and Ambient have significant differences, although matrixes are subdivided into small block, the layout into
+the memory is different. In Ambient all blocks named work group are independent even if they belong to the same process. In ScaLAPACK block will be regrouped
+by process to get a contiguous memory, the multiplication between blocks will be assure by using a correct offset inside the DEGMM call. Therefore, in ambient if we execute a solver from ScaLAPACK,
+one of the first operation to perform will be the rearrangement of the work groups of a same owner into a contiguous block. This operation add a certain overload.
 
@@
 \end{table}
-
-The main advantages in our implementation is the size of coding, it is limited at 7 lines  of code using our framework.
-
------------------------------------------------------------------------------------------------------ \\
-TO DO EXPLAIN THE ADVANTAGE AGAINST   SCALAPACK \\
-IN TERM OF NETWORK, DATA EXCHANGE \\
------------------------------------------------------------------------------------------------------ \\
-
-The results will be compare with the ScaLAPACK implementation, in the section \ref{CUDAIDEOLOGYSECTION}, we announced a compatibility with the ScaLAPACK.
-The representation of the memory of ScaLAPACK and Ambient have significant differences, although matrixes are subdivided into small block, the layout into
-the memory is different. In Ambient all blocks named work group are independent even if they belong to the same process. In ScaLAPACK block will be regrouped
-by process to get a contiguous memory, the multiplication between blocks will be assure by using a correct offset inside the DEGMM call. Therefore, in ambient if we execute a solver from ScaLAPACK,
-one of the first operation to perform will be the rearrangement of the work groups of a same owner into a contiguous block. This operation add a certain overload.
 
@@
 
+\begin{figure}
+\include{PLOT/SinglePDGEMM/SinglePGEMM}
+\end{figure}
+
 \subsection{Second benchmark case}
@@
 \bibliography{biblio}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%% END BIBLIO
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% END BIBLIOr
 

### 2011-08-08 doc/paper_ambient/paper_ambient.tex
minor modifs paper

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%% END INTRODUCTION
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN WORKFLOW
+\section{Ambient workflow}
+
+Alex, your part !
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% END WORKFLOW
 
@@
  One of Ambients' main qualities is the ability to perform a dynamic management of the CPU resources of the cluster.
-To facilitate this, we integrated a light SQL framework [cite something] inside Ambient, which selects the desired number of processors inside the logistics kernel, in order to execute the computational kernel.
+To facilitate this, we integrated a light SQL framework \cite{SQLite} inside Ambient, which selects the desired number of processors inside the logistics kernel, in order to execute the computational kernel.
 

### 2011-08-10 doc/paper_ambient/paper_ambient.tex
Ambient Paper, new section, merge alex section + corrections Adrian

@@
 \newcommand\cf{\textit{c.f. }}
+\renewcommand{\labelitemi}{$\diamond$}
 
@@
 
-Alex, your part !
+In this section we describe the workflow of Ambient applications execution in general terms.
+As the Ambient approach derives from CUDA - first we will look at the simplified workflow of CUDA and Ambient-application's execution:
+
+\begin{table}[h]  
+	\begin{center}
+	 	\begin{tabular}{l l  l l} % L L  L L 
+                        \hline 
+                             \multicolumn{2}{c}{CUDA} &  \multicolumn{2}{c}{Ambient} \\
+                             \hline
+                             \multicolumn{4}{c}{  $\diamond$ Serial code}  \\
+                            $\diamond$ & Kernel encountered                                  &  $\diamond$ & Kernel-pair (logistic/computa- \\
+                               & & & tional) encountered \\
+                               $\diamond$ &Thread block grid construction&     $\diamond$  & Kernel-pair stacked for later \\
+                                &(gridDim, blockDim)  &  &  execution  \\ 
+                               $\diamond$ &  Thread blocks mapping onto de-  &   $\diamond$  &  Serial code   \\
+                                & vice (GPU SMPs) &  & \\
+                               $\diamond$ &  Execution of kernel: GPU SMP  &  $\diamond$&Stack flush signal encountered :  \\
+                               &  executes kernel on every thread& &   execution of kernels  \\
+                               &  of corresponding thread blocks  & & \\
+                                 $\diamond$  &  Serial code &  & \\
+                                  \hline
+        		 \end{tabular}  
+              \caption{Workflow analogy between CUDA and Ambient-application's execution} \label{WORKFLOW}
+	\end{center}
+\end{table}
+
+When the Ambient's flush signal indicates the execution of kernels. It is performed by the followings sequences:
+
+\begin{itemize}
+\item Execution of logistic kernel: allocation of device (MPI group) according to the request. 
+\item Execution of logistic kernel: argument's workgroups (tiles) marking with workDim.
+\item Execution of logistic kernel: selected (pinned) argument's workgroups mapping onto device (MPI processes).
+\item Transfer of workgroups to the target processes.
+\item Execution of computational kernel: MPI process executes kernel thread for work items of assigned work groups.
+\end{itemize}
+
+The main reason of the workflow difference comparing to CUDA is the scale of available resources. 
+As we can use unlimited amount of resources for a kernel's launch - the data becomes the key factor around which we evolve Ambient's design.
+In particular, in order to control the data movements and resource usage we introduce the logistic kernel and making kernel calls non-blocking.
+
+In general, Ambient application operates with abstract types that already have tiled memory structure. Upon the kernel operation 
+call the Ambient framework stacks the operation for later execution and proceeds to the rest of the code.
+
+When the framework encounters the signal to flush the stack of kernels - the actual execution starts. Framework goes through accumulated stack 
+and executes independent operations at once. When executing one operation first the engine goes through the logistic kernel and allocates specified
+ amount of resources for this operation. Next the engine marks argument's tiling using specified parameters and maps resulting workgroups onto allocated processes in a user specified way. 
+Now engine can freely transfer the tiled data from previous owners to the allocated processes changing the tile size if needed. After the transfer is complete the computational kernel kicks in. 
+The computational kernel behaves almost the same way as in CUDA and workgroups are simply being mapped onto threads of MPI process as if it were GPU SMP (the main difference here lies in user being in control of particular process workgroup's ownership).
+
+In further sections we are going to review mechanisms that we have used here in more details.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN DESCRIPTION OF THE FRAMEWORK
-
 \section{Design of the framework}
@@
 \subsection{Execution model and Operations scheduling}
-In a typical HPC application, the programming model imposes a parallel execution of every function of the code, step by step, where the data granularity
-of the problem  has imposed the parallelization.
-In ambient, we developed an scheduler, an  \guillemotleft engine\guillemotright \, that privileged a queue execution model  (promoted  and popularized by the CELL processor), where operations (\eg linear algebra operations \ldots)
-will be stacked up into a list  and then executing according the model of logistics/computation kernel. This model is inspired from computer architecture design;
-the queue data flow execution in a processor \cite{DENNIS-1974}, nevertheless, we are working now on a larger scale where basic instructions ($+,-,*,/$) are  replaced 
-by the pair logistics/computation kernels.
+In a typical HPC application, the programming model imposes a  parallel execution of every function of the code, with data granularity imposing the details 
+of the parallelization.
+In Ambient, we developed an scheduler, an  \guillemotleft engine\guillemotright \, that privileges a queue execution model, where operations (\eg linear algebra operations \ldots)
+are stacked up in a list first, and are executed afterwards according to the model of the logistics/computation kernel. This model is inspired by computer architecture design, i.e.
+the queue data flow execution in a processor \cite{DENNIS-1974}. We are replicating this approach now on a larger scale where basic instructions ($+,-,*,/$) are replaced 
+by the pairs of logistics/computational kernels.
 
-An HPC application can be always decomposed as a set of functions that perform numerical operations.  Start by a basic example, we want to perform the three following algebra operations :
+Any HPC application can be decomposed into a set of functions performing numerical operations.  Starting from basic examples, we want to perform the three following operations :
 
@@
 \text{Operation 1:} & & A + \!\! =  B, \label{3OPERATIONS}  \\
-\text{Operation 2:} & & A \times \!\!= 3,   \nonumber  \\ 
+\text{Operation 2:} & & A \times \!\!= \lambda,   \nonumber  \\ 
 \text{Operation 3:} & & C \times \!\!= D. \nonumber
 \end{eqnarray}
+where $A$, $B$, $C$ and $D$ are some matrices and $\lambda$ is some scalar.
 
-In the usual way, the three operations will be executed in parallel one after one.
-Transposing this example in Ambient's conception means; each operation is now  associated to a logistics and a computational kernel.
-In Practice, considering our addition example(\ref{3OPERATIONS}),  the users will use the following \texttt{push} function:
+Conventionally, these three operations would be executed in parallel one after another.
+Transposing into Ambients' conception means associating each operation with a logistics and a computational kernel.
+Considering the first example of example (\ref{3OPERATIONS}),  a user would invoke the following \texttt{push} function:
 
@@
 
-where we have the appropriated logistics and computational kernel named \texttt{Addition\_logistics} and \texttt{Addition\_computational}  presented in the table \ref{LKERNEL} and \ref{CKERNEL} previously, followed by the two arguments \texttt{\textit{A}} and \texttt{\textit{B}}.
-During the execution of the application, operations will be stack up, and in an ideal case, the content of the stack should be executed at the end of the application.
-Nevertheless, the users have the possibility to execute the content of the stack during the code,
-by the basic following instruction :
+with the appropriate logistics and computational kernels \texttt{Addition\_logistics} and \texttt{Addition\_computational}  (c.f.
+table \ref{LKERNEL} and \ref{CKERNEL}), followed by the arguments \texttt{\textit{A}} and \texttt{\textit{B}}.
+In this way operations, such as those in (\ref{3OPERATIONS}), will stack up during execution of the application.
+At any time then, the operations on the stack can be executed using the basic Ambient instruction
 
 \begin{eqnarray}
-      \texttt{playout().}
+      \texttt{playout(),}
 \end{eqnarray}
 
-This command triggered on the engine of ambient.
-The engine will scan through the stack of operation and identifies the independent operations, and also mark the different dependencies by incrementing a local dedicated counter.
-Then, it adds a  dedicated MPI packet manager\footnote{A packet manager is a tool that manages the communication between processors.} (see section \ref{MPIABSTRACTION})  for every operations and performed all logistics kernels.
-Due to the dependencies, the data will be map in a first MPI group and marked as need in others, with a trace of the previous owner.
-Finally, the independent computational kernels are executed followed by  the depending computational kernels in the appropriated order.
-In case of dependency of the data, the migration of the datas will be done by the packet manager.
-When all computational kernels are over, the stack is cleaned.
-Thus, this method combined to the execution of the computational kernels creates two layers of parallelization : the simultaneous execution of  kernels executed in parallel.
-A schematic of the stack execution is presented on Fig. \ref{EXECUTION}
+although ideally this command would be invoked at the end, when all operations have been pushed onto the stack.
+The \texttt{playout()} command triggers the Ambient engine, scanning through the operation stack, identifying the independent operations and marking dependencies.
+Next, a dedicated MPI packet manager\footnote{a tool managing the communication between processors.} (see section \ref{MPIABSTRACTION})  is added for each operation, and all logistics kernels are executed.
+Where dependencies exist, the relevant data - which is mapped  to some MPI group according to the requirements of the first operation in which it is needed - will be marked as 'needed in other operations' ; those later dependent operations can always trace it to its then-current owner.
+Finally, the computational kernels of independent operations are executed, followed by the dependent computational kernels in the appropriate order.
+In the case of data dependency, the data transfer is handled by the packet manager.
+When all computational kernels have been executed, the stack is cleared.
+A schematic of the stack execution is presented in Fig.
+\ref{EXECUTION} .
+
+Overall, the Ambient approach combines two levels of parallelism: that of each computational kernel and that of several kernels executed in parallel by assigning them to different MPI groups.
 
@@
 
-Now, we  generalized the example \ref{3OPERATIONS} to a vectors of matrices.
-We  add  conditional statements to know which matrices should add and multiply (Table \ref{GRAPH}).
+Next, we generalize example \ref{3OPERATIONS} to deal with vectors of matrices.
+We assume conditional statements determining which matrices should be added and/or multipled (Table \ref{GRAPH}).
 First, from line 1 to 11, as a dry run, the application execution will determine which operations should be performed,
-and it will pushed up needed kernels on the top of stack.
-When the \texttt{playout()} function is called line 12, the engine will execute the stack, by solving dependency, it will  create the directed  acyclic graph of the application, naturally.
-Moreover this example shows the strength of Ambient, we have integrated a double parallelization without any, hard coding of MPI instruction in the original code.
-Thus, a full parallel linear algebra  application can be developed without taking into account the geometry of the problem, which usually done.
-
+pushing the required kernels on the top of stack.
+When the \texttt{playout()} function is called line 12, the engine will execute the stack.
+By resolving dependency, it will create the directed acyclic graph of the application naturally.
 \begin{table}[t] 
@@
                                \tiny{5:}   & \hspace{0.4 cm} \textbf{end if} \\   
-                               \tiny{6:}   & \hspace{0.4 cm} $A[i]\times \!\! =3$,  $\vartriangleright$  push l/c kernel 2, \\                   
+                               \tiny{6:}   & \hspace{0.4 cm} $A[i]\times \!\! =\lambda$,  $\vartriangleright$  push l/c kernel 2, \\                   
                                \tiny{7:}   & \hspace{0.4 cm} \textbf{if}  the conditions on $C[i]$ and  $D[j]$ are verified \textbf{do} \\                               
@@
 
-To conclude this section, unlike a typical parallel execution, this model have several decisive advantages.
-It allows a transparent parallelization of two layers of parallel execution, and most important, it determined the directed  acyclic graph on the full application in order to get the best performance.
+These examples demonstrate the strength of Ambient, allowing for two levels of parallelization without any hard coding of MPI instructions in the original code, with a great degree of flexibility and adaptability.
+Thus, a fully parallel linear algebra  application can be developed without taking into account the geometry of the problem, which usually done.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN PACKET MANAGER 
 \subsection{Packet manager and MPI abstraction} \label{MPIABSTRACTION}
-In a normal HPC application, the  data could be prepared when communications are needed, if the data have more than one type especially. In this case  
-we should prepare a MPI structure and then perform the exchange. These preliminary conversions are painful for the performance. In  Ambient,
- the structure of data layout by work groups necessitates additional metadata to coordinate the good executions of logistics and 
- computational kernels. Particularly, when a remote access is performed in 
-a computational kernel. Thus,  before the block of data (literally the numbers 
-inside the matrix) of the any workgroups,  contiguous data  are included. We called this ensemble \guillemotleft block of data+metadata\guillemotright \, as a netweork packet. It is built
-on a MPI structure. Thus the work groups  are  ready to send or receive without any additional movements or copying. 
-To assure the good execution of the framework, several additional type of packets are  available, they all derived from a standard packet. 
+In standard HPC applications, data may have to be prepared for exchange over the network, particularly if the data is of more than one type; in the latter case,
+a MPI structure is set up, to which the data is then copied, and then this structure is send via network. These preliminary conversions can cut into
+performance. In Ambient, the data layout inside each work group necessitates additional metadata to coordinate the execution of logistics and 
+computational kernels, particularly the latter has to perform a remote access. Thus, before the block of data held by a workgroup, contiguous
+metadata is included. We call this ensemble \guillemotleft block of data+metadata\guillemotright \, as a network packet, which is built
+on a MPI structure. Thereby, work groups are able to send or receive without any additional movements or copying. 
+To assure proper execution of the framework, several additional type of packets are  available, all derived from a standard packet. (MORE INFO HERE. OTHERWISE LAST
+SENTENCE IS REDUNDANT)
 %The control of the packet and MPI exchange is done by a packet manager.
 
-When we execute a computational kernel, we get the data by the Ambient command  \texttt{current(\textit{A})(i,j)} (ref. \ref{CURRENT}), where \textit{A} is a matrix 
-compatible with Ambient, and the indices $i$ and $j$ are the coordinates of the work group (Fig.
-\ref{MPIGROUP}).
-Three possibilities are possibles for getting the data when a process execute this functions
-in a computational kernel.
+When a computational kernel is executed, it can access the any required, such as a block of matrix $A$, by use of the Ambient command  \texttt{current(\textit{A})(i,j)}
+(ref. \ref{CURRENT}), where \textit{A} has to be a matrix  compatible with Ambient (REFERENCE HERE TO SECTION WHERE WE EXPLAIN WHAT AMBIENT-COMPATIBLE
+MATRICES MUST LOOK LIKE), the indices $i$ and $j$ being the coordinates of the work group (Fig. \ref{MPIGROUP}). 
+There are three possibilities for the resultant types of data communication when a process executes this function inside a computational kernel:
+
 \begin{enumerate}
@@
 \item The work group is in the memory space of the present MPI-group (previously created by the associated logistics kernel). 
-The applicant process sends a query request (a specific packet) to its  master process of its MPI group.
-The master process of the group gets the request (the master process knows the repartition of the workgroup inside the MPI-group), it sends a new request to
-the actual owner, when the actual owner gets the request, it send the workgroup needed to the first applicant. 
-\item The work group is inside an other MPI-group. As previously, the applicant process sends to the master process a query  request, and the master process of 
-the actual MPI group  sends an other request to the older owner group. The older master process gets the message and sends a message to actual owner of the data. Then the owner 
-of the data sends the data to the first applicant.
+The process requiring this data sends a query (a specific type of packet) to the master process of its MPI group.
+The master process, which knows which slave process of its' group holds the required work group, sends a new request to
+the that process, which in turn sends the requested workgroup to the requesting process. 
+\item The work group is held inside another MPI-group. As previously, the process which executes the computational kernel sends a query to the master process of its MPI group,
+which in turn sends a request to the master process of the MPI group within which the data is held. That master process then sends a message to the actual owner process,
+which finally sends the data to the initially requesting process.
 \end{enumerate}
-We illustrate the two last previous situations on the figure \ref{MPIEXCHANGE}.
+Case (2) and (3) are illustrated in Fig. \ref{MPIEXCHANGE}.
 \begin{figure}[h]
@@
 	\end{center}
-          \caption{On the left, we illustrate communications inside the same group.
-On the right, we illustrate the communications between  two MPI-groups.
-The dashed circle indicated the previous MPI group.
-In both cases, the masters process have the key
-          role to distribute informations for the exchange. The requests  are arrows, and the orders of operations are marked by numbers.}  \label{MPIEXCHANGE}
+          \caption{On the left: communications inside the same group. On the right: communications between two MPI-groups. The dashed circle indicates the group holding the requested data.
+          In both cases, the master processes have a key role in distributing information for the exchange. Each arrow marks a request, the sequence of requests is enumerated.}  \label{MPIEXCHANGE}
 \end{figure}
 
-In the previous section, we reveled the importance of the communications, theirs perfect execution is relevant to  the packet manager.
-The packet manager  manages the intra communication inside the same MPI group \eg to perform an multiplication, and the inter communication
-between the MPI groups. The packet manager is unique for every MPI-group, it is based on a singleton pattern.
-
- In the previous paragraph, we presented different communication blueprint when datas are requested by a process.
-As communication path between the applicant  and the owner of the data can be more and less complex,  to solve every communication scheme,  we designed the packet manager on two main things :
+Clearly, smooth execution of any communication pattern arising from the great number of possible data dependencies
+is key to a good performance of the Ambient approach. The resolution of communication patters also falls under the purview
+of the packet manager, as it manages both intra- and inter-group communication. Each MPI-group has a unique packet manager,
+identified by a singleton pattern. For the packet manager to be adaptable to any communication pattern, it consists of two parts:
 the delegate pattern and a finite state machine (FSM).
 
-To reach the best performance in HPC,  synchronization operations should be avoided  \cite{EXASCALE-2010}, because it is  a bottleneck.
-Thus, we privileged  MPI asynchronous communication into Ambient (\texttt{mpi\_isend} and \texttt{mpi\_irecv}). Nevertheless, if a message 
-is sent by a asynchronous instruction, an MPI test is necessary to check if the packet can be received.  As we refused synchronization instructions,
-we repeat the process until receiving the message by an infinite loop. Thus we integrated the delegate pattern and the FSM inside the loop.
+To reach the best performance in HPC, bottlenecks arising from synchronization operations should be avoided  \cite{EXASCALE-2010}.
+Thus, MPI asynchronous communication is privileged in Ambient (\texttt{mpi\_isend} and \texttt{mpi\_irecv}). Still, an asynchronously
+sent message  requires a MPI test to check whether the packet can be received.  As we avoid synchronization instructions,
+the receiving process repeatedly checks for the availability of the expected data in an infinite loop, until actually receiving the
+data. Both the delegate pattern and the FSM are placed inside this loop.
 
@@
 	\end{center}
-          \caption{Schematic of the finite sate machine for the closing procedure.
-The states 1 and 2 correspond to open or close.
+          \caption{Schematic of the finite sate machine for the closing procedure. The states 1 and 2 correspond to open or close.
            The transitions are loose, closure or opening, they  depend of the situation of all packets (if they all reach theirs targets or not),
@@
 
-A delegate pattern (C\#) encapsulated methods,    It is similar to pointer of function under C++  but is allows a safer execution, we ported this pattern into Ambient.
-The packet manager class contains a delegate object. The delegate class contains a series of operations pill-up  in a stack. These operations consists of forwarding requests, 
- block of data,  layout of the MPI group (block distribution) and theirs updates. As this delegate is included, into an infinite loop, the packets
-will propagate process by process after every iterations (as shown on Fig. \ref{EXECUTION}). At the end of the delegate execution, a finite state machine checks the states of the packets. 
-If all packets received (data packet, layout packet, \ldots), the FSM (Fig. \ref{FSM}) breaks the infinite loop.
+The delegate pattern (from C\#) encapsulates a method, behaving similar to a function pointer in C++,  but allowing for safer execution.
+The delegate object inside the packet manager class contains a series of operations pil in a stack, such as forwarding requests, 
+data communication and communication of changes in MPI group layout (i.e. block distribution). As this delegate is inside the infinite loop, the packets
+will propagate process by process after every iterations (as shown on Fig. \ref{EXECUTION}). After the delegate execution is done, a finite state machine
+checks the states of the packets.  If all packets (data packet, layout packet, etc.) have been received, the FSM (Fig. \ref{FSM}) breaks the infinite loop.
 
-\subsection{Abstract parallel datatypes} % <- to change ?
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% END PACKET MANAGER
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN ABSTRACT PARALLEL DATATYPES 
+\subsection{Abstract parallel datatypes}
+
+To assure the largest scope to Ambient, we designed the framework on the data structures and not a specific problem. Thus, we enlarge the utilization of the framework
+for a various set of applied computing fields. The connection  between Ambient and the \guillemotleft outside world \guillemotright  is done by an intermediate class that assure the
+connectivity between a class and ambient.
+
+For linear algebra, we designed a matrix class, we contained usual data members for describing the object (number of raws, number of columns, \dots) and  methods for linear
+algebra operations (associated to logistics and computational kernel).This matrix class derived from an other who assured the connection to Ambient.
+When an instance of the matrix class is created, the inheritance mechanics will call constructors, methods in the mother class that  will 
+structure  the memory (tile  layout).
+
+Therefore, we can extend Ambient for new problems. Per example, in fluid mechanics the  two-dimensional finite difference method is very popular for solving the Navier-Stokes  differential equations.
+These equations describes the parameters of the \guillemotleft fluid particle \guillemotright. In this case Ambient will discretize the space and assure, as for a matrix,  the distribution
+ of the data inside the memory.
+Thus, a fluid particle class will contain enough information to assure a good discretization in the cartezian space (as the number of raws and columns inside the matrix class).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% END ABSTRACT PARALLEL DATATYPES
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN VALIDATION
 \subsection{Validation}
-As  validation test, we code a new implementation of PDGEMM using our framework, and we compare to the classical version of PDGEMM of ScaLAPACK (MKL-Version).
-The algorithm of our implementation is based on the classical method using a columns order distribution for the matrices nevertheless, we give a description of the algorithm 
-into the table \ref{AMBIENTDGEMM} to show the modification due to Ambient.
-
-The main advantages in our implementation is the size of coding, it is limited at 7 lines  of code using our framework.
+As a validation test, we code a new implementation of PDGEMM using the Ambient framework, and compare to the ScaLAPACK version of PDGEMM (MKL-Version).
+The algorithm of our implementation is based on the classical method [CITATION NEEDED] using a column order distribution for the matrices (c.f. table \ref{AMBIENTDGEMM} 
+for the modification due to Ambient). The main advantage in our implementation is the required amount of coding, which is limited to seven lines.
 
 ----------------------------------------------------------------------------------------------------- \\
-TO DO EXPLAIN THE ADVANTAGE AGAINST   SCALAPACK \\
+TO DO: EXPLAIN THE ADVANTAGE AGAINST   SCALAPACK \\
 IN TERM OF NETWORK, DATA EXCHANGE \\
@@
 
-The results will be compare with the ScaLAPACK implementation, in the section \ref{CUDAIDEOLOGYSECTION}, we announced a compatibility with the ScaLAPACK.
-The representation of the memory of ScaLAPACK and Ambient have significant differences, although matrixes are subdivided into small block, the layout into
-the memory is different. In Ambient all blocks named work group are independent even if they belong to the same process. In ScaLAPACK block will be regrouped
-by process to get a contiguous memory, the multiplication between blocks will be assure by using a correct offset inside the DEGMM call. Therefore, in ambient if we execute a solver from ScaLAPACK,
-one of the first operation to perform will be the rearrangement of the work groups of a same owner into a contiguous block. This operation add a certain overload.
+We compare the Ambient results with the ScaLAPACK implementation. As related in section \ref{CUDAIDEOLOGYSECTION}, Ambient is compatible with ScaLAPACK,
+despite the difference in memory layout. In Ambient, each work group is principally independent from any other, even if they belong to the same process, while in ScaLAPACK
+blocks are grouped by process s.t. a memory-contiguous object is obtained. Therefore, calling a ScaLAPACK solver from Ambient must entail rearrangement of appropriate
+work groups into a contiguous blocks, and thus a certain overhead. Due to Ambients flexibility in handling and grouping data, from a programming perspective this task
+is simple to accomplish however.
 
@@
 
-The validation test consists of multiplying matrices and saving the results into an output matrix, as follow :
-\begin{eqnarray}
-C = A \times B
-\end{eqnarray}
-This test will be executed with the ambient computational kernel and the Scalapack PDGEMM solver inside Ambient (with the additional workload of rearrangement of the memory)
-As we are using C++ language, the operations $\times$ and $=$ have been overloaded correctly by pushing the good logistics and computational kernels.
-When the calculation is over for both methods results are compared element by element respecting the following test  \cite{Dongarra:1990:SLB:77626.79170} :
+The validation test consists of multiplying matrices and saving the results into an output matrix, $C = A \times B$,
+where we further compare between the output of the Ambient computational kernel and the ScaLAPACK PDGEMM solver.
+Using the C++ language, the operations $\times$ and $=$ have been overloaded appropriately, now containing the "pushing" of the required logistics and computational kernels.
+Afterwards, results are compared element by element using the test  \cite{Dongarra:1990:SLB:77626.79170}  $\left \lvert C^S_{ij} - C^A_{ij} \right \rvert/\left \lvert  \epsilon C^S_{ij} \right \rvert  < n$,
+where  the precision factor $\epsilon$  is the \texttt{double} precision $10^{-16}$, indices  $S$ and $A$ represent the ScaLAPACK and the Ambient version, and $n$ is equal to 16 (decimal digits), the exponent of numerical precision.
 
-\begin{eqnarray}
-\frac{\left \lvert C^S_{ij} - C^A_{ij} \right \rvert}{\left \lvert  \epsilon C^S_{ij} \right \rvert } < n,
-\end{eqnarray}
-
-Where  the precision factor $\epsilon$  is equal to the \texttt{double} precision $10^{-16}$, the indices  $S$ and $A$ represent the ScaLAPACK and the Ambient version,
-indices $i$ and $j$ are the entries of the matrices and $n$ is equal to 16 (decimal digits), the exponent of the numerical precision.
-
-For the validation the size of work groups have been fixed to $128 \times 128$, and the matrices are squares from 256 up to 8196 (by step of 128 elements).
+For the test, the size of work groups has been fixed to $128 \times 128$, and the square matrices $A$, $B$ range from 256 up to 8196 in size (step size of 128).
 The numbers of MPI-Process varied from 2 to 8.
@@
 pseudo-random numbers non-negative, double-precision,  uniformly distributed over the interval $[0.0 , 1.0]$ using a linear congruential algorithm and 
-48-bit integer arithmetic. }  from \texttt{stdlib}.  The validation tests have always been verified successfully . The benchmarks are performed and evaluated in the next sections.
+48-bit integer arithmetic. }  from \texttt{stdlib}.  
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% END VALIDATION
 
@@
 \subsection{First benchmark case}
-The first benchmark consists of evaluating the performance of our parallel matrix multiplication. The initial condition are the same than the validation test. The benchmarks 
-were performed on a AMD-2427 2.2 Ghz ..........
+The first benchmark consists of evaluating the performance of our parallel matrix multiplication to the implementation into the ScaLAPACK-MKL. Both solvers are included as computational kernels
+in Ambient.
+The initial condition are the same than the validation test.
+As the validation test, the matrices size vary from 256 up to 8196 by step of (128 elements).
+A 1-D cyclic  column distribution is
+used for both algorithm. As both kernels are included inside the framework additional workload should take into account and for the ScaLAPACK pDGEMM, we reorganize data to fit the ScaLAPACK
+representation.
+The benchmarks  were performed on AMD Opteron  Magny-Cours) CPUs at 2.1 GHz on the Swiss National Supercomputing Centre facilities.
+The performance is reported in [GFlops], it is simply
+calculated by :
+
+\begin{eqnarray}
+P  = \frac{2n^i}{t},
+\end{eqnarray}
+where $n$ is the size of the square-matrix,  $t$ the full time of calculation and $i$ is equal to 3 for the classical matrix multiplication algorithm.
 
@@
 \include{PLOT/SinglePDGEMM/SinglePGEMM}
+\caption{Performances of PDGEMM using scalapack }
 \end{figure}
@@
 \subsection{Second benchmark case}
+This benchmarks consists to evaluate the performance of the scheduler, especially the repartition of several operation all over the cluster.
+
 \subsection{Third benchmark case}
@@
 \section{Conclusions}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN THANK 
+\thanks{The authors are grateful to W. Sawyer and G. Fourestey from the CSCS staff and V. Rezzonico from EPFL.} 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% END THANK
 

### 2011-08-11 doc/paper_ambient/paper_ambient.tex
paper modifs

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%% END PACKAGE
-\graphicspath{{PLOT/SinglePDGEMM/}}
+\graphicspath{{PLOT/SinglePDGEMM/}{PLOT/VectorPDGEMM/}}
 
@@
 \author[A. Kosenkov, T. Ewart, B. Bauerb, A. Kantian,  M. Troyer, T. Giarmarchy]{Alexandr Kosenkov, Timoth\'ee Ewart, Bela Bauer, Adrian Kantian, Matthias Troyer and Thierry Giamarchy }
-%\author[UNIGE, ETHZ, CSCS]{UNIGE, ETHZ, CSCS}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN AUTHORS AND INSTITUTIONS
@@
                                 &(gridDim, blockDim)  &  &  execution  \\ 
-                               $\diamond$ &  Thread blocks mapping onto de-  &   $\diamond$  &  Serial code   \\
+                               $\diamond$ &  Thread blocks mapping onto de-  &  &  \\
                                 & vice (GPU SMPs) &  & \\
-                               $\diamond$ &  Execution of kernel: GPU SMP  &  $\diamond$&Stack flush signal encountered :  \\
-                               &  executes kernel on every thread& &   execution of kernels  \\
+                               $\diamond$ &  Execution of kernel: GPU SMP & & \\
+                               &  executes kernel on every thread& &    \\
                                &  of corresponding thread blocks  & & \\
-                                 $\diamond$  &  Serial code &  & \\
+                               \multicolumn{4}{c}{  $\diamond$ Serial code}  \\
+                               & &    $\diamond$&Stack flush signal encountered : \\
+                               & & & execution of kernels \\
                                   \hline
@@
 
+In CUDA and OpenCL, the execution model imposes kernel initialization with the appropriate resources (\eg numthread, numblock and grid dimension),
+after which the user does not have to manage the memory access inside GPU memory; further, all threads can readily access any part of the GPU.
+Due to memory distribution in clusters we need to adapt this method of execution.
+Thus, we split this execution into two parts.
+
+First, we prepare the memory, and then execute our "numerical" operation \eg a matrix multiplication in the following sequence \footnote{From here onwards, we will illustrate the execution model through parallel matrix addition and multiplication.}:
+\begin{enumerate}
+	\item A logistic kernel; where we select the number of processors, and map the memory with the datas.
+	\item A computational kernel; where we execute the associated functions (the kernel in CUDA).
+\end{enumerate}
+
+Thus, every operations will have an associated logistics and computational kernel \eg multiplication, addition, or more sophisticated linear algebra operations
+
+\newpage
+
+\begin{figure}[!t]
+	\begin{center}
+           	        \psfrag{Ambient representation}{\tiny{\hspace{-0.4cm}Ambient representation}}
+           	        \psfrag{CUDA/OpenCL representation}{\tiny{\hspace{-0.9cm}CUDA/OpenCL representation}}
+	                 \psfrag{Associated mapping of the data with Ambient}{\tiny{\hspace{+0.15cm}Associated data mapping}}
+	                 \psfrag{1x1 workgroup}{\tiny{\hspace{-0.5cm} \vspace{-0.1cm}1x1 workgroup}}
+	                  \psfrag{of Ambient}{\tiny{\hspace{-0.4cm}\vspace{-0.9cm} of Ambient}}
+	                  \psfrag{Workgroup of 4X4 threads}{\tiny{\hspace{-0.5cm} \vspace{-0.1cm}4x4 workgroup}}
+	                  \psfrag{of CUDA}{\tiny{\hspace{-0.25cm}of CUDA}}
+	                  \psfrag{Scope of GPU thread (left)}{\tiny{\hspace{-0.25cm}GPU scope left}}
+	                  \psfrag{CPU (right)}{\tiny{\hspace{-0.7cm}CPU scope right}}       
+		        \includegraphics[scale=0.44]{FIGURES/CUDAanalogy.eps} 
+	\end{center}
+          \caption{Corresponding of the CUDA representation, to the ambient representation, in terms of the scope of the CPU with the associated data.}
+          \label{CUDAANALOGY}
+\end{figure}
 \begin{table}[h]
@@
 
-In CUDA and OpenCL, the execution model imposes kernel initialization with the appropriate resources (\eg numthread, numblock and grid dimension),
-after which the user does not have to manage the memory access inside GPU memory; further, all threads can readily access any part of the GPU.
-Due to memory distribution in clusters we need to adapt this method of execution.
-Thus, we split this execution into two parts.
-
-First, we prepare the memory, and then execute our "numerical" operation \eg a matrix multiplication in the following sequence \footnote{From here onwards, we will illustrate the execution model through parallel matrix addition and multiplication.}:
-\begin{enumerate}
-	\item A logistic kernel; where we select the number of processors, and map the memory with the datas.
-	\item A computational kernel; where we execute the associated functions (the kernel in CUDA).
-\end{enumerate}
-
-\begin{figure}[!t]
-	\begin{center}
-           	        \psfrag{Ambient representation}{\tiny{\hspace{-0.4cm}Ambient representation}}
-           	        \psfrag{CUDA/OpenCL representation}{\tiny{\hspace{-0.9cm}CUDA/OpenCL representation}}
-	                 \psfrag{Associated mapping of the data with Ambient}{\tiny{\hspace{+0.15cm}Associated data mapping}}
-	                 \psfrag{1x1 workgroup}{\tiny{\hspace{-0.5cm} \vspace{-0.1cm}1x1 workgroup}}
-	                  \psfrag{of Ambient}{\tiny{\hspace{-0.4cm}\vspace{-0.9cm} of Ambient}}
-	                  \psfrag{Workgroup of 4X4 threads}{\tiny{\hspace{-0.5cm} \vspace{-0.1cm}4x4 workgroup}}
-	                  \psfrag{of CUDA}{\tiny{\hspace{-0.25cm}of CUDA}}
-	                  \psfrag{Scope of GPU thread (left)}{\tiny{\hspace{-0.25cm}GPU scope left}}
-	                  \psfrag{CPU (right)}{\tiny{\hspace{-0.7cm}CPU scope right}}       
-		        \includegraphics[scale=0.44]{FIGURES/CUDAanalogy.eps} 
-	\end{center}
-          \caption{Corresponding of the CUDA representation, to the ambient representation, in terms of the scope of the CPU with the associated data.}
-          \label{CUDAANALOGY}
-\end{figure}
-
-Thus, every operations will have an associated logistics and computational kernel \eg multiplication, addition, or more sophisticated linear algebra operations
+\newpage
 
@@
 Switching from one distribution scheme to another, which is a challenge in ScaLAPACK, is easily accomplished with Ambient.
- \begin{table}[t]  
+ \begin{table}[h]
 	\begin{center}
@@
 \item Line 2: the cartesian coordinates of the work group are obtained from the Ambient instruction \texttt{get\_block\_id(A).x}  for $i$ and \texttt{get\_block\_id(A).y}  for $j$. \label{CURRENT}
-\item Line 3 to 5 : we refer the work group $i,j$ to a temporary variable $A_{ref}$ using the Ambient instruction \texttt{current(A)(i,j)}.
-Note that if the processor requires data outside its memory, it will perform a remote access in the background (see section ...
-for details).
+\item Line 3 to 5 : we refer the work group $i,j$ to a temporary variable $A_{ref}$ using the Ambient instruction \texttt{current(A)(i,j)}. Note that if the processor requires data outside its memory,
+ it will perform a remote access in the background (see section \ref{MPIABSTRACTION} for details).
 \item Line 6 to 8 : the addition is performed by looping over the total size of the work group, in parallel for each workgroup (because of the use of the keyword \texttt{pinned}.
@@
 
-\begin{figure}[h]
+\begin{figure}[b]
 	\begin{center}
@@
 on a MPI structure. Thereby, work groups are able to send or receive without any additional movements or copying. 
-To assure proper execution of the framework, several additional type of packets are  available, all derived from a standard packet. (MORE INFO HERE. OTHERWISE LAST
-SENTENCE IS REDUNDANT)
-%The control of the packet and MPI exchange is done by a packet manager.
+To assure proper execution of the framework, several additional type of packets are  available, request packet, layout packet and data-block packet
+they  all derived from a standard packet.
 
 When a computational kernel is executed, it can access the any required, such as a block of matrix $A$, by use of the Ambient command  \texttt{current(\textit{A})(i,j)}
-(ref. \ref{CURRENT}), where \textit{A} has to be a matrix  compatible with Ambient (REFERENCE HERE TO SECTION WHERE WE EXPLAIN WHAT AMBIENT-COMPATIBLE
-MATRICES MUST LOOK LIKE), the indices $i$ and $j$ being the coordinates of the work group (Fig. \ref{MPIGROUP}). 
+(ref.
+\ref{CURRENT}), where \textit{A} has to be a matrix  compatible with Ambient (\cite{ABSTRACTPARADATATYPE}), the indices $i$ and $j$ being the coordinates of the work group (Fig.
+\ref{MPIGROUP}).
 There are three possibilities for the resultant types of data communication when a process executes this function inside a computational kernel:
@@
 of the packet manager, as it manages both intra- and inter-group communication. Each MPI-group has a unique packet manager,
-identified by a singleton pattern. For the packet manager to be adaptable to any communication pattern, it consists of two parts:
-the delegate pattern and a finite state machine (FSM).
+identified by a singleton pattern.
 
@@
 Thus, MPI asynchronous communication is privileged in Ambient (\texttt{mpi\_isend} and \texttt{mpi\_irecv}). Still, an asynchronously
-sent message  requires a MPI test to check whether the packet can be received.  As we avoid synchronization instructions,
+sent message  requires a \texttt{mpi\_test} to check whether the packet can be received.  As we avoid synchronization instructions,
 the receiving process repeatedly checks for the availability of the expected data in an infinite loop, until actually receiving the
-data. Both the delegate pattern and the FSM are placed inside this loop.
+data.
 
-\begin{table}[t] 
+ For the packet manager to be adaptable to any communication pattern, it consists of two parts:
+the delegate pattern and a finite state machine (FSM), both are placed inside this loop.
+
+\begin{table}[b] 
 	\begin{center}
@@
 
-\begin{figure}[h]
-	\begin{center}
-	    \psfrag{1}{\tiny \hspace{-0.1 cm}  1}
-	    \psfrag{2}{\tiny \hspace{-0.1 cm}  2}
-	    \psfrag{Loose}{\tiny \hspace{-0.2 cm}   Loose}
-	    \psfrag{Closure}{\tiny \hspace{-0.2 cm}  Closure}
-	    \psfrag{Open}{\tiny \hspace{-0.4 cm}  Opening}
-             \includegraphics[scale=0.45]{FIGURES/FSM.eps} 
-	\end{center}
-          \caption{Schematic of the finite sate machine for the closing procedure. The states 1 and 2 correspond to open or close.
-           The transitions are loose, closure or opening, they  depend of the situation of all packets (if they all reach theirs targets or not),
-                          the loose transition means the packet has not been received, thus we iterate one more time on the infinite loop.}  \label{FSM}
-\end{figure}
+%\begin{figure}[h]
+%	\begin{center}
+%	    \psfrag{1}{\tiny \hspace{-0.1 cm}  1}
+%	    \psfrag{2}{\tiny \hspace{-0.1 cm}  2}
+%	    \psfrag{Loose}{\tiny \hspace{-0.2 cm}   Loose}
+%	    \psfrag{Closure}{\tiny \hspace{-0.2 cm}  Closure}
+%	    \psfrag{Open}{\tiny \hspace{-0.4 cm}  Opening}
+%          \includegraphics[scale=0.45]{FIGURES/FSM.eps} 
+%	\end{center}
+%          \caption{Schematic of the finite sate machine for the closing procedure. The states 1 and 2 correspond to open or close.
+%           The transitions are loose, closure or opening, they  depend of the situation of all packets (if they all reach theirs targets or not),
+%                          the loose transition means the packet has not been received, thus we iterate one more time on the infinite loop.}  \label{FSM}
+%\end{figure}
 
-The delegate pattern (from C\#) encapsulates a method, behaving similar to a function pointer in C++,  but allowing for safer execution.
-The delegate object inside the packet manager class contains a series of operations pil in a stack, such as forwarding requests, 
-data communication and communication of changes in MPI group layout (i.e. block distribution). As this delegate is inside the infinite loop, the packets
-will propagate process by process after every iterations (as shown on Fig. \ref{EXECUTION}). After the delegate execution is done, a finite state machine
-checks the states of the packets.  If all packets (data packet, layout packet, etc.) have been received, the FSM (Fig. \ref{FSM}) breaks the infinite loop.
+The delegate pattern (from C\#) encapsulates  methods, behaving similar to a function pointer in C++,  but allowing for safer execution.
+The delegate object inside the packet manager class contains a series of functions pile up in a stack, such as forwarding requests, 
+layout and block of data. As this delegate is inside the infinite loop, the delegate execution 
+will propagate the packets process by process after every iterations (as shown on Fig. \ref{EXECUTION}). When the delegate execution is over, a finite state machine
+checks the states of the packets (arrived or not).  If all packets (data packet, layout packet, etc.) have been received, the FSM breaks the infinite loop.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN ABSTRACT PARALLEL DATATYPES 
-\subsection{Abstract parallel datatypes}
+\subsection{Abstract parallel datatypes} \label{ABSTRACTPARADATATYPE}
 
 To assure the largest scope to Ambient, we designed the framework on the data structures and not a specific problem. Thus, we enlarge the utilization of the framework
-for a various set of applied computing fields. The connection  between Ambient and the \guillemotleft outside world \guillemotright  is done by an intermediate class that assure the
+for a various set of applied computing fields. The connection  between Ambient and the \guillemotleft outside world \guillemotright  \, is done by an intermediate class that assure the
 connectivity between a class and ambient.
@@
 As a validation test, we code a new implementation of PDGEMM using the Ambient framework, and compare to the ScaLAPACK version of PDGEMM (MKL-Version).
-The algorithm of our implementation is based on the classical method [CITATION NEEDED] using a column order distribution for the matrices (c.f. table \ref{AMBIENTDGEMM} 
+The algorithm of our implementation is based on the classical method [CITATION NEEDED] using a column order distribution for the matrices (\cf table \ref{AMBIENTDGEMM} 
 for the modification due to Ambient). The main advantage in our implementation is the required amount of coding, which is limited to seven lines.
@@
 \include{PLOT/SinglePDGEMM/SinglePGEMM}
-\caption{Performances of PDGEMM using scalapack }
+\caption{Performances of PDGEMM using ScaLAPACK }
 \end{figure}
@@
 This benchmarks consists to evaluate the performance of the scheduler, especially the repartition of several operation all over the cluster.
+
+\begin{figure}
+\include{PLOT/VectorPDGEMM/VectorPGEMM}
+\caption{Performances of a vector of PDGEMM using ScaLAPACK}
+\end{figure}
 

### 2011-08-11 doc/paper_ambient/paper_ambient.tex
merge modification Adrian into Ambient Paper

@@
 
-In this section we describe the workflow of Ambient applications execution in general terms.
-As the Ambient approach derives from CUDA - first we will look at the simplified workflow of CUDA and Ambient-application's execution:
+[REPHRASED THIS SECTION. I THINK WE SHOULD EXPAND IT A LITTLE BIT, EXPLAINING
+FOR EXAMPLE WHERE IN THE LATER SECTIONS THE VARIOUS ELEMENTS WILL BE EXPLAINED
+(LIKE I DID WITH "PINNED" AND "PLAYOUT", AND GENERALLY EXPLAIN A LITTLE MORE WHAT THE
+BROAD FEATURES AND CAPABILITIES OF AMBIENT ARE (E.G. BOOTSTRAP PARALLELIZATION OF EXISTING
+SERIAL CODE,]
+Before presenting the various elements of Ambient in more detail, we want to describe the
+workflow of execution for an Ambient-parallelized application in general terms. 
+As the Ambient approach derives from CUDA, we consider a simplified workflow of CUDA
+ and Ambient-application's execution side-by-side:
 
@@
                              \multicolumn{4}{c}{  $\diamond$ Serial code}  \\
-                            $\diamond$ & Kernel encountered                                  &  $\diamond$ & Kernel-pair (logistic/computa- \\
+                            $\diamond$ & Kernel encountered                                  &  $\diamond$ & Kernel-pair (logistics/computa- \\
                                & & & tional) encountered \\
@@
         		 \end{tabular}  
-              \caption{Workflow analogy between CUDA and Ambient-application's execution} \label{WORKFLOW}
+              \caption{Workflow analogy between the execution of a CUDA and a Ambient application} \label{WORKFLOW}
 	\end{center}
@@
 
-When the Ambient's flush signal indicates the execution of kernels. It is performed by the followings sequences:
+As soon as Ambient encounters the flush signal (called \texttt{playout()}, see section [APPROPRIATE CITATION HERE]), kernel
+execution is initiated, along the following sequence:
 
 \begin{itemize}
-\item Execution of logistic kernel: allocation of device (MPI group) according to the request. 
-\item Execution of logistic kernel: argument's workgroups (tiles) marking with workDim.
-\item Execution of logistic kernel: selected (pinned) argument's workgroups mapping onto device (MPI processes).
-\item Transfer of workgroups to the target processes.
+\item Execution of logistics kernel: allocation of device (MPI group) according to specifics of request. 
+\item Execution of logistics kernel: argument's work groups (tiles) marking with workDim. [DON'T KNOW WHAT IS MEANT HERE, CANNOT REFORMULATE]
+\item Execution of logistics kernel: selected argument's workgroups  (see keyword \texttt{pinned}, section [APPROPRIATE CITATION HERE])
+mapping onto device (MPI processes).
+\item Transfer of work groups to the target processes.
 \item Execution of computational kernel: MPI process executes kernel thread for work items of assigned work groups.
@@
 
-The main reason of the workflow difference comparing to CUDA is the scale of available resources. 
-As we can use unlimited amount of resources for a kernel's launch - the data becomes the key factor around which we evolve Ambient's design.
-In particular, in order to control the data movements and resource usage we introduce the logistic kernel and making kernel calls non-blocking.
+The main reason for the difference in workflow compared to CUDA is the scale of available resources.  As practically unlimited resources
+are available for a kernel's launch, the data becomes the key factor around which we evolve Ambient's design.
+In particular, in order to control the data movements and resource usage, we introduce the logistics kernel and make kernel calls non-blocking.
 
-In general, Ambient application operates with abstract types that already have tiled memory structure. Upon the kernel operation 
-call the Ambient framework stacks the operation for later execution and proceeds to the rest of the code.
+In general, Ambient applications operate with abstract types that already have tiled memory structure. Upon encountering a kernel operation 
+call, the Ambient framework stacks it for later execution, and proceeds in this way throughout the code. This ideology of design
+allows a high degree of code re-usage from the original serial code, while at the same time providing both flexible and automatized
+tools to manage resources and thus  tune performance.
 
-When the framework encounters the signal to flush the stack of kernels - the actual execution starts. Framework goes through accumulated stack 
-and executes independent operations at once. When executing one operation first the engine goes through the logistic kernel and allocates specified
- amount of resources for this operation. Next the engine marks argument's tiling using specified parameters and maps resulting workgroups onto allocated processes in a user specified way. 
-Now engine can freely transfer the tiled data from previous owners to the allocated processes changing the tile size if needed. After the transfer is complete the computational kernel kicks in. 
-The computational kernel behaves almost the same way as in CUDA and workgroups are simply being mapped onto threads of MPI process as if it were GPU SMP (the main difference here lies in user being in control of particular process workgroup's ownership).
+When the framework encounters the signal to flush the stack of kernels - the actual execution starts, going through the accumulated stack of operations 
+and executes the independent ones at once. When executing one operation, the engine invokes first the associated logistics kernel and allocates a specified
+amount of resources for this operation. Next, the engine marks argument's tiling using specified parameters and maps resulting work groups onto allocated
+processes in a user specified way. Now, the engine can freely transfer the tiled data from previous owners to the allocated processes changing
+the tile size if needed. After the transfer is complete the computational kernel kicks in, behaving in almost the same way as in CUDA, with work groups
+simply being mapped onto threads of MPI process as if they were GPU SMP (the main difference being that the user is in control as to which process owns which workgroup).
 
-In further sections we are going to review mechanisms that we have used here in more details.
+In the following sections, we expand on the role and functionalities of the different elements listed here.
 
@@
 Afterwards, results are compared element by element using the test  \cite{Dongarra:1990:SLB:77626.79170}  $\left \lvert C^S_{ij} - C^A_{ij} \right \rvert/\left \lvert  \epsilon C^S_{ij} \right \rvert  < n$,
-where  the precision factor $\epsilon$  is the \texttt{double} precision $10^{-16}$, indices  $S$ and $A$ represent the ScaLAPACK and the Ambient version, and $n$ is equal to 16 (decimal digits), the exponent of numerical precision.
+where  the precision factor $\epsilon$  is the \texttt{double} precision $10^{-16}$, indices  $S$ and $A$ represent the ScaLAPACK and the Ambient version, and $n=16$, the exponent of numerical precision.
 
@@
 The numbers of MPI-Process varied from 2 to 8.
-Matrices were filled up with pseudo random number generator  \texttt{drand48()}\footnote{This function returns 
-pseudo-random numbers non-negative, double-precision,  uniformly distributed over the interval $[0.0 , 1.0]$ using a linear congruential algorithm and 
+Matrices were filled upusing a pseudo random number generator  \texttt{drand48()}\footnote{This function returns 
+pseudo-random numbers non-negative, double-precision,  uniformly distributed over the interval $[0, 1]$ using a linear congruential algorithm and 
 48-bit integer arithmetic. }  from \texttt{stdlib}.  
@@
 
-\section{Results and discutions}
+\section{Results and discussions}
 \subsection{First benchmark case}
-The first benchmark consists of evaluating the performance of our parallel matrix multiplication to the implementation into the ScaLAPACK-MKL. Both solvers are included as computational kernels
+[STARTED REWRITING THIS; BUT NOT FINISHED]
+The first benchmark consists of evaluating the performance of our parallel matrix multiplication relative to the iScaLAPACK-MKL mplementation. Both solvers are included as computational kernels
 in Ambient.

### 2011-08-12 doc/paper_ambient/paper_ambient.tex
modified the workflow section

@@
 
-[REPHRASED THIS SECTION. I THINK WE SHOULD EXPAND IT A LITTLE BIT, EXPLAINING
-FOR EXAMPLE WHERE IN THE LATER SECTIONS THE VARIOUS ELEMENTS WILL BE EXPLAINED
-(LIKE I DID WITH "PINNED" AND "PLAYOUT", AND GENERALLY EXPLAIN A LITTLE MORE WHAT THE
-BROAD FEATURES AND CAPABILITIES OF AMBIENT ARE (E.G. BOOTSTRAP PARALLELIZATION OF EXISTING
-SERIAL CODE,]
-Before presenting the various elements of Ambient in more detail, we want to describe the
+Before presenting various elements of Ambient in more details, here we describe the
 workflow of execution for an Ambient-parallelized application in general terms. 
-As the Ambient approach derives from CUDA, we consider a simplified workflow of CUDA
- and Ambient-application's execution side-by-side:
+As the Ambient computational architecture derives from CUDA, we will show a simplified workflow of CUDA
+ vs Ambient enabled applications execution side-by-side:
 
@@
 
-As soon as Ambient encounters the flush signal (called \texttt{playout()}, see section [APPROPRIATE CITATION HERE]), kernel
-execution is initiated, along the following sequence:
+As soon as Ambient encounters the flush signal - kernel
+execution is initiated along the following sequence:
 
 \begin{itemize}
-\item Execution of logistics kernel: allocation of device (MPI group) according to specifics of request. 
-\item Execution of logistics kernel: argument's work groups (tiles) marking with workDim. [DON'T KNOW WHAT IS MEANT HERE, CANNOT REFORMULATE]
-\item Execution of logistics kernel: selected argument's workgroups  (see keyword \texttt{pinned}, section [APPROPRIATE CITATION HERE])
-mapping onto device (MPI processes).
+\item Execution of logistics kernel: allocation of device (MPI group) according to specified request. 
+\item Execution of logistics kernel: argument's work groups decomposure according to workDim.
+\item Execution of logistics kernel: selected argument's work groups mapping onto device (MPI processes).
 \item Transfer of work groups to the target processes.
@@
 The main reason for the difference in workflow compared to CUDA is the scale of available resources.  As practically unlimited resources
-are available for a kernel's launch, the data becomes the key factor around which we evolve Ambient's design.
+are available for a kernel's launch, the data becomes the key factor around which Ambient's design evolves.
 In particular, in order to control the data movements and resource usage, we introduce the logistics kernel and make kernel calls non-blocking.
 
-In general, Ambient applications operate with abstract types that already have tiled memory structure. Upon encountering a kernel operation 
-call, the Ambient framework stacks it for later execution, and proceeds in this way throughout the code. This ideology of design
-allows a high degree of code re-usage from the original serial code, while at the same time providing both flexible and automatized
-tools to manage resources and thus  tune performance.
+Generally, Ambient enabled application operates with abstract types that already have tiled memory structure. Upon a kernel operation 
+call, the Ambient framework stacks the operation for later execution and proceeds to the following part of the serial code. This design
+enables high degree of original serial code's reusage while providing powerfull instrumentation capabilities to manage resources and tune performance 
+of the overall application.
 
-When the framework encounters the signal to flush the stack of kernels - the actual execution starts, going through the accumulated stack of operations 
-and executes the independent ones at once. When executing one operation, the engine invokes first the associated logistics kernel and allocates a specified
+When the framework encounters the signal to flush the stack of kernels - the actual execution starts. Framework's engine goes through the accumulated stack 
+and executes all independent operations at once. Executing an operation first the engine invokes the associated logistics kernel and allocates specified
 amount of resources for this operation. Next, the engine marks argument's tiling using specified parameters and maps resulting work groups onto allocated
-processes in a user specified way. Now, the engine can freely transfer the tiled data from previous owners to the allocated processes changing
-the tile size if needed. After the transfer is complete the computational kernel kicks in, behaving in almost the same way as in CUDA, with work groups
-simply being mapped onto threads of MPI process as if they were GPU SMP (the main difference being that the user is in control as to which process owns which workgroup).
+processes in a user specified way. Now the engine can freely transfer the tiled data from previous owners to the allocated processes changing
+the tile size if needed. After the transfer is complete the computational kernel kicks in behaving almost the same way as in CUDA with work groups
+simply being mapped onto threads of MPI process as if they were GPU SMP (mainly, the difference is the user's control over processes/work groups mapping).
 
-In the following sections, we expand on the role and functionalities of the different elements listed here.
+In the following sections, we elaborate on the role and functionalities of the different elements listed here.
 

### 2011-08-12 doc/paper_ambient/paper_ambient.tex
some corrections

@@
                              \multicolumn{4}{c}{  $\diamond$ Serial code}  \\
-                            $\diamond$ & Kernel encountered                                  &  $\diamond$ & Kernel-pair (logistics/computa- \\
+                            $\diamond$ & Kernel encountered                                  &  $\diamond$ & Kernel-pair (logistic/computa- \\
                                & & & tional) encountered \\
@@
 \begin{itemize}
-\item Execution of logistics kernel: allocation of device (MPI group) according to specified request. 
-\item Execution of logistics kernel: argument's work groups decomposure according to workDim.
-\item Execution of logistics kernel: selected argument's work groups mapping onto device (MPI processes).
+\item Execution of logistic kernel: allocation of device (MPI group) according to specified request. 
+\item Execution of logistic kernel: argument's work groups decomposure according to workDim.
+\item Execution of logistic kernel: selected argument's work groups mapping onto device (MPI processes).
 \item Transfer of work groups to the target processes.
@@
 are available for a kernel's launch, the data becomes the key factor around which Ambient's design evolves.
-In particular, in order to control the data movements and resource usage, we introduce the logistics kernel and make kernel calls non-blocking.
+In particular, in order to control the data movements and resource usage, we introduce the logistic kernel and make kernel calls non-blocking.
 
@@
 When the framework encounters the signal to flush the stack of kernels - the actual execution starts. Framework's engine goes through the accumulated stack 
-and executes all independent operations at once. Executing an operation first the engine invokes the associated logistics kernel and allocates specified
+and executes all independent operations at once. Executing an operation first the engine invokes the associated logistic kernel and allocates specified
 amount of resources for this operation. Next, the engine marks argument's tiling using specified parameters and maps resulting work groups onto allocated
@@
 Ambient employs an extension on the CUDA approach to a distributed cluster.
-This is based on the observation that modern clusters employ many thousands of nodes, just like a GPU consists of thousands of small, simple cores, on each of which an runs an independent thread.
+This is based on the observation that modern clusters employ many thousands of nodes, just like a GPU consists of thousands of small, simple cores, on each of which it runs an independent thread.
 In such a comparison however, the technical specifics of the basic cluster unit in terms of processor power, memory bandwidth, network speed, \etc  \, must be considered.
 
-The first thing coming in mind when we compare a  cluster and a GPU is the number of available cores.
-Assimilate a single  CPU core  to  GPU core is reducer  because a CPU has additional capacities than a GPU core.
-Thus, given the entirely different capabilities of a CPU core over a single GPU core, the scope of data for a single CPU (on which a single Ambient "thread" is typically running) in Ambient is rather comparable to that of a CUDA block or an OpenCL work group, the typical  blocks of size is $ 128 \times 128$.
-Of course, the user has full control of the size and can adapt it to the treated problem.
+CPU core has much more additional capabilities comparing to the one of a GPU and therefore we can't simply downcast the CPU core to the GPU one.
+Thus, the scope of the data for a single CPU (on which a single Ambient's thread typically runs) in Ambient is rather comparable to the capacity of CUDA's thread block or an OpenCL work group, the typical block size is $ 128 \times 128$.
+Though, the user has full control of the size and can adapt it for the needs of target problem.
 Like in CUDA, blocks (i.e.
-groups of "threads") and a grid of blocks can then be formed in Ambient.
+groups of threads) and the grid of blocks can be formed in Ambient.
 A basic comparison between CUDA and Ambient is shown in the table \ref{ANALOGY}.
@@
 
-Thus, every operations will have an associated logistics and computational kernel \eg multiplication, addition, or more sophisticated linear algebra operations
+Thus, every operations will have an associated logistic and computational kernel \eg multiplication, addition, or more sophisticated linear algebra operations
 
@@
 			 work group-ID                           &  \texttt{blockIdx}                                 & \texttt{get\_block\_id(A)}   \\  
-			 work group size                        &  \texttt{blockDim}                               & \texttt{get\_dim\_id(A)}      \\  
+			 work group size                        &  \texttt{blockDim}                               & \texttt{get\_mem\_dim(A)}      \\  
 			 grid size                                     &  \texttt{gridDim}                                  & \texttt{get\_grid\_dim(A)}  \\ \hline
@@
 		\caption{Analogy between our framework and  the CUDA and OpenCL ideology.  \texttt{A} is the associated matrix to the  logistic and computational kernel.  
-		The access of the cartesian values are done by the classical operator C/C++ \guillemotleft.\guillemotright     \, \eg \texttt{get\_dim\_id(A).x}  	\label{ANALOGY}}
+		The access of the cartesian values are done by the classical operator C/C++ \guillemotleft.\guillemotright     \, \eg \texttt{get\_mem\_dim(A).x}  	\label{ANALOGY}}
 	\end{center}
@@
 
-\subsubsection{The logistics kernel} The role of the logistics kernel is to initialize both memory and data for execution.
+\subsubsection{The logistic kernel} The role of the logistic kernel is to initialize both memory and data for execution.
 It consists of creating a MPI-group and mapping the data to the group with the appropriate pattern (1D or 2D block-cyclic distribution for a matrix, or alternatively
@@
                            \hline 
-                           \textbf{Logistic kernel }  Algorithm of the addition logistics kernel  &    \\ \hline
+                           \textbf{Logistic kernel }  Algorithm of the addition logistic kernel  &    \\ \hline
                            \textbf{Require}                 A class Matrix  that fits with Ambient framework  &  \\
@@
 
-To illustrate, we describe the example of a simplified\footnote{without any C++ language annotations} logistics kernel of the matrix addition in table {\ref{LKERNEL}}. 
+To illustrate, we describe the example of a simplified\footnote{without any C++ language annotations} logistic kernel of the matrix addition in table {\ref{LKERNEL}}. 
 \begin{itemize}
-\item Line 1:  the signature of the logistics kernel, showing the input/ouput by the keyword\texttt{const}.
+\item Line 1:  the signature of the logistic kernel, showing the input/ouput by the keyword\texttt{const}.
 In this example there are onlz three matrices, but additional arguments like float number, integer \etc can be added.
@@
 
-\subsubsection{The computational kernel}  After the logistics kernel has been executed, we are ready for the computational kernel.
+\subsubsection{The computational kernel}  After the logistic kernel has been executed, we are ready for the computational kernel.
 The mode of execution is similar to the CUDA API with a few specificities.
@@
 (positions, dimensions, \ldots).These are summarized in table \ref{ANALOGY}.
-For illustration, the computational kernel (table \ref{CKERNEL}) of an addition of the two matrices associated with the previous example of the logistics kernel is presented in the table \ref{LKERNEL}.
+For illustration, the computational kernel (table \ref{CKERNEL}) of an addition of the two matrices associated with the previous example of the logistic kernel is presented in the table \ref{LKERNEL}.
 \begin{itemize}
-\item Line 1: the signature of the computational kernel must be the same than that of the associated logistics kernel.
+\item Line 1: the signature of the computational kernel must be the same than that of the associated logistic kernel.
 The only difference is the use of the keyword \texttt{pinned}, indicating an execution over all Ambient work groups.
@@
 %			 work group-ID                                  &  \texttt{blockIdx}                                 & \texttt{get\_block\_id(A)}   \\  
-%			 work group size                               &  \texttt{blockDim}                               & \texttt{get\_dim\_id(A)}      \\  
+%			 work group size                               &  \texttt{blockDim}                               & \texttt{get\_mem\_dim(A)}      \\  
 %			 grid size                                            &  \texttt{gridDim}                                  & \texttt{get\_grid\_dim(A)}  \\ \hline
@@
 \texttt{A} is the associated matrix to the present logistic and computational kernel.
-The access of the cartesian values are done by the classical operator C/C++ \guillemotleft.\guillemotright     \, \eg \texttt{get\_dim\_id(A).x}  }
+The access of the cartesian values are done by the classical operator C/C++ \guillemotleft.\guillemotright     \, \eg \texttt{get\_mem\_dim(A).x}  }
 %		\label{ANALOGY2}
@@
 \subsection{SQL resource management}  \label{SQLSECTION}
- In the previous section, we introduced the logistics kernels and their role in initializing resources for the execution of the computational kernels. 
+ In the previous section, we introduced the logistic kernels and their role in initializing resources for the execution of the computational kernels. 
  One of Ambients' main qualities is the ability to perform a dynamic management of the CPU resources of the cluster.
-To facilitate this, we integrated a light SQL framework \cite{SQLite} inside Ambient, which selects the desired number of processors inside the logistics kernel, in order to execute the computational kernel.
+To facilitate this, we integrated a light SQL framework \cite{SQLite} inside Ambient, which selects the desired number of processors inside the logistic kernel, in order to execute the computational kernel.
 
@@
 This allows for a simple yet flexible management  of the MPI groups.
-These requests (done by the \texttt{scope\_select} Ambient-function) are applied to the logistics/computational addition kernels of our example by the following SQL syntax: \\
+These requests (done by the \texttt{scope\_select} Ambient-function) are applied to the logistic/computational addition kernels of our example by the following SQL syntax: \\
 
@@
 In the Ambient approach, every operation is associated with
-a logistics and a computational kernel, therefore the SQL request will be called several times.
+a logistic and a computational kernel, therefore the SQL request will be called several times.
 To create new groups, processes from the existing group are selected in ascending numerical rank, until all processes of the \texttt{ambient} group
@@
 For example, consider the previous illustration for the SQL request in case of the matrix addition operation ($ A \, +\!\!\!= B$).
-If matrix $A$ has already been created and mapped on specific processors, the logistics kernel can create a MPI group that includes its previous owner group (created in the previous operation), thereby avoiding superfluous data transfer.
+If matrix $A$ has already been created and mapped on specific processors, the logistic kernel can create a MPI group that includes its previous owner group (created in the previous operation), thereby avoiding superfluous data transfer.
 The concrete SQL request would be as follows:
@@
 the queue data flow execution in a processor \cite{DENNIS-1974}. We are replicating this approach now on a larger scale where basic instructions ($+,-,*,/$) are replaced 
-by the pairs of logistics/computational kernels.
+by the pairs of logistic/computational kernels.
 
@@
 Conventionally, these three operations would be executed in parallel one after another.
-Transposing into Ambients' conception means associating each operation with a logistics and a computational kernel.
+Transposing into Ambients' conception means associating each operation with a logistic and a computational kernel.
 Considering the first example of example (\ref{3OPERATIONS}),  a user would invoke the following \texttt{push} function:
@@
 \begin{eqnarray}
-     \texttt{push(Addition\_logistics, Addition\_computional, \textit{A}, \textit{B})},
+     \texttt{push(Addition\_logistic, Addition\_computional, \textit{A}, \textit{B})},
 \end{eqnarray}
 
-with the appropriate logistics and computational kernels \texttt{Addition\_logistics} and \texttt{Addition\_computational}  (c.f.
+with the appropriate logistic and computational kernels \texttt{Addition\_logistic} and \texttt{Addition\_computational}  (c.f.
 table \ref{LKERNEL} and \ref{CKERNEL}), followed by the arguments \texttt{\textit{A}} and \texttt{\textit{B}}.
@@
 The \texttt{playout()} command triggers the Ambient engine, scanning through the operation stack, identifying the independent operations and marking dependencies.
-Next, a dedicated MPI packet manager\footnote{a tool managing the communication between processors.} (see section \ref{MPIABSTRACTION})  is added for each operation, and all logistics kernels are executed.
+Next, a dedicated MPI packet manager\footnote{a tool managing the communication between processors.} (see section \ref{MPIABSTRACTION})  is added for each operation, and all logistic kernels are executed.
 Where dependencies exist, the relevant data - which is mapped  to some MPI group according to the requirements of the first operation in which it is needed - will be marked as 'needed in other operations' ; those later dependent operations can always trace it to its then-current owner.
@@
 Due to the dependency between the kernel 1 and 2, the computational kernel of operation 2 (l/c kernel 2) will be executed later.
-It should be performed in a new MPI group but  with an appropriate SQL request (\ref{SQL2}) in logistics kernel, we can execute the computational kernel where the data are.
+It should be performed in a new MPI group but  with an appropriate SQL request (\ref{SQL2}) in logistic kernel, we can execute the computational kernel where the data are.
 }   \label{EXECUTION}
@@
                                \tiny{13:} & $\vartriangleright$ Note:  $+\!\!=$ and $\times \!\! =$ operators have wrapper functions   \\
-                                                & that push the good logistics and computational kernel. \\                               
+                                                & that push the good logistic and computational kernel. \\                               
                            \end{tabular}   &  \\ \hline
@@
 a MPI structure is set up, to which the data is then copied, and then this structure is send via network. These preliminary conversions can cut into
-performance. In Ambient, the data layout inside each work group necessitates additional metadata to coordinate the execution of logistics and 
+performance. In Ambient, the data layout inside each work group necessitates additional metadata to coordinate the execution of logistic and 
 computational kernels, particularly the latter has to perform a remote access. Thus, before the block of data held by a workgroup, contiguous
@@
 \item The work group is already in the memory of the process. The access is direct.
-\item The work group is in the memory space of the present MPI-group (previously created by the associated logistics kernel). 
+\item The work group is in the memory space of the present MPI-group (previously created by the associated logistic kernel). 
 The process requiring this data sends a query (a specific type of packet) to the master process of its MPI group.
@@
 For linear algebra, we designed a matrix class, we contained usual data members for describing the object (number of raws, number of columns, \dots) and  methods for linear
-algebra operations (associated to logistics and computational kernel).This matrix class derived from an other who assured the connection to Ambient.
+algebra operations (associated to logistic and computational kernel).This matrix class derived from an other who assured the connection to Ambient.
 When an instance of the matrix class is created, the inheritance mechanics will call constructors, methods in the mother class that  will 
@@
 where we further compare between the output of the Ambient computational kernel and the ScaLAPACK PDGEMM solver.
-Using the C++ language, the operations $\times$ and $=$ have been overloaded appropriately, now containing the "pushing" of the required logistics and computational kernels.
+Using the C++ language, the operations $\times$ and $=$ have been overloaded appropriately, now containing the "pushing" of the required logistic and computational kernels.
 Afterwards, results are compared element by element using the test  \cite{Dongarra:1990:SLB:77626.79170}  $\left \lvert C^S_{ij} - C^A_{ij} \right \rvert/\left \lvert  \epsilon C^S_{ij} \right \rvert  < n$,

### 2011-08-12 doc/paper_ambient/paper_ambient.tex
Updated version of paper

@@
 
-Before presenting various elements of Ambient in more details, here we describe the
+Before presenting various elements of Ambient in more detail, this section provides the
 workflow of execution for an Ambient-parallelized application in general terms. 
-As the Ambient computational architecture derives from CUDA, we will show a simplified workflow of CUDA
- vs Ambient enabled applications execution side-by-side:
+As the Ambient computational architecture derives from CUDA, we start with a simplified
+workflow of execution for both CUDA-enabled and Ambient enabled applications side-by-side:
 
@@
 \item Execution of logistic kernel: allocation of device (MPI group) according to specified request. 
-\item Execution of logistic kernel: argument's work groups decomposure according to workDim.
+\item Execution of logistic kernel: argument's work groups decomposition according to workDim.
 \item Execution of logistic kernel: selected argument's work groups mapping onto device (MPI processes).
@@
 
-Generally, Ambient enabled application operates with abstract types that already have tiled memory structure. Upon a kernel operation 
-call, the Ambient framework stacks the operation for later execution and proceeds to the following part of the serial code. This design
-enables high degree of original serial code's reusage while providing powerfull instrumentation capabilities to manage resources and tune performance 
-of the overall application.
+Generally, Ambient enabled applications operate with abstract types that already have tiled memory structure. Upon encountering a kernel operation 
+call, the Ambient framework stacks the operation for later execution and proceeds in this manner through the serial code. This design
+enables a high degree of reusage of the original serial code, while providing powerful instrumentation capabilities to manage resources and tune performance 
+of the application overall.
 
-When the framework encounters the signal to flush the stack of kernels - the actual execution starts. Framework's engine goes through the accumulated stack 
-and executes all independent operations at once. Executing an operation first the engine invokes the associated logistic kernel and allocates specified
-amount of resources for this operation. Next, the engine marks argument's tiling using specified parameters and maps resulting work groups onto allocated
+When the framework encounters the signal to flush the stack of kernels the actual execution starts. The framework's engine goes through the accumulated stack 
+and executes all independent operations at once. Executing an operation, the engine first invokes the associated logistic kernel and allocates a specified
+amount of resources for this operation. Next, the engine marks argument's tiling using specified parameters (???) and maps resulting work groups onto allocated
 processes in a user specified way. Now the engine can freely transfer the tiled data from previous owners to the allocated processes changing
-the tile size if needed. After the transfer is complete the computational kernel kicks in behaving almost the same way as in CUDA with work groups
-simply being mapped onto threads of MPI process as if they were GPU SMP (mainly, the difference is the user's control over processes/work groups mapping).
+the tile size if needed. After the transfer is complete the computational kernel kicks in behaving almost the same way as in CUDA, with work groups
+simply being mapped onto threads of MPI process as if they were GPU SMP (the difference lies mainly in the user's control over the mapping of processes to work groups).
 
@@
 
-CPU core has much more additional capabilities comparing to the one of a GPU and therefore we can't simply downcast the CPU core to the GPU one.
-Thus, the scope of the data for a single CPU (on which a single Ambient's thread typically runs) in Ambient is rather comparable to the capacity of CUDA's thread block or an OpenCL work group, the typical block size is $ 128 \times 128$.
-Though, the user has full control of the size and can adapt it for the needs of target problem.
-Like in CUDA, blocks (i.e.
+A CPU core has much higher capabilities than a GPU core, and thus we can't simply downcast the CPU core to the GPU one.
+Therefore, the scope of the data for a single CPU (on which a single Ambient thread typically runs) in Ambient is rather comparable to the capacity of CUDA's thread block or an OpenCL work group.
+The typical block size is $ 128 \times 128$, although, the user has full control of this size and can adapt it to the needs of target problem.
+As in CUDA, blocks (i.e.
 groups of threads) and the grid of blocks can be formed in Ambient.
@@
 \begin{enumerate}
-	\item A logistic kernel; where we select the number of processors, and map the memory with the datas.
+	\item A logistic kernel; where we select the number of processors, and map the data to memory.
 	\item A computational kernel; where we execute the associated functions (the kernel in CUDA).
@@
 
-Thus, every operations will have an associated logistic and computational kernel \eg multiplication, addition, or more sophisticated linear algebra operations
+Thus, every operation (\eg multiplication, addition, or more sophisticated linear algebra operations) will have an associated logistic and computational kernel
 
@@
 		\end{tabular}
-		\caption{Analogy between our framework and  the CUDA and OpenCL ideology.  \texttt{A} is the associated matrix to the  logistic and computational kernel.  
+		\caption{Analogy between the Ambient framework and  the CUDA and OpenCL ideology.  \texttt{A} is the associated matrix to the  logistic and computational kernel.  
 		The access of the cartesian values are done by the classical operator C/C++ \guillemotleft.\guillemotright     \, \eg \texttt{get\_mem\_dim(A).x}  	\label{ANALOGY}}
@@
 a row/column-wise block distribution for two-dimensional Finite Difference Methods in fluid dynamics).
-The creation of the group is done by an SQL-like request, and the data is mapped into this newly created group using a modular assignment
-function. Both these functionalities are described in section \ref{SQLSECTION}.
+The creation of the group is done by an SQL-like request, and the data is mapped into this newly created group using a modular assignment function.
+Both these functionalities are described in section \ref{SQLSECTION}.
 
@@
                            \textbf{Logistic kernel }  Algorithm of the addition logistic kernel  &    \\ \hline
-                           \textbf{Require}                 A class Matrix  that fits with Ambient framework  &  \\
+                           \textbf{Require}                 A class \texttt{Matrix}  that fits with the Ambient framework  &  \\
                            \begin{tabular}{c l} 
@@
 \begin{itemize}
-\item Line 1:  the signature of the logistic kernel, showing the input/ouput by the keyword\texttt{const}.
-In this example there are onlz three matrices, but additional arguments like float number, integer \etc can be added.
+\item Line 1:  the signature of the logistic kernel, showing the input/ouput by the keyword\texttt{const} (???).
+In this example, there are only two matrices, but additional arguments like float numbers, integers \etc can be added.
 \item Line 2: The SQL-like request selecting the desired number of processors for the operation and performing the required MPI calls. 
@@
 
-We have introduced a keyword (as the \texttt{const} keyword) named \texttt{pinned}.
-This keyword indicates to the framework than the kernel execution must be executed over all Ambient work groups.
+We have introduced a keyword (as the \texttt{const} keyword (???)) named \texttt{pinned}.
+This keyword indicates to the framework that the kernel must be executed over all Ambient work groups.
 Inside a kernel, we employ several Ambient API, analogous to CUDA, to obtain key information about the work groups
@@
  it will perform a remote access in the background (see section \ref{MPIABSTRACTION} for details).
-\item Line 6 to 8 : the addition is performed by looping over the total size of the work group, in parallel for each workgroup (because of the use of the keyword \texttt{pinned}.
+\item Line 6 to 8 : the addition is performed by looping over the total size of the work group, in parallel for each workgroup (because of the use of the keyword \texttt{pinned}).
 \end{itemize}
@@
 		 \end{tabular} 
-		 \caption{Algorithm of the computational kernel associated to the matrix addition.   \label{CKERNEL}}
+		 \caption{Algorithm of the computational kernel associated to matrix addition.   \label{CKERNEL}}
 	\end{center}
@@
 As just shown for the matrix addition, we can recode other linear algebra solvers with the Ambient approach.
-For the presentation of our framework, we will focus on the MPI version of BLAS named pBLAS \cite{SCALAPACK-1992}, as well as a possible alternative to the pDGEMM solver, called ...., using a recursive algorithm.
+For the presentation of our framework, we will focus on the MPI version of BLAS named pBLAS \cite{SCALAPACK-1992}, as well as a possible alternative to the pDGEMM solver, called (???), using a recursive algorithm.
 
@@
 	\begin{center}
-           	         \psfrag{Datas distributed over work group}{\tiny{\hspace{-0.5cm}Datas distributed over work group}}
+           	         \psfrag{Data distributed over work group}{\tiny{\hspace{-0.5cm}Data distributed over work group}}
 	                  \psfrag{Local MPI rank}{\tiny{Local MPI rank}}                
@@
 
-During the execution, many group will be created in practice, and data will have to be moved between groups.
+During execution, many groups will be created in practice, and data will have to be moved between groups.
 To obtain better performance and to reduce the number of data transfers between groups, there are additional possible settings for the SQL request.
@@
          & & \texttt{scope\_select("\textit{N} from ambient as addition\_group }\\ 
-         & & \texttt{	where master is 0 where breakdown contains" + get\_id(\textit{A})}, \nonumber
+         & & \texttt{	where master is 0 where breakdown contains" + get\_id(\textit{A})}), \nonumber
 \end{eqnarray} 
-The function \texttt{get\_id(\textit{A})} will get a tag  of \texttt{\textit{A}} to know is present binding.
+The function \texttt{get\_id(\textit{A})} will get a tag  of \texttt{\textit{A}} to know its present binding.
 
@@
 \subsection{Execution model and Operations scheduling}
-In a typical HPC application, the programming model imposes a  parallel execution of every function of the code, with data granularity imposing the details 
+In a typical HPC application, the programming model imposes parallel execution of every function of the code, with data granularity imposing the details 
 of the parallelization.
 In Ambient, we developed an scheduler, an \guillemotleft engine\guillemotright \, that privileges a queue execution model, where operations (\eg linear algebra operations \ldots)
-are stacked up in a list first, and are executed afterwards according to the model of the logistics/computation kernel. This model is inspired by computer architecture design, i.e.
+are stacked up in a list first, and are executed afterwards according to the model of the logistic/computation kernel. This model is inspired by computer architecture design, i.e.
 the queue data flow execution in a processor \cite{DENNIS-1974}. We are replicating this approach now on a larger scale where basic instructions ($+,-,*,/$) are replaced 
-by the pairs of logistic/computational kernels.
+by pairs of logistic/computational kernels.
 
@@
 
-Conventionally, these three operations would be executed in parallel one after another.
+Conventionally, each of these three operations would be executed in parallel, one after another.
 Transposing into Ambients' conception means associating each operation with a logistic and a computational kernel.
@@
 table \ref{LKERNEL} and \ref{CKERNEL}), followed by the arguments \texttt{\textit{A}} and \texttt{\textit{B}}.
-In this way operations, such as those in (\ref{3OPERATIONS}), will stack up during execution of the application.
+In this way, operations, such as those in (\ref{3OPERATIONS}), will stack up during execution of the application.
 At any time then, the operations on the stack can be executed using the basic Ambient instruction
@@
 \begin{eqnarray}
-      \texttt{playout(),}
+      \texttt{playout()},
 \end{eqnarray}
@@
 Next, a dedicated MPI packet manager\footnote{a tool managing the communication between processors.} (see section \ref{MPIABSTRACTION})  is added for each operation, and all logistic kernels are executed.
-Where dependencies exist, the relevant data - which is mapped  to some MPI group according to the requirements of the first operation in which it is needed - will be marked as 'needed in other operations' ; those later dependent operations can always trace it to its then-current owner.
+Where dependencies exist, the relevant data - which is mapped  to some MPI group according to the requirements of the first operation in which it is needed - will be marked as 'needed in other operations' ; those later dependent operations can always trace this data to its then-current owner.
 Finally, the computational kernels of independent operations are executed, followed by the dependent computational kernels in the appropriate order.
@@
 
-Overall, the Ambient approach combines two levels of parallelism: that of each computational kernel and that of several kernels executed in parallel by assigning them to different MPI groups.
+Overall, the Ambient approach combines two levels of parallelism: that of each computational kernel and that of several such kernels executed in parallel by assigning them to different MPI groups.
 
@@
 First, from line 1 to 11, as a dry run, the application execution will determine which operations should be performed,
-pushing the required kernels on the top of stack.
-When the \texttt{playout()} function is called line 12, the engine will execute the stack.
-By resolving dependency, it will create the directed acyclic graph of the application naturally.
+pushing the required kernels on the top of the stack.
+When the \texttt{playout()} function is called in line 12, the engine will execute the stack.
+By resolving dependencies, it will naturally create the directed acyclic graph of the application.
 \begin{table}[t] 
@@
 These examples demonstrate the strength of Ambient, allowing for two levels of parallelization without any hard coding of MPI instructions in the original code, with a great degree of flexibility and adaptability.
-Thus, a fully parallel linear algebra  application can be developed without taking into account the geometry of the problem, which usually done.
+Thus, a fully parallel linear algebra  application can be developed without taking into account the geometry of the problem, which is usually done.
 
@@
 \begin{enumerate}
-\item The work group is already in the memory of the process. The access is direct.
-\item The work group is in the memory space of the present MPI-group (previously created by the associated logistic kernel). 
+\item \textit{The work group is already in the memory of the process.} The access is direct.
+\item \textit{The work group is in the memory space of the present MPI-group (previously created by the associated logistic kernel).} 
 The process requiring this data sends a query (a specific type of packet) to the master process of its MPI group.
 The master process, which knows which slave process of its' group holds the required work group, sends a new request to
-the that process, which in turn sends the requested workgroup to the requesting process. 
-\item The work group is held inside another MPI-group. As previously, the process which executes the computational kernel sends a query to the master process of its MPI group,
+that process, which in turn sends the requested workgroup to the requesting process. 
+\item \textit{The work group is held inside another MPI-group.} As previously, the process which executes the computational kernel sends a query to the master process of its MPI group,
 which in turn sends a request to the master process of the MPI group within which the data is held. That master process then sends a message to the actual owner process,
@@
 Thus, MPI asynchronous communication is privileged in Ambient (\texttt{mpi\_isend} and \texttt{mpi\_irecv}). Still, an asynchronously
-sent message  requires a \texttt{mpi\_test} to check whether the packet can be received.  As we avoid synchronization instructions,
+sent message requires \texttt{mpi\_test} to check whether the packet can be received.  As we avoid synchronization instructions,
 the receiving process repeatedly checks for the availability of the expected data in an infinite loop, until actually receiving the
@@
                            Algorithm of the packet manager   &    \\ \hline
-                           \textbf{Require}  a MPI process require a block of data &  \\                           
+                           \textbf{Require}  a MPI process requires a block of data &  \\                           
                             \begin{tabular}{c l}                                
                                \tiny{1:}   & \textbf{for}  ;;  \textbf{do} $\vartriangleright$  infinite loop \\
-                               \tiny{2:}   & \hspace{0.2 cm}  \texttt{Delegate()}  $\vartriangleright$ check if  packet has been, \\
-                                                & and received and re-send if needed \\
+                               \tiny{2:}   & \hspace{0.2 cm}  \texttt{Delegate()}  $\vartriangleright$ check if packet has been \\
+                                                & received, and re-send if needed \\
                                \tiny{3:}   & \hspace{0.2 cm}  FSM $\vartriangleright$ break the loop, \\
-                                               & if all packets arrived at destination. \\
+                                               & if all packets have arrived at destination. \\
                                \tiny{4:}   & \textbf{end do} \\
@@
 The delegate pattern (from C\#) encapsulates  methods, behaving similar to a function pointer in C++,  but allowing for safer execution.
-The delegate object inside the packet manager class contains a series of functions pile up in a stack, such as forwarding requests, 
-layout and block of data. As this delegate is inside the infinite loop, the delegate execution 
-will propagate the packets process by process after every iterations (as shown on Fig. \ref{EXECUTION}). When the delegate execution is over, a finite state machine
+The delegate object inside the packet manager class contains a series of functions piled up in a stack, such as forwarding requests, 
+layout, and block of data (???). As this delegate is inside the infinite loop, the delegate execution 
+will propagate the packets process by process after every iteration (as shown in Fig. \ref{EXECUTION}). When the delegate execution is over, a finite state machine
 checks the states of the packets (arrived or not).  If all packets (data packet, layout packet, etc.) have been received, the FSM breaks the infinite loop.
@@
 
-To assure the largest scope to Ambient, we designed the framework on the data structures and not a specific problem. Thus, we enlarge the utilization of the framework
-for a various set of applied computing fields. The connection  between Ambient and the \guillemotleft outside world \guillemotright  \, is done by an intermediate class that assure the
-connectivity between a class and ambient.
+To assure that Ambient has the largest possible scope, we designed the framework on the data structures and not on a specific problem. Thus, we enlarge the utilization of the framework
+to encompass various applied computing fields. The connection  between Ambient and the \guillemotleft outside world \guillemotright  \, is done by an intermediate class that assures the
+connectivity between a class and Ambient.
 
-For linear algebra, we designed a matrix class, we contained usual data members for describing the object (number of raws, number of columns, \dots) and  methods for linear
-algebra operations (associated to logistic and computational kernel).This matrix class derived from an other who assured the connection to Ambient.
-When an instance of the matrix class is created, the inheritance mechanics will call constructors, methods in the mother class that  will 
+For linear algebra, we designed a matrix class which contains the usual data members for describing the object (number of rows, number of columns, \dots) and methods for linear
+algebra operations (associated to logistic and computational kernels). This matrix class derives from another one which ensures the connection to Ambient.
+When an instance of the matrix class is created, the inheritance mechanism will call constructors, methods in the mother class that  will 
 structure the memory (tile  layout).
 
-Therefore, we can extend Ambient for new problems. Per example, in fluid mechanics the  two-dimensional finite difference method is very popular for solving the Navier-Stokes  differential equations.
-These equations describes the parameters of the \guillemotleft fluid particle \guillemotright. In this case Ambient will discretize the space and assure, as for a matrix,  the distribution
+Therefore, we can extend Ambient to new problems. For example, in fluid mechanics the two-dimensional finite difference method is very popular for solving the Navier-Stokes  differential equations.
+These equations describe the parameters of the \guillemotleft fluid particle \guillemotright. In this case Ambient will discretize the space and assure, as for a matrix,  the distribution
  of the data inside the memory.
-Thus, a fluid particle class will contain enough information to assure a good discretization in the cartezian space (as the number of raws and columns inside the matrix class).
+Thus, a fluid particle class will contain enough information to ensure a good discretization in the cartesian space (as the number of rows and columns inside the matrix class) (???).
 

### 2011-08-12 doc/paper_ambient/paper_ambient.tex
live session corrections

@@
 and executes all independent operations at once. Executing an operation, the engine first invokes the associated logistic kernel and allocates a specified
-amount of resources for this operation. Next, the engine marks argument's tiling using specified parameters (???) and maps resulting work groups onto allocated
-processes in a user specified way. Now the engine can freely transfer the tiled data from previous owners to the allocated processes changing
+amount of resources for this operation. Next, the engine sets the argument's (here: the matrices) tiling using user specified work group sizes and maps resulting work groups onto allocated
+processes. Now the engine can freely transfer the tiled data from previous owners to the allocated processes changing
 the tile size if needed. After the transfer is complete the computational kernel kicks in behaving almost the same way as in CUDA, with work groups
@@
 The work group approach imposes the adaptation of the data layout to the framework.
-In linear algebra,  BLAS \cite{LAWSON-1979} is \textit{de facto} the standard for high performance dense Linear Algebra computations.
-It imposes column order on matrix storage, as well as a contiguous storage in memory (figure \ref{MEMMATRIX}).
+The standard math libraries impose both column order as well as a contiguous memory order on matrix storage (figure \ref{MEMMATRIX}).
 
-For the sake of backward compatibility with the standard BLAS-based mathematical libraries (BLAS, LaPACK ScaLAPACK), we keep, the column order, but do not conserve a full contiguous buffer.
-As in the plasma library \cite{LTAIEF-2011}, we select
+Therefore, we keep the column order but do not conserve a full contiguous buffer. As in the PLASMA library \cite{LTAIEF-2011}, we choose
 a tile data layout.
@@
                            \begin{tabular}{c l} 
-                           \tiny{1:} &  Addition\_logistic(Matrix $A$,  Matrix $B$)  \\
+                           \tiny{1:} &  Addition\_logistic(Matrix $A$,  const Matrix $B$)  \\
                            \tiny{2:} &  SQL request for selecting $N$ number of processors   \\
@@
 \begin{itemize}
-\item Line 1:  the signature of the logistic kernel, showing the input/ouput by the keyword\texttt{const} (???).
-In this example, there are only two matrices, but additional arguments like float numbers, integers \etc can be added.
+\item Line 1:  the signature of the logistic kernel.
 \item Line 2: The SQL-like request selecting the desired number of processors for the operation and performing the required MPI calls. 
@@
 
-We have introduced a keyword (as the \texttt{const} keyword (???)) named \texttt{pinned}.
-This keyword indicates to the framework that the kernel must be executed over all Ambient work groups.
+We have introduced a keyword (like the \texttt{const} keyword) named \texttt{pinned}.
+This keyword indicates to the framework that the kernel must be executed over all work groups of the marked argument.
 Inside a kernel, we employ several Ambient API, analogous to CUDA, to obtain key information about the work groups
@@
 As just shown for the matrix addition, we can recode other linear algebra solvers with the Ambient approach.
-For the presentation of our framework, we will focus on the MPI version of BLAS named pBLAS \cite{SCALAPACK-1992}, as well as a possible alternative to the pDGEMM solver, called (???), using a recursive algorithm.
+For the presentation of our framework, we will focus on the MPI version of BLAS named pBLAS \cite{SCALAPACK-1992}, as well as recursive Strassen-like recursive reduced multiplication algorithms.
 
@@
 
-For the packet manager to be adaptable to any communication pattern, it consists of two parts:
-the delegate pattern and a finite state machine (FSM), both are placed inside this loop.
+For the packet manager to be adaptable to any communication pattern, it employs two parts:
+a delegate mechanism and a finite state machine (FSM).
 
@@
 
-The delegate pattern (from C\#) encapsulates  methods, behaving similar to a function pointer in C++,  but allowing for safer execution.
-The delegate object inside the packet manager class contains a series of functions piled up in a stack, such as forwarding requests, 
-layout, and block of data (???). As this delegate is inside the infinite loop, the delegate execution 
-will propagate the packets process by process after every iteration (as shown in Fig. \ref{EXECUTION}). When the delegate execution is over, a finite state machine
-checks the states of the packets (arrived or not).  If all packets (data packet, layout packet, etc.) have been received, the FSM breaks the infinite loop.
+The delegate mechanism serves its design goal - to execute handlers of components listening for the packets arrival, such as layout tables.
+This delegate object inside the packet manager is tied to a list of functions, such as layout forwarding, 
+data block forwarding and control information handlers. As the packet manager checks for completion of the local send process, the finite state machine
+checks the send completion state of the other MPI processes (whether they have sent all their packets or not). If all the packets have been sent, the FSM breaks the infinite loop.
 

### 2011-08-15 doc/paper_ambient/paper_ambient.tex
correcting...

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN AUTHORS AND INSTITUTIONS
-\title{AMBIENT; an  execution and an MPI-CUDA scheduler  }
+\title{Ambient: overview of the workflow and design basics}
 \author[A. Kosenkov, T. Ewart, B. Bauerb, A. Kantian,  M. Troyer, T. Giarmarchy]{Alexandr Kosenkov, Timoth\'ee Ewart, Bela Bauer, Adrian Kantian, Matthias Troyer and Thierry Giamarchy }
@@
 
-Ambient employs an extension on the CUDA approach to a distributed cluster.
+Ambient employs an extension of the CUDA approach onto a distributed cluster.
 This is based on the observation that modern clusters employ many thousands of nodes, just like a GPU consists of thousands of small, simple cores, on each of which it runs an independent thread.
@@
 Therefore, the scope of the data for a single CPU (on which a single Ambient thread typically runs) in Ambient is rather comparable to the capacity of CUDA's thread block or an OpenCL work group.
-The typical block size is $ 128 \times 128$, although, the user has full control of this size and can adapt it to the needs of target problem.
-As in CUDA, blocks (i.e.
-groups of threads) and the grid of blocks can be formed in Ambient.
+The typical block size is $ 128 \times 128$, although, the user has full control of this size and can adapt it to the needs of the target problem.
+As in CUDA, work groups and corresponding grid of groups is formed in Ambient.
 A basic comparison between CUDA and Ambient is shown in the table \ref{ANALOGY}.
@@
 	\end{center}
-          \caption{Representation of the data layout inside the memory for linear algebra. We move from a single contiguous column data layout (left ) to several tile data layouts (right).}
+          \caption{Representation of the data layout inside the memory. Instead of the single contiguos block's layout we use tiled layout.}
           \label{MEMMATRIX}
@@
 
-The work group approach imposes the adaptation of the data layout to the framework.
-The standard math libraries impose both column order as well as a contiguous memory order on matrix storage (figure \ref{MEMMATRIX}).
+The work group approach imposes the data layout of the framework.
+The standard math libraries impose both column order as well as a contiguous memory order on a matrix storage (figure \ref{MEMMATRIX}).
 
@@
 a tile data layout.
-It confers the ability to fit perfectly with the work group representation of Ambient, and further allows an excellent MPI/Thread control (section \ref{MPIABSTRACTION}).
-One of the main features of Ambient is to map this new data layout  to our work group approach, as shown in Fig.
+It allows to avoid penalties due to NUMA nature of memory allocation on the CPU (section \ref{MPIABSTRACTION}) and, naturally, this data layout supports work group representation of execution, as shown in Fig.
 \ref{CUDAANALOGY}.
 
-In CUDA and OpenCL, the execution model imposes kernel initialization with the appropriate resources (\eg numthread, numblock and grid dimension),
-after which the user does not have to manage the memory access inside GPU memory; further, all threads can readily access any part of the GPU.
-Due to memory distribution in clusters we need to adapt this method of execution.
-Thus, we split this execution into two parts.
+In CUDA and OpenCL, the execution model imposes kernel initialization with the appropriate resources (\eg block and grid dimensions),
+after which the user does not have to explicitely manage the memory transfer mechanism throughout the GPU; due to distributive memory of clusters and frequently heterogeneous nature of processing units we need both fine-grained control of work group per process assignement and memory transfer operations.
+That should be done in a simple, automatic yet flexible way.
+In order to achieve that we split the kernel invocation into two parts.
 
-First, we prepare the memory, and then execute our "numerical" operation \eg a matrix multiplication in the following sequence \footnote{From here onwards, we will illustrate the execution model through parallel matrix addition and multiplication.}:
+First, we prepare the memory of corresponding work groups.
+Then we execute our numerical operations \eg a matrix multiplication in the following sequence \footnote{From here onwards, we will illustrate the execution model through parallel matrix addition or multiplication.}:
 \begin{enumerate}
-	\item A logistic kernel; where we select the number of processors, and map the data to memory.
-	\item A computational kernel; where we execute the associated functions (the kernel in CUDA).
+	\item A logistic kernel; where we select the number of processors, and map the data onto processes indices.
+	\item A computational kernel; where we execute required computational operations (similar to the CUDA-kernel).
 \end{enumerate}
 
-Thus, every operation (\eg multiplication, addition, or more sophisticated linear algebra operations) will have an associated logistic and computational kernel
+Thus, every operation (\eg multiplication, addition, or more sophisticated linear algebra operations) will have an associated logistic and computational kernel.
 
@@
 	\end{center}
-          \caption{Corresponding of the CUDA representation, to the ambient representation, in terms of the scope of the CPU with the associated data.}
+          \caption{Similarity between CUDA's thread grid and Ambient's work group grid (with tiles mapped to work groups).}
           \label{CUDAANALOGY}
@@
 	 	\begin{tabular}{c  c  c}
-		 	                                                      & CUDA-OpenCL                                & Ambient            \\  \hline 
-			 Device                                        & Graphic card/Processor (socket)   & Cluster               \\  
+		 	                                                      & CUDA/OpenCL                                & Ambient            \\  \hline 
+			 Device                                        & GPU or CPU (socket)   & Cluster               \\  
 			  Entity  of Calculation               & GPU-Thread                                      & CPU-Thread     \\  
@@
 
-\subsubsection{The logistic kernel} The role of the logistic kernel is to initialize both memory and data for execution.
-It consists of creating a MPI-group and mapping the data to the group with the appropriate pattern (1D or 2D block-cyclic distribution for a matrix, or alternatively
-a row/column-wise block distribution for two-dimensional Finite Difference Methods in fluid dynamics).
-The creation of the group is done by an SQL-like request, and the data is mapped into this newly created group using a modular assignment function.
+\subsubsection{The logistic kernel} The role of the logistic kernel is to initialize memory and work groups for execution.
+It consists of creating an MPI-group and mapping the data to the group with the appropriate pattern (for example 1D/2D block-cyclic distribution or the row/column-wise block distribution).
+The selection of the group is done by issuing an SQL-like request.
+Afterward the data is mapped onto this newly created group using sporadic assignments.
 Both these functionalities are described in section \ref{SQLSECTION}.
 
-In linear algebra, the optimal distribution in terms of load balancing is the 2D block-cyclic distributions, esp.
-in ScaLAPACK, but this will generally depend of the solver used.
-For example, in a QR factorization \cite{DEMMEL-2008} it can be demonstrated that a row-wise distribution is superior to a 2D block-cyclic decomposition for matrices with more rows than columns.
-Switching from one distribution scheme to another, which is a challenge in ScaLAPACK, is easily accomplished with Ambient.
+In linear algebra, the optimal distribution in terms of load balancing is the 2D block-cyclic distributions but in general it depends on the target algorithm.
+For example, for the QR factorization \cite{DEMMEL-2008} it can be demonstrated that the row-wise distribution is superior to the 2D block-cyclic decomposition for matrices with more rows than columns.
+Switching from one distribution scheme to another is a challenge for ScaLAPACK while can be easily accomplished with Ambient.
  \begin{table}[h]
@@
                            \textbf{Logistic kernel }  Algorithm of the addition logistic kernel  &    \\ \hline
-                           \textbf{Require}                 A class \texttt{Matrix}  that fits with the Ambient framework  &  \\
+                           \textbf{Require}                 A class \texttt{Matrix} that adheres the Ambient architecture  &  \\
                            \begin{tabular}{c l} 
                            \tiny{1:} &  Addition\_logistic(Matrix $A$,  const Matrix $B$)  \\
-                           \tiny{2:} &  SQL request for selecting $N$ number of processors   \\
+                           \tiny{2:} &  SQL request for $N$ processors selection   \\
                            \tiny{3:} &  Perform 2/1-D cyclic distribution of the Matrix $A$  $\vartriangleright$ Note: the distribution \\
@@
 
-To illustrate, we describe the example of a simplified\footnote{without any C++ language annotations} logistic kernel of the matrix addition in table {\ref{LKERNEL}}. 
-\begin{itemize}
-\item Line 1:  the signature of the logistic kernel.
-\item Line 2: The SQL-like request selecting the desired number of processors for the operation and performing the required MPI calls. 
-\item Line 3-5: Performing a block-cyclic distribution of the matrixes \texttt{A} and  \texttt{B}. A detailed description of the SQL requests and the distribution functions
-is given in section \ref{SQLSECTION}. Note that the present kernel can be reused for any subsequent operation that uses the same data distribution.
-\end{itemize}
-
 \subsubsection{The computational kernel}  After the logistic kernel has been executed, we are ready for the computational kernel.
@@
 This keyword indicates to the framework that the kernel must be executed over all work groups of the marked argument.
-Inside a kernel, we employ several Ambient API, analogous to CUDA, to obtain key information about the work groups
-(positions, dimensions, \ldots).These are summarized in table \ref{ANALOGY}.
-For illustration, the computational kernel (table \ref{CKERNEL}) of an addition of the two matrices associated with the previous example of the logistic kernel is presented in the table \ref{LKERNEL}.
+Inside the kernel we use Ambient API, analogous to CUDA, in order to obtain necessary information about the work groups (position, dimension and so on).
+Those are partly lister in the table \ref{ANALOGY}.
+For the illustration purpose, the computational kernel (table \ref{CKERNEL}) of an addition of the two matrices associated with previous example of the logistic kernel is presented in the table \ref{LKERNEL}.
 \begin{itemize}
-\item Line 1: the signature of the computational kernel must be the same than that of the associated logistic kernel.
-The only difference is the use of the keyword \texttt{pinned}, indicating an execution over all Ambient work groups.
+\item Line 1: the signature of the computational kernel must be the same as of the associated logistic kernel.
+Here we will also include the use of\texttt{pinned} keyword indicating that the execution should be performed over marked argument's work groups.
 \item Line 2: the cartesian coordinates of the work group are obtained from the Ambient instruction \texttt{get\_block\_id(A).x}  for $i$ and \texttt{get\_block\_id(A).y}  for $j$. \label{CURRENT}
-\item Line 3 to 5 : we refer the work group $i,j$ to a temporary variable $A_{ref}$ using the Ambient instruction \texttt{current(A)(i,j)}. Note that if the processor requires data outside its memory,
- it will perform a remote access in the background (see section \ref{MPIABSTRACTION} for details).
-\item Line 6 to 8 : the addition is performed by looping over the total size of the work group, in parallel for each workgroup (because of the use of the keyword \texttt{pinned}).
+\item Line 3 to 5 : we refer to the work group $i,j$ via temporary variable $A_{ref}$ using the Ambient instruction \texttt{current(A)(i,j)}.
+Note that if the process requires data outside of its scope - the remote access will be performed in the background (see section \ref{MPIABSTRACTION} for details).
+\item Line 6 to 8 : the addition is performed by looping over all elements inside the work group.
 \end{itemize}
@@
 
-As just shown for the matrix addition, we can recode other linear algebra solvers with the Ambient approach.
+As it was shown for the matrix addition, we can recode other linear algebra solvers using Ambient's API.
 For the presentation of our framework, we will focus on the MPI version of BLAS named pBLAS \cite{SCALAPACK-1992}, as well as recursive Strassen-like recursive reduced multiplication algorithms.

### 2011-08-15 doc/paper_ambient/paper_ambient.tex
some more corrections

@@
 \subsection{SQL resource management}  \label{SQLSECTION}
- In the previous section, we introduced the logistic kernels and their role in initializing resources for the execution of the computational kernels. 
- One of Ambients' main qualities is the ability to perform a dynamic management of the CPU resources of the cluster.
-To facilitate this, we integrated a light SQL framework \cite{SQLite} inside Ambient, which selects the desired number of processors inside the logistic kernel, in order to execute the computational kernel.
+ In the previous section, we have introduced the logistic kernel and its role in resources initialization for the execution of the computational kernels. 
+ One of the main Ambient's qualities is the ability to perform dynamic management of the CPU resources on a cluster.
+To facilitate this, we have integrated an SQL parser from \cite{SQLite} into Ambient.
+Using this parser we select the desired number of processors according to supplied SQL string inside the logistic kernel.
 
-At the beginning of the framework execution, a main MPI group is created named \texttt{ambient}.
-To create groups inside this main group we used the abilities of MPI to build nested independent groups of processors  (chapter 6 of  \cite{MPI-2002}).
-We encapsulated these features in the framework, and we control them through the SQL requests.
-This allows for a simple yet flexible management  of the MPI groups.
-These requests (done by the \texttt{scope\_select} Ambient-function) are applied to the logistic/computational addition kernels of our example by the following SQL syntax: \\
+At the beginning of the application's execution global MPI group gets created under the name \texttt{ambient}.
+In order to create nested groups we have used the ability of MPI to construct a new MPI group on top of the parent one (chapter 6 of \cite{MPI-2002}).
+We encapsulated this into the framework controlling it via the mentioned SQL request mechanism.
+It enables simple yet flexible control over the resources.
+An example request for the addition kernel can be described as below (done by the \texttt{scope\_select} function).\\
 
@@
 \end{eqnarray}
-This command selects \texttt{\textit{N}} processors from the original MPI group  \texttt{ambient}. The newly created group is 
-named \texttt{addition\_group}, and its master processor locally has MPI-rank 0 (see Fig. \ref{MPIGROUP}).
+This query will select \texttt{\textit{N}} processors from the global \texttt{ambient} MPI group. The newly created group then is 
+accessible under the name \texttt{addition\_group}, its master process will have local rank 0 (see Fig. \ref{MPIGROUP}).
 
@@
 	\end{center}
-          \caption[]{Creation of a MPI group by the SQL request, with \hbox{\cloud \hspace{0.02cm}} representing a MPI group with its own communicator;  $\bullet$ represents the master 
-          process inside the group whereas the $\circ$ are the associated slave processors inside the corresponding group.
-Also shown is the assignment of the data blocks (each block corresponding to an Ambient work group) inside the newly created MPI group.
-In this example, the matrix is mapped, block by block, on the processors using a 1D column-cyclic distribution.}
+          \caption[]{Creation of an MPI group using SQL request, with \hbox{\cloud \hspace{0.02cm}} representing an MPI group with its own communicator;  $\bullet$ represents the master 
+          process inside the group whereas the $\circ$ are the associated slave processes inside the corresponding group.
+In this example the matrix is mapped block by block onto the processors using 1D column-cyclic distribution.}
           \label{MPIGROUP}
@@
 
-After the SQL request, the data (workgroup) is mapped to the processes inside the newly created MPI group by a distribution function (\eg 1D or 2D block-cyclic distribution), the algorithm of which is described in table \ref{2DCYCLIC} (\cf Fig.
+After SQL request section, the data is being mapped to the processes of the newly created MPI group in terms of work groups.
+Frequently the algorithm would be simply 1D or 2D block-cyclic distribution, the algorithm of which is described in the table \ref{2DCYCLIC} (\cf Fig.
 \ref{MPIGROUP}).
-After this transfer, the processes of the MPI group are the primary owners in terms of data access and execution.
-In the Ambient approach, every operation is associated with
-a logistic and a computational kernel, therefore the SQL request will be called several times.
-To create new groups, processes from the existing group are selected in ascending numerical rank, until all processes of the \texttt{ambient} group
-have been selected. If further groups are required, the selection is restarted from the beginning, thus a processor can belong to multiple groups simultaneously.
+After the actual transfer later those processes will become the primary owners of their data in terms of storage and handling.
+With Ambient every operation is associated with a pair of logistic and computational kernels - that is, there will be multiple SQL requests.
+In order to create new groups, processes from the parent group are being selected in a queue fashion until all processes of the group have been selected.
+If more groups are required - the selection restarts from the beginning of the queue making one process simultaneously a member of several groups.
 
-Every new group will have its own intra-communicator to allow intra-group communication, as well as an inter-communicator to allow for communication between groups, using the master process of the \texttt{ambient} group.
-
-The construction of nested MPI-groups is related to the ability of MPI to create virtual and cartesian topologies  (chapter 7 of  \cite{MPI-2002}).
-However, Ambient allows for the simultaneous construction of the MPI topology and the mapping of the  data (work group).
+Every new group will have its own intra-communicator (allowing intra-group communication) as well as an inter-communicator (allowing communication between groups).
 
@@
                                \tiny{4:} &  \hspace{0.2 cm}  \textbf{for} from $j$  to the number of the work group on $y$ \textbf{do} \\
-                               \tiny{5:} & \hspace{0.4 cm}    Assign the work group $(i,j)$ to the corresponding processor\\
+                               \tiny{5:} & \hspace{0.4 cm}    Assign the work group $(i,j)$ to the corresponding process\\
                                \tiny{6:} &   \hspace{0.2 cm}    \textbf{end for} \\
@@
 For example, consider the previous illustration for the SQL request in case of the matrix addition operation ($ A \, +\!\!\!= B$).
-If matrix $A$ has already been created and mapped on specific processors, the logistic kernel can create a MPI group that includes its previous owner group (created in the previous operation), thereby avoiding superfluous data transfer.
+If matrix $A$ has already been created and mapped on specific processors, the logistic kernel can create an MPI group as a subset or superset of the previous owner group (created in the previous operation), thereby avoiding superfluous data transfer.
 The concrete SQL request would be as follows:
@@
 \end{eqnarray} 
-The function \texttt{get\_id(\textit{A})} will get a tag  of \texttt{\textit{A}} to know its present binding.
+The function \texttt{get\_id(\textit{A})} will get a tag  of \texttt{\textit{A}} to get its current binding.
 
@@
 In a typical HPC application, the programming model imposes parallel execution of every function of the code, with data granularity imposing the details 
-of the parallelization.
-In Ambient, we developed an scheduler, an \guillemotleft engine\guillemotright \, that privileges a queue execution model, where operations (\eg linear algebra operations \ldots)
-are stacked up in a list first, and are executed afterwards according to the model of the logistic/computation kernel. This model is inspired by computer architecture design, i.e.
-the queue data flow execution in a processor \cite{DENNIS-1974}. We are replicating this approach now on a larger scale where basic instructions ($+,-,*,/$) are replaced 
-by pairs of logistic/computational kernels.
+of the parallelization. In Ambient, we have made a scheduler that privileges a queue execution model where operations
+are stacked up in a list first and are executed afterwards according to the respective logistic/computational kernel. This model is inspired by computer architecture design, i.e.
+the queue data flow execution in a processor \cite{DENNIS-1974} (Alex: hell, no!).
 
-Any HPC application can be decomposed into a set of functions performing numerical operations.  Starting from basic examples, we want to perform the three following operations :
+Any HPC application can be decomposed into a set of functions performing numerical operations. Let's start from basic example: we want to perform three following operations :
 
@@
 Conventionally, each of these three operations would be executed in parallel, one after another.
-Transposing into Ambients' conception means associating each operation with a logistic and a computational kernel.
-Considering the first example of example (\ref{3OPERATIONS}),  a user would invoke the following \texttt{push} function:
+Transposing into Ambient's conception means associating each operation with a logistic and a computational kernel.
+Considering the example (\ref{3OPERATIONS}), one would invoke the following \texttt{push} function:
 
@@
 table \ref{LKERNEL} and \ref{CKERNEL}), followed by the arguments \texttt{\textit{A}} and \texttt{\textit{B}}.
-In this way, operations, such as those in (\ref{3OPERATIONS}), will stack up during execution of the application.
-At any time then, the operations on the stack can be executed using the basic Ambient instruction
+This way, operations, such as those in (\ref{3OPERATIONS}), will stack up during execution of the application.
+At any time then, the operations on the stack would be available for the execution through the basic Ambient instruction
 
@@
 although ideally this command would be invoked at the end, when all operations have been pushed onto the stack.
-The \texttt{playout()} command triggers the Ambient engine, scanning through the operation stack, identifying the independent operations and marking dependencies.
+The \texttt{playout()} command triggers the Ambient engine to scan through the operation stack, identify the independent operations and mark dependencies.
 Next, a dedicated MPI packet manager\footnote{a tool managing the communication between processors.} (see section \ref{MPIABSTRACTION})  is added for each operation, and all logistic kernels are executed.

### 2011-08-15 doc/paper_ambient/paper_ambient.tex
some more corrections

@@
                            \textbf{Computatiomal kernel }  Algorithm of the addition computational kernel  &    \\ \hline
-                           \textbf{Require}                 A class Matrix  that fits with Ambient framework  &  \\
+                           \textbf{Require}                 A class Matrix  that adheres the Ambient architecture  &  \\
                            \begin{tabular}{c l} 
@@
 The \texttt{playout()} command triggers the Ambient engine to scan through the operation stack, identify the independent operations and mark dependencies.
-Next, a dedicated MPI packet manager\footnote{a tool managing the communication between processors.} (see section \ref{MPIABSTRACTION})  is added for each operation, and all logistic kernels are executed.
-Where dependencies exist, the relevant data - which is mapped  to some MPI group according to the requirements of the first operation in which it is needed - will be marked as 'needed in other operations' ; those later dependent operations can always trace this data to its then-current owner.
-Finally, the computational kernels of independent operations are executed, followed by the dependent computational kernels in the appropriate order.
+Next, a dedicated MPI packet manager\footnote{a tool managing the communication between processors.} (see section \ref{MPIABSTRACTION})  is added for each operation, and all logistic kernels for independent operations are executed.
+Next the computational kernels of these independent operations are executed.
+Following the completion of those, Ambient takes next chunk of independent operations (whose dependencies were just satisfied) and repeats the cycle transporting the necessary data to the new groups after the execution of logistic kernels is complete.
 In the case of data dependency, the data transfer is handled by the packet manager.
-When all computational kernels have been executed, the stack is cleared.
-A schematic of the stack execution is presented in Fig.
+When all computational kernels have been executed the stack is getting cleaned.
+The schematic overview of the stack execution is presented in Fig.
 \ref{EXECUTION} .
 
-Overall, the Ambient approach combines two levels of parallelism: that of each computational kernel and that of several such kernels executed in parallel by assigning them to different MPI groups.
+In overall, the Ambient employs single kernel multiple data parallelism model that is used over many process groups with different kernels and data.
 
@@
                            \psfrag{l/c kernel 3}{\tiny{l/c kernel 3}}
-                           \psfrag{Abiemt}{\tiny{Ambient}}
-                           \psfrag{schedule}{\tiny{Schedule}}
+                           \psfrag{Ambient}{\tiny{Ambient}}
+                           \psfrag{Schedule}{\tiny{Schedule}}
                            \psfrag{Dependency}{\tiny{Dependency}}
                            \psfrag{Time}{\tiny{Time}}
-                           \psfrag{evolution}{\tiny{evolution}}
+                           \psfrag{Evolution}{\tiny{evolution}}
                            \psfrag{Mapping and execution}{\tiny{Mapping and execution}}                          
@@
 	\end{center}
-          \caption{Model of execution of the ambient scheduler.
-On the classical approach (left) the operations will be executed in parallel step by step whatever the dependencies (marked by the symbol \guillemotleft [\guillemotright \, in the schematic).
-In ambient (right), operations will be associated to the pair logistics/computation (abbreviate l/c) kernel, as operations 1 and 3  (l/c kernels 1 and 3) are independents, the execution of computational kernel will be done, simultaneously.
+          \caption{Ambient scheduler's workflow model.
+On the classical approach (left) the operations will be executed in parallel step by step disregarding the dependencies (marked by the symbol \guillemotleft [\guillemotright \, in the schematic).
+In Ambient (right), operations will be associated to the logistic/computational kernel pair (abbreviate l/c), as operations 1 and 3  (l/c kernels 1 and 3) are independent - the execution of computational kernel will be performed simultaneously.
 Due to the dependency between the kernel 1 and 2, the computational kernel of operation 2 (l/c kernel 2) will be executed later.
-It should be performed in a new MPI group but  with an appropriate SQL request (\ref{SQL2}) in logistic kernel, we can execute the computational kernel where the data are.
+It should be performed in a new MPI group but with an appropriate SQL request (\ref{SQL2}) in logistic kernel we can execute the computational kernel right in the place where the data currently is.
 }   \label{EXECUTION}
@@
 When the \texttt{playout()} function is called in line 12, the engine will execute the stack.
-By resolving dependencies, it will naturally create the directed acyclic graph of the application.
+By resolving dependencies it will naturally create the directed acyclic graph of the application.
 \begin{table}[t] 
@@
                            Determination of the directed  acyclic graph   &    \\ \hline
-                           \textbf{Require}    A class Matrix  that fits with Ambient framework   &  \\  
+                           \textbf{Require}    A class Matrix  that adheres Ambient architecture   &  \\  
                            and  vector of matrices  of size $n$, $A[n]$, $B[n]$, $C[n]$ and $D[n]$ & \\
@@
 	\caption{Generalization of the example \ref{3OPERATIONS} with conditional statements.
-The association of a cumulative stack and  delate stack execution will determine the  directed  acyclic graph.
+The cumulativeness of the stack and delayed manner of execution compromise the directed acyclic graph together.
 \label{GRAPH}}
@@
 
-These examples demonstrate the strength of Ambient, allowing for two levels of parallelization without any hard coding of MPI instructions in the original code, with a great degree of flexibility and adaptability.
+These examples demonstrate the strength of Ambient, allowing two levels of parallelization without any MPI hardcoding inside of the original code, with a great degree of flexibility and adaptability.
 Thus, a fully parallel linear algebra  application can be developed without taking into account the geometry of the problem, which is usually done.

### 2011-08-15 doc/paper_ambient/paper_ambient.tex
some more corrections

@@
 These examples demonstrate the strength of Ambient, allowing two levels of parallelization without any MPI hardcoding inside of the original code, with a great degree of flexibility and adaptability.
-Thus, a fully parallel linear algebra  application can be developed without taking into account the geometry of the problem, which is usually done.
 
@@
 In standard HPC applications, data may have to be prepared for exchange over the network, particularly if the data is of more than one type; in the latter case,
-a MPI structure is set up, to which the data is then copied, and then this structure is send via network. These preliminary conversions can cut into
+an MPI structure is set up, to which the data is then copied, and then this structure is sent via network. These preliminary conversions can cut into
 performance. In Ambient, the data layout inside each work group necessitates additional metadata to coordinate the execution of logistic and 
-computational kernels, particularly the latter has to perform a remote access. Thus, before the block of data held by a workgroup, contiguous
-metadata is included. We call this ensemble \guillemotleft block of data+metadata\guillemotright \, as a network packet, which is built
-on a MPI structure. Thereby, work groups are able to send or receive without any additional movements or copying. 
-To assure proper execution of the framework, several additional type of packets are  available, request packet, layout packet and data-block packet
-they  all derived from a standard packet.
+computational kernels, particularly the latter has to perform a remote access. Thus, the block of data associated with a workgroup exists only in the form that includes meta-data 
+in contiguous memory layout manner. We use this ensemble \guillemotleft block of data + metadata\guillemotright \, directly as a network packet based on top of MPI derived datatypes. 
+Thereby, work groups can move around without additional overhead on copying and any other composure. 
+To assure proper execution of the framework, several additional type of packets are available: control packet, layout packet and data-block packet that derive from one standard packet type.
 
-When a computational kernel is executed, it can access the any required, such as a block of matrix $A$, by use of the Ambient command  \texttt{current(\textit{A})(i,j)}
+During the computational kernel execution, one can access any required remote data, such as a block of matrix $A$, by use of the Ambient modificator  \texttt{current(\textit{A})(i,j)}
 (ref.
-\ref{CURRENT}), where \textit{A} has to be a matrix  compatible with Ambient (\cite{ABSTRACTPARADATATYPE}), the indices $i$ and $j$ being the coordinates of the work group (Fig.
+\ref{CURRENT}), where \textit{A} has to be a matrix or other object that adheres Ambient's architecture (\cite{ABSTRACTPARADATATYPE}), the indices $i$ and $j$ are the coordinates of the work group (Fig.
 \ref{MPIGROUP}).
-There are three possibilities for the resultant types of data communication when a process executes this function inside a computational kernel:
+There are three possibilities of resulting communication while a process issues data via modificators inside computational kernel:
 

### 2011-08-15 doc/paper_ambient/paper_ambient.tex
some changes

@@
 is key to a good performance of the Ambient approach. The resolution of communication patters also falls under the purview
-of the packet manager, as it manages both intra- and inter-group communication. Each MPI-group has a unique packet manager,
-identified by a singleton pattern.
+of the packet manager, as it manages both intra- and inter-group communication. Each MPI-group has its own singleton packet manager.
 
 To reach the best performance in HPC, bottlenecks arising from synchronization operations should be avoided  \cite{EXASCALE-2010}.
-Thus, MPI asynchronous communication is privileged in Ambient (\texttt{mpi\_isend} and \texttt{mpi\_irecv}). Still, an asynchronously
-sent message requires \texttt{mpi\_test} to check whether the packet can be received.  As we avoid synchronization instructions,
-the receiving process repeatedly checks for the availability of the expected data in an infinite loop, until actually receiving the
+Thus, MPI asynchronous communication is privileged in Ambient (\texttt{mpi\_isend} and \texttt{mpi\_irecv}). As we avoid synchronization 
+instructions, the process repeatedly checks for the availability of the incoming data in an infinite loop, until actually receiving the
 data.
 
-For the packet manager to be adaptable to any communication pattern, it employs two parts:
-a delegate mechanism and a finite state machine (FSM).
+For the packet manager to be adaptable to any communication pattern, it employs the delegate mechanism and a finite state machine (FSM).
 
@@
 to encompass various applied computing fields. The connection  between Ambient and the \guillemotleft outside world \guillemotright  \, is done by an intermediate class that assures the
-connectivity between a class and Ambient.
+adherance of class to Ambient's architecture.
 
 For linear algebra, we designed a matrix class which contains the usual data members for describing the object (number of rows, number of columns, \dots) and methods for linear
-algebra operations (associated to logistic and computational kernels). This matrix class derives from another one which ensures the connection to Ambient.
-When an instance of the matrix class is created, the inheritance mechanism will call constructors, methods in the mother class that  will 
-structure the memory (tile  layout).
-
-Therefore, we can extend Ambient to new problems. For example, in fluid mechanics the two-dimensional finite difference method is very popular for solving the Navier-Stokes  differential equations.
-These equations describe the parameters of the \guillemotleft fluid particle \guillemotright. In this case Ambient will discretize the space and assure, as for a matrix,  the distribution
- of the data inside the memory.
-Thus, a fluid particle class will contain enough information to ensure a good discretization in the cartesian space (as the number of rows and columns inside the matrix class) (???).
+algebra operations (associated with logistic and computational kernels). This matrix class derives from another one which ensures the adherence to Ambient's architecture.
 

### 2011-08-15 doc/paper_ambient/paper_ambient.tex
merging two paper_ambient versions

@@
 \title{Ambient: overview of the workflow and design basics}
-\author[A. Kosenkov, T. Ewart, B. Bauerb, A. Kantian,  M. Troyer, T. Giarmarchy]{Alexandr Kosenkov, Timoth\'ee Ewart, Bela Bauer, Adrian Kantian, Matthias Troyer and Thierry Giamarchy }
+\author[A. Kosenkov, T. Ewart, B. Bauerb, A. Kantian,  M. Troyer, T. Giarmarchi]{Alexandr Kosenkov, Timoth\'ee Ewart, Bela Bauer, Adrian Kantian, Matthias Troyer and Thierry Giamarchy }
 
@@
 
-As soon as Ambient encounters the flush signal - kernel
+As soon as Ambient encounters the flush signal [MORE EXPLANATION]- kernel
 execution is initiated along the following sequence:
@@
 \item Execution of logistic kernel: allocation of device (MPI group) according to specified request. 
-\item Execution of logistic kernel: argument's work groups decomposition according to workDim.
-\item Execution of logistic kernel: selected argument's work groups mapping onto device (MPI processes).
+\item Execution of logistic kernel: argument's work groups decomposition according to workDim [MORE EXPLANATION].
+\item Execution of logistic kernel: selected argument's work groups mapping onto device (MPI processes). [MORE EXPLANATION]
 \item Transfer of work groups to the target processes.
@@
 The typical block size is $ 128 \times 128$, although, the user has full control of this size and can adapt it to the needs of the target problem.
-As in CUDA, work groups and corresponding grid of groups is formed in Ambient.
-A basic comparison between CUDA and Ambient is shown in the table \ref{ANALOGY}.
+As in CUDA, work groups and corresponding grid of groups are formed in Ambient.
+A basic comparison between CUDA and Ambient is shown in table \ref{ANALOGY}.
 
@@
 a tile data layout.
-It allows to avoid penalties due to NUMA nature of memory allocation on the CPU (section \ref{MPIABSTRACTION}) and, naturally, this data layout supports work group representation of execution, as shown in Fig.
+It allows to avoid penalties due to NUMA nature of memory allocation on the CPU (section \ref{MPIABSTRACTION}) and, naturally, this data layout supports the work group representation of execution, as shown in Fig.
 \ref{CUDAANALOGY}.
@@
 In CUDA and OpenCL, the execution model imposes kernel initialization with the appropriate resources (\eg block and grid dimensions),
-after which the user does not have to explicitely manage the memory transfer mechanism throughout the GPU; due to distributive memory of clusters and frequently heterogeneous nature of processing units we need both fine-grained control of work group per process assignement and memory transfer operations.
+after which the user does not have to explicitely manage the memory transfer inside the GPU; due to the distributed memory of clusters and the frequently heterogeneous nature of processing units we need both fine-grained control of the work group per process assignement as well as of the memory transfer operations.
 That should be done in a simple, automatic yet flexible way.
-In order to achieve that we split the kernel invocation into two parts.
+In order to achieve this goal we split the kernel invocation into two parts.
 
@@
 \begin{enumerate}
-	\item A logistic kernel; where we select the number of processors, and map the data onto processes indices.
+	\item A logistic kernel; where we select the number of processors, and map the data onto processes indices (???).
 	\item A computational kernel; where we execute required computational operations (similar to the CUDA-kernel).
@@
 The selection of the group is done by issuing an SQL-like request.
-Afterward the data is mapped onto this newly created group using sporadic assignments.
+Afterward the data is mapped onto this newly created group using sporadic (???) assignments.
 Both these functionalities are described in section \ref{SQLSECTION}.
 
-In linear algebra, the optimal distribution in terms of load balancing is the 2D block-cyclic distributions but in general it depends on the target algorithm.
+In linear algebra, the optimal distribution in terms of load balancing often is the 2D block-cyclic distributions, but in general it depends on the target algorithm.
 For example, for the QR factorization \cite{DEMMEL-2008} it can be demonstrated that the row-wise distribution is superior to the 2D block-cyclic decomposition for matrices with more rows than columns.
@@
 Those are partly lister in the table \ref{ANALOGY}.
-For the illustration purpose, the computational kernel (table \ref{CKERNEL}) of an addition of the two matrices associated with previous example of the logistic kernel is presented in the table \ref{LKERNEL}.
+For illustration purposes, the computational kernel (table \ref{CKERNEL}) of an addition of the two matrices associated with previous example of the logistic kernel is presented in the table \ref{LKERNEL}.
 \begin{itemize}
 \item Line 1: the signature of the computational kernel must be the same as of the associated logistic kernel.
-Here we will also include the use of\texttt{pinned} keyword indicating that the execution should be performed over marked argument's work groups.
+Here we will also include the use of the \texttt{pinned} keyword indicating that the execution should be performed over the marked argument's work groups.
 \item Line 2: the cartesian coordinates of the work group are obtained from the Ambient instruction \texttt{get\_block\_id(A).x}  for $i$ and \texttt{get\_block\_id(A).y}  for $j$. \label{CURRENT}
-\item Line 3 to 5 : we refer to the work group $i,j$ via temporary variable $A_{ref}$ using the Ambient instruction \texttt{current(A)(i,j)}.
-Note that if the process requires data outside of its scope - the remote access will be performed in the background (see section \ref{MPIABSTRACTION} for details).
+\item Line 3 to 5 : we refer to the work group $i,j$ via the temporary variable $A_{ref}$ using the Ambient instruction \texttt{current(A)(i,j)}.
+Note that if the process requires data outside of its scope the remote access will be performed in the background (see section \ref{MPIABSTRACTION} for details).
 \item Line 6 to 8 : the addition is performed by looping over all elements inside the work group.
@@
 As it was shown for the matrix addition, we can recode other linear algebra solvers using Ambient's API.
-For the presentation of our framework, we will focus on the MPI version of BLAS named pBLAS \cite{SCALAPACK-1992}, as well as recursive Strassen-like recursive reduced multiplication algorithms.
+For the presentation of our framework, we will focus on the MPI version of BLAS named pBLAS \cite{SCALAPACK-1992}, as well as Strassen-like recursive reduced-multiplication algorithms.
 
@@
 \subsection{SQL resource management}  \label{SQLSECTION}
- In the previous section, we have introduced the logistic kernel and its role in resources initialization for the execution of the computational kernels. 
- One of the main Ambient's qualities is the ability to perform dynamic management of the CPU resources on a cluster.
+ In the previous section, we have introduced the logistic kernel and its role in resource initialization for the execution of the computational kernels. 
+ One of Ambient's main qualities is the ability to perform dynamic management of the CPU resources on a cluster.
 To facilitate this, we have integrated an SQL parser from \cite{SQLite} into Ambient.
-Using this parser we select the desired number of processors according to supplied SQL string inside the logistic kernel.
+Using this parser we select the desired number of processors according the SQL string supplied to the logistic kernel.
 
-At the beginning of the application's execution global MPI group gets created under the name \texttt{ambient}.
-In order to create nested groups we have used the ability of MPI to construct a new MPI group on top of the parent one (chapter 6 of \cite{MPI-2002}).
+At the beginning of the application's execution, the global MPI group is created, nameed \texttt{ambient}.
+In order to create nested groups we have used the ability of MPI to construct a new MPI group on top of the parent group (chapter 6 of \cite{MPI-2002}).
 We encapsulated this into the framework controlling it via the mentioned SQL request mechanism.
 It enables simple yet flexible control over the resources.
-An example request for the addition kernel can be described as below (done by the \texttt{scope\_select} function).\\
+An example request for the addition kernel would be
 
@@
 \end{eqnarray}
-This query will select \texttt{\textit{N}} processors from the global \texttt{ambient} MPI group. The newly created group then is 
+This query, via \texttt{scope\_select}, will select \texttt{\textit{N}} processors from the global \texttt{ambient} MPI group. The newly created group then is 
 accessible under the name \texttt{addition\_group}, its master process will have local rank 0 (see Fig. \ref{MPIGROUP}).
@@
 
-After SQL request section, the data is being mapped to the processes of the newly created MPI group in terms of work groups.
-Frequently the algorithm would be simply 1D or 2D block-cyclic distribution, the algorithm of which is described in the table \ref{2DCYCLIC} (\cf Fig.
+After the SQL request section, the data is being mapped to the processes of the newly created MPI group in terms of work groups.
+Frequently the algorithm would be simply 1D or 2D block-cyclic distribution, the algorithm of which is described in table \ref{2DCYCLIC} (\cf Fig.
 \ref{MPIGROUP}).
-After the actual transfer later those processes will become the primary owners of their data in terms of storage and handling.
-With Ambient every operation is associated with a pair of logistic and computational kernels - that is, there will be multiple SQL requests.
-In order to create new groups, processes from the parent group are being selected in a queue fashion until all processes of the group have been selected.
-If more groups are required - the selection restarts from the beginning of the queue making one process simultaneously a member of several groups.
+After the actual transfer, those processes will become the primary owners of their data in terms of storage and handling.
+With Ambient, every operation is associated with a pair of logistic and computational kernels - that is, there will be multiple SQL requests.
+In order to create new groups, processes from the parent group are being selected in queue fashion until all processes of the group have been selected.
+If more groups are required, the selection restarts from the beginning of the queue, thereby making those processes selected again simultaneous members of several groups.
 
@@
 \end{eqnarray} 
-The function \texttt{get\_id(\textit{A})} will get a tag  of \texttt{\textit{A}} to get its current binding.
+The function \texttt{get\_id(\textit{A})} will get a tag  of \texttt{\textit{A}} to obtain its current binding.
 
@@
 
-Any HPC application can be decomposed into a set of functions performing numerical operations. Let's start from basic example: we want to perform three following operations :
+Any HPC application can be decomposed into a set of functions performing numerical operations. Let's start from a basic example, where we want to perform the three following operations :
 
@@
 \end{eqnarray}
-where $A$, $B$, $C$ and $D$ are some matrices and $\lambda$ is some scalar.
+with $A$, $B$, $C$ and $D$ being some matrices and $\lambda$ being some scalar.
 

### 2011-08-15 doc/paper_ambient/paper_ambient.tex
latest version

@@
 \item Execution of logistic kernel: allocation of device (MPI group) according to specified request. 
-\item Execution of logistic kernel: argument's work groups decomposition according to workDim [MORE EXPLANATION].
-\item Execution of logistic kernel: selected argument's work groups mapping onto device (MPI processes). [MORE EXPLANATION]
+\item Execution of logistic kernel: argument's work groups decomposition according to workDim [INSERT OPENCL REF HERE].
+\item Execution of logistic kernel: selected argument's work groups mapping onto device (MPI processes).
 \item Transfer of work groups to the target processes.
@@
 \begin{enumerate}
-	\item A logistic kernel; where we select the number of processors, and map the data onto processes indices (???).
+	\item A logistic kernel; where we select the number of processes, and map the data onto them.
 	\item A computational kernel; where we execute required computational operations (similar to the CUDA-kernel).
@@
 The selection of the group is done by issuing an SQL-like request.
-Afterward the data is mapped onto this newly created group using sporadic (???) assignments.
+Afterward the data is mapped onto this newly created group using surjective assignments.
 Both these functionalities are described in section \ref{SQLSECTION}.
@@
 of the parallelization. In Ambient, we have made a scheduler that privileges a queue execution model where operations
-are stacked up in a list first and are executed afterwards according to the respective logistic/computational kernel. This model is inspired by computer architecture design, i.e.
-the queue data flow execution in a processor \cite{DENNIS-1974} (Alex: hell, no!).
+are stacked up in a list first and are executed afterwards according to the respective logistic/computational kernel.
 

### 2011-08-15 doc/paper_ambient/paper_ambient.tex
abstract parallel types section

@@
 This way, operations, such as those in (\ref{3OPERATIONS}), will stack up during execution of the application.
-At any time then, the operations on the stack would be available for the execution through the basic Ambient instruction
+At any time then, the operations on the stack would be available for execution through the basic Ambient instruction
 
@@
 Next the computational kernels of these independent operations are executed.
-Following the completion of those, Ambient takes next chunk of independent operations (whose dependencies were just satisfied) and repeats the cycle transporting the necessary data to the new groups after the execution of logistic kernels is complete.
+Following the completion of those, Ambient iterates over the remaining operations and takes the chunk of newly independent operations (whose dependencies were satisfied) and repeats the cycle, transporting the necessary data to the new groups after the execution of logistic kernels is complete.
 In the case of data dependency, the data transfer is handled by the packet manager.
-When all computational kernels have been executed the stack is getting cleaned.
-The schematic overview of the stack execution is presented in Fig.
+When all computational kernels have been executed, the stack is cleaned.
+The schematic overview of stack execution is presented in Fig.
 \ref{EXECUTION} .
 
-In overall, the Ambient employs single kernel multiple data parallelism model that is used over many process groups with different kernels and data.
+Overall, Ambient employs a "single kernel multiple data" parallelism model that is used simultaneously over many process groups with different kernels and data sets.
 
@@
           \caption{Ambient scheduler's workflow model.
-On the classical approach (left) the operations will be executed in parallel step by step disregarding the dependencies (marked by the symbol \guillemotleft [\guillemotright \, in the schematic).
-In Ambient (right), operations will be associated to the logistic/computational kernel pair (abbreviate l/c), as operations 1 and 3  (l/c kernels 1 and 3) are independent - the execution of computational kernel will be performed simultaneously.
+In the classical approach (left) the operations will be executed in parallel step by step disregarding the dependencies (marked by the symbol \guillemotleft [\guillemotright \, in the schematic).
+In Ambient (right), operations will be associated to the logistic/computational kernel pair (abbreviate l/c), as operations 1 and 3  (l/c kernels 1 and 3) are independent - the execution of computational kernels will be performed simultaneously.
 Due to the dependency between the kernel 1 and 2, the computational kernel of operation 2 (l/c kernel 2) will be executed later.
@@
 
-Next, we generalize example \ref{3OPERATIONS} to deal with vectors of matrices.
+Next, we generalize example (\ref{3OPERATIONS}) to deal with vectors of matrices.
 We assume conditional statements determining which matrices should be added and/or multipled (Table \ref{GRAPH}).
@@
 performance. In Ambient, the data layout inside each work group necessitates additional metadata to coordinate the execution of logistic and 
-computational kernels, particularly the latter has to perform a remote access. Thus, the block of data associated with a workgroup exists only in the form that includes meta-data 
-in contiguous memory layout manner. We use this ensemble \guillemotleft block of data + metadata\guillemotright \, directly as a network packet based on top of MPI derived datatypes. 
-Thereby, work groups can move around without additional overhead on copying and any other composure. 
-To assure proper execution of the framework, several additional type of packets are available: control packet, layout packet and data-block packet that derive from one standard packet type.
+computational kernels, particularly if the latter has to perform a remote access. Thus, the block of data associated with a workgroup exists only in a form that includes meta-data 
+in a contiguous memory layout. We use this ensemble \guillemotleft block of data + metadata\guillemotright \, directly as a network packet based on top of MPI derived datatypes. 
+Thereby, work groups can move around without additional copying or any other composition overhead. 
+To assure proper execution of the framework, several additional types of packets are available: control packets, layout packets and data-block packets that derive from one standard packet type.
 
@@
 
-To assure that Ambient has the largest possible scope, we designed the framework on the data structures and not on a specific problem. Thus, we enlarge the utilization of the framework
-to encompass various applied computing fields. The connection  between Ambient and the \guillemotleft outside world \guillemotright  \, is done by an intermediate class that assures the
-adherance of class to Ambient's architecture.
+In order to turn Ambient into general purpose framework, we have decoupled the Ambient data representation from the user datatypes. There are two aspects to the parallel datatypes.
 
-For linear algebra, we designed a matrix class which contains the usual data members for describing the object (number of rows, number of columns, \dots) and methods for linear
-algebra operations (associated with logistic and computational kernels). This matrix class derives from another one which ensures the adherence to Ambient's architecture.
+First is the object's lifetime - as we employ the technique of accumulated execution - we have to unsure that the objects that were created on the stack are still alive at the time of the actual execution.
+For this purpose we introduced livelong class that forces derived data types to dplicate themselves in heap memory upon construction if they were allocated on stack.
+Using this duplicate we can ensure that the object will stay alive until the moment of the actual kernels execution.
+
+Second aspect is the actual connection with the Ambient's work groups engine.
+Each data type that derives from the livelong class has a field pointing to an associated parallel profile of the object.
+This parallel profile object is responsible for the data storage and actual tracking of physical location of the data pieces throughout the cluster's processes.
+The relationship between user type and parallel profile is regulated by means of providing a specialized constructor for the user's data type that sets the memory storage parameters according to the actual user's object parameters in a user defined way.
+
+That is, one is free to design his own data types like parallel sparse matrix alongside with embedded parallel dense matrix type.
+The simpliest example is dense matrix - inside the specialized constructor we simply set the dim.x and dim.y of parallel profile to be matrix cols and matrix rows numbers respectively.
+This setting immidiately sets work group grid of the object considering also provided size of the object's work group.
+Alongside with the dimension other parameters are being set - such as data element size and work block/work item dimensions (the latter can be changed individually for the object during the runtime).
+
+In general, the mechanism described above provides us with all means to adapt the framework in a simple and flexible way to any abstract parallel types that the user might want to use for computations.
 

### 2011-08-15 doc/paper_ambient/paper_ambient.tex
small corrections

@@
 \subsection{Validation}
-As a validation test, we code a new implementation of PDGEMM using the Ambient framework, and compare to the ScaLAPACK version of PDGEMM (MKL-Version).
-The algorithm of our implementation is based on the classical method [CITATION NEEDED] using a column order distribution for the matrices (\cf table \ref{AMBIENTDGEMM} 
-for the modification due to Ambient). The main advantage in our implementation is the required amount of coding, which is limited to seven lines.
+As a validation test, we've coded a new implementation of PDGEMM using the Ambient framework, and compare to the ScaLAPACK version of PDGEMM (MKL-Version).
+The algorithm of our implementation is based on the classical blocking method [CITATION NEEDED] using a column order distribution for the matrices (\cf table \ref{AMBIENTDGEMM} 
+for the modification due to Ambient). One of the obvious advantages in our implementation is the required amount of coding, which is limited to seven lines in this case.
 

### 2011-08-15 doc/paper_ambient/paper_ambient.tex
some corrections to abstract parallel types

@@
 
-In order to turn Ambient into general purpose framework, we have decoupled the Ambient data representation from the user datatypes. There are two aspects to the parallel datatypes.
+In order to turn Ambient into a general purpose framework, we have decoupled the Ambient data representation from the user datatypes. There are two aspects to the parallel datatypes:
 
-First is the object's lifetime - as we employ the technique of accumulated execution - we have to unsure that the objects that were created on the stack are still alive at the time of the actual execution.
-For this purpose we introduced livelong class that forces derived data types to dplicate themselves in heap memory upon construction if they were allocated on stack.
-Using this duplicate we can ensure that the object will stay alive until the moment of the actual kernels execution.
+First is the object's lifetime.
+As we employ the technique of accumulated execution, we have to ensure that the objects that were created on the stack are still alive at the time of the actual execution.
+For this purpose we introduced \texttt { livelong } class that forces derived data types to duplicate themselves in heap memory upon construction if they were allocated on stack.
+Using this duplicate we can ensure that the object will stay alive until the moment of execution of the actual kernels using this data.
 
-Second aspect is the actual connection with the Ambient's work groups engine.
-Each data type that derives from the livelong class has a field pointing to an associated parallel profile of the object.
-This parallel profile object is responsible for the data storage and actual tracking of physical location of the data pieces throughout the cluster's processes.
+The second aspect is the actual connection to Ambient's work groups engine.
+Each data type derived from the livelong class has a field pointing to an associated parallel profile of the object.
+This parallel profile object is responsible for the data storage and actual tracking of the physical location of the data pieces throughout the cluster's processes.
 The relationship between user type and parallel profile is regulated by means of providing a specialized constructor for the user's data type that sets the memory storage parameters according to the actual user's object parameters in a user defined way.
 
-That is, one is free to design his own data types like parallel sparse matrix alongside with embedded parallel dense matrix type.
-The simpliest example is dense matrix - inside the specialized constructor we simply set the dim.x and dim.y of parallel profile to be matrix cols and matrix rows numbers respectively.
-This setting immidiately sets work group grid of the object considering also provided size of the object's work group.
-Alongside with the dimension other parameters are being set - such as data element size and work block/work item dimensions (the latter can be changed individually for the object during the runtime).
+That is, one is free to design his own data types like e.g.
+a parallel sparse matrix type, either alongside the embedded parallel dense matrix type, or to replace it if the user chooses to.
+The simpliest example would be for a dense matrix.
+Inside the specialized constructor we simply set the \texttt{dim.x} and \texttt{dim.y} of the parallel profile to be the numbers of columns and rows respectively.
+Assuming the size of the work group is provided, this immidiately sets the work group grid of the object.
+Alongside with the dimension other parameters are being set - such as the data element size and the work block/work item dimensions (the latter can be changed individually for the object during runtime).
 

### 2011-09-16 doc/paper_ambient/paper_ambient.tex
add a draft of intro in the paper, main idea : tech -> programming model

@@
 \newcommand\cf{\textit{c.f. }}
+\newcommand\df{\textit{d.f. }}
 \renewcommand{\labelitemi}{$\diamond$}
@@
 
-- Stress Performance of Ambient 
-  - in terms of flexible and dynamical resource allocation
-  - Stress performance boost derived from this (we need the benchmarks for that!)
+In the last twenties years, the High Performance Computing (HPC) community has lived a series of technical evolutions that have always influenced the  programming model.
+These new paradigms are always introduced by the new products deliver from the computer industry.  If we examined the two last decades, the most powerful
+machines were based on shared memory architecture(CITE TOP 500), these machines were powerful but expensive and not  configurable, easily. Since, three revolutions appeared.
+
+Firstly, the introduction of modular computer, based  on independent blade system in the nineties , was a  revolution\footnote{The TOP 500 web site inventories the list of
+ the most powerfull computer in the world.
+It delivers statistic among the architecture; we find the SMP system disappeared of the list in 2002} .
+It cuts the price of HPC system, and change the programming  model.
+As the systems are became independents \df the address space is not shared among nodes,  parallel processes have to transmit data over an interconnecting network in order to access
+data that other processes have updated. Several technologies were developed to allow communication between nodes (PVM, NX, Express, P4, \etc), one in particularly were developed as 
+a standard; the Message Passing Interface (MPI).
+This interface has been developed by the participation of more 40 organizations, started working on the standard in 1992, delivered in 1994 for the first version and 1997 for
+the second version. Nowadays, it is the standard for  large grain scalability, it is available under several distributions  from the academic world and HPC constructors.
+
+In the last decade, the distributed memory were confirmed like the technology in HPC architecture machine, 100 \% of the top 500 are now based on this model, as the large scalability applications
+used  a main MPI layer, we can cite as popular   in a few scientist field :  {\sc Quantum ESPRESSO} \cite{QE-2009} in quantum chemistry, Abinit \cite{ABINIT-2005, ABINIT-2009} in quantum mechanic,
+  ScaLAPACK \cite{SCALAPACK-1997} in linear  algebra. Of course, this short list is not exhaustive, and it can be fill up easily in every fields of science and engineering.
+The limitation of this programming model (the scalability) will be the communication overhead and the workload imbalance among process.
+
+Secondly, the multicore technology  introduced by the ship makers the last decade.  Indeed, the progress of the industrial process allowed the integration of several cores into the same die. 
+The technology is available on several processor architecture (Power, X86, \etc). One more time, the programming 
+model was modified. It is curious than multicore technology redirect the shared memory model on the sunset. 
+To program shared memory model, a proved technology exists, it is the POSIX threads library. Two possibilities exist,
+the first one, the programmer utilizes the library directly (pThreads) or he lets the compiler generates a multi-thread executables.  
+Generally, the former option places the burden on the programmer, but when well done, it provides good performance.
+On the other hand,  the second option  named OpenM, easy to program, it is based on preprocessor directive.
+
+As MPI, OpenMP is developed into a consortium, the version 1.0 was released in 1997, the 2.0 in 2005 and new features were integrated in 2008.
+This programming model becomes popular, it is now integrated natively in all compilers and it allows hybrid programs. Hybrid mode MPI/OpenMP
+is the actual programming model in HPC \cite{EXASCALE-2010}.  Per example the performance of the ScaLAPACK library can be improved by a multithread BLAS library (MK, ESSLSMP or ACML). 
+Although this ideology gives results, the threads management  gives an additional workload, when the  grain problem size will become too fine \cite{SMITH-2001}.
+
+And then, the present evolution comes from the introduction of the graphic card into HPC, graphic card exists since the eightiees \cite{iSBX_1983}, the graphic card market 
+was disconnected to the HPC market until a few years ago due to the incredible power of calculation of last generation of GPU. The introducing of the CUDA API allows a real
+opportunity to exploit graphic card into a cluster.
+To summarize the GPU ideology : it is the ability to performed thousand of threads simultaneously, where the thread applied a transformation (kernel)
+on a pixel. Translate to a scientist computational problem, the pixel becomes an entry a tensor (mathematic) or a particle  (kinetic fluid dynamic), \etc.
+
+The classical new programming model introduced by the GPU was an hybrid mode were operation are executed  asynchronously between  multiple CPU's core and GPU.
+The best example of this programming model is the mathematical library MAGMA \cite{MAGMA_2011}.
+
+In this paper, we proposed an alternative approach of programming.  We developed a powerful oriented object framework named Ambient, that transposed  the CUDA ideology to a cluster. 
+As in the CUDA framework, in Ambient developers write kernels which are executed dynamically, and independently if needed, all other the cluster in function of the  resource.
+The needed communications  are assured by a nested MPI layer away from the users. In this paper, we are going to describe the difference of workflow between an Ambient  and a CUDA program, 
+describe the main functionality  of Ambient and perform several  benchmarks.  Until the end of this paper, we will apply and illustrate our framework on basic linear algebra problems.
 

### 2011-09-18 doc/paper_ambient/paper_ambient.tex
paper introduction + more ref

--- 2011-08-15 doc/paper_ambient/paper_ambient.tex
@@
 
-- Stress Performance of Ambient 
-  - in terms of flexible and dynamical resource allocation
-  - Stress performance boost derived from this (we need the benchmarks for that!)
+In the last twenties years, the High Performance Computing (HPC) community has lived a series of technical evolutions that have always influenced the  programming model.
+These new paradigms are always introduced by the new products deliver from the computer companies.  If we examined the two last decades, the most powerful
+machines have moved from shared memory architecture, powerful but expensive and not  configurable, easily,  to distributed memory with multicores processors using
+optional accelerators \cite{Top500}.  This dual transition  material/programming models  were done in three distinctive steps.
+
+Firstly, the introduction of modular computer, based  on independent blade system in the nineties, was a  revolution\footnote{The TOP 500 web site inventories the list of
+ the most powerfull computer in the world.
+It delivers statistic among the architecture; we find the SMP system disappeared of the list in 2002} .
+It cuts the price of HPC system, and change the programming  model.
+ We moved from shared memory API to distributed API. Indeed, as the systems  became independents \df the address space is not shared among nodes, 
+  parallel processes have to transmit data over an interconnecting network in order to access
+data that other processes have updated. Several technologies were designed to allow communication between nodes (PVM, NX, Express, P4, \etc) \cite{Aoyama:1999}, one in particularly were developed as 
+a standard; the Message Passing Interface (MPI) \cite{Snir:1998:MCR:552013}. This interface has been developed by the participation of more 40 organizations, started working on the standard in 1992,
+ delivered in 1994 for the first version and 1997 for  the second version.
+Nowadays, it is the standard for  large grain scalability, it is available under several distributions  from the academic world and HPC constructors.
+
+In the last decade, the distributed memory were confirmed like the technology in HPC architecture machine, 100 \% of the top 500 are now based on this model, as the large scalability applications
+used  a main MPI layer, we can cite as popular   in a few scientist field :  [GIVE ME A MPI CODE INTO ALPS !!!!]   {\sc Quantum ESPRESSO} \cite{QE-2009} in quantum chemistry, Abinit \cite{ABINIT-2005, ABINIT-2009} in
+quantum mechanic,   ScaLAPACK \cite{SCALAPACK-1997} in linear  algebra or SMILE in hypersonic rarefied flow \cite{SMILE-2005}.
+Of course, this short list is not exhaustive, and it can be fill up easily in every fields of
+science and engineering. The limitation of this programming model (the scalability) will be the communication overhead and the workload imbalance among process.
+
+Secondly, the multicore technology  introduced by the ship makers the last decade.  Indeed, the progress of the industrial process allowed the integration of several cores into the same die. 
+This technology is now available on all  processor architecture (Power, X86, ARM, \etc). One more time, the programming 
+model was modified, and it re-actualized the shared memory model.  To program shared memory machine, a proved technology exists, it is the POSIX threads library. 
+The programmers have two possibilities. They utilize the library directly (pThreads) or he lets the compiler generates a multi-thread executables.  
+Generally, the former option places the burden on the programmer, but when well done, it provides good performance.
+On the other hand,  the second option more knew under the  name OpenMP  \cite{openmp:2002}, easier,  is based on  preprocessor directive. This second option
+is clearly the most popular nowadays.
+
+As MPI, OpenMP is developed into a consortium, the version 1.0 was released in 1997, the 2.0 in 2005 and new features were integrated in 2008.
+This programming model becomes popular, it is now integrated natively in all compilers and it allows hybrid programs. Hybrid mode MPI/OpenMP
+is the actual programming model in HPC \cite{EXASCALE-2010}.  Per example the performance of the ScaLAPACK library can be improved by a multithread BLAS library (MKL, ESSLSMP or ACML). 
+Although this ideology gives results, the threads management  gives an additional workload, when the  grain problem size will become too fine \cite{SMITH-2001}.
+
+And then, the present evolution comes from the introduction of the graphic card into HPC, graphic card exists since the eighties \cite{iSBX_1983}, the graphic card market 
+was disconnected to the HPC  until a few years ago due to the incredible power of calculation of last generation of GPU.
+The introducing of the CUDA \cite{CUDA2011} \footnote{An  Open Source alternative to CUDA exists
+it is OpenCL, it also extents the CUDA ideology to  multicore processor.}  API allows a real opportunity to exploit graphic card into a cluster.
+To summarize the GPU ideology : it is the ability to performed thousand of threads simultaneously, where the threads applied a transformation (kernel)
+to a single or a piece of data.
+
+The classical new programming model introduced by the GPU was an hybrid mode were operation are executed  asynchronously between  multiple CPU's core and GPU.
+As the pThreads with OpenMP, the programmer can manage by hand or delegate by preprocessor directives  \cite{PGICUDAFOR2010}.
+The best example of this programming model is the mathematical library MAGMA \cite{MAGMA_2011}.
+
+In this paper, we proposed an alternative approach of programming.  We developed a powerful oriented object framework named Ambient, that transposed  the CUDA ideology to a cluster. 
+As in the CUDA framework, in Ambient, the developer writes kernels which are executed in parallel, dynamically, and independently if needed, all other the resource of the cluster.
+The needed communications  are assured by a nested MPI layer away from the users. From the user point of view, the objective is to write  parallel program without interacting the MPI API.
+
+In this paper, we are going to describe the difference of workflow between an Ambient  and a CUDA program,  describe the blue print of Ambient and evaluated on several  benchmarks.  
+Until the end of this paper, we will apply and illustrate our framework on basic linear algebra problems.
 
@@
 
-As soon as Ambient encounters the flush signal [MORE EXPLANATION]- kernel
+As soon as Ambient encounters the flush signal (contrary to CUDA, the programmer orders the execution   as in queue model) kernel
 execution is initiated along the following sequence:
@@
 \item Execution of logistic kernel: allocation of device (MPI group) according to specified request. 
-\item Execution of logistic kernel: argument's work groups decomposition according to workDim [INSERT OPENCL REF HERE].
+\item Execution of logistic kernel: argument's work groups decomposition according to workDim \cite{OPENCL2011}.
 \item Execution of logistic kernel: selected argument's work groups mapping onto device (MPI processes).

### 2011-09-19 doc/paper_ambient/paper_ambient.tex
paper

@@
 
-In the last decade, the distributed memory were confirmed like the technology in HPC architecture machine, 100 \% of the top 500 are now based on this model, as the large scalability applications
+In the last decade, the distributed memory clusters were confirmed like the technology in HPC architecture machine, 100 \% of the top 500 are now based on this model.
+The large scalability applications
 used  a main MPI layer, we can cite as popular   in a few scientist field :  [GIVE ME A MPI CODE INTO ALPS !!!!]   {\sc Quantum ESPRESSO} \cite{QE-2009} in quantum chemistry, Abinit \cite{ABINIT-2005, ABINIT-2009} in
 quantum mechanic,   ScaLAPACK \cite{SCALAPACK-1997} in linear  algebra or SMILE in hypersonic rarefied flow \cite{SMILE-2005}.
-Of course, this short list is not exhaustive, and it can be fill up easily in every fields of
+Of course, this  list is not exhaustive, and it can be fill up easily in every fields of
 science and engineering. The limitation of this programming model (the scalability) will be the communication overhead and the workload imbalance among process.
@@
 Secondly, the multicore technology  introduced by the ship makers the last decade.  Indeed, the progress of the industrial process allowed the integration of several cores into the same die. 
-This technology is now available on all  processor architecture (Power, X86, ARM, \etc). One more time, the programming 
+This technology is now available on all  processor architectures (Power, X86, ARM, \etc). One more time, the programming 
 model was modified, and it re-actualized the shared memory model.  To program shared memory machine, a proved technology exists, it is the POSIX threads library. 
-The programmers have two possibilities. They utilize the library directly (pThreads) or he lets the compiler generates a multi-thread executables.  
+The programmers have two possibilities. They utilize the library directly (pThreads) or they let the compiler generates a multi-thread executables.  
 Generally, the former option places the burden on the programmer, but when well done, it provides good performance.
-On the other hand,  the second option more knew under the  name OpenMP  \cite{openmp:2002}, easier,  is based on  preprocessor directive. This second option
+On the other hand,  the second option more knew under the  name OpenMP  \cite{openmp:2002}, easier,  is based on  preprocessor directive. This last alternative
 is clearly the most popular nowadays.
@@
 As MPI, OpenMP is developed into a consortium, the version 1.0 was released in 1997, the 2.0 in 2005 and new features were integrated in 2008.
-This programming model becomes popular, it is now integrated natively in all compilers and it allows hybrid programs. Hybrid mode MPI/OpenMP
+It is now integrated natively in all compilers and it allows hybrid programs. Hybrid mode MPI/OpenMP
 is the actual programming model in HPC \cite{EXASCALE-2010}.  Per example the performance of the ScaLAPACK library can be improved by a multithread BLAS library (MKL, ESSLSMP or ACML). 
@@
 
-In this paper, we proposed an alternative approach of programming.  We developed a powerful oriented object framework named Ambient, that transposed  the CUDA ideology to a cluster. 
-As in the CUDA framework, in Ambient, the developer writes kernels which are executed in parallel, dynamically, and independently if needed, all other the resource of the cluster.
-The needed communications  are assured by a nested MPI layer away from the users. From the user point of view, the objective is to write  parallel program without interacting the MPI API.
+The purpose of the  paper is to present a new programming model based on a powerful oriented object (C++) framework named Ambient, that transposed  the CUDA ideology to a cluster. 
+As in the CUDA framework, in Ambient, the developer writes kernels which are executed in parallel under MPI group  all other the resource of the cluster, dynamically, and independently if needed.
+The  communications  are assured by a nested MPI layer away from the users. Our main objective is to allow parallel programming for the user without interacting  with the MPI API.
 
-In this paper, we are going to describe the difference of workflow between an Ambient  and a CUDA program,  describe the blue print of Ambient and evaluated on several  benchmarks.  
-Until the end of this paper, we will apply and illustrate our framework on basic linear algebra problems.
+In this paper, we are going to describe the difference of workflow between an Ambient  and a CUDA program,  describe the general blue print of Ambient and evaluated on several  benchmarks on basic linear algebra.
+Until the end of this paper, we will apply and illustrate our framework on basic linear algebra problems, especially on pBLAS operations.
 
@@
 
-As soon as Ambient encounters the flush signal (contrary to CUDA, the programmer orders the execution   as in queue model) kernel
-execution is initiated along the following sequence:
+As soon as Ambient encounters the flush signal (contrary to CUDA, the programmer decides  the execution start  as in the queue model programming)
+kernel (a logistic kernel prepares the memory, a computational kernel execute pure calculation,   a deeper description of the pair logistics/computational kernels is given in the next section) execution is initiated along the following sequence:
 
@@
 
-When the framework encounters the signal to flush the stack of kernels the actual execution starts. The framework's engine goes through the accumulated stack 
-and executes all independent operations at once. Executing an operation, the engine first invokes the associated logistic kernel and allocates a specified
+The execution of kernels named \guillemotleft  accumulated execution \guillemotright is one of the most important purpose of Ambient because it can be parallel.
+Indeed several kernels can be push into the stack before the execution signal. When the framework encounters the signal to flush the stack of kernels the actual execution starts. 
+The framework's engine goes through the accumulated stack  and executes all independent operations at once.
+Executing an operation, the engine first invokes the associated logistic kernel and allocates a specified
 amount of resources for this operation. Next, the engine sets the argument's (here: the matrices) tiling using user specified work group sizes and maps resulting work groups onto allocated
@@
 A basic comparison between CUDA and Ambient is shown in table \ref{ANALOGY}.
+Moreover the representation of the data by work group fits very well with linear algebra and the matrixes representations (basic types).
 
@@
 The work group approach imposes the data layout of the framework.
-The standard math libraries impose both column order as well as a contiguous memory order on a matrix storage (figure \ref{MEMMATRIX}).
+The standard math libraries impose both column-major-order as well as a contiguous memory order on a matrix storage (figure \ref{MEMMATRIX}).
 
@@
 In CUDA and OpenCL, the execution model imposes kernel initialization with the appropriate resources (\eg block and grid dimensions),
-after which the user does not have to explicitely manage the memory transfer inside the GPU; due to the distributed memory of clusters and the frequently heterogeneous nature of processing units we need both fine-grained control of the work group per process assignement as well as of the memory transfer operations.
+after which the user does not have to explicitly manage the memory transfer inside the GPU; due to the distributed memory of clusters and the frequently heterogeneous nature of processing units we need both fine-grained control of the work group per process assignement as well as of the memory transfer operations.
 That should be done in a simple, automatic yet flexible way.
@@
                                \tiny{1:}   & \textbf{for} from $i$  to the size of the vector \textbf{do} \\
-                               \tiny{2:}   & \hspace{0.2 cm} \textbf{for} from $j$  to the size of the vector \textbf{do} \\                               
+                               \tiny{2:}   & \hspace{0.2 cm} \textbf{for} from $j(0)$  to the size of the vector \textbf{do} \\                               
                                \tiny{3:}   & \hspace{0.4 cm}  \textbf{if}  the conditions on $A[i]$ and  $B[j]$ are verified \textbf{do} \\                               
@@
 
-First is the object's lifetime.
-As we employ the technique of accumulated execution, we have to ensure that the objects that were created on the stack are still alive at the time of the actual execution.
+First is the object's lifetime. Due to the accumulated execution model, objects may be used after their lexical scope has expired.
 For this purpose we introduced \texttt { livelong } class that forces derived data types to duplicate themselves in heap memory upon construction if they were allocated on stack.

### 2011-10-05 doc/paper_ambient/paper_ambient.tex
paper bench
paper ambient merge Adrian corrections, and continuing benchmarks sections

@@
 
-%Nowadays, the numerical simulations get an essential position, as experiments,  for studying physical 
-%properties of many problems. The numerical simulations are presented inside every fields of sciences 
-%and engineering, as classical and quantum mechanic, quantum chemistry,  \etc. As usual, the users always
-%want increase the size of their problems, as billion number of particles for a kinetic simulation, billion 
-%of meshes for a fluid dynamic computation, or else, matrixes of million entries for a quantum mechanic,
-%or the linear algebra problems. Thus, the high performance computing is the solution for solving very large problem,
-%with the peta flops  and the future generation of exa flops computers, the engineers and the scientists will
-%have a quasi unlimited set of resources for modeling, nevertheless, they will be confronted to a challenge, have compatible
-%programs for these machines.
+Over the last twenty years, the High Performance Computing (HPC) community has seen the technical evolution of hardware continuously influence the programming model.
+Over this timespan, the most powerful machines have moved from shared to distributed memory architecture, with networks of
+distributed memory nodes with multicore processors using optional accelerators being the current norm\cite{Top500}.  This transition took place in three distinctive steps.
 
-%A large set of programs; solvers, and software solutions are now availabled to solve large problem in a lot of fields, \eg we can 
-%cite as popular in a few scientist field :  {\sc Quantum ESPRESSO} \cite{QE-2009} in quantum chemistry, Abinit \cite{ABINIT-2005, ABINIT-2009} in quantum mechanic,  ScaLAPACK \cite{SCALAPACK-1997} in linear
-%algebra,  or else SMILE in hypersonic rarefied flow \cite{SMILE-2005}. Of course, this short list is not exhaustive,
-%and it can be fill up easily in every fields of science and engineering.
+First, was introduction of modular nodes based on modular blade systems in the nineties\footnote{SMP systems disappeared from theTOP 500 list 
+of supercomputers in 2002}. Cutting hardware costs, this changed the programming model from shared to distributed memory API, as memory address space was
+not shared among nodes and parallel processes now had to transmit data over an interconnecting network. Several technologies were designed to allow communication
+between nodes (PVM, NX, Express, P4, \etc) \cite{Aoyama:1999}, in particular the Message Passing Interface (MPI) \cite{Snir:1998:MCR:552013}. which has become the
+standard for coarse-grained problem ( \footnote{A number of scientific applications requiring scalability currently employ a main MPI layer,  [GIVE ME A MPI CODE INTO ALPS !!!!]
+{\sc Quantum ESPRESSO} \cite{QE-2009} for quantum chemistry, Abinit \cite{ABINIT-2005, ABINIT-2009} in quantum mechanics, ScaLAPACK \cite{SCALAPACK-1997}
+in linear  algebra or SMILE for hypersonic rarefied flows \cite{SMILE-2005}, to give just a brief, non-exhaustive list.} The factor limiting scalability in this programming model
+is the communication overhead and the workload imbalance among processes.
 
-%If the scientists or the  engineers have the right tools, it will be necessarily  limited by the scalability of the code, in the case of the famous mathematical library ScaLAPACK, the
-%efficiency  and thus the scalability, with the appropriated setting will reach up to 512 processors ($\sim  \mathcal{O}(100)$) \cite{SCALAPACK-1992}, beyond the interest is useless
-%due to the degradation of the performance. In the worst case, the programmers will start  from a white page, and must design and efficient code.
-%Due to the incredible number of cores, and optional accelerators as GPU, FGPA or "CELL like\footnote{Although the CELL processor is now depreciated, it herebies  
-%a way of integrated processor (CPU+GPU), as the prove the last processor products of Intel (core i series) and AMD (fusion).}", the programmers 
-%will have to introduce a mix programming model usually named mix-mode, based on a main MPI layer coupled with one or several additional layers \cite{EXASCALE-2010}.
+Secondly, the introduction of multicore technology has allowed the integration of several CPU cores into the same die, 
+rendering the shared memory model of parallelization once more current. The POSIX threads library is a proven technology in this regard, offering
+two possibilities: either utilize the library directly (using pThreads), or let the compiler generate a multi-threaded executable. The first option places the burden of
+shared memory parallelization on the programmer, but provides good performance when well done, while the second option, called OpenMP \cite{openmp:2002},
+is based on preprocessor directives and easier to use.  It is now integrated natively in all compilers, also allowing for hybrid programming, with hybrid MPI/OpenMP
+being the current programming model in HPC \cite{EXASCALE-2010}.  For example, the performance of the ScaLAPACK library has been shown to improve by
+employing multithreaded BLAS libraries (MKL, ESSLSMP or ACML) [A.K.: NEED CITATIONS HERE.] 
+Although this ideology can yield good results, the threads management imposes an additional workload when the grain problem size becomes too fine \cite{SMITH-2001}.
 
-%The most popular layer for mix-mode is OpenMP, and it is based on the preprocessor directives \texttt{\#pragma}, it demonstrated its performance. It is included successfully inside
-%the  linear algebra libraries like the MKL from Intel or the PESSLSMP from IBM.  This ideology of the utilization of the pre-directives becomes more and more 
-%important, thus it is used for vectorization of loops using the intel compiler and even for GPU with the PGI compiler. Although this ideology gives results, there is 
-%a limitation for the main following reason: the creation/destruction of the threads will be extremely time consuming,
-%when the  grain problem size becomes too fine \cite{SMITH-2001}. To achieve the best performance, the number of thread should be  equal to the number of cores on a socket, 
-%thus the scalability will be increased at maximum by an order of magnitude.  The maximum number of processors now reach, becomes ($\sim \mathcal{O}(1000)$). 
-%This ideology of implementation of the mix mode is more an evolution of the programming model than a revolution.
-It is an fast hand-on to implement that prolongates the life and the performance on a existing code.
-But,
-%it will not prepare at all the code for the future, that will be automatically depreciated  by the new generation of cluster, in term of scalability.
+The latest step in this development is the introduction of graphics processing units (GPUs) into HPC.
+These devices typically perform hundreds or thousands of threads simultaneously, each applying a kernel
+to a single or multiple pieces of data. (???)
+As a framework to enable the use of these GPUs on a cluster, two APIs exist, CUDA \cite{CUDA2011} and OpenCL. Mixed CPU/GPU nodes can be run in hybrid mode, i.e. operations are split up
+and executed asynchronously between multiple CPU cores and the GPUs. As with pThreads, the programmer can manage this by hand or delegate via preprocessor directives \cite{PGICUDAFOR2010}.
+The MAGMA library \cite{MAGMA_2011} would be an example of this model.
 
-%For getting several order of magnitude higher, we need a new programming model. The introduction of the GPU computing by NVidia a few years ago was
-%a real innovation in HPC \cite{NICKOLLS-2010}.
-On a GPU, it is usual to program  thousand of threads.
-Thus, in this paper, we propose a innovative programming model to program on very large distributed cluster.
-We develop
-%a framework named Ambient (C++) where we expend the CUDA programming ideology to  the a full cluster, thus we keep away the MPI programming model from the users.  In a way, we  extended
-%the philosophy of OpenCL, OpenCL extents  the GPU programming model to a processor, assimilate a socket as a GPU, where every cores of the socket are assimilated to a streaming processor.
-%We also developed inside a dynamic stack manager, that stocks and manages the execution of numerical "operations-solver" in functions of the resources of the distributed cluster.
+For the HPC developer, the challenge amidst all these advances is not just to develop code adapted to the latest hardware capabilities 
+and constraints, but to address both the given bottlenecks of the algorithm, the (possibly changing)  bottlenecks of the hardware, and that these 
+two may not be (or stay) congruent. Furthermore, in scientific simulations particularly, the structure of the problem has usually been
+exploited to write a complex code that performs very well - on a single node. The question of how to transpose such code to a large distributed
+cluster while keeping useful and well-performing design is non-trivial, as the direct approach of MPI-parallelizing each operation by hand may
+be prohibitively complex and expensive.
 
-%In this paper, we are going to describe our object oriented framework, and promote a first application based on linear algebra. Our main objective, is to 
-%promote a new technology to develop  parallel application, without any interaction with the MPI and thread layer, and allow, a real access in term of performance to the next generation
-%of clusters. The first possible  targets are applications that require large buffer of contiguous memory: quantum mechanic, linear algebra,  computational fluid dynamics, molecular dynamic \ldots 
-%\, Until the end of this paper, we will focus on the linear algebra problems and their applications.
+The purpose of the paper is to present a new programming model based on a powerful oriented object (C++) framework named Ambient, that
+transposes the CUDA ideology to a cluster. As in the CUDA framework, in Ambient, the developer writes kernels which are executed in parallel under
+MPI groups, allocating cluster resources, dynamically if needed. The communications are assured by a nested MPI layer away from the user.
+Thus, amongst its capabilities, Ambient allows a bootstrap parallelization of existing serial or shared-memory parallel code to distributed-memory systems,
+where the operations that the serial code would perform at each step are collected into a stack. This operation stack is then executed, with Ambient grouping
+dependent and independent operations as necessary, designating MPI processes as appropriate to each of them, and, crucially, managing data storage and
+transfer between MPI processes with an eye towards keeping data volume over the network low and traffic fast. Any algorithm whose basic operations and data
+objects are inhomogenous in terms of time and size requires this dynamically adaptable resource allocation in practice, which Ambient handles largely in the background.
+The scientific algorithm for which Ambient was originally developed, the Density Matrix Renormalization Group (DMRG) [A.K.: INSERT CITATIONS HERE] for example has just this structure,
+as its basic operations are mostly dense matrix-matrix multiplications, with the dense matrices ranging widely in size. The largest ones may necessitate distribution
+over several nodes, while smaller ones may fit easily into the memory of a single node.
 
-In the last twenties years, the High Performance Computing (HPC) community has lived a series of technical evolutions that have always influenced the  programming model.
-These new paradigms are always introduced by the new products deliver from the computer companies.  If we examined the two last decades, the most powerful
-machines have moved from shared memory architecture, powerful but expensive and not  configurable, easily,  to distributed memory with multicores processors using
-optional accelerators \cite{Top500}.  This dual transition  material/programming models  were done in three distinctive steps.
-
-Firstly, the introduction of modular computer, based  on independent blade system in the nineties, was a  revolution\footnote{The TOP 500 web site inventories the list of
- the most powerfull computer in the world.
-It delivers statistic among the architecture; we find the SMP system disappeared of the list in 2002} .
-It cuts the price of HPC system, and change the programming  model.
- We moved from shared memory API to distributed API. Indeed, as the systems  became independents \df the address space is not shared among nodes, 
-  parallel processes have to transmit data over an interconnecting network in order to access
-data that other processes have updated. Several technologies were designed to allow communication between nodes (PVM, NX, Express, P4, \etc) \cite{Aoyama:1999}, one in particularly were developed as 
-a standard; the Message Passing Interface (MPI) \cite{Snir:1998:MCR:552013}. This interface has been developed by the participation of more 40 organizations, started working on the standard in 1992,
- delivered in 1994 for the first version and 1997 for  the second version.
-Nowadays, it is the standard for  large grain scalability, it is available under several distributions  from the academic world and HPC constructors.
-
-In the last decade, the distributed memory clusters were confirmed like the technology in HPC architecture machine, 100 \% of the top 500 are now based on this model.
-The large scalability applications
-used  a main MPI layer, we can cite as popular   in a few scientist field :  [GIVE ME A MPI CODE INTO ALPS !!!!]   {\sc Quantum ESPRESSO} \cite{QE-2009} in quantum chemistry, Abinit \cite{ABINIT-2005, ABINIT-2009} in
-quantum mechanic,   ScaLAPACK \cite{SCALAPACK-1997} in linear  algebra or SMILE in hypersonic rarefied flow \cite{SMILE-2005}.
-Of course, this  list is not exhaustive, and it can be fill up easily in every fields of
-science and engineering. The limitation of this programming model (the scalability) will be the communication overhead and the workload imbalance among process.
-
-Secondly, the multicore technology  introduced by the ship makers the last decade.  Indeed, the progress of the industrial process allowed the integration of several cores into the same die. 
-This technology is now available on all  processor architectures (Power, X86, ARM, \etc). One more time, the programming 
-model was modified, and it re-actualized the shared memory model.  To program shared memory machine, a proved technology exists, it is the POSIX threads library. 
-The programmers have two possibilities. They utilize the library directly (pThreads) or they let the compiler generates a multi-thread executables.  
-Generally, the former option places the burden on the programmer, but when well done, it provides good performance.
-On the other hand,  the second option more knew under the  name OpenMP  \cite{openmp:2002}, easier,  is based on  preprocessor directive. This last alternative
-is clearly the most popular nowadays.
-
-As MPI, OpenMP is developed into a consortium, the version 1.0 was released in 1997, the 2.0 in 2005 and new features were integrated in 2008.
-It is now integrated natively in all compilers and it allows hybrid programs. Hybrid mode MPI/OpenMP
-is the actual programming model in HPC \cite{EXASCALE-2010}.  Per example the performance of the ScaLAPACK library can be improved by a multithread BLAS library (MKL, ESSLSMP or ACML). 
-Although this ideology gives results, the threads management  gives an additional workload, when the  grain problem size will become too fine \cite{SMITH-2001}.
-
-And then, the present evolution comes from the introduction of the graphic card into HPC, graphic card exists since the eighties \cite{iSBX_1983}, the graphic card market 
-was disconnected to the HPC  until a few years ago due to the incredible power of calculation of last generation of GPU.
-The introducing of the CUDA \cite{CUDA2011} \footnote{An  Open Source alternative to CUDA exists
-it is OpenCL, it also extents the CUDA ideology to  multicore processor.}  API allows a real opportunity to exploit graphic card into a cluster.
-To summarize the GPU ideology : it is the ability to performed thousand of threads simultaneously, where the threads applied a transformation (kernel)
-to a single or a piece of data.
-
-The classical new programming model introduced by the GPU was an hybrid mode were operation are executed  asynchronously between  multiple CPU's core and GPU.
-As the pThreads with OpenMP, the programmer can manage by hand or delegate by preprocessor directives  \cite{PGICUDAFOR2010}.
-The best example of this programming model is the mathematical library MAGMA \cite{MAGMA_2011}.
-
-The purpose of the  paper is to present a new programming model based on a powerful oriented object (C++) framework named Ambient, that transposed  the CUDA ideology to a cluster. 
-As in the CUDA framework, in Ambient, the developer writes kernels which are executed in parallel under MPI group  all other the resource of the cluster, dynamically, and independently if needed.
-The  communications  are assured by a nested MPI layer away from the users. Our main objective is to allow parallel programming for the user without interacting  with the MPI API.
-
-In this paper, we are going to describe the difference of workflow between an Ambient  and a CUDA program,  describe the general blue print of Ambient and evaluated on several  benchmarks on basic linear algebra.
-Until the end of this paper, we will apply and illustrate our framework on basic linear algebra problems, especially on pBLAS operations.
+In the following, we will describe the general blueprint of Ambient and its' workflow, the differences to CUDA, and evaluate several basic linear algebra benchmarks.  
+We will apply and illustrate our framework on basic linear algebra problems, especially on pBLAS operations.
 
@@
 
-As soon as Ambient encounters the flush signal (contrary to CUDA, the programmer decides  the execution start  as in the queue model programming)
-kernel (a logistic kernel prepares the memory, a computational kernel execute pure calculation,   a deeper description of the pair logistics/computational kernels is given in the next section) execution is initiated along the following sequence:
+[A.K.: VERY BRIEF DESCRIPTION OF CUDA HERE] Any Ambient-parallelized operation on the other hand is broken down
+in two parts:  first, a logistic kernel sets up the necessary MPI processes and prepares the data. Secondly,
+a computational kernel executes the calculation, performing any necessary communication between MPI processes in the background.
 
+In a code bootstrap-parallelized via Ambient, operations are stacked up symbolically and are actually performed when a flush signal
+is encountered (accumulated execution model); contrary to CUDA, the programmer decides when to start execution.
+Kernel execution is initiated along the following sequence:
+(   a deeper description of the pair logistics/computational kernels is given in the next section)
 \begin{itemize}
 \item Execution of logistic kernel: allocation of device (MPI group) according to specified request. 
-\item Execution of logistic kernel: argument's work groups decomposition according to workDim \cite{OPENCL2011}.
-\item Execution of logistic kernel: selected argument's work groups mapping onto device (MPI processes).
+\item Execution of logistic kernel: different parts of an argument (i.e. data) passed to kernel are mapped to the different 
+        work groups making up the device (i.e. the MPI group).
+\item Execution of logistic kernel: selected argument's work groups are mapped onto device (MPI processes).
 \item Transfer of work groups to the target processes.
@@
 
-The execution of kernels named \guillemotleft  accumulated execution \guillemotright is one of the most important purpose of Ambient because it can be parallel.
-Indeed several kernels can be push into the stack before the execution signal. When the framework encounters the signal to flush the stack of kernels the actual execution starts. 
+The  accumulated execution model is one of the important features of Ambient, it easily accomodates another layer of parallelization, that of operations,
+if that is useful for the problem at hand.
+Indeed several kernels can be push into the stack before the execution signal.
+When the framework encounters the signal to flush the stack of kernels the actual execution starts.
 The framework's engine goes through the accumulated stack  and executes all independent operations at once.
@@
 
------------------------------------------------------------------------------------------------------ \\
-TO DO: EXPLAIN THE ADVANTAGE AGAINST   SCALAPACK \\
-IN TERM OF NETWORK, DATA EXCHANGE \\
------------------------------------------------------------------------------------------------------ \\
-
 We compare the Ambient results with the ScaLAPACK implementation. As related in section \ref{CUDAIDEOLOGYSECTION}, Ambient is compatible with ScaLAPACK,
@@
 [STARTED REWRITING THIS; BUT NOT FINISHED]
-The first benchmark consists of evaluating the performance of our parallel matrix multiplication relative to the iScaLAPACK-MKL mplementation. Both solvers are included as computational kernels
+The first benchmark consists of evaluating the performance of our parallel matrix multiplication relative to the ScaLAPACK-MKL implementation. Both solvers are included as computational kernels
 in Ambient.
@@
 representation.
-The benchmarks  were performed on AMD Opteron  Magny-Cours) CPUs at 2.1 GHz on the Swiss National Supercomputing Centre facilities.
-The performance is reported in [GFlops], it is simply
+The benchmarks  were performed on AMD Opteron  Magny-Cours) CPUs at 2.2 GHz on the Swiss National Supercomputing Centre facilities.
+The peak performance $P_m$ is reported in [GFlops], it is simply
 calculated by :
@@
 \begin{eqnarray}
-P  = \frac{2n^i}{t},
+P_m  = \frac{2n^i}{t},
 \end{eqnarray}
@@
 \subsection{Second benchmark case}
-This benchmarks consists to evaluate the performance of the scheduler, especially the repartition of several operation all over the cluster.
+This benchmarks  evaluates the global performance of the scheduler, especially, when  the independent operations are executed all over the cluster. 
+We now multiply two vectors of matrices pair by pair. Every multiplication is performed by two MPI process and the size of the vectors vary from to
+1 to 48. Thus, the final number of MPI process is equal to 96 which corresponds to fill up 4 calculation nodes.
+
+Before presenting and commenting the results, we would like highlight the readers of the programming of this benchmark. The coding simply consists of few lines
+of declarations, initializations and a basic loop for, as we presented into the algorithm \ref{EXECUTION} . Naturally, coded algorithm  integrates two levels of distributed parallelization 
+without integrate any MPI instructions.
+
+In this benchmark, we  plot the ratio between the performance peak of  vector of matrices $P_{v_m}$ and the associated peak of the machine $P_m$. It is expressed by 
+$R = P_{v_m}/P_m$, or
+
+\begin{eqnarray}
+R = 100\frac{N_m 2n^3}{N_p t4f},
+\end{eqnarray}
+where $N_m$ is the number of matrices, $N_p$ the number of processor, $t$ the full time of execution and $f$ the frequency of the processor. Results are presented
 
 \begin{figure}
-\include{PLOT/VectorPDGEMM/VectorPGEMM}
-\caption{Performances of a vector of PDGEMM using ScaLAPACK}
+\include{PLOT/VectorPDGEMM/VectorWL}
+\caption{something else}
+\end{figure}
+
+\begin{figure}
+\include{PLOT/VectorPDGEMM/VectorWL_2}
+\caption{something else}
 \end{figure}

### 2012-07-04 doc/paper_vli/vli.tex
draft paper vli

\documentclass[11pt]{amsart}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\title{Brief descriptions of the VLI library}
\author{Timothee/Andreas or TAinmdortehaese, that is the question}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle
%\section{}
%\subsection{}

The main purpose of the VLI lib is performed hybrid (CPU/GPU) inner product of polynomial from 1 to 4 variables with large integer coefficient. We will first remember the basics of the inner product of a polynomial, 
and the implementation of the cpu version  The second par will be on the gpu where the inner product and long arithmetic are nested.

\section*{Inner product}

The main object of the library is performed an inner product of vector of polynomials. The polynomials $P$ are defined like :
\begin{eqnarray}
P_a(x,y,\dots) = \sum_i^n \left( \sum_j^n \dots a_{ij} x^i y^j \dots \right)
\end{eqnarray}

To illustrate if we consider a polynomial with two variables,  $n=2$, we get :

\begin{eqnarray}
P_a(x,y) &=&   a_{00} + a_{10}x + a_{20}x^2 + a_{01}y + a_{11} xy + a_{21}x^2y +  a_{02}y^2  \\
              &+ & a_{12}xy^2 + a_{22}x^2y^2 \nonumber \label{poly0}
\end{eqnarray}

 If now we consider 2 vectors of polynomials of $m$ entries  $\boldsymbol{v}_1<P_a>$ and  $\boldsymbol{v}_2<P_b>$ the inner product  consists of the product 
 of every entries of the vectors one by one and a global reduction

 \begin{eqnarray}
 P_c = \sum_i^m \boldsymbol{v}_{1_i}<P_a>  \times  \boldsymbol{v}_{2_i}<P_b> \label{poly1}
\end{eqnarray}

where $times$ indicates the multiplication between two polynomials. We just note remember for the multiplication of the polynomial we get cross term. A cross term coefficient is simply the 
sum of several coefficients resulting from the multiplication. Per example, if we multiply two polynomial as defined in \ref{poly0}, the resulting coefficient $c_{12}xy^2$ will be given by :

\begin{eqnarray}
c_{12}xy^2 = \left(a_{12}b_{00}  +  a_{00}b_{12} + a_{10}b_{02}   + b_{10}a_{02}  +  a_{11}b_{01} + b_{12}a_{01}  \right)xy^2
\end{eqnarray}

These cross terms are not a problem for the cpu version but it is the bottleneck for the gpu version, (see gpu section). In term of  programming and parallelisation, 
for the cpu version the equation  \ref{poly1} can be reduced by a basic loop with a final reduction. These work can be easily done in OpenMP. Every OpenMP threads
will calculate several  polynomial multiplications and make a local reduction. At the end a final reduction is performed sequential by the main thread \footnote{In OpenMP reduction for class is not supported}.

\section*{Details of VLI\_CPU}

The coefficient of the polynomial are represented by a large integer, this number has been designed by the

The Very Large Integer (VLI)  models  basic arithmetic  operation for large integer on CPU (lib\_vli\_cpu) and GPU (lib\_vli\_gpu).  A VLI number is characterized by two parameters, a type and a size.
 The type is a \texttt{unsigned long int}, and the size varies from 2 to 8.
Thus the large integer can vary from 128 to 512 bits, by steps of 64 bits. It's implementation is static and determined during the compile time, it is not dynamic at all, as the well know GMP. Inside the memory every VLI number has contiguous data, 
the number is based on a basic array initialized by the type and size given during the compilation (template parameters).

The basic operations supported by VLI are the additions, the multiplications and the subtractions. For every family of operations, a subset of operations are available :
the subset of the additions are,
\begin{itemize}
\item VLI$<$Type, Size$>$ += VLI$<$Type, Size$>$
\item VLI$<$Type, Size$>$ += \texttt{long} or \texttt{int}
\item VLI$<$Type, Size+1$>$ = VLI$<$Type, Size$>$ + VLI$<$Type, Size$>$
\end{itemize}
the subset of the substractions are,
\begin{itemize}
\item VLI$<$Type, Size$>$ -= VLI$<$Type, Size$>$
\item VLI$<$Type, Size$>$ -= \texttt{long} or \texttt{int}
\end{itemize}
the subset of the  multiplications are,
\begin{itemize}
\item VLI$<$Type, Size$>$  $\times$= VLI$<$Type, Size$>$
\item VLI$<$Type, Size$>$  $\times$= \texttt{long} or \texttt{int}
\item VLI$<$Type, 2$\times$Size$>$ = VLI$<$Type, Size$>$ $\times$ VLI$<$Type, Size$>$
\end{itemize}

\subsection*{Why Assembly ?}

The algorithms of all these operations are based on the childhood knowledge but their realization are done in assembly because a few assembly instructions facilitate the implementation and the performances.
In programming, the basic operators of the addition, subtraction and multiplication between identical type gives the same type.

Per example for an addition, the propagation/subtraction of the  carry/borrow can be done by hand, but its performance will be low compare to a pure assembly version. 
The number of assembly instruction can be reduce by a factor 3 for the additions/substraction, and  2-order for the multiplication.

In classical programming, the addition is realized with bit shift to simulate the carry, added to cast operator between  \texttt{int} and \text{long int}. Oppositely, 
in assembly, the addition with the carry bit (CB) allows a simplified model without hand made carry propagation. \footnote{ The carry bit is a bit of the status register, as the overflow bit, 
it informs  the overflow of an operation and but it can be re-use to propagate the carry. }  It formalized in pseudo-assembly x86-64 by the following sequence:
\begin{table}[htdp]
\begin{center}
\begin{tabular}{ l c c c l c}
 \texttt{add }   &  b[0]  & a[0]    & $\Leftrightarrow$ & a[0]+=b[0]            & \tiny{0}  \\ 
 \texttt{adc }   &  b[1]   & a[1]   & $\Leftrightarrow$ & a[1]+=b[1] + CB  & \tiny{1}  \\  
\end{tabular}
\end{center}
\end{table}%
where the first columns indicates the  mnemonic instructions, \texttt{add} is a basic addition between two number of the same type, if the result is larger than the capacity of the number, the CB is set to one,
and re-use if the \texttt{adc} instruction is performed. In pure assembly, additional \texttt{move} instructions will be necessary to move the data from the global memory, to one of the 16 registers and reciprocally.
We will not detail because it exceeds the purpose of the present document.

The same techniques are performed for all additions/subtractions with respective mnemonic. For the multiplication, we use the mnemonic \texttt{mul} this instruction has the specificity to return the lower and the higher
par of the product into 2 separate registers (the classical operator $\times$ returns only the low part in C/C++/fortran \dots), combined with the additions and our childhood algorithm, we can construct a large multiplication.  
The x86-64 architecture allows 16 registers, as much as possible , we conserved the datas into the registers when we perform a long multiplication to reduce read/write actions to the minimum (1 cycle).
Additional optimizations are done like a full unrolling and assembly tips (xor, conditionals move \dots).

To conclude about assembly, the assembly language is the probably the most stable in term of portability $i386$ are supported since 20 years and 10 years for the x86 (80\% of the top 500 is x86 machine 20 \% power machine).
The VLI lib supported both x86 and power assembly, 99\% of the top 500 cluster.

\subsection*{About the implementation}

Generate by hand all these kernels in assembly can be painful and a source of error. Thus we designed a generator of assembly for every kernels, it generates all versions of a kernels during the compilation.   
It is based of the GNU inline assembly. Typically, the programmer gives a string between asm tag. We use the preprocessor and the package boost preprocessor package. 
(it allows repetition, branching, and more important the construction of string during the precompilation) to generate the strings of the associated kernels.
All kernels into lib\_vli\_cpu use this procedure except the extended multiplication.

\subsection*{Memory layout and SIMD implementation}

It exists at lest two data layout for this problem : the array of structure (AoS) and structure of array (SoA), both have their advantage and disadvantage. In our problem the AoS consists of saving the data in this order  : VLI - Series of VLI (polynomial coefficient)- series of polynomial (the vector), as following  in the memory (contiguous):

\begin{eqnarray}
 \underbrace{
 \underbrace{ \underbrace{a^0_{00} | a^1_{00} | a^2_{00}}_{VLI, a_{00} } || \underbrace{a^0_{10} | a^1_{10} | a^2_{10}}_{VLI, a_{10}} ||   \dots  }_{1^{st} \textrm{polynomial}} 
      |||    \underbrace{ \underbrace{b^0_{00} | a^1_{00} | b^2_{00}}_{VLI, b_{00} } || \underbrace{b^0_{10} | b^1_{10} | b^2_{10}}_{VLI, b_{10}} ||   \dots  }_{2^{nd} \textrm{polynomial} } }_{\textrm{Vector of polynomial}} \nonumber
\end{eqnarray}

The second possibility; SoA interleaves the coefficient of the VLI and the polynomial, they become nested:

\begin{eqnarray}
 \underbrace{
 \underbrace{a^0_{00} | a^0_{10} | \dots || a^1_{00} | a^1_{10} | \dots  }_{1st  \textrm{ nested VLI/Polynomial} } |||  \underbrace{b^0_{00} | b^0_{10} | \dots || b^1_{00} | b^1_{10} | \dots  }_{2nd  \textrm{ nested VLI/Polynomial} }
 }_{\textrm{Vector of polynomial}} \nonumber
\end{eqnarray}

The choice of the memory layout is important because it allows or not the usage of the SIMD instructions. We remember  the SIMD instructions are 128 to 256 bits vector register ($ymm_i$) allows 4 to 8 \texttt{int}  operations simultaneously.
A number of constraints force us to work with the half of the register because we have to manage the carry bit and the extended multiplication by hand. In the previous sections we presented an addition of 128 bits VLI, it took two operations
without \texttt{move}, if with consider the same operation in SIMD (256 bits registers), we will get :

\begin{center}
\begin{tabular}{ c  c c c c c c c c} 
 $ymm_0$   register      &    $a^0_{00}$  & $a^0_{10}$ & $a^0_{20}$  &  $a^0_{30}$ & $a^0_{40}$  &   $a^0_{50}$ &  $a^0_{60}$   & $a^0_{70}$  \\
  shift left    &    $a^0_{10}$  & 0 & $a^0_{30}$  &  0  & $a^0_{50}$  &   0  &  $a^0_{70}$   & 0  \\
  \\
 $ymm_1$   register      &    $b^0_{00}$  & $b^0_{10}$ & $b^0_{20}$  &  $b^0_{30}$ & $b^0_{40}$  &   $b^0_{50}$ &  $b^0_{60}$   & $b^0_{70}$  \\
  shift left     &    $b^0_{10}$  & 0 & $b^0_{30}$  &  0  & $b^0_{50}$  &   0  &  $b^0_{70}$   & 0  \\
\\
$ymm_0 += ymm_1$           &    $c^0_{10}$  & CB & $c^0_{30}$  &  CB  & $c^0_{50}$  &  CB  &  $c^0_{70}$   & CB  \\
shift left  CB register        & CB & 0 & CB & 0 & CB & 0 & CB & 0 \\
\\
$ymm_2$   register      &    $a^1_{00}$  & $a^1_{10}$ & $a^1_{20}$  &  $a^1_{30}$ & $a^1_{40}$  &   $a^1_{50}$ &  $a^1_{60}$   & $a^1_{70}$  \\
shift left     &    $a^1_{10}$  & 0 & $a^1_{30}$  &  0  & $a^1_{50}$  &   0  &  $a^1_{70}$   & 0  \\
 \\
 $ymm_3$   register      &    $b^1_{00}$  & $b^1_{10}$ & $b^1_{20}$  &  $b^1_{30}$ & $b^1_{40}$  &   $b^1_{50}$ &  $b^1_{60}$   & $b^1_{70}$  \\
  shift left   &    $b^1_{10}$  & 0 & $b^1_{30}$  &  0  & $b^1_{50}$  &   0  &  $b^1_{70}$   & 0  \\
\\
$ymm_2 += ymm_3$                                  &    $c^1_{10}$  & CB & $c^1_{30}$  &  CB  & $c^1_{50}$  &  CB  &  $c^1_{70}$   & -  \\
$ymm_2 +=  \textrm{CB register} $            &    $c^1_{10} $  & CB & $c^1_{30}$  &  CB  & $c^1_{50}$  &  CB  &  $c^1_{70}$   & -  \\
 shift right logical          &  0     &    $c^1_{10} $  & 0 & $c^1_{30}$  &  0 & $c^1_{50}$  & 0  &  $c^1_{70}$    \\

\end{tabular}
\end{center}

This operation must be repeat a second time to get the second part of the addition and two time to get the128 bits += operations. At the end we perform 24 additions and 48  shift operations for eight VLI$<4*32, \texttt{\textrm{uint}}>$ += VLI$<4*32, \texttt{\textrm{uint}}>$ operations. Our primary solution (under 64 bits mode,  VLI$<2*64, \texttt{\textrm{uint}}>$ += VLI$<2*64, \texttt{\textrm{uint}}>$) needs 16 operations plus operations and 0 shift.

A second possibility is to work under 31 bits base mode, thus we can make an addition
 VLI$<4*31, \texttt{\textrm{uint}}>$ += VLI$<4*31, \texttt{\textrm{uint}}>$  in  12 additions and 1 shift it will be faster. Nevertheless on monder CPU (Sandybridge), we have 3 APU during the addition of the polynomial as the coefficient are fully independent 
 we can not exclude that the execution of  the addition of the coefficients will be execute in parallel over the APU.

To conclude, if you know consider the multiplication under SIMD execution,

\end{document}

### 2012-07-04 doc/paper_vli/vli.tex
doc vli

@@
 
-To conclude about assembly, the assembly language is the probably the most stable in term of portability $i386$ are supported since 20 years and 10 years for the x86 (80\% of the top 500 is x86 machine 20 \% power machine).
+To conclude about assembly, the assembly language is the probably the most stable in term of portability $i386$ are supported since 20 years and 10 years for the x86 (79\% of the top 500 is x86 machine 20 \% power machine).
 The VLI lib supported both x86 and power assembly, 99\% of the top 500 cluster.
@@
 In the previous sections we presented an addition of 128 bits VLI, it took two operations
-without \texttt{move}, if with consider the same operation in SIMD (256 bits registers), we will get :
+without \texttt{move}.
+The same operation in SIMD (256 bits registers) necessitates a few adaptations as work in base 31 for use the 32 bits as carry, but or number will be limited to 124 bits.
+Thus  we have :
 
@@
  $ymm_0$   register      &    $a^0_{00}$  & $a^0_{10}$ & $a^0_{20}$  &  $a^0_{30}$ & $a^0_{40}$  &   $a^0_{50}$ &  $a^0_{60}$   & $a^0_{70}$  \\
-  shift left    &    $a^0_{10}$  & 0 & $a^0_{30}$  &  0  & $a^0_{50}$  &   0  &  $a^0_{70}$   & 0  \\
-  \\
  $ymm_1$   register      &    $b^0_{00}$  & $b^0_{10}$ & $b^0_{20}$  &  $b^0_{30}$ & $b^0_{40}$  &   $b^0_{50}$ &  $b^0_{60}$   & $b^0_{70}$  \\
-  shift left     &    $b^0_{10}$  & 0 & $b^0_{30}$  &  0  & $b^0_{50}$  &   0  &  $b^0_{70}$   & 0  \\
-\\
-$ymm_0 += ymm_1$           &    $c^0_{10}$  & CB & $c^0_{30}$  &  CB  & $c^0_{50}$  &  CB  &  $c^0_{70}$   & CB  \\
-shift left  CB register        & CB & 0 & CB & 0 & CB & 0 & CB & 0 \\
-\\
+$ymm_0 += ymm_1$    &    $c^0_{00}$  & $c^0_{10}$ & $c^0_{20}$  &  $c^0_{30}$  & $c^0_{40}$  &   $c^0_{50}$  &  $c^0_{60}$   & $c^0_{70}$  \\
+shift right  get CB        &     CB                 & CB                & CB                 & CB                   & CB               &  CB                   & CB                   & CB \\
 $ymm_2$   register      &    $a^1_{00}$  & $a^1_{10}$ & $a^1_{20}$  &  $a^1_{30}$ & $a^1_{40}$  &   $a^1_{50}$ &  $a^1_{60}$   & $a^1_{70}$  \\
-shift left     &    $a^1_{10}$  & 0 & $a^1_{30}$  &  0  & $a^1_{50}$  &   0  &  $a^1_{70}$   & 0  \\
- \\
  $ymm_3$   register      &    $b^1_{00}$  & $b^1_{10}$ & $b^1_{20}$  &  $b^1_{30}$ & $b^1_{40}$  &   $b^1_{50}$ &  $b^1_{60}$   & $b^1_{70}$  \\
-  shift left   &    $b^1_{10}$  & 0 & $b^1_{30}$  &  0  & $b^1_{50}$  &   0  &  $b^1_{70}$   & 0  \\
-\\
-$ymm_2 += ymm_3$                                  &    $c^1_{10}$  & CB & $c^1_{30}$  &  CB  & $c^1_{50}$  &  CB  &  $c^1_{70}$   & -  \\
-$ymm_2 +=  \textrm{CB register} $            &    $c^1_{10} $  & CB & $c^1_{30}$  &  CB  & $c^1_{50}$  &  CB  &  $c^1_{70}$   & -  \\
- shift right logical          &  0     &    $c^1_{10} $  & 0 & $c^1_{30}$  &  0 & $c^1_{50}$  & 0  &  $c^1_{70}$    \\
-
+$ymm_2 += ymm_3$     & $c^1_{00}$  & $c^1_{10}$ & $c^1_{20}$  &  $c^1_{30}$  & $c^1_{40}$  &   $c^1_{50}$  &  $c^1_{60}$   & $c^1_{70}$  \\                            
+$ymm_2 +=  \textrm{CB } $ & $c^1_{00}$  & $c^1_{10}$ & $c^1_{20}$  &  $c^1_{30}$  & $c^1_{40}$  &   $c^1_{50}$  &  $c^1_{60}$   & $c^1_{70}$  \\                            
 \end{tabular}
@@
 
-This operation must be repeat a second time to get the second part of the addition and two time to get the128 bits += operations.
-At the end we perform 24 additions and 48  shift operations for eight VLI$<4*32, \texttt{\textrm{uint}}>$ += VLI$<4*32, \texttt{\textrm{uint}}>$ operations.
-Our primary solution (under 64 bits mode,  VLI$<2*64, \texttt{\textrm{uint}}>$ += VLI$<2*64, \texttt{\textrm{uint}}>$) needs 16 operations plus operations and 0 shift.
+This operation must be repeat two more times to get the 124 bits addition.
+At the end we perform 9 additions and 3  shift operations for eight VLI$<4*31, \texttt{\textrm{uint}}>$ += VLI$<4*31, \texttt{\textrm{uint}}>$ operations.
+ Comparing to the original solution we saved a factor two,  nevertheless on modern CPU (Sandybridge), we have 3 APU  we can not exclude that the execution of  the addition of the polynomial coefficients will be execute in parallel over the APU.
 
-A second possibility is to work under 31 bits base mode, thus we can make an addition
- VLI$<4*31, \texttt{\textrm{uint}}>$ += VLI$<4*31, \texttt{\textrm{uint}}>$  in  12 additions and 1 shift it will be faster.
-Nevertheless on monder CPU (Sandybridge), we have 3 APU during the addition of the polynomial as the coefficient are fully independent
- we can not exclude that the execution of  the addition of the coefficients will be execute in parallel over the APU.
+Below, we should consider the extended  multiplication using SIMD instructions.
+For this multiplication   VLI$<4*64, \texttt{\textrm{ulint}}>$  = VLI$<2*64, \texttt{\textrm{ulint}}>$ $\times$ VLI$<2*64, \texttt{\textrm{ulint}}>$) using the
+childhood  algorithm we need 4 multiplications and 3 additions so 32 multiplications and 24 additions to perform 8 vli multiplications.
+For the SIMD version we need 16 multiplications and 29 additions, 16 copies for the ouput multiplications
+associated to 16 right shift  to extract the higher part on the multiplication and16 mask operator to get the lower part.
+And additional shift to get the carry bit.
+Time for both implementation will be similar.
 
-To conclude, if you know consider the multiplication under SIMD execution,
+As the SIMD implementation does not give enough argument to focus on it we have privileged  the AoS and classic assembly instructions.
+To conclude the new coprocessor of Intel named Mic could reconsider our choice on SIMD
+because this processor support Carry Bit, and extract lower and higher part of the multiplication natively. It also allow selected operations on the register by a judicious system of mask.
 

### 2012-07-16 doc/paper_vli/vli.tex
VLI doc

@@
 
-To illustrate if we consider a polynomial with two variables,  $n=2$, we get :
+ If we consider a polynomial with two variables,  $n=2$, we will get :
 
@@
 
-where $times$ indicates the multiplication between two polynomials. We just note remember for the multiplication of the polynomial we get cross term. A cross term coefficient is simply the 
-sum of several coefficients resulting from the multiplication. Per example, if we multiply two polynomial as defined in \ref{poly0}, the resulting coefficient $c_{12}xy^2$ will be given by :
+where $\times$ indicates the multiplication between two polynomials.
+We should note, every coefficients may be not uniques.
+They can be the sum of several coefficients resulting from the multiplications.
+These coefficients are named cross term. Per example, if we multiply two polynomial as defined in \ref{poly0}, the resulting coefficient $c_{12}xy^2$ will be given by :
 
@@
 
-These cross terms are not a problem for the cpu version but it is the bottleneck for the gpu version, (see gpu section). In term of  programming and parallelisation, 
+These cross terms are not a problem for the cpu version but it is the bottleneck for the gpu version, (see gpu section). In term of  programming and parallelization, 
 for the cpu version the equation  \ref{poly1} can be reduced by a basic loop with a final reduction. These work can be easily done in OpenMP. Every OpenMP threads
-will calculate several  polynomial multiplications and make a local reduction.
-At the end a final reduction is performed sequential by the main thread \footnote{In OpenMP reduction for class is not supported}.
+will calculate several  polynomial multiplications and make a local reduction. At the end, a final reduction is performed  by the main thread \footnote{In OpenMP reduction for class is not supported}.
 
@@
 
-The coefficient of the polynomial are represented by a large integer, this number has been designed by the
-
-The Very Large Integer (VLI)  models  basic arithmetic  operation for large integer on CPU (lib\_vli\_cpu) and GPU (lib\_vli\_gpu).  A VLI number is characterized by two parameters, a type and a size.
- The type is a \texttt{unsigned long int}, and the size varies from 2 to 8.
-Thus the large integer can vary from 128 to 512 bits, by steps of 64 bits.
-It's implementation is static and determined during the compile time, it is not dynamic at all, as the well know GMP.
-Inside the memory every VLI number has contiguous data,
-the number is based on a basic array initialized by the type and size given during the compilation (template parameters).
+The coefficients of the polynomial are represented by a large integer call VLI number.  A VLI number is characterized by two parameters, a Type and a Size.
+ The type is an \texttt{unsigned long int},  the size varies from 2 to 8.
+Thus, the large integer can vary from 128 to 512 bits, by steps of 64 bits.
+It's implementation is static and determined during the compilation time.
+ It is not dynamic at all, as the well know GMP. Inside the memory, the VLI numbers are  contiguous.
 
 The basic operations supported by VLI are the additions, the multiplications and the subtractions. For every family of operations, a subset of operations are available :
-the subset of the additions are,
 \begin{itemize}
@@
 
-The algorithms of all these operations are based on the childhood knowledge but their realization are done in assembly because a few assembly instructions facilitate the implementation and the performances.
+The algorithms of all these operations are based on the childhood knowledge but their realization are done in assembly, because, a few assembly instructions facilitate the implementation and boost the performance.
 In programming, the basic operators of the addition, subtraction and multiplication between identical type gives the same type.
@@
 Per example for an addition, the propagation/subtraction of the  carry/borrow can be done by hand, but its performance will be low compare to a pure assembly version. 
-The number of assembly instruction can be reduce by a factor 3 for the additions/substraction, and  2-order for the multiplication.
+The number of assembly instruction can be reduce by a factor 3 for the additions/subtraction, and  2-order for the multiplication.
 
 In classical programming, the addition is realized with bit shift to simulate the carry, added to cast operator between  \texttt{int} and \text{long int}. Oppositely, 
-in assembly, the addition with the carry bit (CB) allows a simplified model without hand made carry propagation. \footnote{ The carry bit is a bit of the status register, as the overflow bit, 
-it informs  the overflow of an operation and but it can be re-use to propagate the carry. }  It formalized in pseudo-assembly x86-64 by the following sequence:
+in assembly, the addition with the carry bit (CB) allows a simplified model without  carry bit propagation by hand made. \footnote{ The carry bit is a bit of the status register, as the overflow bit, 
+it informs  the overflow of an operation, but it can be re-use to propagate the carry. }  It formalized in pseudo-assembly x86-64 by the following sequence:
 \begin{table}[htdp]
@@
 par of the product into 2 separate registers (the classical operator $\times$ returns only the low part in C/C++/fortran \dots), combined with the additions and our childhood algorithm, we can construct a large multiplication.
-The x86-64 architecture allows 16 registers, as much as possible , we conserved the datas into the registers when we perform a long multiplication to reduce read/write actions to the minimum (1 cycle).
-Additional optimizations are done like a full unrolling and assembly tips (xor, conditionals move \dots).
-
-To conclude about assembly, the assembly language is the probably the most stable in term of portability $i386$ are supported since 20 years and 10 years for the x86 (79\% of the top 500 is x86 machine 20 \% power machine).
-The VLI lib supported both x86 and power assembly, 99\% of the top 500 cluster.
+The x86-64 architecture allows 16 registers, as much as possible , we conserved the datas into the registers when we perform a long multiplication to reduce read/write actions to the minimum (1 cycle for a register).
+Additional optimizations are done like a full unrolling and assembly tips (xor, conditionals move \dots).  The VLI lib supported  x86-64 and Power assembly,
 
@@
 
-Generate by hand all these kernels in assembly can be painful and a source of error.
-Thus we designed a generator of assembly for every kernels, it generates all versions of a kernels during the compilation.
-It is based of the GNU inline assembly. Typically, the programmer gives a string between asm tag. We use the preprocessor and the package boost preprocessor package. 
-(it allows repetition, branching, and more important the construction of string during the precompilation) to generate the strings of the associated kernels.
-All kernels into lib\_vli\_cpu use this procedure except the extended multiplication.
+We intro=duce the assembly using the GNU inline assembly, it consists of a string between ASM tag.
+Generate by hand all these  string (kernels) in assembly can be painful and a source of error (15 000 lines). 
+Thus, we designed a generator of assembly. It generates all versions of a kernels during the compilation.   
+We use the preprocessor and the package boost preprocessor package, it allows repetition, branching, and more
+ important the construction of string during the pre-compilation.  All kernels use this procedure except the extended multiplication for the cpu.
 
@@
 
-It exists at lest two data layout for this problem : the array of structure (AoS) and structure of array (SoA), both have their advantage and disadvantage.
-In our problem the AoS consists of saving the data in this order  : VLI - Series of VLI (polynomial coefficient)- series of polynomial (the vector), as following  in the memory (contiguous):
+It exists at least two data layout for this problem : the array of structure (AoS) and structure of array (SoA), both have their advantage and disadvantage.
+In our problem the AoS consists of saving the data in this order  : VLI - Series of VLI (polynomial coefficient)- series of polynomial (the vector), it can be represent in the memory (contiguous) as following:
 
@@
 
-The second possibility; SoA interleaves the coefficient of the VLI and the polynomial, they become nested:
+The second possibility; SoA interleaves the coefficient of the VLI and the polynomial. They become nested:
 
@@
 The choice of the memory layout is important because it allows or not the usage of the SIMD instructions.
-We remember  the SIMD instructions are 128 to 256 bits vector register ($ymm_i$) allows 4 to 8 \texttt{int}  operations simultaneously.
+We remember,  the SIMD instructions are 128 to 256 bits vector register ($xmm_i$ or $ymm_i$) allows 4 to 8 \texttt{int}  operations simultaneously.
 A number of constraints force us to work with the half of the register because we have to manage the carry bit and the extended multiplication by hand.
@@
 without \texttt{move}.
-The same operation in SIMD (256 bits registers) necessitates a few adaptations as work in base 31 for use the 32 bits as carry, but or number will be limited to 124 bits.
-Thus  we have :
+The same operation in SIMD (256 bits registers) necessitates a few adaptations as work in base 31 (the 32 bits becomes the carry bit), but or number will be limited to 124 bits.
+We get :
 
@@
 \begin{tabular}{ c  c c c c c c c c} 
+\\
  $ymm_0$   register      &    $a^0_{00}$  & $a^0_{10}$ & $a^0_{20}$  &  $a^0_{30}$ & $a^0_{40}$  &   $a^0_{50}$ &  $a^0_{60}$   & $a^0_{70}$  \\
@@
 $ymm_2 +=  \textrm{CB } $ & $c^1_{00}$  & $c^1_{10}$ & $c^1_{20}$  &  $c^1_{30}$  & $c^1_{40}$  &   $c^1_{50}$  &  $c^1_{60}$   & $c^1_{70}$  \\                            
+\\
 \end{tabular}
@@
 
-Below, we should consider the extended  multiplication using SIMD instructions.
-For this multiplication   VLI$<4*64, \texttt{\textrm{ulint}}>$  = VLI$<2*64, \texttt{\textrm{ulint}}>$ $\times$ VLI$<2*64, \texttt{\textrm{ulint}}>$) using the
+Beyond, we should consider the extended  multiplication using SIMD instructions.
+For this multiplication of unsigned number   VLI$<4*64, \texttt{\textrm{ulint}}>$  = VLI$<2*64, \texttt{\textrm{ulint}}>$ $\times$ VLI$<2*64, \texttt{\textrm{ulint}}>$) using the
 childhood  algorithm we need 4 multiplications and 3 additions so 32 multiplications and 24 additions to perform 8 vli multiplications.
@@
 Time for both implementation will be similar.
+We should remember the treatment of the sign,
+ a long multiplication can necessitate branching (cite art of assembly).
+The branching can be remove but it  becomes painful and useless for number larger than 256 bits.
+A SIMD version necessitates no branching but its implementation will
+definitively discredit this technology.
 
-As the SIMD implementation does not give enough argument to focus on it we have privileged  the AoS and classic assembly instructions.
+As the SIMD implementation does not give enough argument to focus on it, we have privileged  the AoS and classic assembly instructions.
 To conclude the new coprocessor of Intel named Mic could reconsider our choice on SIMD
-because this processor support Carry Bit, and extract lower and higher part of the multiplication natively. It also allow selected operations on the register by a judicious system of mask.
+because this processor supports CB, and extraction of the  lower and higher part of the multiplication. It allows selected operations on the registers by a judicious system of mask.
+
+NOTE :  for Andreas and Matthias :
+Max Nvidia just coded $2n <64 bits> =  2n <64 bits>* 2n <64 bits> $ doing this, you get the sign naturally, but the it costs .
+The version  $2n <64 bits> =  n <64 bits> * n <64 bits>$ with branching divides by 1.5 the times of Max.
 

### 2012-07-16 doc/paper_vli/vli.tex
vli doc

@@
 We will first remember the basics of the inner product of a polynomial,
-and the implementation of the cpu version  The second par will be on the gpu where the inner product and long arithmetic are nested.
+and the implementation of the cpu version  The second part will be on the gpu implementation.
 
-\section*{Inner product}
+\section*{Polynomial and Inner product}
 
-The main object of the library is performed an inner product of vector of polynomials. The polynomials $P$ are defined like :
+The library performed inner product of vector of polynomials where polynomials are defined like :
 \begin{eqnarray}
@@
 
- If we consider a polynomial with two variables,  $n=2$, we will get :
+ If we consider a two  polynomial variables where $n=2$, we  have :
 
@@
 
- If now we consider 2 vectors of polynomials of $m$ entries  $\boldsymbol{v}_1<P_a>$ and  $\boldsymbol{v}_2<P_b>$ the inner product  consists of the product 
- of every entries of the vectors one by one and a global reduction
+ If now we consider 2 vectors of polynomials of $m$ entries  $\boldsymbol{v}_1<P_a>$ and  $\boldsymbol{v}_2<P_b>$. The inner product  consists of the product 
+ of every entries  associated to a global reduction.
 
@@
 where $\times$ indicates the multiplication between two polynomials.
-We should note, every coefficients may be not uniques.
-They can be the sum of several coefficients resulting from the multiplications.
+We  note, a coefficient may be not unique.
+It can be the sum of several coefficients resulting from the sum of several multiplications.
 These coefficients are named cross term. Per example, if we multiply two polynomial as defined in \ref{poly0}, the resulting coefficient $c_{12}xy^2$ will be given by :
@@
 
-These cross terms are not a problem for the cpu version but it is the bottleneck for the gpu version, (see gpu section). In term of  programming and parallelization, 
+These cross terms are not a problem for the cpu version but it is the bottleneck for the gpu version  (see gpu section). In term of  programming and parallelization, 
 for the cpu version the equation  \ref{poly1} can be reduced by a basic loop with a final reduction. These work can be easily done in OpenMP. Every OpenMP threads
-will calculate several  polynomial multiplications and make a local reduction. At the end, a final reduction is performed  by the main thread \footnote{In OpenMP reduction for class is not supported}.
+calculate several  polynomial multiplications and make a local reduction. At the end, a final reduction is performed  by the main thread \footnote{In OpenMP reduction for class is not supported}.
 
@@
 Thus, the large integer can vary from 128 to 512 bits, by steps of 64 bits.
-It's implementation is static and determined during the compilation time.
+Its implementation is static and determined during the compilation time.
  It is not dynamic at all, as the well know GMP. Inside the memory, the VLI numbers are  contiguous.
@@
 \end{itemize}
-
+The algorithms of all these operations are based on the childhood knowledge but their realization are done in assembly, because, a few assembly instructions facilitate the implementation and boost the performance.
 \subsection*{Why Assembly ?}
 
-The algorithms of all these operations are based on the childhood knowledge but their realization are done in assembly, because, a few assembly instructions facilitate the implementation and boost the performance.
-In programming, the basic operators of the addition, subtraction and multiplication between identical type gives the same type.
-
-Per example for an addition, the propagation/subtraction of the  carry/borrow can be done by hand, but its performance will be low compare to a pure assembly version. 
+In programming, the basic operators of the addition, subtraction and multiplication between identical type give the same type. 
+Per example, if we need an long addition or long subtraction, the propagation of the  carry/borrow can be done by hand, but the performance will be low compare to a pure assembly version. 
 The number of assembly instruction can be reduce by a factor 3 for the additions/subtraction, and  2-order for the multiplication.
 
-In classical programming, the addition is realized with bit shift to simulate the carry, added to cast operator between  \texttt{int} and \text{long int}. Oppositely, 
-in assembly, the addition with the carry bit (CB) allows a simplified model without  carry bit propagation by hand made. \footnote{ The carry bit is a bit of the status register, as the overflow bit, 
-it informs  the overflow of an operation, but it can be re-use to propagate the carry. }  It formalized in pseudo-assembly x86-64 by the following sequence:
+In assembly, the addition with the carry bit (CB) allows a simplified model without  carry bit propagation by hand\footnote{ The carry bit is a bit of the status register, as the overflow bit, 
+it informs  the overflow of an operation, but it can be re-use to propagate the carry. }.  It is formalized in pseudo-assembly x86-64 by the following sequence:
 \begin{table}[htdp]
@@
 par of the product into 2 separate registers (the classical operator $\times$ returns only the low part in C/C++/fortran \dots), combined with the additions and our childhood algorithm, we can construct a large multiplication.
-The x86-64 architecture allows 16 registers, as much as possible , we conserved the datas into the registers when we perform a long multiplication to reduce read/write actions to the minimum (1 cycle for a register).
-Additional optimizations are done like a full unrolling and assembly tips (xor, conditionals move \dots).  The VLI lib supported  x86-64 and Power assembly,
 
@@
 
+The x86-64 architecture allows 16 registers (32 for power), as much as possible , we conserved the datas into the registers when we perform a long multiplication to reduce read/write actions to the minimum (1 cycle for a register).
+Additional optimizations are done like unrolling and assembly tips (xor,  \dots).  The VLI lib supported  x86-64 and Power assembly,
+
 \subsection*{Memory layout and SIMD implementation}
@@
 It exists at least two data layout for this problem : the array of structure (AoS) and structure of array (SoA), both have their advantage and disadvantage.
-In our problem the AoS consists of saving the data in this order  : VLI - Series of VLI (polynomial coefficient)- series of polynomial (the vector), it can be represent in the memory (contiguous) as following:
+In our problem the AoS consists of saving the data in this order  : VLI - Series of VLI (polynomial coefficient)- series of polynomial (the vector), it can be represented in the memory (contiguous) by:
 
@@
 
-The second possibility; SoA interleaves the coefficient of the VLI and the polynomial. They become nested:
+The second possibility; SoA interleaves the coefficient of the VLI and the polynomial:
 
@@
 The choice of the memory layout is important because it allows or not the usage of the SIMD instructions.
-We remember,  the SIMD instructions are 128 to 256 bits vector register ($xmm_i$ or $ymm_i$) allows 4 to 8 \texttt{int}  operations simultaneously.
-A number of constraints force us to work with the half of the register because we have to manage the carry bit and the extended multiplication by hand.
+The SIMD instructions are 128 to 256 bits vector register ($xmm_i$ or $ymm_i$) allows 4 to 8 \texttt{int}  operations simultaneously.
+A number of constraints force us to work under 31 bits base because we have to manage the carry bit and the extended multiplication by hand.
 In the previous sections we presented an addition of 128 bits VLI, it took two operations
-without \texttt{move}.
-The same operation in SIMD (256 bits registers) necessitates a few adaptations as work in base 31 (the 32 bits becomes the carry bit), but or number will be limited to 124 bits.
-We get :
+without \texttt{move}.   The same operation in SIMD (256 bits registers) necessitates a few adaptations as we work now in base 31 :
 
@@
 For the SIMD version we need 16 multiplications and 29 additions, 16 copies for the ouput multiplications
-associated to 16 right shift  to extract the higher part on the multiplication and16 mask operator to get the lower part.
+associated to 16 right shift  to extract the higher part on the multiplication and 16 mask operator to get the lower part.
 And additional shift to get the carry bit.
-Time for both implementation will be similar.
+Execution time for both implementation will be similar.
 We should remember the treatment of the sign,
  a long multiplication can necessitate branching (cite art of assembly).
-The branching can be remove but it  becomes painful and useless for number larger than 256 bits.
+The branching can be remove but it  becomes painful and useless for number larger than 128 bits.
 A SIMD version necessitates no branching but its implementation will
-definitively discredit this technology.
+definitively discredit this choice.
 
@@
 Max Nvidia just coded $2n <64 bits> =  2n <64 bits>* 2n <64 bits> $ doing this, you get the sign naturally, but the it costs .
-The version  $2n <64 bits> =  n <64 bits> * n <64 bits>$ with branching divides by 1.5 the times of Max.
+The version  $2n <64 bits> =  n <64 bits> * n <64 bits>$ with branching divides by 1.5 the time of Max.
 

### 2012-07-17 doc/paper_vli/vli.tex
doc vli

@@
 
-where $\times$ indicates the multiplication between two polynomials.
-We  note, a coefficient may be not unique.
-It can be the sum of several coefficients resulting from the sum of several multiplications.
+where $\times$ indicates the multiplication between two polynomials.   Note, a result coefficient may be not unique. It can be the sum of several coefficients resulting from several multiplications.
 These coefficients are named cross term. Per example, if we multiply two polynomial as defined in \ref{poly0}, the resulting coefficient $c_{12}xy^2$ will be given by :
@@
 In programming, the basic operators of the addition, subtraction and multiplication between identical type give the same type. 
-Per example, if we need an long addition or long subtraction, the propagation of the  carry/borrow can be done by hand, but the performance will be low compare to a pure assembly version. 
+Per example, if we need a long addition or a long subtraction, the propagation of the  carry/borrow can be done by hand, but the performance will be low compare to a pure assembly version. 
 The number of assembly instruction can be reduce by a factor 3 for the additions/subtraction, and  2-order for the multiplication.
@@
 For the multiplication, we use the mnemonic \texttt{mul} this instruction has the specificity to return the lower and the higher
-par of the product into 2 separate registers (the classical operator $\times$ returns only the low part in C/C++/fortran \dots), combined with the additions and our childhood algorithm, we can construct a large multiplication.
+par of the product into  separate registers (the classical operator $\times$ returns only the low part in C/C++/fortran \dots), combined with the additions and our childhood algorithm, we can construct a large multiplication.
 
@@
 
-We intro=duce the assembly using the GNU inline assembly, it consists of a string between ASM tag.
-Generate by hand all these  string (kernels) in assembly can be painful and a source of error (15 000 lines). 
-Thus, we designed a generator of assembly. It generates all versions of a kernels during the compilation.   
-We use the preprocessor and the package boost preprocessor package, it allows repetition, branching, and more
+We code the assembly using the GNU inline assembly, it consists of a string between ASM tag.
+Generate by hand all these  string (kernels) in assembly can be painful and a source of error ($\sim$10 000 lines). 
+We designed a generator of assembly. It create all versions of a kernels during the compilation.   
+We use the preprocessor and the  boost preprocessor package. It allows repetition, branching, and more
  important the construction of string during the pre-compilation.  All kernels use this procedure except the extended multiplication for the cpu.
@@
 The x86-64 architecture allows 16 registers (32 for power), as much as possible , we conserved the datas into the registers when we perform a long multiplication to reduce read/write actions to the minimum (1 cycle for a register).
-Additional optimizations are done like unrolling and assembly tips (xor,  \dots).  The VLI lib supported  x86-64 and Power assembly,
+Additional optimizations are done like unrolling and assembly tips (xor,  \dots).  The VLI lib supported  x86-64 and Power assembly.
 
@@
 A number of constraints force us to work under 31 bits base because we have to manage the carry bit and the extended multiplication by hand.
-In the previous sections we presented an addition of 128 bits VLI, it took two operations
+In the previous sections we presented an addition of 128 bits VLI.
+It took two operations
 without \texttt{move}.   The same operation in SIMD (256 bits registers) necessitates a few adaptations as we work now in base 31 :
@@
 This operation must be repeat two more times to get the 124 bits addition.
-At the end we perform 9 additions and 3  shift operations for eight VLI$<4*31, \texttt{\textrm{uint}}>$ += VLI$<4*31, \texttt{\textrm{uint}}>$ operations.
+At the end, we perform 9 additions and 3  shift operations for eight VLI$<4*31, \texttt{\textrm{uint}}>$ += VLI$<4*31, \texttt{\textrm{uint}}>$ operations.
  Comparing to the original solution we saved a factor two,  nevertheless on modern CPU (Sandybridge), we have 3 APU  we can not exclude that the execution of  the addition of the polynomial coefficients will be execute in parallel over the APU.
@@
 For this multiplication of unsigned number   VLI$<4*64, \texttt{\textrm{ulint}}>$  = VLI$<2*64, \texttt{\textrm{ulint}}>$ $\times$ VLI$<2*64, \texttt{\textrm{ulint}}>$) using the
-childhood  algorithm we need 4 multiplications and 3 additions so 32 multiplications and 24 additions to perform 8 vli multiplications.
+childhood  algorithm, we need 4 multiplications and 3 additions so 32 multiplications and 24 additions to perform 8 vli multiplications.
 For the SIMD version we need 16 multiplications and 29 additions, 16 copies for the ouput multiplications
@@
 Execution time for both implementation will be similar.
-We should remember the treatment of the sign,
- a long multiplication can necessitate branching (cite art of assembly).
+Nevertheless, we should remember the treatment of the sign,  a long multiplication can necessitate branching (cite art of assembly).
 The branching can be remove but it  becomes painful and useless for number larger than 128 bits.
-A SIMD version necessitates no branching but its implementation will
-definitively discredit this choice.
+A SIMD version necessitates no branching but its implementation will definitively discredit this choice.
 

### 2012-11-11 doc/paper_vli/vli.tex
doc vli

@@
-\documentclass[11pt]{amsart}
+\documentclass[oribibl]{llncs2e/llncs}
 \usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
@@
 %\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
+\usepackage{amssymb,amsmath}
 \usepackage{graphicx}
@@
 
-\title{Brief descriptions of the VLI library}
-\author{Timothee/Andreas or TAinmdortehaese, that is the question}
-%\date{}                                           % Activate to display a given date or no date
+\graphicspath{{WarpTask/figures/}{Res/figures/}} %do not forget the / at the end
+%\include{{llncs2e}}
 
+\title{Large integer arithmetic and polynomial inner product}
+\author{Timoth\'ee Ewart\inst{1}, Andreas Hein$^2$, Matthias Troyer\inst{2} and Thierry Giamarchi\inst{1}}
+
+\institute{Universit\'e de Gen\`eve, \email{timothee.ewart@gmail.com} \thanks{Thank you : Maxim Milakov, Peter Messmer  - NVIDIA, and  Williams Sawyer, Gilles Fourestey - CSCS}  \and Eidgen\"ossische Technische Hochschule Z\"urich }
+
+%\date{}                                           % Activate to display a given date or no dater
 \begin{document}
@@
 
-The main purpose of the VLI lib is performed hybrid (CPU/GPU) inner product of polynomial from 1 to 4 variables with large integer coefficient.
-We will first remember the basics of the inner product of a polynomial,
-and the implementation of the cpu version  The second part will be on the gpu implementation.
+The main purpose of the VLI lib is performed large integer arithmetic based on meta-programming and  hybrid (CPU/GPU) inner product of polynomial up to order 14 with 1 to 4 variables with large integer coefficient 128 to 256 bits.
+We remember the basics of the inner product of a polynomial,  and the implementation of the large integer library for the cpu version  The second part will be on the gpu implementation of the inner product.
+All these algorithms are based on brut force attacks, as their size should not justify a FFT approach \cite{Emeliyanenko:2009}.
 
-\section*{Polynomial and Inner product}
+\section*{Polynomial and Inner product definitions}
 
-The library performed inner product of vector of polynomials where polynomials are defined like :
+The library performed inner product of vector of polynomials from one to four variables where polynomials are defined like (four variables) :
 \begin{eqnarray}
-P_a(x,y,\dots) = \sum_i^n \left( \sum_j^n \dots a_{ij} x^i y^j \dots \right)
+P_a(x,y,z,w) & = & \sum_i^n \sum_j^n  \sum_k^n \sum_l^n  a_{ijkl} x^i y^j z^k w^l , \,\,\,\,\,\, \text{or} \\
+                   & = & \sum_i^n \sum_j^{(n-i)} \sum_k^{(n-i-j)} \sum_l^{(n-i-j-k)}  a_{ijkl} x^i y^j z^k w^l  .
 \end{eqnarray}
-
- If we consider a two  polynomial variables where $n=2$, we  have :
-
-\begin{eqnarray}
-P_a(x,y) &=&   a_{00} + a_{10}x + a_{20}x^2 + a_{01}y + a_{11} xy + a_{21}x^2y +  a_{02}y^2  \\
-              &+ & a_{12}xy^2 + a_{22}x^2y^2 \nonumber \label{poly0}
-\end{eqnarray}
-
- If now we consider 2 vectors of polynomials of $m$ entries  $\boldsymbol{v}_1<P_a>$ and  $\boldsymbol{v}_2<P_b>$. The inner product  consists of the product 
+where  $x,\,y,\,z$ and $w$ are the variables,  $i,\,j,\,k$ and $l$  the indices, $a$  the coefficients, and $n$ the order of the polynomial.
+We name these polynomials max order each and max order combined, respectively.  If now we consider 2 vectors of polynomials of the  same type of $m$ 
+entries  $\boldsymbol{v}_1<P_a>$ and  $\boldsymbol{v}_2<P_b>$. The inner product  consists of the product 
 of every entries  associated to a global reduction.
@@
 
-where $\times$ indicates the multiplication between two polynomials.   Note, a result coefficient may be not unique. It can be the sum of several coefficients resulting from several multiplications.
-These coefficients are named cross term. Per example, if we multiply two polynomial as defined in \ref{poly0}, the resulting coefficient $c_{12}xy^2$ will be given by :
-
-\begin{eqnarray}
-c_{12}xy^2 = \left(a_{12}b_{00}  +  a_{00}b_{12} + a_{10}b_{02}   + b_{10}a_{02}  +  a_{11}b_{01} + b_{12}a_{01}  \right)xy^2
-\end{eqnarray}
-
-These cross terms are not a problem for the cpu version but it is the bottleneck for the gpu version  (see gpu section). In term of  programming and parallelization, 
-for the cpu version the equation  \ref{poly1} can be reduced by a basic loop with a final reduction. These work can be easily done in OpenMP. Every OpenMP threads
+where $\times$ indicates the multiplication between two polynomials. The multiplications occurs basic rules of polynomials multiplication. We note 
+a result coefficient may be not unique. It can be the sum of several coefficients resulting from the sum of several multiplications (contributions). These coefficients are named cross term. 
+These cross terms are not a problem for the cpu version but they are  the bottleneck for the gpu version. 
+In term of  programming and parallelization,  for the cpu version the equation  \ref{poly1} can be reduced by a basic loop with a final reduction.
+These work can be easily done in OpenMP.
+Every OpenMP threads
 calculate several  polynomial multiplications and make a local reduction. At the end, a final reduction is performed  by the main thread \footnote{In OpenMP reduction for class is not supported}.
 
-\section*{Details of VLI\_CPU}
+\section{VLI  CPU library}
 
-The coefficients of the polynomial are represented by a large integer call VLI number.  A VLI number is characterized by two parameters, a Type and a Size.
- The type is an \texttt{unsigned long int},  the size varies from 2 to 8.
-Thus, the large integer can vary from 128 to 512 bits, by steps of 64 bits.
-Its implementation is static and determined during the compilation time.
- It is not dynamic at all, as the well know GMP. Inside the memory, the VLI numbers are  contiguous.
+The coefficients of the polynomial are represented by a large integer named VLI number.  A VLI number is characterized by is Size in bits; it is a  template parameter.
+The size varies from 128 to 512 buts.  Its implementation is static, contrary to the well know GMP \footnote{gmplib.org} and determined during the compilation time. 
+Inside the memory, the VLI numbers are  contiguous. The data type container is an 64 bits unsigned integer
 
-The basic operations supported by VLI are the additions, the multiplications and the subtractions. For every family of operations, a subset of operations are available :
-\begin{itemize}
-\item VLI$<$Type, Size$>$ += VLI$<$Type, Size$>$
-\item VLI$<$Type, Size$>$ += \texttt{long} or \texttt{int}
-\item VLI$<$Type, Size+1$>$ = VLI$<$Type, Size$>$ + VLI$<$Type, Size$>$
-\end{itemize}
-the subset of the substractions are,
-\begin{itemize}
-\item VLI$<$Type, Size$>$ -= VLI$<$Type, Size$>$
-\item VLI$<$Type, Size$>$ -= \texttt{long} or \texttt{int}
-\end{itemize}
-the subset of the  multiplications are,
-\begin{itemize}
-\item VLI$<$Type, Size$>$  $\times$= VLI$<$Type, Size$>$
-\item VLI$<$Type, Size$>$  $\times$= \texttt{long} or \texttt{int}
-\item VLI$<$Type, 2$\times$Size$>$ = VLI$<$Type, Size$>$ $\times$ VLI$<$Type, Size$>$
-\end{itemize}
-The algorithms of all these operations are based on the childhood knowledge but their realization are done in assembly, because, a few assembly instructions facilitate the implementation and boost the performance.
-\subsection*{Why Assembly ?}
+The basic operations supported by VLI are the four basic operations : $+,-,\times, /$, extended multiplication and bit shift operations.
+The library allows  operations between operands of the same type, and  integer.
+For performance issue, we have a specific solver for every operations except division.
+All the algorithms and operations are based on the childhood method but their realizations are done in assembly,
+because, the assembly instructions facilitate the implementation and boost the performance.
+
+\subsection{Assembly}
 
 In programming, the basic operators of the addition, subtraction and multiplication between identical type give the same type. 
-Per example, if we need a long addition or a long subtraction, the propagation of the  carry/borrow can be done by hand, but the performance will be low compare to a pure assembly version. 
-The number of assembly instruction can be reduce by a factor 3 for the additions/subtraction, and  2-order for the multiplication.
-
-In assembly, the addition with the carry bit (CB) allows a simplified model without  carry bit propagation by hand\footnote{ The carry bit is a bit of the status register, as the overflow bit, 
-it informs  the overflow of an operation, but it can be re-use to propagate the carry. }.  It is formalized in pseudo-assembly x86-64 by the following sequence:
-\begin{table}[htdp]
-\begin{center}
-\begin{tabular}{ l c c c l c}
- \texttt{add }   &  b[0]  & a[0]    & $\Leftrightarrow$ & a[0]+=b[0]            & \tiny{0}  \\ 
- \texttt{adc }   &  b[1]   & a[1]   & $\Leftrightarrow$ & a[1]+=b[1] + CB  & \tiny{1}  \\  
-\end{tabular}
-\end{center}
-\end{table}%
-where the first columns indicates the  mnemonic instructions, \texttt{add} is a basic addition between two number of the same type, if the result is larger than the capacity of the number, the CB is set to one,
-and re-use if the \texttt{adc} instruction is performed.
-In pure assembly, additional \texttt{move} instructions will be necessary to move the data from the global memory, to one of the 16 registers and reciprocally.
-We will not detail because it exceeds the purpose of the present document.
-
-The same techniques are performed for all additions/subtractions with respective mnemonic.
-For the multiplication, we use the mnemonic \texttt{mul} this instruction has the specificity to return the lower and the higher
-par of the product into  separate registers (the classical operator $\times$ returns only the low part in C/C++/fortran \dots), combined with the additions and our childhood algorithm, we can construct a large multiplication.
-
-\subsection*{About the implementation}
+Applied the childhood methods necessitate carry\footnote{ The carry bit is a bit of the status register, as the overflow bit, 
+it informs  the overflow of an operation, but it can be re-use to propagate the carry. }. for addition/subtraction and an extended multiplication. In assembly, these features 
+are supported by the pairs of mnemonic; \texttt{addq/adcq} for the addition; \texttt{sbbq/subq} for the subtraction and \texttt{mul} for the extent end multiplication.   
+Combining these mnemonic, we can reproduce easily long arithmetic algorithm, all our work is based on \cite{Hyde:2003:AAL:861534}.
 
@@
 
-The x86-64 architecture allows 16 registers (32 for power), as much as possible , we conserved the datas into the registers when we perform a long multiplication to reduce read/write actions to the minimum (1 cycle for a register).
-Additional optimizations are done like unrolling and assembly tips (xor,  \dots).  The VLI lib supported  x86-64 and Power assembly.
+The x86-64 architecture allows 16 registers (32 for power), as much as possible , we conserved the datas into the registers when we perform a long multiplication to reduce read/write actions to the minimum.
+Additional optimizations are done like full unrolling and assembly tips.
+We  supported  x86-64, Power64 and PTX(GPU) assembly .
 
-\subsection*{Memory layout and SIMD implementation}
+To conclude, GPU has specificities, it does not support carry bit for 64 bits integer, but only 32 bits integer\footnote{Introduce since CUDA 4.2}. 
+The consequence is twice more operations for the additions and one order of magnitude more for the multiplications. 
+PTX has still a nice feature, it allows multiplication-add operations with carry bit support \cite{CUDAasm}.
+
+\subsection{Memory layout and SIMD implementation}
 
 It exists at least two data layout for this problem : the array of structure (AoS) and structure of array (SoA), both have their advantage and disadvantage.
-In our problem the AoS consists of saving the data in this order  : VLI - Series of VLI (polynomial coefficient)- series of polynomial (the vector), it can be represented in the memory (contiguous) by:
-
+In our problem the AoS consists of saving the data in this order  : VLI - Series of VLI (polynomial coefficient)- series of polynomial (the vector).
+The SoA structure interleaves the coefficient of the VLI and the polynomial:
+\begin{figure}
+\begin{minipage}{0.50\linewidth}
 \begin{eqnarray}
+\tiny{
  \underbrace{
  \underbrace{ \underbrace{a^0_{00} | a^1_{00} | a^2_{00}}_{VLI, a_{00} } || \underbrace{a^0_{10} | a^1_{10} | a^2_{10}}_{VLI, a_{10}} ||   \dots  }_{1^{st} \textrm{polynomial}} 
-      |||    \underbrace{ \underbrace{b^0_{00} | a^1_{00} | b^2_{00}}_{VLI, b_{00} } || \underbrace{b^0_{10} | b^1_{10} | b^2_{10}}_{VLI, b_{10}} ||   \dots  }_{2^{nd} \textrm{polynomial} } }_{\textrm{Vector of polynomial}} \nonumber
+      |||    \underbrace{ \underbrace{b^0_{00} | a^1_{00} | b^2_{00}}_{VLI, b_{00} } || \underbrace{b^0_{10} | b^1_{10} | b^2_{10}}_{VLI, b_{10}} ||   \dots  }_{2^{nd} \textrm{polynomial} } }_{\textrm{Vector of polynomial, AoS order}} \nonumber
+}
 \end{eqnarray}
-
-The second possibility; SoA interleaves the coefficient of the VLI and the polynomial:
-
+\end{minipage}
+\begin{minipage}{0.50\linewidth}
 \begin{eqnarray}
+\tiny{
  \underbrace{
  \underbrace{a^0_{00} | a^0_{10} | \dots || a^1_{00} | a^1_{10} | \dots  }_{1st  \textrm{ nested VLI/Polynomial} } |||  \underbrace{b^0_{00} | b^0_{10} | \dots || b^1_{00} | b^1_{10} | \dots  }_{2nd  \textrm{ nested VLI/Polynomial} }
- }_{\textrm{Vector of polynomial}} \nonumber
+ }_{\textrm{Vector of polynomial, SoA Order}} \nonumber }
 \end{eqnarray}
+\end{minipage}
+\caption{Representation of the AoS Order (left) or SoA Order (right) for a vector of polynomial \label{AOSSOA}}
+\end{figure}
 
-The choice of the memory layout is important because it allows or not the usage of the SIMD instructions.
-The SIMD instructions are 128 to 256 bits vector register ($xmm_i$ or $ymm_i$) allows 4 to 8 \texttt{int}  operations simultaneously.
-A number of constraints force us to work under 31 bits base because we have to manage the carry bit and the extended multiplication by hand.
-In the previous sections we presented an addition of 128 bits VLI.
-It took two operations
-without \texttt{move}.   The same operation in SIMD (256 bits registers) necessitates a few adaptations as we work now in base 31 :
+The choice of the memory layout is important because it facilitates  the usage of the SIMD/AVX instructions. We choose the AoS for the following reasons:
+we want reuse the VLI library in other object conception code,  we can not interleave the VLI number with the polynomial like on the figure \ref{AOSSOA}. 
+Thus, the data structure will only concern the very large coefficient.
+For an addition it is a non sense because it does not support carry bit, and less carry bit between entries of the multimedia register, and it  necessitate work under 31 bits to simulate the carry bit.
+The extended multiplication under AVX \texttt{\_\_mm256\_mul\_epu32}  allows only four multiplication of 32 bits, we save a factor two. Nevertheless, on Sandy Bridge, we have 3 ALU, we can not exclude
+the execution of several long multiplication in the same time, thus we loose the benefice of the SIMD version.
+Moreover a large extra work it necessary to prepare the register and shuffle the data into the register which are costly.
+To conclude for our range of operations the extended multiplication varies from 128 to 512 bits , a SIMD version will be not optimum.
+Although, it has been used with success for larger
+integer \cite{SIMD}
 
+\section{A Short introduction on GPU}
+
+A Graphical Portable Unit  is efficient  to address problems that can be expressed as data-parallel computations Ð the same program 
+is executed on many data elements in parallel. Successful usage of GPU in HPC necessitates a good understanding of their design to reach great performance. A GPU is composed 
+of several independent calculation  entities named streaming multiprocessors (SMs typically 4 to 8), where the SMs run hundreds of threads concurrently (typically 256) by group of 32 (warp).
+
+The GPU programming model imposes to partition the problem into coarse sub-problem that can be solved independently in parallel by blocks of threads,
+ and each sub-problem into finer pieces that can be solved cooperatively in parallel by all threads within the block. During the execution, independent blocks will
+ be executed on free SMs, where the program of the block is executed over the threads concurrently way.
+
+In terms of memory all SMs  share  a large  memory (several GigaBytes) with a low performance; approximatively one hundred cycle for read/write operations.
+Inside each SMs additional specials memories (constant, shared, texture, surface) are available. These memoes are characterized by their behavior, they can be read only, shared, cache and more.  
+They have still a common characteristic, a very high performance for read/write (if possible) operations in a few cycles only.
+A good programming practice is to preload data from global memory to shared/texture at once, for subsequent computations.  \label{CUDA_PRACTICE}
+
+Apply the GPU ideology on the polynomial inner product  means every single polynomial multiplication will be execute in a block on a SMs where the multiplication itself will be execute locally in parallel by hundred of threads of the SMs.
+The last step will consist of a global reduction to sum all results polynomials from previous multiplication.
+The division into blocks is natural with the inner product, a block is a single entry of the vectors of the inner product, all the difficulty consists of designing a parallel algorithm of the polynomial multiplication and the final reduction.
+
+\subsection{Inner product booster on gpu}
+
+The first idea coming in mind: calculate every single multiplication per a independent gpu thread,  indeed for a polynomial of type max order each, the total number of multiplication needed between coefficient is equal to $(2n+1)^m$, where $n$ is the order, and $m$ the number of variables.
+This large number of coefficient is great , and compatible with the number of SMs threads.
+Nevertheless,  in the introduction, we demonstrated result coefficients may have several contributions $(n+1)^m$ maximum.
+if the contributions are calculated by several independent threads, it necessitates reduction between threads.
+This solution introduces necessarily  synchronizations, and lower performance.
+
+The peak performance is reached, if the warps run the same amount of work, and consequently a thread should be responsible of the full contributions of an output coefficient.
+To achieve this goal, we defined a task object.
+A task is a set of 32 result coefficients in a specific order.
+A task tells its  warp how calculate coefficient (the thread number one of the warp takes the first coefficient of the task and so on).
+The warps may not have the same number of tasks, as we want balance the amount of work.
+We precede in two  steps, first on cpu, a preliminary work is done. We prepare all informations on the tasks, the second step will be the calculation on GPU.
+
+At the beginning, we create a resulting polynomial  represented by the indices and the contributions. They sorted by descending order on the number of contribution
+and group by 32, this group named a task.
+These tasks are reordered over the warps to equilibrate the amount of work of the warps.
+An illustration of this algorithm is illustrated on FIG \ref{algo_gpu}.
+More the polynomials will have coefficients  more the distribution of the tasks over the warps will be  uniform and the calculations will be efficient. 
+In function of the number of contribution a warp may calculate one task, whereas an other warp may have several tasks to calculate.
+ When the determination of the task's list is over,  we transfer  asynchronously in, or in the global memory, if it does not fit.
+Note that this method does not  change the order of the datas. As for the CPU version we conserve an AoS order.
+
+\begin{figure}[t]
 \begin{center}
-\begin{tabular}{ c  c c c c c c c c} 
-\\
- $ymm_0$   register      &    $a^0_{00}$  & $a^0_{10}$ & $a^0_{20}$  &  $a^0_{30}$ & $a^0_{40}$  &   $a^0_{50}$ &  $a^0_{60}$   & $a^0_{70}$  \\
- $ymm_1$   register      &    $b^0_{00}$  & $b^0_{10}$ & $b^0_{20}$  &  $b^0_{30}$ & $b^0_{40}$  &   $b^0_{50}$ &  $b^0_{60}$   & $b^0_{70}$  \\
-$ymm_0 += ymm_1$    &    $c^0_{00}$  & $c^0_{10}$ & $c^0_{20}$  &  $c^0_{30}$  & $c^0_{40}$  &   $c^0_{50}$  &  $c^0_{60}$   & $c^0_{70}$  \\
-shift right  get CB        &     CB                 & CB                & CB                 & CB                   & CB               &  CB                   & CB                   & CB \\
-$ymm_2$   register      &    $a^1_{00}$  & $a^1_{10}$ & $a^1_{20}$  &  $a^1_{30}$ & $a^1_{40}$  &   $a^1_{50}$ &  $a^1_{60}$   & $a^1_{70}$  \\
- $ymm_3$   register      &    $b^1_{00}$  & $b^1_{10}$ & $b^1_{20}$  &  $b^1_{30}$ & $b^1_{40}$  &   $b^1_{50}$ &  $b^1_{60}$   & $b^1_{70}$  \\
-$ymm_2 += ymm_3$     & $c^1_{00}$  & $c^1_{10}$ & $c^1_{20}$  &  $c^1_{30}$  & $c^1_{40}$  &   $c^1_{50}$  &  $c^1_{60}$   & $c^1_{70}$  \\                            
-$ymm_2 +=  \textrm{CB } $ & $c^1_{00}$  & $c^1_{10}$ & $c^1_{20}$  &  $c^1_{30}$  & $c^1_{40}$  &   $c^1_{50}$  &  $c^1_{60}$   & $c^1_{70}$  \\                            
-\\
+\mbox{
+\includegraphics[scale=0.37, angle=-90]{coeffs.eps} 
+\includegraphics[scale=0.37, angle=-90]{warp.eps} 
+}
+\caption{Left: Results coefficients (red line) sorted in reverse order and group by warp. Right: Balance of the tasks over the warps}
+\label{algo_gpu}
+\end{center}
+\end{figure}
+
+ \begin{table}[t] 
+	\begin{center}
+	 	\begin{tabular}{l l}
+                           \hline 
+                           \textbf{Task execution}  Algorithm &  \\ \hline
+                           \begin{tabular}{c l l} 
+                               \tiny{1:} & VLI tmp   &  $\vartriangleright$ specific to every threads  \\
+                               \tiny{2:} &  \textbf{for} over number of task list size &$\vartriangleright$  Depend of the warp \\
+                               \tiny{3:} &  tmp $\,\, \wedge=$ tmp     & $\vartriangleright$ flush to 0 \\    
+                               \tiny{4:} &   VLI ResCoeff &$\vartriangleright$  thread get an output coeff from a task \\
+                               \tiny{5:} &  \hspace{0.2 cm}  \textbf{for} over the number of contribution  & $\vartriangleright$ each thread loop over its coefficient  \\
+                               \tiny{6:} &  \hspace{0.4 cm}   Indices$_i$ = GetIndices(ResCoeff) &$\vartriangleright$ Determine input indices \\
+                               \tiny{7:} &  \hspace{0.4 cm}   tmp + = CoeffPoly$_1$(indices$_1$) $\times$ CoeffPoly$_2$(indices$_2$) &$\vartriangleright$ Perform the extended MulAdd  \\
+                               \tiny{8:} &  \hspace{0.2 cm}   \textbf{end for} &\\
+                               \tiny{9:} &  SaveSoA(tmp) & $\vartriangleright$ write into the global memory under AoS \\                               
+                               \tiny{10:} &  \textbf{end for} &\\                               
+                          \end{tabular} &  \\ \hline
 		 \end{tabular} 
+		 \caption{Algorithm of the task execution on GPU \label{ALGO}}
 	\end{center}
+\end{table}
 
-This operation must be repeat two more times to get the 124 bits addition.
-At the end, we perform 9 additions and 3  shift operations for eight VLI$<4*31, \texttt{\textrm{uint}}>$ += VLI$<4*31, \texttt{\textrm{uint}}>$ operations.
- Comparing to the original solution we saved a factor two,  nevertheless on modern CPU (Sandybridge), we have 3 APU  we can not exclude that the execution of  the addition of the polynomial coefficients will be execute in parallel over the APU.
+This task's list job executed only once for all polynomial types,  we limit the overhead for the first inner product. As soon as,  all task's list are calculated, 
+we transfer the datas asynchronously on the device (the global memory of the GPU is allocated once, thus we limit the workload of the cuda malloc),
+ and the calculation of all polynomial starts. First, both input polynomials  are load into faster memory (shared or texture) in function of there size. It can be
+ both in the shared, one polynomial into the shared and the second one into the texture or two in the texture memory. 
+The determination of which  fast memory (shared or texture ) is determined during the compilation. Thus we avoid if statement and probable sequentialization of the warps. 
+ we privileged  the texture memory on Kepler architecture, and the shared memory on the Fermi architecture.
 
-Beyond, we should consider the extended  multiplication using SIMD instructions.
-For this multiplication of unsigned number   VLI$<4*64, \texttt{\textrm{ulint}}>$  = VLI$<2*64, \texttt{\textrm{ulint}}>$ $\times$ VLI$<2*64, \texttt{\textrm{ulint}}>$) using the
-childhood  algorithm, we need 4 multiplications and 3 additions so 32 multiplications and 24 additions to perform 8 vli multiplications.
-For the SIMD version we need 16 multiplications and 29 additions, 16 copies for the ouput multiplications
-associated to 16 right shift  to extract the higher part on the multiplication and 16 mask operator to get the lower part.
-And additional shift to get the carry bit.
-Execution time for both implementation will be similar.
-Nevertheless, we should remember the treatment of the sign,  a long multiplication can necessitate branching (cite art of assembly).
-The branching can be remove but it  becomes painful and useless for number larger than 128 bits.
-A SIMD version necessitates no branching but its implementation will definitively discredit this choice.
+When a polynomial multiplication begins,  every warp get corresponding tasks (at least one) from the task's list and loop over its number of task.Then,
+the threads into the warp get their respective output coefficients and iterate over the number of  contribution.
 
-As the SIMD implementation does not give enough argument to focus on it, we have privileged  the AoS and classic assembly instructions.
-To conclude the new coprocessor of Intel named Mic could reconsider our choice on SIMD
-because this processor supports CB, and extraction of the  lower and higher part of the multiplication. It allows selected operations on the registers by a judicious system of mask.
+For every iteration over contribution, the threads determine the input coefficients, read the corresponding data in the fast memory and perform the arithmetic operations : a long multiplication and an optional long addition (if the result coefficient has several contribution).
+The calculation of an output coefficient  are  done into fast memory (shared and texture for the input data, and register for the intermediate result).
+When the loop over contribution terminates.
+We transfer the coefficients into the an intermediate buffer in the global memory, but in SoA (as the Fig.
+\ref{AOSSOA} and table \ref{ALGO}) order to prepare the final reduction.
+The reduction is done using the standard method of Nvidia \cite{CUDAReduction} to achieve the highest performance.
 
-NOTE :  for Andreas and Matthias :
-Max Nvidia just coded $2n <64 bits> =  2n <64 bits>* 2n <64 bits> $ doing this, you get the sign naturally, but the it costs .
-The version  $2n <64 bits> =  n <64 bits> * n <64 bits>$ with branching divides by 1.5 the time of Max.
+To conclude, we developed an optional hybrid mode between CPU/GPU, this  technic is usual in GPU computing \cite{CUDAReduction}.
+We split the inputs vectors of polynomials into two chucks of datas, execute concurrently on CPU and GPU.
+Only at the end, a synchronization during  the memory transfer is done  for the final reduction.
+
+\section{Results}
+
+\begin{figure}[t!]
+\begin{center}
+\mbox{
+\hspace{-0.5cm}
+\includegraphics[scale=0.25, angle=-90]{ME128.eps} 
+\hspace{-0.4cm}
+\includegraphics[scale=0.25, angle=-90]{ME192.eps} 
+\hspace{-0.4cm}
+\includegraphics[scale=0.25, angle=-90]{ME256.eps} 
+}
+\caption{Left: Inner product up to 3 variables with 128 to 256 bits coefficient.
+Middle: Inner product up to 3 variables with 192 to 384 bits coefficient.
+Right: Inner product up to 3 variables with 256 to 512 bits coefficient}
+\label{ResME}
+\end{center}
+\end{figure}
+
+Benchmarks are performed on the facilities of The Swiss Center for Scientific Computing (CSCS), on cray xk6 and xk7 cluster for the GPU and a Sandy Bridge  node 32 logical cores  (HyperThreading on) Xeon E5-2670.
+The size of our vector is equal to 4096, the order of the polynomials varies from 1 to 14, the size of the inputs coefficients of the polynomials varies from 128 to 256 bits, the output polynomial has twice larger coefficient (256 to 192 bits).
+
+Evaluate  integer benchmarks integer are not easy, contrary to float number, it does not exist a universal\footnote{Altough the MIPS exists (Million Instructions Per Seconds), it should be the equivalent of the FLOPS, it's signification is disputable} value as the FLOPS for floating-point operation, because integer operations performed by ALU, and the number of  ALU is not constant between processor, and the execution of the operation are not constant.
+We compute the  result  numbers as GOP/s$=10^{-9} \times batch/t$, where  $batch$ is the number of long integer operations computed during the execution (additions and multiplications), $t$ [s]  is the elapsed time, we note the execution time for GPU contains the memory  transfer.
+
+Results presented on the Figure \ref{ResME} compare our CPU library to GMP (or polynomial is just instantiated with VLI of GMP),  we get an excellent speed up factor from 4 to 11.
+For two raisons, the coefficient of the  polynomial are allocated in the heap, in the case of GMP, they are allocated in the stack, it is well know better performances are obtained with less dynamic memory allocation.
+The second raison, the meta-programming allows the construction of a specific solver for every type of large integer during the compilation, contrary to GMP, we do not have a universal solver (at least for small size), the solver is unique with the maximum optimization (operations are performing in the register only).
+
+The GPU curves show again an excellent speed up to factor 10 to 60 compare to GMP, but with an interesting features.
+We have disappointing results for one variable, a continuous increases of the performance for two variables, and a peak performance for
+three variables around order 8, with a significative drop of the performance.
+On the first point, polynomial on one variable have a small number of coefficient ($Order+1$), in the present condition, we used a very small number of warp, and loose all the performance
+of the GPU.
+For the second and third point is the same cause, as we said in section \ref{CUDA_PRACTICE}, a good practice of programming consists to prefetch the data on faster memory.
+On Kepler, we selected the texture memory because, it
+highly optimized for repetitive and random access contrary to shared memory.
+Its size is still limited to 48 KB maximum.
+For 3 variables the sum of the  size [Byte] of the polynomials is equal to $2(Order+1)^3\times \texttt{sizeof(VLI)}$ which corresponds approximatively
+to the full size of the texture memory.
+When the texture memory is full, memory access are not full cached and the performance droops instantaneously.
+For two variables, the input polynomials are still cached all over the range of Order, thus we get a continuous  increased
+of the performance until probably the maximum of three variables.
+
+%The performance indicator for integer is Millions of Instructions Per Second MIPS\footnote{Also know as Meaningless Indicator of Processor Speed}.
+As 99 \% of the code is spent in the different arithmetic solver, we can easily determine
+%the MIPS of our applications.
+We counted all ASM instructions in our solver and we divided by the full execution time.
+Determine the theoretical peak of the ALU is problematic.
+There number varies from architecture, the cost
+%of integer operations are not the same on ALU [CITE COST GUY] contrary  to FPU.
+Base on the full work of [CITE BENCH GUY] , we determine experimentally the peak performance by saturating the processors
+%by additions and multiplications. We get  38.10 GMIPS for the multiplications and 87.27 GMIPS for the additions, we keep this last value as peak performance indicator.
+
+%We determine experimentally the peak performance  by a benchmark an acceptable based on the benchmark work of [CITE BENCH GUY]
+
+\section{Conclusions}
+
+Turlututu chapeau pointu !  tyrlytyty pointy hat !
+
+\bibliographystyle{plain}	% (uses file "plain.bst")
+\bibliography{biblio/vlibib}
 

### 2012-11-11 doc/paper_vli/vli.tex
doc typo

@@
 
-To conclude, we developed an optional hybrid mode between CPU/GPU, this  technic is usual in GPU computing \cite{CUDAReduction}.
+To conclude, we developed an optional hybrid mode between CPU/GPU, this  technic is usual in GPU computing \cite{magma}.
 We split the inputs vectors of polynomials into two chucks of datas, execute concurrently on CPU and GPU.
@@
 
-Evaluate  integer benchmarks integer are not easy, contrary to float number, it does not exist a universal\footnote{Altough the MIPS exists (Million Instructions Per Seconds), it should be the equivalent of the FLOPS, it's signification is disputable} value as the FLOPS for floating-point operation, because integer operations performed by ALU, and the number of  ALU is not constant between processor, and the execution of the operation are not constant.
+Evaluate  integer benchmarks integer are not easy, contrary to float number, it does not exist a universal\footnote{Altough the MIPS exists (Million Instructions Per Seconds), it should be the equivalent of the FLOPS, it's signification is disputable} value as the FLOPS for floating-point operation, because integer operations performed by ALU, and the number of  ALU is not constant between processor, and the execution of the operation are not constant \cite{ASMcost}.
 We compute the  result  numbers as GOP/s$=10^{-9} \times batch/t$, where  $batch$ is the number of long integer operations computed during the execution (additions and multiplications), $t$ [s]  is the elapsed time, we note the execution time for GPU contains the memory  transfer.

### 2012-11-14 doc/paper_vli/vli.tex
doc vli

@@
 
-\section*{Polynomial and Inner product definitions}
+\section{Art of work}
+
+BLABLABLALBALBALBLALBALBLALBLALBLALBLALBABLALB ALBLALBALBLABLALBLABLALBLA......
+
+\section{Polynomial and Inner product definitions}
 
@@
 Moreover a large extra work it necessary to prepare the register and shuffle the data into the register which are costly.
-To conclude for our range of operations the extended multiplication varies from 128 to 512 bits , a SIMD version will be not optimum.
-Although, it has been used with success for larger
-integer \cite{SIMD}
+For our range of operations the extended multiplication varies from 128 to 512 bits , a SIMD version will be not optimum, although, it has been used with success for larger integer \cite{SIMD}.
+To conclude, our library is build on AoS format.
 

### 2012-11-18 doc/paper_vli/vli.tex
pqper vli

--- 2012-11-11 doc/paper_vli/vli.tex
@@
 
-\title{Large integer arithmetic and polynomial inner product}
-\author{Timoth\'ee Ewart\inst{1}, Andreas Hein$^2$, Matthias Troyer\inst{2} and Thierry Giamarchi\inst{1}}
+\title{Very Large Integer  and polynomial inner product}
+\author{Timoth\'ee Ewart\inst{1}, Andreas Hehn$^2$, Matthias Troyer\inst{2} and Thierry Giamarchi\inst{1}}
 
@@
 %\subsection{}
+%\vspace{-0.65cm}
+\begin{abstract}
+The main purpose of the  VLI C++ library is to perform large integer arithmetic and polynomial inner product.
+Based on meta-programming and  hybrid (CPU/GPU) mode, it performs inner product of polynomial of 1 to 4 variables  up to order 14.
+The coefficients are represented  by large integer of 128 to 256 bits.
+We remember the basics of the inner product with polynomials, and  the assembly implementation of the  library.
+Then, we focus on the gpu implementation of the inner product.
+All these algorithms are based on brute force attacks, as the size of the  problem should not justify a FFT approach.
+We save a factor up to 10 on CPU and 60 on GPU compared to a GMP solution.
+\end{abstract}
+\section{Introduction}
+ A few libraries  LINBOX or NTL  deliver high performance in polynomial solvers.
+These libraries are usually connect ed with GMP\footnote{GMP is the standard library for long arithmetic, gmplib.org} when needed. 
+The introduction of hardware accelerator in HPC a few years ago and more recently GPU  by Nvidia introduce a revolution. It necessitates to adapt library to these new hardwares. 
+The most famous example on this transition in mathematic is the famous blas/lapack libraries to MAGMA \cite{magma}, on the other side, polynomials libraries does not make the transition to GPU.
+During the last few years, authors like \cite{Govindaraju2008, Emeliyanenko:2009, Moreno2010}, write polynomial multiplication  to  this new hardware, successfully.
+All these authors focus on large polynomials,
+ with a FFT approach.
 
-The main purpose of the VLI lib is performed large integer arithmetic based on meta-programming and  hybrid (CPU/GPU) inner product of polynomial up to order 14 with 1 to 4 variables with large integer coefficient 128 to 256 bits.
-We remember the basics of the inner product of a polynomial,  and the implementation of the large integer library for the cpu version  The second part will be on the gpu implementation of the inner product.
-All these algorithms are based on brut force attacks, as their size should not justify a FFT approach \cite{Emeliyanenko:2009}.
+In the present work, we are focusing on the inner product with dense and triangular multivariable polynomials.  The coefficients of the polynomials are large integer. 
+We preferred developed a long integer library  han utilize GMP, because GMP is a dynamic library.
+Moreover, we want a library natively compatible with GPU to allow hybrid calculation. GPU imposes contiguous memory which impossible with a GMP approach.
+Thus, we propose a static library based on contiguous memory where the exact assembly kernels are generated by meta-programming during the compilation, at the time, for CPU and GPU.
+We do not wast time in dynamic memory allocation and let the compiler optimizes deeply the library for both architectures.
 
-\section*{Polynomial and Inner product definitions}
+\section{Polynomial and Inner product definitions}
 
@@
 where  $x,\,y,\,z$ and $w$ are the variables,  $i,\,j,\,k$ and $l$  the indices, $a$  the coefficients, and $n$ the order of the polynomial.
-We name these polynomials max order each and max order combined, respectively.  If now we consider 2 vectors of polynomials of the  same type of $m$ 
+Theses polynomials are named dense and  triangular polynomial, respectively.  If now we consider two vectors of polynomials of the  same type of $m$ 
 entries  $\boldsymbol{v}_1<P_a>$ and  $\boldsymbol{v}_2<P_b>$. The inner product  consists of the product 
 of every entries  associated to a global reduction. 
+ \begin{eqnarray}
+ P_c = \sum_i^m \boldsymbol{v}_{1_i}<P_a>  \times  \boldsymbol{v}_{2_i}<P_b> \label{poly1} \label{InnerProduct}
+\end{eqnarray}
+where $\times$ indicates the multiplication between two polynomials. The multiplications occurs basic rules of polynomials multiplication. 
+A final coefficient may be the sum of several coefficients (named contributions), resulting from the sum of several multiplications, these coefficients are the cross terms. 
+The cross terms are not a problem for the cpu version, as a polynomial multiplication is done by a single thread but they are the bottleneck for the gpu version.
+The number of coefficient for a dense and triangular polynomial are given by $(n+1)^{\kappa}$ and  $(\kappa+n-1)! / ( (\kappa-1)! n! )$, respectively, where $\kappa$ is the number of variable.
 
- \begin{eqnarray}
- P_c = \sum_i^m \boldsymbol{v}_{1_i}<P_a>  \times  \boldsymbol{v}_{2_i}<P_b> \label{poly1}
-\end{eqnarray}
-
-where $\times$ indicates the multiplication between two polynomials. The multiplications occurs basic rules of polynomials multiplication. We note 
-a result coefficient may be not unique. It can be the sum of several coefficients resulting from the sum of several multiplications (contributions). These coefficients are named cross term. 
-These cross terms are not a problem for the cpu version but they are  the bottleneck for the gpu version. 
-In term of  programming and parallelization,  for the cpu version the equation  \ref{poly1} can be reduced by a basic loop with a final reduction.
-These work can be easily done in OpenMP.
-Every OpenMP threads
-calculate several  polynomial multiplications and make a local reduction. At the end, a final reduction is performed  by the main thread \footnote{In OpenMP reduction for class is not supported}.
+In term of  programming and parallelization, at least for the cpu version, the previous equations can be reduced by independent loops with a final reduction.
+ These work can be easily done in OpenMP. Every OpenMP threads calculate several  polynomial multiplications and make a local reduction. At the end, a final reduction is performed  by the main thread.
+ % \footnote{In OpenMP reduction for class is not supported}. Results showed a perfect scalability of the OpenMP code.
 
@@
 
-The coefficients of the polynomial are represented by a large integer named VLI number.  A VLI number is characterized by is Size in bits; it is a  template parameter.
-The size varies from 128 to 512 buts.  Its implementation is static, contrary to the well know GMP \footnote{gmplib.org} and determined during the compilation time. 
-Inside the memory, the VLI numbers are  contiguous. The data type container is an 64 bits unsigned integer
+The coefficients of the polynomial are represented by a large integer named VLI number.  A VLI number is characterized by a size in bits; it is a  template parameter.
+The size varies from 128 to 512 bits.  Its implementation is static, contrary to the well know GMP. 
+Inside the memory, the VLI numbers are  contiguous. The data type container is an 64 bits unsigned integer array.
 
-The basic operations supported by VLI are the four basic operations : $+,-,\times, /$, extended multiplication and bit shift operations.
+The basic operations supported by VLI are the four basic operations : $+,-,\times, /$, extended multiplications and bit shift operations.
 The library allows  operations between operands of the same type, and  integer.
-For performance issue, we have a specific solver for every operations except division.
-All the algorithms and operations are based on the childhood method but their realizations are done in assembly,
-because, the assembly instructions facilitate the implementation and boost the performance.
+For performance issue, we have a specific solver for every operations except  the divisions.
+All the algorithms and operations are based on the childhood methods but the realizations has been done in assembly,
+because, the assembly instructions facilitate the implementation and boost the performances.
 
@@
 
-In programming, the basic operators of the addition, subtraction and multiplication between identical type give the same type. 
-Applied the childhood methods necessitate carry\footnote{ The carry bit is a bit of the status register, as the overflow bit, 
-it informs  the overflow of an operation, but it can be re-use to propagate the carry. }. for addition/subtraction and an extended multiplication. In assembly, these features 
-are supported by the pairs of mnemonic; \texttt{addq/adcq} for the addition; \texttt{sbbq/subq} for the subtraction and \texttt{mul} for the extent end multiplication.   
-Combining these mnemonic, we can reproduce easily long arithmetic algorithm, all our work is based on \cite{Hyde:2003:AAL:861534}.
+As we need large arithmetic, we applied algorithms based on childhood  experience. These methods necessitate carry\footnote{ The carry bit is a bit of the status register, as the overflow bit, 
+it informs  the overflow of an operation, but it can be re-use to propagate the carry. } for addition/subtraction and an extended multiplication for the multiplication.
+Although the hardware supports them; we can not access to these features by the C++ language, the only possibility is the assembly.
 
-We code the assembly using the GNU inline assembly, it consists of a string between ASM tag.
-Generate by hand all these  string (kernels) in assembly can be painful and a source of error ($\sim$10 000 lines). 
-We designed a generator of assembly. It create all versions of a kernels during the compilation.   
-We use the preprocessor and the  boost preprocessor package. It allows repetition, branching, and more
- important the construction of string during the pre-compilation.  All kernels use this procedure except the extended multiplication for the cpu.
+In assembly, these features  are supported by the pairs of mnemonic; \texttt{addq/adcq} for the addition; \texttt{sbbq/subq} for the subtraction and \texttt{mul} for the extentend multiplication.   
+Combining these mnemonic, we reproduce easily long arithmetic algorithm. The ASM work is based on \cite{Hyde:2003:AAL:861534}.
 
-The x86-64 architecture allows 16 registers (32 for power), as much as possible , we conserved the datas into the registers when we perform a long multiplication to reduce read/write actions to the minimum.
-Additional optimizations are done like full unrolling and assembly tips.
-We  supported  x86-64, Power64 and PTX(GPU) assembly .
+We coded the assembly using the GNU inline assembly, it consists of a string between ASM tag.
+Generate by hand all these  assembly strings (kernels)  can be painful and a source of error \footnote{The library has near 8000 lines of assembly cumulated between x86-64, Power64, and PTX} 
+We designed a generator of assembly. It creates, all versions of a kernels during the compilation.   
+It is built on the boost preprocessor package, because we can create, manipulate the strings during the pre-compilation, 
+and generate a specific tuned kernel for every large integer. All kernels use this procedure except the extended multiplication for the cpu.
 
-To conclude, GPU has specificities, it does not support carry bit for 64 bits integer, but only 32 bits integer\footnote{Introduce since CUDA 4.2}. 
-The consequence is twice more operations for the additions and one order of magnitude more for the multiplications. 
-PTX has still a nice feature, it allows multiplication-add operations with carry bit support \cite{CUDAasm}.
+The x86-64 architecture allows 16 registers (32 for power), as much as possible, we conserved the datas into the registers when we perform a long multiplication to reduce read/write actions to the
+stack to the minimum. Additional optimizations have been done like full unrolling and assembly tips. We  supported  x86-64, Power64 and PTX (GPU) assembly.
+Moreover GPU  assembly has specificities, it does not support carry bit for 64 bits integer, but only for 32 bits integer\footnote{Introduce since CUDA 4.2}.  
+It allows multiplication-add 32 bits integers operations with carry bit support \cite{CUDAasm}. 
+Work in 32 bits mode imposes twice more operations for the additions/substractions and one order of magnitude more for the multiplications.
 
@@
 
-It exists at least two data layout for this problem : the array of structure (AoS) and structure of array (SoA), both have their advantage and disadvantage.
+It exists at least two data layouts for this problem: the array of structure (AoS) and structure of array (SoA), both have  advantages and disadvantages.
 In our problem the AoS consists of saving the data in this order  : VLI - Series of VLI (polynomial coefficient)- series of polynomial (the vector).
-The SoA structure interleaves the coefficient of the VLI and the polynomial:
+The SoA structure interleaves the coefficients of the VLI and the polynomials:
 \begin{figure}
@@
 
-The choice of the memory layout is important because it facilitates  the usage of the SIMD/AVX instructions. We choose the AoS for the following reasons:
-we want reuse the VLI library in other object conception code,  we can not interleave the VLI number with the polynomial like on the figure \ref{AOSSOA}. 
-Thus, the data structure will only concern the very large coefficient.
-For an addition it is a non sense because it does not support carry bit, and less carry bit between entries of the multimedia register, and it  necessitate work under 31 bits to simulate the carry bit.
+The choice of the memory layout is important, it facilitates  the usage of the SIMD/AVX instructions. We choose the AoS for the following reasons:
+we want reuse the VLI library in other object conception code, so, we can not interleave the VLI number with the polynomial like on the figure \ref{AOSSOA}. 
+The data structure should only concern the very large coefficient.
+Apply SIMD on AoS, for an addition it is a non sense because there is no specific instruction to propagate carry between entries of the multimedia register.
 The extended multiplication under AVX \texttt{\_\_mm256\_mul\_epu32}  allows only four multiplication of 32 bits, we save a factor two. Nevertheless, on Sandy Bridge, we have 3 ALU, we can not exclude
-the execution of several long multiplication in the same time, thus we loose the benefice of the SIMD version.
-Moreover a large extra work it necessary to prepare the register and shuffle the data into the register which are costly.
-To conclude for our range of operations the extended multiplication varies from 128 to 512 bits , a SIMD version will be not optimum.
-Although, it has been used with success for larger
-integer \cite{SIMD}
+the execution of several long multiplication in the same time, and we could loose the benefice of the SIMD version.
+Moreover, a large extra work is necessary to prepare the registers and shuffle the data into the registers which are expensives.
+For our range of operations the extended multiplication varies from 128 to 512 bits, a SIMD version will be not the optimum solution.
+Though, it has been used with success for larg integer ($>$1024 bits) \cite{SIMD}.
 
@@
 is executed on many data elements in parallel. Successful usage of GPU in HPC necessitates a good understanding of their design to reach great performance. A GPU is composed 
-of several independent calculation  entities named streaming multiprocessors (SMs typically 4 to 8), where the SMs run hundreds of threads concurrently (typically 256) by group of 32 (warp).
+of several independent calculation  entities named streaming multiprocessors (SMs typically a tenth), where the SMs run hundreds of threads concurrently (typically 256) by group of 32 (warp).
+The GPU programming model imposes to divide the problem into coarse sub-problem that can be solved independently in parallel by blocks of threads,
+ and each sub-problem into finer pieces that can be solved cooperatively in parallel by all threads within the block. 
+% During the execution, independent blocks will be executed on free SMs, where the program of the block is executed over the threads concurrently way.
 
-The GPU programming model imposes to partition the problem into coarse sub-problem that can be solved independently in parallel by blocks of threads,
- and each sub-problem into finer pieces that can be solved cooperatively in parallel by all threads within the block. During the execution, independent blocks will
- be executed on free SMs, where the program of the block is executed over the threads concurrently way.
-
-In terms of memory all SMs  share  a large  memory (several GigaBytes) with a low performance; approximatively one hundred cycle for read/write operations.
-Inside each SMs additional specials memories (constant, shared, texture, surface) are available. These memoes are characterized by their behavior, they can be read only, shared, cache and more.  
-They have still a common characteristic, a very high performance for read/write (if possible) operations in a few cycles only.
+In terms of memory all SMs  share  a large  memory (several GigaBytes) with  low performance; approximatively one hundred cycle for read/write operations.
+Inside each SMs additional specials memories (constant, shared, texture, surface) are available. These memories are characterized by their behavior, they can be read only, shared, cache and more.  
+They have still a common characteristic, a very high performance for read/write  operations in a few cycles only.
 A good programming practice is to preload data from global memory to shared/texture at once, for subsequent computations.  \label{CUDA_PRACTICE}
 
-Apply the GPU ideology on the polynomial inner product  means every single polynomial multiplication will be execute in a block on a SMs where the multiplication itself will be execute locally in parallel by hundred of threads of the SMs.
-The last step will consist of a global reduction to sum all results polynomials from previous multiplication.
-The division into blocks is natural with the inner product, a block is a single entry of the vectors of the inner product, all the difficulty consists of designing a parallel algorithm of the polynomial multiplication and the final reduction.
+Apply the GPU ideology on the polynomial inner product  means the polynomial multiplications will be execute in parallel on a SMs.
+Then a global reduction will sum all results polynomials.
+The division into blocks is natural, a block is a single entry of the vectors (equation \ref{InnerProduct}), all the difficulty consists of designing a parallel algorithm of the polynomial multiplication and the final reduction.
 
 \subsection{Inner product booster on gpu}
-
-The first idea coming in mind: calculate every single multiplication per a independent gpu thread,  indeed for a polynomial of type max order each, the total number of multiplication needed between coefficient is equal to $(2n+1)^m$, where $n$ is the order, and $m$ the number of variables.
-This large number of coefficient is great , and compatible with the number of SMs threads.
-Nevertheless,  in the introduction, we demonstrated result coefficients may have several contributions $(n+1)^m$ maximum.
-if the contributions are calculated by several independent threads, it necessitates reduction between threads.
-This solution introduces necessarily  synchronizations, and lower performance.
-
-The peak performance is reached, if the warps run the same amount of work, and consequently a thread should be responsible of the full contributions of an output coefficient.
-To achieve this goal, we defined a task object.
-A task is a set of 32 result coefficients in a specific order.
-A task tells its  warp how calculate coefficient (the thread number one of the warp takes the first coefficient of the task and so on).
-The warps may not have the same number of tasks, as we want balance the amount of work.
-We precede in two  steps, first on cpu, a preliminary work is done. We prepare all informations on the tasks, the second step will be the calculation on GPU.
-
-At the beginning, we create a resulting polynomial  represented by the indices and the contributions. They sorted by descending order on the number of contribution
-and group by 32, this group named a task.
-These tasks are reordered over the warps to equilibrate the amount of work of the warps.
-An illustration of this algorithm is illustrated on FIG \ref{algo_gpu}.
-More the polynomials will have coefficients  more the distribution of the tasks over the warps will be  uniform and the calculations will be efficient. 
-In function of the number of contribution a warp may calculate one task, whereas an other warp may have several tasks to calculate.
- When the determination of the task's list is over,  we transfer  asynchronously in, or in the global memory, if it does not fit.
-Note that this method does not  change the order of the datas. As for the CPU version we conserve an AoS order.
 
@@
 }
-\caption{Left: Results coefficients (red line) sorted in reverse order and group by warp. Right: Balance of the tasks over the warps}
+\caption{Left: Results coefficients (red line) sorted in reverse order and group by warp. Right: Balance of the tasks over the warps.}
 \label{algo_gpu}
@@
 
+The first idea coming in mind: calculate every single multiplication per an independent gpu thread,  indeed for a dense polynomial, the number of coefficient can be large and compatible with the number of SMs threads.
+Nevertheless,  the result coefficients may have several contributions.
+If the contributions are calculated by several threads, it necessitates reduction between threads.
+This solution introduces   synchronizations, and lower performance.
+The peak performance will be reached, if the warps run the same amount of work, and every single thread into a warp should be responsible of the full contributions of an output coefficient.
+To achieve this goal, we defined a task object.
+A task is a set of 32 result coefficients in a specific order.
+A task tells a  warp how calculates coefficient (the thread number one of the warp takes the first coefficient of the task and so on).
+The warps may not have the same number of tasks, as we equilibrate the workload.
+We proceeded in two  steps; first on cpu, where a preliminary work is done, we prepare all informations on the tasks, and then, the calculation on GPU.
+
+For the first step, we create a resulting polynomial  represented by the indices and the contributions. They are sorted by descending order on the number of contribution
+and group by 32 as a task. These tasks are reordered over the warps to equilibrate the amount of work. An illustration of this algorithm is illustrated on the figure \ref{algo_gpu}. 
+More the polynomials will have coefficients  more the distribution of the tasks over the warps will be  uniform and the calculations will be efficient. 
+We fix, the maximum number of wrap to eight (8 warps - 256 threads), as far as possible, we assign at least two tasks per warps.
+Thus, in function of the number of coefficients a warp may calculate one task, whereas an other warp may have several tasks to calculate.
+ When the determination of the task's list is over,  we transfer  asynchronously the tasks lists in the global memory.
+This method does not  change the order of the datas. As for the CPU version, we have an AoS order.
  \begin{table}[t] 
@@
                            \begin{tabular}{c l l} 
-                               \tiny{1:} & VLI tmp   &  $\vartriangleright$ specific to every threads  \\
+                               \tiny{1:} & VLI tmp   &  $\vartriangleright$ Specific to every threads  \\
                                \tiny{2:} &  \textbf{for} over number of task list size &$\vartriangleright$  Depend of the warp \\
-                               \tiny{3:} &  tmp $\,\, \wedge=$ tmp     & $\vartriangleright$ flush to 0 \\    
-                               \tiny{4:} &   VLI ResCoeff &$\vartriangleright$  thread get an output coeff from a task \\
-                               \tiny{5:} &  \hspace{0.2 cm}  \textbf{for} over the number of contribution  & $\vartriangleright$ each thread loop over its coefficient  \\
+                               \tiny{3:} &  tmp $\,\, \wedge=$ tmp     & $\vartriangleright$ Flush to 0 \\    
+                               \tiny{4:} &   VLI ResCoeff &$\vartriangleright$  Thread gets a result coefficient from a task \\
+                               \tiny{5:} &  \hspace{0.2 cm}  \textbf{for} over the number of contribution  & $\vartriangleright$ Each thread loop over its coefficient  \\
                                \tiny{6:} &  \hspace{0.4 cm}   Indices$_i$ = GetIndices(ResCoeff) &$\vartriangleright$ Determine input indices \\
@@
                                \tiny{8:} &  \hspace{0.2 cm}   \textbf{end for} &\\
-                               \tiny{9:} &  SaveSoA(tmp) & $\vartriangleright$ write into the global memory under AoS \\                               
+                               \tiny{9:} &  SaveSoA(tmp) & $\vartriangleright$ Write into the global memory under AoS \\                               
                                \tiny{10:} &  \textbf{end for} &\\                               
@@
 		 \end{tabular} 
-		 \caption{Algorithm of the task execution on GPU \label{ALGO}}
+		 \caption{Algorithm of the task execution on GPU for the polynomial multiplications. \label{ALGO}}
 	\end{center}
@@
 
-This task's list job executed only once for all polynomial types,  we limit the overhead for the first inner product. As soon as,  all task's list are calculated, 
-we transfer the datas asynchronously on the device (the global memory of the GPU is allocated once, thus we limit the workload of the cuda malloc),
- and the calculation of all polynomial starts. First, both input polynomials  are load into faster memory (shared or texture) in function of there size. It can be
- both in the shared, one polynomial into the shared and the second one into the texture or two in the texture memory. 
-The determination of which  fast memory (shared or texture ) is determined during the compilation. Thus we avoid if statement and probable sequentialization of the warps. 
- we privileged  the texture memory on Kepler architecture, and the shared memory on the Fermi architecture.
+This task's list job is executed only once for all polynomial types,  we limit the overhead for the first inner product. As soon as,  all task's list are calculated, 
+we transfer the datas of the polynomials asynchronously on the device (the global memory of the GPU is allocated once, we limit the workload of the cuda free/malloc),
+ and the calculation of all polynomial multiplication start.
 
-When a polynomial multiplication begins,  every warp get corresponding tasks (at least one) from the task's list and loop over its number of task.Then,
-the threads into the warp get their respective output coefficients and iterate over the number of  contribution.
-
-For every iteration over contribution, the threads determine the input coefficients, read the corresponding data in the fast memory and perform the arithmetic operations : a long multiplication and an optional long addition (if the result coefficient has several contribution).
+ First, both input polynomials  are cached into the texture memory. 
+ %faster memory (shared or texture) in function of there size. 
+% we privileged  the texture memory on Kepler architecture. %, and the shared memory on the Fermi architecture.
+When a polynomial multiplication begins,  every warp get corresponding tasks (at least one) from the task's list and loop over its number of task. Then,
+the threads into a warp gets their respective output coefficients and iterate over the number of  contribution.
+For every iteration over contribution, the threads determine the input coefficients, read the corresponding data in the fast memory and perform the arithmetic operations: a extended multiplication and an optional long addition (if the result coefficient has several contributions).
 The calculation of an output coefficient  are  done into fast memory (shared and texture for the input data, and register for the intermediate result).
-When the loop over contribution terminates.
-We transfer the coefficients into the an intermediate buffer in the global memory, but in SoA (as the Fig.
-\ref{AOSSOA} and table \ref{ALGO}) order to prepare the final reduction.
+When the loop over contribution terminates,  we transfer the coefficients into the an intermediate buffer in the global memory, but in SoA  order to prepare the final reduction. 
 The reduction is done using the standard method of Nvidia \cite{CUDAReduction} to achieve the highest performance.
 
-To conclude, we developed an optional hybrid mode between CPU/GPU, this  technic is usual in GPU computing \cite{CUDAReduction}.
+%  It can be both in the shared, one polynomial into the shared and the second one into the texture or two in the texture memory. 
+%The determination of which  fast memory (shared or texture) is determined during the compilation. Thus we avoid if statement and probable sequentialization of the warps.
+
+To conclude, This  algorithm has the great advantage to work well with dense and sparse polynomials.
+The main difficult consists  to design an efficient function "GetIndices" that determine the contributions.
+ We also developed an optional hybrid mode between CPU/GPU, this  technic is usual in GPU computing \cite{magma}.
 We split the inputs vectors of polynomials into two chucks of datas, execute concurrently on CPU and GPU.
-Only at the end, a synchronization during  the memory transfer is done  for the final reduction.
+Only at the end, a synchronization during  the memory transfer is done  by the final reduction.
 
@@
 \hspace{-0.5cm}
-\includegraphics[scale=0.25, angle=-90]{ME128.eps} 
+\includegraphics[scale=0.37, angle=-90]{ME128.eps} 
 \hspace{-0.4cm}
-\includegraphics[scale=0.25, angle=-90]{ME192.eps} 
-\hspace{-0.4cm}
-\includegraphics[scale=0.25, angle=-90]{ME256.eps} 
+\includegraphics[scale=0.37, angle=-90]{MC128.eps} 
+%\hspace{-0.4cm}
+%\includegraphics[scale=0.25, angle=-90]{ME256.eps} 
 }
-\caption{Left: Inner product up to 3 variables with 128 to 256 bits coefficient.
-Middle: Inner product up to 3 variables with 192 to 384 bits coefficient.
-Right: Inner product up to 3 variables with 256 to 512 bits coefficient}
+\caption{Left: Dense polynomial inner product up to 3 variables with 128 to 256 bits coefficients.
+Right: Triangular polynomial inner product up to 4 variables with 128 to 256 bits coefficient.
+The 4 variables curves miss points due to not enough memory on the node.
+Larger coefficients give the same results.
+Size of the vector 4096.}
 \label{ResME}
@@
 
-Benchmarks are performed on the facilities of The Swiss Center for Scientific Computing (CSCS), on cray xk6 and xk7 cluster for the GPU and a Sandy Bridge  node 32 logical cores  (HyperThreading on) Xeon E5-2670.
-The size of our vector is equal to 4096, the order of the polynomials varies from 1 to 14, the size of the inputs coefficients of the polynomials varies from 128 to 256 bits, the output polynomial has twice larger coefficient (256 to 192 bits).
+Benchmarks are performed on the facilities of The Swiss Center for Scientific Computing (CSCS), on cray  xk7 cluster for the GPU and a Sandy Bridge  node 32 logical cores  (HyperThreading on) Xeon E5-2670.
+The size of our vector is equal to 4096, the order of the polynomials varies from 1 to 14, the size of the inputs coefficients of the polynomials varies from 128 to 256 bits, the output polynomial has twice larger coefficient (256 to 512 bits).
 
-Evaluate  integer benchmarks integer are not easy, contrary to float number, it does not exist a universal\footnote{Altough the MIPS exists (Million Instructions Per Seconds), it should be the equivalent of the FLOPS, it's signification is disputable} value as the FLOPS for floating-point operation, because integer operations performed by ALU, and the number of  ALU is not constant between processor, and the execution of the operation are not constant.
-We compute the  result  numbers as GOP/s$=10^{-9} \times batch/t$, where  $batch$ is the number of long integer operations computed during the execution (additions and multiplications), $t$ [s]  is the elapsed time, we note the execution time for GPU contains the memory  transfer.
+Evaluate  integer benchmarks integer are not easy, contrary to float number, because integer operations performed by ALU,  the number of  ALU is not constant between processors, and the execution of the operations are  not constant \cite{ASMcost}.
+We compute the  result  numbers as operations \footnote{A operation can be a long multiplication, addition and so on}   per seconds $GOP/s=10^{-9} \times batch/t$, where  $batch$ is the number of long integer operations computed during the execution (additions and multiplications), $t$ [s]  is the elapsed time, for more realistic comparison the execution time for GPU contains the memory  transfer.
 
-Results presented on the Figure \ref{ResME} compare our CPU library to GMP (or polynomial is just instantiated with VLI of GMP),  we get an excellent speed up factor from 4 to 11.
+ The Figure \ref{ResME} presents the performance of a dense and triangular polynomials inner product.
+It compares our CPU library to GMP ( where polynomials are just instantiated with VLI or GMP),  we get an excellent speed up factor from 4 to 11.
 For two raisons, the coefficient of the  polynomial are allocated in the heap, in the case of GMP, they are allocated in the stack, it is well know better performances are obtained with less dynamic memory allocation.
-The second raison, the meta-programming allows the construction of a specific solver for every type of large integer during the compilation, contrary to GMP, we do not have a universal solver (at least for small size), the solver is unique with the maximum optimization (operations are performing in the register only).
+The second raison, the meta-programming allows the construction of a specific solver for every type of large integer during the compilation, contrary to GMP, we do not have a universal solver (as GMP  for small size), the solver is unique with the maximum optimizations.
+% (operations are performing in the register only, no stack call).
 
-The GPU curves show again an excellent speed up to factor 10 to 60 compare to GMP, but with an interesting features.
-We have disappointing results for one variable, a continuous increases of the performance for two variables, and a peak performance for
-three variables around order 8, with a significative drop of the performance.
-On the first point, polynomial on one variable have a small number of coefficient ($Order+1$), in the present condition, we used a very small number of warp, and loose all the performance
-of the GPU.
-For the second and third point is the same cause, as we said in section \ref{CUDA_PRACTICE}, a good practice of programming consists to prefetch the data on faster memory.
-On Kepler, we selected the texture memory because, it
-highly optimized for repetitive and random access contrary to shared memory.
-Its size is still limited to 48 KB maximum.
-For 3 variables the sum of the  size [Byte] of the polynomials is equal to $2(Order+1)^3\times \texttt{sizeof(VLI)}$ which corresponds approximatively
-to the full size of the texture memory.
-When the texture memory is full, memory access are not full cached and the performance droops instantaneously.
-For two variables, the input polynomials are still cached all over the range of Order, thus we get a continuous  increased
-of the performance until probably the maximum of three variables.
+The GPU curves show again an excellent speed up to factor 10 to 60 maximum compared to GMP, but with an interesting features. 
+For both kind of polynomials, the performances for one variable are low, because we have a small number of threads during the execution of the GPU algorithms.
+As the number of threads is fixed by the number of coefficient, $2n$ for a dense polynomial,  the maximum number of threads will be 28
+ which corresponds to one warp.
+We do not utilize at all the full power of the GPU.
+For two and more variables, whatever the polynomials, the performance are like a parabola.
+The position of the maximum is specific to the type of polynomial.
+We obtained the maximum of performance, if the following conditions are respected.
+The number of running threads is maximum; 256 threads, and both inputs polynomials
+are cached into the texture memory. On Kepler the size of the texture memory is equal to 48 kB, this memory  is highly optimized for repetitive and random access contrary to shared memory
+ (where bank memory conflict can appear).
+Moreover, for large order, the result polynomial have  more coefficients, so the balance of the tasks over the warp is good, and consequently the efficiency too.
 
-%The performance indicator for integer is Millions of Instructions Per Second MIPS\footnote{Also know as Meaningless Indicator of Processor Speed}.
-As 99 \% of the code is spent in the different arithmetic solver, we can easily determine
-%the MIPS of our applications.
-We counted all ASM instructions in our solver and we divided by the full execution time.
-Determine the theoretical peak of the ALU is problematic.
-There number varies from architecture, the cost
-%of integer operations are not the same on ALU [CITE COST GUY] contrary  to FPU.
-Base on the full work of [CITE BENCH GUY] , we determine experimentally the peak performance by saturating the processors
-%by additions and multiplications. We get  38.10 GMIPS for the multiplications and 87.27 GMIPS for the additions, we keep this last value as peak performance indicator.
+\begin{figure}[t!]
+\begin{center}
+\mbox{
+\hspace{-0.5cm}
+\includegraphics[scale=0.37, angle=-90]{ME128MIPS.eps} 
+\hspace{-0.4cm}
+\includegraphics[scale=0.37, angle=-90]{MC128MIPS.eps} 
+%\hspace{-0.4cm}
+%\includegraphics[scale=0.25, angle=-90]{ME256.eps} 
+}
+\caption{Dense and triangular polynomial inner product in [GIPS]. From 1 to 4 variables with 128 to 256 bits coefficients. Size of the vector 4096.}
+\label{ResMEMIPS}
+\end{center}
+\end{figure}
 
-%We determine experimentally the peak performance  by a benchmark an acceptable based on the benchmark work of [CITE BENCH GUY]
+For the small orders the number of GPU threads is lower than 256 but it will increase with the order until a maximum.
+After, for the larger orders, although the number of threads is 256, the polynomials will be not full cache into the texture memory.
+So the execution kernels calls data from the global memory that reduce the performance.
+ All this behavior show the difference between a CPU and a GPU.
+A CPU has really more transistor for the  data caching and flow control, it assures a regular and optimum access to the data.
+For the GPU,
+ the programmer manages by hand the flow control, in return, it will have more transistor for the data processing, and consequently the performance.
 
-\section{Conclusions}
+Although translate the results into MIPS is hard, we did it, at least on GPU because the GPU kernel of the extended multiplication utilizes only one PTX instruction (\texttt{madc}).
+Counting the total number of this instruction during
+the inner product, and dividing by the elapsed time, we obtain an estimation in GIPS (Giga Instruction Per Seconds).
+A  profiling of the code with CUDA profiler indicates the bottle neck of the application is the product of polynomial, so we do not take into account   the transfer of the data and the ASM instructions into the reduction.
+Therefore, we underestimate the performance of the library.
+The peak performance of  a sample card based on Kepler GK110 is evaluated easily from CUDA documentation, for 32-bit integer multiply-add on 14 SMs * 700 MHz * 32 instructions per clock per SM = 314 GIPS.
+The results are reported in the figure \ref{ResMEMIPS}, we obtain 67\% of the Kepler card, that is a good result.
+The results are better for the dense polynomial because the coefficient \texttt{GetIndices} of the GPU algorithms is more time consuming for the triangular polynomials.
 
-Turlututu chapeau pointu !  tyrlytyty pointy hat !
+\section{Conclusions and future work}
 
-\bibliographystyle{plain}	% (uses file "plain.bst")
+The VLI library gives efficient solvers for the 128 to 512 bits integer associated to polynomial with 1 to 4 variables.
+We obtain a comfortable speed up until 60 compare to GMP for the our specific range.
+We aware the comparisons to GMP were not fair.
+GMP excels in thousand bits arithmetic, where our library can not work.
+As future work, we would like to implement, new kinds of polynomials, and a better management
+of  memory to maintain the performance for  polynomial inner product with large coefficients on GPU.
+
+\bibliographystyle{llncs2e/splncs}
 \bibliography{biblio/vlibib}

### 2012-11-19 doc/paper_vli/vli.tex
VLI paper

@@
 \begin{abstract}
-The main purpose of the  VLI C++ library is to perform large integer arithmetic and polynomial inner product.
-Based on meta-programming and  hybrid (CPU/GPU) mode, it performs inner product of polynomial of 1 to 4 variables  up to order 14.
+The main purpose of the  VLI C++ library is to perform large integer arithmetic and  inner product of polynomial.
+Based on meta-programming and  hybrid (CPU/GPU) mode, it performs inner product, where polynomials have  one to four variables  up to order fourteen.
 The coefficients are represented  by large integer of 128 to 256 bits.
-We remember the basics of the inner product with polynomials, and  the assembly implementation of the  library.
+In this paper, we remember the basics of the inner product with polynomials, and  the assembly implementation of the  library.
 Then, we focus on the gpu implementation of the inner product.
@@
 The most famous example on this transition in mathematic is the famous blas/lapack libraries to MAGMA \cite{magma}, on the other side, polynomials libraries does not make the transition to GPU.
-During the last few years, authors like \cite{Govindaraju2008, Emeliyanenko:2009, Moreno2010}, write polynomial multiplication  to  this new hardware, successfully.
+During the last few years, authors like \cite{Govindaraju2008, Emeliyanenko:2009, Moreno2010}, wrote polynomial multiplication  to  this new hardware, successfully.
 All these authors focus on large polynomials,
@@
 In the present work, we are focusing on the inner product with dense and triangular multivariable polynomials.  The coefficients of the polynomials are large integer. 
-We preferred developed a long integer library  han utilize GMP, because GMP is a dynamic library.
-Moreover, we want a library natively compatible with GPU to allow hybrid calculation. GPU imposes contiguous memory which impossible with a GMP approach.
+We preferred develop a long integer library  han utilize GMP, because GMP is a dynamic library.
+Moreover, we want a library natively compatible with GPU to allow hybrid calculation. GPU imposes contiguous memory which is impossible with a GMP approach.
 Thus, we propose a static library based on contiguous memory where the exact assembly kernels are generated by meta-programming during the compilation, at the time, for CPU and GPU.
-We do not wast time in dynamic memory allocation and let the compiler optimizes deeply the library for both architectures.
+We do not wast time in the dynamic memory allocation and let the compiler optimizes deeply the library for both architectures.
 
@@
 where $\times$ indicates the multiplication between two polynomials. The multiplications occurs basic rules of polynomials multiplication. 
-A final coefficient may be the sum of several coefficients (named contributions), resulting from the sum of several multiplications, these coefficients are the cross terms. 
+A final coefficient may be the sum of several coefficients (named contributions), resulting from the sum of several multiplications. These coefficients are the cross terms. 
 The cross terms are not a problem for the cpu version, as a polynomial multiplication is done by a single thread but they are the bottleneck for the gpu version.
-The number of coefficient for a dense and triangular polynomial are given by $(n+1)^{\kappa}$ and  $(\kappa+n-1)! / ( (\kappa-1)! n! )$, respectively, where $\kappa$ is the number of variable.
+The number of coefficient for a dense and triangular polynomial are given by $(n+1)^{\kappa}$ and  $(\kappa+n-1)! / ( (\kappa-1)! n! )$, where $\kappa$ is the number of variable.
 
@@
 
-As we need large arithmetic, we applied algorithms based on childhood  experience. These methods necessitate carry\footnote{ The carry bit is a bit of the status register, as the overflow bit, 
+As we need carry\footnote{ The carry bit is a bit of the status register, as the overflow bit, 
 it informs  the overflow of an operation, but it can be re-use to propagate the carry. } for addition/subtraction and an extended multiplication for the multiplication.
@@
 
-The choice of the memory layout is important, it facilitates  the usage of the SIMD/AVX instructions. We choose the AoS for the following reasons:
-we want reuse the VLI library in other object conception code, so, we can not interleave the VLI number with the polynomial like on the figure \ref{AOSSOA}. 
+The choice of the memory layout is important. It facilitates  the usage of the SIMD/AVX instructions. We choose the AoS for the following reasons:
+we want reuse the VLI library in other object conception code, so we can not interleave the VLI number with the polynomial. 
 The data structure should only concern the very large coefficient.
@@
 the execution of several long multiplication in the same time, and we could loose the benefice of the SIMD version.
-Moreover, a large extra work is necessary to prepare the registers and shuffle the data into the registers which are expensives.
+Moreover, a large extra work is necessary to prepare the registers and shuffle the data which is expensive.
 For our range of operations the extended multiplication varies from 128 to 512 bits, a SIMD version will be not the optimum solution.
@@
 
-\section{A Short introduction on GPU}
+\section{A Short introduction to GPU}
 
@@
 Then a global reduction will sum all results polynomials.
-The division into blocks is natural, a block is a single entry of the vectors (equation \ref{InnerProduct}), all the difficulty consists of designing a parallel algorithm of the polynomial multiplication and the final reduction.
+The division into blocks is natural, a block is a single entry of the vectors (equation \ref{InnerProduct}).
+All the difficultis consists of designing a parallel algorithm of the polynomial multiplication and the final reduction.
 
@@
 
-The first idea coming in mind: calculate every single multiplication per an independent gpu thread,  indeed for a dense polynomial, the number of coefficient can be large and compatible with the number of SMs threads.
+The first idea coming in mind: calculate every single multiplication per an independent gpu thread,  indeed for a  polynomial, the number of coefficient can be large and compatible with the number of SMs threads.
 Nevertheless,  the result coefficients may have several contributions.
@@
 
-For the first step, we create a resulting polynomial  represented by the indices and the contributions. They are sorted by descending order on the number of contribution
+For the first step, we create a fictive resulting polynomial  represented by the indices and the contributions. They are sorted by descending order on the number of contribution
 and group by 32 as a task. These tasks are reordered over the warps to equilibrate the amount of work. An illustration of this algorithm is illustrated on the figure \ref{algo_gpu}. 
 More the polynomials will have coefficients  more the distribution of the tasks over the warps will be  uniform and the calculations will be efficient. 
-We fix, the maximum number of wrap to eight (8 warps - 256 threads), as far as possible, we assign at least two tasks per warps.
+We fix, the maximum number of warp to eight (8 warps - 256 threads), as far as possible, we assign at least two tasks per warps.
 Thus, in function of the number of coefficients a warp may calculate one task, whereas an other warp may have several tasks to calculate.
@@
 For every iteration over contribution, the threads determine the input coefficients, read the corresponding data in the fast memory and perform the arithmetic operations: a extended multiplication and an optional long addition (if the result coefficient has several contributions).
-The calculation of an output coefficient  are  done into fast memory (shared and texture for the input data, and register for the intermediate result).
+The calculation of an output coefficient  are  done into fast memory (texture for the input data, and register for the intermediate result).
 When the loop over contribution terminates,  we transfer the coefficients into the an intermediate buffer in the global memory, but in SoA  order to prepare the final reduction. 
@@
 
-To conclude, This  algorithm has the great advantage to work well with dense and sparse polynomials.
+To conclude, this  algorithm has the great advantage to work well with dense and sparse polynomials.
 The main difficult consists  to design an efficient function "GetIndices" that determine the contributions.
@@
 Benchmarks are performed on the facilities of The Swiss Center for Scientific Computing (CSCS), on cray  xk7 cluster for the GPU and a Sandy Bridge  node 32 logical cores  (HyperThreading on) Xeon E5-2670.
-The size of our vector is equal to 4096, the order of the polynomials varies from 1 to 14, the size of the inputs coefficients of the polynomials varies from 128 to 256 bits, the output polynomial has twice larger coefficient (256 to 512 bits).
+The size of our vector is equal to 4096, the order of the polynomials varies from 1 to 14, the size of the inputs coefficients of the polynomials varies from 128 to 256 bits, the output polynomial has twice larger coefficients (256 to 512 bits).
 
-Evaluate  integer benchmarks integer are not easy, contrary to float number, because integer operations performed by ALU,  the number of  ALU is not constant between processors, and the execution of the operations are  not constant \cite{ASMcost}.
+Evaluate  integer benchmarks integer are not easy, contrary to float number,  the integers operations are performed by ALU,  the number of  ALU  between processors, and the execution of the operations are  not constant \cite{ASMcost}.
 We compute the  result  numbers as operations \footnote{A operation can be a long multiplication, addition and so on}   per seconds $GOP/s=10^{-9} \times batch/t$, where  $batch$ is the number of long integer operations computed during the execution (additions and multiplications), $t$ [s]  is the elapsed time, for more realistic comparison the execution time for GPU contains the memory  transfer.
@@
  which corresponds to one warp.
-We do not utilize at all the full power of the GPU.
+We do not utilize the capacity of the GPU.
 For two and more variables, whatever the polynomials, the performance are like a parabola.
@@
 Although translate the results into MIPS is hard, we did it, at least on GPU because the GPU kernel of the extended multiplication utilizes only one PTX instruction (\texttt{madc}).
-Counting the total number of this instruction during
+Counting the total number of this  ASM instruction during
 the inner product, and dividing by the elapsed time, we obtain an estimation in GIPS (Giga Instruction Per Seconds).
@@
 As future work, we would like to implement, new kinds of polynomials, and a better management
-of  memory to maintain the performance for  polynomial inner product with large coefficients on GPU.
+of  memory to maintain the performance of the inner product on GPU.
 

### 2012-11-20 doc/paper_vli/Res/MIPS/mainME.cpp
paper vli

@@ -2,13 +2,14 @@
 #include <math.h>
 
 double foo( int Order, int var, int mul, int add){
-    return 4096*pow((Order+1),2*var)*mul; //  +4095*pow((2*Order+1),var)*add;
+    return 4096*pow((Order+1),2*var)*mul +4095*pow((2*Order+1),var)*add;
 }
 
 int main(){
 
 for ( int i=0; i < 15; ++i)
-     std::cout << i << " " <<  foo(i,1,32,0) << " " <<  foo(i,2,32,0) <<  " " <<  foo(i,3,32,0) << std::endl;
+     std::cout << i << " " <<  foo(i,1,32,0) << " " <<  foo(i,2,32,0)<<  " " <<  foo(i,3,32,0) <<  " " <<  foo(i,4,32,0) << std::endl;
+    // std::cout << i << " " <<  foo(i,1,1,1) << " " <<  foo(i,2,1,1)<<  " " <<  foo(i,3,1,1) <<  " " <<  foo(i,4,1,1) << std::endl;
 
 
 

### 2012-11-20 doc/paper_vli/vli.tex
paper vli
vli doc

@@
 %\include{{llncs2e}}
+\title{Very Large Integer  and polynomial inner product}
+\author{Timoth\'ee Ewart\inst{1}\thanks{Acknowledgments: Maxim Milakov, Peter Messmer  - NVIDIA, and  Williams Sawyer, Gilles Fourestey - CSCS}  , Andreas Hehn$^2$, Matthias Troyer\inst{2} and Thierry Giamarchi\inst{1}}
 
-\title{Very Large Integer  and polynomial inner product}
-\author{Timoth\'ee Ewart\inst{1}, Andreas Hehn$^2$, Matthias Troyer\inst{2} and Thierry Giamarchi\inst{1}}
-
-\institute{Universit\'e de Gen\`eve, \email{timothee.ewart@gmail.com} \thanks{Thank you : Maxim Milakov, Peter Messmer  - NVIDIA, and  Williams Sawyer, Gilles Fourestey - CSCS}  \and Eidgen\"ossische Technische Hochschule Z\"urich }
+\institute{Universit\'e de Gen\`eve, \email{timothee.ewart@gmail.com}  \and Eidgen\"ossische Technische Hochschule Z\"urich }
 
@@
 \begin{abstract}
-The main purpose of the  VLI C++ library is to perform large integer arithmetic and  inner product of polynomial.
-Based on meta-programming and  hybrid (CPU/GPU) mode, it performs inner product, where polynomials have  one to four variables  up to order fourteen.
+The main purpose of the  Very Large Integer (VLI/C++) library is to perform large integer arithmetic and  inner product of polynomial.
+Based on meta-programming and  hybrid (CPU/GPU) mode, it performs inner product  from 1 to 4 variables  up to order 14.
 The coefficients are represented  by large integer of 128 to 256 bits.
@@
 Then, we focus on the gpu implementation of the inner product.
-All these algorithms are based on brute force attacks, as the size of the  problem should not justify a FFT approach.
+All our algorithms are based on brute force attacks, as the size of the  problem should not justify a FFT approach.
 We save a factor up to 10 on CPU and 60 on GPU compared to a GMP solution.
@@
  A few libraries  LINBOX or NTL  deliver high performance in polynomial solvers.
-These libraries are usually connect ed with GMP\footnote{GMP is the standard library for long arithmetic, gmplib.org} when needed. 
-The introduction of hardware accelerator in HPC a few years ago and more recently GPU  by Nvidia introduce a revolution. It necessitates to adapt library to these new hardwares. 
-The most famous example on this transition in mathematic is the famous blas/lapack libraries to MAGMA \cite{magma}, on the other side, polynomials libraries does not make the transition to GPU.
+These libraries are  connected with GMP\footnote{GMP is the standard library for long arithmetic, gmplib.org} when needed. 
+The introduction of hardware accelerator in HPC a few years ago and more recently GPU  by Nvidia introduce a revolution in programming model. It necessitates to adapt library to these new hardwares. 
+The most famous example on this transition in mathematic is the famous blas/lapack libraries to MAGMA on the other side, polynomials libraries does not make the transition to GPU.
 During the last few years, authors like \cite{Govindaraju2008, Emeliyanenko:2009, Moreno2010}, wrote polynomial multiplication  to  this new hardware, successfully.
@@
 In the present work, we are focusing on the inner product with dense and triangular multivariable polynomials.  The coefficients of the polynomials are large integer. 
-We preferred develop a long integer library  han utilize GMP, because GMP is a dynamic library.
+We preferred develop a long integer library  than utilize GMP, because GMP is a dynamic library.
 Moreover, we want a library natively compatible with GPU to allow hybrid calculation. GPU imposes contiguous memory which is impossible with a GMP approach.
@@
 where  $x,\,y,\,z$ and $w$ are the variables,  $i,\,j,\,k$ and $l$  the indices, $a$  the coefficients, and $n$ the order of the polynomial.
-Theses polynomials are named dense and  triangular polynomial, respectively.  If now we consider two vectors of polynomials of the  same type of $m$ 
+These polynomials are named dense and  triangular polynomial, respectively.  If now we consider two vectors of polynomials of the  same type of $m$ 
 entries  $\boldsymbol{v}_1<P_a>$ and  $\boldsymbol{v}_2<P_b>$. The inner product  consists of the product 
@@
 In term of  programming and parallelization, at least for the cpu version, the previous equations can be reduced by independent loops with a final reduction.
- These work can be easily done in OpenMP. Every OpenMP threads calculate several  polynomial multiplications and make a local reduction. At the end, a final reduction is performed  by the main thread.
+The CPU parallelization has been done with OpenMP.
+The OpenMP threads calculate several  polynomial multiplications and make a local reduction.
+At the end, a final reduction is performed  by the main thread.
  % \footnote{In OpenMP reduction for class is not supported}. Results showed a perfect scalability of the OpenMP code.
@@
 The coefficients of the polynomial are represented by a large integer named VLI number.  A VLI number is characterized by a size in bits; it is a  template parameter.
-The size varies from 128 to 512 bits.  Its implementation is static, contrary to the well know GMP. 
+The size varies from 128 to 512 bits.  Its implementation is static, contrary to GMP. 
 Inside the memory, the VLI numbers are  contiguous. The data type container is an 64 bits unsigned integer array.
@@
 
-As we need carry\footnote{ The carry bit is a bit of the status register, as the overflow bit, 
+We applied algorithms based on childhood  experience for large arithmetic. These methods necessitate carry\footnote{ The carry bit is a bit of the status register, as the overflow bit, 
 it informs  the overflow of an operation, but it can be re-use to propagate the carry. } for addition/subtraction and an extended multiplication for the multiplication.
@@
 
-In assembly, these features  are supported by the pairs of mnemonic; \texttt{addq/adcq} for the addition; \texttt{sbbq/subq} for the subtraction and \texttt{mul} for the extentend multiplication.   
+In assembly, these traits  are supported by the pairs of mnemonic; \texttt{addq/adcq} for the addition; \texttt{sbbq/subq} for the subtraction and \texttt{mul} for the extentend multiplication.   
 Combining these mnemonic, we reproduce easily long arithmetic algorithm. The ASM work is based on \cite{Hyde:2003:AAL:861534}.
@@
 
-The x86-64 architecture allows 16 registers (32 for power), as much as possible, we conserved the datas into the registers when we perform a long multiplication to reduce read/write actions to the
-stack to the minimum. Additional optimizations have been done like full unrolling and assembly tips. We  supported  x86-64, Power64 and PTX (GPU) assembly.
-Moreover GPU  assembly has specificities, it does not support carry bit for 64 bits integer, but only for 32 bits integer\footnote{Introduce since CUDA 4.2}.  
+The x86-64 architecture allows 16 registers (32 for power), as much as possible, in the solver, we conserved the datas into the registers to reduce memory read/writes to the the minimum.
+ Additional optimizations have been done like full unrolling or assembly tips. We  supported  x86-64, Power64 and PTX (GPU) assembly.
+Moreover PTX has specificities, it does not support carry bit for 64 bits integer, but only for 32 bits integer\footnote{Introduce since CUDA 4.2}.  
 It allows multiplication-add 32 bits integers operations with carry bit support \cite{CUDAasm}. 
@@
 
-The choice of the memory layout is important. It facilitates  the usage of the SIMD/AVX instructions. We choose the AoS for the following reasons:
+The choice of the memory layout is important. It may facilitate  the usage of the SIMD/AVX instructions. We choose the AoS for the following reasons:
 we want reuse the VLI library in other object conception code, so we can not interleave the VLI number with the polynomial. 
 The data structure should only concern the very large coefficient.
-Apply SIMD on AoS, for an addition it is a non sense because there is no specific instruction to propagate carry between entries of the multimedia register.
+Apply SIMD on AoS, for an addition is a non sense because there is no specific instruction to propagate carry between entries of the multimedia register.
 The extended multiplication under AVX \texttt{\_\_mm256\_mul\_epu32}  allows only four multiplication of 32 bits, we save a factor two. Nevertheless, on Sandy Bridge, we have 3 ALU, we can not exclude
-the execution of several long multiplication in the same time, and we could loose the benefice of the SIMD version.
-Moreover, a large extra work is necessary to prepare the registers and shuffle the data which is expensive.
+the execution of several long multiplication in the same time, and we lose the benefice of the SIMD version.
+Moreover, a large extra work is necessary to prepare the registers and shuffle the data into the registers which are expensive.
 For our range of operations the extended multiplication varies from 128 to 512 bits, a SIMD version will be not the optimum solution.
@@
 The division into blocks is natural, a block is a single entry of the vectors (equation \ref{InnerProduct}).
-All the difficultis consists of designing a parallel algorithm of the polynomial multiplication and the final reduction.
+All the difficulties consists of designing a parallel algorithm of the polynomial multiplication and the final reduction.
 
@@
 
-The first idea coming in mind: calculate every single multiplication per an independent gpu thread,  indeed for a  polynomial, the number of coefficient can be large and compatible with the number of SMs threads.
+The first idea coming in mind: calculate every single multiplication per an independent gpu thread,  indeed for a dense polynomial, the number of coefficient can be large and compatible with the number of SMs threads.
 Nevertheless,  the result coefficients may have several contributions.
 If the contributions are calculated by several threads, it necessitates reduction between threads.
-This solution introduces   synchronizations, and lower performance.
+This solution introduces   synchronizations, and low performance.
 The peak performance will be reached, if the warps run the same amount of work, and every single thread into a warp should be responsible of the full contributions of an output coefficient.
@@
 When the loop over contribution terminates,  we transfer the coefficients into the an intermediate buffer in the global memory, but in SoA  order to prepare the final reduction. 
-The reduction is done using the standard method of Nvidia \cite{CUDAReduction} to achieve the highest performance.
+The reduction is done using the standard method of N/vidia \cite{CUDAReduction} to achieve the highest performance.
 
@@
 To conclude, this  algorithm has the great advantage to work well with dense and sparse polynomials.
-The main difficult consists  to design an efficient function "GetIndices" that determine the contributions.
- We also developed an optional hybrid mode between CPU/GPU, this  technic is usual in GPU computing \cite{magma}.
+The main difficult consists  to design an efficient function "GetIndices" that determines the contributions.
+ We also developed an optional hybrid mode between CPU/GPU, this  technic is usual in GPU computing.
 We split the inputs vectors of polynomials into two chucks of datas, execute concurrently on CPU and GPU.
@@
 Larger coefficients give the same results.
-Size of the vector 4096.}
+Size of the vector 4096.
+For four variables we are limited by the memory of the node.
+GMP gives similar results whatever the polynomials.}
 \label{ResME}
@@
 
-Evaluate  integer benchmarks integer are not easy, contrary to float number,  the integers operations are performed by ALU,  the number of  ALU  between processors, and the execution of the operations are  not constant \cite{ASMcost}.
+Evaluate  integer benchmarks  are not easy, contrary to float number integer operations are performed by ALU,  the number of  ALU  between processors and the execution time of the operations are  not constants \cite{ASMcost}.
 We compute the  result  numbers as operations \footnote{A operation can be a long multiplication, addition and so on}   per seconds $GOP/s=10^{-9} \times batch/t$, where  $batch$ is the number of long integer operations computed during the execution (additions and multiplications), $t$ [s]  is the elapsed time, for more realistic comparison the execution time for GPU contains the memory  transfer.
@@
  which corresponds to one warp.
-We do not utilize the capacity of the GPU.
+We do not utilize at all the capacity of the GPU.
 For two and more variables, whatever the polynomials, the performance are like a parabola.
@@
 }
-\caption{Dense and triangular polynomial inner product in [GIPS]. From 1 to 4 variables with 128 to 256 bits coefficients. Size of the vector 4096.}
+\caption{Inner products of  dense and triangular polynomial   in [GIPS]. From 1 to 4 variables with 128 to 256 bits coefficients. Size of the vector 4096.}
 \label{ResMEMIPS}
@@
 
-For the small orders the number of GPU threads is lower than 256 but it will increase with the order until a maximum.
-After, for the larger orders, although the number of threads is 256, the polynomials will be not full cache into the texture memory.
-So the execution kernels calls data from the global memory that reduce the performance.
- All this behavior show the difference between a CPU and a GPU.
+For the small orders the number of GPU threads is lower than 256 but it will increase with the order until the maximum.
+After, for the large orders, although the number of threads is 256, the polynomials will be not full cache into the texture memory.
+So the execution kernel calls data from the global memory that reduces the performance.
+ All this behavior shows the differences between a CPU and a GPU.
 A CPU has really more transistor for the  data caching and flow control, it assures a regular and optimum access to the data.
@@
 As future work, we would like to implement, new kinds of polynomials, and a better management
-of  memory to maintain the performance of the inner product on GPU.
+of  memory to maintain the performance on GPU.
 

### 2012-11-21 doc/paper_vli/vli.tex
VLI paper corrections

@@
 %\include{{llncs2e}}
-\title{Very Large Integer  and polynomial inner product}
+\title{Very Large Integers  and polynomial inner product}
 \author{Timoth\'ee Ewart\inst{1}\thanks{Acknowledgments: Maxim Milakov, Peter Messmer  - NVIDIA, and  Williams Sawyer, Gilles Fourestey - CSCS}  , Andreas Hehn$^2$, Matthias Troyer\inst{2} and Thierry Giamarchi\inst{1}}
@@
 \begin{abstract}
-The main purpose of the  Very Large Integer (VLI/C++) library is to perform large integer arithmetic and  inner product of polynomial.
-Based on meta-programming and  hybrid (CPU/GPU) mode, it performs inner product  from 1 to 4 variables  up to order 14.
-The coefficients are represented  by large integer of 128 to 256 bits.
-In this paper, we remember the basics of the inner product with polynomials, and  the assembly implementation of the  library.
-Then, we focus on the gpu implementation of the inner product.
-All our algorithms are based on brute force attacks, as the size of the  problem should not justify a FFT approach.
+The main purpose of the  Very Large Integer (VLI/C++) library is to perform large integer arithmetic and  inner product of polynomials. Based on meta-programming of assembly,
+and  hybrid (CPU/GPU) mode, it performs inner product of dense and triangular polynomials  of 1 to 4 variables  up to order 14. The coefficients are represented  by large integers of 128 to 256 bits. 
+In this paper, we remember the basics of the inner product with polynomials, and  we present the implementation of the library in assembly.
+Then, we focus on the GPU implementation of the inner product.
+All our algorithms are based on brute force attacks, because the problem is too small for a FFT approach to be valuable.
 We save a factor up to 10 on CPU and 60 on GPU compared to a GMP solution.
+We get 67 \% of the peak performance of the GPU.
 \end{abstract}
 \section{Introduction}
- A few libraries  LINBOX or NTL  deliver high performance in polynomial solvers.
+ A few libraries like  LINBOX or NTL  provide high performance in polynomial solvers.
 These libraries are  connected with GMP\footnote{GMP is the standard library for long arithmetic, gmplib.org} when needed. 
-The introduction of hardware accelerator in HPC a few years ago and more recently GPU  by Nvidia introduce a revolution in programming model. It necessitates to adapt library to these new hardwares. 
-The most famous example on this transition in mathematic is the famous blas/lapack libraries to MAGMA on the other side, polynomials libraries does not make the transition to GPU.
-During the last few years, authors like \cite{Govindaraju2008, Emeliyanenko:2009, Moreno2010}, wrote polynomial multiplication  to  this new hardware, successfully.
-All these authors focus on large polynomials,
- with a FFT approach.
+The introduction of hardware accelerators in HPC a few years ago, and more recently of GPU  by NVIDIA, introduce a revolution in the programming models. 
+It is therefore necessary  to adapt the libraries to these new hardwares.  The most famous example on this transition is the famous blas/lapack libraries to MAGMA. 
+Polynomials libraries, however  did not yet  make the transition to GPU.
+During the last few years, the polynomial multiplication was successfully implement to this new hardware
+\cite{Govindaraju2008, Emeliyanenko:2009, Moreno2010}. All these works focus on large polynomials,
+ using a FFT approach.
 
-In the present work, we are focusing on the inner product with dense and triangular multivariable polynomials.  The coefficients of the polynomials are large integer. 
-We preferred develop a long integer library  than utilize GMP, because GMP is a dynamic library.
+In the present work, we are focusing on the inner product of dense and triangular multivariable polynomials.  The coefficients of the polynomials are large integers. 
+We preferred to develop a long integer library  rather than utilizing GMP, because GMP is a dynamic library.
 Moreover, we want a library natively compatible with GPU to allow hybrid calculation. GPU imposes contiguous memory which is impossible with a GMP approach.
-Thus, we propose a static library based on contiguous memory where the exact assembly kernels are generated by meta-programming during the compilation, at the time, for CPU and GPU.
-We do not wast time in the dynamic memory allocation and let the compiler optimizes deeply the library for both architectures.
+Thus, we propose a static library based on contiguous memory where the exact assembly kernels are generated by meta-programming during the compilation for CPU and GPU.
+We do not waste time in the dynamic memory allocation, and let the compiler optimize deeply the library for both architectures.
 
-\section{Polynomial and Inner product definitions}
+\section{Polynomial and inner product definitions}
 
-The library performed inner product of vector of polynomials from one to four variables where polynomials are defined like (four variables) :
+The library performs the inner product of vectors of polynomials from one to four variables, where polynomials are defined like (four variables) :
 \begin{eqnarray}
-P_a(x,y,z,w) & = & \sum_i^n \sum_j^n  \sum_k^n \sum_l^n  a_{ijkl} x^i y^j z^k w^l , \,\,\,\,\,\, \text{or} \\
-                   & = & \sum_i^n \sum_j^{(n-i)} \sum_k^{(n-i-j)} \sum_l^{(n-i-j-k)}  a_{ijkl} x^i y^j z^k w^l  .
+P_a(x,y,z,w) & = & \sum_{i=0}^n \sum_{j=0}^n  \sum_{k=0}^n \sum_{l=0}^n  a_{ijkl} x^i y^j z^k w^l , \,\,\,\,\,\, \text{or} \\
+                   & = & \sum_{i=0}^n \sum_{j=0}^{(n-i)} \sum_{k=0}^{(n-i-j)} \sum_{l=0}^{(n-i-j-k)}  a_{ijkl} x^i y^j z^k w^l  .
 \end{eqnarray}
 where  $x,\,y,\,z$ and $w$ are the variables,  $i,\,j,\,k$ and $l$  the indices, $a$  the coefficients, and $n$ the order of the polynomial.
-These polynomials are named dense and  triangular polynomial, respectively.  If now we consider two vectors of polynomials of the  same type of $m$ 
-entries  $\boldsymbol{v}_1<P_a>$ and  $\boldsymbol{v}_2<P_b>$. The inner product  consists of the product 
-of every entries  associated to a global reduction. 
+The polynomials of the form  (1) and (2) are named dense and triangular respectively. Consider now two vectors of polynomials of the  same type with $m$ 
+entries  $\boldsymbol{v}_1<P_a>$ and  $\boldsymbol{v}_2<P_b>$. The inner product  consists in the product 
+of every entries  associated with a global reduction. 
  \begin{eqnarray}
@@
 \end{eqnarray}
-where $\times$ indicates the multiplication between two polynomials. The multiplications occurs basic rules of polynomials multiplication. 
+where $\times$ indicates the multiplication between two polynomials. The multiplications use basic rules of polynomials multiplication. 
 A final coefficient may be the sum of several coefficients (named contributions), resulting from the sum of several multiplications. These coefficients are the cross terms. 
-The cross terms are not a problem for the cpu version, as a polynomial multiplication is done by a single thread but they are the bottleneck for the gpu version.
-The number of coefficient for a dense and triangular polynomial are given by $(n+1)^{\kappa}$ and  $(\kappa+n-1)! / ( (\kappa-1)! n! )$, where $\kappa$ is the number of variable.
+The cross terms are not a problem for the CPU version, as a polynomial multiplication is done by a single thread, but they are the bottleneck for the GPU version.
+The number of coefficient for a dense and triangular polynomial is given by $(n+1)^{\kappa}$ and  $(\kappa+n-1)! / ( (\kappa-1)! n! )$ respectively, where $\kappa$ is the number of variables.
 
-In term of  programming and parallelization, at least for the cpu version, the previous equations can be reduced by independent loops with a final reduction.
+In terms of  programming and parallelization, at least for the CPU version, the previous equations can be reduced to independent loops with a final reduction.
 The CPU parallelization has been done with OpenMP.
@@
 
-The coefficients of the polynomial are represented by a large integer named VLI number.  A VLI number is characterized by a size in bits; it is a  template parameter.
-The size varies from 128 to 512 bits.  Its implementation is static, contrary to GMP. 
-Inside the memory, the VLI numbers are  contiguous. The data type container is an 64 bits unsigned integer array.
+The coefficients of the polynomials are represented by a large integer named VLI number.  A VLI number is characterized by a size in bits; the size is a  template parameter, which
+varies from 128 to 512 bits.  Its implementation is static, contrary to GMP. 
+Inside the memory, the VLI numbers are  contiguous. The data type container is a 64 bits unsigned integer array.
 
-The basic operations supported by VLI are the four basic operations : $+,-,\times, /$, extended multiplications and bit shift operations.
-The library allows  operations between operands of the same type, and  integer.
-For performance issue, we have a specific solver for every operations except  the divisions.
-All the algorithms and operations are based on the childhood methods but the realizations has been done in assembly,
-because, the assembly instructions facilitate the implementation and boost the performances.
+The basic operations supported by VLI are the four basic operations: $+,-,\times, /$, extended multiplications and bit shift.
+The library allows  operations between operands of the same type, and with basic  integer.
+For performance reasons, we have a specific solver for every operation except  the division.
+All the algorithms and operations are based on the elementary  methods that we learn
+in our the childhood but the realizations has been done in assembly,
+because the assembly instructions facilitate the implementation and boost the performances.
 
@@
 
-We applied algorithms based on childhood  experience for large arithmetic. These methods necessitate carry\footnote{ The carry bit is a bit of the status register, as the overflow bit, 
-it informs  the overflow of an operation, but it can be re-use to propagate the carry. } for addition/subtraction and an extended multiplication for the multiplication.
-Although the hardware supports them; we can not access to these features by the C++ language, the only possibility is the assembly.
+We applied algorithms based on childhood  experience for large arithmetic. These methods necessitate carry\footnote{ The carry bit is a bit of the register of status, as the overflow bit, 
+it informs  the overflow of an operation, but it can be re-used to propagate the carry. } for addition/subtraction and an extended multiplication for the multiplication.
+Although the hardware supports them, we can not access  these features by the C++ language. The only possibility is the assembly.
 
-In assembly, these traits  are supported by the pairs of mnemonic; \texttt{addq/adcq} for the addition; \texttt{sbbq/subq} for the subtraction and \texttt{mul} for the extentend multiplication.   
-Combining these mnemonic, we reproduce easily long arithmetic algorithm. The ASM work is based on \cite{Hyde:2003:AAL:861534}.
+In assembly, these operations  are supported by the pair of mnemonics; \texttt{addq/adcq} for the addition; \texttt{sbbq/subq} for the subtraction, and \texttt{mul} for the extentend multiplication.
+Combining these mnemonics, we reproduce easily long arithmetic algorithm, as described into the famous  ASM book of \cite{Hyde:2003:AAL:861534}.
 
-We coded the assembly using the GNU inline assembly, it consists of a string between ASM tag.
-Generate by hand all these  assembly strings (kernels)  can be painful and a source of error \footnote{The library has near 8000 lines of assembly cumulated between x86-64, Power64, and PTX} 
-We designed a generator of assembly. It creates, all versions of a kernels during the compilation.   
-It is built on the boost preprocessor package, because we can create, manipulate the strings during the pre-compilation, 
-and generate a specific tuned kernel for every large integer. All kernels use this procedure except the extended multiplication for the cpu.
+We coded the assembly using the GNU inline assembly which consists of strings between ASM tags.
+Generating by hand all these  assembly strings (kernels)  can be painful and a source of error \footnote{The library has near 8000 lines of assembly cumulated between x86-64, Power64, and PTX} 
+We designed a generator of assembly. It creates all versions of a kernel during the compilation.   
+It is built on the boost preprocessor package, because we can thus create and manipulate the strings during the pre-compilation, 
+and generate a specifically tuned kernel for every large integer. All kernels use this procedure except the extended multiplication for the CPU.
 
-The x86-64 architecture allows 16 registers (32 for power), as much as possible, in the solver, we conserved the datas into the registers to reduce memory read/writes to the the minimum.
+The x86-64 architecture allows 16 registers (32 for power). As much as possible, in the solver, we conserve the data in the registers in order to reduce memory read/writes to the  minimum.
  Additional optimizations have been done like full unrolling or assembly tips. We  supported  x86-64, Power64 and PTX (GPU) assembly.
-Moreover PTX has specificities, it does not support carry bit for 64 bits integer, but only for 32 bits integer\footnote{Introduce since CUDA 4.2}.  
-It allows multiplication-add 32 bits integers operations with carry bit support \cite{CUDAasm}. 
-Work in 32 bits mode imposes twice more operations for the additions/substractions and one order of magnitude more for the multiplications.
+ PTX does not support carry bit for 64 bits integer, but only for 32 bits integer\footnote{Introduced since CUDA 4.2}.  
+It allows multiplication-addition of 32 bits integers  with carry bit support \cite{CUDAasm}. 
+Work in 32 bits mode imposes twice as much operations for the additions/substractions, and one order of magnitude more for the multiplications.
 
@@
 
-It exists at least two data layouts for this problem: the array of structure (AoS) and structure of array (SoA), both have  advantages and disadvantages.
-In our problem the AoS consists of saving the data in this order  : VLI - Series of VLI (polynomial coefficient)- series of polynomial (the vector).
+There exists at least two data layouts for this problem: the array of structure (AoS) and structure of array (SoA).
+Both have  advantages and disadvantages.
+In our problem the AoS consists in saving the data in this order : VLI - series of VLI (polynomial coefficient)- series of polynomials (the vector).
 The SoA structure interleaves the coefficients of the VLI and the polynomials:
@@
  \underbrace{ \underbrace{a^0_{00} | a^1_{00} | a^2_{00}}_{VLI, a_{00} } || \underbrace{a^0_{10} | a^1_{10} | a^2_{10}}_{VLI, a_{10}} ||   \dots  }_{1^{st} \textrm{polynomial}} 
-      |||    \underbrace{ \underbrace{b^0_{00} | a^1_{00} | b^2_{00}}_{VLI, b_{00} } || \underbrace{b^0_{10} | b^1_{10} | b^2_{10}}_{VLI, b_{10}} ||   \dots  }_{2^{nd} \textrm{polynomial} } }_{\textrm{Vector of polynomial, AoS order}} \nonumber
+      |||    \underbrace{ \underbrace{b^0_{00} | a^1_{00} | b^2_{00}}_{VLI, b_{00} } || \underbrace{b^0_{10} | b^1_{10} | b^2_{10}}_{VLI, b_{10}} ||   \dots  }_{2^{nd} \textrm{polynomial} } }_{\textrm{Vector of polynomials, AoS order}} \nonumber
 }
@@
  \underbrace{a^0_{00} | a^0_{10} | \dots || a^1_{00} | a^1_{10} | \dots  }_{1st  \textrm{ nested VLI/Polynomial} } |||  \underbrace{b^0_{00} | b^0_{10} | \dots || b^1_{00} | b^1_{10} | \dots  }_{2nd  \textrm{ nested VLI/Polynomial} }
- }_{\textrm{Vector of polynomial, SoA Order}} \nonumber }
+ }_{\textrm{Vector of polynomials, SoA Order}} \nonumber }
 \end{eqnarray}
@@
 The choice of the memory layout is important. It may facilitate  the usage of the SIMD/AVX instructions. We choose the AoS for the following reasons:
-we want reuse the VLI library in other object conception code, so we can not interleave the VLI number with the polynomial. 
-The data structure should only concern the very large coefficient.
+we want to reuse the VLI library in other objects conception code, so we can not interleave the VLI number with the polynomial. 
  Apply SIMD on AoS, for an addition is a non sense because there is no specific instruction to propagate carry between entries of the multimedia register.
-The extended multiplication under AVX \texttt{\_\_mm256\_mul\_epu32}  allows only four multiplication of 32 bits, we save a factor two. Nevertheless, on Sandy Bridge, we have 3 ALU, we can not exclude
+The extended multiplication under AVX \texttt{\_\_mm256\_mul\_epu32}  allows only four multiplications of 32 bits, we save a factor two.
+Nevertheless, on Sandy Bridge, we have 3 ALU, we can not exclude
 the execution of several long multiplication in the same time, and we lose the benefice of the SIMD version.
-Moreover, a large extra work is necessary to prepare the registers and shuffle the data into the registers which are expensive.
-For our range of operations the extended multiplication varies from 128 to 512 bits, a SIMD version will be not the optimum solution.
-Though, it has been used with success for larg integer ($>$1024 bits) \cite{SIMD}.
+Moreover, a large extra work is necessary to prepare the registers and shuffle the data into the registers which is expensive.
+For our range of operations the extended multiplication varies from 128 to 512 bits, a SIMD version will not be the optimal solution.
+Though, it has been used with success for large integers ($>$1024 bits) \cite{SIMD}.
 
-\section{A Short introduction to GPU}
+\section{A short introduction to GPU}
 
-A Graphical Portable Unit  is efficient  to address problems that can be expressed as data-parallel computations Ð the same program 
-is executed on many data elements in parallel. Successful usage of GPU in HPC necessitates a good understanding of their design to reach great performance. A GPU is composed 
+A Graphical Portable Unit  is efficient  to address problems which can be expressed as data-parallel computations Ð the same program 
+is executed on many data elements in parallel. Successful usage of GPU in HPC requires a good understanding of their design to reach great performance. A GPU is composed 
 of several independent calculation  entities named streaming multiprocessors (SMs typically a tenth), where the SMs run hundreds of threads concurrently (typically 256) by group of 32 (warp).
@@
 
-In terms of memory all SMs  share  a large  memory (several GigaBytes) with  low performance; approximatively one hundred cycle for read/write operations.
-Inside each SMs additional specials memories (constant, shared, texture, surface) are available. These memories are characterized by their behavior, they can be read only, shared, cache and more.  
-They have still a common characteristic, a very high performance for read/write  operations in a few cycles only.
-A good programming practice is to preload data from global memory to shared/texture at once, for subsequent computations.  \label{CUDA_PRACTICE}
+In terms of memory, all SMs  share  a large  memory (several GigaBytes) with  low performance; approximatively one hundred cycles for read/write operations.
+Inside each SM additional special memories (constant, shared, texture, surface) are available. These memories are characterized by their behavior: they can be read only, shared, cache and more.  
+They have still a common characteristics, a very high performance for read/write  operations in a few cycles only.
+A good programming practice is to preload data from global memory in shared/texture at once, for subsequent computations.  \label{CUDA_PRACTICE}
 
-Apply the GPU ideology on the polynomial inner product  means the polynomial multiplications will be execute in parallel on a SMs.
-Then a global reduction will sum all results polynomials.
-The division into blocks is natural, a block is a single entry of the vectors (equation \ref{InnerProduct}).
-All the difficulties consists of designing a parallel algorithm of the polynomial multiplication and the final reduction.
+Applying the GPU ideology to the polynomial inner product  means that the polynomial multiplications will be executed in parallel on a SM.
+Then a global reduction will sum all resulting polynomials.
+The division into blocks is natural, a block being a single entry of the vector (equation \ref{InnerProduct}).
+All the difficulties consists in designing a parallel algorithm for the polynomial multiplication and the final reduction.
 
-\subsection{Inner product booster on gpu}
+\subsection{Inner product booster on GPU}
 
@@
 \caption{Left: Results coefficients (red line) sorted in reverse order and group by warp. Right: Balance of the tasks over the warps.}
-\label{algo_gpu}
+\label{algo_GPU}
 \end{center}
@@
 
-The first idea coming in mind: calculate every single multiplication per an independent gpu thread,  indeed for a dense polynomial, the number of coefficient can be large and compatible with the number of SMs threads.
+The first idea coming in mind: calculate every single multiplication on an independent GPU thread.
+Indeed for a dense polynomial, the number of coefficients can be large and compatible with the number of SMs threads.
 Nevertheless,  the result coefficients may have several contributions.
-If the contributions are calculated by several threads, it necessitates reduction between threads.
-This solution introduces   synchronizations, and low performance.
-The peak performance will be reached, if the warps run the same amount of work, and every single thread into a warp should be responsible of the full contributions of an output coefficient.
-To achieve this goal, we defined a task object.
+If the contributions are calculated by several threads, this requires reduction between threads.
+This solution introduces   synchronizations, and leads to low performance.
+The peak performance will be reached if the warps run the same amount of work, and every single thread in a warp is responsible for all contributions to an output coefficient.
+In order to achieve this goal, we defined a task object.
 A task is a set of 32 result coefficients in a specific order.
-A task tells a  warp how calculates coefficient (the thread number one of the warp takes the first coefficient of the task and so on).
+A task tells a  warp how to calculate coefficients (the thread number one of the warp takes the first coefficient of the task and so on).
 The warps may not have the same number of tasks, as we equilibrate the workload.
-We proceeded in two  steps; first on cpu, where a preliminary work is done, we prepare all informations on the tasks, and then, the calculation on GPU.
+We proceed in two  steps: first on CPU, where a preliminary work is done, we prepare all information on the tasks, and then, the calculation on GPU.
 
-For the first step, we create a fictive resulting polynomial  represented by the indices and the contributions. They are sorted by descending order on the number of contribution
-and group by 32 as a task. These tasks are reordered over the warps to equilibrate the amount of work. An illustration of this algorithm is illustrated on the figure \ref{algo_gpu}. 
-More the polynomials will have coefficients  more the distribution of the tasks over the warps will be  uniform and the calculations will be efficient. 
-We fix, the maximum number of warp to eight (8 warps - 256 threads), as far as possible, we assign at least two tasks per warps.
-Thus, in function of the number of coefficients a warp may calculate one task, whereas an other warp may have several tasks to calculate.
- When the determination of the task's list is over,  we transfer  asynchronously the tasks lists in the global memory.
-This method does not  change the order of the datas. As for the CPU version, we have an AoS order.
+For the first step, we create a fictive resulting polynomial  represented by the indices and the contributions. They are sorted by descending order of the number of contributions
+and grouped by 32 as a task. These tasks are reordered over the warps to equilibrate the amount of work. An illustration of this algorithm is given on  figure \ref{algo_GPU}. 
+The more coefficients the polynomials have, the more uniform the distribution of the tasks over the warps will be, and the more efficient the calculation.
+We fix the maximum number of warp to eight (8 warps - 256 threads), as far as possible, we assign at least two tasks per warp.
+Thus, depending on  the number of coefficients a warp may calculate one task, whereas another warp may have several tasks to calculate.
+ When the determination of the task's list is over,  we transfer  asynchronously the tasks list to the global memory.
+This method does not  change the ordering of the data. As for the CPU version, we have an AoS order.
  \begin{table}[t] 
@@
                            \begin{tabular}{c l l} 
-                               \tiny{1:} & VLI tmp   &  $\vartriangleright$ Specific to every threads  \\
-                               \tiny{2:} &  \textbf{for} over number of task list size &$\vartriangleright$  Depend of the warp \\
+                               \tiny{1:} & VLI tmp   &  $\vartriangleright$ Specific to every thread  \\
+                               \tiny{2:} &  \textbf{for} over number of task list size &$\vartriangleright$  Depend on the warp \\
                                \tiny{3:} &  tmp $\,\, \wedge=$ tmp     & $\vartriangleright$ Flush to 0 \\    
                                \tiny{4:} &   VLI ResCoeff &$\vartriangleright$  Thread gets a result coefficient from a task \\
-                               \tiny{5:} &  \hspace{0.2 cm}  \textbf{for} over the number of contribution  & $\vartriangleright$ Each thread loop over its coefficient  \\
+                               \tiny{5:} &  \hspace{0.2 cm}  \textbf{for} over the number of contributions  & $\vartriangleright$ Each thread loops over its coefficients \\
                                \tiny{6:} &  \hspace{0.4 cm}   Indices$_i$ = GetIndices(ResCoeff) &$\vartriangleright$ Determine input indices \\
@@
 
-This task's list job is executed only once for all polynomial types,  we limit the overhead for the first inner product. As soon as,  all task's list are calculated, 
-we transfer the datas of the polynomials asynchronously on the device (the global memory of the GPU is allocated once, we limit the workload of the cuda free/malloc),
+This task's list job is executed for all polynomial types only once. All the informations are kept into the global memory and reuse in case of.  As soon as,  all task's lists are calculated, 
+we transfer the datas of the polynomials asynchronously on the device (the global memory of the GPU is allocated only once and reuse),
  and the calculation of all polynomial multiplication start.
@@
  First, both input polynomials  are cached into the texture memory. 
- %faster memory (shared or texture) in function of there size. 
-% we privileged  the texture memory on Kepler architecture. %, and the shared memory on the Fermi architecture.
-When a polynomial multiplication begins,  every warp get corresponding tasks (at least one) from the task's list and loop over its number of task. Then,
-the threads into a warp gets their respective output coefficients and iterate over the number of  contribution.
-For every iteration over contribution, the threads determine the input coefficients, read the corresponding data in the fast memory and perform the arithmetic operations: a extended multiplication and an optional long addition (if the result coefficient has several contributions).
-The calculation of an output coefficient  are  done into fast memory (texture for the input data, and register for the intermediate result).
-When the loop over contribution terminates,  we transfer the coefficients into the an intermediate buffer in the global memory, but in SoA  order to prepare the final reduction. 
-The reduction is done using the standard method of N/vidia \cite{CUDAReduction} to achieve the highest performance.
+When a polynomial multiplication begins,  every warp gets the corresponding tasks (at least one) from the task's list and loops over its number of tasks. Then,
+the threads into a warp get their respective output coefficients and iterate over the number of  contributions.
+For every iteration over contributions, the threads determine the input coefficients, read the corresponding data in the fast memory ,and perform the arithmetic operations: an extended multiplication and an optional long addition (if the result coefficient has several contributions).
+The calculation of an output coefficient  is  done into fast memory (texture for the input data, and register for the intermediate result).
+When the loop over contributions terminates,  we transfer the coefficients into the an intermediate buffer in the global memory, but in SoA  order to prepare the final reduction. 
+The reduction is done using the standard method of NVIDIA \cite{CUDAReduction} to achieve the highest performance.
 
@@
 
-To conclude, this  algorithm has the great advantage to work well with dense and sparse polynomials.
-The main difficult consists  to design an efficient function "GetIndices" that determines the contributions.
- We also developed an optional hybrid mode between CPU/GPU, this  technic is usual in GPU computing.
-We split the inputs vectors of polynomials into two chucks of datas, execute concurrently on CPU and GPU.
+To conclude, this  algorithm has the great advantage of working well with dense and sparse polynomials. 
+The main difficulty consists  in designing an efficient function GetIndices which determines the contributions.
+ We also developed an optional hybrid mode between CPU/GPU. This  technique is usual in GPU computing.
+  We split the input vectors of polynomials into two chuncks of data, execute concurrently on CPU and GPU. 
   Only at the end, a synchronization during  the memory transfer is done  by the final reduction.
@@
 }
-\caption{Left: Dense polynomial inner product up to 3 variables with 128 to 256 bits coefficients.
-Right: Triangular polynomial inner product up to 4 variables with 128 to 256 bits coefficient.
-The 4 variables curves miss points due to not enough memory on the node.
-Larger coefficients give the same results.
+\caption{Left: Inner product of dense polynomials with up to 3 variables and 126 to 256 bits coefficients.
+Right: Inner product of triangular polynomials with up to 3 variables and 126 to 256 bits coefficients.
 Size of the vector 4096.
@@
 
-Benchmarks are performed on the facilities of The Swiss Center for Scientific Computing (CSCS), on cray  xk7 cluster for the GPU and a Sandy Bridge  node 32 logical cores  (HyperThreading on) Xeon E5-2670.
-The size of our vector is equal to 4096, the order of the polynomials varies from 1 to 14, the size of the inputs coefficients of the polynomials varies from 128 to 256 bits, the output polynomial has twice larger coefficients (256 to 512 bits).
+Benchmarks are performed on the facilities of the Swiss Center for Scientific Computing (CSCS), on a Cray  xk7 cluster for the GPU and a Sandy Bridge  node with 32 logical cores  (HyperThreading on) Xeon E5-2670.
+The size of our vector is equal to 4096, the order of the polynomials varies from 1 to 14, the size of the input coefficients of the polynomials varies from 128 to 256 bits, the output polynomial has twice larger coefficients (256 to 512 bits).
 
-Evaluate  integer benchmarks  are not easy, contrary to float number integer operations are performed by ALU,  the number of  ALU  between processors and the execution time of the operations are  not constants \cite{ASMcost}.
-We compute the  result  numbers as operations \footnote{A operation can be a long multiplication, addition and so on}   per seconds $GOP/s=10^{-9} \times batch/t$, where  $batch$ is the number of long integer operations computed during the execution (additions and multiplications), $t$ [s]  is the elapsed time, for more realistic comparison the execution time for GPU contains the memory  transfer.
+Evaluating  integer benchmarks  is not easy, contrary to float number integer operations are performed by ALU (and not on the single FPU),  the number of  ALU  between processors and the execution time of the operations are  not constants \cite{ASMcost} (constant for FPU).
+We compute the  result  numbers as operations \footnote{An operation can be a long multiplication, addition and so on}   per seconds $GOP/s=10^{-9} \times batch/t$, where  $batch$ is the number of long integer operations computed during the execution (additions and multiplications), $t$ [s]  is the elapsed time.
+For more realistic comparison the execution time for GPU contains the memory  transfer.
 
- The Figure \ref{ResME} presents the performance of a dense and triangular polynomials inner product.
-It compares our CPU library to GMP ( where polynomials are just instantiated with VLI or GMP),  we get an excellent speed up factor from 4 to 11.
-For two raisons, the coefficient of the  polynomial are allocated in the heap, in the case of GMP, they are allocated in the stack, it is well know better performances are obtained with less dynamic memory allocation.
-The second raison, the meta-programming allows the construction of a specific solver for every type of large integer during the compilation, contrary to GMP, we do not have a universal solver (as GMP  for small size), the solver is unique with the maximum optimizations.
+  Figure \ref{ResME} presents the performance of a dense and triangular polynomials inner product.
+It compares our CPU library to GMP (where polynomials are just instantiated with VLI or GMP).
+We get an excellent speed up factor from 4 to 11, for two raisons.
+Coefficients of the  polynomial are allocated in the heap, while in the case of GMP, they are allocated in the stack.
+It is well know that better performances are obtained with less dynamic memory allocation.
+The second reason is that  the meta-programming allows the construction of a specific solver for every type of large integer during the compilation, contrary to GMP, we do not have a universal solver (as GMP  for small size), the solver being unique with the maximum optimizations.
 % (operations are performing in the register only, no stack call).
 
-The GPU curves show again an excellent speed up to factor 10 to 60 maximum compared to GMP, but with an interesting features. 
-For both kind of polynomials, the performances for one variable are low, because we have a small number of threads during the execution of the GPU algorithms.
-As the number of threads is fixed by the number of coefficient, $2n$ for a dense polynomial,  the maximum number of threads will be 28
+The GPU curves show again an excellent speed up  factor of 10 to 60 maximum, compared to GMP, but with an interesting feature. 
+For both kind of polynomials, the performance for one variable is low, because we have a small number of threads during the execution of the GPU algorithms.
+As the number of threads is fixed by the number of coefficients, $2n$ for a dense polynomial,  the maximum number of threads will be 28
  which corresponds to one warp.
 We do not utilize at all the capacity of the GPU.
-For two and more variables, whatever the polynomials, the performance are like a parabola.
-The position of the maximum is specific to the type of polynomial.
-We obtained the maximum of performance, if the following conditions are respected.
-The number of running threads is maximum; 256 threads, and both inputs polynomials
-are cached into the texture memory. On Kepler the size of the texture memory is equal to 48 kB, this memory  is highly optimized for repetitive and random access contrary to shared memory
+For two and more variables, whatever the polynomials, the performance is  like a parabola.
+The position of the maximum depends on  to the type of polynomial.
+We obtained the maximum of performance  if the following conditions are satisfied.
+The number of running threads is maximum; (256 threads), and both input
+  polynomials  are cached into the texture memory.
+On Kepler the size of the texture memory is equal to 48 kB.
+This memory  is highly optimized for repetitive and random access contrary to shared memory
  (where bank memory conflict can appear).
-Moreover, for large order, the result polynomial have  more coefficients, so the balance of the tasks over the warp is good, and consequently the efficiency too.
+Moreover, for large order, the result polynomials have  more coefficients, so the balance of the tasks over the warp is good, and consequently the efficiency too.
 
@@
 After, for the large orders, although the number of threads is 256, the polynomials will be not full cache into the texture memory.
-So the execution kernel calls data from the global memory that reduces the performance.
+So the execution kernel calls data from the global memory which reduces the performance.
  All this behavior shows the differences between a CPU and a GPU.
-A CPU has really more transistor for the  data caching and flow control, it assures a regular and optimum access to the data.
+A CPU has really more transistor for the  data caching and flow control, it ensures a regular and optimum access to the data.
 For the GPU,
- the programmer manages by hand the flow control, in return, it will have more transistor for the data processing, and consequently the performance.
+ the programmer manages by hand the flow control. In return, it will have more transistor for the data processing, and consequently the performance.
 
-Although translate the results into MIPS is hard, we did it, at least on GPU because the GPU kernel of the extended multiplication utilizes only one PTX instruction (\texttt{madc}).
-Counting the total number of this ASM instruction during
-the inner product, and dividing by the elapsed time, we obtain an estimation in GIPS (Giga Instruction Per Seconds).
-A  profiling of the code with CUDA profiler indicates the bottle neck of the application is the product of polynomial, so we do not take into account   the transfer of the data and the ASM instructions into the reduction.
+Although translating the results into MIPS is difficult, we did it, at least on GPU because the GPU kernel of the extended multiplication utilizes only one PTX instruction (\texttt{madc}).
+Counting the total number of  ASM instructions during  the inner product, and dividing by the elapsed time, we obtain an estimation in GIPS (Giga Instructions Per Second).
+A  profiling of the code with the CUDA profiler indicates that the bottleneck of the application is the product of polynomials, so we do not take into account   the transfer of the data and the ASM belonging  to the reduction.
 Therefore, we underestimate the performance of the library.
-The peak performance of  a sample card based on Kepler GK110 is evaluated easily from CUDA documentation, for 32-bit integer multiply-add on 14 SMs * 700 MHz * 32 instructions per clock per SM = 314 GIPS.
-The results are reported in the figure \ref{ResMEMIPS}, we obtain 67\% of the Kepler card, that is a good result.
-The results are better for the dense polynomial because the coefficient \texttt{GetIndices} of the GPU algorithms is more time consuming for the triangular polynomials.
+The peak performance of  a sample card based on Kepler GK110 is evaluated easily from CUDA documentation: for 32-bit integer multiply-add on 14 SMs * 700 MHz * 32 instructions per clock per SM = 314 GIPS.
+The results are reported in the figure \ref{ResMEMIPS}.
+We obtain 67\% of the Kepler card, which is a good result.
+The results are better for the dense polynomials because the function \texttt{GetIndices}  of the GPU algorithms is more time consuming for the triangular polynomials.
 
@@
 
-The VLI library gives efficient solvers for the 128 to 512 bits integer associated to polynomial with 1 to 4 variables.
-We obtain a comfortable speed up until 60 compare to GMP for the our specific range.
+The VLI library provides efficient solvers for the 128 to 512 bits integer associated with polynomials of 1 to 4 variables. We obtain a comfortable speed up as high as  60 compared 
+to GMP for the our specific range.
 We aware the comparisons to GMP were not fair.
-GMP excels in thousand bits arithmetic, where our library can not work.
+We are working with a hundred of bits without dynamic memory allocation,  while GMP excels in thousand bits arithmetic, where our library can not work.
 As future work, we would like to implement, new kinds of polynomials, and a better management
-of  memory to maintain the performance on GPU.
+of  the memory to maintain the performance on GPU.
 

### 2012-11-22 doc/paper_vli/vli.tex
doc vli

@@
 it informs  the overflow of an operation, but it can be re-used to propagate the carry. } for addition/subtraction and an extended multiplication for the multiplication.
-Although the hardware supports them, we can not access  these features by the C++ language. The only possibility is the assembly.
+Although the hardware supports them, we can not access  these features by the C++ language. The only possibility is the assembly language.
 
-In assembly, these operations  are supported by the pair of mnemonics; \texttt{addq/adcq} for the addition; \texttt{sbbq/subq} for the subtraction, and \texttt{mul} for the extentend multiplication.
+In assembly, these operations  are supported by the pair of mnemonics; \texttt{addq/adcq} for the addition; \texttt{sbbq/subq} for the subtraction, and \texttt{mul} for the extended multiplication.   
 Combining these mnemonics, we reproduce easily long arithmetic algorithm, as described into the famous  ASM book of \cite{Hyde:2003:AAL:861534}.
@@
 We coded the assembly using the GNU inline assembly which consists of strings between ASM tags.
-Generating by hand all these  assembly strings (kernels)  can be painful and a source of error \footnote{The library has near 8000 lines of assembly cumulated between x86-64, Power64, and PTX} 
+Generating by hand all these  assembly strings (kernels)  can be painful and a source of error\footnote{The library has near 10000 lines of assembly cumulated between x86-64, Power64, and PTX} 
 We designed a generator of assembly. It creates all versions of a kernel during the compilation.   
-It is built on the boost preprocessor package, because we can thus create and manipulate the strings during the pre-compilation, 
-and generate a specifically tuned kernel for every large integer. All kernels use this procedure except the extended multiplication for the CPU.
+It is built on the preprocessor with the boost preprocessor package. We can thus create and manipulate the ASM strings during the precompilation, 
+and generate a specifically tuned kernel for every large integer. We give a small example of a long addition with the generator into the table \ref{ASMGENERATOR}.
+
+ \begin{table}[h] 
+	\begin{center}
+	 	\begin{tabular}{l l}
+                           \hline 
+                           \textbf{Assembly generator}      \hspace{8.5 cm}  ASM  x86-64 \\ \hline
+                           \begin{tabular}{l l l l } 
+                               \tiny{1:} & begin asm(\texttt{"}     &  $\vartriangleright$ Start ASM string & asm(\texttt{"} \\
+                               \tiny{2:} &  \textbf{boost  repeat} over n  &$\vartriangleright$  Start the repetition over n from 0 to n-1 &   \\
+                               \tiny{3:} &   \hspace{0.2 cm}  boost  if(n = 0) &  $\vartriangleright$   if n = 0 &  \hspace{0.2 cm}  \texttt{"}\texttt{addq} 0x0(rsi), rax\texttt{"}\\    
+                               \tiny{4:} &   \hspace{0.4 cm}  boost stringize(\texttt{addq}  $r_i$, $r_j$)  &$\vartriangleright$   Become \texttt{"addq} \dots &  \hspace{0.2 cm}    \texttt{"}\texttt{adcq}  0x8(rsi), rax\texttt{"}\\
+                               \tiny{5:} &   \hspace{0.2 cm}  boost  else  & $\vartriangleright$   if n != 0 &  \hspace{0.2 cm}  \texttt{"}\texttt{adcq}  0x10(rsi), rax\texttt{"} \\ 
+                               \tiny{6:} &   \hspace{0.4 cm}  boost stringize(\texttt{adcq}  $r_i$, $r_j$)  &$\vartriangleright$   Become \texttt{"adcq} \dots &  \hspace{0.2 cm}   \texttt{"}\texttt{adcq}  0x18(rsi), rax\texttt{"}\\
+                               \tiny{7:} &  \textbf{end boost repeat} &  &   \\                            
+                               \tiny{8:} &  \texttt{"})  &  $\vartriangleright$ end ASM string & \texttt{"})\\
+                          \end{tabular} &  \\ \hline
+		 \end{tabular} 
+		 \caption{Basic example of the ASM generator for a long addition (n=4), the preprocessor with the boost preprocessor macros generate the effective string for the ASM block.
+ASM \texttt{move} are not included, as the correct use of the registers.
+It exceeds  the purpose of the present document \label{ASMGENERATOR}.
+}
+	\end{center}
+\end{table}
 
@@
 It allows multiplication-addition of 32 bits integers  with carry bit support \cite{CUDAasm}. 
-Work in 32 bits mode imposes twice as much operations for the additions/substractions, and one order of magnitude more for the multiplications.
+Work in 32 bits mode imposes twice as much operations for the additions/subtractions, and one order of magnitude more for the multiplications.
 
@@
 In our problem the AoS consists in saving the data in this order : VLI - series of VLI (polynomial coefficient)- series of polynomials (the vector).
-The SoA structure interleaves the coefficients of the VLI and the polynomials:
-\begin{figure}
-\begin{minipage}{0.50\linewidth}
-\begin{eqnarray}
-\tiny{
- \underbrace{
- \underbrace{ \underbrace{a^0_{00} | a^1_{00} | a^2_{00}}_{VLI, a_{00} } || \underbrace{a^0_{10} | a^1_{10} | a^2_{10}}_{VLI, a_{10}} ||   \dots  }_{1^{st} \textrm{polynomial}} 
-      |||    \underbrace{ \underbrace{b^0_{00} | a^1_{00} | b^2_{00}}_{VLI, b_{00} } || \underbrace{b^0_{10} | b^1_{10} | b^2_{10}}_{VLI, b_{10}} ||   \dots  }_{2^{nd} \textrm{polynomial} } }_{\textrm{Vector of polynomials, AoS order}} \nonumber
-}
-\end{eqnarray}
-\end{minipage}
-\begin{minipage}{0.50\linewidth}
-\begin{eqnarray}
-\tiny{
- \underbrace{
- \underbrace{a^0_{00} | a^0_{10} | \dots || a^1_{00} | a^1_{10} | \dots  }_{1st  \textrm{ nested VLI/Polynomial} } |||  \underbrace{b^0_{00} | b^0_{10} | \dots || b^1_{00} | b^1_{10} | \dots  }_{2nd  \textrm{ nested VLI/Polynomial} }
- }_{\textrm{Vector of polynomials, SoA Order}} \nonumber }
-\end{eqnarray}
-\end{minipage}
-\caption{Representation of the AoS Order (left) or SoA Order (right) for a vector of polynomial \label{AOSSOA}}
-\end{figure}
+The SoA structure interleaves the coefficients of the VLI and the polynomials.
+
+%\begin{figure}
+%\begin{minipage}{0.50\linewidth}
+%\begin{eqnarray}
+%\tiny{
+% \underbrace{
+% \underbrace{ \underbrace{a^0_{00} | a^1_{00} | a^2_{00}}_{VLI, a_{00} } || \underbrace{a^0_{10} | a^1_{10} | a^2_{10}}_{VLI, a_{10}} ||   \dots  }_{1^{st} \textrm{polynomial}} 
+%      |||    \underbrace{ \underbrace{b^0_{00} | a^1_{00} | b^2_{00}}_{VLI, b_{00} } || \underbrace{b^0_{10} | b^1_{10} | b^2_{10}}_{VLI, b_{10}} ||   \dots  }_{2^{nd} \textrm{polynomial} } }_{\textrm{Vector of polynomials, AoS order}} \nonumber
+%}
+%\end{eqnarray}
+%\end{minipage}
+%\begin{minipage}{0.50\linewidth}
+%\begin{eqnarray}
+%\tiny{
+% \underbrace{
+% \underbrace{a^0_{00} | a^0_{10} | \dots || a^1_{00} | a^1_{10} | \dots  }_{1st  \textrm{ nested VLI/Polynomial} } |||  \underbrace{b^0_{00} | b^0_{10} | \dots || b^1_{00} | b^1_{10} | \dots  }_{2nd  \textrm{ nested VLI/Polynomial} }
+% }_{\textrm{Vector of polynomials, SoA Order}} \nonumber }
+%\end{eqnarray}
+%\end{minipage}
+%\caption{Representation of the AoS Order (left) or SoA Order (right) for a vector of polynomial \label{AOSSOA}}
+%\end{figure}
 
@@
 For our range of operations the extended multiplication varies from 128 to 512 bits, a SIMD version will not be the optimal solution.
-Though, it has been used with success for large integers ($>$1024 bits) \cite{SIMD}.
+Though, it has been used with sucess for  integers higher than 1024 bits \cite{SIMD}.
 
@@
 
-Applying the GPU ideology to the polynomial inner product  means that the polynomial multiplications will be executed in parallel on a SM.
+Applying the GPU ideology to the polynomial inner product  means that the polynomial multiplications will be executed in parallel on the SMs.
 Then a global reduction will sum all resulting polynomials.
@@
 As the number of threads is fixed by the number of coefficients, $2n$ for a dense polynomial,  the maximum number of threads will be 28
- which corresponds to one warp.
-We do not utilize at all the capacity of the GPU.
-For two and more variables, whatever the polynomials, the performance is  like a parabola.
-The position of the maximum depends on  to the type of polynomial.
-We obtained the maximum of performance  if the following conditions are satisfied.
-The number of running threads is maximum; (256 threads), and both input
+ which corresponds to one warp. We do not utilize at all the capacity of the GPU. For two and more variables, whatever the polynomials, we have a maximum. 
+  We obtained the maximum of performance  if the following conditions are satisfied. The number of running threads is maximum; (256 threads), and both input
   polynomials  are cached into the texture memory.

### 2012-12-28 doc/paper_vli/vli.tex
VLI paper, integrate CSCS corrections

@@
 %\include{{llncs2e}}
-\title{Very Large Integers  and polynomial inner product}
-\author{Timoth\'ee Ewart\inst{1}\thanks{Acknowledgments: Maxim Milakov, Peter Messmer  - NVIDIA, and  Williams Sawyer, Gilles Fourestey - CSCS}  , Andreas Hehn$^2$, Matthias Troyer\inst{2} and Thierry Giamarchi\inst{1}}
+\title{Very Large Integers  and polynomial inner products}
+\author{Timoth\'ee Ewart\inst{1}\thanks{Acknowledgments: Maxim Milakov, Peter Messmer:   NVIDIA,   Williams Sawyer, Gilles Fourestey:  CSCS, HP2C funding, hp2c.ch}  , Andreas Hehn$^2$, Matthias Troyer\inst{2} and Thierry Giamarchi\inst{1}}
 
@@
 \begin{abstract}
-The main purpose of the  Very Large Integer (VLI/C++) library is to perform large integer arithmetic and  inner product of polynomials. Based on meta-programming of assembly,
-and  hybrid (CPU/GPU) mode, it performs inner product of dense and triangular polynomials  of 1 to 4 variables  up to order 14. The coefficients are represented  by large integers of 128 to 256 bits. 
-In this paper, we remember the basics of the inner product with polynomials, and  we present the implementation of the library in assembly.
-Then, we focus on the GPU implementation of the inner product.
-All our algorithms are based on brute force attacks, because the problem is too small for a FFT approach to be valuable.
-We save a factor up to 10 on CPU and 60 on GPU compared to a GMP solution.
-We get 67 \% of the peak performance of the GPU.
+The main purpose of the  Very Large Integer (VLI/C++) library is to perform large integer arithmetic and  inner products of polynomials. Based on meta-programming of assembly language,
+and  hybrid (CPU/GPU) mode, it performs inner products of dense and triangular polynomials  of 1 to 4 variables  up to order 14 where the coefficients are represented  by large integers of 128 to 256 bits.
+In this paper, we present the basics of the inner products of polynomials, and  the implementation of the library in assembly language for the CPU and the GPU.
+All our algorithms are based on brute force attacks, because the problem is too small for a FFT approach.
+We save a factor up to 10 on CPU, and 60 on GPU compared to a GMP solution,
+and attain 67 \% of the peak performance of the GPU.
 \end{abstract}
 \section{Introduction}
- A few libraries like  LINBOX or NTL  provide high performance in polynomial solvers.
-These libraries are  connected with GMP\footnote{GMP is the standard library for long arithmetic, gmplib.org} when needed. 
+ A few libraries like  LINBOX or NTL  provide high performance for polynomial solvers.
+These libraries utilize with GMP\footnote{GMP is the standard library for long arithmetic, gmplib.org} when needed. 
 The introduction of hardware accelerators in HPC a few years ago, and more recently of GPU  by NVIDIA, introduce a revolution in the programming models. 
-It is therefore necessary  to adapt the libraries to these new hardwares.  The most famous example on this transition is the famous blas/lapack libraries to MAGMA. 
-Polynomials libraries, however  did not yet  make the transition to GPU.
-During the last few years, the polynomial multiplication was successfully implement to this new hardware
+It is therefore necessary  to adapt the libraries to these new hardwares.  One example on this transition is the well know  blas/lapack libraries to MAGMA. 
+Polynomials libraries, however  have not yet  do the transition to GPU.
+During the last few years, the polynomial multiplication was successfully implemented on this new hardware
 \cite{Govindaraju2008, Emeliyanenko:2009, Moreno2010}. All these works focus on large polynomials,
@@
 
-In the present work, we are focusing on the inner product of dense and triangular multivariable polynomials.  The coefficients of the polynomials are large integers. 
-We preferred to develop a long integer library  rather than utilizing GMP, because GMP is a dynamic library.
-Moreover, we want a library natively compatible with GPU to allow hybrid calculation. GPU imposes contiguous memory which is impossible with a GMP approach.
+In the present work, we focus on the inner products of dense and triangular multivariable polynomials.  The coefficients of the polynomials are large integers. 
+We preferred to develop a long integer library  rather than utilizing GMP, because GMP is a dynamic allocated memory library.
+Moreover, we want a library natively compatible with GPU to allow hybrid calculation. GPUs impose contiguous memory which is impossible with a GMP approach due to the dynamic allocated memory.
 Thus, we propose a static library based on contiguous memory where the exact assembly kernels are generated by meta-programming during the compilation for CPU and GPU.
-We do not waste time in the dynamic memory allocation, and let the compiler optimize deeply the library for both architectures.
+We do not waste time in the dynamic memory allocation, and let the compiler optimize  the library for both architectures.
 
-\section{Polynomial and inner product definitions}
+\section{Polynomial and inner products definitions}
 
-The library performs the inner product of vectors of polynomials from one to four variables, where polynomials are defined like (four variables) :
+The library performs the inner products of vectors of polynomials from one to four variables, where polynomials are defined like (four variables) :
 \begin{eqnarray}
@@
 The polynomials of the form  (1) and (2) are named dense and triangular respectively. Consider now two vectors of polynomials of the  same type with $m$ 
-entries  $\boldsymbol{v}_1<P_a>$ and  $\boldsymbol{v}_2<P_b>$. The inner product  consists in the product 
-of every entries  associated with a global reduction. 
+entries  $\boldsymbol{v}_1<P_a>$ and  $\boldsymbol{v}_2<P_b>$. The inner products  consist of the product 
+of all entries  associated with a global reduction. 
  \begin{eqnarray}
- P_c = \sum_i^m \boldsymbol{v}_{1_i}<P_a>  \times  \boldsymbol{v}_{2_i}<P_b> \label{poly1} \label{InnerProduct}
+ P_c = \sum_{i=0}^m \boldsymbol{v}_{1_i}<P_a>  \times  \boldsymbol{v}_{2_i}<P_b> \label{poly1} \label{InnerProduct}
 \end{eqnarray}
-where $\times$ indicates the multiplication between two polynomials. The multiplications use basic rules of polynomials multiplication. 
-A final coefficient may be the sum of several coefficients (named contributions), resulting from the sum of several multiplications. These coefficients are the cross terms. 
+where $\times$ indicates the multiplication between two polynomials. The multiplications use usual rules of polynomials multiplication. 
+A final coefficient may be the sum of several coefficients (called contributions), resulting from the sum of several multiplications. These coefficients are the cross terms. 
 The cross terms are not a problem for the CPU version, as a polynomial multiplication is done by a single thread, but they are the bottleneck for the GPU version.
@@
 
-In terms of  programming and parallelization, at least for the CPU version, the previous equations can be reduced to independent loops with a final reduction.
+In terms of  programming and parallelization, in the CPU version, the previous equations can be reduced to independent loops with a final reduction.
 The CPU parallelization has been done with OpenMP.
@@
 
-The coefficients of the polynomials are represented by a large integer named VLI number.  A VLI number is characterized by a size in bits; the size is a  template parameter, which
+The coefficients of the polynomials are represented by a large integer named the VLI number.  A VLI number is characterized by a size in bits; the size is a  template parameter, and
 varies from 128 to 512 bits.  Its implementation is static, contrary to GMP. 
-Inside the memory, the VLI numbers are  contiguous. The data type container is a 64 bits unsigned integer array.
+Inside the memory, the VLI numbers are  contiguous. The data type container is a 64 bit unsigned integer array.
 
 The basic operations supported by VLI are the four basic operations: $+,-,\times, /$, extended multiplications and bit shift.
-The library allows  operations between operands of the same type, and with basic  integer.
-For performance reasons, we have a specific solver for every operation except  the division.
-All the algorithms and operations are based on the elementary  methods that we learn
-in our the childhood but the realizations has been done in assembly,
-because the assembly instructions facilitate the implementation and boost the performances.
+The library allows  operations between operands of the same type, and with basic  integers.
+For performance reasons, we have a specific solver for every operation except  for division.
+All the algorithms and operations are based on basic arithmetic  but the implementations have been done in assembly language to  boost  performance.
 
@@
 
-We applied algorithms based on childhood  experience for large arithmetic. These methods necessitate carry\footnote{ The carry bit is a bit of the register of status, as the overflow bit, 
-it informs  the overflow of an operation, but it can be re-used to propagate the carry. } for addition/subtraction and an extended multiplication for the multiplication.
-Although the hardware supports them, we can not access  these features by the C++ language. The only possibility is the assembly language.
+We applied algorithms based on childhood  experience for large arithmetic. The implementation necessitates a "carry bit" and an extended multiplication. The hardware supports
+the carry bit in the status register, as the extended multiplications, but they are only accessible by the assembly language.
 
@@
 We coded the assembly using the GNU inline assembly which consists of strings between ASM tags.
-Generating by hand all these  assembly strings (kernels)  can be painful and a source of error\footnote{The library has near 10000 lines of assembly cumulated between x86-64, Power64, and PTX} 
-We designed a generator of assembly. It creates all versions of a kernel during the compilation.   
-It is built on the preprocessor with the boost preprocessor package. We can thus create and manipulate the ASM strings during the precompilation, 
+Generating by hand all these  assembly strings (kernels)  can be painful and a source of error-prone\footnote{The library has near 10000 lines of assembly in total for x86-64, Power64, and PTX} 
+We designed a assembly generator. It creates all versions of a kernel during the compilation.   
+It is built on Boost preprocessor package. We can thus create and manipulate the ASM strings during the precompilation, 
 and generate a specifically tuned kernel for every large integer. We give a small example of a long addition with the generator into the table \ref{ASMGENERATOR}.
@@
 		 \end{tabular} 
-		 \caption{Basic example of the ASM generator for a long addition (n=4), the preprocessor with the boost preprocessor macros generate the effective string for the ASM block.
-ASM \texttt{move} are not included, as the correct use of the registers.
-It exceeds  the purpose of the present document \label{ASMGENERATOR}.
-}
+		 \caption{Simplified  example of the ASM generator for a long addition (n=4), the Boost preprocessor macros generate the effective string for the ASM block.
+ASM \texttt{move} are not included.
+\label{ASMGENERATOR}  }
 	\end{center}
@@
 
-The x86-64 architecture allows 16 registers (32 for power). As much as possible, in the solver, we conserve the data in the registers in order to reduce memory read/writes to the  minimum.
- Additional optimizations have been done like full unrolling or assembly tips. We  supported  x86-64, Power64 and PTX (GPU) assembly.
- PTX does not support carry bit for 64 bits integer, but only for 32 bits integer\footnote{Introduced since CUDA 4.2}.  
-It allows multiplication-addition of 32 bits integers  with carry bit support \cite{CUDAasm}. 
-Work in 32 bits mode imposes twice as much operations for the additions/subtractions, and one order of magnitude more for the multiplications.
+The x86-64 architecture allows 16 registers (32 for Power architecture).
+As much as possible, in the solver, we conserve the data in the registers in order to reduce memory read/writes to the  minimum.
+Additional optimizations have been done like full unrolling or assembly tips.
+We  supported  x86-64, Power64 and PTX (GPU) assembly.
+ PTX does not support carry bit for 64-bit integer, but only for 32-bit integer\footnote{Introduced since CUDA 4.2}.  
+It allows multiplication-addition of 32-bit integer  with carry bit support \cite{CUDAasm}. 
+Calculation in 32-bit mode imposes twice as much operations for the additions/subtractions, and one order of magnitude more for the multiplications.
 
@@
 The choice of the memory layout is important. It may facilitate  the usage of the SIMD/AVX instructions. We choose the AoS for the following reasons:
-we want to reuse the VLI library in other objects conception code, so we can not interleave the VLI number with the polynomial. 
- Apply SIMD on AoS, for an addition is a non sense because there is no specific instruction to propagate carry between entries of the multimedia register.
-The extended multiplication under AVX \texttt{\_\_mm256\_mul\_epu32}  allows only four multiplications of 32 bits, we save a factor two.
+we want to reuse the VLI library without polynomial in other code, so we can not interleave the VLI number with the polynomial. 
+ Applying SIMD to the AoS, for an addition is a nonsense because there is no specific instruction to propagate carry bit between entries of the multimedia register.
+The extended multiplication under AVX \texttt{\_\_mm256\_mul\_epu32}  allows only four multiplications of 32-bit, saving a factor of two.
 Nevertheless, on Sandy Bridge, we have 3 ALU, we can not exclude
-the execution of several long multiplication in the same time, and we lose the benefice of the SIMD version.
-Moreover, a large extra work is necessary to prepare the registers and shuffle the data into the registers which is expensive.
+the execution of several long multiplication at the same time, and we lose the benefit of the SIMD version.
+Moreover, there is a large overhead  to prepare the registers and shuffle the data into the registers.
 For our range of operations the extended multiplication varies from 128 to 512 bits, a SIMD version will not be the optimal solution.
-Though, it has been used with sucess for  integers higher than 1024 bits \cite{SIMD}.
+However SIMD has been used with success for  integers larger  than 1024-bit \cite{SIMD}.
 
@@
 
-A Graphical Portable Unit  is efficient  to address problems which can be expressed as data-parallel computations Ð the same program 
+A Graphical Processing Unit  is efficient  to address problems which can be expressed as data-parallel computations Ð the same program 
 is executed on many data elements in parallel. Successful usage of GPU in HPC requires a good understanding of their design to reach great performance. A GPU is composed 
-of several independent calculation  entities named streaming multiprocessors (SMs typically a tenth), where the SMs run hundreds of threads concurrently (typically 256) by group of 32 (warp).
-The GPU programming model imposes to divide the problem into coarse sub-problem that can be solved independently in parallel by blocks of threads,
+of several independent calculation  entities named streaming multiprocessors (SMs typically 4 to 12), where the SMs run hundreds of threads concurrently (typically 256) by group of 32 (warp).
+The GPU programming model divides the problem into coarse sub-problem that can be solved independently in parallel by blocks of threads,
  and each sub-problem into finer pieces that can be solved cooperatively in parallel by all threads within the block. 
@@
 
-In terms of memory, all SMs  share  a large  memory (several GigaBytes) with  low performance; approximatively one hundred cycles for read/write operations.
-Inside each SM additional special memories (constant, shared, texture, surface) are available. These memories are characterized by their behavior: they can be read only, shared, cache and more.  
-They have still a common characteristics, a very high performance for read/write  operations in a few cycles only.
+In terms of memory, all SMs  share  a large  memory (several GigaBytes) with  low performance but  high performance compared to CPU main memory; approximatively one hundred cycles for read/write operations.
+Inside each SM additional special memories (constant, shared, texture, surface) are available.
+These memories are characterized by their behavior: they can be read only, shared, cache and more.
+They still have common characteristic: a very high performance for read/write  operations in a few cycles only.
 A good programming practice is to preload data from global memory in shared/texture at once, for subsequent computations.  \label{CUDA_PRACTICE}
 
-Applying the GPU ideology to the polynomial inner product  means that the polynomial multiplications will be executed in parallel on the SMs.
+Applying the GPU ideology to the polynomial inner products  means that the polynomial multiplications will be executed in parallel on the SMs.
 Then a global reduction will sum all resulting polynomials.
 The division into blocks is natural, a block being a single entry of the vector (equation \ref{InnerProduct}).
-All the difficulties consists in designing a parallel algorithm for the polynomial multiplication and the final reduction.
+The difficulty consists of designing a parallel algorithm for the polynomial multiplication and the final reduction.
 
@@
 
-The first idea coming in mind: calculate every single multiplication on an independent GPU thread.
+The first idea comes to mind: calculate every single multiplication on an independent GPU thread.
 Indeed for a dense polynomial, the number of coefficients can be large and compatible with the number of SMs threads.
-Nevertheless,  the result coefficients may have several contributions.
+Nevertheless,  the resulting coefficients may have several contributions.
 If the contributions are calculated by several threads, this requires reduction between threads.
-This solution introduces   synchronizations, and leads to low performance.
-The peak performance will be reached if the warps run the same amount of work, and every single thread in a warp is responsible for all contributions to an output coefficient.
-In order to achieve this goal, we defined a task object.
-A task is a set of 32 result coefficients in a specific order.
+This solution introduces   synchronizations and leads to low performance.
+The peak performance will be attained if the warps have the same amount of work, and every  thread in a warp is responsible for all contributions to an output coefficient.
+In order to achieve this goal, we define a task object.
+A task is a set of 32 resulting coefficients in a specific order.
 A task tells a  warp how to calculate coefficients (the thread number one of the warp takes the first coefficient of the task and so on).
-The warps may not have the same number of tasks, as we equilibrate the workload.
-We proceed in two  steps: first on CPU, where a preliminary work is done, we prepare all information on the tasks, and then, the calculation on GPU.
+The warps may not have the same number of tasks, as we balance the workload.
+We proceed in two  steps: first on CPU, where a preliminary work is done, namely all information on the tasks is determined, and then, the calculation on GPU.
 
-For the first step, we create a fictive resulting polynomial  represented by the indices and the contributions. They are sorted by descending order of the number of contributions
-and grouped by 32 as a task. These tasks are reordered over the warps to equilibrate the amount of work. An illustration of this algorithm is given on  figure \ref{algo_GPU}. 
+For the first step, we create a dummy resulting polynomial  represented by the indices and the contributions. They are sorted by descending order of the number of contributions
+and grouped by 32 as a task. These tasks are reordered over the warps to balance the amount of work. An illustration of this algorithm is given on  figure \ref{algo_GPU}. 
 The more coefficients the polynomials have, the more uniform the distribution of the tasks over the warps will be, and the more efficient the calculation.
-We fix the maximum number of warp to eight (8 warps - 256 threads), as far as possible, we assign at least two tasks per warp.
+We fix the maximum number of warps to eight (8 warps/256 threads), as far as possible, we assign at least two tasks per warp.
 Thus, depending on  the number of coefficients a warp may calculate one task, whereas another warp may have several tasks to calculate.
- When the determination of the task's list is over,  we transfer  asynchronously the tasks list to the global memory.
+ When the determination of the list of the tasks is over,  we transfer  asynchronously it to the global memory.
 This method does not  change the ordering of the data. As for the CPU version, we have an AoS order.
@@
 
-This task's list job is executed for all polynomial types only once. All the informations are kept into the global memory and reuse in case of.  As soon as,  all task's lists are calculated, 
-we transfer the datas of the polynomials asynchronously on the device (the global memory of the GPU is allocated only once and reuse),
- and the calculation of all polynomial multiplication start.
+This tasks determination is executed for all polynomial types only once.
+All the information is kept into the global memory and reuse if we need to calculate several times the same kind of inner products.
+ As soon as,  all task of lists are calculated,  we transfer the data of the polynomials asynchronously onto the device (the global memory of the GPU is allocated only once and reuse),
+ and the calculation of all polynomial multiplications start.
 
@@
 When a polynomial multiplication begins,  every warp gets the corresponding tasks (at least one) from the task's list and loops over its number of tasks. Then,
-the threads into a warp get their respective output coefficients and iterate over the number of  contributions.
-For every iteration over contributions, the threads determine the input coefficients, read the corresponding data in the fast memory ,and perform the arithmetic operations: an extended multiplication and an optional long addition (if the result coefficient has several contributions).
-The calculation of an output coefficient  is  done into fast memory (texture for the input data, and register for the intermediate result).
-When the loop over contributions terminates,  we transfer the coefficients into the an intermediate buffer in the global memory, but in SoA  order to prepare the final reduction. 
+the threads in a warp get their respective output coefficients and iterate over the number of  contributions.
+For every iteration over contributions, the threads determine the input coefficients, read the corresponding data in the fast memory, and perform the arithmetic operations: an extended multiplication and an optional long addition (if the result coefficient has several contributions).
+The calculation of an output coefficient  is  done in fast memory (texture for the input data, and register for the intermediate result).
+When the loop over contributions terminates,  we transfer the coefficients into the an intermediate buffer in the global memory, but in SoA  order for the final reduction. 
 The reduction is done using the standard method of NVIDIA \cite{CUDAReduction} to achieve the highest performance.
@@
 To conclude, this  algorithm has the great advantage of working well with dense and sparse polynomials. 
-The main difficulty consists  in designing an efficient function GetIndices which determines the contributions.
+The main difficulty consists  of designing an efficient function GetIndices which determines the contributions.
  We also developed an optional hybrid mode between CPU/GPU. This  technique is usual in GPU computing.
-  We split the input vectors of polynomials into two chuncks of data, execute concurrently on CPU and GPU. 
-  Only at the end, a synchronization during  the memory transfer is done  by the final reduction.
+  We split the input vectors of polynomials into two chunks of data, execute concurrently on CPU and GPU. 
+  A synchronization  is performed during  the memory transfer   by the final reduction, at the end of the calculation.
 
@@
 
-Benchmarks are performed on the facilities of the Swiss Center for Scientific Computing (CSCS), on a Cray  xk7 cluster for the GPU and a Sandy Bridge  node with 32 logical cores  (HyperThreading on) Xeon E5-2670.
-The size of our vector is equal to 4096, the order of the polynomials varies from 1 to 14, the size of the input coefficients of the polynomials varies from 128 to 256 bits, the output polynomial has twice larger coefficients (256 to 512 bits).
+Benchmarks are performed on the facilities of the Swiss Center for Scientific Computing (CSCS), on a Cray  XK7 cluster for the GPU, and a Sandy Bridge  node with 32 logical cores  (HyperThreading on) Xeon E5-2670.
+The size of our vector is equal to 4096, the order of the polynomials varies from 1 to 14, the size of the input coefficients of the polynomials varies from 128 to 256 bits, the output polynomial has double width-coefficients (256 to 512 bits).
 
-Evaluating  integer benchmarks  is not easy, contrary to float number integer operations are performed by ALU (and not on the single FPU),  the number of  ALU  between processors and the execution time of the operations are  not constants \cite{ASMcost} (constant for FPU).
+Evaluating  integer benchmarks  is not easy: unlike the floating point, integer operations are performed by ALU (and not on the single FPU),  the number of  ALU  between processors (Intel, AMD, IBM) and the execution time of the operations are  not constant \cite{ASMcost} (constant for FPU).
 We compute the  result  numbers as operations \footnote{An operation can be a long multiplication, addition and so on}   per seconds $GOP/s=10^{-9} \times batch/t$, where  $batch$ is the number of long integer operations computed during the execution (additions and multiplications), $t$ [s]  is the elapsed time.
@@
 
-  Figure \ref{ResME} presents the performance of a dense and triangular polynomials inner product.
-It compares our CPU library to GMP (where polynomials are just instantiated with VLI or GMP).
-We get an excellent speed up factor from 4 to 11, for two raisons.
+  Figure \ref{ResME} presents the performance of a dense and triangular polynomials inner products.
+It compares our CPU library to GMP (where polynomials are  instantiated with VLI or GMP).
+We get an excellent speed-up factor from 4 to 11, for two reasons.
 Coefficients of the  polynomial are allocated in the heap, while in the case of GMP, they are allocated in the stack.
 It is well know that better performances are obtained with less dynamic memory allocation.
-The second reason is that  the meta-programming allows the construction of a specific solver for every type of large integer during the compilation, contrary to GMP, we do not have a universal solver (as GMP  for small size), the solver being unique with the maximum optimizations.
-% (operations are performing in the register only, no stack call).
+The second reason is that  the meta-programming allows the construction of a specific solver for every type of large integer during the compilation, thus, we do not have a universal solver, the solver being unique with the maximum optimizations.
+Moreover operations are performed in the registers only, no stack call.
 
 The GPU curves show again an excellent speed up  factor of 10 to 60 maximum, compared to GMP, but with an interesting feature. 
-For both kind of polynomials, the performance for one variable is low, because we have a small number of threads during the execution of the GPU algorithms.
+For both kinds of polynomials, the performance for one variable is low, because we have a small number of threads during the execution of the GPU algorithms.
 As the number of threads is fixed by the number of coefficients, $2n$ for a dense polynomial,  the maximum number of threads will be 28
- which corresponds to one warp. We do not utilize at all the capacity of the GPU. For two and more variables, whatever the polynomials, we have a maximum. 
-  We obtained the maximum of performance  if the following conditions are satisfied. The number of running threads is maximum; (256 threads), and both input
+ which corresponds to one warp. We do not utilize the full capacity of the GPU. For two and more variables, whatever the polynomials, 
+  we attain the maximum performance,  if the following conditions are satisfied: The number of running threads is maximum (256 threads), and both input
   polynomials  are cached into the texture memory.
 On Kepler the size of the texture memory is equal to 48 kB.
-This memory  is highly optimized for repetitive and random access contrary to shared memory
- (where bank memory conflict can appear).
-Moreover, for large order, the result polynomials have  more coefficients, so the balance of the tasks over the warp is good, and consequently the efficiency too.
+This memory  is highly optimized for repetitive and random access in contrast to shared memory
+ where bank memory conflict can appear.
+Moreover, for large order, the resulting polynomials have  more coefficients, so the balance of the tasks over the warp is good, and consequently the efficiency too.
 
@@
 For the small orders the number of GPU threads is lower than 256 but it will increase with the order until the maximum.
-After, for the large orders, although the number of threads is 256, the polynomials will be not full cache into the texture memory.
+After that, for the large orders, although the number of threads is 256, the polynomials will be not full cache into the texture memory.
 So the execution kernel calls data from the global memory which reduces the performance.
- All this behavior shows the differences between a CPU and a GPU.
-A CPU has really more transistor for the  data caching and flow control, it ensures a regular and optimum access to the data.
-For the GPU,
- the programmer manages by hand the flow control. In return, it will have more transistor for the data processing, and consequently the performance.
+ All this behavior shows the differences between a CPU and a GPU. A CPU has  more transistors for the  data caching and flow control, it ensures a regular and optimal access to the data. For the GPU,
+ the programmer manages the flow control by hand. In return, it will have more transistors for the data processing, and consequently more performance.
 
-Although translating the results into MIPS is difficult, we did it, at least on GPU because the GPU kernel of the extended multiplication utilizes only one PTX instruction (\texttt{madc}).
-Counting the total number of  ASM instructions during  the inner product, and dividing by the elapsed time, we obtain an estimation in GIPS (Giga Instructions Per Second).
+Although translating the results into MIPS is difficult, we attempt it, at least on the GPU, because the GPU kernel of the extended multiplication utilizes only one PTX instruction (\texttt{madc}).
+Counting the total number of  ASM instructions during  the inner products, and dividing by the elapsed time, we obtain an estimation in GIPS (Giga Instructions Per Second).
 A  profiling of the code with the CUDA profiler indicates that the bottleneck of the application is the product of polynomials, so we do not take into account   the transfer of the data and the ASM belonging  to the reduction.
@@
 The results are reported in the figure \ref{ResMEMIPS}.
-We obtain 67\% of the Kepler card, which is a good result.
+We obtain 67\% of the peak performance of the Kepler card, which is a good result.
 The results are better for the dense polynomials because the function \texttt{GetIndices}  of the GPU algorithms is more time consuming for the triangular polynomials.
@@
 to GMP for the our specific range.
-We aware the comparisons to GMP were not fair.
-We are working with a hundred of bits without dynamic memory allocation,  while GMP excels in thousand bits arithmetic, where our library can not work.
+The limited performances of GMP for small size were one of the argument to develop the VLI library, although GMP excels in thousand bits arithmetic where our library can not operate.
 As future work, we would like to implement, new kinds of polynomials, and a better management 

### 2013-01-04 doc/paper_vli/vli.tex
vli paper

@@
 The main purpose of the  Very Large Integer (VLI/C++) library is to perform large integer arithmetic and  inner products of polynomials. Based on meta-programming of assembly language,
-and  hybrid (CPU/GPU) mode, it performs inner products of dense and triangular polynomials  of 1 to 4 variables  up to order 14 where the coefficients are represented  by large integers of 128 to 256 bits.
+and  hybrid (CPU/GPU) mode, it performs inner products of dense and triangular polynomials  of 1 to 4 variables  up to order 14 where the coefficients are represented  by large integers of 128 to 512 bits.
 In this paper, we present the basics of the inner products of polynomials, and  the implementation of the library in assembly language for the CPU and the GPU.
@@
 Moreover, we want a library natively compatible with GPU to allow hybrid calculation. GPUs impose contiguous memory which is impossible with a GMP approach due to the dynamic allocated memory.
-Thus, we propose a static library based on contiguous memory where the exact assembly kernels are generated by meta-programming during the compilation for CPU and GPU.
+Thus, we propose a  library based on contiguous memory where the exact assembly kernels are generated by meta-programming during the compilation for CPU and GPU.
 We do not waste time in the dynamic memory allocation, and let the compiler optimize  the library for both architectures.
@@
 
-\subsection{Assembly}
+\subsection{Assembly language}
 
@@
 
-We coded the assembly using the GNU inline assembly which consists of strings between ASM tags.
+We coded the kernels using the GNU inline assembly which consists of strings between ASM tags.
 Generating by hand all these  assembly strings (kernels)  can be painful and a source of error-prone\footnote{The library has near 10000 lines of assembly in total for x86-64, Power64, and PTX} 
@@
                            \begin{tabular}{l l l l } 
-                               \tiny{1:} & begin asm(\texttt{"}     &  $\vartriangleright$ Start ASM string & asm(\texttt{"} \\
+                               \tiny{1:} &  asm(\texttt{"}     &  $\vartriangleright$ Start ASM string & asm(\texttt{"} \\
                                \tiny{2:} &  \textbf{boost  repeat} over n  &$\vartriangleright$  Start the repetition over n from 0 to n-1 &   \\
@@
 As much as possible, in the solver, we conserve the data in the registers in order to reduce memory read/writes to the  minimum.
-Additional optimizations have been done like full unrolling or assembly tips.
-We  supported  x86-64, Power64 and PTX (GPU) assembly.
+Additional optimizations have been done like full unrolling and assembly tips.
+We  supported  x86-64, Power64 and PTX (GPU) assembly language.
  PTX does not support carry bit for 64-bit integer, but only for 32-bit integer\footnote{Introduced since CUDA 4.2}.  
@@
 Both have  advantages and disadvantages.
-In our problem the AoS consists in saving the data in this order : VLI - series of VLI (polynomial coefficient)- series of polynomials (the vector).
+In our problem the AoS consists in saving the data in this order : VLI - series of VLI (polynomial coefficient) - series of polynomials (the vector).
 The SoA structure interleaves the coefficients of the VLI and the polynomials.
@@
 we want to reuse the VLI library without polynomial in other code, so we can not interleave the VLI number with the polynomial. 
- Applying SIMD to the AoS, for an addition is a nonsense because there is no specific instruction to propagate carry bit between entries of the multimedia register.
-The extended multiplication under AVX \texttt{\_\_mm256\_mul\_epu32}  allows only four multiplications of 32-bit, saving a factor of two.
-Nevertheless, on Sandy Bridge, we have 3 ALU, we can not exclude
+Moreover  applying SIMD to the AoS, for an addition is a nonsense because there is no specific instruction to propagate carry bit between entries of the multimedia register.
+The extended multiplication under AVX \texttt{\_\_mm256\_mul\_epu32}  allows only four multiplications of 32-bit, saving a factor of two, nevertheless, on Sandy Bridge, we have 3 ALU, we can not exclude
 the execution of several long multiplication at the same time, and we lose the benefit of the SIMD version.
@@
 the threads in a warp get their respective output coefficients and iterate over the number of  contributions.
-For every iteration over contributions, the threads determine the input coefficients, read the corresponding data in the fast memory, and perform the arithmetic operations: an extended multiplication and an optional long addition (if the result coefficient has several contributions).
+For every iteration over contributions, the threads determine the input coefficients, read the corresponding data in the texture memory, and perform the arithmetic operations: an extended multiplication and an optional long addition (if the result coefficient has several contributions).
 The calculation of an output coefficient  is  done in fast memory (texture for the input data, and register for the intermediate result).
@@
 To conclude, this  algorithm has the great advantage of working well with dense and sparse polynomials. 
-The main difficulty consists  of designing an efficient function GetIndices which determines the contributions.
+The main difficulty consists  of designing an efficient function GetIndices (\ref{ALGO}) which determines the contributions.
  We also developed an optional hybrid mode between CPU/GPU. This  technique is usual in GPU computing.
@@
 It compares our CPU library to GMP (where polynomials are  instantiated with VLI or GMP).
-We get an excellent speed-up factor from 4 to 11, for two reasons.
+We get an excellent speed-up factor from 4 to 11, for three reasons.
 Coefficients of the  polynomial are allocated in the heap, while in the case of GMP, they are allocated in the stack.
@@
 The second reason is that  the meta-programming allows the construction of a specific solver for every type of large integer during the compilation, thus, we do not have a universal solver, the solver being unique with the maximum optimizations.
-Moreover operations are performed in the registers only, no stack call.
+Thirdly, the operations are performed in the registers only, no stack call.
 
@@
 Although translating the results into MIPS is difficult, we attempt it, at least on the GPU, because the GPU kernel of the extended multiplication utilizes only one PTX instruction (\texttt{madc}).
-Counting the total number of  ASM instructions during  the inner products, and dividing by the elapsed time, we obtain an estimation in GIPS (Giga Instructions Per Second).
+Counting the total number of  this ASM instructions during  the inner products, and dividing by the elapsed time, we obtain an estimation in GIPS (Giga Instructions Per Second).
 A  profiling of the code with the CUDA profiler indicates that the bottleneck of the application is the product of polynomials, so we do not take into account   the transfer of the data and the ASM belonging  to the reduction.
@@
 The VLI library provides efficient solvers for the 128 to 512 bits integer associated with polynomials of 1 to 4 variables. We obtain a comfortable speed up as high as  60 compared 
-to GMP for the our specific range.
+to GMP for a specific range.
 The limited performances of GMP for small size were one of the argument to develop the VLI library, although GMP excels in thousand bits arithmetic where our library can not operate.

### 2013-01-08 doc/paper_vli/vli.tex
some changes and TODO comments

@@
 and  hybrid (CPU/GPU) mode, it performs inner products of dense and triangular polynomials  of 1 to 4 variables  up to order 14 where the coefficients are represented  by large integers of 128 to 512 bits.
-In this paper, we present the basics of the inner products of polynomials, and  the implementation of the library in assembly language for the CPU and the GPU.
+In this paper, we present the basics of the inner products of polynomials, and  the implementation of the library for the CPU and the GPU.
 All our algorithms are based on brute force attacks, because the problem is too small for a FFT approach.
@@
 \section{Introduction}
+% TODO references LINBOX, NTL
  A few libraries like  LINBOX or NTL  provide high performance for polynomial solvers.
@@
 The introduction of hardware accelerators in HPC a few years ago, and more recently of GPU  by NVIDIA, introduce a revolution in the programming models. 
-It is therefore necessary  to adapt the libraries to these new hardwares.  One example on this transition is the well know  blas/lapack libraries to MAGMA. 
-Polynomials libraries, however  have not yet  do the transition to GPU.
+It is therefore necessary  to adapt the libraries to these new hardwares.
+A prominent example of such an adaption are the MAGMA libraries, which port the well known BLAS/LAPACK packages to GPU accelerators.
+Polynomials libraries, however,  have not done the transition to GPU yet.
 During the last few years, the polynomial multiplication was successfully implemented on this new hardware
@@
 In the present work, we focus on the inner products of dense and triangular multivariable polynomials.  The coefficients of the polynomials are large integers. 
-We preferred to develop a long integer library  rather than utilizing GMP, because GMP is a dynamic allocated memory library.
-Moreover, we want a library natively compatible with GPU to allow hybrid calculation. GPUs impose contiguous memory which is impossible with a GMP approach due to the dynamic allocated memory.
-Thus, we propose a  library based on contiguous memory where the exact assembly kernels are generated by meta-programming during the compilation for CPU and GPU.
-We do not waste time in the dynamic memory allocation, and let the compiler optimize  the library for both architectures.
+In contrast to GMP which dynamically adapts the size of the integers, the presented long integer library is based on static integer sizes.
+Moreover, our library is natively compatible with GPU to allow hybrid calculation. GPUs impose contiguous memory which is impossible with a GMP approach due to the dynamic allocated memory.
+Thus, we propose a  library based on contiguous memory where the exact assembly kernels are generated during the compilation for CPU and GPU.
+The static integer sizes allow the compiler to aggressivly optimize the non-assembly parts of the library and avoids costly heap allocations for the data.
 
@@
 \end{eqnarray}
-where  $x,\,y,\,z$ and $w$ are the variables,  $i,\,j,\,k$ and $l$  the indices, $a$  the coefficients, and $n$ the order of the polynomial.
+where  $x,\,y,\,z$ and $w$ are symbolic variables,  $i,\,j,\,k$ and $l$  the indices, $a$  the coefficients, and $n$ the order of the polynomial.
 The polynomials of the form  (1) and (2) are named dense and triangular respectively. Consider now two vectors of polynomials of the  same type with $m$ 
@@
 where $\times$ indicates the multiplication between two polynomials. The multiplications use usual rules of polynomials multiplication. 
-A final coefficient may be the sum of several coefficients (called contributions), resulting from the sum of several multiplications. These coefficients are the cross terms. 
+A final coefficient may be the sum of several coefficients (which we call contributions in the following passage), resulting from the sum of several multiplications.
+These coefficients are the cross terms.
 The cross terms are not a problem for the CPU version, as a polynomial multiplication is done by a single thread, but they are the bottleneck for the GPU version.
@@
 
-The coefficients of the polynomials are represented by a large integer named the VLI number.  A VLI number is characterized by a size in bits; the size is a  template parameter, and
-varies from 128 to 512 bits.  Its implementation is static, contrary to GMP. 
+The coefficients of the polynomials are represented by a large integer named the VLI number.  A VLI number is characterized by a size in bits.
+Our current implementation allows for integer sizes from 128 to 512 bits, but is easily extendable to larger integers.
 Inside the memory, the VLI numbers are  contiguous. The data type container is a 64 bit unsigned integer array.
@@
 For performance reasons, we have a specific solver for every operation except  for division.
-All the algorithms and operations are based on basic arithmetic  but the implementations have been done in assembly language to  boost  performance.
+All the algorithms and operations are based on basic arithmetic but have been implemented in assembly language to make use of hardware features which are not accessible by plain C++.
 
@@
 
-We applied algorithms based on childhood  experience for large arithmetic. The implementation necessitates a "carry bit" and an extended multiplication. The hardware supports
+We applied algorithms based on childhood  experience for large arithmetic. The implementation necessitates a carry bit and an extended multiplication. The hardware supports
 the carry bit in the status register, as the extended multiplications, but they are only accessible by the assembly language.
 
-In assembly, these operations  are supported by the pair of mnemonics; \texttt{addq/adcq} for the addition; \texttt{sbbq/subq} for the subtraction, and \texttt{mul} for the extended multiplication.   
+In assembly, these operations  are supported by the pair of mnemonics; \texttt{addq/adcq} for the addition; \texttt{subq/sbbq} for the subtraction, and \texttt{mul} for the extended multiplication.   
 Combining these mnemonics, we reproduce easily long arithmetic algorithm, as described into the famous  ASM book of \cite{Hyde:2003:AAL:861534}.
 
-We coded the kernels using the GNU inline assembly which consists of strings between ASM tags.
-Generating by hand all these  assembly strings (kernels)  can be painful and a source of error-prone\footnote{The library has near 10000 lines of assembly in total for x86-64, Power64, and PTX} 
+We coded the kernels using the GNU inline assembly.
+Generating by hand all these  assembly strings (kernels)  can be painful and error-prone\footnote{The library has near 10000 lines of assembly in total for x86-64, Power64, and PTX}.
 We designed a assembly generator. It creates all versions of a kernel during the compilation.   
@@
 The x86-64 architecture allows 16 registers (32 for Power architecture).
-As much as possible, in the solver, we conserve the data in the registers in order to reduce memory read/writes to the  minimum.
+We try to reuse the data in the registers as much as possible in order to reduce memory read/writes to the minimum.
 Additional optimizations have been done like full unrolling and assembly tips.
 We  supported  x86-64, Power64 and PTX (GPU) assembly language.
- PTX does not support carry bit for 64-bit integer, but only for 32-bit integer\footnote{Introduced since CUDA 4.2}.  
-It allows multiplication-addition of 32-bit integer  with carry bit support \cite{CUDAasm}. 
-Calculation in 32-bit mode imposes twice as much operations for the additions/subtractions, and one order of magnitude more for the multiplications.
+ PTX does not support carry bit for 64-bit integers, but only for 32-bit integers\footnote{Introduced since CUDA 4.2}.  
+It allows a fused multiplication-addition of 32-bit integer  with carry bit support \cite{CUDAasm}. 
+The calculation in 32-bit mode imposes twice as much operations for the additions/subtractions, and one order of magnitude more for the multiplications.
 
@@
 
+% TODO: which problem are you talking about?
 There exists at least two data layouts for this problem: the array of structure (AoS) and structure of array (SoA).
 Both have  advantages and disadvantages.
-In our problem the AoS consists in saving the data in this order : VLI - series of VLI (polynomial coefficient) - series of polynomials (the vector).
+In our problem the AoS consists in saving the data in this hierarchy: VLI - series of VLI (polynomial coefficient) - series of polynomials (the vector).
 The SoA structure interleaves the coefficients of the VLI and the polynomials.
@@
 
-The choice of the memory layout is important. It may facilitate  the usage of the SIMD/AVX instructions. We choose the AoS for the following reasons:
-we want to reuse the VLI library without polynomial in other code, so we can not interleave the VLI number with the polynomial. 
-Moreover  applying SIMD to the AoS, for an addition is a nonsense because there is no specific instruction to propagate carry bit between entries of the multimedia register.
+The choice of the memory layout is important.
+It may facilitate  the usage of the SIMD instructions such as the AVX instructions of the Sandy Bridge x86-64 architecture.
+We choose the AoS for the following reasons:
+Applying SIMD to the AoS, for an addition is a nonsense because there is no specific instruction to propagate carry bit between entries of the SIMD register.
 The extended multiplication under AVX \texttt{\_\_mm256\_mul\_epu32}  allows only four multiplications of 32-bit, saving a factor of two, nevertheless, on Sandy Bridge, we have 3 ALU, we can not exclude
@@
 However SIMD has been used with success for  integers larger  than 1024-bit \cite{SIMD}.
+% TODO somehow the problem with SIMD operations is not entirely clear yet. You mention the issues, but I think you should elaborate them a little more.
 
@@
 
-A Graphical Processing Unit  is efficient  to address problems which can be expressed as data-parallel computations Ð the same program 
+A Graphical Processing Unit  is efficient  to address problems which can be expressed as data-parallel computations, where the same stream of instructions
 is executed on many data elements in parallel. Successful usage of GPU in HPC requires a good understanding of their design to reach great performance. A GPU is composed 
-of several independent calculation  entities named streaming multiprocessors (SMs typically 4 to 12), where the SMs run hundreds of threads concurrently (typically 256) by group of 32 (warp).
+of several independent calculation  entities named streaming multiprocessors (SMs typically 4 to 12), where the SMs run hundreds of threads concurrently (typically 256) in groups of 32 threads (called ``warps'').
 The GPU programming model divides the problem into coarse sub-problem that can be solved independently in parallel by blocks of threads,
@@
 
-Evaluating  integer benchmarks  is not easy: unlike the floating point, integer operations are performed by ALU (and not on the single FPU),  the number of  ALU  between processors (Intel, AMD, IBM) and the execution time of the operations are  not constant \cite{ASMcost} (constant for FPU).
+Evaluating  integer benchmarks  is not easy: unlike the floating point, integer operations are performed by ALU (and not on the single FPU),  the number of  ALUs  between processors (Intel, AMD, IBM) and the execution time of the operations are  not constant \cite{ASMcost} (constant for FPU).
 We compute the  result  numbers as operations \footnote{An operation can be a long multiplication, addition and so on}   per seconds $GOP/s=10^{-9} \times batch/t$, where  $batch$ is the number of long integer operations computed during the execution (additions and multiplications), $t$ [s]  is the elapsed time.
@@
 We get an excellent speed-up factor from 4 to 11, for three reasons.
-Coefficients of the  polynomial are allocated in the heap, while in the case of GMP, they are allocated in the stack.
-It is well know that better performances are obtained with less dynamic memory allocation.
+Coefficients of the  polynomial are allocated in one large block on the heap, while in the case of GMP, each coefficient needs to be allocated independently.
+It is well known that better performances are obtained with less dynamic memory allocations.
 The second reason is that  the meta-programming allows the construction of a specific solver for every type of large integer during the compilation, thus, we do not have a universal solver, the solver being unique with the maximum optimizations.
 Thirdly, the operations are performed in the registers only, no stack call.
+%TODO what do you mean by the last sentence?
 
@@
 This memory  is highly optimized for repetitive and random access in contrast to shared memory
- where bank memory conflict can appear.
-Moreover, for large order, the resulting polynomials have  more coefficients, so the balance of the tasks over the warp is good, and consequently the efficiency too.
+ where bank memory conflicts may appear.
+Moreover, for high orders the resulting polynomials have more coefficients which eases the the work balance between different threads and consequently improves efficiency.
 
@@
 For the small orders the number of GPU threads is lower than 256 but it will increase with the order until the maximum.
-After that, for the large orders, although the number of threads is 256, the polynomials will be not full cache into the texture memory.
-So the execution kernel calls data from the global memory which reduces the performance.
+After that, for the large orders the polynomials will be too large to be cached by texture memory properly. So the execution kernel calls data from the global memory which reduces the performance.
  All this behavior shows the differences between a CPU and a GPU. A CPU has  more transistors for the  data caching and flow control, it ensures a regular and optimal access to the data. For the GPU,
@@
 to GMP for a specific range.
-The limited performances of GMP for small size were one of the argument to develop the VLI library, although GMP excels in thousand bits arithmetic where our library can not operate.
+The limited performances of GMP for small size were the main argument to develop the VLI library, although GMP excels in thousand bits arithmetic where our library can not operate.
 As future work, we would like to implement, new kinds of polynomials, and a better management 

### 2013-01-09 doc/paper_vli/vli.tex
paper vli, modifs

@@
 
-\subsection{Memory layout and SIMD implementation}
+\subsection{Memory layout and SIMD possibility}
 
@@
                                \tiny{2:} &  \textbf{for} over number of task list size &$\vartriangleright$  Depend on the warp \\
-                               \tiny{3:} &  tmp $\,\, \wedge=$ tmp     & $\vartriangleright$ Flush to 0 \\    
-                               \tiny{4:} &   VLI ResCoeff &$\vartriangleright$  Thread gets a result coefficient from a task \\
-                               \tiny{5:} &  \hspace{0.2 cm}  \textbf{for} over the number of contributions  & $\vartriangleright$ Each thread loops over its coefficients \\
-                               \tiny{6:} &  \hspace{0.4 cm}   Indices$_i$ = GetIndices(ResCoeff) &$\vartriangleright$ Determine input indices \\
-                               \tiny{7:} &  \hspace{0.4 cm}   tmp + = CoeffPoly$_1$(indices$_1$) $\times$ CoeffPoly$_2$(indices$_2$) &$\vartriangleright$ Perform the extended MulAdd  \\
-                               \tiny{8:} &  \hspace{0.2 cm}   \textbf{end for} &\\
-                               \tiny{9:} &  SaveSoA(tmp) & $\vartriangleright$ Write into the global memory under AoS \\                               
-                               \tiny{10:} &  \textbf{end for} &\\                               
+                               \tiny{3:} &   VLI ResCoeff &$\vartriangleright$  Thread gets a result coefficient from a task \\
+                               \tiny{4:} &  \hspace{0.2 cm}  \textbf{for} over the number of contributions  & $\vartriangleright$ Each thread loops over its coefficients \\
+                               \tiny{5:} &  \hspace{0.4 cm}   Indices$_i$ = GetIndices(ResCoeff) &$\vartriangleright$ Determine input indices \\
+                               \tiny{6:} &  \hspace{0.4 cm}   tmp + = CoeffPoly$_1$(indices$_1$) $\times$ CoeffPoly$_2$(indices$_2$) &$\vartriangleright$ Perform the extended MulAdd  \\
+                               \tiny{7:} &  \hspace{0.2 cm}   \textbf{end for} &\\
+                               \tiny{8:} &  SaveSoA(tmp) & $\vartriangleright$ Write into the global memory under AoS \\                               
+                               \tiny{9:} &  \textbf{end for} &\\                               
                           \end{tabular} &  \\ \hline
@@
 
-To conclude, this  algorithm has the great advantage of working well with dense and sparse polynomials. 
+To conclude, this  algorithm has the great advantage of working well with dense and triangular polynomials. 
 The main difficulty consists  of designing an efficient function GetIndices (\ref{ALGO}) which determines the contributions.
@@
 The second reason is that  the meta-programming allows the construction of a specific solver for every type of large integer during the compilation, thus, we do not have a universal solver, the solver being unique with the maximum optimizations.
-Thirdly, the operations are performed in the registers only, no stack call.
-%TODO what do you mean by the last sentence?
+Thirdly, the datas stay in the registers during an arithmetic operation.
 

### 2013-01-13 doc/paper_vli/vli.tex
refactoring VLI paper

@@
 \begin{abstract}
-The main purpose of the  Very Large Integer (VLI/C++) library is to perform large integer arithmetic and  inner products of polynomials. Based on meta-programming of assembly language,
-and  hybrid (CPU/GPU) mode, it performs inner products of dense and triangular polynomials  of 1 to 4 variables  up to order 14 where the coefficients are represented  by large integers of 128 to 512 bits.
+The main purpose of the  Very Large Integer (VLI/C++/ASM) library is to perform large integer arithmetic and  inner products of polynomials. Based on meta-programming of assembly language,
+and  hybrid (CPU/GPU) mode, it performs inner products of dense and triangular polynomials  of 1 to 4 variables  where the coefficients are represented  by large integers of 128 to 512 bits. 
 In this paper, we present the basics of the inner products of polynomials, and  the implementation of the library for the CPU and the GPU.
@@
 % TODO references LINBOX, NTL
- A few libraries like  LINBOX or NTL  provide high performance for polynomial solvers.
-These libraries utilize with GMP\footnote{GMP is the standard library for long arithmetic, gmplib.org} when needed. 
-The introduction of hardware accelerators in HPC a few years ago, and more recently of GPU  by NVIDIA, introduce a revolution in the programming models. 
-It is therefore necessary  to adapt the libraries to these new hardwares.
+
+The Very Large Integer library performs  standards  operations for large integer arithmetic and polynomial. This library also provides optimized implementations of inner products  for GPU.
+This library is composed of two distinct modules,  but enough closed  to work together by the mechanism of template instantiation. VLI is a fix size library from 128 to 512 bits,
+as the polynomial module  which is limited for dense and triangular polynomials  up to 4 variables.
+The core of VLI is build under ASM language to provide the best performance.
+Every arithmetic operations
+are executed by a specific kernel generated specifically during the compilation by an ASM generator. Moreover, VLI allows multithreading and hybrid  GPU usage for inner products of polynomials.
+Multithreading is build on OpenMP library and GPU module under CUDA. As the CPU, the GPU is a fixe size library, and ASM kernels are also determinated specifically during the compilation.
+
+In large arithmetic, the standard library is GMP\footnote{GMP is the standard library for long arithmetic, gmplib.org}.
+GMP is polyvalent, nevertheless this polyvalence is attained by the usage of  dynamic memory.
+It is well know dynamic allocation is time consuming, and it can become a bottle neck for small operations which are in the working range.  
+Alternatively for polynomial arithmetic, a few libraries like  LINBOX or NTL  provide a solution but  they use GMP when needed. 
+Although hardware accelerators were introduced  in HPC a few years ago, as the GPUs  by NVIDIA. Only a few libraries have been adapted 
 A prominent example of such an adaption are the MAGMA libraries, which port the well known BLAS/LAPACK packages to GPU accelerators.
-Polynomials libraries, however,  have not done the transition to GPU yet.
+The polynomials libraries, however,  have not done the transition to these new hardwares yet.
 During the last few years, the polynomial multiplication was successfully implemented on this new hardware
-\cite{Govindaraju2008, Emeliyanenko:2009, Moreno2010}. All these works focus on large polynomials,
- using a FFT approach.
+\cite{Govindaraju2008, Emeliyanenko:2009, Moreno2010}. All these works focus on large polynomials, using a FFT approach.
 
-In the present work, we focus on the inner products of dense and triangular multivariable polynomials.  The coefficients of the polynomials are large integers. 
-In contrast to GMP which dynamically adapts the size of the integers, the presented long integer library is based on static integer sizes.
-Moreover, our library is natively compatible with GPU to allow hybrid calculation. GPUs impose contiguous memory which is impossible with a GMP approach due to the dynamic allocated memory.
-Thus, we propose a  library based on contiguous memory where the exact assembly kernels are generated during the compilation for CPU and GPU.
-The static integer sizes allow the compiler to aggressivly optimize the non-assembly parts of the library and avoids costly heap allocations for the data.
+In this work, we are going to highlight the speedup given by a tiny ASM library without any dynamic memory, suggest an alternative for small size integer. 
+And propose an efficient implementation of inner product on GPU with the latest CUDA PTX features.
+
+%In the present work, we focus on the inner products of dense and triangular multivariable polynomials.  The coefficients of the polynomials are large integers. 
+%In contrast to GMP which dynamically adapts the size of the integers, the presented long integer library is based on static integer sizes.
+%Moreover, our library is natively compatible with GPU to allow hybrid calculation. GPUs impose contiguous memory which is impossible with a GMP approach due to the dynamic allocated memory.
+%Thus, we propose a  library based on contiguous memory where the exact assembly kernels are generated during the compilation for CPU and GPU.
+%The static integer sizes allow the compiler to aggressivly optimize the non-assembly parts of the library and avoids costly heap allocations for the data.
 
@@
 
-The library performs the inner products of vectors of polynomials from one to four variables, where polynomials are defined like (four variables) :
+The library performs standards operations for polynomials (up to four variables) and optimized solvers for  the inner products of vectors of polynomials (presented in the next section).
+The polynomials are defined like  e.g. our variables
 \begin{eqnarray}
@@
 The polynomials of the form  (1) and (2) are named dense and triangular respectively. Consider now two vectors of polynomials of the  same type with $m$ 
-entries  $\boldsymbol{v}_1<P_a>$ and  $\boldsymbol{v}_2<P_b>$. The inner products  consist of the product 
+entries  $\boldsymbol{v}_1$ and  $\boldsymbol{v}_2$. The inner products  consist of the product 
 of all entries  associated with a global reduction. 
  \begin{eqnarray}
- P_c = \sum_{i=0}^m \boldsymbol{v}_{1_i}<P_a>  \times  \boldsymbol{v}_{2_i}<P_b> \label{poly1} \label{InnerProduct}
+ P_c = \sum_{i=0}^m \boldsymbol{v}_{1_i} \times  \boldsymbol{v}_{2_i}  \label{poly1} \label{InnerProduct}
 \end{eqnarray}
@@
 the carry bit in the status register, as the extended multiplications, but they are only accessible by the assembly language.
-
 In assembly, these operations  are supported by the pair of mnemonics; \texttt{addq/adcq} for the addition; \texttt{subq/sbbq} for the subtraction, and \texttt{mul} for the extended multiplication.   
@@
 We designed a assembly generator. It creates all versions of a kernel during the compilation.   
-It is built on Boost preprocessor package. We can thus create and manipulate the ASM strings during the precompilation, 
+It is built on Boost preprocessor package. We can thus create and manipulate the ASM strings during the compilation, 
 and generate a specifically tuned kernel for every large integer. We give a small example of a long addition with the generator into the table \ref{ASMGENERATOR}.
@@
 
-% TODO: which problem are you talking about?
-There exists at least two data layouts for this problem: the array of structure (AoS) and structure of array (SoA).
-Both have  advantages and disadvantages.
-In our problem the AoS consists in saving the data in this hierarchy: VLI - series of VLI (polynomial coefficient) - series of polynomials (the vector).
-The SoA structure interleaves the coefficients of the VLI and the polynomials.
+We have privileged the  array  of structure  (SoA ) detrimental to the  structure of array (SoA)
+The AoS consists in saving the data in this hierarchy: VLI - series of VLI (polynomial coefficient) - series of polynomials (the vector), where  the SoA structure interleaves the coefficients of the VLI and the polynomials.
+A representation of both memory layout is presented on figure \ref{AOSSOA}.
 
-%\begin{figure}
-%\begin{minipage}{0.50\linewidth}
-%\begin{eqnarray}
-%\tiny{
-% \underbrace{
-% \underbrace{ \underbrace{a^0_{00} | a^1_{00} | a^2_{00}}_{VLI, a_{00} } || \underbrace{a^0_{10} | a^1_{10} | a^2_{10}}_{VLI, a_{10}} ||   \dots  }_{1^{st} \textrm{polynomial}} 
-%      |||    \underbrace{ \underbrace{b^0_{00} | a^1_{00} | b^2_{00}}_{VLI, b_{00} } || \underbrace{b^0_{10} | b^1_{10} | b^2_{10}}_{VLI, b_{10}} ||   \dots  }_{2^{nd} \textrm{polynomial} } }_{\textrm{Vector of polynomials, AoS order}} \nonumber
-%}
-%\end{eqnarray}
-%\end{minipage}
-%\begin{minipage}{0.50\linewidth}
-%\begin{eqnarray}
-%\tiny{
-% \underbrace{
-% \underbrace{a^0_{00} | a^0_{10} | \dots || a^1_{00} | a^1_{10} | \dots  }_{1st  \textrm{ nested VLI/Polynomial} } |||  \underbrace{b^0_{00} | b^0_{10} | \dots || b^1_{00} | b^1_{10} | \dots  }_{2nd  \textrm{ nested VLI/Polynomial} }
-% }_{\textrm{Vector of polynomials, SoA Order}} \nonumber }
-%\end{eqnarray}
-%\end{minipage}
-%\caption{Representation of the AoS Order (left) or SoA Order (right) for a vector of polynomial \label{AOSSOA}}
-%\end{figure}
+\begin{figure}
+    \begin{minipage}{0.50\linewidth}
+        \begin{eqnarray}
+        \tiny{
+                 \underbrace{
+                 \underbrace{ \underbrace{a^0_{00} | a^1_{00} | a^2_{00}}_{VLI, a_{00} } || \underbrace{a^0_{10} | a^1_{10} | a^2_{10}}_{VLI, a_{10}} ||   \dots  }_{1^{st} \textrm{polynomial}} 
+                  |||    \underbrace{ \underbrace{b^0_{00} | a^1_{00} | b^2_{00}}_{VLI, b_{00} } || \underbrace{b^0_{10} | b^1_{10} | b^2_{10}}_{VLI, b_{10}} ||   \dots  }_{2^{nd} \textrm{polynomial} } }_{\textrm{Vector of polynomials, AoS order}} \nonumber
+            }
+        \end{eqnarray}
+    \end{minipage}
+\begin{minipage}{0.50\linewidth}
+    \begin{eqnarray}
+    \tiny{
+        \underbrace{
+        \underbrace{a^0_{00} | a^0_{10} | \dots || a^1_{00} | a^1_{10} | \dots  }_{1st  \textrm{ nested VLI/Polynomial} } |||  \underbrace{b^0_{00} | b^0_{10} | \dots || b^1_{00} | b^1_{10} | \dots  }_{2nd  \textrm{ nested VLI/Polynomial} }
+        }_{\textrm{Vector of polynomials, SoA Order}} \nonumber }
+        \end{eqnarray}
+\end{minipage}
+\caption{Representation of the AoS Order (left) or SoA Order (right) for a vector of polynomial \label{AOSSOA}.  The SoA structure limits the usage of the VLI module alone.}
+\end{figure}
 
-The choice of the memory layout is important.
+We have choose the SoA for the two following reasons :
+
+\begin{enumerate}
+\item SoA order allows the natural propagation of the carry bit between the entries of the VLI with the pair of mnemonics \texttt{addq/adcq}, as the utilization of the extended multiplication without any padding.
+\item The VLI and the polynomial module of the library can be separated and use independently. Thinks impossible if the coefficients are interleaved.
+\end{enumerate}
+
+The disagreement  of this choice, the SoA order does not facilitate the usage of the SIMD instruction such as the AVX instructions of the Sandy Bridge x86-64 architecture.
+Applying SIMD to the AoS, for an addition or a subtraction is a nonsense because there is no specific instruction to propagate carry bit between entries of the SIMD register.
+It could be  apply  on multiplications for integer  larger than 1024-bit \cite{SIMD}, unfortunately, it does not correspond to our working range. Consequently, we do not 
+use the multimedia instruction into the ASM kernels.
+
+%The choice of the memory layout is important.
 It may facilitate  the usage of the SIMD instructions such as the AVX instructions of the Sandy Bridge x86-64 architecture.
 We choose the AoS for the following reasons:
-Applying SIMD to the AoS, for an addition is a nonsense because there is no specific instruction to propagate carry bit between entries of the SIMD register.
-The extended multiplication under AVX \texttt{\_\_mm256\_mul\_epu32}  allows only four multiplications of 32-bit, saving a factor of two, nevertheless, on Sandy Bridge, we have 3 ALU, we can not exclude
-the execution of several long multiplication at the same time, and we lose the benefit of the SIMD version.
+%Applying SIMD to the AoS, for an addition is a nonsense because there is no specific instruction to propagate carry bit between entries of the SIMD register.
+%The extended multiplication under AVX \texttt{\_\_mm256\_mul\_epu32}  allows only four multiplications of 32-bit, saving a factor of two, nevertheless, on Sandy Bridge, we have 3 ALU, we can not exclude
+%the execution of several long multiplication at the same time, and we lose the benefit of the SIMD version.
 Moreover, there is a large overhead  to prepare the registers and shuffle the data into the registers.
@@
 
-\section{A short introduction to GPU}
+%\section{A short introduction to GPU}
 
-A Graphical Processing Unit  is efficient  to address problems which can be expressed as data-parallel computations, where the same stream of instructions
-is executed on many data elements in parallel. Successful usage of GPU in HPC requires a good understanding of their design to reach great performance. A GPU is composed 
-of several independent calculation  entities named streaming multiprocessors (SMs typically 4 to 12), where the SMs run hundreds of threads concurrently (typically 256) in groups of 32 threads (called ``warps'').
-The GPU programming model divides the problem into coarse sub-problem that can be solved independently in parallel by blocks of threads,
- and each sub-problem into finer pieces that can be solved cooperatively in parallel by all threads within the block.
-% During the execution, independent blocks will be executed on free SMs, where the program of the block is executed over the threads concurrently way.
+%A Graphical Processing Unit  is efficient  to address problems which can be expressed as data-parallel computations, where the same stream of instructions
+%is executed on many data elements in parallel. Successful usage of GPU in HPC requires a good understanding of their design to reach great performance. A GPU is composed 
+%of several independent calculation  entities named streaming multiprocessors (SMs typically 4 to 12), where the SMs run hundreds of threads concurrently (typically 256) in groups of 32 threads (called ``warps'').
+%The GPU programming model divides the problem into coarse sub-problem that can be solved independently in parallel by blocks of threads,
+% and each sub-problem into finer pieces that can be solved cooperatively in parallel by all threads within the block.
 
-In terms of memory, all SMs  share  a large  memory (several GigaBytes) with  low performance but  high performance compared to CPU main memory; approximatively one hundred cycles for read/write operations.
+%In terms of memory, all SMs  share  a large  memory (several GigaBytes) with  low performance but  high performance compared to CPU main memory; approximatively one hundred cycles for read/write operations.
 Inside each SM additional special memories (constant, shared, texture, surface) are available.
@@
 They still have common characteristic: a very high performance for read/write  operations in a few cycles only.
-A good programming practice is to preload data from global memory in shared/texture at once, for subsequent computations.  \label{CUDA_PRACTICE}
+%A good programming practice is to preload data from global memory in shared/texture at once, for subsequent computations.  \label{CUDA_PRACTICE}
 
-Applying the GPU ideology to the polynomial inner products  means that the polynomial multiplications will be executed in parallel on the SMs.
+%Applying the GPU ideology to the polynomial inner products  means that the polynomial multiplications will be executed in parallel on the SMs.
 Then a global reduction will sum all resulting polynomials.
@@
 
-\subsection{Inner product booster on GPU}
+\section{Inner product booster on GPU}
+
+%The first idea comes to mind: calculate every single multiplication on an independent GPU thread.
+Indeed for a dense polynomial, the number of coefficients can be large and compatible with the number of SMs threads.
+Nevertheless, the resulting coefficients may have several contributions.
+If the contributions are calculated by several threads, this requires reduction between threads.
+This solution introduces   synchronizations and leads to low performance.
+
+The GPU algorithm of the inner product is divided into two parts.
+A preliminary step on the CPU,  where we prepare the calculation blue print of the  contribution.
+And the calculation itself on the GPU
+
+\begin{enumerate}
+
+\item  First, we create a dummy resulting polynomial  represented by the indices and the contributions. They are sorted by descending order of the number of contributions
+and grouped by 32 as a task.
+A task is a set of 32 resulting coefficients in a specific order.
+It tells a  warp how to calculate coefficients (the thread number one of the warp takes the first coefficient of the task and so on).
+These tasks are reordered over the warps to balance the amount of work. An illustration of this algorithm is given on  figure \ref{algo_GPU}. 
+The more coefficients the polynomials have, the more uniform the distribution of the tasks over the warps will be, and the more efficient the calculation.
+We fix the maximum number of warps to eight (8 warps/256 threads), as far as possible, we assign at least two tasks per warp.
+Thus, depending on  the number of coefficients a warp may calculate one task, whereas another warp may have several tasks to calculate.
+When the determination of the list of the tasks is over,  we transfer  asynchronously  to the global memory the data of the polynomial and all the tasks.
+This method does not  change the ordering of the data, ss for the CPU version, we have an AoS order.
+But we have now informations how calculate the polynomial into a efficient way.
+In terms of memory of the GPU, we allocate allocated only once, and reuse inner products after inner products., thus we save time.
 
@@
 
-The first idea comes to mind: calculate every single multiplication on an independent GPU thread.
-Indeed for a dense polynomial, the number of coefficients can be large and compatible with the number of SMs threads.
-Nevertheless, the resulting coefficients may have several contributions.
-If the contributions are calculated by several threads, this requires reduction between threads.
-This solution introduces   synchronizations and leads to low performance.
-The peak performance will be attained if the warps have the same amount of work, and every  thread in a warp is responsible for all contributions to an output coefficient.
-In order to achieve this goal, we define a task object.
-A task is a set of 32 resulting coefficients in a specific order.
-A task tells a  warp how to calculate coefficients (the thread number one of the warp takes the first coefficient of the task and so on).
-The warps may not have the same number of tasks, as we balance the workload.
-We proceed in two  steps: first on CPU, where a preliminary work is done, namely all information on the tasks is determined, and then, the calculation on GPU.
+\item  Second, at the beginning we  cached both input polynomials into the texture memory. 
+When a polynomial multiplication begins,  every warp gets the corresponding tasks (at least one) from the task's list and loops over its number of tasks. Then,
+the threads in a warp get their respective output coefficients and iterate over the number of  contributions.
+For every iteration over contributions, the threads determine the input coefficients, read the corresponding data in the texture memory, and perform the arithmetic operations: an extended multiplication and an optional long addition (if the result coefficient has several contributions).
+The calculation of an output coefficient  is  done in fast memory (texture for the input data, and register for the intermediate result).
+When the loop over contributions terminates,  we transfer the coefficients into the an intermediate buffer in the global memory, but in SoA  order for the final reduction. 
+The reduction is done using the standard method of NVIDIA \cite{CUDAReduction} to achieve the highest performance.
 
-For the first step, we create a dummy resulting polynomial  represented by the indices and the contributions. They are sorted by descending order of the number of contributions
-and grouped by 32 as a task. These tasks are reordered over the warps to balance the amount of work. An illustration of this algorithm is given on  figure \ref{algo_GPU}. 
-The more coefficients the polynomials have, the more uniform the distribution of the tasks over the warps will be, and the more efficient the calculation.
-We fix the maximum number of warps to eight (8 warps/256 threads), as far as possible, we assign at least two tasks per warp.
-Thus, depending on  the number of coefficients a warp may calculate one task, whereas another warp may have several tasks to calculate.
- When the determination of the list of the tasks is over,  we transfer  asynchronously it to the global memory.
-This method does not  change the ordering of the data. As for the CPU version, we have an AoS order.
  \begin{table}[t] 
@@
 
-This tasks determination is executed for all polynomial types only once.
-All the information is kept into the global memory and reuse if we need to calculate several times the same kind of inner products.
- As soon as,  all task of lists are calculated,  we transfer the data of the polynomials asynchronously onto the device (the global memory of the GPU is allocated only once and reuse),
- and the calculation of all polynomial multiplications start.
+\end{enumerate}
 
- First, both input polynomials  are cached into the texture memory. 
-When a polynomial multiplication begins,  every warp gets the corresponding tasks (at least one) from the task's list and loops over its number of tasks. Then,
-the threads in a warp get their respective output coefficients and iterate over the number of  contributions.
-For every iteration over contributions, the threads determine the input coefficients, read the corresponding data in the texture memory, and perform the arithmetic operations: an extended multiplication and an optional long addition (if the result coefficient has several contributions).
-The calculation of an output coefficient  is  done in fast memory (texture for the input data, and register for the intermediate result).
-When the loop over contributions terminates,  we transfer the coefficients into the an intermediate buffer in the global memory, but in SoA  order for the final reduction. 
-The reduction is done using the standard method of NVIDIA \cite{CUDAReduction} to achieve the highest performance.
+This algorithm gives good  performance on a GPU because the warps have approximatively the same amount of work,  and every thread in a warp will be responsible for all contributions to an output coefficient.
+The contributions are not dispatched between the threads, thus we do not  introduces   synchronizations and leads to low performance. 
+This  algorithm has the great advantage of working well with dense - triangular polynomials, and it could be extended to sparse polynomials, as structure of the polynomial is not primordial.
+The main difficulty consists  of designing an efficient function GetIndices (\ref{ALGO}) which determines the contributions.
 
-%  It can be both in the shared, one polynomial into the shared and the second one into the texture or two in the texture memory. 
-%The determination of which  fast memory (shared or texture) is determined during the compilation. Thus we avoid if statement and probable sequentialization of the warps.
-
-To conclude, this  algorithm has the great advantage of working well with dense and triangular polynomials. 
-The main difficulty consists  of designing an efficient function GetIndices (\ref{ALGO}) which determines the contributions.
- We also developed an optional hybrid mode between CPU/GPU. This  technique is usual in GPU computing.
+To conclude,  We also developed an optional hybrid mode between CPU/GPU. This  technique is usual in GPU computing.
 We split the input vectors of polynomials into two chunks of data, execute concurrently on CPU and GPU. 
@@
 
-\begin{figure}[t!]
-\begin{center}
-\mbox{
-\hspace{-0.5cm}
-\includegraphics[scale=0.37, angle=-90]{ME128.eps} 
-\hspace{-0.4cm}
-\includegraphics[scale=0.37, angle=-90]{MC128.eps} 
-%\hspace{-0.4cm}
-%\includegraphics[scale=0.25, angle=-90]{ME256.eps} 
-}
-\caption{Left: Inner product of dense polynomials with up to 3 variables and 126 to 256 bits coefficients.
-Right: Inner product of triangular polynomials with up to 3 variables and 126 to 256 bits coefficients.
-Size of the vector 4096.
-For four variables we are limited by the memory of the node.
-GMP gives similar results whatever the polynomials.}
-\label{ResME}
-\end{center}
-\end{figure}
-
 Benchmarks are performed on the facilities of the Swiss Center for Scientific Computing (CSCS), on a Cray  XK7 cluster for the GPU, and a Sandy Bridge  node with 32 logical cores  (HyperThreading on) Xeon E5-2670.
@@
 
-Evaluating  integer benchmarks  is not easy: unlike the floating point, integer operations are performed by ALU (and not on the single FPU),  the number of  ALUs  between processors (Intel, AMD, IBM) and the execution time of the operations are  not constant \cite{ASMcost} (constant for FPU).
 We compute the  result  numbers as operations \footnote{An operation can be a long multiplication, addition and so on}   per seconds $GOP/s=10^{-9} \times batch/t$, where  $batch$ is the number of long integer operations computed during the execution (additions and multiplications), $t$ [s]  is the elapsed time.
 For more realistic comparison the execution time for GPU contains the memory  transfer.
+Contrary to Floating-point, we can not present Integers operations as FLOPS, in fact nteger operations are performed by ALU (and not on the FPU),  the number of  ALUs  between processors (Intel, AMD, IBM) and the execution time of the operations are  not constant \cite{ASMcost} (quasi - constant for FPU).
+We could calculate a MIPS (Million Integer Per Seconds) by counting all the integer instruction into our code and divide by the elapse time.
+But, we can not compare to any peak of performance, because the integer peak is really different from addition to multiplication, as well show by \cite{Vladimirov2012}.
 
@@
 \hspace{-0.5cm}
+\includegraphics[scale=0.37, angle=-90]{ME128.eps} 
+\hspace{-0.4cm}
+\includegraphics[scale=0.37, angle=-90]{MC128.eps} 
+%\hspace{-0.4cm}
+%\includegraphics[scale=0.25, angle=-90]{ME256.eps} 
+}
+\caption{Left: Inner product of dense polynomials with up to 3 variables and 126 to 256 bits coefficients.
+Right: Inner product of triangular polynomials with up to 3 variables and 126 to 256 bits coefficients.
+Size of the vector 4096.
+For four variables we are limited by the memory of the node.
+GMP gives similar results whatever the polynomials.}
+\label{ResME}
+\end{center}
+\end{figure}
+
+\begin{figure}[t!]
+\begin{center}
+\mbox{
+\hspace{-0.5cm}
 \includegraphics[scale=0.37, angle=-90]{ME128MIPS.eps} 
@@
 
-Although translating the results into MIPS is difficult, we attempt it, at least on the GPU, because the GPU kernel of the extended multiplication utilizes only one PTX instruction (\texttt{madc}).
+Although translating the results into MIPS is difficult, we attempt it, at least on the GPU, because the GPU kernel of the extended multiplication utilizes only one PTX instruction (\texttt{madc}), we can compare to a theoretical peak.
  Counting the total number of  this ASM instructions during  the inner products, and dividing by the elapsed time, we obtain an estimation in GIPS (Giga Instructions Per Second).
@@
 The limited performances of GMP for small size were the main argument to develop the VLI library, although GMP excels in thousand bits arithmetic where our library can not operate.
-As future work, we would like to implement, new kinds of polynomials, and a better management 
-of  the memory to maintain the performance on GPU.
+As future work, we would like to implement, new kinds of polynomials, and a better management  of  the memory to maintain the performance on GPU.
+And them, we would like explore new  SIMD architecture Xeon PHI where the carry bit is implemented.
 

### 2013-01-14 doc/paper_vli/vli.tex
morning reading paper VLI

@@
 %\section{}
-%\subsection{}
+%\subsection{} giamarchi
 %\vspace{-0.65cm}
@@
 
-The Very Large Integer library performs  standards  operations for large integer arithmetic and polynomial. This library also provides optimized implementations of inner products  for GPU.
-This library is composed of two distinct modules,  but enough closed  to work together by the mechanism of template instantiation. VLI is a fix size library from 128 to 512 bits,
-as the polynomial module  which is limited for dense and triangular polynomials  up to 4 variables.
-The core of VLI is build under ASM language to provide the best performance.
+The Very Large Integer library performs  standards  operations for large integer arithmetic and polynomial.
+This library also provides optimized implementations of inner products  of polynomials for GPU.
+This library is composed of two distinct modules,  but enough closed  to work together by the mechanism of template instantiation. VLI is a fix size library from 128 to 512 bits.
+The polynomial module  is limited for dense and triangular polynomials  up to 4 variables.
+The core of VLI is build under assembly language to provide the best performance.
 Every arithmetic operations
-are executed by a specific kernel generated specifically during the compilation by an ASM generator. Moreover, VLI allows multithreading and hybrid  GPU usage for inner products of polynomials.
-Multithreading is build on OpenMP library and GPU module under CUDA. As the CPU, the GPU is a fixe size library, and ASM kernels are also determinated specifically during the compilation.
+are executed by a specific kernel generated specifically during the compilation by an assembly language generator.
+Moreover, VLI allows multithreading and hybrid  GPU usage for the inner products of polynomials.
+Multithreading is build with OpenMP library whereas the GPU module utilizes CUDA. As on the CPU, PTX kernels are generated specifically during the compilation.
 
@@
 GMP is polyvalent, nevertheless this polyvalence is attained by the usage of  dynamic memory.
-It is well know dynamic allocation is time consuming, and it can become a bottle neck for small operations which are in the working range.  
-Alternatively for polynomial arithmetic, a few libraries like  LINBOX or NTL  provide a solution but  they use GMP when needed. 
+It is well know dynamic allocation is time consuming, and it can become a bottle neck for small operations, which correspond to our working range.  
+Alternatively, for polynomial arithmetic, a few libraries like  LINBOX or NTL  provide a solution but  they use GMP when needed. 
 Although hardware accelerators were introduced  in HPC a few years ago, as the GPUs  by NVIDIA. Only a few libraries have been adapted 
@@
 
-In this work, we are going to highlight the speedup given by a tiny ASM library without any dynamic memory, suggest an alternative for small size integer. 
-And propose an efficient implementation of inner product on GPU with the latest CUDA PTX features.
+In this work, we are going to highlight the speedup given by a tiny ASM library without any dynamic memory,  suggest an alternative at GMP for small size integer. 
+We also propose an efficient implementation of the inner products for  dense and triangular multivariable polynomials  on GPU with the latest CUDA PTX features.
 
@@
 
-The library performs standards operations for polynomials (up to four variables) and optimized solvers for  the inner products of vectors of polynomials (presented in the next section).
-The polynomials are defined like  e.g. our variables
+The library performs standards operations for polynomials (up to four variables) and optimized solvers for  the inner products of vectors of polynomials (presented in the next sections).
+The polynomials are defined like  $e.g.$ our variables
 \begin{eqnarray}
@@
 
-\section{VLI  CPU library}
+\section{Very Large Integer library}
 
@@
 We coded the kernels using the GNU inline assembly.
-Generating by hand all these  assembly strings (kernels)  can be painful and error-prone\footnote{The library has near 10000 lines of assembly in total for x86-64, Power64, and PTX}.
+Generating by hand all these  assembly strings (kernels)  can be painful and error-prone\footnote{The library has near 10 000 lines of assembly in total for x86-64, Power64, and PTX}.
 We designed a assembly generator. It creates all versions of a kernel during the compilation.   
@@
 
-\subsection{Memory layout and SIMD possibility}
+\subsection{Memory layout}
 
@@
 
-We have choose the SoA for the two following reasons :
+We have chosen the SoA for the two following reasons :
 
@@
 The disagreement  of this choice, the SoA order does not facilitate the usage of the SIMD instruction such as the AVX instructions of the Sandy Bridge x86-64 architecture.
-Applying SIMD to the AoS, for an addition or a subtraction is a nonsense because there is no specific instruction to propagate carry bit between entries of the SIMD register.
+Applying SIMD to the AoS for an addition/subtraction is a nonsense because there is no specific instruction to propagate carry bit between entries of the SIMD register.
 It could be  apply  on multiplications for integer  larger than 1024-bit \cite{SIMD}, unfortunately, it does not correspond to our working range. Consequently, we do not 
@@
 The GPU algorithm of the inner product is divided into two parts.
-A preliminary step on the CPU,  where we prepare the calculation blue print of the  contribution.
-And the calculation itself on the GPU
+A preliminary step on the CPU executed only once,  where we prepare the execution blue print of the  contribution.
+And the second step the calculation itself on the GPU
 
@@
 Thus, depending on  the number of coefficients a warp may calculate one task, whereas another warp may have several tasks to calculate.
-When the determination of the list of the tasks is over,  we transfer  asynchronously  to the global memory the data of the polynomial and all the tasks.
-This method does not  change the ordering of the data, ss for the CPU version, we have an AoS order.
-But we have now informations how calculate the polynomial into a efficient way.
-In terms of memory of the GPU, we allocate allocated only once, and reuse inner products after inner products., thus we save time.
+When the determination of the task lists is over,  we transfer  asynchronously  to the global memory the data of the polynomial and all the task lists.
+This method does not  change the ordering of the data, as for the CPU version, we have an AoS order.
+In terms of memory of the GPU, we allocate allocated only once, and reuse inner products after inner products, thus we do not wast time on successive reallocations of the memory.
 
@@
 \end{figure}
-
-\item  Second, at the beginning we  cached both input polynomials into the texture memory. 
-When a polynomial multiplication begins,  every warp gets the corresponding tasks (at least one) from the task's list and loops over its number of tasks. Then,
-the threads in a warp get their respective output coefficients and iterate over the number of  contributions.
-For every iteration over contributions, the threads determine the input coefficients, read the corresponding data in the texture memory, and perform the arithmetic operations: an extended multiplication and an optional long addition (if the result coefficient has several contributions).
-The calculation of an output coefficient  is  done in fast memory (texture for the input data, and register for the intermediate result).
-When the loop over contributions terminates,  we transfer the coefficients into the an intermediate buffer in the global memory, but in SoA  order for the final reduction. 
-The reduction is done using the standard method of NVIDIA \cite{CUDAReduction} to achieve the highest performance.
 
@@
 
+\item  Second, at the beginning we  cached both input polynomials into the texture memory. 
+When a polynomial multiplication begins,  every warp gets the corresponding tasks (at least one) from the task's list and loops over its number of tasks. Then,
+the threads in a warp get their respective output coefficients and iterate over the number of  contributions.
+For every iteration over contributions, the threads determine the input coefficients, read the corresponding data in the texture memory, and perform the arithmetic operations: an extended multiplication and an optional long addition (if the result coefficient has several contributions).
+The calculation of an output coefficient  is  done in fast memory (texture for the input data, and register for the intermediate result).
+When the loop over contributions terminates,  we transfer the coefficients into the an intermediate buffer in the global memory, but in SoA  order for the final reduction. 
+The reduction is done using the standard method of NVIDIA \cite{CUDAReduction} to achieve the highest performance.
+
 \end{enumerate}
 
-This algorithm gives good  performance on a GPU because the warps have approximatively the same amount of work,  and every thread in a warp will be responsible for all contributions to an output coefficient.
+This algorithm is efficient on a GPU because the warps have approximatively the same amount of work,  and every thread in a warp will be responsible for all contributions to an output coefficient.
 The contributions are not dispatched between the threads, thus we do not  introduces   synchronizations and leads to low performance. 
 This  algorithm has the great advantage of working well with dense - triangular polynomials, and it could be extended to sparse polynomials, as structure of the polynomial is not primordial.
-The main difficulty consists  of designing an efficient function GetIndices (\ref{ALGO}) which determines the contributions.
+The main difficulty consists  of designing an efficient function \texttt{GetIndices} (table \ref{ALGO}) which determines the contributions.
 
@@
 As future work, we would like to implement, new kinds of polynomials, and a better management  of  the memory to maintain the performance on GPU.
-And them, we would like explore new  SIMD architecture Xeon PHI where the carry bit is implemented.
+And then, we would like explore new  the SIMD architecture Xeon PHI
+where the carry bit is natively implemented.
 

### 2013-01-14 doc/paper_vli/vli.tex
fix AoS SoA paper

@@
 
-We have privileged the  array  of structure  (SoA ) detrimental to the  structure of array (SoA)
+We have privileged the  array  of structure  (AoS ) detrimental to the  structure of array (SoA)
 The AoS consists in saving the data in this hierarchy: VLI - series of VLI (polynomial coefficient) - series of polynomials (the vector), where  the SoA structure interleaves the coefficients of the VLI and the polynomials.
@@
 
-We have chosen the SoA for the two following reasons :
+We have chosen the AoS for the two following reasons :
 
 \begin{enumerate}
-\item SoA order allows the natural propagation of the carry bit between the entries of the VLI with the pair of mnemonics \texttt{addq/adcq}, as the utilization of the extended multiplication without any padding.
+\item AoS order allows the natural propagation of the carry bit between the entries of the VLI with the pair of mnemonics \texttt{addq/adcq}, as the utilization of the extended multiplication without any padding.
 \item The VLI and the polynomial module of the library can be separated and use independently. Thinks impossible if the coefficients are interleaved.
@@
 
-The disagreement  of this choice, the SoA order does not facilitate the usage of the SIMD instruction such as the AVX instructions of the Sandy Bridge x86-64 architecture.
+The disagreement  of this choice, the AoS order does not facilitate the usage of the SIMD instruction such as the AVX instructions of the Sandy Bridge x86-64 architecture.
 Applying SIMD to the AoS for an addition/subtraction is a nonsense because there is no specific instruction to propagate carry bit between entries of the SIMD register.

### 2013-01-15 doc/paper_vli/vli.tex
paper vi

--- 2013-01-14 doc/paper_vli/vli.tex
@@
 
-The Very Large Integer library performs  standards  operations for large integer arithmetic and polynomial.
-This library also provides optimized implementations of inner products  of polynomials for GPU.
-This library is composed of two distinct modules,  but enough closed  to work together by the mechanism of template instantiation. VLI is a fix size library from 128 to 512 bits.
-The polynomial module  is limited for dense and triangular polynomials  up to 4 variables.
-The core of VLI is build under assembly language to provide the best performance.
+The Very Large Integer library realizes large arithmetic , polynomials  and vector operations............ There are three 3 different modules.
+
+The large integer module models large signed integer from 128 to 512 bits.
+The operations supported are bits operations , the four basic operators $(\times,+,/,-)$ , extended multiplications and bit shift operations.
+.....SOMETHING ELSE .......
+
+The polynomial module models polynomials up to 4 variables for dense and triangular polynomials, As previously, the polynomials usual  operations between polynomials and operations with monomials.
+.....SOMETHING ELSE .......
+
+The vector of polynomial module provides a container for polynomials and allows support for optimized inners products.
+
+Modules allow flexibility, they can be interchangeable.
+........
+polynomial can have large integer coefficient or other coefficient .....
+All modules combined, we can have access to specialized GPU solver for inner product.
+
+We goal of this library is perfumed efficient arithmetic for polynomial with large coefficient and execute efficient inner product on GPU........ ANDRES SHOUD WE TALK ON VLI INT SE OR QE ....?
+
+ Computer sciences provide solution for large arithmetic, and polynomial operations.
+For large arithmetic the most famous is GMP .....
+WRITE GMP is polyvalent but polyvalence obtained by dynamic memory.....
+ Polynomial operations can be performed by libraries as LINBOX and NTL.
+.........
+these libraries allows manipulation but with GMP, it not performable for small coefficient ......
+The GPU is a revolution is programming model tell the LAPACK -> MAGMA .....
+...... polynomial ported only for large size with FFT approach inefficient for pb .... It does not exists optimized polynomials inner products for GPU ......
+
+%standards  operations for large integer arithmetic and polynomial. This library also provides optimized implementations of inner products  for GPU.
+%This library is composed of two distinct modules,  but enough closed  to work together by the mechanism of template instantiation. VLI is a fix size library from 128 to 512 bits,
+%as the polynomial module  which is limited for dense and triangular polynomials  up to 4 variables.
+The core of VLI is build under ASM language to provide the best performance.
 Every arithmetic operations
-are executed by a specific kernel generated specifically during the compilation by an assembly language generator.
-Moreover, VLI allows multithreading and hybrid  GPU usage for the inner products of polynomials.
-Multithreading is build with OpenMP library whereas the GPU module utilizes CUDA. As on the CPU, PTX kernels are generated specifically during the compilation.
+%are executed by a specific kernel generated specifically during the compilation by an ASM generator. Moreover, VLI allows multithreading and hybrid  GPU usage for inner products of polynomials.
+%Multithreading is build on OpenMP library and GPU module under CUDA. As the CPU, the GPU is a fixe size library, and ASM kernels are also determinated specifically during the compilation.
 
-In large arithmetic, the standard library is GMP\footnote{GMP is the standard library for long arithmetic, gmplib.org}.
+%In large arithmetic, the standard library is GMP\footnote{GMP is the standard library for long arithmetic, gmplib.org}.
 GMP is polyvalent, nevertheless this polyvalence is attained by the usage of  dynamic memory.
-It is well know dynamic allocation is time consuming, and it can become a bottle neck for small operations, which correspond to our working range.  
-Alternatively, for polynomial arithmetic, a few libraries like  LINBOX or NTL  provide a solution but  they use GMP when needed. 
-Although hardware accelerators were introduced  in HPC a few years ago, as the GPUs  by NVIDIA. Only a few libraries have been adapted 
-A prominent example of such an adaption are the MAGMA libraries, which port the well known BLAS/LAPACK packages to GPU accelerators.
-The polynomials libraries, however,  have not done the transition to these new hardwares yet.
-During the last few years, the polynomial multiplication was successfully implemented on this new hardware
-\cite{Govindaraju2008, Emeliyanenko:2009, Moreno2010}. All these works focus on large polynomials, using a FFT approach.
+%It is well know dynamic allocation is time consuming, and it can become a bottle neck for small operations which are in the working range.  
+%Alternatively for polynomial arithmetic, a few libraries like  LINBOX or NTL  provide a solution but  they use GMP when needed. 
+%Although hardware accelerators were introduced  in HPC a few years ago, as the GPUs  by NVIDIA. Only a few libraries have been adapted 
+%A prominent example of such an adaption are the MAGMA libraries, which port the well known BLAS/LAPACK packages to GPU accelerators.
+%The polynomials libraries, however,  have not done the transition to these new hardwares yet.
+%During the last few years, the polynomial multiplication was successfully implemented on this new hardware
+%\cite{Govindaraju2008, Emeliyanenko:2009, Moreno2010}. All these works focus on large polynomials, using a FFT approach.
 
-In this work, we are going to highlight the speedup given by a tiny ASM library without any dynamic memory,  suggest an alternative at GMP for small size integer. 
-We also propose an efficient implementation of the inner products for  dense and triangular multivariable polynomials  on GPU with the latest CUDA PTX features.
+%In this work, we are going to highlight the speedup given by a tiny ASM library without any dynamic memory, suggest an alternative for small size integer. 
+%And propose an efficient implementation of inner product on GPU with the latest CUDA PTX features.
 
-%In the present work, we focus on the inner products of dense and triangular multivariable polynomials.  The coefficients of the polynomials are large integers. 
-%In contrast to GMP which dynamically adapts the size of the integers, the presented long integer library is based on static integer sizes.
-%Moreover, our library is natively compatible with GPU to allow hybrid calculation. GPUs impose contiguous memory which is impossible with a GMP approach due to the dynamic allocated memory.
-%Thus, we propose a  library based on contiguous memory where the exact assembly kernels are generated during the compilation for CPU and GPU.
-%The static integer sizes allow the compiler to aggressivly optimize the non-assembly parts of the library and avoids costly heap allocations for the data.
+%\section{Polynomial and inner products definitions}
 
-\section{Polynomial and inner products definitions}
+%The library performs standards operations for polynomials (up to four variables) and optimized solvers for  the inner products of vectors of polynomials (presented in the next section).
+%The polynomials are defined like  e.g. our variables
+%\begin{eqnarray}
+%P_a(x,y,z,w) & = & \sum_{i=0}^n \sum_{j=0}^n  \sum_{k=0}^n \sum_{l=0}^n  a_{ijkl} x^i y^j z^k w^l , \,\,\,\,\,\, \text{or} \\
+%                   & = & \sum_{i=0}^n \sum_{j=0}^{(n-i)} \sum_{k=0}^{(n-i-j)} \sum_{l=0}^{(n-i-j-k)}  a_{ijkl} x^i y^j z^k w^l  .
+%\end{eqnarray}
+%where  $x,\,y,\,z$ and $w$ are symbolic variables,  $i,\,j,\,k$ and $l$  the indices, $a$  the coefficients, and $n$ the order of the polynomial.
+%The polynomials of the form  (1) and (2) are named dense and triangular respectively. Consider now two vectors of polynomials of the  same type with $m$ 
+%entries  $\boldsymbol{v}_1$ and  $\boldsymbol{v}_2$. The inner products  consist of the product 
+%of all entries  associated with a global reduction. 
+ %\begin{eqnarray}
+% P_c = \sum_{i=0}^m \boldsymbol{v}_{1_i} \times  \boldsymbol{v}_{2_i}  \label{poly1} \label{InnerProduct}
+%\end{eqnarray}
+%where $\times$ indicates the multiplication between two polynomials. The multiplications use usual rules of polynomials multiplication. 
+%A final coefficient may be the sum of several coefficients (which we call contributions in the following passage), resulting from the sum of several multiplications.
+These coefficients are the cross terms.
+%The cross terms are not a problem for the CPU version, as a polynomial multiplication is done by a single thread, but they are the bottleneck for the GPU version.
+%The number of coefficient for a dense and triangular polynomial is given by $(n+1)^{\kappa}$ and  $(\kappa+n-1)! / ( (\kappa-1)! n! )$ respectively, where $\kappa$ is the number of variables.
 
-The library performs standards operations for polynomials (up to four variables) and optimized solvers for  the inner products of vectors of polynomials (presented in the next sections).
-The polynomials are defined like  $e.g.$ our variables
-\begin{eqnarray}
-P_a(x,y,z,w) & = & \sum_{i=0}^n \sum_{j=0}^n  \sum_{k=0}^n \sum_{l=0}^n  a_{ijkl} x^i y^j z^k w^l , \,\,\,\,\,\, \text{or} \\
-                   & = & \sum_{i=0}^n \sum_{j=0}^{(n-i)} \sum_{k=0}^{(n-i-j)} \sum_{l=0}^{(n-i-j-k)}  a_{ijkl} x^i y^j z^k w^l  .
-\end{eqnarray}
-where  $x,\,y,\,z$ and $w$ are symbolic variables,  $i,\,j,\,k$ and $l$  the indices, $a$  the coefficients, and $n$ the order of the polynomial.
-The polynomials of the form  (1) and (2) are named dense and triangular respectively. Consider now two vectors of polynomials of the  same type with $m$ 
-entries  $\boldsymbol{v}_1$ and  $\boldsymbol{v}_2$. The inner products  consist of the product 
-of all entries  associated with a global reduction. 
- \begin{eqnarray}
- P_c = \sum_{i=0}^m \boldsymbol{v}_{1_i} \times  \boldsymbol{v}_{2_i}  \label{poly1} \label{InnerProduct}
-\end{eqnarray}
-where $\times$ indicates the multiplication between two polynomials. The multiplications use usual rules of polynomials multiplication. 
-A final coefficient may be the sum of several coefficients (which we call contributions in the following passage), resulting from the sum of several multiplications.
-These coefficients are the cross terms.
-The cross terms are not a problem for the CPU version, as a polynomial multiplication is done by a single thread, but they are the bottleneck for the GPU version.
-The number of coefficient for a dense and triangular polynomial is given by $(n+1)^{\kappa}$ and  $(\kappa+n-1)! / ( (\kappa-1)! n! )$ respectively, where $\kappa$ is the number of variables.
-
-In terms of  programming and parallelization, in the CPU version, the previous equations can be reduced to independent loops with a final reduction.
-The CPU parallelization has been done with OpenMP.
+%In terms of  programming and parallelization, in the CPU version, the previous equations can be reduced to independent loops with a final reduction.
+%The CPU parallelization has been done with OpenMP.
 The OpenMP threads calculate several  polynomial multiplications and make a local reduction.
 At the end, a final reduction is performed  by the main thread.
+
  % \footnote{In OpenMP reduction for class is not supported}. Results showed a perfect scalability of the OpenMP code.
 
-\section{Very Large Integer library}
+%\section{VLI  CPU library}
 
-The coefficients of the polynomials are represented by a large integer named the VLI number.  A VLI number is characterized by a size in bits.
-Our current implementation allows for integer sizes from 128 to 512 bits, but is easily extendable to larger integers.
-Inside the memory, the VLI numbers are  contiguous. The data type container is a 64 bit unsigned integer array.
+%The coefficients of the polynomials are represented by a large integer named the VLI number.  A VLI number is characterized by a size in bits.
+%Our current implementation allows for integer sizes from 128 to 512 bits, but is easily extendable to larger integers.
+%Inside the memory, the VLI numbers are  contiguous. The data type container is a 64 bit unsigned integer array.
 
-The basic operations supported by VLI are the four basic operations: $+,-,\times, /$, extended multiplications and bit shift.
+%The basic operations supported by VLI are the four basic operations: $+,-,\times, /$, extended multiplications and bit shift.
 The library allows  operations between operands of the same type, and with basic  integers.
-For performance reasons, we have a specific solver for every operation except  for division.
-All the algorithms and operations are based on basic arithmetic but have been implemented in assembly language to make use of hardware features which are not accessible by plain C++.
+%For performance reasons, we have a specific solver for every operation except  for division.
+%All the algorithms and operations are based on basic arithmetic but have been implemented in assembly language to make use of hardware features which are not accessible by plain C++.
 
-\subsection{Assembly language}
+\section{Implementation of the library}
+\subsubsection{memory layout}
 
-We applied algorithms based on childhood  experience for large arithmetic. The implementation necessitates a carry bit and an extended multiplication. The hardware supports
+.... contrary to GMP  explain a static version of large arithmetic can be faster than a dynamic one....   ....   VLI number are base on 64-bit integer array .... size compile time polymorphism ...
+.... polynomials constructed over boost::array ............. the vector of polynomial is std::vector..... combined together we have a contiguous memory, and more precisely  SoA (show the figure). 
+...... It exists a second order name AoS, it does not allow a construction into module ............
+
+\begin{figure}
+    \begin{minipage}{0.50\linewidth}
+        \begin{eqnarray}
+        \tiny{
+                 \underbrace{
+                 \underbrace{ \underbrace{a^0_{00} | a^1_{00} | a^2_{00}}_{VLI, a_{00} } || \underbrace{a^0_{10} | a^1_{10} | a^2_{10}}_{VLI, a_{10}} ||   \dots  }_{1^{st} \textrm{polynomial}} 
+                  |||    \underbrace{ \underbrace{b^0_{00} | a^1_{00} | b^2_{00}}_{VLI, b_{00} } || \underbrace{b^0_{10} | b^1_{10} | b^2_{10}}_{VLI, b_{10}} ||   \dots  }_{2^{nd} \textrm{polynomial} } }_{\textrm{Vector of polynomials, AoS order}} \nonumber
+            }
+        \end{eqnarray}
+    \end{minipage}
+\begin{minipage}{0.50\linewidth}
+    \begin{eqnarray}
+    \tiny{
+        \underbrace{
+        \underbrace{a^0_{00} | a^0_{10} | \dots || a^1_{00} | a^1_{10} | \dots  }_{1st  \textrm{ nested VLI/Polynomial} } |||  \underbrace{b^0_{00} | b^0_{10} | \dots || b^1_{00} | b^1_{10} | \dots  }_{2nd  \textrm{ nested VLI/Polynomial} }
+        }_{\textrm{Vector of polynomials, SoA Order}} \nonumber }
+        \end{eqnarray}
+\end{minipage}
+\caption{Representation of the AoS Order (left) or SoA Order (right) for a vector of polynomial \label{AOSSOA}.  The SoA structure limits the usage of the VLI module alone.}
+\end{figure}
+
+......start transition ...........  Polynomials operations necessitates large arithmetic operations for the coefficients , ..... coefficient operations must be fast, we need ASM .....
+
+\subsubsection{Kernels and assembly  language}
+
+We applied algorithms based on childhood  experience for large arithmetic........... for the specific range classical operations are enough ........
+The implementation necessitates a carry bit and an extended multiplication. The hardware supports
 the carry bit in the status register, as the extended multiplications, but they are only accessible by the assembly language.
@@
 
-\subsection{Memory layout}
-
-We have privileged the  array  of structure  (SoA ) detrimental to the  structure of array (SoA)
-The AoS consists in saving the data in this hierarchy: VLI - series of VLI (polynomial coefficient) - series of polynomials (the vector), where  the SoA structure interleaves the coefficients of the VLI and the polynomials.
-A representation of both memory layout is presented on figure \ref{AOSSOA}.
-
-\begin{figure}
-    \begin{minipage}{0.50\linewidth}
-        \begin{eqnarray}
-        \tiny{
-                 \underbrace{
-                 \underbrace{ \underbrace{a^0_{00} | a^1_{00} | a^2_{00}}_{VLI, a_{00} } || \underbrace{a^0_{10} | a^1_{10} | a^2_{10}}_{VLI, a_{10}} ||   \dots  }_{1^{st} \textrm{polynomial}} 
-                  |||    \underbrace{ \underbrace{b^0_{00} | a^1_{00} | b^2_{00}}_{VLI, b_{00} } || \underbrace{b^0_{10} | b^1_{10} | b^2_{10}}_{VLI, b_{10}} ||   \dots  }_{2^{nd} \textrm{polynomial} } }_{\textrm{Vector of polynomials, AoS order}} \nonumber
-            }
-        \end{eqnarray}
-    \end{minipage}
-\begin{minipage}{0.50\linewidth}
-    \begin{eqnarray}
-    \tiny{
-        \underbrace{
-        \underbrace{a^0_{00} | a^0_{10} | \dots || a^1_{00} | a^1_{10} | \dots  }_{1st  \textrm{ nested VLI/Polynomial} } |||  \underbrace{b^0_{00} | b^0_{10} | \dots || b^1_{00} | b^1_{10} | \dots  }_{2nd  \textrm{ nested VLI/Polynomial} }
-        }_{\textrm{Vector of polynomials, SoA Order}} \nonumber }
-        \end{eqnarray}
-\end{minipage}
-\caption{Representation of the AoS Order (left) or SoA Order (right) for a vector of polynomial \label{AOSSOA}.  The SoA structure limits the usage of the VLI module alone.}
-\end{figure}
-
-We have chosen the SoA for the two following reasons :
-
-\begin{enumerate}
-\item SoA order allows the natural propagation of the carry bit between the entries of the VLI with the pair of mnemonics \texttt{addq/adcq}, as the utilization of the extended multiplication without any padding.
-\item The VLI and the polynomial module of the library can be separated and use independently. Thinks impossible if the coefficients are interleaved.
-\end{enumerate}
-
+..... We does not use SIMD .... because the data order does not help .....
+....  simplify .......
 The disagreement  of this choice, the SoA order does not facilitate the usage of the SIMD instruction such as the AVX instructions of the Sandy Bridge x86-64 architecture.
-Applying SIMD to the AoS for an addition/subtraction is a nonsense because there is no specific instruction to propagate carry bit between entries of the SIMD register.
+Applying SIMD to the AoS, for an addition or a subtraction is a nonsense because there is no specific instruction to propagate carry bit between entries of the SIMD register.
 It could be  apply  on multiplications for integer  larger than 1024-bit \cite{SIMD}, unfortunately, it does not correspond to our working range. Consequently, we do not 
 use the multimedia instruction into the ASM kernels.
+............
 
-%The choice of the memory layout is important.
+\section{Advance solver : Inner products}
+
+..... Introduce the capacity of the lib to perform inner product ................
+...... make a short math description .........
+
+%The library performs standards operations for polynomials (up to four variables) and optimized solvers for  the inner products of vectors of polynomials (presented in the next section).
+%The polynomials are defined like  e.g. our variables
+%\begin{eqnarray}
+%P_a(x,y,z,w) & = & \sum_{i=0}^n \sum_{j=0}^n  \sum_{k=0}^n \sum_{l=0}^n  a_{ijkl} x^i y^j z^k w^l , \,\,\,\,\,\, \text{or} \\
+%                   & = & \sum_{i=0}^n \sum_{j=0}^{(n-i)} \sum_{k=0}^{(n-i-j)} \sum_{l=0}^{(n-i-j-k)}  a_{ijkl} x^i y^j z^k w^l  .
+%\end{eqnarray}
+%where  $x,\,y,\,z$ and $w$ are symbolic variables,  $i,\,j,\,k$ and $l$  the indices, $a$  the coefficients, and $n$ the order of the polynomial.
+%The polynomials of the form  (1) and (2) are named dense and triangular respectively. Consider now two vectors of polynomials of the  same type with $m$ 
+%entries  $\boldsymbol{v}_1$ and  $\boldsymbol{v}_2$. The inner products  consist of the product 
+%of all entries  associated with a global reduction. 
+ %\begin{eqnarray}
+% P_c = \sum_{i=0}^m \boldsymbol{v}_{1_i} \times  \boldsymbol{v}_{2_i}  \label{poly1} \label{InnerProduct}
+%\end{eqnarray}
+%where $\times$ indicates the multiplication between two polynomials. % The multiplications use usual rules of polynomials multiplication.
+
+\subsubsection{CPU implementation}
+........ explain quickly ..... just loop other coefficient .... with the OMP /// ......
+%In terms of  programming and parallelization, in the CPU version, the previous equations can be reduced to independent loops with a final reduction.
+%The CPU parallelization has been done with OpenMP.
+The OpenMP threads calculate several  polynomial multiplications and make a local reduction.
+At the end, a final reduction is performed  by the main thread.
+
+....... transisition explain the pb of the cross term ..... 
+%A final coefficient may be the sum of several coefficients (which we call contributions in the following passage), resulting from the sum of several multiplications.
+These coefficients are the cross terms.
+%The cross terms are not a problem for the CPU version, as a polynomial multiplication is done by a single thread, but they are the bottleneck for the GPU version.
+%The number of coefficient for a dense and triangular polynomial is given by $(n+1)^{\kappa}$ and  $(\kappa+n-1)! / ( (\kappa-1)! n! )$ respectively, where $\kappa$ is the number of variables. 
+....... end pb .......
+..... GPU is interesting because we have a lot of threads available, we can calculate contribution but in a correct way to avoid pb .............
+
+%\subsection{Memory layout and SIMD possibility}
+
+%We have privileged the  array  of structure  (SoA ) detrimental to the  structure of array (SoA)
+%The AoS consists in saving the data in this hierarchy: VLI - series of VLI (polynomial coefficient) - series of polynomials (the vector), where  the SoA structure interleaves the coefficients of the VLI and the polynomials.
+%A representation of both memory layout is presented on figure \ref{AOSSOA}.
+
+%We have choose the SoA for the two following reasons :
+
+%\begin{enumerate}
+%\item SoA order allows the natural propagation of the carry bit between the entries of the VLI with the pair of mnemonics \texttt{addq/adcq}, as the utilization of the extended multiplication without any padding.
+%\item The VLI and the polynomial module of the library can be separated and use independently. Thinks impossible if the coefficients are interleaved.
+%\end{enumerate}
+
+%%The choice of the memory layout is important.
 It may facilitate  the usage of the SIMD instructions such as the AVX instructions of the Sandy Bridge x86-64 architecture.
 We choose the AoS for the following reasons:
-%Applying SIMD to the AoS, for an addition is a nonsense because there is no specific instruction to propagate carry bit between entries of the SIMD register.
-%The extended multiplication under AVX \texttt{\_\_mm256\_mul\_epu32}  allows only four multiplications of 32-bit, saving a factor of two, nevertheless, on Sandy Bridge, we have 3 ALU, we can not exclude
-%the execution of several long multiplication at the same time, and we lose the benefit of the SIMD version.
+%%Applying SIMD to the AoS, for an addition is a nonsense because there is no specific instruction to propagate carry bit between entries of the SIMD register.
+%%The extended multiplication under AVX \texttt{\_\_mm256\_mul\_epu32}  allows only four multiplications of 32-bit, saving a factor of two, nevertheless, on Sandy Bridge, we have 3 ALU, we can not exclude
+%%the execution of several long multiplication at the same time, and we lose the benefit of the SIMD version.
 Moreover, there is a large overhead  to prepare the registers and shuffle the data into the registers.
@@
 However SIMD has been used with success for  integers larger  than 1024-bit \cite{SIMD}.
-% TODO somehow the problem with SIMD operations is not entirely clear yet. You mention the issues, but I think you should elaborate them a little more.
+%% TODO somehow the problem with SIMD operations is not entirely clear yet. You mention the issues, but I think you should elaborate them a little more.
 
@@
 
-\section{Inner product booster on GPU}
+%The library performs standards operations for polynomials (up to four variables) and optimized solvers for  the inner products of vectors of polynomials (presented in the next section).
+%The polynomials are defined like  e.g. our variables
+%\begin{eqnarray}
+%P_a(x,y,z,w) & = & \sum_{i=0}^n \sum_{j=0}^n  \sum_{k=0}^n \sum_{l=0}^n  a_{ijkl} x^i y^j z^k w^l , \,\,\,\,\,\, \text{or} \\
+%                   & = & \sum_{i=0}^n \sum_{j=0}^{(n-i)} \sum_{k=0}^{(n-i-j)} \sum_{l=0}^{(n-i-j-k)}  a_{ijkl} x^i y^j z^k w^l  .
+%\end{eqnarray}
+%where  $x,\,y,\,z$ and $w$ are symbolic variables,  $i,\,j,\,k$ and $l$  the indices, $a$  the coefficients, and $n$ the order of the polynomial.
+%The polynomials of the form  (1) and (2) are named dense and triangular respectively. Consider now two vectors of polynomials of the  same type with $m$ 
+%entries  $\boldsymbol{v}_1$ and  $\boldsymbol{v}_2$. The inner products  consist of the product 
+%of all entries  associated with a global reduction. 
+ %\begin{eqnarray}
+% P_c = \sum_{i=0}^m \boldsymbol{v}_{1_i} \times  \boldsymbol{v}_{2_i}  \label{poly1} \label{InnerProduct}
+%\end{eqnarray}
+%where $\times$ indicates the multiplication between two polynomials. The multiplications use usual rules of polynomials multiplication. 
+%A final coefficient may be the sum of several coefficients (which we call contributions in the following passage), resulting from the sum of several multiplications.
+These coefficients are the cross terms.
+%The cross terms are not a problem for the CPU version, as a polynomial multiplication is done by a single thread, but they are the bottleneck for the GPU version.
+%The number of coefficient for a dense and triangular polynomial is given by $(n+1)^{\kappa}$ and  $(\kappa+n-1)! / ( (\kappa-1)! n! )$ respectively, where $\kappa$ is the number of variables.
+
+%In terms of  programming and parallelization, in the CPU version, the previous equations can be reduced to independent loops with a final reduction.
+%The CPU parallelization has been done with OpenMP.
+The OpenMP threads calculate several  polynomial multiplications and make a local reduction.
+At the end, a final reduction is performed  by the main thread.
+
+ % \footnote{In OpenMP reduction for class is not supported}. Results showed a perfect scalability of the OpenMP code.
 
@@
 
+\subsubsection{GPU implementation}
+
 The GPU algorithm of the inner product is divided into two parts.
-A preliminary step on the CPU executed only once,  where we prepare the execution blue print of the  contribution.
-And the second step the calculation itself on the GPU
+A preliminary step on the CPU,  where we prepare the calculation blue print of the  contribution.
+And the calculation itself on the GPU
 
@@
 
-To conclude,  We also developed an optional hybrid mode between CPU/GPU. This  technique is usual in GPU computing.
+To conclude, ........ both method can be mix ........ We also developed an optional hybrid mode between CPU/GPU. This  technique is usual in GPU computing.
 We split the input vectors of polynomials into two chunks of data, execute concurrently on CPU and GPU. 

### 2013-01-16 doc/paper_vli/vli.tex
itemize paper

@@
 
-The Very Large Integer library realizes large arithmetic , polynomials  and vector operations............ There are three 3 different modules.
+\begin{itemize}
+\item The Very Large Integer library realizes large arithmetic , polynomials  and vector operation.  There are three 3 different modules. 
+\item The large integer module models large signed integer from 128 to 512 bits.
+The operations supported are bits operations , the four basic operators $(\times,+,/,-)$ , extended multiplications and bit shift operations.
+\item The polynomial module models polynomials up to 4 variables for dense and triangular polynomials, As previously, the polynomials usual  operations between polynomials and operations with monomials.
+\item The vector of polynomial module provides a container for polynomials and allows support for optimized inners products.
+\item Modules allow flexibility, they can be interchangeable. 
+\item polynomial can have large integer coefficient or other coefficient 
+\item All modules combined, we can have access to specialized GPU solver for inner product.
+\end{itemize}
+--------------------------------------------------------
+\begin{itemize}
+\item  We goal of this library is perfumed efficient arithmetic for polynomial with large coefficient and execute efficient inner product on GPU.
+\item give an efficient library for the HP2C initiative code .e.g 
+\end{itemize}
+--------------------------------------------------------
+\begin{itemize}
 
-The large integer module models large signed integer from 128 to 512 bits.
-The operations supported are bits operations , the four basic operators $(\times,+,/,-)$ , extended multiplications and bit shift operations.
-.....SOMETHING ELSE .......
-
-The polynomial module models polynomials up to 4 variables for dense and triangular polynomials, As previously, the polynomials usual  operations between polynomials and operations with monomials.
-.....SOMETHING ELSE .......
-
-The vector of polynomial module provides a container for polynomials and allows support for optimized inners products.
-
-Modules allow flexibility, they can be interchangeable.
-........
-polynomial can have large integer coefficient or other coefficient .....
-All modules combined, we can have access to specialized GPU solver for inner product.
-
-We goal of this library is perfumed efficient arithmetic for polynomial with large coefficient and execute efficient inner product on GPU........ ANDRES SHOUD WE TALK ON VLI INT SE OR QE ....?
-
- Computer sciences provide solution for large arithmetic, and polynomial operations.
-For large arithmetic the most famous is GMP .....
-WRITE GMP is polyvalent but polyvalence obtained by dynamic memory.....
- Polynomial operations can be performed by libraries as LINBOX and NTL.
-.........
-these libraries allows manipulation but with GMP, it not performable for small coefficient ......
-The GPU is a revolution is programming model tell the LAPACK -> MAGMA .....
-...... polynomial ported only for large size with FFT approach inefficient for pb .... It does not exists optimized polynomials inner products for GPU ......
+\item Computer sciences provide solution for large arithmetic, and polynomial operations. For large arithmetic the standard is GMP
+\item  GMP is polyvalent but polyvalence obtained by dynamic memory 
+\item Polynomial operations can be performed by libraries as LINBOX and NTL
+\item these libraries allows manipulation but with GMP, it not performable for small coefficient
+\item  The GPU is a revolution is programming model tell the LAPACK -> MAGMA 
+\item polynomial ported only for large size with FFT approach inefficient for pb 
+\item It does not exists optimized polynomials inner products for GPU
+\end{itemize}
 
@@
 \subsubsection{memory layout}
-
-.... contrary to GMP  explain a static version of large arithmetic can be faster than a dynamic one....   ....   VLI number are base on 64-bit integer array .... size compile time polymorphism ...
-.... polynomials constructed over boost::array ............. the vector of polynomial is std::vector..... combined together we have a contiguous memory, and more precisely  SoA (show the figure). 
-...... It exists a second order name AoS, it does not allow a construction into module ............
+\begin{itemize}
+\item contrary to GMP  explain a static version of large arithmetic can be faster than a dynamic one
+\item   VLI number are base on 64-bit integer array, size compile time polymorphism 
+\item  polynomials constructed over boost::array, the vector of polynomial is std::vector
+\item combined together we have a contiguous memory, and more precisely  SoA (show the figure). 
+\item  It exists a second order name AoS, it does not allow a construction into module 
+\end{itemize}
 
@@
 
-......start transition ...........  Polynomials operations necessitates large arithmetic operations for the coefficients , ..... coefficient operations must be fast, we need ASM .....
+\begin{itemize}
+\item Polynomials operations necessitates large arithmetic operations for the coefficients ,
+\item  coefficient operations must be fast, we need ASM 
+\end{itemize}
 
@@
 
-We applied algorithms based on childhood  experience for large arithmetic........... for the specific range classical operations are enough ........
+We applied algorithms based on childhood  experience for large arithmetic, for the specific range classical algorithms are enough.
 The implementation necessitates a carry bit and an extended multiplication. The hardware supports
@@
 
-..... We does not use SIMD .... because the data order does not help .....
-....  simplify .......
-The disagreement  of this choice, the SoA order does not facilitate the usage of the SIMD instruction such as the AVX instructions of the Sandy Bridge x86-64 architecture.
+\begin{itemize}
+\item  We does not use SIMD 
+\item because the data order does not help
+\item 
+The disagreement  of this choice, the AoS order does not facilitate the usage of the SIMD instruction such as the AVX instructions of the Sandy Bridge x86-64 architecture.
 Applying SIMD to the AoS, for an addition or a subtraction is a nonsense because there is no specific instruction to propagate carry bit between entries of the SIMD register.
@@
 use the multimedia instruction into the ASM kernels.
-............
+\end{itemize}
 
@@
 
-..... Introduce the capacity of the lib to perform inner product ................
-...... make a short math description .........
+\begin{itemize}
+\item Introduce the capacity of the lib to perform inner product 
+\item make a short math description 
+\end{itemize}
 
@@
 \subsubsection{CPU implementation}
-........ explain quickly ..... just loop other coefficient .... with the OMP /// ......
+\begin{itemize}
+\item   explain quickly ..... just loop other coefficient 
+\item  with the OMP 
+\end{itemize}
 %In terms of  programming and parallelization, in the CPU version, the previous equations can be reduced to independent loops with a final reduction.
@@
 
+-----------------------------------------------------------------------
+
 ....... transisition explain the pb of the cross term .....
+
+-----------------------------------------------------------------------
+
 %A final coefficient may be the sum of several coefficients (which we call contributions in the following passage), resulting from the sum of several multiplications.
@@
 %The number of coefficient for a dense and triangular polynomial is given by $(n+1)^{\kappa}$ and  $(\kappa+n-1)! / ( (\kappa-1)! n! )$ respectively, where $\kappa$ is the number of variables.
-....... end pb .......
-..... GPU is interesting because we have a lot of threads available, we can calculate contribution but in a correct way to avoid pb .............
+
+\begin{itemize}
+\item  GPU is interesting because we have a lot of threads available, 
+\item we can calculate contribution but in a correct way to avoid pb 
+\item need an algo fits GPU constraints
+\end{itemize}
 
@@
 
-To conclude, ........ both method can be mix ........ We also developed an optional hybrid mode between CPU/GPU. This  technique is usual in GPU computing.
-We split the input vectors of polynomials into two chunks of data, execute concurrently on CPU and GPU. 
-A synchronization  is performed during  the memory transfer   by the final reduction, at the end of the calculation.
+\begin{itemize}
+\item both algo can be mix -> hybrid mode
+\item a few line to explain how does it work
+\end{itemize}
+%We also developed an optional hybrid mode between CPU/GPU. This  technique is usual in GPU computing.
+%We split the input vectors of polynomials into two chunks of data, execute concurrently on CPU and GPU. 
+%A synchronization  is performed during  the memory transfer   by the final reduction, at the end of the calculation.
 
-\section{Results}
+\section{benchmarks}
 

### 2013-01-16 doc/paper_vli/vli.tex
paper

@@
 
+ THE LIB DOES
 \begin{itemize}
-\item The Very Large Integer library realizes large arithmetic , polynomials  and vector operation.  There are three 3 different modules. 
+\item The Very Large Integer library realizes large arithmetic , polynomials  and vector operation.  There are three 3 different modules interchangeable. 
+\end{itemize}
+
+\begin{enumerate}
+\item  \begin{itemize}
 \item The large integer module models large signed integer from 128 to 512 bits.
 The operations supported are bits operations , the four basic operators $(\times,+,/,-)$ , extended multiplications and bit shift operations.
+\item compile time polymorphism
+\end{itemize}
+
+\item \begin{itemize}
 \item The polynomial module models polynomials up to 4 variables for dense and triangular polynomials, As previously, the polynomials usual  operations between polynomials and operations with monomials.
+\item polynomial can have large integer coefficient or other coefficient e.g. GMP
+\end{itemize}
+
+\item \begin{itemize}
 \item The vector of polynomial module provides a container for polynomials and allows support for optimized inners products.
-\item Modules allow flexibility, they can be interchangeable. 
-\item polynomial can have large integer coefficient or other coefficient 
-\item All modules combined, we can have access to specialized GPU solver for inner product.
+\item  modules combined, we can have access to specialized GPU solver for inner product.
+\end{itemize}
+\end{enumerate}
+
+--------------------------------------------------------
+
+THE OBJECTIVE
+
+\begin{itemize}
+\item  We goal of this library is performed efficient arithmetic for polynomial with large coefficient and execute efficient inner product : multithread and/or GPU.
+\item give an efficient library for the HP2C initiative code .e.g  SE
 \end{itemize}
 --------------------------------------------------------
+
+OTHER LIBRAIRIES
+
 \begin{itemize}
-\item  We goal of this library is perfumed efficient arithmetic for polynomial with large coefficient and execute efficient inner product on GPU.
-\item give an efficient library for the HP2C initiative code .e.g 
-\end{itemize}
---------------------------------------------------------
-\begin{itemize}
-
 \item Computer sciences provide solution for large arithmetic, and polynomial operations. For large arithmetic the standard is GMP
@@
 
+TRANSITION
 \begin{itemize}
@@
 \item  coefficient operations must be fast, we need ASM 
+\item this memory layout facilitates classical ASM 
 \end{itemize}

### 2013-01-16 doc/paper_vli/vli.tex
paper

@@
 \end{itemize}
-
+Description of the 3 modules : vli, poly and vector
 \begin{enumerate}
@@
 \end{itemize}
-
 \item \begin{itemize}
@@
 \item  We goal of this library is performed efficient arithmetic for polynomial with large coefficient and execute efficient inner product : multithread and/or GPU.
-\item give an efficient library for the HP2C initiative code .e.g  SE
+\item give an efficient library for the HP2C initiative code .e.g  HTSE
 \end{itemize}
@@
 \section{Implementation of the library}
-\subsubsection{memory layout}
+\subsubsection{module VLI}
 \begin{itemize}
@@
 \item   VLI number are base on 64-bit integer array, size compile time polymorphism 
-\item  polynomials constructed over boost::array, the vector of polynomial is std::vector
-\item combined together we have a contiguous memory, and more precisely  SoA (show the figure). 
-\item  It exists a second order name AoS, it does not allow a construction into module 
+\item VLI module performs arithmetic operations, it needs performable kernels
+\item ASM Kernel
+
+%\item  polynomials constructed over boost::array, the vector of polynomial is std::vector
+%\item combined together we have a contiguous memory, and more precisely  SoA (show the figure). 
+%\item  It exists a second order name AoS, it does not allow a construction into module
+
 \end{itemize}
 
-\begin{figure}
-    \begin{minipage}{0.50\linewidth}
-        \begin{eqnarray}
-        \tiny{
-                 \underbrace{
-                 \underbrace{ \underbrace{a^0_{00} | a^1_{00} | a^2_{00}}_{VLI, a_{00} } || \underbrace{a^0_{10} | a^1_{10} | a^2_{10}}_{VLI, a_{10}} ||   \dots  }_{1^{st} \textrm{polynomial}} 
-                  |||    \underbrace{ \underbrace{b^0_{00} | a^1_{00} | b^2_{00}}_{VLI, b_{00} } || \underbrace{b^0_{10} | b^1_{10} | b^2_{10}}_{VLI, b_{10}} ||   \dots  }_{2^{nd} \textrm{polynomial} } }_{\textrm{Vector of polynomials, AoS order}} \nonumber
-            }
-        \end{eqnarray}
-    \end{minipage}
-\begin{minipage}{0.50\linewidth}
-    \begin{eqnarray}
-    \tiny{
-        \underbrace{
-        \underbrace{a^0_{00} | a^0_{10} | \dots || a^1_{00} | a^1_{10} | \dots  }_{1st  \textrm{ nested VLI/Polynomial} } |||  \underbrace{b^0_{00} | b^0_{10} | \dots || b^1_{00} | b^1_{10} | \dots  }_{2nd  \textrm{ nested VLI/Polynomial} }
-        }_{\textrm{Vector of polynomials, SoA Order}} \nonumber }
-        \end{eqnarray}
-\end{minipage}
-\caption{Representation of the AoS Order (left) or SoA Order (right) for a vector of polynomial \label{AOSSOA}.  The SoA structure limits the usage of the VLI module alone.}
-\end{figure}
-
-TRANSITION
-\begin{itemize}
-\item Polynomials operations necessitates large arithmetic operations for the coefficients ,
-\item  coefficient operations must be fast, we need ASM 
-\item this memory layout facilitates classical ASM 
-\end{itemize}
-
-\subsubsection{Kernels and assembly  language}
-
+----------------------------------------------- \\
+COPY/PAST from the LAST ASM\\
+-----------------------------------------------\\
 We applied algorithms based on childhood  experience for large arithmetic, for the specific range classical algorithms are enough.
@@
 \item  We does not use SIMD 
-\item because the data order does not help
+\item because the contiguous data does not help at all for CB propagation
 \item 
-The disagreement  of this choice, the AoS order does not facilitate the usage of the SIMD instruction such as the AVX instructions of the Sandy Bridge x86-64 architecture.
-Applying SIMD to the AoS, for an addition or a subtraction is a nonsense because there is no specific instruction to propagate carry bit between entries of the SIMD register.
-It could be  apply  on multiplications for integer  larger than 1024-bit \cite{SIMD}, unfortunately, it does not correspond to our working range. Consequently, we do not 
-use the multimedia instruction into the ASM kernels.
+%The disagreement  of this choice, the AoS order does not facilitate the usage of the SIMD instruction such as the AVX instructions of the Sandy Bridge x86-64 architecture.
+%Applying SIMD to the AoS, for an addition or a subtraction is a nonsense because there is no specific instruction to propagate carry bit between entries of the SIMD register.
+%It could be  apply  on multiplications for integer  larger than 1024-bit \cite{SIMD}, unfortunately, it does not correspond to our working range. Consequently, we do not 
+%use the multimedia instruction into the ASM kernels.
+\end{itemize}
+
+\subsubsection{module polynomial}
+
+short description of both type of polynomial
+
+\begin{itemize}
+\item construct on boost array
+\item compile time polymorphism for dense, triangular
+\item works whatever the coefficient
+\item description of the implementation of the basic operations, the perf depends of coefficients implementation
+\end{itemize}
+
+\subsubsection{module vector}
+\begin{itemize}
+\item +- std::vector
+\end{itemize}
+
+\subsubsection{memory layout}
+
+All elements modules combined together give the memory layout
+
+\begin{figure}
+   \begin{minipage}{0.50\linewidth}
+        \begin{eqnarray}
+        \tiny{
+                \underbrace{
+                 \underbrace{ \underbrace{a^0_{00} | a^1_{00} | a^2_{00}}_{VLI, a_{00} } || \underbrace{a^0_{10} | a^1_{10} | a^2_{10}}_{VLI, a_{10}} ||   \dots  }_{1^{st} \textrm{polynomial}} 
+                  |||    \underbrace{ \underbrace{b^0_{00} | a^1_{00} | b^2_{00}}_{VLI, b_{00} } || \underbrace{b^0_{10} | b^1_{10} | b^2_{10}}_{VLI, b_{10}} ||   \dots  }_{2^{nd} \textrm{polynomial} } }_{\textrm{Vector of polynomials, AoS order}} \nonumber
+            }
+        \end{eqnarray}
+    \end{minipage}
+\begin{minipage}{0.50\linewidth}
+   \begin{eqnarray}
+    \tiny{
+        \underbrace{
+        \underbrace{a^0_{00} | a^0_{10} | \dots || a^1_{00} | a^1_{10} | \dots  }_{1st  \textrm{ nested VLI/Polynomial} } |||  \underbrace{b^0_{00} | b^0_{10} | \dots || b^1_{00} | b^1_{10} | \dots  }_{2nd  \textrm{ nested VLI/Polynomial} }
+        }_{\textrm{Vector of polynomials, SoA Order}} \nonumber }
+       \end{eqnarray}
+\end{minipage}
+\caption{Representation of the AoS Order (left) or SoA Order (right) for a vector of polynomial \label{AOSSOA}.  The SoA structure limits the usage of the VLI module alone.}
+\end{figure}
+
+\begin{itemize}
+\item object conception imposes SoA
+\item AoS is the good for SIMD ( global solution)
+\item  with the flexibility of module we lost the SIMD 
 \end{itemize}
@@
 \begin{itemize}
-\item Introduce the capacity of the lib to perform inner product 
+\item Introduce the capacity of the lib to perform inner product  as external solver
 \item make a short math description 

### 2013-01-16 doc/paper_vli/vli.tex
paper

@@
 
---------------------------------------------------------
-
 THE OBJECTIVE
@@
 \end{itemize}
---------------------------------------------------------
 
@@
 \begin{itemize}
-\item  We does not use SIMD 
+\item We does not use SIMD for kernels 
 \item because the contiguous data does not help at all for CB propagation 
-\item 
+\item It could be  apply  on multiplications for integer  larger than 1024-bit \cite{SIMD}, unfortunately, it does not correspond to our working range.
+
+%Consequently, we do not  use the multimedia instruction into the ASM kernels.
+
 %The disagreement  of this choice, the AoS order does not facilitate the usage of the SIMD instruction such as the AVX instructions of the Sandy Bridge x86-64 architecture.
@@
 \begin{itemize}
-\item +- std::vector
+\item derived from  std::vector
 \end{itemize}
@@
 \item object conception imposes SoA
-\item AoS is the good for SIMD ( global solution)
-\item  with the flexibility of module we lost the SIMD 
+\item AoS is the good for SIMD ( global solution) but for this solution modules disappear, we loose all the flexibility
+\item  Do SoA to AoS reorganization of the data costs too much, Maybe intel Xeon phi could be a solution in the future, as gpu we could develop a booster
 \end{itemize}
@@
 \item   explain quickly ..... just loop other coefficient 
-\item  with the OMP 
+\item  explain the implementation of threadings basic OMP
 \end{itemize}
@@
 
+\subsubsection{hybri-mode}
+
 \begin{itemize}
-\item both algo can be mix -> hybrid mode
+\item both algo can be mix  $=>$ hybrid mode
 \item a few line to explain how does it work

### 2013-01-22 doc/paper_vli/vli2.tex
rough structure sketch of the new version of the vli paper

\documentclass[oribibl]{llncs2e/llncs}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
\usepackage{graphicx}

\author{Timoth\'ee Ewart\inst{1}\thanks{Acknowledgments: Maxim Milakov, Peter Messmer:   NVIDIA,   Williams Sawyer, Gilles Fourestey:  CSCS, HP2C funding, hp2c.ch}  , Andreas Hehn$^2$, Matthias Troyer\inst{2} and Thierry Giamarchi\inst{1}}

\institute{Universit\'e de Gen\`eve, \email{timothee.ewart@gmail.com}  \and Eidgen\"ossische Technische Hochschule Z\"urich }

\title{VLI - a library for large integer and polynomial arithemtic}
\author{Timoth\'ee Ewart\inst{1}\thanks{Acknowledgments: Maxim Milakov, Peter Messmer:   NVIDIA,   Williams Sawyer, Gilles Fourestey:  CSCS, HP2C funding, hp2c.ch}  , Andreas Hehn$^2$, Matthias Troyer\inst{2} and Thierry Giamarchi\inst{1}}

\institute{Universit\'e de Gen\`eve, \email{timothee.ewart@gmail.com}  \and Eidgen\"ossische Technische Hochschule Z\"urich }

\begin{document}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
We present a high-performance C++ library for 128 to 512 bit integer arithemtics and symbolic polynomial computations.
While both parts of the library can be used independently optimized kernels for symbolic polynomials with large integer coefficients are provided.
%Even though the application developed as a special purpose library for high-temperature series expansions, it is also of great interest for other applications due to its modular design.
%The kernels for the large integer operations are implemented in \verb|x86-64| and \verb|power64| assembly.
%The polynomials may have up to 4 symbolic variables and have a compile time fixed truncation order.
Since our target application is high-temperature series expansions (HTSE) we implemented a special routine to perform inner products of vectors of polynomials with large integer coefficients.
Here we employ hybrid CPU/GPU approach to make optimal use of todays supercomputers, where we attain 67\% of the peak performance of the current NVIDIA Kepler GPU.
Compared to a solution using the GNU Multiple Precision Arithmetic Library (GMP) we save a factor up to 10 on the CPU and
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
\begin{itemize}
\item {\bf Why did we develop the library?}
\item Applications of large integers and polynomials in many fields of science, e.g.\ cryptography
\item our intention: special purpose library for high-temperature series expansions.
\item HTSE require symbolic polynomials with arbitrary precise coefficients. Coefficients can be represented as integers.
\item maximal integer size known beforehand
\item Hot spot of HTSE inner products of vectors of such polynomials.
\item {\bf Problems with existing libraries:}
\item GMP integers are dynamic in size and optimized to cover a large range of large integers up to several thousand bits. We need only up to 512 bits.
\item inner products of polynomials are well suited for GPUs $\rightarrow$ specialized GPU kernels required.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Large integers}
\begin{itemize}
\item {\bf Description}
\item signed integer arithmetics
\item fixed size (128-512bit)
\item all standard integer operations (arithmetic, comparison and bit operations)
\item in addition to that: extended multiplication which doubles the number of bits, extended add which increases the size by one 64bit.
\item {\bf Implementation details}
\item data is stored as 64 bit integer arrays.
\item stack based $\rightarrow$ cheap memory allocation
\item optimized ASM kernels to use hardware features not accessible from C++ (\verb|x86_64| carry addition \verb|adcq|, 64bit multiplication \verb|mulq| which calculates low and hi 64bit part of the result.)
\item ASM code generated using \verb|BOOST_PP| to compactify the code and minimize bugs.
\end{itemize}

\section{Polynomials}
\begin{itemize}
\item {\bf Description}
\item dense symbolic polynomial in 1-4 variables.
\begin{equation}
    p(x,y,z,a) = \sum_{i,j,k,l} c_{ijkl} x^i y^j z^k a^l
\end{equation}
\item either ``dense'' (${i,j,k,l} \le N$) or ``triangular'' ($i+j+k+l \le N$) structure.
\item any arithmetic type as coefficient $c_{ijkl}$
\item addition, subtraction, multiplication with other polynomials and monomials
\item {\bf Implementation details}
\item structure and truncation order $N$ is compile time fixed 
\item The symbolic variable names are given as template parameter
\item Automatic matching of variable names of different polynomial types using template meta programming.
\end{itemize}

\section{Optimized inner product}
\paragraph{CPU}
\begin{itemize}
\item The hot spot of HTSE are inner products of large vectors with symbolic polynomials as components. The coefficients of the polynomials are large integers.
\item We continue to use the single large integer routines
\item No SIMD, since the number of operations for a multiplication increases by a factor TODO due to the manual handling of the carry and separate lo/hi word multiplication.
\item OpenMP parallelization
\end{itemize}
\paragraph{GPU}
\begin{itemize}
\item Many small tasks of the same form. $\rightarrow$ well suited for GPUs.
\item {\bf Implementation details}
\item Each resulting coefficient of the polynomial $c_{I,J,K,L}$ may has several e
\end{itemize}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Benchmarks}
\begin{itemize}
\item Benchmarks performed
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusion}
\begin{itemize}
\item TODO
\item Licence?
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2013-01-22 doc/paper_vli/vli2.tex
abstract draft

@@
 \begin{abstract}
-We present a high-performance C++ library for 128 to 512 bit integer arithemtics and symbolic polynomial computations.
+We present a high-performance C++ library for 128 to 512 bit integer arithmetic and symbolic polynomial computations.
 While both parts of the library can be used independently optimized kernels for symbolic polynomials with large integer coefficients are provided.
@@
 %The polynomials may have up to 4 symbolic variables and have a compile time fixed truncation order.
-Since our target application is high-temperature series expansions (HTSE) we implemented a special routine to perform inner products of vectors of polynomials with large integer coefficients.
-Here we employ hybrid CPU/GPU approach to make optimal use of todays supercomputers, where we attain 67\% of the peak performance of the current NVIDIA Kepler GPU.
-Compared to a solution using the GNU Multiple Precision Arithmetic Library (GMP) we save a factor up to 10 on the CPU and
+%Since our main target applications is high-temperature series expansions (HTSE) we implemented a special routine to perform inner products of vectors of polynomials with large integer coefficients.
+Our main target application is high-temperature series expansions (HTSE) which require inner products of large vectors of polynomials with large integer coefficients.
+For this purpose we implemented a tunable hybrid CPU/GPU inner product function using OpenMP and NVIDIA CUDA.
+This way we make optimal use of todays and upcoming hybrid supercomputers and attain 67\% of the peak performance of the current NVIDIA Kepler GPU.
+Compared to a pure CPU solution using the GNU Multiple Precision Arithmetic Library (GMP) we gain a speedup of 10x for a pure CPU inner product and 60x using a GPU accelerator.
 \end{abstract}
@@
 \item {\bf Description}
-\item signed integer arithmetics
+\item signed integer arithmetic
 \item fixed size (128-512bit)
@@
 \item TODO
-\item Licence?
+\item License?
 \end{itemize}

### 2013-01-22 doc/paper_vli/vli2.tex
minor update

@@
 While both parts of the library can be used independently optimized kernels for symbolic polynomials with large integer coefficients are provided.
-%Even though the application developed as a special purpose library for high-temperature series expansions, it is also of great interest for other applications due to its modular design.
-%The kernels for the large integer operations are implemented in \verb|x86-64| and \verb|power64| assembly.
-%The polynomials may have up to 4 symbolic variables and have a compile time fixed truncation order.
-%Since our main target applications is high-temperature series expansions (HTSE) we implemented a special routine to perform inner products of vectors of polynomials with large integer coefficients.
 Our main target application is high-temperature series expansions (HTSE) which require inner products of large vectors of polynomials with large integer coefficients.
@@
 \item all standard integer operations (arithmetic, comparison and bit operations)
-\item in addition to that: extended multiplication which doubles the number of bits, extended add which increases the size by one 64bit.
+\item in addition to that: fused multiply-add, extended multiplication which doubles the number of bits, extended add which increases the size by one 64bit.
 \item {\bf Implementation details}
@@
 \item {\bf Description}
-\item dense symbolic polynomial in 1-4 variables.
+\item symbolic polynomial in 1-4 variables.
 \begin{equation}
@@
 \item {\bf Implementation details}
-\item Each resulting coefficient of the polynomial $c_{I,J,K,L}$ may has several e
+\item Each resulting coefficient of the polynomial $c_{IJKL}$ may has several e
 \end{itemize}

### 2013-01-24 doc/paper_vli/vli2.tex
work on the vli paper

@@
 \usepackage{graphicx}
-
-\author{Timoth\'ee Ewart\inst{1}\thanks{Acknowledgments: Maxim Milakov, Peter Messmer:   NVIDIA,   Williams Sawyer, Gilles Fourestey:  CSCS, HP2C funding, hp2c.ch}  , Andreas Hehn$^2$, Matthias Troyer\inst{2} and Thierry Giamarchi\inst{1}}
-
-\institute{Universit\'e de Gen\`eve, \email{timothee.ewart@gmail.com}  \and Eidgen\"ossische Technische Hochschule Z\"urich }
+\usepackage{color}
+\graphicspath{{WarpTask/figures/}{Res/figures/}} %do not forget the / at the end
 
@@
 \begin{abstract}
-We present a high-performance C++ library for 128 to 512 bit integer arithmetic and symbolic polynomial computations.
-While both parts of the library can be used independently optimized kernels for symbolic polynomials with large integer coefficients are provided.
-Our main target application is high-temperature series expansions (HTSE) which require inner products of large vectors of polynomials with large integer coefficients.
-For this purpose we implemented a tunable hybrid CPU/GPU inner product function using OpenMP and NVIDIA CUDA.
-This way we make optimal use of todays and upcoming hybrid supercomputers and attain 67\% of the peak performance of the current NVIDIA Kepler GPU.
-Compared to a pure CPU solution using the GNU Multiple Precision Arithmetic Library (GMP) we gain a speedup of 10x for a pure CPU inner product and 60x using a GPU accelerator.
+We present a high-performance C++ library for high but fixed precision
+(128 to 512 bit) integer arithmetic and symbolic polynomial
+computations. While the large integer and polynomial computation parts
+of the library can be used independently optimized kernels for symbolic
+polynomials with large integer coefficients are provided. The kernels
+were manually optimized in assembly language for the x86-64 and power64
+architectures. Our main target application is high-temperature series
+expansions which require inner products of large vectors of polynomials
+with large integer coefficients. For this purpose we implemented a
+tunable hybrid CPU/GPU inner product function using OpenMP and NVIDIA
+CUDA with inline PTX assembly. This way we make optimal use of today's
+and upcoming hybrid supercomputers and attain 67\% of the peak
+performance of the current NVIDIA Kepler GPU. Compared to a pure CPU
+solution using the GNU Multiple Precision Arithmetic Library (GMP) we
+gain a speedup of 10x for a pure CPU inner product and 60x using a GPU
+accelerator.
 \end{abstract}
@@
 \item all standard integer operations (arithmetic, comparison and bit operations)
-\item in addition to that: fused multiply-add, extended multiplication which doubles the number of bits, extended add which increases the size by one 64bit.
+\item in addition to that: fused multiply-add, extended multiplication which doubles the number of bits, extended add which increases the size by one 64bit word.
 \item {\bf Implementation details}
 \item data is stored as 64 bit integer arrays.
+\item two-complement representation
 \item stack based $\rightarrow$ cheap memory allocation
+\item addition and multiplication using the standard schoolbook algorithms.
+\item Numbers are two small for Toom\cite{Toom} or FFT based approaches like the Sch\"onhage-Strassen algorithm\cite{Schonhage}.
 \item optimized ASM kernels to use hardware features not accessible from C++ (\verb|x86_64| carry addition \verb|adcq|, 64bit multiplication \verb|mulq| which calculates low and hi 64bit part of the result.)
-\item ASM code generated using \verb|BOOST_PP| to compactify the code and minimize bugs.
+\item TODO which special features do we use on power64?
+\item assembly code generated using \verb|BOOST_PP| to compactify the code and minimize bugs.
 \end{itemize}
@@
 \item structure and truncation order $N$ is compile time fixed 
+\item function hooks to allow custom optimized kernels for certain coefficient types
 \item The symbolic variable names are given as template parameter
 \item Automatic matching of variable names of different polynomial types using template meta programming.
+\end{itemize}
+
+\section{Overview of the GPU architecture}
+\begin{itemize}
+\item A GPU has multiple streaming multiprocessors (SM)
+\item They execute instructions in a SIMT (single-instruction, multiple-thread).
+\item A group of 32 threads (called ``warp'') execute the instruction stream in a lockstep fashion on multiple data, if the threads follow different execution paths all branches need to be executed serially.
+\item SM switches between up to 64 warps to hide latencies. $\rightarrow$ To gain good performance thousands of threads are necessary.
+\item NVIDIA CUDA resembles this hardware properties by threads that are organized in so called blocks, which consist of a few hundred threads.
+A thread block is always scheduled on a single SM, but it is possible to schedule multiple thread blocks on a single SM.
+\item Memory hierarchy: large (several GB) slow global memory accessible from all threads; small (a few KB) very fast shared memory which is shared among all threads of a thread block; The registers of the SM are split among all threads hosted by the SM.
+\item Special purpose memories: Texture memory: globally shared, read-only, cache optimized for 2D access
 \end{itemize}
@@
 \item The hot spot of HTSE are inner products of large vectors with symbolic polynomials as components. The coefficients of the polynomials are large integers.
-\item We continue to use the single large integer routines
-\item No SIMD, since the number of operations for a multiplication increases by a factor TODO due to the manual handling of the carry and separate lo/hi word multiplication.
-\item OpenMP parallelization
+\item We continue to use the single large integer kernels
+\item OpenMP parallelization: split vector into chunks, each thread does part of the inner product
+\item No SIMD, since we had to use 32bit words and the number of operations for a multiplication increases by a factor TODO due to the manual handling of the carry.
 \end{itemize}
@@
 \item {\bf Implementation details}
-\item Each resulting coefficient of the polynomial $c_{IJKL}$ may has several e
+\item Several coefficients $a_{ijkl}$, $b_{i'j'k'l'}$ of the polynomials to be multiplied contribute to the same coefficient $c_{IJKL}$ of the resulting polynomial
+\begin{equation}
+    c_{IJKL} = \sum_{i,j,k,l} a_{ijkl} \cdot b_{I-i,J-j,K-k,L-l}
+\end{equation}
+\item The number of terms in this sum depends on the orders $I,J,K,L$ of the resulting coefficient $\rightarrow$ load inbalance
+\item We group 32 coefficients of the result into a task and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}). {\color{red} TODO: Tim please elaborate}
+\begin{figure}[t]
+    \begin{center}
+    \mbox{
+        \includegraphics[scale=0.37, angle=-90]{coeffs.eps} 
+        \includegraphics[scale=0.37, angle=-90]{warp.eps} 
+    }
+    \caption{Left: Number of contributions to each coefficient of the resulting polynomial (red line) sorted in reverse order and split into tasks. Right: Load balancing of the tasks over the warps.}
+    \label{fig:GPU_load_balance}
+    \end{center}
+\end{figure}
+\item Coalesced memory access?
+\item Using NVIDIA parallel thread execution (PTX) assembly language to benefit from hardware carry bit propagation \verb|addc.cc| and fused multiply add with carry bit propagation \verb|madc.cc|.
 \end{itemize}
@@
 \section{Benchmarks}
+\paragraph{CPU}
 \begin{itemize}
-\item Benchmarks performed
+\item pure CPU benchmarks were done on a Sandy bridge Intel Xeon E5-2670 with 32 logical cores.
+\end{itemize}
+\paragraph{GPU, hybrid}
+\begin{itemize}
+\item GPU benchmarks were performed on Todi, a Cray XK7 with NVIDIA Tesla K20X GPUs and AMD Opteron CPUs, at the Swiss Center for Scientific Computing (CSCS)
 \end{itemize}
@@
 \begin{itemize}
-\item TODO
+\item We 
 \item License?
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\bibliographystyle{llncs2e/splncs}
+\bibliography{biblio/vlibib}
 

### 2013-01-24 doc/paper_vli/vli2.tex
wip vli paper

@@
 
-\section{Overview of the GPU architecture}
-\begin{itemize}
-\item A GPU has multiple streaming multiprocessors (SM)
-\item They execute instructions in a SIMT (single-instruction, multiple-thread).
-\item A group of 32 threads (called ``warp'') execute the instruction stream in a lockstep fashion on multiple data, if the threads follow different execution paths all branches need to be executed serially.
-\item SM switches between up to 64 warps to hide latencies. $\rightarrow$ To gain good performance thousands of threads are necessary.
-\item NVIDIA CUDA resembles this hardware properties by threads that are organized in so called blocks, which consist of a few hundred threads.
-A thread block is always scheduled on a single SM, but it is possible to schedule multiple thread blocks on a single SM.
-\item Memory hierarchy: large (several GB) slow global memory accessible from all threads; small (a few KB) very fast shared memory which is shared among all threads of a thread block; The registers of the SM are split among all threads hosted by the SM.
-\item Special purpose memories: Texture memory: globally shared, read-only, cache optimized for 2D access
-\end{itemize}
-
 \section{Optimized inner product}
@@
 \item {\bf Implementation details}
-\item Several coefficients $a_{ijkl}$, $b_{i'j'k'l'}$ of the polynomials to be multiplied contribute to the same coefficient $c_{IJKL}$ of the resulting polynomial
+\item Element-wise polynomial multiplication + reduction at the end.
+\item Each thread calculates one coefficient $c_{IJKL}$ the product polynomial.
+\item Several coefficients $a_{ijkl}$, $b_{i'j'k'l'}$ of the polynomials to be multiplied contribute to the same coefficient $c_{IJKL}$
 \begin{equation}
@@
 \item The number of terms in this sum depends on the orders $I,J,K,L$ of the resulting coefficient $\rightarrow$ load inbalance
-\item We group 32 coefficients of the result into a task and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}). {\color{red} TODO: Tim please elaborate}
+\item We group 32 coefficients of the result into a task and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}).
 \begin{figure}[t]
@@
 \end{figure}
-\item Coalesced memory access?
+\item {\color{red} TODO: Tim please elaborate}
+%\item Coalesced memory access?
 \item Using NVIDIA parallel thread execution (PTX) assembly language to benefit from hardware carry bit propagation \verb|addc.cc| and fused multiply add with carry bit propagation \verb|madc.cc|.

### 2013-01-25 doc/paper_vli/vli2.tex
paper:

@@
 \item The number of terms in this sum depends on the orders $I,J,K,L$ of the resulting coefficient $\rightarrow$ load inbalance
-\item We group 32 coefficients of the result into a task and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}).
+%\item {\color{blue} We design an execution plan.
+We group the coefficients of a fictive result polynomial by packet of 32 } into a task and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}).
+\item We group the coefficients by 32 into a task and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}).
+{\color{blue} Preparation of the task is done on CPU}
+
 \begin{figure}[t]
@@
 \end{figure}
-\item {\color{red} TODO: Tim please elaborate}
+\item {\color{blue} During the execution of a task by a warp,  each threads into the warp get the indices of the fictive coefficient, determine the necessary  contributions and  then calculate the output coefficient}
+\item {\color{blue} Reduction is performed under NVIDA standard}
+\item {\color{blue} This algorithm is efficient whatever the type of polynomial. But the termination of the contributions of an output coefficient can be tricky}
 %\item Coalesced memory access?
+\item {\color{blue}  Karatsuba algorithm \cite{Karatsuba1963} for the largest extended multiplication (256 bit to 512 bit) on GPU}
 \item Using NVIDIA parallel thread execution (PTX) assembly language to benefit from hardware carry bit propagation \verb|addc.cc| and fused multiply add with carry bit propagation \verb|madc.cc|.

### 2013-01-25 doc/paper_vli/vli2.tex
vli paper wip

@@
 \item {\bf Problems with existing libraries:}
-\item GMP integers are dynamic in size and optimized to cover a large range of large integers up to several thousand bits. We need only up to 512 bits.
+\item GMP integers \cite{GMP} are dynamic in size and optimized to cover a large range of large integers up to several thousand bits. We need only up to 512 bits.
 \item inner products of polynomials are well suited for GPUs $\rightarrow$ specialized GPU kernels required.
@@
 \item addition, subtraction, multiplication with other polynomials and monomials
+\item special multiplication which will double the truncation order
 \item {\bf Implementation details}
@@
 \section{Optimized inner product}
+The hot spot of HTSE are inner products of large vectors with symbolic polynomials as components. The coefficients of the polynomials are large integers.
+All polynomials are assumed to have the same truncation order.
+The inner product will result in a polynomial of twice the order of the original polynomials
+with coefficients of twice the width of the original coefficients, i.e.\ 256 and
+512 bit, respectively.
 \paragraph{CPU}
 \begin{itemize}
-\item The hot spot of HTSE are inner products of large vectors with symbolic polynomials as components. The coefficients of the polynomials are large integers.
 \item We continue to use the single large integer kernels
-\item OpenMP parallelization: split vector into chunks, each thread does part of the inner product
+\item OpenMP parallelization: split vectors into chunks, each thread does part of the inner product. Very regular problem, good load balance.
 \item No SIMD, since we had to use 32bit words and the number of operations for a multiplication increases by a factor TODO due to the manual handling of the carry.
@@
 We group the coefficients of a fictive result polynomial by packet of 32 } into a task and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}).
-\item We group the coefficients by 32 into a task and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}).
+\item We group the coefficients by 32 into a task and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}), i.e.\ each thread calculates one coefficient.
+Since the differences within a warp are rather small all threads of the warp will finish
+
 {\color{blue} Preparation of the task is done on CPU}
@@
 \item {\color{blue}  Karatsuba algorithm \cite{Karatsuba1963} for the largest extended multiplication (256 bit to 512 bit) on GPU}
-\item Using NVIDIA parallel thread execution (PTX) assembly language to benefit from hardware carry bit propagation \verb|addc.cc| and fused multiply add with carry bit propagation \verb|madc.cc|.
+\item Using NVIDIA parallel thread execution (PTX) assembly language to benefit from hardware carry bit propagation \verb|addc.cc| and fused multiply-add with carry bit propagation \verb|madc.cc|.
 \end{itemize}
@@
 \item pure CPU benchmarks were done on a Sandy bridge Intel Xeon E5-2670 with 32 logical cores.
-\end{itemize}
-\paragraph{GPU, hybrid}
-\begin{itemize}
 \item GPU benchmarks were performed on Todi, a Cray XK7 with NVIDIA Tesla K20X GPUs and AMD Opteron CPUs, at the Swiss Center for Scientific Computing (CSCS)
+\item Testcase: inner product of two vectors dimension 4096, with ``dense'' and
+``triangular'' polynomials of order 1 to 14 with 128 and 256 bit integer
+coefficients. Which will result in polynomials of order 2 to 28 with 256 and 512 bit integer coefficiencts, respectively.
+\item We estimate the performance in integer operations per second (IOP/s).
+It
+\item For the GPU performance we only consider the polynomial polynomial multiplications and neglect the reduction at the end of the inner product.
+
+\begin{figure}[t!]
+    \begin{center}
+    \mbox{
+        \hspace{-0.5cm}
+        \includegraphics[scale=0.37, angle=-90]{ME128MIPS.eps} 
+        \hspace{-0.4cm}
+        \includegraphics[scale=0.37, angle=-90]{MC128MIPS.eps} 
+        %\hspace{-0.4cm}
+        %\includegraphics[scale=0.25, angle=-90]{ME256.eps} 
+    }
+    \caption{Performance of the inner product of two vectors of dimension 4096 with ``dense'' (left) and ``triangular'' (right) polynomials having 128 bit coefficients on a NVIDIA Tesla K20X (GK110).}
+    \label{ResMEMIPS}
+    \end{center}
+\end{figure}  
+\item We reach up to $200$ GIOP/s using 32 bit multiply-add (\verb|madc|)
+\item Theoretical maximum for the NVIDIA Tesla K20X are $14 \, \mathrm{SMs} \cdot 732 \, \mathrm{MHz} \cdot 32 \, \mathrm{ IOP / cycle / SM} = 327 \, \mathrm{GIOP/s}$.
+(Multiply-add throughput taken from \cite{CUDAProgramming}).
+\item We reach 61\% of the peak.
+
 \end{itemize}

### 2013-01-25 doc/paper_vli/vli2.tex
vli paper wip

@@
 \usepackage{graphicx}
+\usepackage{tabularx}
+%\usepackage{multicolumn}
 \usepackage{color}
@@
 CUDA with inline PTX assembly. This way we make optimal use of today's
-and upcoming hybrid supercomputers and attain 67\% of the peak
+and upcoming hybrid supercomputers and attain $65\%$ of the peak
 performance of the current NVIDIA Kepler GPU. Compared to a pure CPU
@@
 \item all standard integer operations (arithmetic, comparison and bit operations)
-\item in addition to that: fused multiply-add, extended multiplication which doubles the number of bits, extended add which increases the size by one 64bit word.
+\item in addition to that: multiply-add, extended multiplication which doubles the number of bits, extended add which increases the size by one 64bit word.
 \item {\bf Implementation details}
@@
 We group the coefficients of a fictive result polynomial by packet of 32 } into a task and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}).
-\item We group the coefficients by 32 into a task and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}), i.e.\ each thread calculates one coefficient.
-Since the differences within a warp are rather small all threads of the warp will finish
-
+\item Since threads within a warp are executed in a lockstep manner, we try to give all threads within one warp the same amount of work.
+Such that a thread has to wait as little as possible for the other threads of the warp to finish.
+\item We group the coefficients by 32 into a task and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}), i.e.\ each thread calculates one coefficient of the resulting polynomial.
+\item Since the work load differences within a warp are small, the 
 {\color{blue} Preparation of the task is done on CPU}
@@
 \item {\color{blue}  Karatsuba algorithm \cite{Karatsuba1963} for the largest extended multiplication (256 bit to 512 bit) on GPU}
-\item Using NVIDIA parallel thread execution (PTX) assembly language to benefit from hardware carry bit propagation \verb|addc.cc| and fused multiply-add with carry bit propagation \verb|madc.cc|.
+\item Using NVIDIA parallel thread execution (PTX) assembly language to benefit from hardware carry bit propagation \verb|addc.cc| and multiply-add with carry bit propagation \verb|madc.cc|.
 \end{itemize}
@@
 \item pure CPU benchmarks were done on a Sandy bridge Intel Xeon E5-2670 with 32 logical cores.
+\begin{table}
+\begin{center}
+\begin{tabular}{l|cc|c}
+ Operation & Execution & time [s] & Speed-up\\
+   & \, GMP\,  & \, VLI\,  & \\
+ \hline
+\verb|add|, 128 bit & 1.11 & 0.33 & \\
+\verb|add|, 192 bit & 1.25 & 0.45 & \\
+\verb|add|, 256 bit & 1.11 & 0.81 & \\
+\verb|add|, 384 bit & & & \\
+\verb|add|, 448 bit & & & \\
+\verb|add|, 512 bit & & & \\
+\hline
+\verb|mul|, 128 bit $\rightarrow$ 256 bit & 1.36 & 0.93 & 1.46x \\
+\verb|mul|, 192 bit $\rightarrow$ 384 bit & 3.21 & 1.18 & 2.72x \\
+\verb|mul|, 256 bit $\rightarrow$ 512 bit & 3.74 & 2.04 & 1.83x \\ 
+\end{tabular}
+\caption{Exection time for $10^8$ operations.}
+\end{center}
+\end{table}
 \item GPU benchmarks were performed on Todi, a Cray XK7 with NVIDIA Tesla K20X GPUs and AMD Opteron CPUs, at the Swiss Center for Scientific Computing (CSCS)
@@
 It
-\item For the GPU performance we only consider the polynomial polynomial multiplications and neglect the reduction at the end of the inner product.
+\item For the GPU performance estimate we only consider the instructions for the polynomial-polynomial multiplications and neglect the reduction at the end of the inner product.
 
@@
 \end{figure}  
-\item We reach up to $200$ GIOP/s using 32 bit multiply-add (\verb|madc|)
-\item Theoretical maximum for the NVIDIA Tesla K20X are $14 \, \mathrm{SMs} \cdot 732 \, \mathrm{MHz} \cdot 32 \, \mathrm{ IOP / cycle / SM} = 327 \, \mathrm{GIOP/s}$.
+\item We reach up to $214$ GIOP/s using 32 bit multiply-add (\verb|madc|)
+\item Theoretical maximum for the NVIDIA Tesla K20X are Number of Multiprocessors (SMX) $\cdot$ Clock frequency $\cdot$ Instruction throughput of \verb|madc|  =  $14 \, \mathrm{SMX} \cdot 732 \, \mathrm{MHz} \cdot 32 \, \mathrm{ IOP / cycle / SMX} = 327 \, \mathrm{GIOP/s}$.
 (Multiply-add throughput taken from \cite{CUDAProgramming}).
-\item We reach 61\% of the peak.
+\item We reach $65\%$ of the peak.
 

### 2013-01-26 doc/paper_vli/vli2.tex
table paper
vli paper wip

@@
 \item Since the work load differences within a warp are small, the 
-{\color{blue} Preparation of the task is done on CPU}
+%{\color{blue} Preparation of the task is done on CPU}
 
@@
 \section{Benchmarks}
-\paragraph{CPU}
+\paragraph{Simple integer operations}
 \begin{itemize}
@@
      Operation & Execution & time [s] & Speed-up\\
-   & \, GMP\,  & \, VLI\,  & \\
+       & GMP & VLI & \\
      \hline
-\verb|add|, 128 bit & 1.11 & 0.33 & \\
-\verb|add|, 192 bit & 1.25 & 0.45 & \\
-\verb|add|, 256 bit & 1.11 & 0.81 & \\
-\verb|add|, 384 bit & & & \\
-\verb|add|, 448 bit & & & \\
-\verb|add|, 512 bit & & & \\
+    \verb|add|, 128 bit & 1.11 & 0.33 & 3.36x \\
+    \verb|add|, 192 bit & 1.21 & 0.45 & 2.68x \\
+    \verb|add|, 256 bit & 1.11 & 0.81 & 1.37x \\
+    \verb|add|, 320 bit & 1.25 & 0.81 & 1.54x \\
+    \verb|add|, 384 bit & 1.25 & 0.97 & 1.28x \\
+    \verb|add|, 448 bit & 1.34 & 1.01 & 1.32x \\
+    \verb|add|, 512 bit & 1.31 & 1.07 & 1.25x \\
     \hline
@@
     \end{tabular}
-\caption{Exection time for $10^8$ operations.}
+    \caption{Execution time comparison of simple integer operations using the GNU Multiprecision Library and the VLI library for $10^8$ operations.}
     \end{center}
 \end{table}
+\end{itemize}
+
+\paragraph{Optimized inner product with GPU accelerator}
+\begin{itemize}
 \item GPU benchmarks were performed on Todi, a Cray XK7 with NVIDIA Tesla K20X GPUs and AMD Opteron CPUs, at the Swiss Center for Scientific Computing (CSCS)
@@
 \item We estimate the performance in integer operations per second (IOP/s).
-It
 \item For the GPU performance estimate we only consider the instructions for the polynomial-polynomial multiplications and neglect the reduction at the end of the inner product.
@@
 \item We reach up to $214$ GIOP/s using 32 bit multiply-add (\verb|madc|)
-\item Theoretical maximum for the NVIDIA Tesla K20X are Number of Multiprocessors (SMX) $\cdot$ Clock frequency $\cdot$ Instruction throughput of \verb|madc|  =  $14 \, \mathrm{SMX} \cdot 732 \, \mathrm{MHz} \cdot 32 \, \mathrm{ IOP / cycle / SMX} = 327 \, \mathrm{GIOP/s}$.
+\item Theoretical maximum for the NVIDIA Tesla K20X is given by Number of Multiprocessors (SMX) $\cdot$ Clock frequency $\cdot$ Instruction throughput of \verb|madc|  =  $14 \, \mathrm{SMX} \cdot 732 \, \mathrm{MHz} \cdot 32 \, \mathrm{ IOP / cycle / SMX} = 327 \, \mathrm{GIOP/s}$.
 (Multiply-add throughput taken from \cite{CUDAProgramming}).
 \item We reach $65\%$ of the peak.
-
+\item Poor performance for low orders as well as for polynomials in 1 symbolic variable, since too little coefficients need to be calculated to saturate the number of available threads.
+\item Maximal performance for ``dense'' polynomial in 3 variables with a truncation at 7th order, where we employ $(2\cdot7+1)^3 = 3375$ threads.
+Using more threads may yield an even better performance, but we start to exceed the size of the texture memory cache and have to load data from the device memory more often, which causes the decrease of performance for higher orders.
+\item Polynomials in 4 symbolic variables only up to 6th order due to memory limitations on the device.
+Could be overcome by splitting the vectors and performing the inner product on these parts sequentially.
+\item Performance for ``triangular'' polynomials rises lower with increasing order, because the number of threads used grows slower with the order than for ``dense'' polynomials.
 \end{itemize}

### 2013-01-26 doc/paper_vli/vli2.tex
vli paper wip

@@
 
-\title{VLI - a library for large integer and polynomial arithemtic}
+\title{VLI - a library for high precision integer and polynomial arithemtic}
 \author{Timoth\'ee Ewart\inst{1}\thanks{Acknowledgments: Maxim Milakov, Peter Messmer:   NVIDIA,   Williams Sawyer, Gilles Fourestey:  CSCS, HP2C funding, hp2c.ch}  , Andreas Hehn$^2$, Matthias Troyer\inst{2} and Thierry Giamarchi\inst{1}}
@@
 The inner product will result in a polynomial of twice the order of the original polynomials
-with coefficients of twice the width of the original coefficients, i.e.\ 256 and
-512 bit, respectively.
+with coefficients of twice the width of the original coefficients.
 \paragraph{CPU}
 \begin{itemize}
-\item We continue to use the single large integer kernels
+\item We use the single large integer kernels
 \item OpenMP parallelization: split vectors into chunks, each thread does part of the inner product. Very regular problem, good load balance.
-\item No SIMD, since we had to use 32bit words and the number of operations for a multiplication increases by a factor TODO due to the manual handling of the carry.
+\item A SIMD implementation to perform 4 large integer operations in parallel on current \verb|x86_64| architectures is not useful, since we had to use 32bit integer operations, which increases the number of multiply instructions for a large integer multiplication by a factor of 4 compared to the 64bit based implementation.
+Using current Streaming SIMD Extensions (SSE) allows only two 32bit multiplications yielding 64bit results in a SIMD fashion.
+Thus a SIMD implementation using SSE will be two times slower than our serial version using 64 bit integers.
+We also neglected the carry bit propagation, which needs to be done manually in SSE, while the sequential version takes advantage of the hardware support for carry bit propagation.
+\item The upcoming AVX2 will double the vector width and support four 32bit integer multiplications, but no carry bit support.
 \end{itemize}
@@
 \item {\bf Implementation details}
+\item based on 32bit integers due to architecture.
 \item Element-wise polynomial multiplication + reduction at the end.
@@
 We group the coefficients of a fictive result polynomial by packet of 32 } into a task and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}).
-\item Since threads within a warp are executed in a lockstep manner, we try to give all threads within one warp the same amount of work.
-Such that a thread has to wait as little as possible for the other threads of the warp to finish.
-\item We group the coefficients by 32 into a task and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}), i.e.\ each thread calculates one coefficient of the resulting polynomial.
-\item Since the work load differences within a warp are small, the 
-%{\color{blue} Preparation of the task is done on CPU}
-
+\item Since threads within a warp are executed in a lockstep manner, we try to give all threads within a warp the same amount of work.
+This way the threads have to wait as little as possible for the other threads of the warp to finish.
+\item We group the coefficients by 32 into a ``task'' and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}), i.e.\ each thread calculates one coefficient of the resulting polynomial.
 \begin{figure}[t]
@@
     \begin{center}
-    \begin{tabular}{l|cc|c}
+    \begin{tabular}{l|cc|r}
      Operation & Execution & time [s] & Speed-up\\
@@
      \hline
-    \verb|add|, 128 bit & 1.11 & 0.33 & 3.36x \\
+    \verb|add|, 128 bit & 1.11 & 0.33 & {\bf 3.36x} \\
     \verb|add|, 192 bit & 1.21 & 0.45 & 2.68x \\
@@
     \verb|mul|, 128 bit $\rightarrow$ 256 bit & 1.36 & 0.93 & 1.46x \\
-    \verb|mul|, 192 bit $\rightarrow$ 384 bit & 3.21 & 1.18 & 2.72x \\
+    \verb|mul|, 192 bit $\rightarrow$ 384 bit & 3.21 & 1.18 & {\bf 2.72x} \\
     \verb|mul|, 256 bit $\rightarrow$ 512 bit & 3.74 & 2.04 & 1.83x \\ 
@@
 \item For the GPU performance estimate we only consider the instructions for the polynomial-polynomial multiplications and neglect the reduction at the end of the inner product.
+\begin{figure}[t!]
+    \begin{center}
+    \mbox{
+        \hspace{-0.5cm}
+        \includegraphics[scale=0.37, angle=-90]{ME128.eps} 
+        \hspace{-0.4cm}
+        \includegraphics[scale=0.37, angle=-90]{MC128.eps} 
+    }
+    \caption{Left: Inner product of dense polynomials with up to 3 variables and 126 to 256 bits coefficients.
+Right: Inner product of triangular polynomials with up to 3 variables and 126 to 256 bits coefficients.
+Size of the vector 4096.
+For four variables we are limited by the memory of the node.
+GMP gives similar results whatever the polynomials.}
+    \label{ResME}
+    \end{center}
+\end{figure}
 
@@
     \caption{Performance of the inner product of two vectors of dimension 4096 with ``dense'' (left) and ``triangular'' (right) polynomials having 128 bit coefficients on a NVIDIA Tesla K20X (GK110).}
-    \label{ResMEMIPS}
+    \label{fig:ResMEMIPS}
     \end{center}
@@
 \item We reach $65\%$ of the peak.
-\item Poor performance for low orders as well as for polynomials in 1 symbolic variable, since too little coefficients need to be calculated to saturate the number of available threads.
+\item Poor performance for low orders as well as for polynomials in 1 symbolic variable, since too little coefficients need to be calculated to saturate the number of available threads (Fig.
+\ref{fig:ResMEMIPS}).
 \item Maximal performance for ``dense'' polynomial in 3 variables with a truncation at 7th order, where we employ $(2\cdot7+1)^3 = 3375$ threads.
@@
 \begin{itemize}
-\item We 
+\item We presented a C++ library
 \item License?

### 2013-01-26 doc/paper_vli/vli2.tex
vli paper wip

@@
 \begin{itemize}
-\item {\bf Description}
+\item {\bf Description:}
 \item signed integer arithmetic
@@
 \item in addition to that: multiply-add, extended multiplication which doubles the number of bits, extended add which increases the size by one 64bit word.
-\item {\bf Implementation details}
+\item {\bf Implementation details:}
 \item data is stored as 64 bit integer arrays.
@@
 \begin{itemize}
-\item {\bf Description}
+\item {\bf Description:}
 \item symbolic polynomial in 1-4 variables.
@@
 \item special multiplication which will double the truncation order
-\item {\bf Implementation details}
+\item {\bf Implementation details:}
 \item structure and truncation order $N$ is compile time fixed 
@@
 \item Many small tasks of the same form. $\rightarrow$ well suited for GPUs.
-\item {\bf Implementation details}
+\item {\bf Implementation details:}
 \item based on 32bit integers due to architecture.
@@
     \verb|add|, 448 bit & 1.34 & 1.01 & 1.32x \\
-    \verb|add|, 512 bit & 1.31 & 1.07 & 1.25x \\
+    \verb|add|, 512 bit & 1.31 & 1.07 & {\bf 1.25x} \\
     \hline
@@
     \caption{Execution time comparison of simple integer operations using the GNU Multiprecision Library and the VLI library for $10^8$ operations.}
+    \label{tab:vli_vs_gmp}
     \end{center}
 \end{table}
+\item Comparing the execution time of simple addition and multiplication operations our implementation for fixed size integers is between $25\%$ and $236\%$ faster for the addition and up to $172\%$ faster for the multiplication (table \ref{tab:vli_vs_gmp}).
+\item Note that the reported time is the pure execution time of the operation. All variables were allocated before.
+Since our integer library is stack based we do not need expensive calls to \verb|malloc|, which will give an additional speedup for the inner product.
 \end{itemize}

### 2013-01-26 doc/paper_vli/vli2.tex
vli paper wip

@@
 
-\title{VLI - a library for high precision integer and polynomial arithemtic}
+\title{VLI - a library for high precision integer and polynomial arithmetic}
 \author{Timoth\'ee Ewart\inst{1}\thanks{Acknowledgments: Maxim Milakov, Peter Messmer:   NVIDIA,   Williams Sawyer, Gilles Fourestey:  CSCS, HP2C funding, hp2c.ch}  , Andreas Hehn$^2$, Matthias Troyer\inst{2} and Thierry Giamarchi\inst{1}}
@@
 \end{equation}
-\item The number of terms in this sum depends on the orders $I,J,K,L$ of the resulting coefficient $\rightarrow$ load inbalance
+\item The number of terms in this sum depends on the orders $I,J,K,L$ of the resulting coefficient $\rightarrow$ load in-balance
 %\item {\color{blue} We design an execution plan.
@@
 \end{figure}
-\item {\color{blue} During the execution of a task by a warp,  each threads into the warp get the indices of the fictive coefficient, determine the necessary  contributions and  then calculate the output coefficient}
-\item {\color{blue} Reduction is performed under NVIDA standard}
-\item {\color{blue} This algorithm is efficient whatever the type of polynomial. But the termination of the contributions of an output coefficient can be tricky}
+\item Each thread of the warp gets the indices of the resulting coefficient and determines the necessary terms to calculate the output coefficient.
+\item Reduction is implemented according to the best practices by NVIDIA (TODO reference)
+%\item {\color{blue} This algorithm is efficient whatever the type of polynomial. But the termination of the contributions of an output coefficient can be tricky}
 %\item Coalesced memory access?
-\item {\color{blue}  Karatsuba algorithm \cite{Karatsuba1963} for the largest extended multiplication (256 bit to 512 bit) on GPU}
+\item We use the Karatsuba algorithm \cite{Karatsuba1963} for the largest extended multiplication (256 bit to 512 bit) on GPU, because an integer addition requires substantially less cycles than multiplication the GPU.
 \item Using NVIDIA parallel thread execution (PTX) assembly language to benefit from hardware carry bit propagation \verb|addc.cc| and multiply-add with carry bit propagation \verb|madc.cc|.
@@
 ``triangular'' polynomials of order 1 to 14 with 128 and 256 bit integer
-coefficients. Which will result in polynomials of order 2 to 28 with 256 and 512 bit integer coefficiencts, respectively.
+coefficients. Which will result in polynomials of order 2 to 28 with 256 and 512 bit integer coefficients, respectively.
 \item We estimate the performance in integer operations per second (IOP/s).
@@
     }
-    \caption{Left: Inner product of dense polynomials with up to 3 variables and 126 to 256 bits coefficients.
+    \caption{Comparison of the inner product with and without GPU accelerator.
+Left: Inner product of dense polynomials with up to 3 variables and 126 to 256 bits coefficients.
 Right: Inner product of triangular polynomials with up to 3 variables and 126 to 256 bits coefficients.
 Size of the vector 4096.
-For four variables we are limited by the memory of the node.
-GMP gives similar results whatever the polynomials.}
+GMP gives similar results independent of the polynomial structure.}
     \label{ResME}
@@
 \begin{itemize}
-\item We presented a C++ library
+\item TODO
 \item License?

### 2013-01-28 doc/paper_vli/vli2.tex
vli paper wip

@@
 architectures. Our main target application is high-temperature series
-expansions which require inner products of large vectors of polynomials
+expansions which requires inner products of large vectors of polynomials
 with large integer coefficients. For this purpose we implemented a
@@
 \section{Large integers}
-\begin{itemize}
-\item {\bf Description:}
-\item signed integer arithmetic
-\item fixed size (128-512bit)
-\item all standard integer operations (arithmetic, comparison and bit operations)
-\item in addition to that: multiply-add, extended multiplication which doubles the number of bits, extended add which increases the size by one 64bit word.
-\item {\bf Implementation details:}
-\item data is stored as 64 bit integer arrays.
-\item two-complement representation
-\item stack based $\rightarrow$ cheap memory allocation
-\item addition and multiplication using the standard schoolbook algorithms.
-\item Numbers are two small for Toom\cite{Toom} or FFT based approaches like the Sch\"onhage-Strassen algorithm\cite{Schonhage}.
-\item optimized ASM kernels to use hardware features not accessible from C++ (\verb|x86_64| carry addition \verb|adcq|, 64bit multiplication \verb|mulq| which calculates low and hi 64bit part of the result.)
-\item TODO which special features do we use on power64?
-\item assembly code generated using \verb|BOOST_PP| to compactify the code and minimize bugs.
-\end{itemize}
+The library provides a class template \verb|integer<n_bits>| to represent signed integers with a fixed size of $128 \le n \le 512$ bits.
+Objects of this class behave like the regular C++ type $int$.
+We implemented all standard integer operations starting from basic arithmetic operations, comparion operations and bit operations, like bit-shift or bit-wise logical operations.
+In addition to these operations we also provide a multiply-add function, an extended multiplcation which doubles the number of bits when two integers of the same size get multiplied and an extended add which increases the size of the integer by one 64bit word.
+The data of the integer is stored as a 64 bit unsigned integer array, where we use a standard two-complement represnetation of the number.
+Due to the fixed size of the integer, we are able to allocate the memory for the data on the stack.
+This way we save expensive heap allocations, which would require a manual memory pool management to get efficient allocations.
+The basic arithmetic operations like addition and multiplication use standard schoolbook algorithms, since the integers are too small for Toom\cite{Toom} or FFT based approaches like the Sch\"onhage-Strassen algorithm\cite{Schonhage}.
+We implemented the schoolbook algorithms in optimized assembly code to make optimal use of the hardware features that are not accessible from C++, like the carry addition \verb|adcq| or the 64bit multiplication \verb|mulq| which calculates the low and hi 64 bit part of the result on the \verb|x86-64| architecture.
+Since writing assembly code can result in very long error-prone code, we generate most parts of the assembly code with the C++ preprocessor using the \verb|BOOST_PP| library.
+
+%\begin{itemize}
+%\item {\bf Description:}
+%\item signed integer arithmetic
+%\item fixed size (128-512bit)
+%\item all standard integer operations (arithmetic, comparison and bit operations)
+%\item in addition to that: multiply-add, extended multiplication which doubles the number of bits, extended add which increases the size by one 64bit word.
+%\item {\bf Implementation details:}
+%\item data is stored as 64 bit integer arrays.
+%\item two-complement representation
+%\item stack based $\rightarrow$ cheap memory allocation
+%\item addition and multiplication using the standard schoolbook algorithms.
+%\item Numbers are two small for Toom\cite{Toom} or FFT based approaches like the Sch\"onhage-Strassen algorithm\cite{Schonhage}.
+%\item optimized ASM kernels to use hardware features not accessible from C++ (\verb|x86_64| carry addition \verb|adcq|, 64bit multiplication \verb|mulq| which calculates low and hi 64bit part of the result.)
+%\item TODO which special features do we use on power64?
+%\item assembly code generated using \verb|BOOST_PP| to compactify the code and minimize bugs.
+%\end{itemize}
 
 \section{Polynomials}
+The second part of the library is a template class 
+\begin{verbatim}
+    polynomial<CoeffType, Structure<N>, Var0, Var1, Var2, Var3>
+\end{verbatim}
+for symbolic polynomials in 1 to 4 variables
+\begin{equation}
+    p(x,y,z,a) = \sum_{i,j,k,l} c_{ijkl} x^i y^j z^k a^l
+\end{equation}
+having either a ``dense'' (${i,j,k,l} \le N$) or a ``triangular'' ($i+j+k+l \le N$) structure.
+The truncation order $N$ is fixed at compile-time.
+As coefficient $c_{ijkl}$ of the polynomial any arithemetic type may be used.
+
 \begin{itemize}
@@
 \item symbolic polynomial in 1-4 variables.
-\begin{equation}
-    p(x,y,z,a) = \sum_{i,j,k,l} c_{ijkl} x^i y^j z^k a^l
-\end{equation}
 \item either ``dense'' (${i,j,k,l} \le N$) or ``triangular'' ($i+j+k+l \le N$) structure.

### 2013-01-28 doc/paper_vli/vli2.tex
vli paper wip

@@
 \section{Large integers}
-The library provides a class template \verb|integer<n_bits>| to represent signed integers with a fixed size of $128 \le n \le 512$ bits.
-Objects of this class behave like the regular C++ type $int$.
+The library provides a template class \verb|integer<n>| to represent signed integers with a fixed size of $128 \le n \le 512$ bits.
+Objects of this class behave like the regular C++ type \verb|int|.
 We implemented all standard integer operations starting from basic arithmetic operations, comparion operations and bit operations, like bit-shift or bit-wise logical operations.
-In addition to these operations we also provide a multiply-add function, an extended multiplcation which doubles the number of bits when two integers of the same size get multiplied and an extended add which increases the size of the integer by one 64bit word.
-The data of the integer is stored as a 64 bit unsigned integer array, where we use a standard two-complement represnetation of the number.
+In addition to these operations we also provide a multiply-add function, an extended multiplcation which doubles the number of bits when two integers of the same size get multiplied and an extended add which increases the size of the integer by one 64 bit word.
+
+The data of the integer is stored as a 64 bit unsigned integer array, where we use a standard two-complement representation of the number.
 Due to the fixed size of the integer, we are able to allocate the memory for the data on the stack.
@@
 The basic arithmetic operations like addition and multiplication use standard schoolbook algorithms, since the integers are too small for Toom\cite{Toom} or FFT based approaches like the Sch\"onhage-Strassen algorithm\cite{Schonhage}.
-We implemented the schoolbook algorithms in optimized assembly code to make optimal use of the hardware features that are not accessible from C++, like the carry addition \verb|adcq| or the 64bit multiplication \verb|mulq| which calculates the low and hi 64 bit part of the result on the \verb|x86-64| architecture.
+We implemented the schoolbook algorithms in optimized assembly code to make optimal use of the hardware features that are not accessible from C++, like the carry addition \verb|adcq| or the 64 bit multiplication \verb|mulq| which calculates the low and hi 64 bit part of the result on the \verb|x86-64| architecture.
 Since writing assembly code can result in very long error-prone code, we generate most parts of the assembly code with the C++ preprocessor using the \verb|BOOST_PP| library.
@@
 \begin{equation}
-    p(x,y,z,a) = \sum_{i,j,k,l} c_{ijkl} x^i y^j z^k a^l
+    p(x,y,z,a) = \sum_{i,j,k,l} c_{ijkl} \cdot x^i y^j z^k a^l
 \end{equation}
@@
 The truncation order $N$ is fixed at compile-time.
-As coefficient $c_{ijkl}$ of the polynomial any arithemetic type may be used.
+The coefficients $c_{ijkl}$ of the polynomial may have any C++ type/class supporting basic arithmetic operations.
+The polynomial itself supports additions, subtractions and multiplications with other polynomials or monomials.
+We also provide a special multiplication function returning a polynomial with truncation order $2N$, such that no terms are dropped.
+It is also possible to mix and match polynomials having different sets of variables.
+The coefficients will be automatically mapped to the corresponding symbols during compile time.
+All operations call functions with default implementations for all coefficient types.
+These functions may be overloaded by the user to provide optimized implementations for specific coefficient types.
+The user may for example add a specialized implementation for \verb|float| using Streaming SIMD Extension intrinsics without touching the library itself.
+Except for the general default implementation of the operations the library provides optimized routines for large integer coefficients \verb|integer<n>|.
 
-\begin{itemize}
-\item {\bf Description:}
-\item symbolic polynomial in 1-4 variables.
-\item either ``dense'' (${i,j,k,l} \le N$) or ``triangular'' ($i+j+k+l \le N$) structure.
-\item any arithmetic type as coefficient $c_{ijkl}$
-\item addition, subtraction, multiplication with other polynomials and monomials
-\item special multiplication which will double the truncation order
-\item {\bf Implementation details:}
-\item structure and truncation order $N$ is compile time fixed 
-\item function hooks to allow custom optimized kernels for certain coefficient types
-\item The symbolic variable names are given as template parameter
-\item Automatic matching of variable names of different polynomial types using template meta programming.
-\end{itemize}
+%\begin{itemize}
+%\item {\bf Description:}
+%\item symbolic polynomial in 1-4 variables.
+%\item either ``dense'' (${i,j,k,l} \le N$) or ``triangular'' ($i+j+k+l \le N$) structure.
+%\item any arithmetic type as coefficient $c_{ijkl}$
+%\item addition, subtraction, multiplication with other polynomials and monomials
+%\item special multiplication which will double the truncation order
+%\item {\bf Implementation details:}
+%\item structure and truncation order $N$ is compile time fixed 
+%\item function hooks to allow custom optimized kernels for certain coefficient types
+%\item The symbolic variable names are given as template parameter
+%\item Automatic matching of variable names of different polynomial types using template meta programming.
+%\end{itemize}
 
 \section{Optimized inner product}
-The hot spot of HTSE are inner products of large vectors with symbolic polynomials as components. The coefficients of the polynomials are large integers.
-All polynomials are assumed to have the same truncation order.
-The inner product will result in a polynomial of twice the order of the original polynomials
-with coefficients of twice the width of the original coefficients.
+Since the hot spots of our main target application are inner products of large vectors with symbolic polynomials as components, where the coefficients of the polynomials are large integers, we implemented an optimized inner product function for this special case.
+All polynomials of the vector are assumed to have the same structure and truncation order $N$ and have the same \verb|integer<n>| type as coefficients.
+The inner product will result in a polynomial of twice the order of the original polynomials with coefficients of twice the width \verb|integer<2n>| of the original coefficients.
+To employ today's supercomputers as efficiently as possible the inner product is a hybrid CPU/GPU implementation,
+where the inner product the vectors are split into a part calculated on the CPU and a part calculated on the GPU.
+The ratio of these two parts can be set when building the library.
+It is also possible to deactivate either part completely and compile a pure CPU version.
+
 \paragraph{CPU}
-\begin{itemize}
-\item We use the single large integer kernels
-\item OpenMP parallelization: split vectors into chunks, each thread does part of the inner product. Very regular problem, good load balance.
-\item A SIMD implementation to perform 4 large integer operations in parallel on current \verb|x86_64| architectures is not useful, since we had to use 32bit integer operations, which increases the number of multiply instructions for a large integer multiplication by a factor of 4 compared to the 64bit based implementation.
+On the CPU we perform the inner product by element-wise polynomial-polynomial multiplications using the implementation of the polynomial.
+This operation can easily parallelized using OpenMP by splitting the vector and distributing equal-sized chunks among the available threads.
+Since all polynomials are of the same structure the work is well balanced between the threads.
+
+The highly parallel structure of the operation suggests a SIMD approach.
+However to perform 4 large integer operations in parallel on current \verb|x86-64| architectures is not advisable,
+since the kernels need to be based 32 bit integer operations instead of 64 bit operations which are available for the sequential version.
+This increases the number of required multiply instructions for a large integer multiplication by a factor of 4.
 Using current Streaming SIMD Extensions (SSE) allows only two 32bit multiplications yielding 64bit results in a SIMD fashion.
 Thus a SIMD implementation using SSE will be two times slower than our serial version using 64 bit integers.
-We also neglected the carry bit propagation, which needs to be done manually in SSE, while the sequential version takes advantage of the hardware support for carry bit propagation.
-\item The upcoming AVX2 will double the vector width and support four 32bit integer multiplications, but no carry bit support.
-\end{itemize}
+Furthermore we neglected the carry bit propagation, which needs to be done manually in SSE,
+while the sequential version takes advantage of the hardware support for carry bit propagation.
+
+The upcoming AVX2 instruction set will feature integer SIMD operations with twice the vector width of SSE.
+However, it will neither support any 64 bit integer operations nor any carry bit propagation.
+Therefore it will at most be as fast as our sequential version.
+
+%\begin{itemize}
+%\item We use the single large integer kernels
+%\item OpenMP parallelization: split vectors into chunks, each thread does part of the inner product. Very regular problem, good load balance.
+%\item A SIMD implementation to perform 4 large integer operations in parallel on current \verb|x86_64| architectures is not useful, since we had to use 32bit integer operations, which increases the number of multiply instructions for a large integer multiplication by a factor of 4 compared to the 64bit based implementation.
+%Using current Streaming SIMD Extensions (SSE) allows only two 32bit multiplications yielding 64bit results in a SIMD fashion.
+%Thus a SIMD implementation using SSE will be two times slower than our serial version using 64 bit integers.
+%We also neglected the carry bit propagation, which needs to be done manually in SSE, while the sequential version takes advantage of the hardware support for carry bit propagation.
+%\item The upcoming AVX2 will double the vector width and support four 32bit integer multiplications, but no carry bit support.
+%\end{itemize}
 \paragraph{GPU}
@@
 \item TODO
+\item Extend integers up to 2048 bits.
+\item Allow more polynomial structures
 \item License?

### 2013-01-29 doc/paper_vli/vli2.tex
vli paper wip

@@
-\documentclass[oribibl]{llncs2e/llncs}
+\documentclass[oribibl,a4paper]{llncs2e/llncs}
 \usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
@@
 \title{VLI - a library for high precision integer and polynomial arithmetic}
-\author{Timoth\'ee Ewart\inst{1}\thanks{Acknowledgments: Maxim Milakov, Peter Messmer:   NVIDIA,   Williams Sawyer, Gilles Fourestey:  CSCS, HP2C funding, hp2c.ch}  , Andreas Hehn$^2$, Matthias Troyer\inst{2} and Thierry Giamarchi\inst{1}}
+\author{Timoth\'ee Ewart\inst{1}\thanks{Acknowledgments: Maxim Milakov, Peter Messmer:   NVIDIA,   Williams Sawyer, Gilles Fourestey:  CSCS, HP2C funding, hp2c.ch}  , Andreas Hehn$^2$ and Matthias Troyer\inst{2}}
 
@@
 
-\paragraph{CPU}
 On the CPU we perform the inner product by element-wise polynomial-polynomial multiplications using the implementation of the polynomial.
@@
 
-The highly parallel structure of the operation suggests a SIMD approach.
-However to perform 4 large integer operations in parallel on current \verb|x86-64| architectures is not advisable,
-since the kernels need to be based 32 bit integer operations instead of 64 bit operations which are available for the sequential version.
-This increases the number of required multiply instructions for a large integer multiplication by a factor of 4.
-Using current Streaming SIMD Extensions (SSE) allows only two 32bit multiplications yielding 64bit results in a SIMD fashion.
+Even though the parallel structure of the operation suggests a SIMD approach performing 4 large integer operations in parallel on current \verb|x86-64| architectures is not advisable.
+Using the Streaming SIMD Extensions (SSE) all kernels need to be based on 32 bit integer operations instead of 64 bit operations which are available for the sequential version.
+This reduction of word size increases the number of required multiply instructions for a large integer multiplication by a factor of 4.
+Furthermore, the current SSE instruction set only supports an integer SIMD multiplication for two 32bit numbers, each yielding a 64 bit result.
 Thus a SIMD implementation using SSE will be two times slower than our serial version using 64 bit integers.
-Furthermore we neglected the carry bit propagation, which needs to be done manually in SSE,
-while the sequential version takes advantage of the hardware support for carry bit propagation.
+Note that we also neglected the carry bit propagation of the required additions.
+While the sequential version benefits of the hardware support for carry bit propagation, an SSE implementation needs to handle the carry manually.
 
 The upcoming AVX2 instruction set will feature integer SIMD operations with twice the vector width of SSE.
-However, it will neither support any 64 bit integer operations nor any carry bit propagation.
-Therefore it will at most be as fast as our sequential version.
+However, it will neither support 64 bit integer multiplications nor carry bit propagation. Therefore it will at most be as fast as our sequential version.
 
@@
 \paragraph{GPU}
+On the GPU the picture is slightly different. The SIMT
 \begin{itemize}
@@
 \begin{figure}[t]
-    \begin{center}
+    \centering
     \mbox{
@@
     \label{fig:GPU_load_balance}
-    \end{center}
 \end{figure}
@@
 \begin{table}
-    \begin{center}
+    \centering
     \begin{tabular}{l|cc|r}
@@
     \label{tab:vli_vs_gmp}
-    \end{center}
 \end{table}
@@
 \begin{figure}[t!]
-    \begin{center}
+    \centering
     \mbox{
@@
     \label{ResME}
-    \end{center}
 \end{figure}
@@
     \label{fig:ResMEMIPS}
-    \end{center}
 \end{figure}  

### 2013-01-30 doc/paper_vli/vli2.tex
update table paper + figure

@@
     \hline
-    \verb|add|, 128 bit & 1.11 & 0.33 & {\bf 3.36x} \\
-    \verb|add|, 192 bit & 1.21 & 0.45 & 2.68x \\
-    \verb|add|, 256 bit & 1.11 & 0.81 & 1.37x \\
-    \verb|add|, 320 bit & 1.25 & 0.81 & 1.54x \\
-    \verb|add|, 384 bit & 1.25 & 0.97 & 1.28x \\
-    \verb|add|, 448 bit & 1.34 & 1.01 & 1.32x \\
-    \verb|add|, 512 bit & 1.31 & 1.07 & {\bf 1.25x} \\
+   \verb|add|, 128 bit & 1.11 & 0.28 & {\bf 3.96x} \\
+   \verb|add|, 192 bit & 1.21 & 0.42 & 2.88x \\
+   \verb|add|, 256 bit & 1.11 & 0.77 & 1.44x \\
+   \verb|add|, 320 bit & 1.25 & 0.94 & 1.32x \\
+   \verb|add|, 384 bit & 1.25 & 0.91 & 1.37x \\
+   \verb|add|, 448 bit & 1.34 & 0.97 & 1.38x \\
+   \verb|add|, 512 bit & 1.31 & 1.04 & {\bf 1.26x} \\
    \hline
@@
 \end{table}
-\item Comparing the execution time of simple addition and multiplication operations our implementation for fixed size integers is between $25\%$ and $236\%$ faster for the addition and up to $172\%$ faster for the multiplication (table \ref{tab:vli_vs_gmp}).
+
+\item Comparing the execution time of simple addition and multiplication operations our implementation for fixed size integers is between $26\%$ and $296\%$ faster for the addition and up to $172\%$ faster for the multiplication (table \ref{tab:vli_vs_gmp}).
 \item Note that the reported time is the pure execution time of the operation. All variables were allocated before.

### 2013-01-31 doc/paper_vli/vli2.tex
vli paper wip

@@
 Objects of this class behave like the regular C++ type \verb|int|.
-We implemented all standard integer operations starting from basic arithmetic operations, comparion operations and bit operations, like bit-shift or bit-wise logical operations.
-In addition to these operations we also provide a multiply-add function, an extended multiplcation which doubles the number of bits when two integers of the same size get multiplied and an extended add which increases the size of the integer by one 64 bit word.
+We implemented all standard integer operations starting from basic arithmetic operations, comparison operations and bit operations, like bit-shift or bit-wise logical operations.
+In addition to these operations we also provide a multiply-add function, an extended multiplication which doubles the number of bits when two integers of the same size get multiplied and an extended add which increases the size of the integer by one 64 bit word.
 
@@
 The upcoming AVX2 instruction set will feature integer SIMD operations with twice the vector width of SSE.
-However, it will neither support 64 bit integer multiplications nor carry bit propagation. Therefore it will at most be as fast as our sequential version.
+However, it will neither support 64 bit integer multiplications nor carry bit propagation.
+Therefore it will be at most as fast as our sequential version.
 
@@
 \paragraph{GPU}
-On the GPU the picture is slightly different. The SIMT
-\begin{itemize}
-\item Many small tasks of the same form. $\rightarrow$ well suited for GPUs.
-\item {\bf Implementation details:}
-\item based on 32bit integers due to architecture.
-\item Element-wise polynomial multiplication + reduction at the end.
-\item Each thread calculates one coefficient $c_{IJKL}$ the product polynomial.
-\item Several coefficients $a_{ijkl}$, $b_{i'j'k'l'}$ of the polynomials to be multiplied contribute to the same coefficient $c_{IJKL}$
-\begin{equation}
-    c_{IJKL} = \sum_{i,j,k,l} a_{ijkl} \cdot b_{I-i,J-j,K-k,L-l}
+While on the CPU a SIMD approach is not promising, the highly parallel structure of the problem is well suited for GPU accelerators.
+In a nutshell GPUs offer a hybrid between SIMD and massively multithreaded execution,
+where the GPU schedules thousands of threads in so called warps. A warp is a group of usually 32 threads which are executed in a lockstep manner.
+We implemented the inner product in NVIDIA CUDA to exploit this powerful architecture.
+Our target hardware is the NVIDIA Tesla K20X.
+We perform element-wise polynomial multiplications where each thread calculates one coefficient of the product polynomials
+\begin{equation} \label{coefficient_sum}
+    c_{IJKL} = \sum_{i,j,k,l} a_{ijkl} \cdot b_{I-i,J-j,K-k,L-l} \,
 \end{equation}
-\item The number of terms in this sum depends on the orders $I,J,K,L$ of the resulting coefficient $\rightarrow$ load in-balance
-%\item {\color{blue} We design an execution plan.
-We group the coefficients of a fictive result polynomial by packet of 32 } into a task and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}).
-\item Since threads within a warp are executed in a lockstep manner, we try to give all threads within a warp the same amount of work.
-This way the threads have to wait as little as possible for the other threads of the warp to finish.
-\item We group the coefficients by 32 into a ``task'' and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}), i.e.\ each thread calculates one coefficient of the resulting polynomial.
+where  $a_{ijkl}$, $b_{i'j'k'l'}$ are the coefficients of the polynomials to be multiplied.
+Once all coefficients are calculated we perform a reduction over the result vector to obtain the final result of the inner product.
+This way we avoid race conditions and minimize the number of synchronization points.
+However, this method leads to a load-inbalance since the number of terms in the sum (Eq.\ \ref{coefficient_sum}) depends on the orders $I,J,K,L$ of the resulting coefficient.
+To overcome this problem we set up an execution plan before the actual calculation.
+Therefor we determine the number of terms of each coefficient of the product polynomial and sort them according to this number and assemble groups of 32 coefficients.
+Starting with the most expensive group the groups get scheduled on the available warps balancing the work among the available warps (Fig.\ \ref{fig:GPU_load_balance}).
+Note that by sorting the coefficients we also minimize the work-load difference within a warp, such that the threads within the warp idle as little as possible until the other threads finished their lockstep calculation.
+
 \begin{figure}[t]
@@
     }
-    \caption{Left: Number of contributions to each coefficient of the resulting polynomial (red line) sorted in reverse order and split into tasks. Right: Load balancing of the tasks over the warps.}
+    \caption{Left: Number of contributions to each coefficient of the resulting polynomial (red line) sorted and split into tasks. Right: Load balancing of the tasks over the warps.}
     \label{fig:GPU_load_balance}
 \end{figure}
-\item Each thread of the warp gets the indices of the resulting coefficient and determines the necessary terms to calculate the output coefficient.
+
+%The data of the polynomials to be multiplied is copied asynchonously from the host to the device's texture memory.
+Once the \verb|inner_product| function is called the vectors to be multiplied are copied asynchronously from the host to the device.
+We store the data in the texture memory to take advantage of the texture cache which optimized for 2D access.
+% TODO maybe something about the size and how we use it exactly.
+The intermediate results, i.e. the vector of the product polynomials, is written to the global memory.
+At this point we change the memory layout from the usual Array (vector) of Array (polynomial) of Structures (large integer) to an Array of Structures of Arrays where the data segments of the large integer coefficients are interleaved within the polynomial (Fig.\ \ref{fig:GPU_data_layout}).
+\begin{figure}[t]
+    \centering
+    \mbox{
+    }
+    \caption{Data layout of the intermediate data on the GPU.}
+    \label{fig:GPU_data_layout}
+\end{figure}
+Thus we can access the coefficients in a coalesced way during the reduction at the end of the calculation, where each thread requires first the least significant word of the coefficient, then the next higher significant word and so on.
+Once the reduction is completed the result is transferred from the device memory back to the host using the regular memory layout.
+
+Like for the SSE approach on the CPU our kernels on the GPU rely on 32 bit unsigned integer arithmetic.
+However, on the GPU we were able to profit from hardwired addition with carry \verb|addc.cc| and multiply-add with carry \verb|madc.cc| using NVIDIA parallel thread execution (PTX) inline assembly.
+On the GPU we also apply the Karatsuba algorithm \cite{Karatsuba1963}, which corresponds to the Toom-2 method, for the 256 bit integer multiplication because it uses less multiplications and more additions on the words than the schoolbook algorithm.
+While this advantage is negligible on the CPU, we save TODO\% on the GPU, since an integer addition requires substantially less cycles than an integer multiplication on the NVIDIA Kepler.
+%Unlike on the CPU where we operate on 64 bit words, we e
+%Internally the kernels operate on 32 bit unsigned integers
+\begin{itemize}
+%\item Many small tasks of the same form. $\rightarrow$ well suited for GPUs.
+%\item {\bf Implementation details:}
+\item based on 32bit integers due to architecture.
+%\item Element-wise polynomial multiplication + reduction at the end.
+%\item Each thread calculates one coefficient $c_{IJKL}$ the product polynomial.
+%\item Several coefficients $a_{ijkl}$, $b_{i'j'k'l'}$ of the polynomials to be multiplied contribute to the same coefficient $c_{IJKL}$
+%\begin{equation}
+%    c_{IJKL} = \sum_{i,j,k,l} a_{ijkl} \cdot b_{I-i,J-j,K-k,L-l}
+%\end{equation}
+%\item The number of terms in this sum depends on the orders $I,J,K,L$ of the resulting coefficient $\rightarrow$ load in-balance
+%\item {\color{blue} We design an execution plan.
+We group the coefficients of a fictive result polynomial by packet of 32 } into a task and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}).
+%\item Since threads within a warp are executed in a lockstep manner, we try to give all threads within a warp the same amount of work.
+This way the threads have to wait as little as possible for the other threads of the warp to finish.
+%\item We group the coefficients by 32 into a ``task'' and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}), i.e.\ each thread calculates one coefficient of the resulting polynomial.
+%\item Each thread of the warp gets the indices of the resulting coefficient and determines the necessary terms to calculate the output coefficient.
 \item Reduction is implemented according to the best practices by NVIDIA (TODO reference)
-%\item {\color{blue} This algorithm is efficient whatever the type of polynomial. But the termination of the contributions of an output coefficient can be tricky}
-%\item Coalesced memory access?
-\item We use the Karatsuba algorithm \cite{Karatsuba1963} for the largest extended multiplication (256 bit to 512 bit) on GPU, because an integer addition requires substantially less cycles than multiplication the GPU.
+%%\item {\color{blue} This algorithm is efficient whatever the type of polynomial. But the termination of the contributions of an output coefficient can be tricky}
+%\item We use the Karatsuba algorithm \cite{Karatsuba1963} for the largest extended multiplication (256 bit to 512 bit) on GPU, because an integer addition requires substantially less cycles than multiplication the GPU.
 \item Using NVIDIA parallel thread execution (PTX) assembly language to benefit from hardware carry bit propagation \verb|addc.cc| and multiply-add with carry bit propagation \verb|madc.cc|.

### 2013-02-01 doc/paper_vli/vli2.tex
added memory layout pictures

@@
 We implemented all standard integer operations starting from basic arithmetic operations, comparison operations and bit operations, like bit-shift or bit-wise logical operations.
-In addition to these operations we also provide a multiply-add function, an extended multiplication which doubles the number of bits when two integers of the same size get multiplied and an extended add which increases the size of the integer by one 64 bit word.
+In addition to these operations we also provide a multiply-add function, an extended multiplication which doubles the number of bits when two integers of the same size get multiplied and an extended add which increases the size of the integer by one 64 bit segment.
 
@@
 The basic arithmetic operations like addition and multiplication use standard schoolbook algorithms, since the integers are too small for Toom\cite{Toom} or FFT based approaches like the Sch\"onhage-Strassen algorithm\cite{Schonhage}.
+%An exception is the 256 bit multiplication on the GPU, where we use the Karatsuba algorithm \cite{Karatsuba1963}.
 We implemented the schoolbook algorithms in optimized assembly code to make optimal use of the hardware features that are not accessible from C++, like the carry addition \verb|adcq| or the 64 bit multiplication \verb|mulq| which calculates the low and hi 64 bit part of the result on the \verb|x86-64| architecture.
@@
 Using the Streaming SIMD Extensions (SSE) all kernels need to be based on 32 bit integer operations instead of 64 bit operations which are available for the sequential version.
-This reduction of word size increases the number of required multiply instructions for a large integer multiplication by a factor of 4.
+This reduction of segment size increases the number of required multiply instructions for a large integer multiplication by a factor of 4.
 Furthermore, the current SSE instruction set only supports an integer SIMD multiplication for two 32bit numbers, each yielding a 64 bit result.
@@
 The intermediate results, i.e. the vector of the product polynomials, is written to the global memory.
-At this point we change the memory layout from the usual Array (vector) of Array (polynomial) of Structures (large integer) to an Array of Structures of Arrays where the data segments of the large integer coefficients are interleaved within the polynomial (Fig.\ \ref{fig:GPU_data_layout}).
+At this point we change the memory layout from the usual Array (vector) of Array (polynomial) of Structures (large integer) to an Array of Structures of Arrays where the data segments of the large integer coefficients are interleaved within the polynomial (Fig.\ \ref{fig:GPU_data_layout}), such that the least significant segments of all coefficients are continuous in memory followed by the next more signficiant segments.
 \begin{figure}[t]
     \centering
-    \mbox{
+\includegraphics[scale=0.9]{memory_layout.eps} \\
+    \vspace{0.5cm}
+%    \hline
+\includegraphics[scale=0.9]{memory_layout_gpu.eps} 
+    \caption{Memory layout of a vector of polynomials with 192 bit integer coefficients.
+             Each coefficient $\textsf{c}_{ijkl}$ consists of three 64 bit segments \textsf{s}.
+             Top: Regular memory layout used on the CPU. Each coefficient is stored contiguously.
+             Bottom: Memory layout used for intermediate results on the GPU. The coefficients are stored in an interleaved way grouping the segments according to their significance.
+             Note that we will have 32 bit segments on the GPU, which doubles the number of segments in return.
     }
-    \caption{Data layout of the intermediate data on the GPU.}
     \label{fig:GPU_data_layout}
 \end{figure}
-Thus we can access the coefficients in a coalesced way during the reduction at the end of the calculation, where each thread requires first the least significant word of the coefficient, then the next higher significant word and so on.
+This memory layout allows for a efficient coalesced access during the reduction at the end of the calculation.
 Once the reduction is completed the result is transferred from the device memory back to the host using the regular memory layout.
@@
 Like for the SSE approach on the CPU our kernels on the GPU rely on 32 bit unsigned integer arithmetic.
-However, on the GPU we were able to profit from hardwired addition with carry \verb|addc.cc| and multiply-add with carry \verb|madc.cc| using NVIDIA parallel thread execution (PTX) inline assembly.
-On the GPU we also apply the Karatsuba algorithm \cite{Karatsuba1963}, which corresponds to the Toom-2 method, for the 256 bit integer multiplication because it uses less multiplications and more additions on the words than the schoolbook algorithm.
-While this advantage is negligible on the CPU, we save TODO\% on the GPU, since an integer addition requires substantially less cycles than an integer multiplication on the NVIDIA Kepler.
-%Unlike on the CPU where we operate on 64 bit words, we e
-%Internally the kernels operate on 32 bit unsigned integers
-\begin{itemize}
+However, on the GPU we were able to profit from hard-wired addition with carry \verb|addc.cc| and multiply-add with carry \verb|madc.cc| using NVIDIA parallel thread execution (PTX) inline assembly.
+Contrary to the CPU where we used only schoolbook algorithms, we employ the Karatsuba algorithm \cite{karatsuba1963}, which corresponds to the Toom-2 method, for the 256 bit integer multiplication on the GPU.
+Even though the method does not require less operations than the schoolbook algorithm it is advantageous on the GPU, because it uses less multiplications and more additions on the 32 bit segments.
+While on the CPU these operations are almost equally expensive, the addition on the NVIDIA Kepler has a 5 times higher throughput than the multiplication \cite{CUDAProgramming}.
+
+%\begin{itemize}
 %\item Many small tasks of the same form. $\rightarrow$ well suited for GPUs.
 %\item {\bf Implementation details:}
-\item based on 32bit integers due to architecture.
+%\item based on 32bit integers due to architecture.
 %\item Element-wise polynomial multiplication + reduction at the end.
@@
 %\item Each thread of the warp gets the indices of the resulting coefficient and determines the necessary terms to calculate the output coefficient.
-\item Reduction is implemented according to the best practices by NVIDIA (TODO reference)
+%\item Reduction is implemented according to the best practices by NVIDIA (TODO reference)
 %%\item {\color{blue} This algorithm is efficient whatever the type of polynomial. But the termination of the contributions of an output coefficient can be tricky}
 %\item We use the Karatsuba algorithm \cite{Karatsuba1963} for the largest extended multiplication (256 bit to 512 bit) on GPU, because an integer addition requires substantially less cycles than multiplication the GPU.
-\item Using NVIDIA parallel thread execution (PTX) assembly language to benefit from hardware carry bit propagation \verb|addc.cc| and multiply-add with carry bit propagation \verb|madc.cc|.
-\end{itemize}
+%\item Using NVIDIA parallel thread execution (PTX) assembly language to benefit from hardware carry bit propagation \verb|addc.cc| and multiply-add with carry bit propagation \verb|madc.cc|.
+%\end{itemize}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-02-01 doc/paper_vli/vli2.tex
minor fixes

@@
 The basic arithmetic operations like addition and multiplication use standard schoolbook algorithms, since the integers are too small for Toom\cite{Toom} or FFT based approaches like the Sch\"onhage-Strassen algorithm\cite{Schonhage}.
-%An exception is the 256 bit multiplication on the GPU, where we use the Karatsuba algorithm \cite{Karatsuba1963}.
+An exception is the 256 bit multiplication on the GPU, where we use the Karatsuba algorithm \cite{Karatsuba1963} as we will explain later.
 We implemented the schoolbook algorithms in optimized assembly code to make optimal use of the hardware features that are not accessible from C++, like the carry addition \verb|adcq| or the 64 bit multiplication \verb|mulq| which calculates the low and hi 64 bit part of the result on the \verb|x86-64| architecture.
@@
              Bottom: Memory layout used for intermediate results on the GPU. The coefficients are stored in an interleaved way grouping the segments according to their significance.
-             Note that we will have 32 bit segments on the GPU, which doubles the number of segments in return.
+             Note that on the GPU we will have segments of only 32 bit width, which doubles the number of segments in return.
     }
@@
 However, on the GPU we were able to profit from hard-wired addition with carry \verb|addc.cc| and multiply-add with carry \verb|madc.cc| using NVIDIA parallel thread execution (PTX) inline assembly.
-Contrary to the CPU where we used only schoolbook algorithms, we employ the Karatsuba algorithm \cite{karatsuba1963}, which corresponds to the Toom-2 method, for the 256 bit integer multiplication on the GPU.
+Contrary to the CPU where we used only schoolbook algorithms, we employ the Karatsuba algorithm \cite{Karatsuba1963} for the 256 bit integer multiplication on the GPU.
 Even though the method does not require less operations than the schoolbook algorithm it is advantageous on the GPU, because it uses less multiplications and more additions on the 32 bit segments.

### 2013-02-01 doc/paper_vli/vli2.tex
vli paper wip

@@
 
+% TODO write something on the number of threads used
+
 %\begin{itemize}
@@
 \paragraph{Simple integer operations}
-\begin{itemize}
-\item pure CPU benchmarks were done on a Sandy bridge Intel Xeon E5-2670 with 32 logical cores.
+We benchmarked the large integer part of the library on an Intel Sandy bridge Xeon E5-2670 node with 32 logical cores
+and compared against the commonly used GNU Multiple Precision Arithemtic library (GMP).
+The results are shown in table \ref{tab:vli_vs_gmp}.
 \begin{table}
@@
 \end{table}
+Comparing the execution time of simple addition and multiplication operations our implementation for fixed size integers is between $26\%$ and $296\%$ faster for the addition and up to $172\%$ faster for the multiplication.
+Note that the reported time is the pure execution time of the operation. All variables were allocated before.
+The allocation will 
+Since our integer library is stack based we do not need expensive calls to \verb|malloc| like GMP, which will give an additional speedup for the inner product, where multiple allocations are necessary.
 
-\item Comparing the execution time of simple addition and multiplication operations our implementation for fixed size integers is between $26\%$ and $296\%$ faster for the addition and up to $172\%$ faster for the multiplication (table \ref{tab:vli_vs_gmp}).
-\item Note that the reported time is the pure execution time of the operation. All variables were allocated before.
-Since our integer library is stack based we do not need expensive calls to \verb|malloc|, which will give an additional speedup for the inner product.
-\end{itemize}
+%\begin{itemize}
+%\item pure CPU benchmarks were done on a Sandy bridge Intel Xeon E5-2670 with 32 logical cores.
+%\item Comparing the execution time of simple addition and multiplication operations our implementation for fixed size integers is between $26\%$ and $296\%$ faster for the addition and up to $172\%$ faster for the multiplication (table \ref{tab:vli_vs_gmp}).
+%\item Note that the reported time is the pure execution time of the operation. All variables were allocated before.
+%Since our integer library is stack based we do not need expensive calls to \verb|malloc|, which will give an additional speedup for the inner product.
+%\end{itemize}
 
 \paragraph{Optimized inner product with GPU accelerator}
+The benchmarks for the inner product were again performed on the Intel Sandy bridge node for the pure CPU version of the inner product.
+The GPU benchmarks were performed on Todi, a Cray XK7 with NVIDIA Tesla K20X GPUs and AMD Opteron CPUs, at the Swiss Center for Scientific Computing (CSCS).
+The test cases are inner products of two vectors of dimension 4096 with ``dense'' and ``triangular'' polynomials of order 1 to 14 with 128 and 256 bit integer coefficients.
+Since the inner product will double the order of the polynomial and the width of the large integer coefficients, the test cases will result in polynomials of order 2 to 28 with 256 and 512 bit integer coefficients, respectively.
+
+Figure \ref{fig:performance_large_int_ops} shows the performance of the inner product for the various test cases in number of large integer operations per second.
+Monovariant polynomials show a rather poor performance of at most $1.6 \text{G large integer OP/s}$ on the CPU.
+
 \begin{itemize}
-\item GPU benchmarks were performed on Todi, a Cray XK7 with NVIDIA Tesla K20X GPUs and AMD Opteron CPUs, at the Swiss Center for Scientific Computing (CSCS)
-\item Testcase: inner product of two vectors dimension 4096, with ``dense'' and
-``triangular'' polynomials of order 1 to 14 with 128 and 256 bit integer
-coefficients. Which will result in polynomials of order 2 to 28 with 256 and 512 bit integer coefficients, respectively.
+%\item GPU benchmarks were performed on Todi, a Cray XK7 with NVIDIA Tesla K20X GPUs and AMD Opteron CPUs, at the Swiss Center for Scientific Computing (CSCS)
+%\item Testcase: inner product of two vectors dimension 4096, with ``dense'' and
+%``triangular'' polynomials of order 1 to 14 with 128 and 256 bit integer
+%coefficients. Which will result in polynomials of order 2 to 28 with 256 and 512 bit integer coefficients, respectively.
 \item We estimate the performance in integer operations per second (IOP/s).
@@
 GMP gives similar results independent of the polynomial structure.}
-    \label{ResME}
+    \label{fig:performance_large_int_ops}
 \end{figure}

### 2013-05-16 doc/ambient/tensor_ws/tikz/energy_contraction.tex
tikz diagram for ambient poster

\documentclass{standalone}

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amscd}
\usepackage{color}
\usepackage{amsthm}

\usepackage{tikz}
\usetikzlibrary{calc}
\usetikzlibrary{arrows,shapes,positioning,decorations.pathmorphing}

\begin{document}

\tikzset{spin/.style={circle=2pt,draw=black!100,fill=blue!20,inner sep=3pt}}
\tikzset{hamil/.style={rectangle=2pt,draw=black!100,fill=red!40,inner sep=3pt}}
\tikzset{coarse grain/.style={isosceles triangle,isosceles triangle stretches,draw, fill=green!15, anchor=center, inner sep=3pt, minimum width=.8cm, shape border rotate=270, thick}}
\tikzset{arrow down/.style={single arrow, draw, fill, shape border rotate=270, anchor=center, minimum height=.4cm, single arrow head extend=.2cm, inner sep=8pt}}

\begin{tikzpicture}

\foreach \i in {1,...,5} {
	\node[spin] (M1 \i) at (\i,0) {};
%	\draw[thick] (M1 \i.south) -- +(0,-.4);
	\node[hamil] (H \i) at (\i,-.5) {};
	\draw[thick] (M1 \i) -- (H \i);
	\node[spin] (M2 \i) at (\i,-1) {};
	\draw[thick] (H \i) -- (M2 \i);
}
\foreach \i [evaluate = \i as \j using int(\i+1)] in {1,...,4} {
	\draw[thick] (M1 \i) -- (M1 \j);
	\draw[thick] (H \i) -- (H \j);
	\draw[thick] (M2 \i) -- (M2 \j);
}

%\draw[dashed,thick] (M1 5.east) -- +(.4,0);
%\draw[dashed,thick] (M1 1.west) -- +(-.4,0);

\end{tikzpicture}

\end{document}

### 2014-04-01 doc/ambient/paper_ambient/paper_ambient.tex
Initial version, svn r4227

--- 2011-10-05 doc/paper_ambient/paper_ambient.tex
@@
 between nodes (PVM, NX, Express, P4, \etc) \cite{Aoyama:1999}, in particular the Message Passing Interface (MPI) \cite{Snir:1998:MCR:552013}. which has become the
-standard for coarse-grained problem ( \footnote{A number of scientific applications requiring scalability currently employ a main MPI layer,  [GIVE ME A MPI CODE INTO ALPS !!!!]
-{\sc Quantum ESPRESSO} \cite{QE-2009} for quantum chemistry, Abinit \cite{ABINIT-2005, ABINIT-2009} in quantum mechanics, ScaLAPACK \cite{SCALAPACK-1997}
-in linear  algebra or SMILE for hypersonic rarefied flows \cite{SMILE-2005}, to give just a brief, non-exhaustive list.} The factor limiting scalability in this programming model
+standard for coarse-grained problem ( \footnote{A number of scientific applications requiring scalability currently employ a main MPI layer, 
+{ ALPS \cite{ALPS2011} for strongly correlated quantum mechanical systems ,\sc Quantum ESPRESSO} \cite{QE-2009} for quantum chemistry, Abinit \cite{ABINIT-2005, ABINIT-2009} in quantum mechanics, ScaLAPACK \cite{SCALAPACK-1997} in linear  algebra or SMILE for hypersonic rarefied flows \cite{SMILE-2005}, to give just a brief, non-exhaustive list.} The factor limiting scalability in this programming model
 is the communication overhead and the workload imbalance among processes.
@@
 \subsection{First benchmark case}
-[STARTED REWRITING THIS; BUT NOT FINISHED]
 The first benchmark consists of evaluating the performance of our parallel matrix multiplication relative to the ScaLAPACK-MKL implementation. Both solvers are included as computational kernels
@@
 \begin{figure}
-\include{PLOT/SinglePDGEMM/SinglePGEMM}
-\caption{Performances of PDGEMM using ScaLAPACK }
+\include{PLOT/SinglePDGEMM/SinglePGEMMWK}
+\caption{Performances of PDGEMM using Ambient  \label{PLOTSINGLE0}}
 \end{figure}
+
+\begin{figure}
+\include{PLOT/SinglePDGEMM/SinglePGEMMWKSCALA}
+\caption{Comparing Ambient's PDGEMM against ScaLAPACK  \label{PLOTSINGLEq}}
+\end{figure}
+
+The benchmark of our Ambient pDGEMM is presented on the figure \ref{PLOTSINGLE0}
 
@@
 This benchmarks  evaluates the global performance of the scheduler, especially, when  the independent operations are executed all over the cluster. 
-We now multiply two vectors of matrices pair by pair. Every multiplication is performed by two MPI process and the size of the vectors vary from to
+We now multiply two vectors of matrices pair by pair. Every multiplication is performed by 2 MPI process and the size of the vectors vary from to
 1 to 48. Thus, the final number of MPI process is equal to 96 which corresponds to fill up 4 calculation nodes.
@@
 Before presenting and commenting the results, we would like highlight the readers of the programming of this benchmark. The coding simply consists of few lines
-of declarations, initializations and a basic loop for, as we presented into the algorithm \ref{EXECUTION} . Naturally, coded algorithm  integrates two levels of distributed parallelization 
+of declarations, initializations and a basic loop for, as we presented into the algorithm \ref{EXECUTION}. Naturally, coded algorithm  integrates two levels of distributed parallelization 
 without integrate any MPI instructions.
@@
 In this benchmark, we  plot the ratio between the performance peak of  vector of matrices $P_{v_m}$ and the associated peak of the machine $P_m$. It is expressed by 
-$R = P_{v_m}/P_m$, or
+$R = P_{v_m}/P_m$, or developed by :
 
 \begin{eqnarray}
-R = 100\frac{N_m 2n^3}{N_p t4f},
+R = 100\frac{2n^3 N_m }{4f t N_p},
 \end{eqnarray}
-where $N_m$ is the number of matrices, $N_p$ the number of processor, $t$ the full time of execution and $f$ the frequency of the processor. Results are presented
+where $N_m$ is the number of matrices, $N_p$ the number of processor, $t$ the full time of execution and $f$ the frequency of the processor. Results are presented on the figure \ref{PLOTVECTOR0}
 
 \begin{figure}
-\include{PLOT/VectorPDGEMM/VectorWL}
-\caption{something else}
+\include{PLOT/VectorPDGEMM/VectorWL0}
+\caption{Peak performance ration of a variable vector of matrices. \label{PLOTVECTOR0}}
 \end{figure}

### 2014-04-01 doc/ambient/tensor_ws/tensor_ws.tex
Initial version, svn r4227

\documentclass[a4paper]{article}

%% do not edit the following macro
\usepackage{graphics}
\newcommand{\abs}[4]{
\begin{center}
  {\large\bf #1}

   \vspace*{.6em}
  {#2}

   \vspace*{0.2em}
  {\it #3}

\end{center}
{#4}
}

\begin{document}
\abs{%
% 1. Title of Your Talk
Exposure of Dataflow parallelism in Density Matrix Renormalization Group}{%
%%%% 2. List of Authors
Alexandr Kosenkov$^1$, Michele Dolfi$^1$, Timothee Ewart$^2$, Adrian Kantian$^{2}$, Bela Bauer$^{3}$, Matthias Troyer$^1$}{%
%%%% 3. List of Affiliations
$^1$Theoretische Physik, ETH Zurich, 8093 Zurich, Switzerland\\
$^2$DPMC-MaNEP, University of Geneva, 1211 Geneva, Switzerland\\
$^3$Microsoft Research, Station Q, University of California, Santa Barbara, CA 93106, USA}{%
%%%% 4. Abstract

\begin{itemize}
\item Density Matrix Renormalization Group application deals with tensor operations using complex matrices objects. The application is automatically parallelized using Ambient: dataflow distributed memory framework for linear algebra operations.
\end{itemize}

}

\end{document}

### 2014-04-01 benchmarks/mkl/scalapack/CPar.h

@@ -1,0 +1,200 @@
+
+/*
+C++ style = reference, pointers are evil, but we used fortran library !!!
+*/
+extern "C" void blacs_pinfo_( int* mypnum, int* nprocs); 
+extern "C" void blacs_gridinfo_( int* nContxt, int* nRows, int* nCols, int* nMyRow, int* nMyCol);
+extern "C" void blacs_gridinit_(  int* nContxt, char * order,  int* np_row,  int* np_col);
+extern "C" void blacs_gridexit_( int* nContxt);
+extern "C" void blacs_exit_( int* error_code);
+extern "C" void blacs_get_(int* Contxt, int* what, int* val); 
+extern "C" int numroc_(int* nN, int* nB, int* nIProc, int* nISrcProc, int* nNProcs);
+extern "C" void pdgesvd_(char *jobu, char *jobvt,
+		int *m, int *n,
+		double *a, int *ia, int *ja, int *desca,
+		double *s,
+		double *u, int *iu, int *ju, int *descu,
+		double *vt, int *ivt, int *jvt, int *descvt,
+		double *work, int *lwork,double *rwork, int *info);
+
+
+class CGrid
+{
+public:
+	//First Constructor by hand !
+	CGrid(int nRowsGrid, int nColsGrid, int nDesiredBlockSiwe):_nRowsGrid(nRowsGrid),_nColsGrid(nColsGrid),
+								   _nDesiredBlockSize(nDesiredBlockSiwe),_nMyRow(0),_nMyCol(0){}
+	
+	//Second Constructor automatic !
+	CGrid(int nNumProcs, int nDesiredBlockSiwe):_nMyRow(0),_nMyCol(0)
+	{
+		_nDesiredBlockSize = nDesiredBlockSiwe;
+		//Cast C++ style
+		_nRowsGrid = static_cast<int>(floor(sqrt(nNumProcs)));
+		
+		while (nNumProcs % _nRowsGrid != 0)
+			_nRowsGrid--;
+
+		_nColsGrid = nNumProcs/_nRowsGrid;
+	};
+	
+
+	~CGrid(){}	
+	
+	void InitBlockSize(int nM, int nN)
+	{
+		int zero = 0;
+	
+		if (nM < _nRowsGrid*_nDesiredBlockSize)
+		{
[…]

### 2014-04-01 src/ambient/regression/functional/algo.exp.cpp

@@ -1,0 +1,54 @@
+#include "params.hpp"
+
+template<class T, std::size_t Size>
+struct helper_test{
+     /* only for TE (complex) */
+     static void test(){ }
+};
+
+template<class T, std::size_t Size>
+struct helper_test<std::complex<T>, Size >{
+     static void test(){
+         typedef alps::numeric::matrix<std::complex<T>> sMatrix;
+         typedef ambient::numeric::tiles<ambient::numeric::matrix<std::complex<T> > > pMatrix;
+     
+         std::complex<T> a; 
+         ambient::numeric::kernels::detail::randomize(a);
+     
+         pMatrix pA(Size, Size), pB(Size, Size);
+         sMatrix sA(Size, Size), sB(Size, Size);
+        
+         generate(pA);
+         sA = cast<sMatrix>(pA);
+     
+         sB = exp(sA,a);
+         pB = exp(pA,a);
+     
+         BOOST_CHECK(pB==sB);
+     }
+};
+
+BOOST_AUTO_TEST_CASE_TEMPLATE( EXP_ALGO, T, test_types) {
+    helper_test<typename T::value_type, T::valuex>::test(); 
+}
+
+BOOST_AUTO_TEST_CASE_TEMPLATE( EXP_ALGO_HERMITIAN, T, test_types) {
+    typedef alps::numeric::matrix<typename T::value_type> sMatrix;
+    typedef ambient::numeric::tiles<ambient::numeric::matrix<typename T::value_type> > pMatrix;
+
+    typename T::value_type a; 
+    ambient::numeric::kernels::detail::randomize(a);
+
+    int Size = T::valuex; 
+
+    pMatrix pA(Size, Size), pB(Size, Size);
+    sMatrix sA(Size, Size), sB(Size, Size);
+   
+    generate_hermitian(pA);
+    sA = cast<sMatrix>(pA);
+
+    sB = exp_hermitian(sA,a);
+    pB = exp_hermitian(pA,a);
+
+    BOOST_CHECK(pB==sB);
+}

### 2014-04-01 src/ambient/regression/performance/gemm.blas.cpp

@@ -1,0 +1,53 @@
+#include "params.hpp"
+
+namespace Random{
+   struct random {
+       random(){};
+       double operator()(){return drand48();} 
+       int IntRd(){return rand();}
+   };
+}
+
+BOOST_AUTO_TEST_CASE_TEMPLATE(gemm_blas, T, test_types){
+    size_t x = get_input_x<T>();
+    size_t y = get_input_y<T>();
+    size_t nthreads = get_input_threads<T>();
+
+    Random::random Rd;
+    omp_set_num_threads(nthreads);
+    typename T::value_type* ad;
+    typename T::value_type* bd;
+    typename T::value_type* cd;
+    
+    int m,n,k;
+    int lda,ldb,ldc;
+    double alpha(1.0), beta(1.0);
+
+    m = (int)x;
+    n = (int)y;
+    k = (int)y;
+
+    lda = ldb = ldc = (int)x;
+
+    ad = new typename T::value_type[x*y]; 
+    bd = new typename T::value_type[x*y]; 
+    cd = new typename T::value_type[x*y]; 
+  
+    memset((void*)cd,0,x*y*sizeof(typename T::value_type));
+   
+    for(int i(0); i< x*y ; ++i){
+        ad[i] = Rd();
+        bd[i] = Rd();
+    }
+       
+    ambient::async_timer time("blas");
+    time.begin();
+    ambient::numeric::kernels::helper_blas<typename T::value_type>::gemm("N","N", &m, &n, &k, &alpha, ad, &lda, bd, &ldb, &beta, cd, &ldc);
+    time.end();
+
+    report(time, GFlopsGemm, x, y, nthreads);
+    delete [] ad;   
+    delete [] bd;   
+    delete [] cd;   
+}
+

### 2014-04-01 src/ambient/regression/performance/qr.ambient.cpp

@@ -1,0 +1,38 @@
+#include "params.hpp"
+
+BOOST_AUTO_TEST_CASE_TEMPLATE( test, T, test_types){
+    typedef ambient::dim2 dim;
+    typedef alps::numeric::matrix<typename T::value_type> sMatrix;
+    typedef ambient::numeric::tiles<ambient::numeric::matrix<typename T::value_type> > pMatrix;
+
+    size_t x = get_input_x<T>();
+    size_t y = get_input_y<T>();
+    size_t nthreads = get_input_threads<T>();
+
+    pMatrix pA(x, y);
+    pMatrix pQ(x, y);
+    pMatrix pR(x, y);
+
+    sMatrix sA(x, y);
+    sMatrix sQ(x, y);
+    sMatrix sR(x, y);
+
+    generate(pA);
+    sA = cast<sMatrix>(pA);
+
+    ambient::sync();
+    qr(sA, sQ, sR);
+    qr(pA, pQ, pR); 
+
+    ambient::async_timer time("ambient");
+    time.begin();
+    ambient::sync();
+    time.end();
+
+    report(time, GFlopsGemm, x, y, nthreads);
+    ambient::sync();
+
+    BOOST_CHECK(sQ == pQ);
+    BOOST_CHECK(sR == pR);
+}
+

### 2014-04-01 src/ambient/src/ambient/channels/mpi/request.h

@@ -1,0 +1,60 @@
+/*
+ * Ambient, License - Version 1.0 - May 3rd, 2012
+ *
+ * Permission is hereby granted, free of charge, to any person or organization
+ * obtaining a copy of the software and accompanying documentation covered by
+ * this license (the "Software") to use, reproduce, display, distribute,
+ * execute, and transmit the Software, and to prepare derivative works of the
+ * Software, and to permit third-parties to whom the Software is furnished to
+ * do so, all subject to the following:
+ *
+ * The copyright notices in the Software and this entire statement, including
+ * the above license grant, this restriction and the following disclaimer,
+ * must be included in all copies of the Software, in whole or in part, and
+ * all derivative works of the Software, unless such copies or derivative
+ * works are solely in the form of machine-executable object code generated by
+ * a source language processor.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
+ * DEALINGS IN THE SOFTWARE.
+ */
+
+#ifndef AMBIENT_CHANNELS_MPI_REQUEST
+#define AMBIENT_CHANNELS_MPI_REQUEST
+
+namespace ambient { namespace channels { namespace mpi {
+
+    class request_impl : public memory::use_bulk_new<request_impl> {
+    public:
+        request_impl(){}
+        request_impl(void(*impl)(request_impl*), typename channel::scalar_type& v, int target, int tag = 0);
+        request_impl(void(*impl)(request_impl*), typename channel::block_type& r, int target, int tag = 0);
+        inline bool operator()();
+        void* data;
+        int extent;
+        int target;
+        MPI_Request mpi_request;
+        void(*impl)(request_impl*);
+        bool once;
+        int tag;
+    };
+
+    class request {
[…]

### 2014-04-01 src/ambient/src/ambient/channels/mpi/request.hpp

@@ -1,0 +1,74 @@
+/*
+ * Ambient, License - Version 1.0 - May 3rd, 2012
+ *
+ * Permission is hereby granted, free of charge, to any person or organization
+ * obtaining a copy of the software and accompanying documentation covered by
+ * this license (the "Software") to use, reproduce, display, distribute,
+ * execute, and transmit the Software, and to prepare derivative works of the
+ * Software, and to permit third-parties to whom the Software is furnished to
+ * do so, all subject to the following:
+ *
+ * The copyright notices in the Software and this entire statement, including
+ * the above license grant, this restriction and the following disclaimer,
+ * must be included in all copies of the Software, in whole or in part, and
+ * all derivative works of the Software, unless such copies or derivative
+ * works are solely in the form of machine-executable object code generated by
+ * a source language processor.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
+ * DEALINGS IN THE SOFTWARE.
+ */
+
+namespace ambient { namespace channels { namespace mpi {
+
+    // type information required //
+    inline request_impl::request_impl(void(*impl)(request_impl*), typename channel::scalar_type& v, int target, int tag)
+    : extent(sizeof(typename channel::scalar_type::numeric_union)/sizeof(double)), 
+      data(&v.v),
+      target(target),
+      impl(impl),
+      tag(tag),
+      once(false)
+    {
+    }
+    // type information required //
+    inline request_impl::request_impl(void(*impl)(request_impl*), typename channel::block_type& r, int target, int tag)
+    : extent(r.spec.extent/sizeof(double)), 
+      data(r.data),
+      target(target),
+      impl(impl),
+      tag(tag),
+      once(false)
+    {
+    }
+    inline bool request_impl::operator()(){
[…]

### 2014-04-01 src/ambient/src/ambient/interface/kernel.hpp

@@ -1,0 +1,118 @@
+/*
+ * Ambient, License - Version 1.0 - May 3rd, 2012
+ *
+ * Permission is hereby granted, free of charge, to any person or organization
+ * obtaining a copy of the software and accompanying documentation covered by
+ * this license (the "Software") to use, reproduce, display, distribute,
+ * execute, and transmit the Software, and to prepare derivative works of the
+ * Software, and to permit third-parties to whom the Software is furnished to
+ * do so, all subject to the following:
+ *
+ * The copyright notices in the Software and this entire statement, including
+ * the above license grant, this restriction and the following disclaimer,
+ * must be included in all copies of the Software, in whole or in part, and
+ * all derivative works of the Software, unless such copies or derivative
+ * works are solely in the form of machine-executable object code generated by
+ * a source language processor.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
+ * DEALINGS IN THE SOFTWARE.
+ */
+
+#ifndef AMBIENT_INTERFACE_KERNELS
+#define AMBIENT_INTERFACE_KERNELS
+#include "ambient/utils/timings.hpp"
+
+namespace ambient {
+
+    using ambient::controllers::ssm::functor;
+    using ambient::memory::instr_bulk;
+
+    template<typename FP, FP fp> struct kernel_inliner{};
+    #include "ambient/interface/pp/kernel_inliner.pp.hpp"
+
+    template<class K>
+    class kernel : public functor {
+    public:
+        #define inliner kernel_inliner<typename K::ftype,&K::c>
+        inline void operator delete (void* ptr){ }
+        inline void* operator new (size_t size){
[…]

### 2014-04-01 src/ambient/src/ambient/interface/pp/kernel_inliner.pp.hpp

@@ -1,0 +1,476 @@
+/*
+ * Ambient, License - Version 1.0 - May 3rd, 2012
+ *
+ * Permission is hereby granted, free of charge, to any person or organization
+ * obtaining a copy of the software and accompanying documentation covered by
+ * this license (the "Software") to use, reproduce, display, distribute,
+ * execute, and transmit the Software, and to prepare derivative works of the
+ * Software, and to permit third-parties to whom the Software is furnished to
+ * do so, all subject to the following:
+ *
+ * The copyright notices in the Software and this entire statement, including
+ * the above license grant, this restriction and the following disclaimer,
+ * must be included in all copies of the Software, in whole or in part, and
+ * all derivative works of the Software, unless such copies or derivative
+ * works are solely in the form of machine-executable object code generated by
+ * a source language processor.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
+ * DEALINGS IN THE SOFTWARE.
+ */
+
+template< typename T0 , void(*fp)( T0& )>
+struct kernel_inliner<void(*)( T0& ), fp> {
+    typedef T0 t0;
+    static const int arity = 1;
+    static inline void latch(functor* o, T0& arg0 ){
+        if(ctxt.tunable()){
+            info<T0>::typed::template score<0>(arg0);
+            ctxt.schedule();
+        }
+        if(ctxt.remote()){
+            info<T0>::typed::template modify_remote<0>(arg0);
+            return;
+        }else if(ctxt.local()){
+            info<T0>::typed::template modify_local<0>(arg0, o);
+        }else{
+            info<T0>::typed::template modify<0>(arg0, o);
+        }
[…]

### 2014-04-01 src/ambient/src/ambient/interface/typed.hpp

@@ -1,0 +1,378 @@
+/*
+ * Ambient, License - Version 1.0 - May 3rd, 2012
+ *
+ * Permission is hereby granted, free of charge, to any person or organization
+ * obtaining a copy of the software and accompanying documentation covered by
+ * this license (the "Software") to use, reproduce, display, distribute,
+ * execute, and transmit the Software, and to prepare derivative works of the
+ * Software, and to permit third-parties to whom the Software is furnished to
+ * do so, all subject to the following:
+ *
+ * The copyright notices in the Software and this entire statement, including
+ * the above license grant, this restriction and the following disclaimer,
+ * must be included in all copies of the Software, in whole or in part, and
+ * all derivative works of the Software, unless such copies or derivative
+ * works are solely in the form of machine-executable object code generated by
+ * a source language processor.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
+ * DEALINGS IN THE SOFTWARE.
+ */
+
+#ifndef AMBIENT_INTERFACE_TYPED
+#define AMBIENT_INTERFACE_TYPED
+
+#define EXTRACT(var) T* var = (T*)m->arguments[arg];
+
+namespace ambient { namespace numeric {
+    template <class T, class Allocator> class matrix;
+    template <class T> class transpose_view;
+    template <class T> class diagonal_matrix;
+} }
+
+namespace ambient {
+    template<typename T> class default_allocator;
+    using ambient::controllers::ssm::functor;
+    using ambient::models::ssm::history;
+    using ambient::models::ssm::revision;
+    using ambient::memory::instr_bulk;
+
+    // {{{ compile-time type info: singular types + inplace and future specializations
+    template <typename T> struct singular_info {
[…]

### 2014-04-01 src/ambient/src/ambient/memory/data_bulk.hpp

@@ -1,0 +1,68 @@
+/*
+ * Ambient, License - Version 1.0 - May 3rd, 2012
+ *
+ * Permission is hereby granted, free of charge, to any person or organization
+ * obtaining a copy of the software and accompanying documentation covered by
+ * this license (the "Software") to use, reproduce, display, distribute,
+ * execute, and transmit the Software, and to prepare derivative works of the
+ * Software, and to permit third-parties to whom the Software is furnished to
+ * do so, all subject to the following:
+ *
+ * The copyright notices in the Software and this entire statement, including
+ * the above license grant, this restriction and the following disclaimer,
+ * must be included in all copies of the Software, in whole or in part, and
+ * all derivative works of the Software, unless such copies or derivative
+ * works are solely in the form of machine-executable object code generated by
+ * a source language processor.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
+ * DEALINGS IN THE SOFTWARE.
+ */
+
+#ifndef AMBIENT_MEMORY_DATA_BULK_HPP
+#define AMBIENT_MEMORY_DATA_BULK_HPP
+
+namespace ambient { namespace memory {
+
+    inline data_bulk& data_bulk::instance(){
+        static data_bulk singleton; return singleton;
+    }
+
+    inline data_bulk::data_bulk(){
+        this->limit = ambient::isset("AMBIENT_BULK_LIMIT") ? ambient::getint("AMBIENT_BULK_LIMIT") : AMBIENT_MAX_SID;
+        this->reuse_enabled = ambient::isset("AMBIENT_BULK_REUSE") ? true : false; 
+        this->reset_enabled = ambient::isset("AMBIENT_BULK_DEALLOCATE") ? true : false; 
[…]

### 2014-04-01 src/ambient/src/ambient/memory/region.hpp

@@ -1,0 +1,119 @@
+/*
+ * Ambient, License - Version 1.0 - May 3rd, 2012
+ *
+ * Permission is hereby granted, free of charge, to any person or organization
+ * obtaining a copy of the software and accompanying documentation covered by
+ * this license (the "Software") to use, reproduce, display, distribute,
+ * execute, and transmit the Software, and to prepare derivative works of the
+ * Software, and to permit third-parties to whom the Software is furnished to
+ * do so, all subject to the following:
+ *
+ * The copyright notices in the Software and this entire statement, including
+ * the above license grant, this restriction and the following disclaimer,
+ * must be included in all copies of the Software, in whole or in part, and
+ * all derivative works of the Software, unless such copies or derivative
+ * works are solely in the form of machine-executable object code generated by
+ * a source language processor.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
+ * DEALINGS IN THE SOFTWARE.
+ */
+
+#ifndef AMBIENT_MEMORY_REGION
+#define AMBIENT_MEMORY_REGION
+
+namespace ambient { namespace memory {
+
+    constexpr size_t paged(size_t size, size_t page){  return page * (size_t)((size+page-1)/page); } // (page example: 4096)
+    constexpr size_t aligned_64(size_t size){ return 64 * (size_t)((size+63)/64); }
+    constexpr size_t aligned_8(size_t size){ return 8 * (size_t)((size+7)/8); }
+    template<size_t S> constexpr size_t aligned_64(){ return 64 * (size_t)((S+63)/64); }
+
+    template<size_t S, class Factory>
+    class serial_region {
+    public:
+        serial_region(){
+            this->buffer = NULL;
[…]

### 2014-04-01 src/ambient/src/ambient/numeric/kernels/math.hpp

@@ -1,0 +1,177 @@
+/*
+ * Ambient, License - Version 1.0 - May 3rd, 2012
+ *
+ * Permission is hereby granted, free of charge, to any person or organization
+ * obtaining a copy of the software and accompanying documentation covered by
+ * this license (the "Software") to use, reproduce, display, distribute,
+ * execute, and transmit the Software, and to prepare derivative works of the
+ * Software, and to permit third-parties to whom the Software is furnished to
+ * do so, all subject to the following:
+ *
+ * The copyright notices in the Software and this entire statement, including
+ * the above license grant, this restriction and the following disclaimer,
+ * must be included in all copies of the Software, in whole or in part, and
+ * all derivative works of the Software, unless such copies or derivative
+ * works are solely in the form of machine-executable object code generated by
+ * a source language processor.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
+ * DEALINGS IN THE SOFTWARE.
+ */
+
+#ifndef AMBIENT_NUMERIC_MATRIX_KERNELS_MATH
+#define AMBIENT_NUMERIC_MATRIX_KERNELS_MATH
+
+#define PLASMA_IB        64
+#define PlasmaNoTrans    111
+#define PlasmaTrans      112
+#define PlasmaConjTrans  113
+#define PlasmaUpper      121
+#define PlasmaLower      122
+#define PlasmaUpperLower 123
+#define PlasmaLeft       141
+#define PlasmaRight      142
+#define PlasmaForward    391
+#define PlasmaColumnwise 401
+
+typedef int PLASMA_enum;
+
+template <PLASMA_enum TR>
+struct trans_type{
+    const static int value = TR;
+};
+
[…]

### 2014-04-01 src/ambient/src/ambient/numeric/kernels/utils.hpp

@@ -1,0 +1,89 @@
+/*
+ * Ambient, License - Version 1.0 - May 3rd, 2012
+ *
+ * Permission is hereby granted, free of charge, to any person or organization
+ * obtaining a copy of the software and accompanying documentation covered by
+ * this license (the "Software") to use, reproduce, display, distribute,
+ * execute, and transmit the Software, and to prepare derivative works of the
+ * Software, and to permit third-parties to whom the Software is furnished to
+ * do so, all subject to the following:
+ *
+ * The copyright notices in the Software and this entire statement, including
+ * the above license grant, this restriction and the following disclaimer,
+ * must be included in all copies of the Software, in whole or in part, and
+ * all derivative works of the Software, unless such copies or derivative
+ * works are solely in the form of machine-executable object code generated by
+ * a source language processor.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
+ * DEALINGS IN THE SOFTWARE.
+ */
+
+#ifndef AMBIENT_NUMERIC_MATRIX_KERNELS_UTILS
+#define AMBIENT_NUMERIC_MATRIX_KERNELS_UTILS
+
+#include <cstring>
+
+namespace ambient {
+
+    template <typename T>
+    inline void memcpy(T* dd, T *sd, size_t w, T alfa){
+        std::memcpy(dd, sd, w);
+    }
+
+    template <typename T>
+    inline void memscal(T* dd, T *sd, size_t w, T alfa){
+        int z = w/sizeof(T);
+        do{ *dd++ = alfa*(*sd++); }while(--z > 0); // note: dd != sd
+    }
+
+    template <typename T>
+    inline void memscala(T* dd, T *sd, size_t w, T alfa){
[…]

### 2014-04-01 src/ambient/src/ambient/numeric/matrix.hpp

@@ -1,0 +1,30 @@
+/*
+ * Ambient, License - Version 1.0 - May 3rd, 2012
+ *
+ * Permission is hereby granted, free of charge, to any person or organization
+ * obtaining a copy of the software and accompanying documentation covered by
+ * this license (the "Software") to use, reproduce, display, distribute,
+ * execute, and transmit the Software, and to prepare derivative works of the
+ * Software, and to permit third-parties to whom the Software is furnished to
+ * do so, all subject to the following:
+ *
+ * The copyright notices in the Software and this entire statement, including
+ * the above license grant, this restriction and the following disclaimer,
+ * must be included in all copies of the Software, in whole or in part, and
+ * all derivative works of the Software, unless such copies or derivative
+ * works are solely in the form of machine-executable object code generated by
+ * a source language processor.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
+ * DEALINGS IN THE SOFTWARE.
+ */
+
+#include "ambient/numeric/matrix/matrix.hpp"
+#include "ambient/numeric/matrix/diagonal_matrix.hpp"
+#include "ambient/numeric/matrix/tiles.hpp"
+#include "ambient/numeric/matrix/utils.hpp"

### 2014-04-01 src/ambient/src/ambient/numeric/traits.hpp

@@ -1,0 +1,38 @@
+/*
+ * Ambient, License - Version 1.0 - May 3rd, 2012
+ *
+ * Permission is hereby granted, free of charge, to any person or organization
+ * obtaining a copy of the software and accompanying documentation covered by
+ * this license (the "Software") to use, reproduce, display, distribute,
+ * execute, and transmit the Software, and to prepare derivative works of the
+ * Software, and to permit third-parties to whom the Software is furnished to
+ * do so, all subject to the following:
+ *
+ * The copyright notices in the Software and this entire statement, including
+ * the above license grant, this restriction and the following disclaimer,
+ * must be included in all copies of the Software, in whole or in part, and
+ * all derivative works of the Software, unless such copies or derivative
+ * works are solely in the form of machine-executable object code generated by
+ * a source language processor.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
+ * DEALINGS IN THE SOFTWARE.
+ */
+
+#ifndef AMBIENT_NUMERIC_TRAITS
+#define AMBIENT_NUMERIC_TRAITS
+
+namespace ambient { namespace numeric { namespace traits {
+
+    template<class T> struct real_type { typedef typename real_type<typename T::value_type>::type type; };
+    template<>        struct real_type<double> { typedef double type; };
+    template<class T> struct real_type<std::complex<T> > { typedef T type; };
+
+} } }
+
+#endif

### 2014-04-01 src/ambient/src/ambient/utils/dim2.h

@@ -1,0 +1,108 @@
+/*
+ * Ambient, License - Version 1.0 - May 3rd, 2012
+ *
+ * Permission is hereby granted, free of charge, to any person or organization
+ * obtaining a copy of the software and accompanying documentation covered by
+ * this license (the "Software") to use, reproduce, display, distribute,
+ * execute, and transmit the Software, and to prepare derivative works of the
+ * Software, and to permit third-parties to whom the Software is furnished to
+ * do so, all subject to the following:
+ *
+ * The copyright notices in the Software and this entire statement, including
+ * the above license grant, this restriction and the following disclaimer,
+ * must be included in all copies of the Software, in whole or in part, and
+ * all derivative works of the Software, unless such copies or derivative
+ * works are solely in the form of machine-executable object code generated by
+ * a source language processor.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
+ * DEALINGS IN THE SOFTWARE.
+ */
+
+#ifndef AMBIENT_UTILS_DIM2
+#define AMBIENT_UTILS_DIM2
+
+namespace ambient{
+
+    class dim2 {
+    public:
+        size_t x, y;
+        dim2() 
+        : x(0), y(0) 
+        {
+        }
+
+        dim2(size_t x, size_t y) 
+        : x(x), y(y) 
+        {
+        }
+
+        size_t max() const {
+            return std::max(this->x, this->y);
+        }
+
+        size_t min() const {
+            return std::min(this->x, this->y);
+        }
+
+        size_t square() const {
+            return this->x * this->y;
+        }
+
+        dim2& operator=(int value){
+            this->x = this->y = value;
+            return *this;
+        }
+
[…]

### 2014-04-01 src/ambient/src/ambient/utils/enable_omp.hpp

@@ -1,0 +1,42 @@
+/*
+ * Ambient, License - Version 1.0 - May 3rd, 2012
+ *
+ * Permission is hereby granted, free of charge, to any person or organization
+ * obtaining a copy of the software and accompanying documentation covered by
+ * this license (the "Software") to use, reproduce, display, distribute,
+ * execute, and transmit the Software, and to prepare derivative works of the
+ * Software, and to permit third-parties to whom the Software is furnished to
+ * do so, all subject to the following:
+ *
+ * The copyright notices in the Software and this entire statement, including
+ * the above license grant, this restriction and the following disclaimer,
+ * must be included in all copies of the Software, in whole or in part, and
+ * all derivative works of the Software, unless such copies or derivative
+ * works are solely in the form of machine-executable object code generated by
+ * a source language processor.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
+ * DEALINGS IN THE SOFTWARE.
+ */
+
+#ifndef AMBIENT_UTILS_ENABLE_OMP
+#define AMBIENT_UTILS_ENABLE_OMP
+
+#include <omp.h>
+#define AMBIENT_THREADING_TAGLINE "using openmp"
+#define AMBIENT_THREAD_ID omp_get_thread_num()
+#define AMBIENT_PRAGMA(a) _Pragma( #a )
+#define AMBIENT_THREAD AMBIENT_PRAGMA(omp task untied)
+#define AMBIENT_PARALLEL_FOR(...) AMBIENT_PRAGMA(omp parallel for schedule(dynamic, 1)) for(__VA_ARGS__)
[…]

### 2014-04-01 src/ambient/src/ambient/utils/mutex.hpp

@@ -1,0 +1,87 @@
+/*
+ * Ambient, License - Version 1.0 - May 3rd, 2012
+ *
+ * Permission is hereby granted, free of charge, to any person or organization
+ * obtaining a copy of the software and accompanying documentation covered by
+ * this license (the "Software") to use, reproduce, display, distribute,
+ * execute, and transmit the Software, and to prepare derivative works of the
+ * Software, and to permit third-parties to whom the Software is furnished to
+ * do so, all subject to the following:
+ *
+ * The copyright notices in the Software and this entire statement, including
+ * the above license grant, this restriction and the following disclaimer,
+ * must be included in all copies of the Software, in whole or in part, and
+ * all derivative works of the Software, unless such copies or derivative
+ * works are solely in the form of machine-executable object code generated by
+ * a source language processor.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
+ * DEALINGS IN THE SOFTWARE.
+ */
+
+#ifndef AMBIENT_UTILS_MUTEX
+#define AMBIENT_UTILS_MUTEX
+
+namespace ambient { 
+
+    // note: classes below are dervied from boost
+    // and need to be reworked for cilk and openmp
+
+    template <typename M>
+    class guard {
+    private:
+        M& mtx;
+        guard(const guard &);
+        void operator=(const guard &);
+    public:
+        explicit guard(M& nmtx) : mtx(nmtx){ mtx.lock(); }
+        ~guard(){ mtx.unlock(); }
+    };
+
+    class mutex {
+    private:
+        pthread_mutex_t m;
+    public:
+        mutex(mutex const&) = delete;
+        mutex& operator=(mutex const&) = delete;
+
+        mutex(){
+            int const res = pthread_mutex_init(&m,NULL);
[…]

### 2014-04-01 src/ambient/src/ambient/utils/timings.hpp

@@ -1,0 +1,111 @@
+/*
+ * Ambient, License - Version 1.0 - May 3rd, 2012
+ *
+ * Permission is hereby granted, free of charge, to any person or organization
+ * obtaining a copy of the software and accompanying documentation covered by
+ * this license (the "Software") to use, reproduce, display, distribute,
+ * execute, and transmit the Software, and to prepare derivative works of the
+ * Software, and to permit third-parties to whom the Software is furnished to
+ * do so, all subject to the following:
+ *
+ * The copyright notices in the Software and this entire statement, including
+ * the above license grant, this restriction and the following disclaimer,
+ * must be included in all copies of the Software, in whole or in part, and
+ * all derivative works of the Software, unless such copies or derivative
+ * works are solely in the form of machine-executable object code generated by
+ * a source language processor.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
+ * DEALINGS IN THE SOFTWARE.
+ */
+
+#ifndef AMBIENT_UTILS_TIMINGS
+#define AMBIENT_UTILS_TIMINGS
+#include "ambient/ambient.hpp"
+#include "ambient/utils/io.hpp"
+
+namespace ambient {
+
+    void sync();
+
+    #if defined(__APPLE__) && defined(AMBIENT_OMP)
+    struct time {
+        static double start(){ return omp_get_wtime(); }
+        static double stop() { return omp_get_wtime(); }
+    };
+    #elif defined(__APPLE__)
+    struct time {
+        static double start(){ return 0.; }
+        static double stop() { return 0.; }
+    };
+    #else
+    #define BILLION 0x3B9ACA00
+    struct time {
+        time(): thread_(pthread_self()){}
[…]

### 2014-04-01 src/ambient/src/boost/numeric/bindings/lapack/computational/hgeqz.hpp

@@ -1,0 +1,524 @@
+//
+// Copyright (c) 2002--2010
+// Toon Knapen, Karl Meerbergen, Kresimir Fresl,
+// Thomas Klimpel and Rutger ter Borg
+//
+// Distributed under the Boost Software License, Version 1.0.
+// (See accompanying file LICENSE_1_0.txt or copy at
+// http://www.boost.org/LICENSE_1_0.txt)
+//
+// THIS FILE IS AUTOMATICALLY GENERATED
+// PLEASE DO NOT EDIT!
+//
+
+#ifndef BOOST_NUMERIC_BINDINGS_LAPACK_COMPUTATIONAL_HGEQZ_HPP
+#define BOOST_NUMERIC_BINDINGS_LAPACK_COMPUTATIONAL_HGEQZ_HPP
+
+#include <boost/assert.hpp>
+#include <boost/numeric/bindings/begin.hpp>
+#include <boost/numeric/bindings/detail/array.hpp>
+#include <boost/numeric/bindings/is_column_major.hpp>
+#include <boost/numeric/bindings/is_complex.hpp>
+#include <boost/numeric/bindings/is_mutable.hpp>
+#include <boost/numeric/bindings/is_real.hpp>
+#include <boost/numeric/bindings/lapack/workspace.hpp>
+#include <boost/numeric/bindings/remove_imaginary.hpp>
+#include <boost/numeric/bindings/size.hpp>
+#include <boost/numeric/bindings/stride.hpp>
+#include <boost/numeric/bindings/traits/detail/utils.hpp>
+#include <boost/numeric/bindings/value_type.hpp>
+#include <boost/static_assert.hpp>
+#include <boost/type_traits/is_same.hpp>
+#include <boost/type_traits/remove_const.hpp>
+#include <boost/utility/enable_if.hpp>
+
+//
+// The LAPACK-backend for hgeqz is the netlib-compatible backend.
+//
+#include <boost/numeric/bindings/lapack/detail/lapack.h>
+#include <boost/numeric/bindings/lapack/detail/lapack_option.hpp>
+
+namespace boost {
+namespace numeric {
+namespace bindings {
+namespace lapack {
+
+//
+// The detail namespace contains value-type-overloaded functions that
+// dispatch to the appropriate back-end LAPACK-routine.
+//
+namespace detail {
+
+//
+// Overloaded function for dispatching to
+// * netlib-compatible LAPACK backend (the default), and
+// * float value-type.
+//
[…]

### 2014-04-01 src/dmrg/applications/examples/dmrg_current/dmrg_current.cpp

@@ -1,0 +1,72 @@
+/*****************************************************************************
+ *
+ * ALPS MPS DMRG Project
+ *
+ * Copyright (C) 2013 Institute for Theoretical Physics, ETH Zurich
+ *               2011-2013 by Michele Dolfi <dolfim@phys.ethz.ch>
+ * 
+ * This software is part of the ALPS Applications, published under the ALPS
+ * Application License; you can use, redistribute it and/or modify it under
+ * the terms of the license, either version 1 or (at your option) any later
+ * version.
+ * 
+ * You should have received a copy of the ALPS Application License along with
+ * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
+ * available from http://alps.comp-phys.org/.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
+ * DEALINGS IN THE SOFTWARE.
+ *
+ *****************************************************************************/
+
+#include <iostream>
+#include <fstream>
+
+#include "dmrg/utils/DmrgParameters.h"
+
+#include "dmrg/sim/symm_factory.h"
+
+int main(int argc, char ** argv)
+{
+    try {
+        if (argc != 3)
+        {
+            maquis::cout << "Usage: <parms> <model_parms>" << std::endl;
+            exit(1);
+        }
+        
+        maquis::cout.precision(10);
+        
+        /// Loading parameters
+        std::ifstream param_file(argv[1]);
+        if (!param_file) {
+            maquis::cerr << "Could not open parameter file." << std::endl;
+            exit(1);
+        }
+        DmrgParameters parms(param_file);
+        
+        /// Loading model
+        std::ifstream model_file(argv[2]);
+        if (!model_file) {
[…]

### 2014-04-01 src/dmrg/applications/multigrid/mg_dmrg.cpp

@@ -1,0 +1,109 @@
+/*****************************************************************************
+ *
+ * ALPS MPS DMRG Project
+ *
+ * Copyright (C) 2013 Institute for Theoretical Physics, ETH Zurich
+ *               2011-2013 by Michele Dolfi <dolfim@phys.ethz.ch>
+ * 
+ * This software is part of the ALPS Applications, published under the ALPS
+ * Application License; you can use, redistribute it and/or modify it under
+ * the terms of the license, either version 1 or (at your option) any later
+ * version.
+ * 
+ * You should have received a copy of the ALPS Application License along with
+ * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
+ * available from http://alps.comp-phys.org/.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
+ * DEALINGS IN THE SOFTWARE.
+ *
+ *****************************************************************************/
+
+#include <cmath>
+#include <iterator>
+#include <iostream>
+#include <sys/stat.h>
+
+#include "utils/data_collector.hpp"
+
+#include "dmrg/utils/DmrgParameters.h"
+#include "mg_dmrg.h"
+#include "utils/timings.h"
+
+#include <boost/function.hpp>
+
+// factory
+void mg_dmrg(DmrgParameters & parms, ModelParameters & model)
+{
+    std::map<std::string, boost::function<void (DmrgParameters & p, ModelParameters & m)> > factory_map;
+    
+#ifdef HAVE_TrivialGroup
+    factory_map["none"] = run_mg_dmrg<TrivialGroup>;
+#endif
+#ifdef HAVE_U1
+    factory_map["u1"] = run_mg_dmrg<U1>;
+#endif
+#ifdef HAVE_TwoU1
[…]

### 2014-04-01 src/dmrg/framework/dmrg/block_matrix/block_matrix.hpp

@@ -1,0 +1,550 @@
+/*****************************************************************************
+ *
+ * ALPS MPS DMRG Project
+ *
+ * Copyright (C) 2013 Institute for Theoretical Physics, ETH Zurich
+ *               2011-2011 by Bela Bauer <bauerb@phys.ethz.ch>
+ * 
+ * This software is part of the ALPS Applications, published under the ALPS
+ * Application License; you can use, redistribute it and/or modify it under
+ * the terms of the license, either version 1 or (at your option) any later
+ * version.
+ * 
+ * You should have received a copy of the ALPS Application License along with
+ * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
+ * available from http://alps.comp-phys.org/.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
+ * DEALINGS IN THE SOFTWARE.
+ *
+ *****************************************************************************/
+
+#include "dmrg/block_matrix/block_matrix.h"
+
+#include "utils/function_objects.h"
+#include "utils/bindings.hpp"
+
+#include <alps/type_traits/is_complex.hpp>
+
+#include <boost/serialization/serialization.hpp>
+#include <boost/ptr_container/serialize_ptr_vector.hpp>
+
+template<class Matrix, class SymmGroup>
+block_matrix<Matrix, SymmGroup>::block_matrix() 
+{
+}
+
+template<class Matrix, class SymmGroup>
+block_matrix<Matrix, SymmGroup>::block_matrix(Index<SymmGroup> rows,
+                                              Index<SymmGroup> cols)
+: rows_(rows)
+, cols_(cols)
+{
+    assert(rows_.size() == cols_.size());
[…]

### 2014-04-01 src/dmrg/framework/dmrg/block_matrix/detail/alps.hpp

@@ -1,0 +1,63 @@
+/*****************************************************************************
+ *
+ * ALPS MPS DMRG Project
+ *
+ * Copyright (C) 2013 Institute for Theoretical Physics, ETH Zurich
+ *               2011-2012 by Michele Dolfi <dolfim@phys.ethz.ch>
+ * 
+ * This software is part of the ALPS Applications, published under the ALPS
+ * Application License; you can use, redistribute it and/or modify it under
+ * the terms of the license, either version 1 or (at your option) any later
+ * version.
+ * 
+ * You should have received a copy of the ALPS Application License along with
+ * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
+ * available from http://alps.comp-phys.org/.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
+ * DEALINGS IN THE SOFTWARE.
+ *
+ *****************************************************************************/
+
+#ifndef MAQUIS_BLOCK_MATRIX_DEATAIL_ALPS_HPP
+#define MAQUIS_BLOCK_MATRIX_DEATAIL_ALPS_HPP
+
+#include <alps/numeric/matrix.hpp>
+#include <alps/numeric/matrix/algorithms.hpp>
+#include <alps/numeric/diagonal_matrix.hpp>
+#include "dmrg/block_matrix/detail/alps_detail.hpp"
+#include "dmrg/block_matrix/detail/one_matrix.hpp"
+#include "utils/traits.hpp"
+
+namespace maquis { namespace traits {
+
+    template <typename T, typename MemoryBlock> 
[…]

### 2014-04-01 src/dmrg/framework/dmrg/block_matrix/detail/alps_detail_inline.hpp

@@ -1,0 +1,247 @@
+/*****************************************************************************
+ *
+ * MAQUIS DMRG Project
+ *
+ * Copyright (C) 2011-2012 by Bela Bauer <bauerb@phys.ethz.ch>
+ *                            Michele Dolfi <dolfim@phys.ethz.ch>
+ *                            Timothee Ewart <timothee.ewart@gmail.com>
+ *                            Alexandr Kosenkov <alex.kosenkov@gmail.com>
+ *
+ *****************************************************************************/
+
+#ifndef MAQUIS_BLOCK_MATRIX_DEATAIL_ALPS_MATRIX_DETAIL_HPP
+#define MAQUIS_BLOCK_MATRIX_DEATAIL_ALPS_MATRIX_DETAIL_HPP
+
+#include <x86intrin.h>
+#include <immintrin.h>
+#include <alps/numeric/is_zero.hpp>
+
+template<class T, class SymmGroup>
+class block_matrix;
+
+namespace maquis { namespace dmrg { namespace detail {
+        
+    template<class InputIterator, class OutputIterator, class T>
+    void iterator_axpy(InputIterator in1, InputIterator in2,
+                       OutputIterator out1, T val)
+    {
+        std::transform(in1, in2, out1, out1, boost::lambda::_1*val+boost::lambda::_2);
+    }
+    
+    inline void iterator_axpy(double const * in1, double const * in2,
+                              double * out1, double val)
+    {
+        fortran_int_t one = 1, diff = in2-in1;
+#if defined(BIND_FORTRAN_LOWERCASE) || defined(__xlC__)
+        daxpy(&diff, &val, in1, &one, out1, &one);
+#else
+        daxpy_(&diff, &val, in1, &one, out1, &one);
+        /*
+        if ((std::size_t)in1%32 != 0) {
+            std::cout << "in1 not aligned\n";
+            exit(1);
+        }
+        if ((std::size_t)out1%32 != 0) {
+            std::cout << "out1 not aligned\n";
+            exit(1);
+        }
+        fortran_int_t ndiv4 = diff/4;
+        __m256d alfa = _mm256_broadcast_sd(&val);
+        for(fortran_int_t n=0; n < ndiv4; ++n)
+        {
+            __m256d x = _mm256_load_pd(in1 + 4*n);
+                    x = _mm256_mul_pd(alfa, x);
+
+            __m256d y = _mm256_load_pd(out1 + 4*n);
+
[…]

### 2014-04-01 src/dmrg/framework/dmrg/block_matrix/grouped_symmetry.h

@@ -1,0 +1,78 @@
+/*****************************************************************************
+ *
+ * ALPS MPS DMRG Project
+ *
+ * Copyright (C) 2013 Institute for Theoretical Physics, ETH Zurich
+ *               2013-2013 by Michele Dolfi <dolfim@phys.ethz.ch>
+ * 
+ * This software is part of the ALPS Applications, published under the ALPS
+ * Application License; you can use, redistribute it and/or modify it under
+ * the terms of the license, either version 1 or (at your option) any later
+ * version.
+ * 
+ * You should have received a copy of the ALPS Application License along with
+ * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
+ * available from http://alps.comp-phys.org/.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
+ * DEALINGS IN THE SOFTWARE.
+ *
+ *****************************************************************************/
+
+#ifndef MAQUIS_DMRG_BLOCK_MATRIX_GROUPED_SYMMETRY_H
+#define MAQUIS_DMRG_BLOCK_MATRIX_GROUPED_SYMMETRY_H
+
+#include "dmrg/block_matrix/symmetry.h"
+#include "dmrg/block_matrix/indexing.h"
+
+//// TRAITS
+
+template <class SymmGroup>
+struct grouped_symmetry;
+
+template <>
+struct grouped_symmetry<TrivialGroup> {
+    typedef TrivialGroup type;
+};
+
+template <>
+struct grouped_symmetry<U1> {
+    typedef TwoU1 type;
+};
+
+//// GROUPPING FUNCTIONS
+
+inline TrivialGroup::charge group(TrivialGroup::charge c1, TrivialGroup::charge c2)
+{
[…]

### 2014-04-01 src/dmrg/framework/dmrg/models/coded/factory_lattice.hpp

@@ -1,0 +1,53 @@
+/*****************************************************************************
+ *
+ * ALPS MPS DMRG Project
+ *
+ * Copyright (C) 2013 Institute for Theoretical Physics, ETH Zurich
+ *               2011-2011 by Michele Dolfi <dolfim@phys.ethz.ch>
+ * 
+ * This software is part of the ALPS Applications, published under the ALPS
+ * Application License; you can use, redistribute it and/or modify it under
+ * the terms of the license, either version 1 or (at your option) any later
+ * version.
+ * 
+ * You should have received a copy of the ALPS Application License along with
+ * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
+ * available from http://alps.comp-phys.org/.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
+ * DEALINGS IN THE SOFTWARE.
+ *
+ *****************************************************************************/
+
+#ifndef MAQUIS_DMRG_MODELS_CODED_FACTORY_LATTICE_H
+#define MAQUIS_DMRG_MODELS_CODED_FACTORY_LATTICE_H
+
+#include "dmrg/models/coded/lattice.hpp"
+
+inline boost::shared_ptr<lattice_impl>
+coded_lattice_factory(BaseParameters & parms)
+{
+    typedef boost::shared_ptr<lattice_impl> impl_ptr;
+    if (parms["LATTICE"] == std::string("periodic chain lattice"))
+        return impl_ptr(new ChainLattice(parms, true));
+    else if (parms["LATTICE"] == std::string("chain lattice"))
+        return impl_ptr(new ChainLattice(parms, false));
+    else if (parms["LATTICE"] == std::string("open chain lattice"))
[…]

### 2014-04-01 src/dmrg/framework/dmrg/models/continuum/factory_lattice.hpp

@@ -1,0 +1,42 @@
+/*****************************************************************************
+ *
+ * ALPS MPS DMRG Project
+ *
+ * Copyright (C) 2013 Institute for Theoretical Physics, ETH Zurich
+ *               2011-2011 by Michele Dolfi <dolfim@phys.ethz.ch>
+ * 
+ * This software is part of the ALPS Applications, published under the ALPS
+ * Application License; you can use, redistribute it and/or modify it under
+ * the terms of the license, either version 1 or (at your option) any later
+ * version.
+ * 
+ * You should have received a copy of the ALPS Application License along with
+ * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
+ * available from http://alps.comp-phys.org/.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
+ * DEALINGS IN THE SOFTWARE.
+ *
+ *****************************************************************************/
+
+#include "dmrg/models/continuum/lattice.hpp"
+
+inline boost::shared_ptr<lattice_impl> cont_lattice_factory (BaseParameters & parms)
+{
+    typedef boost::shared_ptr<lattice_impl> impl_ptr;
+    std::string lat_str = parms.is_set("lattice") ? "lattice" : "LATTICE";
+    if (parms[lat_str] == std::string("continuous_chain")
+        || parms[lat_str] == std::string("continuous_left_chain")
+        || parms[lat_str] == std::string("continuous_center_chain"))
+        return impl_ptr(new ContChain(parms, false));
+    else if (parms[lat_str] == std::string("periodic_continuous_chain"))
+        return impl_ptr(new ContChain(parms, true));
+    else {
+        throw std::runtime_error("Don't know this lattice!");
+    }
+}

### 2014-04-01 src/dmrg/framework/dmrg/models/continuum/factory_none.hpp

@@ -1,0 +1,44 @@
+/*****************************************************************************
+ *
+ * ALPS MPS DMRG Project
+ *
+ * Copyright (C) 2013 Institute for Theoretical Physics, ETH Zurich
+ *               2011-2011 by Michele Dolfi <dolfim@phys.ethz.ch>
+ * 
+ * This software is part of the ALPS Applications, published under the ALPS
+ * Application License; you can use, redistribute it and/or modify it under
+ * the terms of the license, either version 1 or (at your option) any later
+ * version.
+ * 
+ * You should have received a copy of the ALPS Application License along with
+ * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
+ * available from http://alps.comp-phys.org/.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
+ * DEALINGS IN THE SOFTWARE.
+ *
+ *****************************************************************************/
+
+#include "dmrg/models/continuum/models_none.hpp"
+
+template<class Matrix>
+struct cont_model_factory<Matrix, TrivialGroup> {
+    static typename model_traits<Matrix, TrivialGroup>::model_ptr parse
+    (Lattice const & lattice, BaseParameters & model)
+    {
+        std::string model_str = model.is_set("model") ? "model" : "MODEL";
+        if (model[model_str] == std::string("optical_lattice"))
+            return typename model_traits<Matrix, TrivialGroup>::model_ptr(
+                        new OpticalLatticeNull<Matrix>(lattice, model)
+                   );
+        else {
+            throw std::runtime_error("Don't know this model!");
+            return typename model_traits<Matrix, TrivialGroup>::model_ptr();
+        }
+    }
+};

### 2014-04-01 src/dmrg/framework/dmrg/mp_tensors/mpo.h

@@ -1,0 +1,387 @@
+/*****************************************************************************
+ *
+ * ALPS MPS DMRG Project
+ *
+ * Copyright (C) 2013 Institute for Theoretical Physics, ETH Zurich
+ *               2011-2011 by Bela Bauer <bauerb@phys.ethz.ch>
+ * 
+ * This software is part of the ALPS Applications, published under the ALPS
+ * Application License; you can use, redistribute it and/or modify it under
+ * the terms of the license, either version 1 or (at your option) any later
+ * version.
+ * 
+ * You should have received a copy of the ALPS Application License along with
+ * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
+ * available from http://alps.comp-phys.org/.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
+ * DEALINGS IN THE SOFTWARE.
+ *
+ *****************************************************************************/
+ 
+#ifndef MPO_H
+#define MPO_H
+
+#include <vector>
+#include <set>
+
+#include "dmrg/mp_tensors/mpotensor.h"
+
+template<class Matrix, class SymmGroup>
+class MPO : public std::vector<MPOTensor<Matrix, SymmGroup> >
+{
+public:
+    typedef MPOTensor<Matrix, SymmGroup> elem_type;
+
+    MPO() { }
+    
+    MPO(std::size_t L, elem_type elem = elem_type())
+    : std::vector<elem_type>(L, elem)
+    { }
+    
+    std::size_t length() const { return this->size(); }
+    
+    void compress(double cutoff)
+    {
+        calc_charges();
+        
+        for (int p = 0; p < this->size()-1; ++p) {
[…]

### 2014-04-01 src/dmrg/framework/dmrg/mp_tensors/mps_initializers.h

@@ -1,0 +1,555 @@
+/*****************************************************************************
+ *
+ * ALPS MPS DMRG Project
+ *
+ * Copyright (C) 2013 Institute for Theoretical Physics, ETH Zurich
+ *               2011-2013 by Bela Bauer <bauerb@phys.ethz.ch>
+ *                            Michele Dolfi <dolfim@phys.ethz.ch>
+ * 
+ * This software is part of the ALPS Applications, published under the ALPS
+ * Application License; you can use, redistribute it and/or modify it under
+ * the terms of the license, either version 1 or (at your option) any later
+ * version.
+ * 
+ * You should have received a copy of the ALPS Application License along with
+ * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
+ * available from http://alps.comp-phys.org/.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
+ * DEALINGS IN THE SOFTWARE.
+ *
+ *****************************************************************************/
+
+#ifndef MPS_INITIALIZER_H
+#define MPS_INITIALIZER_H
+
+#include <fstream>
+#include <sstream>
+#include <algorithm>
+
+#include <boost/tokenizer.hpp>
+
+#include "dmrg/utils/DmrgParameters.h"
+#include "dmrg/mp_tensors/mps_sectors.h"
+#include "dmrg/mp_tensors/compression.h"
+#include "dmrg/mp_tensors/state_mps.h"
+#include "dmrg/models/chem/pg_util.h"
+
+#include "dmrg/mp_tensors/mps.h"
+#include "dmrg/mp_tensors/mps_mpo_ops.h"
+
+template<class Matrix, class SymmGroup>
[…]

### 2014-04-01 src/dmrg/framework/dmrg/mp_tensors/mps_mpo_ops.h

@@ -1,0 +1,351 @@
+/*****************************************************************************
+ *
+ * ALPS MPS DMRG Project
+ *
+ * Copyright (C) 2013 Institute for Theoretical Physics, ETH Zurich
+ *               2011-2011 by Bela Bauer <bauerb@phys.ethz.ch>
+ * 
+ * This software is part of the ALPS Applications, published under the ALPS
+ * Application License; you can use, redistribute it and/or modify it under
+ * the terms of the license, either version 1 or (at your option) any later
+ * version.
+ * 
+ * You should have received a copy of the ALPS Application License along with
+ * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
+ * available from http://alps.comp-phys.org/.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
+ * DEALINGS IN THE SOFTWARE.
+ *
+ *****************************************************************************/
+
+#ifndef MPS_MPO_OPS_H
+#define MPS_MPO_OPS_H
+
+#include "dmrg/mp_tensors/mps.h"
+#include "dmrg/mp_tensors/mpo.h"
+
+#include "dmrg/mp_tensors/special_mpos.h"
+#include "dmrg/mp_tensors/contractions.h"
+
+#include "dmrg/utils/utils.hpp"
+#include "utils/traits.hpp"
+
+template<class Matrix, class SymmGroup>
+std::vector<Boundary<Matrix, SymmGroup> >
+left_mpo_overlaps(MPS<Matrix, SymmGroup> const & mps, MPO<Matrix, SymmGroup> const & mpo)
+{
+    assert(mpo.length() == mps.length());
+    std::size_t L = mps.length();
[…]

### 2014-04-01 src/dmrg/framework/dmrg/mp_tensors/ts_ops.h

@@ -1,0 +1,220 @@
+/*****************************************************************************
+ *
+ * ALPS MPS DMRG Project
+ *
+ * Copyright (C) 2013 Institute for Theoretical Physics, ETH Zurich
+ *               2013-2013 by Sebastian Keller <sebkelle@phys.ethz.ch>
+ *
+ * 
+ * This software is part of the ALPS Applications, published under the ALPS
+ * Application License; you can use, redistribute it and/or modify it under
+ * the terms of the license, either version 1 or (at your option) any later
+ * version.
+ * 
+ * You should have received a copy of the ALPS Application License along with
+ * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
+ * available from http://alps.comp-phys.org/.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
+ * DEALINGS IN THE SOFTWARE.
+ *
+ *****************************************************************************/
+
+#ifndef MAQUIS_DMRG_TS_OPS_H
+#define MAQUIS_DMRG_TS_OPS_H
+
+#include "dmrg/mp_tensors/mpstensor.h"
+#include "dmrg/mp_tensors/mpotensor.h"
+#include "dmrg/block_matrix/indexing.h"
+#include "dmrg/block_matrix/multi_index.h"
+#include "dmrg/block_matrix/block_matrix.h"
+#include "dmrg/block_matrix/block_matrix_algorithms.h"
+
+template<class MPOMatrix, class MPSMatrix, class SymmGroup>
+MPOTensor<MPSMatrix, SymmGroup> make_twosite_mpo(MPOTensor<MPOMatrix, SymmGroup> const & mpo1,
+                                                 MPOTensor<MPOMatrix, SymmGroup> const & mpo2,
[…]

### 2014-04-01 src/dmrg/framework/dmrg/utils/archive.h

@@ -1,0 +1,97 @@
+/*****************************************************************************
+ *
+ * ALPS MPS DMRG Project
+ *
+ * Copyright (C) 2013 Institute for Theoretical Physics, ETH Zurich
+ *               2011-2011 by Bela Bauer <bauerb@phys.ethz.ch>
+ *               2011-2012 by Michele Dolfi <dolfim@phys.ethz.ch>
+ * 
+ * This software is part of the ALPS Applications, published under the ALPS
+ * Application License; you can use, redistribute it and/or modify it under
+ * the terms of the license, either version 1 or (at your option) any later
+ * version.
+ * 
+ * You should have received a copy of the ALPS Application License along with
+ * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
+ * available from http://alps.comp-phys.org/.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
+ * DEALINGS IN THE SOFTWARE.
+ *
+ *****************************************************************************/
+
+#ifndef STORAGE_ARCHIVE_H
+#define STORAGE_ARCHIVE_H
+
+#ifdef USE_AMBIENT
+#include "ambient/ambient.hpp"
+#endif
+
+#include <boost/utility.hpp>
+#include <alps/hdf5.hpp>
+#include <alps/utility/encode.hpp>
+
+namespace storage {
+
+    inline std::string once(std::string fp){
+        #ifdef USE_AMBIENT
+        if(!ambient::master() && !ambient::ctxt.scoped()) return fp+"."+std::to_string(ambient::rank());
+        #endif
+        return fp;
+    }
+
+    inline void uniq(std::string fp){
+        #ifdef USE_AMBIENT
[…]

### 2014-04-01 src/dmrg/framework/dmrg/utils/time_limit_exception.h

@@ -1,0 +1,58 @@
+/*****************************************************************************
+ *
+ * ALPS MPS DMRG Project
+ *
+ * Copyright (C) 2013 Institute for Theoretical Physics, ETH Zurich
+ *               2013-2013 by Michele Dolfi <dolfim@phys.ethz.ch>
+ * 
+ * This software is part of the ALPS Applications, published under the ALPS
+ * Application License; you can use, redistribute it and/or modify it under
+ * the terms of the license, either version 1 or (at your option) any later
+ * version.
+ * 
+ * You should have received a copy of the ALPS Application License along with
+ * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
+ * available from http://alps.comp-phys.org/.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
+ * DEALINGS IN THE SOFTWARE.
+ *
+ *****************************************************************************/
+
+#ifndef MAQUIS_DMRG__UTILS_TIME_LIMIT_EXCEPTION_H
+#define MAQUIS_DMRG__UTILS_TIME_LIMIT_EXCEPTION_H
+
+#include <string>
+#include <boost/lexical_cast.hpp>
+#include <stdexcept>
+
+namespace dmrg {
+    class time_limit : public std::runtime_error {
+    public:
+        time_limit() : std::runtime_error("time limit reached") {}
+        
+        time_limit(int sw, int st)
+        : std::runtime_error(  std::string("time limit reached. current status is [ ")
+                             + std::string("sweep=") + boost::lexical_cast<std::string>(sw) + std::string(", ")
+                             + std::string("site=" ) + boost::lexical_cast<std::string>(st) + std::string(" ]."))
+        , sweep_(sw)
[…]

### 2014-04-01 src/dmrg/lib/python/pydmrg/load.py

@@ -1,0 +1,120 @@
+#############################################################################
+#
+# MAQUIS DMRG Project
+#
+# Copyright (C) 2013-2013 by Michele Dolfi <dolfim@phys.ethz.ch>
+#
+#############################################################################
+
+import pyalps
+from pyalps.load import *
+try:
+    from pyalps.ngs import archive
+except:
+    from pyalps.hdf5 import archive
+
+from copy import deepcopy
+
+def ReadMeasurements(ar, measurements, path, props):
+    ret = []
+    for meas in measurements:
+        if meas in ar.list_children(path):
+            try:
+                d = pyalps.DataSet()
+                d.props = deepcopy(props)
+                d.props['observable'] = meas
+                d.props['hdf5_path'] = path+'/'+meas
+                d.y = ar[ d.props['hdf5_path']+'/mean/value' ]
+                if 'labels' in ar.list_children(d.props['hdf5_path']):
+                    d.x = parse_labels(ar[ d.props['hdf5_path']+'/labels' ])
+                else:
+                    d.x = range(len(d.y))
+                ret.append(d)
+            except Exception, e:
+                print "Could not load " + meas
+                print e
+                pass
+    return ret
+
+def ReadDMRGSweeps(ar, measurements, props, path='/simulation', selector=None):
+    ret = []
+    for sweep_name in ar.list_children(path):
+        if sweep_name == 'iteration':  ## new iterarion format
+            for sweep_num in ar.list_children(path+'/iteration'):
+                sweep = int(sweep_num)
+                ret_sweep = []
+                p_sweep = path+'/iteration/'+sweep_num
+                props_sweep = deepcopy(props)
+                props_sweep['sweep'] = sweep
+                if 'parameters' in ar.list_children(p_sweep):
+                    props_sweep.update( dict(ar[p_sweep+'/parameters']) )
+                if selector==None or selector(props_sweep):
+                    ret_sweep += ReadMeasurements(ar, measurements, p_sweep+'/results', props_sweep)
+                if len(ret_sweep) > 0:
+                    ret.append(ret_sweep)
+        elif 'sweep' in sweep_name:  ## old iterarion format
+            sweep = eval(sweep_name.strip('sweep'))
+            ret_sweep = []
+            p_sweep = path+'/'+sweep_name
+            props_sweep = deepcopy(props)
+            props_sweep['sweep'] = sweep
+            if 'parameters' in ar.list_children(p_sweep):
+                props_sweep.update( dict(ar[p_sweep+'/parameters']) )
[…]

### 2014-04-01 src/dmrg/regression/applications/dmrg/test_ff_ss.py

@@ -1,0 +1,64 @@
+#!/usr/bin/env alpspython
+
+from pydmrg import apptest
+import sys, os
+
+testname       = os.path.splitext( os.path.basename(sys.argv[0]) )[0]
+reference_dir  = os.path.join( os.path.dirname(os.path.abspath(__file__)), 'ref/' )
+reference_file = os.path.join(reference_dir, testname+'.h5')
+
+class mytest(apptest.DMRGTestBase):
+    testname = testname
+    reference_file = reference_file
+    
+    inputs   = {
+                'parms': {
+                            'nsweeps'                    : 3,
+                            'nmainsweeps'                : 1,
+                            'ngrowsweeps'                : 1,
+                            
+                            'max_bond_dimension'         : 200,
+                            
+                            'truncation_initial'         : 0.0001,
+                            'truncation_final'           : 1e-12,
+                            
+                            'ietl_jcd_maxiter'           : 6,
+                            
+                            'alpha_initial'              : 0.001,
+                            'alpha_main'                 : 1e-6,
+                            'alpha_final'                : 0,
+                            
+                            'resultfile'                 : testname+'.out.h5',
+                            'chkpfile'                   : testname+'.out.ckp.h5',
+                            
+                            'optimization'               : 'singlesite',
+                          },
+                'model': {
+                            'LATTICE'                   : 'open square lattice',
+                            'L'                         : 5,
+                            'W'                         : 5,
+
+                            'MODEL'                     : 'spinless fermions',
+                            't'                         : 1,
+
+                            'CONSERVED_QUANTUMNUMBERS'  : 'N',
+                            'N_total'                   : 10,
+                            
+                            'MEASURE_LOCAL[Local density]' : 'n',
+                            'ENABLE_MEASURE[Entropy]'      : 1,
+                          },
+                  }
+    observables = [
+                    apptest.observable_test.reference_file('Energy',        file=reference_file),
+                    apptest.observable_test.reference_file('Entropy',       file=reference_file),
+                    apptest.observable_test.reference_file('Local density', file=reference_file),
+
+                    apptest.observable_test.reference_file('Energy',         file=os.path.join(reference_dir, 'test_ff.diag.h5'),
+                                                            tolerance=1e-4),
+                    apptest.observable_test.reference_file('Local density',  file=os.path.join(reference_dir, 'test_ff.diag.h5'),
+                                                            tolerance=1e-4),
+                  ]
+
+
+if __name__ == '__main__':
+    apptest.main()

### 2014-04-01 src/dmrg/regression/applications/dmrg_tevol/test_pure_bosons_diss_dm_none.py

@@ -1,0 +1,89 @@
+#!/usr/bin/env alpspython
+
+import sys
+
+from pydmrg import apptest
+import sys, os
+
+testname       = os.path.splitext( os.path.basename(sys.argv[0]) )[0]
+reference_dir  = os.path.join( os.path.dirname(os.path.abspath(__file__)), 'ref/' )
+reference_file = os.path.join(reference_dir, testname+'.h5')
+
+parms = {
+            'nsweeps_img'                : 0,
+            'nsweeps'                    : 50,
+            
+            'max_bond_dimension'         : 100,
+            
+            'truncation_final'           : 1e-10,
+            
+            'dt'                         : 0.1,
+            'te_type'                    : 'nn',
+            'te_optim'                   : 1,
+            'expm_method'                : 'geev',
+            
+            'measure_each'               : 10,
+            'chkp_each'                  : 50,
+            
+            'resultfile'                 : testname+'.out.h5',
+            'chkpfile'                   : testname+'.out.ckp.h5',
+            
+            'init_state'                 : 'basis_state_dm',
+            'init_basis_state'           : '0,1,0',
+            
+            'always_measure'             : 'Local density',
+            
+            'symmetry'                   : 'none',
+            'model_library'              : 'coded',
+            
+            'COMPLEX'                    : 1,
+        }
+
+model = {
+            'LATTICE'                   : 'open chain lattice',
+            'L'                         : 3,
+            
+            'MODEL'                     : 'super boson Hubbard',
+            'Nmax'                      : 3,
+            't'                         : 1,
+            'U'                         : 1,
+            'Gamma1a'                   : 0.2,
+            
+            'ENABLE_MEASURE[Local density]' : 1,
+        }
+
+class mytest_2nd(apptest.DMRGTestBase):
+    testname = testname + '_2nd'
+    reference_file = os.path.join(reference_dir, testname+'.h5')
+    
+    inputs   = {
+                'parms': dict( parms.items() + {'te_order': 'second'}.items() ),
+                'model': model,
+                  }
+    observables = [
+                    apptest.observable_test.reference_file('Local density', file=reference_file,
+                                                            load_type = 'iterations'),
+                                                            
+                    apptest.observable_test.reference_file('Local density', load_type = 'iterations', tolerance=0.01,
+                                                           file=os.path.join(reference_dir, 'pure_bosons_diss.diag.h5')),
+                  ]
+
+class mytest_4th(apptest.DMRGTestBase):
+    testname = testname + '_4th'
[…]

### 2014-04-01 src/dmrg/regression/mp_tensors/dmrg.cpp

@@ -1,0 +1,36 @@
+#include <cmath>
+#include <iterator>
+#include <iostream>
+
+#include "dmrg/block_matrix/detail/alps.hpp"
+typedef alps::numeric::matrix<double> Matrix;
+
+#include "dmrg/block_matrix/indexing.h"
+#include "dmrg/mp_tensors/mps.h"
+#include "dmrg/mp_tensors/mpo.h"
+#include "dmrg/mp_tensors/contractions.h"
+#include "dmrg/mp_tensors/mps_mpo_ops.h"
+
+#include "dmrg/mp_tensors/special_mpos.h"
+
+#include "dmrg/optimize/optimize.h"
+
+typedef TrivialGroup grp;
+
+typedef std::vector<MPOTensor<Matrix, grp> > mpo_t;
+typedef Boundary<Matrix, grp> boundary_t;
+
+int main()
+{
+    Index<grp> phys; phys.insert(std::make_pair(TrivialGroup::Plus, 2));
+    
+    int L = 32, M = 10;
+    MPS<Matrix, grp> mps(L, M, phys);
+    
+    MPOTensor<Matrix, grp> id_mpo = identity_mpo<Matrix>(mps[0].site_dim());
+    MPOTensor<Matrix, grp> sz_mpo = s12_sz_mpo<Matrix>(mps[0].site_dim());
+    
+    MPO<Matrix, grp> szsz = s12_ising<Matrix>(L, -1, 1);
+//    ss_optimize<Matrix, grp>(mps, MPO<Matrix, grp>(L, id_mpo));
+    ss_optimize<Matrix, grp>(mps, szsz, 2, 1e-14, 1000);
+}

### 2014-04-01 src/dmrg/regression/tests/mp_tensors/apply_op.cpp

@@ -1,0 +1,396 @@
+/*****************************************************************************
+ *
+ * ALPS MPS DMRG Project
+ *
+ * Copyright (C) 2013 Institute for Theoretical Physics, ETH Zurich
+ *               2011-2013 by Michele Dolfi <dolfim@phys.ethz.ch>
+ * 
+ * This software is part of the ALPS Applications, published under the ALPS
+ * Application License; you can use, redistribute it and/or modify it under
+ * the terms of the license, either version 1 or (at your option) any later
+ * version.
+ * 
+ * You should have received a copy of the ALPS Application License along with
+ * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
+ * available from http://alps.comp-phys.org/.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
+ * DEALINGS IN THE SOFTWARE.
+ *
+ *****************************************************************************/
+
+#define BOOST_TEST_MAIN
+
+#include <boost/test/included/unit_test.hpp>
+#include <boost/test/floating_point_comparison.hpp>
+#include <boost/mpl/list.hpp>
+
+#include <iterator>
+#include <iostream>
+
+
+using std::cerr;
+using std::cout;
+using std::endl;
+
+#include "dmrg/block_matrix/detail/alps.hpp"
+
+#include "dmrg/utils/DmrgParameters.h"
+
+#include "dmrg/block_matrix/indexing.h"
+#include "dmrg/mp_tensors/mps.h"
+#include "dmrg/mp_tensors/mps_initializers.h"
+#include "dmrg/mp_tensors/mps_mpo_ops.h"
+#include "dmrg/models/generate_mpo.hpp"
[…]

### 2014-04-01 src/dmrg/regression/tests/qrdecomp/qrdecomp.cpp

@@ -1,0 +1,196 @@
+/*****************************************************************************
+ *
+ * ALPS MPS DMRG Project
+ *
+ * Copyright (C) 2013 Institute for Theoretical Physics, ETH Zurich
+ *               2011-2013 by Michele Dolfi <dolfim@phys.ethz.ch>
+ * 
+ * This software is part of the ALPS Applications, published under the ALPS
+ * Application License; you can use, redistribute it and/or modify it under
+ * the terms of the license, either version 1 or (at your option) any later
+ * version.
+ * 
+ * You should have received a copy of the ALPS Application License along with
+ * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
+ * available from http://alps.comp-phys.org/.
+ *
+ * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
+ * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
+ * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
+ * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
+ * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
+ * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
+ * DEALINGS IN THE SOFTWARE.
+ *
+ *****************************************************************************/
+
+#define BOOST_TEST_MAIN
+
+#include <boost/test/included/unit_test.hpp>
+#include <boost/test/floating_point_comparison.hpp>
+
+#include <iterator>
+#include <iostream>
+
+
+using std::cerr;
+using std::cout;
+using std::endl;
+
+#include "dmrg/block_matrix/detail/alps.hpp"
+
+#include "dmrg/utils/DmrgParameters.h"
+
+#include "dmrg/block_matrix/indexing.h"
+#include "dmrg/mp_tensors/mps.h"
+#include "dmrg/mp_tensors/mps_initializers.h"
+#include "dmrg/mp_tensors/mps_mpo_ops.h"
+#include "dmrg/models/generate_mpo.hpp"
[…]

### 2020-09-23 dmrg/doc/qcmaquis_manual.tex
added QCMaquis user manual with some updates

\documentclass[bibliography=totoc,12pt,a4paper]{scrartcl}

\usepackage[nott]{kpfonts}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
% \usepackage[libertine]{newtxmath}
\usepackage{graphicx}% Include figure files
\usepackage{xcolor}
\usepackage{bm}% bold math
\usepackage[colorlinks=true]{hyperref} % web links
\definecolor{linkcolour}{rgb}{0,0.2,0.6}
\definecolor{citecolour}{rgb}{0,0.8,0.2}
\hypersetup{colorlinks,breaklinks,urlcolor=linkcolour,linkcolor=linkcolour,citecolor=citecolour}
\usepackage{framed}
\usepackage{longtable}
\usepackage[bf,singlelinecheck=off]{caption}
\usepackage{amsmath}
\usepackage{listings}
%\lstset{breaklines=true}
\usepackage{multirow}
% \usepackage[top=2.25cm, bottom=2.25cm, left=2cm, right=2.25cm]{geometry}
\usepackage{booktabs}
\usepackage{chemformula}
\usepackage{placeins}
\usepackage[numbers,sort&compress]{natbib}
\setcounter{secnumdepth}{4}
% \renewcommand{\baselinestretch}{1.5}
\newcommand{\relv}{release version 3.0}
\newcommand{\mol}{\textsc{OpenMolcas}}
\newcommand{\qcm}{\textsc{QCMaquis}}
\newcommand{\scine}{SCINE}
\newcommand{\hostp}{\mol}
\newcommand{\molbuild}{\texttt{my-Molcas-build}}
\newcommand{\molsrc}{\texttt{my-Molcas-src}}
\newcommand{\qcmsrc}{\texttt{my-QCMaquis-src}}
\newcommand{\qcmbuild}{\texttt{my-QCMaquis-build}}
%\setlength{\parindent}{1.0cm}
%\setlength{\parskip}{1.00ex plus0.5ex minus0.5ex}
% \setlength{\abovecaptionskip}{2.00ex plus0.25ex minus0.25ex}
% \setlength{\belowcaptionskip}{\abovecaptionskip}
%\setlength{\textfloatsep}{3.0ex plus0.5ex minus0.5ex}
%\setlength{\intextsep}{3.0ex plus0.5ex minus0.5ex}

\newcommand{\tableoptionskip}{0.3em}
\newcommand{\myemph}[1]{\textbf{#1}}

% \usepackage{bibentry}
%% required for the bibentry package

\lstdefinelanguage{molcas}
{
keywordsprefix=\&,alsoletter=\&,%
keywords=[1]{GATEWAY,SEWARD,SCF,DMRGSCF,MPSSI},
morekeywords=[2]{ActiveSpaceOptimizer,DMRGSettings,EndDMRGSettings,OOptimizationSettings,EndOOptimizationSettings,NEVPT2Prep,EvRDM,HDF5,DMRG,EJOB,SPIN,AMFI,Angmom},
keywordstyle=[1]\color{green!20!red},
keywordstyle=[2]\color{green!20!blue},
sensitive=false,
morecomment=[l]{*},
morecomment=[s]{/*}{*/},
}

\lstdefinelanguage{qcmaquis}
{
%
keywords=[1]{L,nelec,spin,irrep,nsweeps,max_bond_dimension,integral_file,chkpfile,resultfile,init_state,LATTICE,lattice_library,MODEL,model_library},
morekeywords=[2]{conv_thresh,integral_cutoff,truncation_initial,truncation_final,symmetry,MEASURE},
keywordstyle=[1]\color{green!20!red},
keywordstyle=[2]\color{green!20!blue},
commentstyle=\color{gray},
sensitive=false,
morecomment=[l]{//},
morecomment=[s]{/*}{*/},
}

\lstset{numbers=left,numberstyle=\tiny,numberblanklines=true,basicstyle=\ttfamily\small,frame=single,showlines=true,flexiblecolumns=true,breaklines=true,postbreak={\llap{\textcolor{red}{$\lhook\joinrel\longrightarrow$\kern2.1em}}},escapechar=|}

\newcommand{\todo}[1]{\textcolor{red}{#1}}
\newcommand{\kwd}[1]{\texttt{#1}}

\title{A quick user guide to the\\ \scine-\qcm\ software suite for \mol}
\author{Leon Freitag, Sebastian Keller, Stefan Knecht, Yingjin Ma,\\ Christopher Stein and Markus Reiher}
\publishers{ETH Z\"urich, Laboratorium f\"ur Physikalische Chemie, Vladimir-Prelog-Weg 2,\\ CH-8093 Z\"urich}
\date{\today}

\begin{document}

\bibliographystyle{achemso}
% \nobibliography{qcmaquis_manual}

\pagenumbering{roman}
\maketitle
\thispagestyle{empty}
\begin{center}
\end{center}
\vspace{-1.85cm}
\centerline{\large{\relv}}

\vspace{2cm}
\noindent We kindly request that, for reproducibility reasons, any use of the \qcm\ software suite for  density matrix renormalization group (DMRG) calculations in \mol\ that results in published material should cite the set-up steered by settings and warm-up procedures described in:
\begin{framed}
		\noindent {\textbf{Check for a preprint on \href{http://arxiv.org/}{arXiv.org} (to appear soon).}}\\
	\noindent Freitag~L.; Keller~S.; Knecht~S.; Lindh~R.; Ma~Y.; Stein~C. J.; Reiher~M. \emph{in preparation}.
\end{framed}

The DMRG calculations are then conducted with the software \qcm\ that requires a citation. It is described in the following paper:
\begin{framed}
%\noindent {\textbf{Check for the journal article asap on \href{http://arxiv.org/pdf/1510.02026.pdf}{arXiv.org}.}}\\
\noindent Keller,~S.; Dolfi,~M.; Troyer,~M.; Reiher,~M. \emph{J. Chem. Phys.}
  \textbf{2015}, \emph{143}, 244118, \href{https://doi.org/10.1063/1.4939000}{doi:10.1063/1.4939000}.
\end{framed}

\qcm\ builds upon the ALPS MPS project. The \href{http://alps.comp-phys.org/static/mps\_doc/index.html}{ALPS MPS codes} implement the DMRG algorithm for variational ground and low-lying excited state search as well as time evolution of arbitrary one- and two-dimensional models in a matrix-product-state representation. They have been developed at ETH Zurich by Michele Dolfi and Bela Bauer in the group of Matthias Troyer with contributions from Sebastian Keller and Alexandr Kosenkov and at the University of Geneva by Timoth{\'e}e Ewart and Adrian Kantian in the group of Thierry Giamarchi.

\begin{framed}
\noindent \textbf{For further information on the ALPS project, please visit \href{alps.comp-phys.org}{alps.comp-phys.org}.}\\[2ex]
\textbf{Refer to the original ALPS MPS paper:}\\
M. Dolfi, B. Bauer, S. Keller, A. Kosenkov, T. Ewart, A. Kantian, T. Giamarchi, M. Troyer, \textit{Comp. Phys. Commun.} \textbf{2014}, 12, 3430. \href{https://doi.org/10.1016/j.cpc.2014.08.019}{doi:10.1016/j.cpc.2014.08.019}
\end{framed}

ALPS is a general open-source framework for the description of strongly correlated many-particle systems.
\begin{framed}
B. Bauer, \textit{et al.} (ALPS Collaboration), The ALPS project release 2.0: open source software for strongly correlated systems, \textit{J. Stat. Mech.} \textbf{2011} P05001. \href{http://dx.doi.org/10.1088/1742-5468/2011/05/P05001}{http://dx.doi.org/10.1088/1742-5468/2011/05/P05001}.
\end{framed}

%The NEVPT2 program builds on the NEVPT2 code originally authored by Celestino Angeli, Renzo Cimiraglia and Jean-Paul Malrieu.
%\begin{framed}
% \noindent Cite the following papers when performing NEVPT2 calculations within the \mol{} software suite:
%%  \bibentry{Freitag_JChemTheoryComput_Multireference_2017}
%
%Freitag,~L.; Knecht,~S.; Angeli,~C.; Reiher,~M. \emph{J. Chem. Theory Comput.}
%  \textbf{2017}, \emph{13}, 451--459, \href{https://doi.org/10.1021/acs.jctc.6b00778}{doi:10.1021/acs.jctc.6b00778}
%
%Angeli,~C.; Cimiraglia,~R.; Malrieu,~J.-P. \emph{J. Chem. Phys.} \textbf{2002},
%  \emph{117}, 9138--9153, \href{https://doi.org/10.1063/1.1515317}{doi:10.1063/1.1515317}
%%  \bibentry{Angeli_JChemPhys_nelectron_2002}
%\end{framed}
%
%The MPS-SI program exploits the framework of the CASSI code in \hostp.
%\begin{framed}
% \noindent Cite the following papers when performing MPS-SI calculations within the \mol{} software suite:
%
% Knecht,~S.;~Keller~S.;~Autschbach~J.;~Reiher~M., \emph{J.~Chem.~Theory~Comput.} \textbf{2016}, \emph{12} 5881--5894, \href{http://dx.doi.org/10.1021/acs.jctc.6b00889}{http://dx.doi.org/10.1021/acs.jctc.6b00889}
%
% Malmqvist,~P.-{\AA}.;~Roos,~B.~O., \emph{Chem. Phys. Lett.} \textbf{1989},
%   \emph{155}, 189--194
%\end{framed}

\clearpage

\tableofcontents
\clearpage

\pagenumbering{arabic}
\section{Introduction to the \qcm\ Software Suite}

\subsection{Overview and Goals}
The computational cost of traditional full CI and full CI-based multiconfigurational calculations (such as CASSCF) scales exponentially with the number of active orbitals, thus calculations beyond a certain limit (e. g. 18 electrons in 18 orbitals for CASSCF) become prohibitively expensive. A possible approach to overcome the scaling limit is to employ methods such as density matrix renormalization group (DMRG) \cite{Schollwock_AnnPhys_densitymatrix_2011,Marti_ZPhysChem_Density_2010,Chan2011,Chan_JChemPhys_Matrix_2016} to solve the full CI problem approximately.

In DMRG, usually the wave function is represented in form of a matrix product state
(MPS). The \qcm\ software suite solves a full CI problem approximately by means of
an efficient optimization of an MPS wave function based on a second-generation DMRG algorithm \cite{Keller_JChemPhys_efficient_2015}.
The quantum-chemical operators are represented as matrix product operators (MPOs) which provides the necessary flexibility to accommodate Abelian and non-Abelian symmetries as well as the implementation of non-relativistic and relativistic quantum chemical Hamiltonians \cite{maquis-rel}, respectively,
in a unified framework. We have implemented the special unitary group of degree 2 (SU(2)) in the MPO representation of the non-relativistic Hamiltonian to ensure
spin conservation \cite{Keller_JChemPhys_Spinadapted_2016}.
The current implementation of \qcm\ allows for efficient full-CI-type calculations of active space sizes beyond capabilities (``CAS(18,18)") of standard CI approaches.

The \qcm\ software suite comes with a Fortran/Python-based interface \cite{interface} to the \mol\ \cite{Aquilante_JComputChem_Molcas_2016} framework, where we have implemented a state-specific and state-average DMRG self-consistent field (DMRG-SCF) algorithm, the possibility to include solvent effects in DMRG calculations, analytic gradients for state-specific calculations, state interaction in a nonorthogonal basis\cite{Knecht_JChemTheoryComput_nonorthogonal_2016} and n-electron valence state perturbation theory employing the DMRG reference wave function.\cite{Freitag_JChemTheoryComput_Multireference_2017} These developments allow for, among others, structure optimizations, spin-orbit coupling calculations, evaluation of transition densities between non-orthogonal states for electronic and magnetic properties, as well as to account for dynamic correlation by means of the NEVPT2 approach.

\begin{framed}
	\large{\textbf{Advice:} The break-even point wrt computational cost for a DMRG-CI/-SCF calculation compared to a ``traditional" CAS-CI/-SCF calculation is approximately reached for a CAS(14,14) space. For active spaces smaller than CAS(14,14) we recommend to choose a traditional \mol\ CAS approach.}
\end{framed}

\subsection{What features are included in this release?}\label{sec:features-v1.0}

\subsubsection{Updates in the \relv}
In the \relv, \qcm{} features:
\begin{itemize}
  \item a completely rewritten \qcm{}-\mol{} interface that does not require Python
  \item simple and straightforward to use Fortran, C and C++ API for interfacing with other quantum chemistry packages (e.\,g.\ BAGEL or Serenity)
  \item simplified installation procedure with less package dependencies
  \item Bug fix for calculation of excited states with Fiedler ordering
  \item Reworked distributed 4-RDM calculation in NEVPT2 with improved error handling
\end{itemize}

\subsubsection{Features in the \qcm{} standalone version}\label{sec:features-stand}

\begin{itemize}
	\item Optimization of spin-adapted SU(2) MPS wave functions with the DMRG algorithm (DMRG-CI calculations)
	\item Non-relativistic and scalar-relativistic quantum-chemical Hamiltonians
	\item Calculation of reduced density matrices of up to fourth order and transition reduced density matrices of up to third order
	\item Calculation of excited states
	\item A tool set to analyze the MPS wave function and its quantum entanglement
\end{itemize}

\subsubsection{Features in  \qcm\ in \mol}\label{sec:features-qcm-mol}

\begin{itemize}
	\item DMRG-CI and DMRG-SCF calculations w/wo reaction field (e.g. PCM)
	\item State-specific and state-averaged DMRG-SCF calculations
	\item Analytic gradients for state-specific DMRG-SCF calculations
  \item MPS state interaction (MPS-SI) for the calculation of spin-orbit coupling matrix elements,
  electronic and magnetic properties
  \item state-specific and quasi-degenerate (multi-state) DMRG-NEVPT2 calculations
\end{itemize}

% \subsection{Organization of this document}
%
% In the following we list and briefly summarize the remaining sections of this document.
%
% \begin{itemize}
% 	\item Section \ref{sec:install-req}, \nameref{sec:install-req},  guides through the software requirements and the registration process for \qcm.
% 	\item Section \ref{sec:installation}, \nameref{sec:installation}, guides through the installation process for \qcm\ (and possibly the host program \mol).
% 	\item Section \ref{sec:maintenance}, \nameref{sec:maintenance}, summarizes general information concerning the \qcm\ software suite including how to ask for user support, retreive future patches for the code and where to send bug reports.
% 	\item Section \ref{sec:general-consid}, \nameref{sec:general-consid}, introduces the basic workflow for \qcm\ DMRG calculations and provides details about memory requirements as well as input and output data required and generated by \qcm, respectively.
% 	\item Section \ref{sec:input}, \nameref{sec:input}, provides in the first part a list of keywords and options to control \qcm\ DMRG calculation in \mol. The second part explains in detail all mandatory and optional \qcm\ keywords and their usage. In addition we introduce several tools that are part of the \qcm\ software suite which can be used to analyze and visualize the resulting DMRG wave function(s).
% 	\item Section \ref{sec:input-examples}, \nameref{sec:input-examples}, then shows two example input files for \qcm\ DMRG calculations in \mol\ and with the \qcm\ standalone suite, respectively.
% \end{itemize}

\subsection{Licenses}\label{sec:license}
The \qcm{}-\mol{} interface and the NEVPT2 program are licensed under GNU LGPL, version 2.1.
The \qcm{} software suite and the ALPS library are distributed under \href{http://alps.comp-phys.org/static/software/applications/LICENSE.txt}{ALPS Application License version 1.0} and the \href{http://alps.comp-phys.org/static/software/ALPS/LICENSE.txt}{ALPS Library License version 1.0}.
\clearpage
\newpage
\section{Software Requirements \& Download}\label{sec:install-req}

\subsection{Prerequisites}
In order to install either \qcm\ or the quantum chemistry package \mol\ with DMRG support through the \qcm\ software suite \cite{Keller_JChemPhys_efficient_2015,interface}, requires the following libraries/programs:

\begin{itemize}
 \item Git
%  \item Python version 2.\texttt{x} with \texttt{x}\ $\geq 7$, with SciPy and NumPy
 \item HDF5 (\url{http://www.hdfgroup.org/HDF5}) with Python2 bindings (\texttt{h5py})
 \item SZIP
 \item GNU Scientific Library GSL (\url{http://www.gnu.org/software/gsl/})
 \item CMake version $\geq$3.7 (\url{https://cmake.org/})
 \item Boost version $\geq$1.56 (\url{https://github.com/boostorg/boost})
 \item Python 3.x with \texttt{h5py} package installed\\
(optional, required only for MPSSI and distributed RDM calculations for NEVPT2 or the python tools)
\end{itemize}
\vspace{0.01cm}
\noindent
Please make sure that these libraries/programs are available and their location is visible in your \texttt{\$PATH}\ and
\texttt{\$LD\_LIBRARY\_PATH}, respectively.
Note that these libraries are \emph{NOT} part of the installation package of the \qcm\ software suite (see Section \ref{sec:installation}\
for further details).

\subsection{Downloading \qcm{}}\label{sec:reg}

\begin{framed}
\noindent \textbf{Advice:} We recommend to install \qcm{} with the \mol-\qcm{} interface. \qcm{} installed with the \mol-\qcm{} interface is also fully usable as a standalone version.
\end{framed}

%%% Obtaining QCMaquis from scine.ethz.ch
If you want to install QCMaquis with the \mol-\qcm{} interface, please proceed to Section~\ref{sec:install-qcm-mol}. No manual download is necessary: \qcm{} and \mol-\qcm{} interface will be automatically downloaded installed during the compilation and installation procedure of \mol{}.

If you want to download and install a standalone version of \qcm{}, it may be obtained from \url{https://scine.ethz.ch/download/qcmaquis}. After downloading, please proceed to Section~\ref{sec:install-qcm-stand} for installation instructions.

\clearpage
\newpage
\section{Installation}\label{sec:installation}

% QCMaquis in Molcas
\subsection{Installation of the \mol-\qcm{} interface}\label{sec:install-qcm-mol}

% \noindent In the following steps \ref{subsubsec:download}-\ref{subsubsec:runtime}\ we describe how to successfully build and install the \mol\ program together with the \qcm\ software suite.
The installation of \qcm\ has been tested for different
operating systems and compiler/math libraries environments found in Section~\ref{subsec:supportedOS}.
While other combinations might work equally well they are \emph{not} officially supported.

\noindent The installation of the \qcm\ software suite within \mol\ will
comprise several libraries which are \emph{automatically downloaded and installed} during the \mol\ build
process provided that DMRG support within \mol\ is requested by the user.
The list of external libraries comprises:
\begin{itemize}
\item Optionally: the NEVPT2 program
\end{itemize}
All of the above libraries will be installed locally in the user-defined build folder \texttt{my-Molcas-build} of \mol.

\subsubsection{Download and build folder setup}
Download \mol{} from the official \mol{} GitLab repository:

\begin{verbatim}
git clone git@gitlab.com:Molcas/OpenMolcas.git /path/to/my-Molcas-src
\end{verbatim}

\noindent Create a build folder \texttt{\molbuild} -- note that this folder \emph{does not} necessarily have to be a subfolder
of \texttt{my-Molcas-src} -- and change to this new folder:
\begin{verbatim}
mkdir /path/to/my-Molcas-build && cd /path/to/my-Molcas-build
\end{verbatim}

\subsubsection{Configuration}\label{subsubsec:configure}

To install \qcm{} and the \qcm{}-\mol{} interface, one may follow the standard \mol{} installation procedure and enable the CMake option \kwd{DMRG} (and \kwd{NEVPT2} for the NEVPT2 program).

% Below we present some example configurations for the GNU compiler collection and the Intel compiler.

\paragraph{Configuration with the GNU compiler suite}\label{sec:gnu-conf}$\;$\\

\noindent To configure \mol\ and \qcm\ with the GNU compiler suite type
\begin{verbatim}
FC=gfortran CC=gcc CXX=g++ cmake -DDMRG=ON -DLINALG=MKL \
  -DQCMaquis_NAME="Your Name" -DQCMaquis_EMAIL="your@email.com" \
  /path/to/my-Molcas-src
\end{verbatim}

\noindent where we assumed (\texttt{-DLINALG=MKL}) that the MKL libraries are available (recommended option).
If the MKL libraries are not available internal math libraries of \mol\ can be requested with \texttt{-DLINALG=Internal} (default option for \texttt{LINALG} in \mol) or \texttt{-DLINALG=Runtime} (to use a system-wide BLAS or LAPACK installation). Please also specify your name and e-mail address in the \texttt{-DQCMaquis\_NAME} and \texttt{-DQCMaquis\_EMAIL} CMake options, otherwise the download of \qcm{} will not work.

\begin{framed}
	\noindent\large{Note: We strongly recommend to configure \qcm\ (and \mol) with the Intel Math Kernel Library (MKL) to ensure the best numerical performance. See Section \ref{subsec:supportedOS}\ for further details.}
\end{framed}
% \paragraph{Configuration with the Intel compiler suite}\label{sec:intel-conf}$\;$\\
\begin{framed}
 \noindent Note: Intel compiler is currently not supported. Running a binary compiled with Intel compilers may result in segmentation faults. We are currently investigating the issue.
\end{framed}

\paragraph{Upgrading an existing \mol{} installation}\label{sec:build-upgrade}$\;$\\

If you're upgrading from an \mol{} installation compiled with an older release of \qcm{}, you need to add an additional command line flag when invoking \texttt{cmake}, namely
\begin{verbatim}
  -DQCMaquis_UPDATE=ON
\end{verbatim}

This flag activates the \qcm{} update for the current and all the following \texttt{cmake} runs. To switch off \qcm{} updates, \texttt{cmake} may be run with
\begin{verbatim}
  -DQCMaquis_UPDATE=OFF
\end{verbatim}

% \noindent To configure \mol\ and \qcm\ with the Intel compiler suite including MKL type
% \begin{verbatim}
% FC=ifort CC=icc CXX=icpc cmake -DDMRG=ON -DLINALG=MKL \
%   -DQCMaquis_NAME="Your Name" -DQCMaquis_EMAIL="your@email.com" \
%   /path/to/my-Molcas-src
% \end{verbatim}

\paragraph{Additional options}$\;$\\
Following additional options can be switched on by passing additional flags to \texttt{cmake}:
\begin{itemize}
 \item \texttt{-DNEVPT2=ON}: Downloads and compiles the NEVPT2 program and its interface with \mol\ and \qcm.
 \item \texttt{-DOPENMP=ON}: Enable OpenMP parallelization for MOLCAS and NEVPT2 (see Section \ref{sec:parallel}). Note that \qcm{} is built with OpenMP support \emph{regardless} of this flag.
\end{itemize}

Other options to enable/disable various options in \mol{} may be passed to \texttt{cmake} at this step. Please consult the \mol{} manual for details.
% \paragraph{Additional \mol{} options}$\;$\\
% By running
% \begin{verbatim}
%  ccmake /path/to/my-Molcas-src
% \end{verbatim}
% in the build directory you may obtain a text user interface where several options for building \mol{}, along with the ones mentioned above, may be conveniently set. Use \texttt{ccmake} to set other various options for \mol{}: for details please consult the \mol{} manual.
\begin{framed}
 \noindent Note: it is currently discouraged to install the \qcm{}-\mol{} interface together with the \textsc{Block} or \textsc{CheMPS2} interfaces available in \mol{}. This setup has not been tested and may lead to an unusable installation. A fix allowing different interfaces to be installed at the same time is currently under development.
\end{framed}

\paragraph{MPI parallel and shared-memory OMP parallel configurations}\label{sec:parallel}$\;$\\

\noindent Installing an MPI-parallel version of \mol\ with DMRG support is possible although
\qcm\ itself is by default shared-memory OMP but \emph{not} yet MPI-parallelized.
To configure \mol\ for an MPI-installation type
\begin{verbatim}
FC=mpif90 CC=mpicc CXX=mpiCXX cmake -DDMRG=ON -DLINALG=MKL \
  -DQCMaquis_NAME="Your Name" -DQCMaquis_EMAIL="your@email.com" \
  /path/to/my-Molcas-src
\end{verbatim}

\vspace{2ex}

A shared-memory OMP parallelized version of \mol\ can be activated with the option
\texttt{-DOPENMP=ON}\ passed to \texttt{cmake} during the configuration step, for example
\begin{verbatim}
FC=... CC=... CXX=... cmake -DDMRG=ON -DLINALG=MKL \
  -DQCMaquis_NAME="Your Name" -DQCMaquis_EMAIL="your@email.com" \
  /path/to/my-Molcas-src
\end{verbatim}

\noindent It should work with either compiler suite, GNU or Intel, but the user may want to consult the \mol\ user manual for further information.

\vspace{2ex}

In order to exploit the shared-memory OMP parallelization of \qcm\ which is \emph{enabled by default}
the user is strongly encouraged to set at runtime the environment
variable \texttt{export QCMaquis\_CPUS=XX} or \texttt{export OMP\_NUM\_THREADS=XX}
where \texttt{XX}\ specifies the number of shared-memory cores to be used. The default is to use a single core.

\subsubsection{Building and installation}\label{subsubsec:build}

After a successful configuration, type
\begin{verbatim}
make
\end{verbatim}
\noindent or
\begin{verbatim}
make -j8
\end{verbatim}
to compile \mol\ (in parallel on 8 cores) and install all its components in the build folder \texttt{\molbuild}.
In the same folder \qcm\ as well as the required \textsc{Boost} and \textsc{Alps}
libraries will be downloaded and installed, respectively.
The installation process thus requires a working internet connection.

\subsubsection{Setting up the runtime environment}\label{subsubsec:runtime}

After having successfully passed the \qcm\ installation step as indicated by the CMake message
\begin{verbatim}
[xxx%] Installation of QCMaquis, ALPS and Boost was successful!
\end{verbatim}
set up your \mol{} environment by setting \texttt{MOLCAS} environment variable
\begin{verbatim}
export MOLCAS=/path/to/my-MOLCAS-build
\end{verbatim}
and then set up the \qcm{} runtime environment by running the following shell script
\begin{verbatim}
source $MOLCAS/qcmaquis/bin/qcmaquis.sh
\end{verbatim}
You might want to include the above lines in your local \texttt{\$HOME/.bashrc} to set up your environment automatically every time your session is started.

\subsection{Supported operating systems and compiler environments}\label{subsec:supportedOS}

\qcm\ in \hostp\ is tested and officially supported for Linux (x86\_64)
and MacOS X (10.12.x and 10.13.x ) operating systems. We support GCC compilers starting from 5.x up to 7.x along with the MKL (Linux, x86\_64) and MKL/Accelerate (Mac OS X) math libraries, respectively, to successfully build the \qcm\ software suite within \hostp.
Note that other combinations might work equally well but they are not officially supported at present.

\noindent See the \mol\ manual at \url{www.molcas.org}\ for details on
supported operating systems and compiler environments of a ``plain" \mol\ installation.

%\begin{table}[htbp]
%\begin{center}
%\caption{\label{comb}Supported operating system/compiler/math library combinations for
%the combined installation of \mol\ and \qcm.}
% \begin{tabular}{lll}
%  \toprule
%operating system & compiler & math library\\
%\midrule
%Mac OS X (10.11 ``El Capitan") & GNU 5.2 & ``Accelerate" (Xcode)\\
%Mac OS X (10.11 ``El Capitan") & Intel XE 15 (patch 0) & MKL 11.2 \\
%x86\_64                        & GNU 6.2.0             & MKL 2018.1 \\
%x86\_64                        & Intel XE 15 (patch 0) & MKL 11.2 \\
%x86\_64                        & GNU 5.4               & MKL 15.0.090 \\
%x86\_64                        & GNU 4.8               & MKL 11.1 \\
%x86\_64                        & GNU 4.7.2             & MKL 11.1 \\
%x86\_64                        & Intel XE 13           & MKL 11.1\\
%x86\_64                        & Intel XE 13 (sp 1)    & MKL 11.1 \\
%x86\_64                        & OpenMPI 1.6.5 + Intel XE 13 (sp 1)    & MKL 11.1 \\
%\bottomrule
% \end{tabular}
%\end{center}
%\end{table}

\subsection{Tests and verification of the installation}

To verify that your installation of \mol\ and \qcm\ was successful, run the following command
\begin{verbatim}
pymolcas verify qcmaquis
\end{verbatim}
\noindent The message
\begin{verbatim}
Verification has been completed
\end{verbatim}
\noindent indicates that all tests passed correctly and your installation is good to go for production work.
\vspace{2ex}

\noindent Test inputs are distributed with \mol{} and can be found in the \texttt{qcmaquis} subdirectory of the \mol{} test directory.
They may also serve as sample inputs for your actual calculation.  Table \ref{tab:inputex}\ comprises a short summary of
the tests for \qcm\ covering different computational aspects within the \mol\ framework.

\begin{longtable}{lll}
\caption{\label{tab:inputex}List of \qcm{} test inputs in \mol{}.}
\\
  \toprule
file & type & features\\
  \midrule
\endfirsthead

\multicolumn{3}{c}%
{{\bfseries \tablename\ \thetable{} -- continued from previous page}} \\
\toprule
file & type & features\\
\midrule
\endhead
\midrule
001.input & State-specific & Ground-state DMRG-SCF optimization of N$_2$\\
002.input & State-average & Two-state (S$_0$/S$_1$) DMRG-SCF of N$_2$\\
003.input & DMRG-CI     & DMRG-CI with a dynamic list of renormalized states $m$\\
004.input & State-specific & Structure optimization of the ground state of dioxetanone\\
005.input & State-specific & Structure optimization of the S$_1$ state of N$_2$\\
006.input & PCM model      & DMRG-SCF for a non equilibrium state\\
007.input & MPS-SI & Calculation of singlet-triplet spin-orbit matrix elements\\
008.input & State-specific & DMRG-SCF with Fiedler ordering and CI-DEAS start guess\\
009.input & NEVPT2 & DMRG-NEVPT2 energy calculation of the CH$_2$ triplet state\\
010.input & MC-PDFT & DMRG-MC-PDFT energy calculation of the ground state of N$_2^+$\\
\bottomrule
\end{longtable}

\subsection{Maintenance}\label{sec:maintenance}

\subsubsection{Updates and patches}\label{sec:patches-qcm}

New versions of the \qcm\ software suite as well as possible bug fixes will be announced
on the \href{https://www.scine.ethz.ch/download/qcmaquis}{SCINE \qcm\ homepage}. They can be then downloaded from the SCINE website or by rerunning CMake in the \mol{} directory.

\subsubsection{Reporting bugs and user support}\label{sec:bugs-support}

The \qcm\ program suite is distributed to the \mol\ community with no obligations on the side of the authors.
The authors thus take no responsibility for the performance of the code nor for the correctness of the results.
This distribution policy gives the authors no responsibility for problems experienced by the users when using the \qcm\ program in \mol.

%Bug reports are to be reported via the mailing list \href{mailto:dmrg@phys.chem.ethz.ch}{dmrg@phys.chem.ethz.ch} and will be dealt with by one of the authors, although no responsibility on the promptness of this response is given.
Bugs/suggestions for improvements are to be reported as follows:

\begin{itemize}
  \item For issues regarding usage of \qcm\ in \mol{}, open an issue in the \mol{} bug tracker at
    \url{https://gitlab.com/Molcas/OpenMolcas/issues}
  \item For issues regarding the usage of \qcm{} standalone version, please send an e-mail to \href{mailto:dmrg@phys.chem.ethz.ch}{dmrg@phys.chem.ethz.ch}.
\end{itemize}

Any issue will be dealt with by one of the authors, although no guarantee on the promptness of response is given.
In general, serious bugs that have been reported and fixed will lead to a new patch of the \qcm\ program,
announced and distributed via the \href{https://www.scine.ethz.ch/download/qcmaquis}{SCINE \qcm\ homepage}.

\clearpage
\newpage

\section{Running a \qcm\ DMRG Calculation}\label{sec:general-consid}

Running a \qcm\ DMRG calculation --- either DMRG-CI or DMRG-SCF --- is most easily achieved with the \qcm-\mol{} interface \cite{interface}, which integrates the DMRG calculation into the \mol{} workflow. Usage of the \qcm-\mol{} interface described in Section~\ref{sec:input}. In a few advanced cases the user might want to run \qcm{} directly, which will be described in Section~\ref{sec:workflow-qcm}.

\subsection{Keywords and Options for \hostp}\label{sec:input}

\subsubsection{The DMRGSCF module}\label{sec:dmrgscf}

\begin{framed}
\noindent Cite the following papers when performing DMRG-SCF calculations within the \mol{} software suite:
\begin{enumerate}
\item Freitag,~L.; Keller,~S.; Knecht,~S.; Lindh,~R.; Ma,~Y.; Stein,~C.~J.;
  Reiher,~M. \textit{in
		preparation} \textbf{Check for a preprint on \href{http://arxiv.org/}{arXiv.org} (to appear soon).}
\item{Keller,~S.;~Reiher,~M., \emph{J. Chem. Phys.} \textbf{2016}, \emph{144}, 134101\\ \href{https://doi.org/10.1063/1.4944921}{doi:10.1063/1.4944921}}
\item Keller,~S.;~Dolfi,~M.;~Troyer,~M.;~Reiher,~M., \emph{J. Chem. Phys.}
  \textbf{2015}, \emph{143}, 244118\\ \href{https://doi.org/10.1063/1.4939000}{doi:10.1063/1.4939000}
\item{Knecht~S.;~Hedeg{\aa}rd E. D.;~Keller S.;~Kovyrshin A.;~Ma  Y.;~Muolo A.;~Stein C. J.;~Reiher M.,
	\emph{Chimia} \textbf{2016}, \emph{70} ~244--251, \href{https://doi.org/10.2533/chimia.2016.244}{doi:10.2533/chimia.2016.244}.
}
  \end{enumerate}
\end{framed}

The \qcm-\mol{} interface introduces a new \hostp\ module, \texttt{DMRGSCF} which may be used to perform DMRG-CI or DMRG-SCF wave function optimizations. A \texttt{DMRGSCF} input (see example in Listing~\ref{lst:dmrgscf}) consists of the mandatory keyword \kwd{ActiveSpaceOptimizer} and the two input blocks \kwd{DMRGSettings\ldots{}EndDMRGSettings} and \kwd{OOptimizationSettings\ldots{}EndOOptimizationSettings}. Optional keywords within the \texttt{DMRGSCF} input section are \kwd{Fiedler} and \kwd{CIDEAS}, respectively.

\paragraph{The \kwd{ActiveSpaceOptimizer} keyword}\mbox{}\\
The \kwd{ActiveSpaceOptimizer} keyword specifies the CI solver program to be used in the \kwd{DMRGSCF} module. Currently, only one value, namely\\
\noindent\kwd{ActiveSpaceOptimizer=QCMaquis}\\
is supported, which, obviously, calls the \qcm{} program. In the future, this option will be used to enable interfaces to other DMRG programs such as \textsc{Block} or \textsc{CheMPS2}.

\paragraph{The \kwd{Fiedler} keyword}\mbox{}\\
The \kwd{Fiedler} keyword, i.e.,\\
\noindent\kwd{Fiedler=on}\\
enables a state-specific orbital ordering for the MPS optimization
by exploiting concepts from graph theory. The ordering follows from the elements of the
Fiedler vector \cite{fied73,fied75} which is the eigenvector corresponding to the second
lowest eigenvalue of the so-called graph Laplacian.
Further details concerning our implementation of the Fiedler-vector ordering within \qcm\ can be found in Ref.\cite{interface}.

\paragraph{The \kwd{CIDEAS} keyword}\mbox{}\\
The \kwd{CIDEAS} keyword, i.e.,\\
\noindent\kwd{CIDEAS=on}\\
enables a more advanced algorithm to construct a suitable initial MPS (see the keyword \kwd{init\_state} in Table \ref{tab:standard_kw_auto} for other options) provided by the configuration interaction dynamically extended active space (CI-DEAS) approach \cite{Legeza2003,fiedler}. The CI-DEAS protocol can be interpreted as an orbital entanglement entropy guided configuration selection and the quality of this initial guess depends on the quality of the initial CAS vector.
Further details concerning our implementation of the CI-DEAS approach within \qcm\ can be found in Ref.\cite{interface}.
\begin{framed}
\noindent \textbf{Warning!} The CI-DEAS functionality is currently restricted to calculations performed with C$_1$\ symmetry. Support for other point group symmetries will be available in due time.

\noindent \textbf{Note}: The \kwd{CIDEAS} option requires to set the ``HF occupation" for each state in the \texttt{OOptimizationSettings}\ input section by means of the \kwd{SOCC}\ keyword (see Table \ref{table:opt-keyword}).
\end{framed}

\paragraph{The \kwd{DMRGSettings} block}\mbox{}\\
The \kwd{DMRGSettings} block contains DMRG-specific options, which will be passed on to the \qcm{} program. The mandatory keywords, which must be present in each calculation, are summarized in Table~\ref{tab:standard_kw}. In addition to these keywords, any \qcm{} keyword (see Table~\ref{tab:advanced_kw} may be used in this section.
% In addition, keyword \kwd{Compress} may be used to enable wavefunction compression, e.\,g.\ for a subsequent NEVPT2 calculation.

\begin{table}[h]
 \caption{\label{tab:standard_kw}Mandatory and special \qcm{} keywords to be specified in the \kwd{DMRGSettings} block of the \kwd{DMRGSCF} input.}
\begin{tabular}{ll}
 \toprule
 keyword & description \\
 \midrule

 \multirow{3}{*}{\texttt{max\_bond\_dimension}} & \multirow{3}{11cm}{Maximum number of renormalized states
(commonly referred to as $m$-value or maximum bond dimension)
kept during each microiteration step of a sweep.}\\
& \\
& \\
\multirow{4}{*}{\texttt{nsweeps}} & \multirow{4}{11cm}{Maximum number of DMRG sweeps. Please be aware that \texttt{nsweeps} sets
the number of combined forward and backward sweeps. Thus, the actual number of sweeps is $2\ \times $ \texttt{nsweeps}.} \\
& \\
& \\
& \\
%   \cmidrule(rl){1-2}
%  \multirow{6}{*}{\texttt{Compress} (optional)} & \multirow{5}{12cm}{Compress the MPS obtained from the wavefunction optimization to an MPS with a smaller maximum number of renormalized states (provided as a parameter to the option) and use the compressed MPS in a subsequent 4-RDM evaluation. As the 4-RDM evaluation scales as $m^{3}L^8$, this is a convenient way to save computational cost in DMRG-NEVPT2 calculations.}\\
%   & \\
%   & \\
%   & \\
%   & \\
%   & \\
%   & \textbf{default:} \texttt{Compress=0}, which disables the compression. \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\begin{center}
\caption{Inoperative \kwd{RASSCF} keywords in the \kwd{OOptimizationSettings} block.}\label{table:disabled-kw}
\begin{tabular}{l@{\hspace{1.25cm}}l}
\toprule
keyword & description \\
\midrule
\multirow{2}{*}{\texttt{RASScf}} & \multirow{3}{12cm}{Keywords are unavailable as excitation level restrictions between subspaces are unsupported in DMRG calculations.}\\
\multirow{2}{*}{\texttt{GASScf}}  & \\
 & \\
\multirow{4}{*}{\texttt{TIGHt}} & \multirow{4}{12cm}{This keyword is not available because the Jacobi-Davidson diagonalization is
independent and can be controlled with the \texttt{ietl\_jcd\_tol} and \texttt{ietl\_jcd\_maxiter} parameters (see
Section \ref{sec:qcmaquis-kw}) in the \kwd{DMRGSettings} section.}\\
 & \\
 & \\
 & \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

\paragraph{The \kwd{OOptimizationSettings} block}\mbox{}\\
The \kwd{OOptimizationSettings} block contains general, non DMRG-specific options required for the wavefunction optimisation (such as number of the active electrons, active orbital specification etc.) and resemble the input for an analogous CASCI or CASSCF calculation with the \mol{} \kwd{RASSCF} module. Most of the \kwd{RASSCF} keywords are accepted, with the exception of the keywords in Table~\ref{table:disabled-kw}. Please consult the \kwd{RASSCF} entry of the \mol{} manual for a detailed description of the input.

In addition to the standard \kwd{RASSCF} keywords, several optional keywords are available, listed in Table~\ref{table:opt-keyword}.

% \begin{longtable}{l@{\hspace{1.25cm}}l}
\begin{table}[h]
\caption{Optional non-standard keywords in the \kwd{OOptimizationSettings} block.}\label{table:opt-keyword}
\begin{tabular}{ll}
\\
  \toprule
keyword & description \\
  \midrule
% \endfirsthead
%
% \multicolumn{2}{c}%
% {{\bfseries \tablename\ \thetable{} -- continued from previous page}} \\
% \toprule
% keyword & description \\
% \midrule
% \endhead
\multirow{4}{*}{\texttt{FCIDUMP}} & \multirow{4}{12cm}{Skip the optimization and write out the transformed active MO integrals to a
\kwd{FCIDUMP} file in \kwd{\$WorkDir} (see Ref. \citenum{fcidump} for a detailed description of the format) which can be used in subsequent \qcm\ DMRG calculations.}\\
 & \\
 & \\
 & \\
\multirow{6}{*}{\texttt{SOCCupy}} & \multirow{6}{12cm}{Initial electronic configuration for the calculated state(s). This keyword is equivalent to the \texttt{hf\_occ} card in the \qcm\ input (see Section
\ref{sec:qcmaquis-kw}), but allows input for multiple states. The occupation is inserted as a string (strings) of aliases of occupations of the active (RAS2) orbitals with the aliases $2 \,=\,$ full, $u \,=\,$ up, $d \,=\,$ down, $0 \,=\,$ empty. For several states, the occupation strings for each state are separated by newlines.}\\
 & \\
 & \\
 & \\
 & \\
 & \\
 & \\
 \cmidrule(rl){1-2}
 \multicolumn{2}{l}{Optional options for DMRG-NEVPT2 calculations}\\
 \cmidrule(rl){1-2}
 \multirow{3}{*}{\texttt{NEVPT2prep}} & \multirow{3}{12cm}{Prepare QCMaquis input templates for distributed 3-TDM or 4-RDM evaluation for a subsequent DMRG-NEVPT2 calculation. More about external RDM evaluation in Section~\ref{sec:4rdm-distributed}.} \\
  & \\
  & \\
\bottomrule
% \end{longtable}
\end{tabular}
\end{table}

\subsubsection{DMRG calculations with the RASSCF module}

\begin{framed}
  \noindent \textbf{Note}: The input format for DMRG/DMRGSCF calculations with the \kwd{RASSCF} module of \mol{} described in this section is \textbf{deprecated} and will be \textbf{unavailable} in future versions of \mol{}!

  \noindent Instead, please use the \kwd{DMRGSCF} module as described in Section~\ref{sec:dmrgscf}.
\end{framed}

Instead of using the new \kwd{DMRGSCF} module, DMRG calculations with \qcm{} are also supported from within the \kwd{RASSCF} module directly. Compared to the \kwd{DMRGSCF} input, the \kwd{RASSCF} input changes as follows:
\begin{samepage}
  \begin{itemize}
  \item The \kwd{DMRG} keyword must be specified in the \kwd{RASSCF} input.
  \item The \kwd{DMRGSettings} block from \kwd{DMRGSCF} input is delimited with \kwd{RGInput\ldots{}EndRG} instead of \kwd{DMRGSettings\ldots{}EndDMRGSettings}
  \item The \kwd{OOptimizationSettings} block is just the normal \kwd{RASSCF} input, i.\,e.\ it is not delimited with \kwd{OOptimizationSettings\ldots{}EndOOptimizationSettings}
  \end{itemize}
An input example is provided in Listing~\ref{lst:rasscf}.
\end{samepage}

\subsubsection{Molcas environment variables}

\noindent As described in Section \ref{sec:parallel}\ \qcm\ is built by default with a shared-memory
OMP parallelization. To speedup calculations the user can thus set at runtime the environment variable \texttt{OMP\_NUM\_THREADS} to the number of shared-memory cores to be used. The default is to use all possible cores.

\FloatBarrier
\newpage
\subsubsection{Input Examples}
\begin{lstlisting}[language=molcas,caption={(File \texttt{001.input} in the test directory) Input example for a DMRG-SCF calculation with the \kwd{DMRGSCF} module (\ch{N2}, 6 orbitals, with symmetry).},label=lst:dmrgscf]
&GATEWAY
 coord
 2
Angstrom
 N  0.000000  0.000000  -0.54880
 N  0.000000  0.000000   0.54880
 basis=cc-pvdz
&SEWARD
&SCF
&DMRGSCF
ActiveSpaceOptimizer=QCMaquis
DMRGSettings
  nsweeps            = 4
  max_bond_dimension = 100
EndDMRGSettings
OOptimizationSettings
  inactive = 2 0 0 0 2 0 0 0
  RAS2     = 1 1 1 0 1 1 1 0
  LINEAR
EndOOptimizationSettings
\end{lstlisting}
\vspace{-0.5em}
\begin{lstlisting}[language=molcas,caption={Input example for a DMRG-SCF calculation with the \kwd{RASSCF} module (\ch{N2}, 6 orbitals, with symmetry)},label=lst:rasscf]
&GATEWAY
 coord
 2
Angstrom
 N 0.000000  0.000000  -0.54880
 N 0.000000  0.000000   0.54880
 basis=cc-pvdz
&SEWARD
&SCF
&RASSCF
DMRG
RGINPUT
  nsweeps            = 4
  max_bond_dimension = 100
EndRG
inactive = 2 0 0 0 2 0 0 0
RAS2     = 1 1 1 0 1 1 1 0
LINEAR
\end{lstlisting}

\clearpage
\newpage

\section{State interaction with DMRG (MPS-SI)}\label{sec:mps-si}

\begin{framed}
 \noindent Cite the following papers when performing MPS-SI calculations within the \mol{} software suite:
\begin{enumerate}
\item Knecht,~S.;~Keller~S.;~Autschbach~J.;~Reiher~M., \emph{J.~Chem.~Theory~Comput.} \textbf{2016}, \emph{12} 5881--5894, \href{https://doi.org/10.1021/acs.jctc.6b00889}{doi:10.1021/acs.jctc.6b00889}
 \item Malmqvist,~P.-{\AA}.;~Roos,~B.~O., \emph{Chem. Phys. Lett.} \textbf{1989}, \emph{155}, 189--194\\ \href{https://doi.org/10.1016/0009-2614(89)85347-3}{doi:10.1016/0009-2614(89)85347-3}
\end{enumerate}
\end{framed}

The state-interaction module for MPS wave functions (MPS-SI) introduced in
Ref.~\cite{knec16b} is based on the \texttt{CASSI/RASSI} framework \cite{malm89,malm02} of \hostp.
MPS-SI is a generalized state-interaction approach for
both \textit{nonorthogonal} and \textit{orthonormal} spin-free MPS wave
functions which enables the evaluation of arbitrary one- and
two-particle transition matrix elements as well as, for example, matrix
elements of the spin-orbit coupling operator.
For instance, diagonalization of the spin-orbit Hamiltonian matrix
yields spin-orbit coupled wave functions as linear combinations of the
uncoupled, spin-pure MPS states. The latter can (but do not have to)
be obtained as results from one or several DMRG-SCF orbital
optimization calculations as described in Section \ref{sec:dmrgscf}.

Following the work of Malmqvist \cite{malm86}, the central element of the MPS-SI
algorithm is the transformation of the bra and ket MPS wave functions
to a biorthonormal basis representation.
It is important to note that the latter transformation is not needed
if the MPS wave functions considered for state interaction
share a common MO basis. In this particular case, the MPS-SI program directly
proceeds with the calculation of the reduced (transition) one- and two-particle
density matrices.
We emphasize that our approach is applicable to the general case with MPS
wave functions built from mutually nonorthogonal molecular orbital bases.
It therefore provides the desired flexibility
to find the best individual molecular orbital basis to represent
wave functions of different spin and/or spatial symmetry.
After solving a generalized eigenvalue
equation of the form
\begin{equation}\label{eq:eigproblem}
{\textbf{Hc}} = E {\textbf{Sc}}\ ,
\end{equation}
with the Hamiltonian matrix \textbf{H} expressed in the basis of the DMRG-SCF MPS
wave functions and the wave function overlap matrix \textbf{S},
a set of fully orthogonal and non-interacting states are obtained
as linear combinations of the DMRG-SCF MPS wave functions with
the expansion coefficients given by \textbf{c}.

Moreover, it is possible to ``dress" the diagonal elements of the Hamiltonian in Eq.~\eqref{eq:eigproblem} for \kwd{MPS-SI} by adding
a correlation-correction term obtained, for example, from a preceding NEVPT2
calculation (see Section \ref{sec:nevpt2}), by either using the \texttt{HDIAG} keyword within the \kwd{RASSI}\ module or provide the \texttt{nevpt2.h5}\ wave function file as input option to the \kwd{\&MPSSI} module (see the description of the keyword \kwd{FILE} in the \hostp\ manual for the \kwd{MPSSI} module for details on the correct syntax).

\subsection{Running MPS-SI}\label{sec:run-mpssi}

An MPS-SI calculation requires reference DMRG-SCF or NEVPT2
wave functions as obtained with the \kwd{DMRGSCF} (see Section \ref{sec:dmrgscf}) or \kwd{NEVPT2} modules (see Section \ref{sec:nevpt2}), respectively.
\begin{framed}
 \noindent Note: Currently, the MPS-SI implementation supports all point groups implemented in \hostp\
 if there is only one set of input states (i.e., all states have the same spatial \underline{and} spin symmetry). By contrast, if there is more than one set of input states of different spin symmetry, the \kwd{MPS-SI} module
 supports only calculations in C$_1$\ symmetry. The latter restriction will be removed in a forthcoming release.
\end{framed}
Listing \ref{lst:mpssi} shows an input example for an MPS-SI calculation (a calculation of spin-orbit coupling matrix elements for the \ch{CH2} molecule) where
the important keywords are highlighted in blue. A detailed explanation of the input is given below.

\begin{lstlisting}[language=molcas,caption={(File \texttt{007.input} in the QCMaquis test directory) Input example for a one-shot MPS-SI spin-orbit calculation (Singlet and triplet state, \ch{CH2} molecule, 6 electrons in 6 orbitals, C1 symmetry).},label=lst:mpssi]
&GATEWAY
  coord
  3
  CH2 Triplet coordinates in Angstrom
  C 0.000000 0.000000 0.000000
  H 0.000000 0.000000 1.077500
  H 0.784304 0.000000 -0.738832
  basis=ANO-RCC-MB
  Group=Nosym
  AMFI
  ANGMOM
  0.0 0.0 0.0
&SEWARD
&DMRGSCF
  ActiveSpaceOptimizer=QCMaquis
  DMRGSettings
    max_bond_dimension=1024
    nsweeps=10
  EndDMRGSettings
  OOptimizationSettings
    Spin=3
    Inactive=1
    Ras2=6
    NActEl=6,0,0
  EndOOptimizationSettings
  >> COPY $Project.JobIph JOBOLD
  >> COPY $Project.dmrgscf.h5 $Project.trip.h5
  >> EXEC mv $CurrDir/$Project.checkpoint_state.0.h5 $CurrDir/$Project.trip.checkpoint_state.0.h5 |\label{line:mv}|
  >> EXEC $MOLCAS/Tools/qcmaquis/qcm_checkpoint_rename.py $Project.trip.h5 -q |\label{line:rename}|
&DMRGSCF
ActiveSpaceOptimizer=QCMaquis
  DMRGSettings
    max_bond_dimension=1024
    nsweeps=10
  EndDMRGSettings
  OOptimizationSettings
    Spin=1
    Inactive=1
    Ras2=6
    NActEl=6,0,0
    JobIph
  EndOOptimizationSettings
>> COPY $Project.dmrgscf.h5 $Project.sing.h5
>> EXEC mv $CurrDir/$Project.checkpoint_state.0.h5 $CurrDir/$Project.sing.checkpoint_state.0.h5 |\label{line:mv2}|
>> EXEC $MOLCAS/Tools/qcmaquis/qcm_checkpoint_rename.py $Project.sing.h5 -q |\label{line:rename2}|
&MPSSI
  Nrof
  2 1 1
  1
  1
  FILE
  2
  $Project.trip.h5
  $Project.sing.h5
  EJOB
  SOCOupling
  0.0001
  SPIN
\end{lstlisting}

In order to calculate spin-orbit matrix elements within the \kwd{MPS-SI} framework, one needs to add the keyword \texttt{AMFI} (and optionally \texttt{Angmom}) in the \kwd{\&GATEWAY} input section (This is no different from a standard RASSI calculation). Then, we perform two DMRG-SCF calculations with the \kwd{\&DMRGSCF} module (see Section \ref{sec:dmrgscf})\ to calculate the lowest-lying triplet and singlet state of CH$_2$.
At the end of each DMRG-SCF step, we save the \texttt{dmrgscf.h5} files for each spin state for later usage in \kwd{\&MPSSI}.

The largest differences to standard RASSI calculations are found in lines~\ref{line:mv} and \ref{line:rename} in Listing \ref{lst:mpssi}. Line \ref{line:mv}, in addition to the \kwd{dmrgscf.h5} file, also saves the \qcm{} checkpoint folder \texttt{\$Project.checkpoint\_state.0.h5} under a different name, which would otherwise be overwritten in the subsequent DMRG-SCF calculation. The information about the \qcm{} checkpoint name is saved in the \texttt{dmrgscf.h5} file, hence the reference to the new name must be updated in the \kwd{dmrgscf.h5} file with an utility script named \kwd{qcm\_checkpoint\_rename.py} in line~\ref{line:rename}. In the second DMRG-SCF calculation, these steps are repeated in lines~\ref{line:mv2} and \ref{line:rename2}.

The MPS-SI framework is driven by the \kwd{\&MPSSI} module, which derives from the \mol{} \kwd{\&RASSI} module and has the same input syntax. The spin-orbit matrix-element calculation is triggered with \texttt{SPIN}. Note, that we use the keyword \texttt{EJOB}\ which prevents the MPS-SI program to calculate any two-particle reduced (transition) density matrices which are not only rather expensive to calculate compared to the one-electron reduced (transition) density matrices but also are of no further usage for the calculation of spin-orbit matrix-elements. The diagonal values for the Hamiltonian in Eq.\eqref{eq:eigproblem}\ are taken in this case from the respective \texttt{\$Project.xxx.h5} files (with \texttt{xxx=sing,trip}).
\begin{framed}
 \noindent Note: For MPS-SI, the wavefunctions from the RASSCF/DMRG-SCF calculations \emph{MUST} be provided as \texttt{dmrgscf.h5} files. Old-style \texttt{JobIph} files are not supported.
\end{framed}

\subsection{Input options for MPS-SI}\label{sec:input-mpssi}

\begin{table}[h]
  \begin{center}
    \caption{\label{table:mpssi-keyword}\kwd{\&MPSSI} module keywords not present in the \mol{} \kwd{\&RASSI} module.}
    \begin{tabular}{l@{\hspace{0.9cm}}l}
      \toprule
      keyword & description \\
      \midrule
      \multirow{2}{*}{\texttt{QDSC}} &  \multirow{2}{12.2cm}{Read the Hermitian effective Hamiltonian from a preceding QD-SC-NEVPT2 calculation for state mixing in MPS-SI.} \\
      &\\[\tableoptionskip]
      \multirow{2}{*}{\texttt{QDPC}} & \multirow{2}{12.2cm}{Same as above, but for the QD-PC-NEVPT2 effective Hamiltonian.}\\
            &\\
      \bottomrule
    \end{tabular}
  \end{center}
\end{table}
\FloatBarrier

\clearpage
\newpage

\section{The NEVPT2 module}\label{sec:nevpt2}

\begin{framed}
 \noindent Cite the following papers when performing NEVPT2 calculations within the \hostp\ software suite:
%  \bibentry{Freitag_JChemTheoryComput_Multireference_2017}
\begin{enumerate}
\item Freitag,~L.; Knecht,~S.; Angeli,~C.; Reiher,~M. \emph{J. Chem. Theory Comput.}
  \textbf{2017}, \emph{13}, 451--459\\ \href{https://doi.org/10.1021/acs.jctc.6b00778}{doi:10.1021/acs.jctc.6b00778}
\item Angeli,~C.; Cimiraglia,~R.; Malrieu,~J.-P. \emph{J. Chem. Phys.} \textbf{2002},
  \emph{117}, 9138\\ \href{https://doi.org/10.1063/1.1515317}{doi:10.1063/1.1515317}
%  \bibentry{Angeli_JChemPhys_nelectron_2002}
\end{enumerate}
\end{framed}

NEVPT2 is a second-order perturbation theory with a CAS (or a CAS-like) reference wavefunction originally developed by Angeli et al.~\cite{Angeli_JChemPhys_Introduction_2001,Angeli_ChemPhysLett_Nelectron_2001,Angeli_JChemPhys_nelectron_2002,Angeli_JChemPhys_quasidegenerate_2004} In contrast to CASPT2, it uses a Dyall Hamiltonian\cite{Dyall_JChemPhys_choice_1995} as the zeroth-order Hamiltonian and is therefore inherently free of intruder states and parameters such as the IPEA shift. NEVPT2 exists in two formulations -- the strongly- (SC-) and the partially-contracted NEVPT2 (PC-NEVPT2), which differ in the basis of the first-order wavefunction expansion. The implementation in the NEVPT2 module is based on the original NEVPT2 implementation by Angeli et al.~\cite{Angeli_JChemPhys_nelectron_2002,Angeli_JChemPhys_quasidegenerate_2004}, with the implementation of the \qcm{} DMRG reference wave function and Cholesky decomposition for the two-electron integrals\cite{chimia,Freitag_JChemTheoryComput_Multireference_2017}. For excited states both single-state and multi-state calculations with the QD-NEVPT2 approach\cite{Angeli_JChemPhys_quasidegenerate_2004} are supported.

\subsection{Running NEVPT2}

Prior to running a NEVPT2 calculation, one must obtain a reference wavefunction with the \kwd{RASSCF} or \kwd{DMRGSCF} module and perform an integral transformation with the \kwd{MOTRA} module.

Currently, the implementation supports \emph{only} QCMaquis DMRG reference wavefunctions (support for CASSCF reference wavefunctions will be added in the near future). It is nevertheless possible to run NEVPT2 with a CASSCF reference wavefunction by performing a DMRG-CI calculation with a sufficiently large $m$ value using the CASSCF converged orbitals. For example, an $m$ value of 2000 recovers the exact CASCI energy up to $5\times{}10^{-8}$\,a.\,u. for active spaces of up to 14 orbitals. Listing~\ref{lst:nevpt2} shows an input example for a NEVPT2 calculation, the important keywords are highlighted in \textcolor{green!20!blue}{blue}, and below we will explain the input example in more detail.

\begin{lstlisting}[language=molcas,caption={(\texttt{009.input} in the QCMaquis test subdirectory) Input example for a one-shot NEVPT2 calculation (Singlet \ch{CH2}, 6 electrons in 6 orbitals, RICD)},label=lst:nevpt2]
&GATEWAY
  coord
  3
  CH2 Triplet coordinates in Angstrom
  C      0.000000  0.000000  0.000000
  H      0.000000  0.000000  1.077500
  H      0.784304  0.000000 -0.738832
  basis=cc-pVTZ
  Group=Nosym
  RICD
  CDTH=1.0E-7
&SEWARD
&DMRGSCF
  ActiveSpaceOptimizer=QCMaquis
  DMRGSettings
      max_bond_dimension=128
      nsweeps=5
  EndDMRGSettings
  OOptimizationSettings
      Spin=3
      Inactive=1
      Ras2=6
      NActEl=6,0,0
      NEVPT2Prep
      EvRDM
  EndOOptimizationSettings
&MOTRA
  Frozen=0
  CTOnly
  Kpq
  HDF5
&NEVPT2
\end{lstlisting}

First, one performs a DMRG-SCF calculation with the keywords \kwd{NEVPT2Prep} and (optionally) \kwd{EvRDM}, which enable the evaluation of the 4-RDMs (and, in case of multiple states, also t-3RDMs) required by NEVPT2. If \kwd{EvRDM} is specified, the RDMs are evaluated immediately, otherwise an input file is prepared for separate RDM calculations with QCMaquis (see also Table~\ref{table:nevpt2-keyword}). The latter option is particularly useful for large active spaces, where offloaded evaluation can be performed in a massively parallel manner, which will be explained in detail in Section~\ref{sec:4rdm-distributed}.

\begin{framed}
 \noindent Note: The computational cost of the RDM evaluation grows as $N^8$ with the number of active orbitals, therefore we strongly recommend to use the distributed RDM evaluation for active spaces larger than 10-11 orbitals described in Section~\ref{sec:4rdm-distributed}.
\end{framed}

Second, one must perform an integral transformation with the \kwd{MOTRA} module. If no Cholesky decomposition or RICD is used in the calculation, the only mandatory keyword is \kwd{HDF5}, which enables the write-out of the transformed integrals in the HDF5 format required by the \kwd{NEVPT2} module. If Cholesky decomposition is used, one additionally needs to add the keys \kwd{CTOnly} and \kwd{Kpq}.

\begin{framed}
 \noindent Note: The Cholesky decomposition currently does not support symmetry, and the support for frozen orbitals in \kwd{MOTRA} with Cholesky is untested, hence also the keyword \kwd{Frozen=0} is recommended!
\end{framed}

Finally, one calls the NEVPT2 module with \kwd{\&NEVPT2}. It has no mandatory options, but options described in Section~\ref{sec:nevpt2-opts} can be specified. If a distributed RDM evaluation is performed, the \kwd{NEVPT2} module must be called in a separate calculation.

\subsection{Options to the \kwd{NEVPT2} module}\label{sec:nevpt2-opts}
\begin{longtable}{l@{\hspace{0.9cm}}l}
\caption{\label{table:nevpt2-keyword}NEVPT2 keywords.}
\\
  \toprule
keyword & description \\
  \midrule
\endfirsthead

\multicolumn{2}{c}%
{{\bfseries \tablename\ \thetable{} -- continued from previous page}} \\
\toprule
keyword & description \\
\midrule
\endhead
      \texttt{STATes} &  Number of electronic states to calculate. Default: 1 \\[\tableoptionskip]
      \multirow{2}{*}{\texttt{NOMS}} &  \multirow{2}{13.2cm}{Omit the QD-NEVPT2 calculation and perform single-state NEVPT2 calculations instead.} \\
      &\\[\tableoptionskip]
      \multirow{5}{*}{\texttt{MULT}} & \multirow{5}{13.2cm}{Select specific states to perform QD-NEVPT2 calculation. Followed by a list of whitespace-separated state numbers, preceded by their total amount. Example: \kwd{MULT=3 1 2 4} for states 1, 2, 4 of a preceeding DMRG-SCF calculation of 4 roots (or more). \kwd{MULT=ALL} includes all states and is the default.}\\
      &\\
      &\\
      &\\
      &\\[\tableoptionskip]
      \multirow{3}{*}{\texttt{FILE}}&\multirow{3}{13.2cm}{Specify the path to a \kwd{JobIph} or \kwd{.h5} file with the reference wavefunction. By default, the reference wavefunction is read from \kwd{JOBIPH}.}\\
      &\\
      &\\[\tableoptionskip]
      \multirow{9}{*}{\texttt{FROZen}}&\multirow{9}{13.2cm}{Specify the number of frozen orbitals. The number of frozen orbitals may be specified in two ways: if only one number $n$ is specified, then all orbitals from 1 to $n$ are frozen. Otherwise, it is possible to freeze specific orbitals with the \texttt{SELECT} keyword which follows the \texttt{FROZEN} keyword. In this case, the total number of frozen orbitals followed by the space-separated list of frozen orbitals must be entered. Note that if symmetry is used, the orbital numbering for all symmetries is still consecutive, e.\,g. the 1\textsuperscript{st} orbital of symmetry 2 is has the number $m+1$ if there are $m$ orbitals in symmetry 1.}\\
      &\\
      &\\
      &\\
      &\\
      &\\
      &\\
      &\\
      &\\[\tableoptionskip]
      \multirow{3}{*}{\texttt{NOPC}}&\multirow{3}{13.2cm}{Disable the PC-NEVPT2 calculation. If the option is not present (default), both SC-NEVPT2 and PC-NEVPT2 calculations are performed.}\\
      &\\
      &\\[\tableoptionskip]
      \multirow{5}{*}{\texttt{SKIPk}}&\multirow{5}{13.2cm}{Skip the calculation of Koopmans' matrices. Requires a file named \texttt{nevpt.h5} obtained from a previous calculation in the scratch directory. May be useful to restart a previous crashed calculation if it crashed past the calculation of Koopmans' matrices, and may save some computational time, especially for large active spaces.}\\
      &\\
      &\\
      &\\
      &\\
      \cmidrule(rl){1-2}
      \multicolumn{2}{l}{New options concerning 4-RDM calculations (since \relv)}\\
      \cmidrule(rl){1-2}
      \multirow{4}{*}{\texttt{RDMRead}}&\multirow{4}{13.2cm}{Do not calculate the 4-RDM, but rather read it from QCMaquis result files \texttt{\$Project.results$\_$state.X.h5} for state \texttt{X}. Useful if the previous calculation crashed but the 4-RDM evaluation step has succeeded. Do NOT use it if you are using the distributed 4-RDM calculation.}\\
      &\\
      &\\
      &\\[\tableoptionskip]
      \multirow{8}{*}{\texttt{DISTributedRDM}}&\multirow{8}{13.2cm}{Enable reading the distributed RDMs, calculated separately. This keyword should be followed by another line, which specifies the path to the folder with the calculation results. The 4-RDM will then be read from QCMaquis HDF5 files found in \kwd{$\langle$path$\rangle$/4rdm-scratch.$\langle$state$\rangle$/parts/part-*/}\kwd{
      \$Project.results$\_$state.$\langle$state$\rangle$.h5}.
      More about distributed n-RDM evaluation can be found in Section~\ref{sec:4rdm-distributed}: if the scripts presented in this section are used, the path should be specified as \texttt{\$WorkDir}.}\\
      &\\
      &\\
      &\\
      &\\
      &\\
      &\\
      &\\
      \bottomrule
%    \end{tabular}
%  \end{center}
\end{longtable}
\FloatBarrier
\subsection{Massively parallel distributed 4-RDM calculations}\label{sec:4rdm-distributed}
The computational cost of the RDM evaluation grows as $N^8$ with the number of active orbitals, and becomes prohibitively expensive even for moderate active spaces. Therefore we provide an (experimental) python utility \texttt{jobmanager.py} for distributed massively parallel 4-RDM calculations. With distributed 4-RDM calculations, active spaces of up to 22 orbitals can be employed in DMRG-NEVPT2 calculations without any approximation to the 4-RDM.

\texttt{jobmanager.py} splits the evaluation of the 4-RDM $G_{ijklmnop}$ into four-index subblocks with indices $i,j,k,l$. Due to permutational symmetry and the properties of the creation and annihilation operators, $i >= j >= k >= l$ and no more than two indexes are equal (pairwise equality $i=j$ and $k=l$ is allowed). The script prepares input files and, if requested, submits a separate job for each subblock, and merges the subblocks into the full matrix once the jobs are finished. The script is expected to be run on a head node of a distrubuted computing system with a batch system: \href{https://www.ibm.com/support/knowledgecenter/en/SSETD4/product_welcome_platform_lsf.html}{LSF} has been tested, but any batch system which supports the \href{http://www.drmaa.org/}{DRMAA} library, such as Slurm or PBS, should work. If no support for DRMAA is found, the script still may be used to prepare the input files for each subblock, which then may be submitted manually.

Note that the DMRG-SCF/NEVPT2 calculation need not be performed on the same system as the 4-RDM evaluation.
\subsubsection*{How to run NEVPT2 calculations with distributed 4-RDM evaluation}
\subsubsection{Prerequisites}
\begin{itemize}
 \item Python $>= 2.7.9$ (3.x is also supported)
 \item (optional) DRMAA library compatible with your batch submission system\\
 (e.\,g.\ \href{https://github.com/IBMSpectrumComputing/lsf-drmaa}{LSF-DRMAA})
 \item (optional) \href{https://github.com/drmaa-python/drmaa-python.github.com}{Python DRMAA}
 \item (optional) GNU Parallel
\end{itemize}
 If your system administrator has not set up DRMAA and Python DRMAA, you might need to download and install these libraries yourself. After the installation, the environment variable \kwd{DRMAA\_LIBRARY\_PATH} must be set to the path to \kwd{libdrmaa.so} and, if Python does not find the DRMAA Python binding, also \kwd{PYTHONPATH} to the path of the Python DRMAA library.
\subsubsection{Workflow}\label{sec:4rdm-workflow}
\begin{enumerate}
 \item Run DMRGSCF and MOTRA calculations as shown in e.\,g. Listing~\ref{lst:nevpt2}, but \myemph{omit} the \kwd{\&NEVPT2} keyword. The \kwd{NEVPT2Prep} keyword creates QCMaquis input templates and the MPS checkpoint files required for a later 4-RDM and/or t-3RDM evaluation.
 \item Copy the \kwd{\$MOLCAS/Tools/distributed-4rdm/prepare\_rdm\_template.sh} script to the MOLCAS scratch directory and run it. The script will create subdirectories named \kwd{4rdm-scratch.$\langle$state$\rangle$} for each state.

 If you wish to perform the 4-RDM evaluation on a different machine (e.\,g.\ a cluster), copy the subdirectory for each state to that machine.
 If you do not wish to evaluate the 4-RDM for all states, pass the list of desired states as parameters to the \kwd{prepare\_rdm\_template.sh} script. For example,
 \kwd{./prepare\_rdm\_template.sh \{0..2\}} or \kwd{./prepare\_rdm\_template.sh 0 1 2} will create the scratch directories for states from 0 to 2 (note that QCMaquis starts counting states with 0).

%  \item For each state, create a directory on the machine where the RDM calculation will be performed with subdirectories \kwd{parts} and \kwd{template}. Copy the \kwd{jobmanager.py} script to this directory, and following files and subdirectories from the DMRGSCF/MOTRA scratch directory to the \kwd{template} subdirectory:
%  \begin{itemize}
%   \item \kwd{\$MOLCAS\_Project.checkpoint\_state.$\langle$state$\rangle$.0.0.h5}
%   \item \kwd{FCIDUMP}
%   \item \kwd{meas-4rdm.$\langle$state$\rangle$.in}
%   \item \kwd{submit-4rdm.sh} (from \kwd{\$MOLCAS/template-files/distributed-4rdm})
%  \end{itemize}
%  \item Change to the \kwd{template} subdirectory and edit the \kwd{submit-4rdm.sh} file. Change the state number (line starting with \kwd{state=}) and the number of processors per job. Rename the \kwd{meas-4rdm.$\langle$state$\rangle$.in} file to \kwd{meas-4rdm-template.in}.
 \item \textbf{If you have installed and working DRMAA setup:}\\
  For each state, change to the \kwd{4rdm-scratch.$\langle$state$\rangle$} subdirectory and run\\
 \kwd{nohup jobmanager.py \&}\\
 (Login to the machine where you evaluate the 4-RDM before if you wish to run the evaluation on a different machine.) This will create a subdirectory for each batch job (corresponding for each four-index 4-RDM subblock) and submit the jobs. The script will stay in the background until all the jobs have completed.
 The script also accepts the following job-specific options:
 \begin{itemize}
  \item \kwd{-t HH:MM:SS}: set the maximum walltime per job. Default is 24h.
  \item \kwd{-n NCPU}: run each job in an SMP parallelised fashion and set the number of CPU cores per job. Default is 1 core. For large active spaces, it is recommended to use several cores (e.\,g.\ 16 or 24, or as much as is available on a single node on your cluster).
 \end{itemize}
 \textbf{If you do NOT have DRMAA installed and working:}\\
 Run the \texttt{jobmanager.py} script with the \texttt{-n} option:\\
 \kwd{jobmanager.py -n}\\
 This will create subfolders for each 4-RDM block and prepare all the necessary input scripts, but will not submit them to the batch system. Now you may manually submit the scripts from the subfolders \kwd{parts/part-*}.

 \item If you ran the distributed 4-RDM calculation on a different machine, copy the \kwd{4rdm-scratch.$\langle$state$\rangle$} back to \mol{} \kwd{\$WorkDir}.
 \item Create an input file with the input to the \kwd{\&NEVPT2} module and run it. The keyword \kwd{DISTributedRDM} followed by the path to \kwd{4rdm-scratch.$\langle$state$\rangle$} folders (in our case, \kwd{\$WorkDir}) is \emph{mandatory}.
\end{enumerate}

\subsubsection{Troubleshooting}
The \kwd{jobmanager.py} script is experimental, and also batch jobs in queuing systems are prone to crash, therefore we provide a mechanism to identify and restart the crashed batch jobs. The NEVPT2 program will check if the 4-RDM calculation has been finished correctly. If some 4-RDM values are missing, the NEVPT2 program will stop with an error. In this case several options are available:
\begin{itemize}
 \item[] \textbf{If DRMAA has been used:}
 \item If the \kwd{jobmanager.py} finishes without errors, it will produce two files, \kwd{successlist} and \kwd{faillist} with the list of successful and failed batch jobs, respectively. In this case, the failed jobs may be restarted using the restart mode of \kwd{jobmanager.py}, which is invoked with\\
 \kwd{nohup jobmanager.py -r successlist faillist \&}\label{item:jobm-restart}
 \item If the \kwd{jobmanager.py} finishes with an error, the \kwd{successlist} and \kwd{faillist} will be either nonexistent or empty. Note that this does NOT necessarily mean that the jobs have failed: in our tests, certain configurations of the queuing system may lead to the crash of the \kwd{jobmanager.py} script after the successful completion of the jobs.\\

 \textbf{If DRMAA has not been used and the script was run with the \texttt{-n} switch:}\\
 \item In this case the user is advised to check manually the subfolders for each 4-RDM subblock for the existence of \texttt{\$Project.results$\_$state.X.h5} files. The files should exist and the command\\
 \kwd{h5dump \$Project.results$\_$state.X.h5 | grep fourpt}\\
 should not yield an empty result -- otherwise the corresponding calculation should be rerun.
\end{itemize}

\subsubsection{Transition 3-RDM distributed calculations}
\kwd{jobmanager.py} also supports distributed calculations of t-3RDMs (required for QD-NEVPT2). The split evaluation is similar to that of the 4-RDMs, and Section~\ref{sec:4rdm-workflow} can be followed with the following differences:
\begin{enumerate}
 \item The t-3RDM evaluation requires two states instead of one. Run the \\
 \kwd{prepare\_rdm\_template.sh} script with the \kwd{-3} parameter.
 \item Launch the \kwd{jobmanager.py} script with the \kwd{-3} parameter.
\end{enumerate}

\clearpage
\newpage

\section{\qcm{} Standalone Version}

Section \ref{sec:install-qcm-stand}\ describes in details the installation process for a standalone version of \qcm.
\begin{framed}
\noindent Note: If you have installed \qcm{} with the \qcm{}-\mol{} interface, you may skip this Section.
\end{framed}

% To install \qcm\ as external module of the quantum chemistry software \mol\ proceed to Section \ref{sec:install-qcm-mol}.

% QCMaquis standalone

\subsection{Installation}\label{sec:install-qcm-stand}

\noindent In the following steps \ref{subsubsec:download-stand}-\ref{subsubsec:runtime-stand}\ we describe how to successfully build and install  the \qcm\ software suite in a standalone manner.
The installation of \qcm\ has been tested for different operating systems and compiler/math libraries environments. Their list can be found in Section \ref{subsec:supportedOS}.
While other combinations might work equally well they are \emph{not} officially supported.

The following libraries are required and need to be installed manually by the user (but are usually provided by the system package manager)
\begin{itemize}
 \item GSL
 \item HDF5
 \item SZIP
 \item Boost
\end{itemize}
Additionally, \qcm{} distribution contains a customised version of the ALPS 2.3.0 library, which will be downloaded and installed automatically.

\subsubsection{Download \qcm}\label{subsubsec:download-stand}

As mentioned in Section~\ref{sec:reg}, \qcm{} standalone version may be downloaded from \texttt{\url{https://scine.ethz.ch/download/qcmaquis}}.

\subsubsection{Setting up a build folder}

Extract the downloaded archive \texttt{public.tar.bz2} into a new directory \texttt{\qcmsrc}.
\noindent Create a build folder \texttt{\qcmbuild} -- note that this folder \emph{does not} necessarily have to be a subfolder
of \qcmsrc\ -- and change to this new folder:
\begin{verbatim}
mkdir /path/to/my-QCMaquis-build && cd /path/to/my-QCMaquis-build
\end{verbatim}

\subsubsection{Configuration}\label{subsubsec:configure-stand}

\qcm{} supports operating systems, C++ compilers and math libraries as listed in Section~\ref{subsec:supportedOS}. Below we will show the configuration steps for
the most popular compiler suites GNU (Section \ref{sec:gnu-conf-stand}) and Intel (Section \ref{sec:intel-conf-stand}), respectively.
How to setup and use a shared-memory OMP installation of \qcm\ is described in Section \ref{sec:parallel-stand}.

\paragraph{Configuration with the GNU compiler suite}\label{sec:gnu-conf-stand}$\;$\\

\noindent To configure \qcm\ with the GNU compiler suite type
\begin{verbatim}
CC=gcc CXX=g++ cmake /path/to/my-QCMaquis-src
\end{verbatim}

\noindent where we assumed that the Intel Math Kernel Library (MKL) is available (recommended option).
If the MKL libraries are not available \qcm\ will search for other suitable math libraries installed on the operating system. If none are found the configuration step will stop with an appropriate message.

% \paragraph{Configuration with the Intel compiler suite}\label{sec:intel-conf-stand}$\;$\\
%
% \noindent To configure \qcm\ with the Intel compiler suite including MKL type
% \begin{verbatim}
% FC=ifort CC=icc CXX=icpc cmake -DQCM_standalone=ON /path/to/my-QCMaquis-src
% \end{verbatim}

\begin{framed}
 \noindent Note: Intel compiler is currently not supported. Running a binary compiled with Intel compilers may result in segmentation faults. We are currently investigating the issue.
\end{framed}

\paragraph{Shared-memory OMP parallel configuration}\label{sec:parallel-stand}$\;$\\

By default \qcm\ is built as shared-memory OMP parallelized version which should work with either compiler suite, GNU or Intel.

\vspace{2ex}

In order to exploit the shared-memory OMP parallelization of \qcm\ the user is strongly encouraged to set at runtime the environment
variable
\begin{verbatim}
export OMP_NUM_THREADS=XX
\end{verbatim}
where \texttt{XX}\ specifies the number of shared-memory cores to be used. The default (\emph{depending on the operating system!!!}) is to use all available cores.

\subsubsection{Building and installation}\label{subsubsec:build-stand}

After a successful configuration, type
\begin{verbatim}
make
\end{verbatim}
\noindent or
\begin{verbatim}
make -j8
\end{verbatim}
to compile \qcm\ (in parallel on 8 cores) and install all its components in the build folder \texttt{\qcmbuild}.

\subsection{Keywords and Options}\label{sec:qcmaquis-kw}

In the following we describe (i) compulsory keywords (Section \ref{sec:compul-qcm}), (ii)
optional keywords (Section \ref{sec:optional-qcm}) for \qcm\ DMRG calculations as well as (iii)
keywords for property calculations (Section \ref{sec:prop-qcm}).
Most of the \qcm\ keywords have default settings that guarantee convergence in the general case and are inserted automatically
by the host program (\hostp\ in this case). A reasonable choice of default values for optional keywords is given in our
example \qcm\ input file in Section \ref{sec:input-qcm}.

\begin{framed}
\noindent\textbf{Note:}\ \textsc{Maquis} has many features beyond quantum chemistry, e.g. related to solid state physics.
Some keywords listed in the example file in Section \ref{sec:input-qcm} (e.\,g.\ \kwd{LATTICE}, \kwd{lattice\_library} etc.)\ are therefore explained only briefly in the following
and are not to be changed from the default values (see Listing~\ref{lst:qcm}), if a quantum chemical calculation is desired.
\end{framed}

\subsubsection{Compulsory keywords}\label{sec:compul-qcm}

The keywords in Tables \ref{tab:standard_kw} and \ref{tab:standard_kw_auto} have to be set for every DMRG calculation since they may crucially affect the accuracy of the final result.
Their choice depends for example on the molecule under consideration
(i.\,e. whether you expect strong static electron correlation and/or dynamical correlation to play a major role),
the nature of the reference orbitals (Hartree-Fock orbitals, natural orbitals of some kind, \ldots), the size of the
active space, and many other aspects.

% \noindent Some of the compulsory keywords listed in Table \ref{standard_kw}\ are indeed automatically set if you run a \qcm\ DMRG calculation through the \qcm\ host program driver which is the case for \mol-\qcm\ DMRG
% calculations. In this case skip the upper part of Table \ref{standard_kw}\ and proceed immediately to the lower part
% marked by ``{\color{red}{Keywords \emph{NOT} set by the host program \mol}}".

\begin{table}
  \caption{\label{tab:standard_kw_auto}Compulsory \qcm{} keywords automatically set by the \qcm{}-\mol{} interface.}
  \begin{tabular}{p{3.6cm}l}
  \toprule
keyword & description \\
  \midrule

\texttt{nelec} & Total number of electrons. \\[\tableoptionskip]
\multirow{3}{*}{\texttt{irrep}} & \multirow{3}{12cm}{Irreducible representation of the point group symmetry of the target state. \textbf{Note:} Counting starts with 0 which has to be the totally symmetric representation.} \\
& \\
& \\[\tableoptionskip]
\multirow{2}{*}{\texttt{spin}} & \multirow{2}{12cm}{Total spin ($2\times S$) of the target state, for example:\\
\texttt{spin=0} (singlet), \texttt{spin=1} (doublet), \texttt{spin=2} (triplet), \ldots}\\
& \\[\tableoptionskip]
\texttt{L} & Length of lattice = number of orbitals in the active space. \\[\tableoptionskip]
\multirow{2}{*}{\texttt{integral\_file}} & \multirow{2}{12cm}{Path and filename of the integral file, for example\\
\texttt{integral\_file = /path/to/file/FCIDUMP}} \\
& \\[\tableoptionskip]
\texttt{chkpfile} & Path and name of the folder in which the MPS is stored. \\[\tableoptionskip]
\texttt{resultfile} & Path and filename of the result file. \\[\tableoptionskip]
\multirow{3}{*}{\texttt{n\_ortho\_states}} &\multirow{3}{12cm}{If an excited state calculation is desired, the number of states the current wave function is to be orthogonalized against shall be specified here.} \\
& \\
& \\[\tableoptionskip]
\multirow{3}{*}{\texttt{ortho\_states}} & \multirow{3}{12cm}{Path(s) and filename(s) of the MPS checkpoint file(s) that store the lower lying states to which the current MPS shall be orthogonal to.} \\
& \\
& \\[\tableoptionskip]
\multirow{9}{*}{\texttt{init\_state}} & \multirow{9}{12cm}{Possible options are \texttt{default}, \texttt{thin} and \texttt{hf}.
The \texttt{default} and \texttt{thin} initializations fill the initial MPS with random numbers, the difference being
that a singular value decomposition reduces the bond dimension to \texttt{init\_bond\_dimension} in the case of
\texttt{thin}. Usage of \texttt{hf} generates an MPS consisting of only the determinant defined on the \texttt{hf\_occ}
card. Note that the CI-DEAS procedure \cite{Legeza2003,fiedler,interface} (as invoked by \texttt{dmrginit.py}, see Section
\ref{sec:pytools-qcm}) behaves like a restart from the newly generated CI-DEAS MPS.\\
% {\textbf{default in \mol: \texttt{init\_state = "default"}}}
}\\
& \\
& \\
& \\
& \\
& \\
& \\
& \\
& \\[\tableoptionskip]
\multirow{7}{*}{\begin{minipage}{5cm}
                  \texttt{%
                  LATTICE\\
                  lattice\_library\\
                  MODEL\\
                  model\_library\\}
                \end{minipage}}
 & \multirow{3}{11cm}{These options set the Hamiltonian and the lattice type in \qcm{} calculations. For a quantum chemical calculation, these must be set to (see also Listing~\ref{lst:qcm})}\\
 & \\
 & \\
  & \texttt{LATTICE = "orbitals"}\\
  & \texttt{lattice\_library = "coded"}\\
  & \texttt{MODEL = "quantum\_chemistry"}\\
  & \texttt{model\_library = "coded"}\\
\bottomrule
\end{tabular}
\end{table}
\FloatBarrier
\begin{framed}
\noindent\textbf{Note:}\ We strongly encourage any new user of \qcm\ to carefully read first Ref.\ \citenum{Keller_CHIMIA_Determining_2014}\ which gives a detailed introduction to
DMRG calculations in quantum chemistry. Further quantum chemical DMRG sample calculations starting from different computational setups are discussed for
example in Refs. \cite{interface,ru-dmrg,fde-dmrg,srdft-dmrg}.
\end{framed}
\subsubsection{Optional keywords}\label{sec:optional-qcm}

The keywords summarized in Table \ref{tab:advanced_kw} may be exploited by the more experienced user but can be safely
ignored by those who just want to get started. They may affect the convergence and accuracy of the final result, though.
For the inexperienced user however, we advise to not change these settings and accept the default values provided by \qcm.

% \vspace{3ex}

\begin{longtable}{ll}
\caption{Optional keywords for \qcm\ calculations.}\label{tab:advanced_kw}
\\
  \toprule
keyword & description \\
  \midrule
\endfirsthead

\multicolumn{2}{c}%
{{\bfseries \tablename\ \thetable{} -- continued from previous page}} \\
\toprule
keyword & description \\
\midrule
\endhead

\multirow{6}{*}{\texttt{orbital\_order}} & \multirow{6}{12cm}{Manual ordering of the orbitals along the one dimensional lattice. The order has to be entered as a string of comma separated orbital numbers. We recommend the Fiedler ordering based on the mutual information \cite{interface,fiedler} which can be obtained by means of the python script \texttt{fiedler.py} (see Section \ref{sec:pytools-qcm}).\\
	\textbf{default}: \texttt{orbital\_order = "1,2,3,4,5,6,\ldots"}} \\
& \\
& \\
& \\
& \\
& \\[\tableoptionskip]
\multirow{4}{*}{\texttt{sweep\_bond\_dimensions}} & \multirow{4}{12cm}{Sets \texttt{max\_bond\_dimension}\ for each sweep separately.\\
	\emph{Note}: Replaces \texttt{max\_bond\_dimension} which does \emph{NOT} need to be specified in this case.\\
	Example (\texttt{nsweeps=3}): \texttt{sweep\_bond\_dimensions}=``300,400,500"} \\
& \\
& \\
& \\[\tableoptionskip]
\multirow{2}{*}{\texttt{init\_bond\_dimension}} & \multirow{2}{12cm}{Adjusts the maximal bond dimension of the MPS produced by the
CI-DEAS procedure \cite{Legeza2003,fiedler,interface}.} \\
& \\[\tableoptionskip]
\multirow{7}{*}{\texttt{conv\_thresh}} &\multirow{7}{12cm}{Sets the energy convergence threshold (in Hartree). If the lowest energy from the previous sweep differs from the lowest energy of the current sweep by less than \texttt{conv\_thresh}, the DMRG calculation stops.\\
		\emph{Note}: Requires to set also \texttt{truncation\_final} and \texttt{ietl\_jcd\_tol}.\\
	{Numerical format}: \texttt{xe-y} with \texttt{x}\ and \texttt{y} being integers. \\
	Example: \texttt{conv\_thresh = 1e-6}.}\\
& \\
& \\
& \\
& \\
& \\
& \\[\tableoptionskip]
\multirow{3}{*}{\texttt{ietl\_jcd\_tol}} & \multirow{3}{12cm}{Convergence threshold for the Jacobi-Davidson diagonalization.\\
	{Numerical format}: \texttt{xe-y} with \texttt{x}\ and \texttt{y} being integers. \\
	Example: \texttt{ietl\_jcd\_tol = 1e-6}.} \\
& \\
& \\[\tableoptionskip]
\multirow{2}{*}{\texttt{ietl\_jcd\_maxiter}} & \multirow{2}{12cm}{Maximum number of iterations in the Jacobi-Davidson diagonalization.} \\
& \\ \pagebreak
\multirow{6}{*}{\texttt{truncation\_initial}} & \multirow{6}{12cm}{If during the \texttt{ngrowsweeps}, the sum of the discarded singular values for $m$ retained states is lower than the value defined here, more block states will be discarded until the discarded sum increases to \texttt{truncation\_initial}.\\
		{Numerical format}: \texttt{xe-y} with \texttt{x}\ and \texttt{y} being integers. \\
		Example: \texttt{truncation\_initial = 1e-6}.}\\
& \\
& \\
& \\
& \\
& \\[\tableoptionskip]
\multirow{6}{*}{\texttt{truncation\_final}} & \multirow{6}{12cm}{If during the \texttt{nmainsweeps}, the sum of the discarded singular values for $m$ retained states is lower than the value defined here, more block states will be discarded until the discarded sum increases to \texttt{truncation\_final}.\\
	{Numerical format}: \texttt{xe-y} with \texttt{x}\ and \texttt{y} being integers. \\
	Example: \texttt{truncation\_final = 1e-6}.}\\
& \\
& \\
& \\
& \\
& \\[\tableoptionskip]
\multirow{2}{*}{\texttt{measure\_each}} &\multirow{2}{12cm}{Tells the program to compute the expectation values every\\$2\ \times$
	\texttt{measure\_each} sweeps.} \\
& \\[\tableoptionskip]
\multirow{10}{*}{\texttt{symmetry}} & \multirow{10}{11cm}{Defines the total symmetry group. Default is the combined spin- (SU2) and point
symmetry group (PG) \texttt{su2u1pg}, where the \texttt{pg}-suffix should be omitted for better performance if the molecule is C1-symmetric.
For test purposes, it is possible to switch off spin-adaptation, again with or without point group symmetry: \texttt{2u1(pg)} .
In the latter case the keywords \texttt{spin} and \texttt{nelec} (see Section \ref{sec:compul-qcm}) have no meaning.
Instead \texttt{u1\_total\_charge1} and \texttt{u1\_total\_charge2} corresponding to the number of up and down electrons have to be specified.} \\
& \\
& \\
& \\
& \\
& \\
& \\
& \\
& \\
& \\[\tableoptionskip]
\multirow{2}{*}{\texttt{chkp\_each}} & \multirow{2}{11cm}{Tells the program to update the checkpoint file every\\ $2\ \times$
\texttt{chkp\_each} sweeps.} \\
& \\[\tableoptionskip]
\multirow{9}{*}{\texttt{hf\_occ}} & \multirow{9}{11cm}{Occupation of the starting orbitals (e.g. Hartree-Fock occupation) to be
entered as a comma separated string of occupation aliases. The aliases are defined as follows: 4~=~full, 3~=~up,
2~=~down, 1~=~empty. This information has to be entered in case of \texttt{hf} as \texttt{init\_state} and for the
CI-DEAS procedure \cite{Legeza2003,fiedler,interface} as invoked by \texttt{dmrginit.py}.\\
Example: \texttt{hf\_occ = "4,4,4,2,2,1"}\ for a CAS(8,6) triplet state setup where the unpaired electrons are ``located"
in orbital \# 4 and \# 5.}\\
& \\
& \\
& \\
& \\
& \\
& \\
& \\
& \\[\tableoptionskip]
\multirow{2}{*}{\texttt{nmainsweeps}} & \multirow{2}{11cm}{Number of sweeps in which \texttt{truncation\_final} is used in the singular value decomposition.} \\
& \\[\tableoptionskip]
\multirow{2}{*}{\texttt{ngrowsweeps}} &\multirow{2}{11cm}{Number of sweeps in which \texttt{truncation\_initial} is used in the singular value decomposition.}\\
& \\[\tableoptionskip]
\texttt{storagedir} &Scratch directory for temporary files.\\
\bottomrule
\end{longtable}

\subsubsection{Keywords for expectation value calculations}\label{sec:prop-qcm}

\qcm\ can (in principle) compute expectation values for any one- or two-particle operator that can be formulated in
second quantization. Table \ref{tab:expval}\ comprises a list of the available property keywords available in \qcm.
% For further updates/other properties please contact \href{mailto:dmrg@phys.chem.ethz.ch}{dmrg@phys.chem.ethz.ch}.

The one-particle reduced density matrix as well as the one-particle spin-density matrix are implicitly computed from the
expectation values of some of the operators contained in \texttt{MEASURE[ChemEntropy]}.

If a property is available only in 2U1 symmetry, the SU2U1 checkpoint may be transformed into 2U1 group with the \texttt{mps\_transform(\_pg)} tool (see Section~\ref{sec:tools-qcm})

\begin{longtable}{ll}
 \caption{\label{tab:expval}Expectation value calculations available in the release version of \qcm.}
\\
  \toprule
keyword & description \\
  \midrule
\endfirsthead

\multicolumn{2}{c}%
{{\bfseries \tablename\ \thetable{} -- continued from previous page}} \\
\toprule
keyword & description \\
\midrule
\endhead
% \begin{tabular}{p{5.6cm}l}

\multirow{3}{*}{\texttt{MEASURE[ChemEntropy]}} & \multirow{3}{11cm}{All expectation values over the operators required to
calculate the mutual information (as specified in Ref.~\cite{entanglement}) will be computed. Available only for SU2 symmetry.} \\
 & \\
 & \\[\tableoptionskip]
\multirow{3}{*}{\texttt{MEASURE[1rdm]}} & \multirow{3}{11cm}{Computes the one-particle reduced density matrix, without the additional correlators
contained in the \texttt{ChemEntropy} measurement.}\\
 & \\
 & \\[\tableoptionskip]
\multirow{10}{*}{\begin{minipage}{5cm}
                 \texttt{MEASURE[2rdm]}\\
                 \texttt{MEASURE[3rdm]}\\
                 \texttt{MEASURE[4rdm]}\\
                \end{minipage}
                }& \multirow{10}{11cm}{Computes the two-particle, three-particle and four-particle reduced density matrix, respectively. In case of 4-RDM, \texttt{MEASURE[4rdm]="p4:p3:p1:p2@\textit{l,k,i,j}"} allows to compute only a 4-index sub-block of the 4-RDM with the indices \texttt{\textit{i,j,k,l}}. This functionality is used in the distributed RDM calculations described in Section~\ref{sec:4rdm-distributed}. Analogously, for 3-RDM a two-index subblock may be calculated with \texttt{MEASURE[3rdm]="p1:p2@\textit{j,i}"}.\\ 3-RDM and 4-RDM calculations are available only in 2U1(pg) symmetry.}\\
& \\
& \\
&\\
&\\
&\\
&\\
&\\
&\\
&\\[\tableoptionskip]
\multirow{6}{*}{\begin{minipage}{5.6cm}
%                  \texttt{MEASURE[trans1rdm]="\textit{chk}"}\\ %% trans1rdm_aa, _bb, _ab, _ba ??
                 \texttt{MEASURE[trans2rdm]="\textit{chk}"}\\
                 \texttt{MEASURE[trans3rdm]="\textit{chk}"}\\
                \end{minipage}} & \multirow{6}{11cm}{Computes the transition two- or three-particle reduced density matrix with the current state and the state specified in the \texttt{\textit{chk}} checkpoint file. For t-3RDM calculations, two-index subblocks can be calculated with \texttt{MEASURE[trans3rdm]="\textit{chk};p1:p2@\textit{j,i}"}. Available only for 2U1(pg) symmetry.}\\
 & \\
 & \\
 & \\
 & \\
 & \\ \pagebreak
\multirow{4}{*}{\texttt{MEASURE\_LOCAL[\textit{name}]="\textit{op}"}} &
            \multirow{4}{11cm}{Computes $\langle \psi | op_i | \psi \rangle, \, i = 1 \ldots L$. \texttt{Nup}, \texttt{Ndown} and \texttt{Nup*Ndown}
            are meaningful choices for \textit{op}. Available for \texttt{2u1(pg) only.}\\
            \emph{Note}: \texttt{\textit{name}}\ defines the name under which the expectation values are stored on the \texttt{resultfile}.}\\
 & \\
 & \\
 & \\[\tableoptionskip]
\multirow{8}{5.6cm}{\texttt{MEASURE\_HALF}\texttt{\_CORRELATIONS} \texttt{[\textit{name}]="\textit{op$_1$:op$_2$}"}} &
            \multirow{8}{11cm}{Computes $\langle \psi | op_{1i} op_{2j}| \psi \rangle, \, i = 1 \ldots L, j=i+1 \ldots L$.
            \texttt{Nup}, \texttt{Ndown}, \texttt{Nup*Ndown}, \texttt{cdag\_up}, \texttt{cdag\_down}, \texttt{c\_up}, \texttt{c\_down},
            \texttt{cdag\_up*Ndown},
            \texttt{c\_up*Ndown},
            \texttt{cdag\_down*Nup},
            \texttt{c\_down*Nup},
            \texttt{cdag\_up*cdag\_down},
            \texttt{c\_up*c\_down},
            \texttt{cdag\_up*c\_down},
            and \texttt{cdag\_down*c\_up},
            as \textit{op$_1$} and \textit{op$_2$} are recognized by the program. Available for \texttt{2u1(pg)} only.\\
            \emph{Note}: \texttt{\textit{name}}\ defines the name under which the expectation values are stored on the \texttt{resultfile}.}\\
%\multirow{2}{*}{= "\textit{op$_1$:op$_2$}"}} & \\
 & \\
 & \\
 & \\
 & \\
 & \\
 & \\
 & \\
\bottomrule
\end{longtable}
\FloatBarrier

\subsection{QCMaquis input example}\label{sec:input-qcm}

Listing~\ref{lst:qcm} provides an input example for a DMRG-CASCI calculation. It assumes that an integral file \texttt{FCIDUMP} has been generated and placed in the same directory with the input file. The mandatory keywords are indicated in red, and the optional keywords in blue. The \texttt{FCIDUMP} for this particular example has been generated by the \mol-\qcm{} interface with the \texttt{001.input} input file from the \mol-\qcm{} test directory.

\begin{lstlisting}[language=qcmaquis,caption={Input example for a DMRG-CASCI(8,8) standalone \qcm{} calculation},label=lst:qcm]
 //active-space dependent parameters
L     = 8
nelec = 8
// target symmetry (spin and spatial) of the wave function:
// 2*spin = 0 (singlet) + totally symmetric point group irrep
spin  = 0
irrep = 0
//parameters to control the actual DMRG calculation
nsweeps = 8
max_bond_dimension = 256
conv_thresh = 1e-6
// initialization procedure
init_state = 'default'
// technical parameters
symmetry = 'su2u1pg'
integral_cutoff = 1e-40
truncation_initial = 1e-50
truncation_final = 1e-7
chkpfile = 'chkp.h5'
resultfile = 'result.h5'
integral_file = "FCIDUMP"
storagedir = '/scratch/$USER/boundaries'
LATTICE = "orbitals"
lattice_library = "coded"
MODEL = "quantum_chemistry"
model_library = "coded"
//all expectation value calculations required for entropy measures
MEASURE[ChemEntropy]
\end{lstlisting}

%\clearpage
\clearpage
\newpage
\section{Additional Considerations}

\subsection{Files required and written by \qcm}\label{sec:file-req-qcm}

\qcm\ requires only two files to start a calculation:
\begin{enumerate}
 \item an input file, see Listing~\ref{lst:qcm} for example
 \item an integral file in the \texttt{FCIDUMP} format as described in Ref. \citenum{fcidump}.
\end{enumerate}

When used with the \qcm{}-\mol{} interface, both files are produced by the interface from the \mol{} input.

Besides standard output, \qcm\ produces two types of files:
\begin{enumerate}
 \item a HDF5 \emph{result file} in which all information on, e.g. the energies and expectation values is stored. The file name is specified as the \kwd{resultfile} option in \qcm{} input file. In \qcm{}-\mol{} calculations, the result files are named \texttt{\$Project.result\_state.}(state\#)\texttt{.h5}, and are stored in the same directory as the \mol{} input file. Note that state numbering begins with 0.
 \item a \emph{checkpoint folder} which contains the matrix product state (MPS) wave function. The checkpoint folder name is controlled with the \kwd{chkpfile} option in \qcm{} input file, and In \qcm{}-\mol{} calculations, the checkpoint folders are named \texttt{\$Project.checkpoint\_state.}(state\#)\texttt{.h5}. The checkpoint is required for a later restart of the calculation, while the tools and the analysis scripts
described in Sections \ref{sec:tools-qcm} and \ref{sec:pytools-qcm}, respectively, require either a \texttt{result-file} or a
\texttt{checkpoint-folder} or both.
\end{enumerate}

\subsection{Typical workflow for a standalone QCMaquis DMRG calculation}\label{sec:workflow-qcm}
The usual workflow to set up, run and analyze a DMRG calculation proceeds as follows:
\begin{itemize}
   \item prepare an FCIDUMP integral file using the \mol-\qcm\ host
   program driver \cite{interface}) starting from a set of previously computed reference orbitals
   \item prepare an input file \texttt{dmrg-input} starting for example from the template input file provided in Listing~\ref{lst:qcm} and adjust all molecular system and wave function specific parameters (for example \texttt{nelec, spin,
   irrep, L, \ldots}, see Sections \ref{sec:compul-qcm}-\ref{sec:prop-qcm} for a list of all compulsory and optional input arguments)
   \item run the DMRG calculation with\\
\texttt{dmrg dmrg-input (| tee out)}
   \item compute expectation values with\\
\texttt{dmrg\_meas dmrg-input}
   \item analyze the results using the \textsc{Python} tools provided with the \qcm\ package (see Sections \ref{sec:tools-qcm}\ and
   \ref{sec:pytools-qcm}, respectively, for a list of utility programs and scripts)
\end{itemize}

\subsection{Memory management and memory requirements}\label{sec:memory-qcm}

Note that if you run a \mol-\qcm\ DMRG calculation, the memory consumption and memory allocation
of \qcm\ is entirely \emph{disconnected}\ from that of \mol. This means that the maximum memory specified within \mol{} (e.\,g.\ with the \texttt{MOLCAS\_MEM} environment variable) has no influence on the \qcm{} memory usage. For DMRG-SCF calculations, the \mol{} memory usage may be kept low (i.\,e. at around 4 GB regardless of the size of the active space), since CI vector expansions (which grow roughly factorially with the number of electrons and $N_\text{act}$) need not be stored in memory anymore. On the other hand, the calculation of the reduced two-particle density matrix is quite memory-intensive with the current implementation: for a 20 orbital active space it currently requires $\approx$16\,GB of memory, and an active space of 72 orbitals requires about 900\,GB of memory. (Theoretically, the memory requirement scales as $N_\text{act}^4$, but with a large prefactor). Note that in a QCMaquis standalone run, the memory problem may be alleviated by running \texttt{dmrg\_meas} on a different machine.

% \subsection{\qcm\ tools}\label{sec:tools-qcm}
%
% \qcm\ comes with several tools that allow for example further manipulation of the MPS or to acquire additional wave function analysis information.
% The tools \texttt{det2mps\_"symmetry"}, \texttt{mps2ci} and \texttt{mps\_transform(\_"pg")} will be briefly described in the following.
% These tools are provided in the \texttt{\molbuild/qcmaquis/bin} folder.
% \begin{table}[h]
%   \caption{Overview of \qcm\ tools.}
% \begin{tabular}{ll}
%       \toprule
%       {tool} & {description} \\
%       \midrule
%
%                         \multirow{6}{*}{\texttt{mps\_transform(\_pg)}} & \multirow{6}{11cm}{This tool allows for a transformation of an \texttt{su2u1(pg)} MPS wave function to \texttt{2u1(pg)} symmetry.\\
%                         	\emph{Command line}:\\ \texttt{mps\_transform(\_pg) chkpfile}\\
%                         	\emph{Note}: for an \texttt{su2u1(pg)} MPS with S\ $ > 0$\  \texttt{2u1(pg)} MPSs for all S$_z$ components are generated.}\\
%                         & \\
%                         & \\
%                         & \\
%                         & \\
%                         & \\
%                         \cmidrule(rl){1-2}
%       \multirow{14}{*}{\texttt{det2mps\_{symmetry}}} & \multirow{14}{11cm}{This tool generates determinants based on the
%       CI-DEAS procedure \cite{Legeza2003,fiedler} and inserts them in an MPS from which a new DMRG calculation can be started. Starting from such an MPS is likely to  improve convergence behaviour and is less prone to get stuck in local minima. The current implementation is described in Ref. \cite{interface}. The number of determinants will be chosen such that the maximal bond order at any site does not exceed the value set in \texttt{init\_bond\_dimension}. The new MPS
%       will be stored in the checkpoint folder specified in \texttt{chkpfile}.\\
%       \emph{Command line}:\\ \texttt{det2mps\_"symmetry" dmrg-input} \\\emph{Note}:  "\texttt{symmetry}" must equal the total symmetry specified in the input file \texttt{dmrg-input}.} \\
%       & \\
%      & \\
%      & \\
%      & \\
%      & \\
%      & \\
%      & \\
%      & \\
%      & \\
%      & \\
%      & \\
%      & \\
%      & \\
%       \cmidrule(rl){1-2}
%            \multirow{6}{*}{\texttt{mps\_overlap\_{symmetry}(\_pg)}} & \multirow{6}{11cm}{This tool calculates the overlap
%            	between two MPS $\Theta_\mathrm{1}$\ and $\Theta_\mathrm{2}$, respectively, according to $\langle \Theta_\mathrm{1} | \Theta_\mathrm{2} \rangle $.\\ \emph{Command line}:\\
%            	\texttt{mps\_overlap\_"{symmetry}"(\_pg) chkpfile\_1 chkpfile\_2}\\
%            	\emph{Note}:  "\texttt{symmetry}" must equal the full symmetry specified in the input file \texttt{dmrg-input}.} \\
%            & \\
%            & \\
%            & \\
%            & \\
%            & \\
%             \cmidrule(rl){1-2}
% \multirow{11}{*}{\texttt{mps2ci\_2u1(pg)}} & \multirow{11}{11cm}{Given a text file \texttt{determant\_list.txt}\ containing a list of determinant strings, this tool
% 					  calculates the CI-coefficients of the respective determinants
% 					  \cite{determinants}. The determinant strings have to encode the occupation of the orbitals as described for the \texttt{hf\_occ} keyword (4~=~full, 3~=~up, 2~=~down, 1~=~empty).\\
% 					  \emph{Command line}:\\ \texttt{mps2ci\_2u1(pg) chkpfile determinant\_list.txt}\\
% 					  \emph{Note}: with the conversion tool \texttt{mps\_transform(\_pg)} this analysis becomes possible also for MPS of the full symmetry \texttt{su2u1(pg)}.}\\
%                                                & \\
%                                                & \\
%                                                & \\
%                                                & \\
%                                                & \\
%                                                & \\
%                                                & \\
%                                                & \\
%                                                & \\
%                                                & \\
%       \bottomrule
% \end{tabular}
% \end{table}
% \FloatBarrier
\subsection{\qcm\ \text{python} scripts for wave function analysis and visualization}\label{sec:pytools-qcm}

The python scripts of \qcm\ are helpful to analyze and visualize the results of a DMRG calculation. Their usage should be evident from the documentation strings in the \textsc{Python} files.

The only script currently supported, \texttt{mutinf.py} is described below. Old scripts described in the previous versions of the manual have been moved to the\\ \kwd{dmrg/python/legacy} subdirectory, are deprecated and require Python 2.
%%% Commented out the documentation of old scripts until they get ported to Python 3
% However, those that are most frequently used will be briefly explained here. All scripts except \texttt{dmrginit.py}\ take the \qcm\ HDF5 result file as input. They are located in the folder\\
% \texttt{\$MOLCAS/qcmaquis/lib/python/pyeval}.
\begin{longtable}{ll}
  \caption{Overview of \qcm\ \textsc{Python} analysis and visualization scripts.}\\
      \toprule
      {script} & {description} \\
      \midrule
            \endfirsthead
            \multicolumn{2}{c}%
            {{\bfseries \tablename\ \thetable{} -- continued from previous page}} \\
            \toprule
            script & description \\
            \midrule
            \endhead
%                   \multirow{5}{*}{\texttt{sweeps.py}} & \multirow{5}{11cm}{Plots the energy for each microiteration.\\
%                   	\emph{Command line}:\\ \texttt{sweeps.py resultfile}\\
%                   	\emph{Note}: use this tool to check the convergence wrt the number of sweeps.} \\
%                   & \\
%                   & \\
%                   & \\
%                   & \\
%                   \cmidrule(rl){1-2}
%       \multirow{11}{*}{\texttt{dmrginit.py}} & \multirow{11}{11cm}{Starts a \qcm\ DMRG calculation with \texttt{max\_bond\_dimension =
%       200} and \texttt{nsweeps = 2}, measures the entropy information  \cite{entropy, entanglement} from this unconverged calculation and based on this, generates a new MPS with the CI-DEAS procedure according to all settings specified in the input file \texttt{dmrg-input}. We recommend to use this script for the preparation of calculations for active spaces that are larger than those that can be handled with traditional CAS methods.\\
%        \emph{Command line}:\\ \texttt{dmrginit.py dmrg-input}
%        } \\
%       & \\
%       & \\
%       & \\
%       & \\
%       & \\
%       & \\
%       & \\
%       & \\
%       & \\
%       & \\
%       \cmidrule(rl){1-2}
%             \multirow{10}{*}{\texttt{fiedler.py}} & \multirow{10}{11cm}{Optimizes the ordering based on entropy information as proposed in Ref. \citenum{fiedler}. The current implementation is described in Ref. \citenum{interface}. The ordering ensures that highly entangled orbitals are close to each other in the one dimensional lattice. The first ordering in the output ignores the point group symmetry of the orbitals, while the second version orders the orbitals within each irreducible representation and then reorders these symmetry blocks.\\
%             	\emph{Command line}:\\ \texttt{fiedler.py resultfile}\\
%             	\emph{Note}: we recommend to use the second option.} \\
%             & \\
%             & \\
%             & \\
%             & \\
%             & \\
%             & \\
%             & \\
%             & \\
%             & \\
%             & \\
%             \cmidrule(rl){1-2}
%                                                 \multirow{4}{*}{\texttt{input.py}} & \multirow{4}{11cm}{This script recovers the complete \qcm\ input file \texttt{dmrg-input} from a given \texttt{resultfile}.\\
%                                                 	\emph{Command line}:\\ \texttt{input.py resultfile}}\\
%                                                 & \\
%                                                 & \\
%                                                 &\\
%                                                 \cmidrule(rl){1-2}
      \multirow{9}{*}{\texttt{mutinf.py}} & \multirow{9}{11cm}{Produces mutual information plots \cite{entanglement}\ given that an expectation value calculation for \texttt{MEASURE[ChemEntropy]}\ has been performed.\\
      \emph{Usage:} \texttt{mutinf.py resultfile}\\
      \emph{Note}: if orbital images (in .png format) named  1.png, 2.png, \ldots, L.png (with L being the number of active orbitals) are present in the same folder where \texttt{mutinf.py}\ is executed, they can be added to the mutual information plot by providing the optional argument \texttt{-i}  to \texttt{mutinf.py}.}\\
      & \\
      & \\
      & \\
      & \\
      & \\
      & \\
      & \\
      & \\
      \bottomrule
\end{longtable}

\clearpage
\newpage

\bibliography{qcmaquis_manual}

\end{document}

### 2021-02-22 dmrg/doc/qcmaquis_manual.tex

@@
 keywords=[1]{GATEWAY,SEWARD,SCF,DMRGSCF,MPSSI},
-morekeywords=[2]{ActiveSpaceOptimizer,DMRGSettings,EndDMRGSettings,OOptimizationSettings,EndOOptimizationSettings,NEVPT2Prep,EvRDM,HDF5,DMRG,EJOB,SPIN,AMFI,Angmom},
+morekeywords=[2]{ActiveSpaceOptimizer,DMRGSettings,EndDMRGSettings,OOptimizationSettings,EndOOptimizationSettings,NEVPT2Prep,HDF5,DMRG,EJOB,SPIN,AMFI,Angmom},
 keywordstyle=[1]\color{green!20!red},
@@
 
-First, one performs a DMRG-SCF calculation with the keywords \kwd{NEVPT2Prep} and (optionally) \kwd{EvRDM}, which enable the evaluation of the 4-RDMs (and, in case of multiple states, also t-3RDMs) required by NEVPT2.
-If \kwd{EvRDM} is specified, the RDMs are evaluated immediately, otherwise an input file is prepared for separate RDM calculations with QCMaquis (see also Table~\ref{table:nevpt2-keyword}).
-The latter option is particularly useful for large active spaces, where offloaded evaluation can be performed in a massively parallel manner, which will be explained in detail in Section~\ref{sec:4rdm-distributed}.
-
-\begin{framed}
- \noindent Note: The computational cost of the RDM evaluation grows as $N^8$ with the number of active orbitals, therefore we strongly recommend to use the distributed RDM evaluation for active spaces larger than 10-11 orbitals described in Section~\ref{sec:4rdm-distributed}.
+First, one performs a DMRG-SCF calculation with the keyword \kwd{NEVPT2Prep}, which enables the evaluation of the 4-RDMs (and, in case of multiple states, also t-3RDMs) required by NEVPT2.
+The RDMs can be evaluated separately in a massively parallel manner, which is explained in detail in Section~\ref{sec:4rdm-distributed}.
+
+\begin{framed}
+ \noindent Note: The computational cost of the RDM evaluation grows as $N^8$ with the number of active orbitals, therefore we strongly recommend to use the distributed RDM evaluation for active spaces larger than 11-12 orbitals described in Section~\ref{sec:4rdm-distributed}.
 \end{framed}

### 2021-07-30 dmrg/doc/qcmaquis_manual.tex

@@
 \newpage
+
+\subsection{QCMaquis pre-Born--Oppenheimer module}\label{sec:preBO}
+
+The pre-Born--Oppenheimer (pre-BO) module within QCMaquis provides a tool to perform
+DMRG-CASCI calculations of \textit{multicomponent} systems consisting of multiple spin-$\frac12$
+particle types.
+
+QCMaquis follows the formatting of the integral file shown in
+Listing~\ref{lst:prebodump}. The first line with only a single float,
+contains a constant shift in energy, e.g., point-charge repulsion of classical
+nuclei.
+The following lines contain the terms in the Hamiltonian where pairs of numbers
+$m-n$ describe a single second-quantization operator. The first number denotes
+the particle type (starting with 0) and the second number the orbital index of
+the given particle type (also starting with 0 for all types). 
+We assume physics notation and normal-ordering, i.e., the following would be a
+valid term
+\begin{equation}
+  \left(\int \phi_{1-1}(1)\phi_{0-1}(2)g(1,2)\phi_{1-0}(1)\phi_{0-1}(2)\right)
+  a^\dagger_{1-1} a^\dagger_{0-1} a_{0-1} a_{1-0}\,.
+\end{equation}
+The second and third lines contain single-paricle type \textit{hopping} terms.
+The fourth line stands for a single-particle type interaction, and the fifth
+line for an interaction between different particle types. 
+Note here that the aforementioned convention implies that operators belonging to
+the same particle types in a given string of operators must be either at
+positions 1 and 4 or 2 and 3.
+
+\begin{lstlisting}[language=qcmaquis,caption={Pre-BO integral file
+format.},label=lst:prebodump]
+1.145901278714102                       
+   0-0         0-0                                      -2.451508889071943
+   1-0         1-0                                      2.30233235666604
+   0-2         0-0         0-1         0-1              -0.08690785254423253
+   1-1         0-1         0-1         1-0              4.887021698358933e-05
+\end{lstlisting}
+
+Listing~\ref{lst:qcm} shows an input example for a pre-BO DMRG-FCI calculation. 
+It assumes that a pre-BO integral file with has been generated. Additionally,
+the 1-particle RDM and the mutual information can be measured and printed to a
+file.
+
+\begin{lstlisting}[language=qcmaquis,caption={Input example for a pre-BO DMRG-FCI
+calculation},label=lst:preboinput]
+nsweeps                         = 6
+max_bond_dimension              = 1000
+
+// Pre-BO specific input
+// String with the number of parciles for each type
+PreBO_ParticleTypeVector        = "4 1"
+// Spin-1/2 fermion (1) or spin-0 boson (0)
+// Attention: the boson-related functionality has not been fully tested, yet.
+PreBO_FermionOrBosonVector      = "1 1"
+// Number of orbitals for each type.
+PreBO_OrbitalVector             = "3 3"
+// Number of alpha-beta particles for each type
+// Note: for spin-0 bosons it would be a single number.
+PreBO_InitialStateVector        = "2 2 1 0"
+
+optimization                    = twosite 
+symmetry                        = nu1
+
+model_library                   = coded
+lattice_library                 = coded
+LATTICE                         = "preBO lattice"
+
+// Your model
+MODEL                           = PreBO
+init_state                      = 'default'
+
+// Hamiltonian file
+integral_file                   = FCIDUMP.E_4_H1_1
+integral_cutoff                 = 1.0E-10
+
+MEASURE[1rdm]          = 1
+MEASURE[mutinf]        = 1
+\end{lstlisting}
+
+%\clearpage
+\clearpage
+\newpage
+
 \section{Additional Considerations}

### 2021-08-24 dmrg/doc/qcmaquis_manual.tex

@@
 
-\subsection{QCMaquis input example}\label{sec:input-qcm}
+\subsection{QCMaquis input example}
+\label{sec:input-qcm}
 
@@
 \begin{lstlisting}[language=qcmaquis,caption={Input example for a DMRG-CASCI(8,8) standalone \qcm{} calculation},label=lst:qcm]
- //active-space dependent parameters
+  // Lattice size (= number of orbitals)
   L     = 8
+  // Number of electrons
   nelec = 8
-// target symmetry (spin and spatial) of the wave function:
-// 2*spin = 0 (singlet) + totally symmetric point group irrep
+  // Target symmetry (spin and spatial) of the wave function:
+  // 2*spin = 0 (singlet)
   spin  = 0
+  // Wavefunction irreducible representation (0 == totally symmetric)
   irrep = 0
-//parameters to control the actual DMRG calculation
+  // Parameters to control the actual DMRG calculation
   nsweeps              = 8
@@
   conv_thresh          = 1e-6
-// initialization procedure
+  // MPS initialization
   init_state = 'default'
-// technical parameters
+  // Wave function symmetry (su2u1pg for spin-adapted DMRG + point group)
   symmetry             = 'su2u1pg'
+  // Threshold below which integrals inside the integral_file are neglected
   integral_cutoff      = 1e-40
-truncation_initial = 1e-50
+  // Threshold for the SVD truncation for the initial sweeps
+  truncation_initial   = 1e-16
+  // Threshold for the SVD truncation for the final sweeps
   truncation_final     = 1e-7
+  // HDF5 file where the MPS will be stored
   chkpfile             = 'chkp.h5'
+  // HDF5 file where the results of the DMRG calculations will be stored
   resultfile           = 'result.h5'
+  // File with the one- and two-body integrals
   integral_file        = "FCIDUMP"
+  // Path to the scratch folder
   storagedir           = '/scratch/$USER/boundaries'
+  // Lattice-related parameters (not to be changed for qc calculations)
   LATTICE              = "orbitals"
@@
   model_library        = "coded"
-//all expectation value calculations required for entropy measures
-MEASURE[ChemEntropy]
+  // List of measurements to be performed at the end of the optimization
+  MEASURE[ChemEntropy] = 1
 \end{lstlisting}
 
-%\clearpage
-\clearpage
-\newpage
+\subsection{\qcm\ examples library}
+\label{sec:examples}
+
+We provide in the folder \texttt{example} a set of example input files to run the different kinds of DMRG-based calculations that are supported by \qcm{}.
+These example calculations are organized in different folder, one per feature provided by \qcm, as listed below:
+
+\begin{itemize}
+ \item \texttt{iTD-DMRG}: example calculation (\texttt{H2\_2e4o.iTD.SS.inp}) of an imaginary-time DMRG calculation.
+The integral file is generated for the H$_2$ molecule and based on the 6-31G basis set.
+ Note that the calculation should be run with the executable \texttt{applications/evolve/dmrg\_evolve}, which is generated by setting the \texttt{BUILD\_DMRG\_EVOLVE} CMake flag to \texttt{ON}.
+ We provide also the input for running a conventional \texttt{DMRG} calculation (\texttt{H2\_2e4o.TI.SS.inp}), which must be run with the \texttt{applications/dmrg/dmrg} executable and should return the same energy as the iTD-DMRG simulation.
+ \item \texttt{TD-DMRG}: example calculation of a real-time DMRG calculation.
+The integral file is associated with the $\pi$ orbital space (CAS(6,6)) of benzene calculated with the 6-31G* basis set.
+As already noted above, the calculations should be run with the \texttt{applications/evolve/dmrg\_evolve} executable.
+ Note that the initial state for the propagation is a random MPS (the \texttt{init\_state} keyword is set to \texttt{default}).
+ Therefore, the calculation does not correspond to any physically meaningful non-equilibrium process. 
+ However, this calculation can be used to assess that the \texttt{dmrg\_evolve} executable has been compiled correctly by verifying that the energy is conserved along the propagation.
+ For this example calculation we provide the input files for both the single-site (\texttt{Benzene\_6e6o.TD.SS.inp}) and the two-site (\texttt{Benzene\_6e6o.TD.TS.inp}) propagators.
+\end{itemize}
+
+For all input files, we provide also the corresponding output file that can be used as a reference to ensure that the \qcm\ executables have been compiled correctly.
 
@@
 
-The pre-Born--Oppenheimer (pre-BO) module within QCMaquis provides a tool to perform
-DMRG-CASCI calculations of \textit{multicomponent} systems consisting of multiple spin-$\frac12$
-particle types.
-
-QCMaquis follows the formatting of the integral file shown in
-Listing~\ref{lst:prebodump}. The first line with only a single float,
-contains a constant shift in energy, e.g., point-charge repulsion of classical
-nuclei.
-The following lines contain the terms in the Hamiltonian where pairs of numbers
-$m-n$ describe a single second-quantization operator. The first number denotes
-the particle type (starting with 0) and the second number the orbital index of
-the given particle type (also starting with 0 for all types). 
-We assume physics notation and normal-ordering, i.e., the following would be a
-valid term
+The \qcm\ pre-Born--Oppenheimer (pre-BO) module provides a tool to perform DMRG-CASCI calculations for \textit{multicomponent} Hamiltonians consisting of multiple spin-$\frac12$ particle types.
+
+QCMaquis follows the formatting of the integral file shown in Listing~\ref{lst:prebodump}.
+The first line, with only a single float, contains a constant shift in energy, e.g., point-charge repulsion of classical nuclei.
+The following lines contain the terms in the Hamiltonian where pairs of numbers $m-n$ describe a single second-quantization operator.
+The first number denotes the particle type (starting with 0) and the second number the orbital index associated with the given particle type (also starting with 0 for all types).
+We assume physics notation and normal-ordering, i.e., the following would be a valid term
+
 \begin{equation}
@@
 \end{equation}
+
 The second and third lines contain single-paricle type \textit{hopping} terms.
@@
 \qcm\ requires only two files to start a calculation:
+
 \begin{enumerate}
@@
 
-\subsection{Typical workflow for a standalone QCMaquis DMRG calculation}\label{sec:workflow-qcm}
+\subsection{Typical workflow for a standalone QCMaquis DMRG calculation}
+\label{sec:workflow-qcm}
+
 The usual workflow to set up, run and analyze a DMRG calculation proceeds as follows:
+
 \begin{itemize}

### 2022-02-15 dmrg/framework/dmrg/models/chem/cideas/ci_generator.hpp
Include vDMRG code

@@ -33,7 +33,7 @@
 #include <algorithm>
 #include <alps/numeric/matrix.hpp>
 
-#include "dmrg/models/lattice.h"
+#include "dmrg/models/lattice/lattice.h"
 #include "dmrg/models/measurements/chementropy.h"
 #include "determinant.hpp"
 

### 2022-02-15 dmrg/framework/dmrg/models/generate_mpo/1D_mpo_maker.hpp

@@ -4,6 +4,7 @@
  *
  * Copyright (C) 2015 Laboratory of Physical Chemistry, ETH Zurich
  *               2015-2015 by Sebastian Keller <sebkelle@phys.ethz.ch>
+ *               2021 by Alberto Baiardi <abaiardi@ethz.ch>
  *
  * This software is part of the ALPS Applications, published under the ALPS
  * Application License; you can use, redistribute it and/or modify it under
@@ -26,6 +27,14 @@
 
 #ifndef ONED_MPO_MAKER_H
 #define ONED_MPO_MAKER_H
+
+#include <utility>
+#include "dmrg/mp_tensors/mpo.h"
+#include "dmrg/models/model.h"
+#include "dmrg/models/term_descriptor.h"
+#include "dmrg/models/generate_mpo/utils.hpp"
+#include "dmrg/models/lattice/lattice.h"
+#include "dmrg/models/OperatorHandlers/OpTable.h"
 
 namespace generate_mpo
 {
@@ -224,6 +233,32 @@ namespace generate_mpo
         term_descriptor<typename Matrix::value_type> term = arrange_operators(positions, operators, tag_handler);
         term.coeff *= scale;
         return sign_and_fill(term, ident, fill, tag_handler, lat);
+    }
+
+    /** @brief Enum class used in the following class */
+    enum class IonizedOrbital { Up, Down };
+
+    /**
+     * @brief Routine that builds the MPO representation of the destruction operator.
+     * TODO: Check if it works also for the TwoU1 and SU2U1 symmetry.
+     */
+    template<class Matrix, class SymmGroup>
+    MPO<Matrix, SymmGroup>
+    make_destroy_mpo(const Lattice& lat, Model<Matrix, SymmGroup>& model, int iOrb, IonizedOrbital alphaOrBeta=IonizedOrbital::Up)
+    {
+      using tag_type = typename OPTable<Matrix, SymmGroup>::tag_type;
+      std::vector<int> pos(1, iOrb);
+      auto name = alphaOrBeta == IonizedOrbital::Up ? "destroy_up" : "destroy_down";
+      tag_type opDestroy = model.get_operator_tag(name, lat.template get_prop<typename SymmGroup::subcharge>("type", pos[0]));
+      tag_type ops_[1] = {opDestroy};
+      std::vector<tag_type> ops(ops_, ops_+1);
[…]

### 2022-02-15 dmrg/framework/dmrg/utils/parallel/placement.hpp

@@ -4,6 +4,7 @@
  *
  * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
  *               2013-2013 by Alexandr Kosenkov <alex.kosenkov@gmail.com>
+ *               2019 by Alberto Baiardi <abaiardi@ethz.ch>
  * 
  * This software is part of the ALPS Applications, published under the ALPS
  * Application License; you can use, redistribute it and/or modify it under
@@ -29,113 +30,108 @@
 
 namespace parallel {
 
-    inline void make_consistent(std::vector<int>& p_new, int b2, std::vector<int>& e_new,
-                                const std::vector<int>& p_old, int b1, std::vector<int>& e_old)
-    {
-        // exceptions check
-        for(int i = 0; i < e_old.size(); i++) if(b1 == e_old[i]) return;
-        for(int i = 0; i < e_new.size(); i++) if(b2 == e_new[i]) return;
-        // common check
-        for(int i = 0; i < p_new.size(); i++){
-            if(p_new[i] == p_old[b1]){
-                e_old.push_back(b1);
-                for(int ii = i; ii < p_new.size(); ii++) 
-                if(p_new[ii] == p_old[b1]) p_new[ii] = -1;
-                return;
-            }
+inline void make_consistent(std::vector<int>& p_new, int b2, std::vector<int>& e_new,
+                            const std::vector<int>& p_old, int b1, std::vector<int>& e_old)
+{
+    auto it_old = std::find(e_old.begin(), e_old.end(), b1);
+    auto it_new = std::find(e_new.begin(), e_new.end(), b2);
+    if (it_old != e_old.end() && it_new != e_new.end()) {
+        auto it_inner = std::find(p_new.begin(), p_new.end(), p_old[b1]);
+        if (it_inner != p_new.end()) {
+            e_old.push_back(b1);
+            std::replace(p_new.begin(), p_new.end(), p_old[b1], -1);
         }
-        // overlap check
-        if(p_new[b2] != -1){
+        else if(p_new[b2] != -1) {
             e_new.push_back(b2);
             p_new[b2] = -1;
-            return;
-        }
-
-        p_new[b2] = p_old[b1];
-    }
-
-    inline void make_complete(std::vector<int>& placement){
-        for(int i = 0; i < placement.size(); i++)
-        if(placement[i] == -1){
-            for(int k = 0; k < placement.size(); k++){
-                bool found = false;
[…]

### 2022-02-15 dmrg/doc/qcmaquis_manual.tex

@@
 
+% -- Loads packages --
 \usepackage[nott]{kpfonts}
+\usepackage{amsmath}
+\usepackage{bm}
+\usepackage{booktabs}
+\usepackage[bf,singlelinecheck=off]{caption}
 \usepackage[T1]{fontenc}
+\usepackage{framed}
+\usepackage{graphicx}
+\usepackage[colorlinks=true]{hyperref} % web links
 \usepackage[utf8]{inputenc}
-% \usepackage[libertine]{newtxmath}
-\usepackage{graphicx}% Include figure files
+\usepackage{listings}
+\usepackage{longtable}
+\usepackage{multirow}
+\usepackage[numbers,sort&compress]{natbib}
+\usepackage{placeins}
 \usepackage{xcolor}
-\usepackage{bm}% bold math
-\usepackage[colorlinks=true]{hyperref} % web links
+
+% Definition of colours for hyperref
 \definecolor{linkcolour}{rgb}{0,0.2,0.6}
@@
 \hypersetup{colorlinks,breaklinks,urlcolor=linkcolour,linkcolor=linkcolour,citecolor=citecolour}
-\usepackage{framed}
-\usepackage{longtable}
-\usepackage[bf,singlelinecheck=off]{caption}
-\usepackage{amsmath}
-\usepackage{listings}
-%\lstset{breaklines=true}
-\usepackage{multirow}
-% \usepackage[top=2.25cm, bottom=2.25cm, left=2cm, right=2.25cm]{geometry}
-\usepackage{booktabs}
-\usepackage{chemformula}
-\usepackage{placeins}
-\usepackage[numbers,sort&compress]{natbib}
+
 \setcounter{secnumdepth}{4}
-% \renewcommand{\baselinestretch}{1.5}
+
+% -- New commands --
 \newcommand{\relv}{release version 3.0}
@@
 \newcommand{\qcmbuild}{\texttt{my-QCMaquis-build}}
-%\setlength{\parindent}{1.0cm}
-%\setlength{\parskip}{1.00ex plus0.5ex minus0.5ex}
-% \setlength{\abovecaptionskip}{2.00ex plus0.25ex minus0.25ex}
-% \setlength{\belowcaptionskip}{\abovecaptionskip}
-%\setlength{\textfloatsep}{3.0ex plus0.5ex minus0.5ex}
-%\setlength{\intextsep}{3.0ex plus0.5ex minus0.5ex}
+\newcommand{\kwd}[1]{\texttt{#1}}
 
@@
 \newcommand{\myemph}[1]{\textbf{#1}}
-
-% \usepackage{bibentry}
-%% required for the bibentry package
 
@@
 
-\lstset{numbers=left,numberstyle=\tiny,numberblanklines=true,basicstyle=\ttfamily\small,frame=single,showlines=true,flexiblecolumns=true,breaklines=true,postbreak={\llap{\textcolor{red}{$\lhook\joinrel\longrightarrow$\kern2.1em}}},escapechar=|}
-
-\newcommand{\todo}[1]{\textcolor{red}{#1}}
-\newcommand{\kwd}[1]{\texttt{#1}}
+\lstset{
+	numbers=left,
+	numberstyle=\tiny,
+	numberblanklines=true,
+	basicstyle=\ttfamily\small,
+	frame=single,
+	showlines=true,
+	flexiblecolumns=true,
+	breaklines=true,
+	postbreak={\llap{\textcolor{red}{$\lhook\joinrel\longrightarrow$\kern2.1em}}},
+	escapechar=|
+}
 
 \title{A quick user guide to the\\ \scine-\qcm\ software suite for \mol}
-\author{Leon Freitag, Sebastian Keller, Stefan Knecht, Yingjin Ma,\\ Christopher Stein and Markus Reiher}
+\author{Alberto Baiardi, Robin Feldmann, Leon Freitag, Sebastian Keller, \\ Stefan Knecht, Yingjin Ma, Christopher Stein and Markus Reiher}
 \publishers{ETH Z\"urich, Laboratorium f\"ur Physikalische Chemie, Vladimir-Prelog-Weg 2,\\ CH-8093 Z\"urich}
@@
 \vspace{2cm}
+
 \noindent We kindly request that, for reproducibility reasons, any use of the \qcm\ software suite for density matrix renormalization group (DMRG) calculations in \mol\ that results in published material should cite the set-up steered by settings and warm-up procedures described in:
+
 \begin{framed}
   \noindent {\textbf{Check for a preprint on \href{http://arxiv.org/}{arXiv.org} (to appear soon).}}\\
-	\noindent Freitag~L.; Keller~S.; Knecht~S.; Lindh~R.; Ma~Y.; Stein~C. J.; Reiher~M. \emph{in preparation}.
+  \noindent Freitag~L.; Knecht~S.; Lindh~R.; Ma~Y.; Stein~C. J.; Reiher~M. \emph{in preparation}.
 \end{framed}
 
-The DMRG calculations are then conducted with the software \qcm\ that requires a citation. It is described in the following paper:
+\noindent The DMRG calculations are then conducted with the software \qcm\ that requires a citation.
+It is described in the following paper:
+
 \begin{framed}
-%\noindent {\textbf{Check for the journal article asap on \href{http://arxiv.org/pdf/1510.02026.pdf}{arXiv.org}.}}\\
 \noindent Keller,~S.; Dolfi,~M.; Troyer,~M.; Reiher,~M. \emph{J. Chem. Phys.}
@@
 
+The reference works for the vibrational DMRG method are the following:
+
+\begin{framed}
+  \noindent Baiardi,~A.; Stein,~C.J.; Barone,~V.; Reiher,~M. \emph{J. Chem. Theory Comput.}
+  \textbf{2017}, \emph{13}, 3764,   \href{https://doi.org/10.1021/acs.jctc.7b00329}{doi:10.1021/acs.jctc.7b00329}. \\
+  \noindent Glaser,~N.; Baiardi~A.; Reiher,~M. ``Tensor Network States for Vibrational Spectroscopy'', \href{https://arxiv.org/abs/2109.08961}{arXiv.org}.
+\end{framed}
+
+The reference paper Nuclear-Electron All-Particle DMRG method (see Section~\ref{sec:preBO}) is the following:
+
+\begin{framed}
+  \noindent Muolo,~A.; Baiardi,~A.; Feldmann,~R.; Reiher,~M. \emph{J. Chem. Theory Phys.}
+	\textbf{2020}, \emph{152}, 204103, \href{https://doi.org/10.1063/5.0007166}{doi:10.1063/5.0007166}.
+\end{framed}
+
 \qcm\ builds upon the ALPS MPS project. 
-The \href{http://alps.comp-phys.org/static/mps\_doc/index.html}{ALPS MPS codes} implement the DMRG algorithm for variational ground and low-lying excited state search as well as time evolution of arbitrary one- and two-dimensional models in a matrix-product-state representation.
+The \href{http://alps.comp-phys.org/static/mps\_doc/index.html}{ALPS MPS code} implements the DMRG algorithm for variational ground and low-lying excited state search as well as time evolution of arbitrary one- and two-dimensional models in a matrix-product-state representation.
 They have been developed at ETH Zurich by Michele Dolfi and Bela Bauer in the group of Matthias Troyer with contributions from Sebastian Keller and Alexandr Kosenkov and at the University of Geneva by Timoth{\'e}e Ewart and Adrian Kantian in the group of Thierry Giamarchi.
@@
 ALPS is a general open-source framework for the description of strongly correlated many-particle systems.
+
 \begin{framed}
@@
 \pagenumbering{arabic}
+
 \section{Introduction to the \qcm\ Software Suite}
@@
 \subsection{Overview and Goals}
-The computational cost of traditional full CI and full CI-based multiconfigurational calculations (such as CASSCF) scales exponentially with the number of active orbitals, thus calculations beyond a certain limit (e.
-g.
-18 electrons in 18 orbitals for CASSCF) become prohibitively expensive.
-A possible approach to overcome the scaling limit is to employ methods such as density matrix renormalization group (DMRG) \cite{Schollwock_AnnPhys_densitymatrix_2011,Marti_ZPhysChem_Density_2010,Chan2011,Chan_JChemPhys_Matrix_2016} to solve the full CI problem approximately.
-
-In DMRG, usually the wave function is represented in form of a matrix product state
-(MPS). The \qcm\ software suite solves a full CI problem approximately by means of
-an efficient optimization of an MPS wave function based on a second-generation DMRG algorithm \cite{Keller_JChemPhys_efficient_2015}.
-The quantum-chemical operators are represented as matrix product operators (MPOs) which provides the necessary flexibility to accommodate Abelian and non-Abelian symmetries as well as the implementation of non-relativistic and relativistic quantum chemical Hamiltonians \cite{maquis-rel}, respectively,
-in a unified framework. We have implemented the special unitary group of degree 2 (SU(2)) in the MPO representation of the non-relativistic Hamiltonian to ensure
+
+The computational cost of traditional full CI and full CI-based multiconfigurational calculations (such as CASSCF) scales exponentially with the number of active orbitals.
+Calculations beyond a certain limit (e. g. 18 electrons in 18 orbitals for CASSCF) become, therefore, prohibitively expensive. 
+A possible approach to overcome the scaling limit is to employ methods such as density matrix renormalization group (DMRG) \cite{Schollwock_AnnPhys_densitymatrix_2011,Marti_ZPhysChem_Density_2010,Chan2011,Chan_JChemPhys_Matrix_2016,Baiardi2020_Review} to solve the full CI problem efficiently.
+
+In DMRG, the wave function is represented in form of a matrix product state (MPS). 
+The \qcm\ software suite solves the full CI problem approximately by means of an efficient optimization of an MPS wave function based on the so-called second-generation DMRG algorithm \cite{Keller_JChemPhys_efficient_2015}.
+The quantum-chemical operators are represented as matrix product operators (MPOs), which provides the necessary flexibility to accommodate Abelian and non-Abelian symmetries, as well as the implementation of non-relativistic and relativistic quantum chemical Hamiltonians \cite{maquis-rel}, respectively, in a unified framework.
+We leveraged the generality of the MPO/MPS-based framework to further extend \qcm\ to support Born--Oppenheimer\cite{Baiardi2017_vDMRG} and pre--Born--Oppenheimer\cite{Muolo2020_NEAP-DMRG} vibrational Hamiltonians.
+We have implemented the special unitary group of degree 2 (SU(2)) in the MPO representation of the non-relativistic Hamiltonian to ensure
 spin conservation \cite{Keller_JChemPhys_Spinadapted_2016}.
-The current implementation of \qcm\ allows for efficient full-CI-type calculations of active space sizes beyond capabilities (``CAS(18,18)") of standard CI approaches.
-
-The \qcm\ software suite comes with a Fortran/Python-based interface \cite{interface} to the \mol\ \cite{Aquilante_JComputChem_Molcas_2016} framework, where we have implemented a state-specific and state-average DMRG self-consistent field (DMRG-SCF) algorithm, the possibility to include solvent effects in DMRG calculations, analytic gradients for state-specific calculations, state interaction in a nonorthogonal basis\cite{Knecht_JChemTheoryComput_nonorthogonal_2016} and n-electron valence state perturbation theory employing the DMRG reference wave function.\cite{Freitag_JChemTheoryComput_Multireference_2017} These developments allow for, among others, structure optimizations, spin-orbit coupling calculations, evaluation of transition densities between non-orthogonal states for electronic and magnetic properties, as well as to account for dynamic correlation by means of the NEVPT2 approach.
+The current implementation of \qcm\ allows for efficient full-CI-type calculations of active space sizes well beyond capabilities (``CAS(18,18)") of standard CI approaches.
+
+The \qcm\ software suite comes with a Fortran/Python-based interface \cite{interface} to the \mol\ \cite{Aquilante_JComputChem_Molcas_2016} framework, where we have implemented state-specific and state-average DMRG self-consistent field (DMRG-SCF) algorithms, the possibility to include solvent effects in DMRG calculations, analytic gradients for state-specific calculations, state interaction in a nonorthogonal basis,\cite{Knecht_JChemTheoryComput_nonorthogonal_2016} and n-electron valence state perturbation theory employing the DMRG reference wave function.\cite{Freitag_JChemTheoryComput_Multireference_2017}
+These developments allow for, among others, structure optimizations, spin-orbit coupling calculations, evaluation of transition densities between non-orthogonal states for electronic and magnetic properties, as well as to account for dynamic correlation by means of the NEVPT2 approach.
 
@@
   \large{\textbf{Advice:} The break-even point wrt computational cost for a DMRG-CI/-SCF calculation compared to a ``traditional" CAS-CI/-SCF calculation is approximately reached for a CAS(14,14) space.
-For active spaces smaller than CAS(14,14) we recommend to choose a traditional \mol\ CAS approach.}
+For active spaces smaller than CAS(14,14) we recommend to choose the conventional \mol\ CAS algorithm.}
 \end{framed}
 
-\subsection{What features are included in this release?}\label{sec:features-v1.0}
+\subsection{What features are included in this release?}
+\label{sec:features-v1.0}
 
@@
 
-\subsection{Licenses}\label{sec:license}
+\subsection{Licenses}
+\label{sec:license}
+
 The \qcm{}-\mol{} interface and the NEVPT2 program are licensed under GNU LGPL, version 2.1.
 The \qcm{} software suite and the ALPS library are distributed under \href{http://alps.comp-phys.org/static/software/applications/LICENSE.txt}{ALPS Application License version 1.0} and the \href{http://alps.comp-phys.org/static/software/ALPS/LICENSE.txt}{ALPS Library License version 1.0}.
+
 \clearpage
 \newpage
-\section{Software Requirements \& Download}\label{sec:install-req}
+\section{Software Requirements \& Download}
+\label{sec:install-req}
 
 \subsection{Prerequisites}
-In order to install either \qcm\ or the quantum chemistry package \mol\ with DMRG support through the \qcm\ software suite \cite{Keller_JChemPhys_efficient_2015,interface}, requires the following libraries/programs:
+
+In order to install either \qcm\ or the quantum chemistry package \mol\ with DMRG support through the \qcm\ software suite \cite{Keller_JChemPhys_efficient_2015,interface}, the following libraries/programs are required:
 
@@
  \item Git
-%  \item Python version 2.\texttt{x} with \texttt{x}\ $\geq 7$, with SciPy and NumPy
  \item HDF5 (\url{http://www.hdfgroup.org/HDF5}) with Python2 bindings (\texttt{h5py})
@@
 \end{itemize}
-\vspace{0.01cm}
-\noindent
-Please make sure that these libraries/programs are available and their location is visible in your \texttt{\$PATH}\ and
-\texttt{\$LD\_LIBRARY\_PATH}, respectively.
-Note that these libraries are \emph{NOT} part of the installation package of the \qcm\ software suite (see Section \ref{sec:installation}\
-for further details).
+
+\noindent Please make sure that these libraries/programs are available and their location is visible in your \texttt{\$PATH}\ and \texttt{\$LD\_LIBRARY\_PATH}, respectively.
+Note that these libraries are \emph{NOT} part of the installation package of the \qcm\ software suite (see Section~\ref{sec:installation} for further details).
 
@@
 \begin{framed}
-\noindent \textbf{Advice:} We recommend to install \qcm{} with the \mol-\qcm{} interface. \qcm{} installed with the \mol-\qcm{} interface is also fully usable as a standalone version.
+\noindent \textbf{Advice:} We recommend to install \qcm{} with the \mol-\qcm{} interface.
+\qcm{} installed with the \mol-\qcm{} interface is also fully usable as a standalone version.
 \end{framed}
@@
 
-If you want to download and install a standalone version of \qcm{}, it may be obtained from \url{https://scine.ethz.ch/download/qcmaquis}.
-After downloading, please proceed to Section~\ref{sec:install-qcm-stand} for installation instructions.
+If you want to download and install a standalone version of \qcm{}, it may be obtained from the github repository \url{https://github.com/qcscine/qcmaquis}.
 
@@
 \newpage
-\section{Installation}\label{sec:installation}
-
-% QCMaquis in Molcas
-\subsection{Installation of the \mol-\qcm{} interface}\label{sec:install-qcm-mol}
-
-% \noindent In the following steps \ref{subsubsec:download}-\ref{subsubsec:runtime}\ we describe how to successfully build and install the \mol\ program together with the \qcm\ software suite.
-The installation of \qcm\ has been tested for different
-operating systems and compiler/math libraries environments found in Section~\ref{subsec:supportedOS}.
+
+\section{Installation}
+\label{sec:installation}
+
+\subsection{Installation of the \mol-\qcm{} interface}
+\label{sec:install-qcm-mol}
+
+The installation of \qcm\ has been tested for different operating systems and compiler/math libraries environments found in Section~\ref{subsec:supportedOS}.
 While other combinations might work equally well they are \emph{not} officially supported.
 
-\noindent The installation of the \qcm\ software suite within \mol\ will
-comprise several libraries which are \emph{automatically downloaded and installed} during the \mol\ build
-process provided that DMRG support within \mol\ is requested by the user.
+\noindent The installation of the \qcm\ software suite within \mol\ will comprise several libraries which are \emph{automatically downloaded and installed} during the \mol\ build process provided that DMRG support within \mol\ is requested by the user.
 The list of external libraries comprises:
+
 \begin{itemize}
@@
 \end{itemize}
+
 All of the above libraries will be installed locally in the user-defined build folder \texttt{my-Molcas-build} of \mol.
@@
 \subsubsection{Download and build folder setup}
+
 Download \mol{} from the official \mol{} GitLab repository:
@@
 of \texttt{my-Molcas-src} -- and change to this new folder:
+
 \begin{verbatim}
@@
 
-\subsubsection{Configuration}\label{subsubsec:configure}
+\subsubsection{Configuration}
+\label{subsubsec:configure}
 
@@
 
-\paragraph{Configuration with the GNU compiler suite}\label{sec:gnu-conf}$\;$\\
+\paragraph{Configuration with the GNU compiler suite}
+\label{sec:gnu-conf}$\;$\\
 
 \noindent To configure \mol\ and \qcm\ with the GNU compiler suite type
+
 \begin{verbatim}
   FC=gfortran CC=gcc CXX=g++ cmake -DDMRG=ON -DLINALG=MKL \
-  -DQCMaquis_NAME="Your Name" -DQCMaquis_EMAIL="your@email.com" \
     /path/to/my-Molcas-src
@@
 If the MKL libraries are not available internal math libraries of \mol\ can be requested with \texttt{-DLINALG=Internal} (default option for \texttt{LINALG} in \mol) or \texttt{-DLINALG=Runtime} (to use a system-wide BLAS or LAPACK installation).
-Please also specify your name and e-mail address in the \texttt{-DQCMaquis\_NAME} and \texttt{-DQCMaquis\_EMAIL} CMake options, otherwise the download of \qcm{} will not work.
+%Please also specify your name and e-mail address in the \texttt{-DQCMaquis\_NAME} and \texttt{-DQCMaquis\_EMAIL} CMake options, otherwise the download of \qcm{} will not work.
 
@@
 \end{framed}
-% \paragraph{Configuration with the Intel compiler suite}\label{sec:intel-conf}$\;$\\
+
 \begin{framed}
- \noindent Note: Intel compiler is currently not supported. Running a binary compiled with Intel compilers may result in segmentation faults. We are currently investigating the issue.
+  \noindent Note: Intel compiler is currently not supported. Running a binary compiled with Intel compilers may result in segmentation faults.
+  We are currently investigating the issue.
 \end{framed}
 
-\paragraph{Upgrading an existing \mol{} installation}\label{sec:build-upgrade}$\;$\\
+\paragraph{Upgrading an existing \mol{} installation}
+\label{sec:build-upgrade}$\;$\\
 
 If you're upgrading from an \mol{} installation compiled with an older release of \qcm{}, you need to add an additional command line flag when invoking \texttt{cmake}, namely
+
 \begin{verbatim}
@@
 This flag activates the \qcm{} update for the current and all the following \texttt{cmake} runs. To switch off \qcm{} updates, \texttt{cmake} may be run with
+
 \begin{verbatim}
@@
 \paragraph{Additional options}$\;$\\
+
 Following additional options can be switched on by passing additional flags to \texttt{cmake}:
+
 \begin{itemize}
  \item \texttt{-DNEVPT2=ON}: Downloads and compiles the NEVPT2 program and its interface with \mol\ and \qcm.
- \item \texttt{-DOPENMP=ON}: Enable OpenMP parallelization for MOLCAS and NEVPT2 (see Section \ref{sec:parallel}). Note that \qcm{} is built with OpenMP support \emph{regardless} of this flag.
+ \item \texttt{-DOPENMP=ON}: Enables OpenMP parallelization for MOLCAS and NEVPT2 (see Section \ref{sec:parallel}).
+       Note that \qcm{} is built with OpenMP support \emph{regardless} of this flag.
 \end{itemize}
@@
 Other options to enable/disable various options in \mol{} may be passed to \texttt{cmake} at this step. Please consult the \mol{} manual for details.
+
 % \paragraph{Additional \mol{} options}$\;$\\
@@
 Use \texttt{ccmake} to set other various options for \mol{}: for details please consult the \mol{} manual.
+
 \begin{framed}
@@
 
-\noindent Installing an MPI-parallel version of \mol\ with DMRG support is possible although
-\qcm\ itself is by default shared-memory OMP but \emph{not} yet MPI-parallelized.
+\noindent Installing an MPI-parallel version of \mol\ with DMRG support is possible although \qcm\ itself is by default shared-memory OMP but \emph{not} yet MPI-parallelized.
 To configure \mol\ for an MPI-installation type
+
 \begin{verbatim}
@@
 
-A shared-memory OMP parallelized version of \mol\ can be activated with the option
-\texttt{-DOPENMP=ON}\ passed to \texttt{cmake} during the configuration step, for example
+A shared-memory OMP parallelized version of \mol\ can be activated with the option \texttt{-DOPENMP=ON}\ passed to \texttt{cmake} during the configuration step, for example
+
 \begin{verbatim}
@@
 
-In order to exploit the shared-memory OMP parallelization of \qcm\ which is \emph{enabled by default}
-the user is strongly encouraged to set at runtime the environment
-variable \texttt{export QCMaquis\_CPUS=XX} or \texttt{export OMP\_NUM\_THREADS=XX}
-where \texttt{XX}\ specifies the number of shared-memory cores to be used. The default is to use a single core.
+In order to exploit the shared-memory OMP parallelization of \qcm\ which is \emph{enabled by default} the user is strongly encouraged to set at runtime the environment variable \texttt{export QCMaquis\_CPUS=XX} or \texttt{export OMP\_NUM\_THREADS=XX}
+where \texttt{XX}\ specifies the number of shared-memory cores to be used.
+The default is to use a single core.
 
@@
 \end{verbatim}
+
 \noindent or
+
 \begin{verbatim}
@@
 \end{verbatim}
+%
 to compile \mol\ (in parallel on 8 cores) and install all its components in the build folder \texttt{\molbuild}.
-In the same folder \qcm\ as well as the required \textsc{Boost} and \textsc{Alps}
-libraries will be downloaded and installed, respectively.
+In the same folder \qcm\ will be downloaded and installed.
 The installation process thus requires a working internet connection.
@@
 After having successfully passed the \qcm\ installation step as indicated by the CMake message
+
 \begin{verbatim}
@@
 \end{verbatim}
+
 set up your \mol{} environment by setting \texttt{MOLCAS} environment variable
+
 \begin{verbatim}
@@
 \end{verbatim}
+
 and then set up the \qcm{} runtime environment by running the following shell script
+
 \begin{verbatim}
@@
 \end{verbatim}
+
 You might want to include the above lines in your local \texttt{\$HOME/.bashrc} to set up your environment automatically every time your session is started.
@@
 
-\qcm\ in \hostp\ is tested and officially supported for Linux (x86\_64)
-and MacOS X (10.12.x and 10.13.x ) operating systems.
-We support GCC compilers starting from 5.x up to 7.x along with the MKL (Linux, x86\_64) and MKL/Accelerate (Mac OS X) math libraries, respectively, to successfully build the \qcm\ software suite within \hostp.
+\qcm\ in \hostp\ is tested and officially supported for Linux (x86\_64) and MacOS X (10.12.x and 10.13.x ) operating systems.
+We support GCC compilers starting from 5.x up to 9.x along with the MKL (Linux, x86\_64) and MKL/Accelerate (Mac OS X) math libraries, respectively, to successfully build the \qcm\ software suite within \hostp.
 Note that other combinations might work equally well but they are not officially supported at present.
 
-\noindent See the \mol\ manual at \url{www.molcas.org}\ for details on
-supported operating systems and compiler environments of a ``plain" \mol\ installation.
+\noindent See the \mol\ manual at \url{www.molcas.org}\ for details on supported operating systems and compiler environments of a ``plain" \mol\ installation.
 
@@
 To verify that your installation of \mol\ and \qcm\ was successful, run the following command
+
 \begin{verbatim}
@@
 \end{verbatim}
+
 \noindent The message
+
 \begin{verbatim}
@@
 \end{verbatim}
-\noindent indicates that all tests passed correctly and your installation is good to go for production work.
+
+\noindent indicates that all tests passed correctly and your installation is ready for production work.
 \vspace{2ex}
@@
 \noindent Test inputs are distributed with \mol{} and can be found in the \texttt{qcmaquis} subdirectory of the \mol{} test directory.
-They may also serve as sample inputs for your actual calculation.  Table \ref{tab:inputex}\ comprises a short summary of
-the tests for \qcm\ covering different computational aspects within the \mol\ framework.
-
-\begin{longtable}{lll}
-\caption{\label{tab:inputex}List of \qcm{} test inputs in \mol{}.}
-\\
-  \toprule
+They may also serve as sample inputs for your actual calculation. 
+Table \ref{tab:inputex}\ comprises a short summary of the tests for \qcm\ covering different computational aspects within the \mol\ framework.
+
+\begin{table}
+  \begin{tabular}{lll}
+    \hline \hline
     file & type & features\\
-  \midrule
-\endfirsthead
-
-\multicolumn{3}{c}%
-{{\bfseries \tablename\ \thetable{} -- continued from previous page}} \\
-\toprule
-file & type & features\\
-\midrule
-\endhead
-\midrule
+    \hline
     001.input & State-specific & Ground-state DMRG-SCF optimization of N$_2$\\
@@
     010.input & MC-PDFT        & DMRG-MC-PDFT energy calculation of the ground state of N$_2^+$\\
-\bottomrule
-\end{longtable}
+    \hline \hline
+  \end{tabular}
+  \caption{List of \qcm{} test inputs in \mol{}.}
+  \label{tab:inputex}
+\end{table}
 
@@
 \begin{itemize}
-  \item For issues regarding usage of \qcm\ in \mol{}, open an issue in the \mol{} bug tracker at
-    \url{https://gitlab.com/Molcas/OpenMolcas/issues}
-  \item For issues regarding the usage of \qcm{} standalone version, please send an e-mail to \href{mailto:dmrg@phys.chem.ethz.ch}{dmrg@phys.chem.ethz.ch}.
+  \item For issues regarding usage of \qcm\ in \mol{}, open an issue in the \mol{} bug tracker at \url{https://gitlab.com/Molcas/OpenMolcas/issues}
+  \item For issues regarding the usage of \qcm{} standalone version, please open an issue on the \href{https://github.com/qcscine/qcmaquis}{\qcm\ GitHub repository}.
 \end{itemize}
@@
 Any issue will be dealt with by one of the authors, although no guarantee on the promptness of response is given.
-In general, serious bugs that have been reported and fixed will lead to a new patch of the \qcm\ program,
-announced and distributed via the \href{https://www.scine.ethz.ch/download/qcmaquis}{SCINE \qcm\ homepage}.
+In general, serious bugs that have been reported and fixed will lead to a new patch of the \qcm\ program, announced and distributed via the \href{https://github.com/qcscine/qcmaquis}{\qcm\ GihHub repository}.
 
@@
 
-\section{Running a \qcm\ DMRG Calculation}\label{sec:general-consid}
-
-Running a \qcm\ DMRG calculation --- either DMRG-CI or DMRG-SCF --- is most easily achieved with the \qcm-\mol{} interface \cite{interface}, which integrates the DMRG calculation into the \mol{} workflow.
+\section{Running a \qcm\ DMRG Calculation}
+\label{sec:general-consid}
+
+Running a \qcm\ DMRG calculation --- either DMRG-CI or DMRG-SCF --- is most easily achieved with the \qcm-\mol{} interface\cite{interface}, which integrates the DMRG calculation into the \mol{} workflow.
 Usage of the \qcm-\mol{} interface described in Section~\ref{sec:input}.
@@
 
-\subsection{Keywords and Options for \hostp}\label{sec:input}
-
-\subsubsection{The DMRGSCF module}\label{sec:dmrgscf}
+\subsection{Keywords and Options for \hostp}
+\label{sec:input}
+
+\subsubsection{The DMRGSCF module}
+\label{sec:dmrgscf}
 
@@
     \item Freitag,~L.; Keller,~S.; Knecht,~S.; Lindh,~R.; Ma,~Y.; Stein,~C.~J.;
-  Reiher,~M. \textit{in
-		preparation} \textbf{Check for a preprint on \href{http://arxiv.org/}{arXiv.org} (to appear soon).}
-\item{Keller,~S.;~Reiher,~M., \emph{J. Chem. Phys.} \textbf{2016}, \emph{144}, 134101\\ \href{https://doi.org/10.1063/1.4944921}{doi:10.1063/1.4944921}}
+    Reiher,~M. \textit{in preparation} \textbf{Check for a preprint on \href{http://arxiv.org/}{arXiv.org} (to appear soon).}
+    \item{Keller,~S.;~Reiher,~M., \emph{J. Chem. Phys.} \textbf{2016}, \emph{144}, 134101\\   
+      \href{https://doi.org/10.1063/1.4944921}{doi:10.1063/1.4944921}}
     \item Keller,~S.;~Dolfi,~M.;~Troyer,~M.;~Reiher,~M., \emph{J. Chem. Phys.}
@@
 	\item{Knecht~S.;~Hedeg{\aa}rd E. D.;~Keller S.;~Kovyrshin A.;~Ma  Y.;~Muolo A.;~Stein C. J.;~Reiher M.,
-	\emph{Chimia} \textbf{2016}, \emph{70} ~244--251, \href{https://doi.org/10.2533/chimia.2016.244}{doi:10.2533/chimia.2016.244}.
-}
+	  \emph{Chimia} \textbf{2016}, \emph{70} ~244--251,
+	  \href{https://doi.org/10.2533/chimia.2016.244}{doi:10.2533/chimia.2016.244}.}
   \end{enumerate}
@@
 
-The \qcm-\mol{} interface introduces a new \hostp\ module, \texttt{DMRGSCF} which may be used to perform DMRG-CI or DMRG-SCF wave function optimizations.
+The \qcm-\mol{} interface introduces a new \hostp\ module, \texttt{DMRGSCF}, which may be used to perform DMRG-CI or DMRG-SCF wave function optimizations.
 A \texttt{DMRGSCF} input (see example in Listing~\ref{lst:dmrgscf}) consists of the mandatory keyword \kwd{ActiveSpaceOptimizer} and the two input blocks \kwd{DMRGSettings\ldots{}EndDMRGSettings} and \kwd{OOptimizationSettings\ldots{}EndOOptimizationSettings}.
@@
 \paragraph{The \kwd{ActiveSpaceOptimizer} keyword}\mbox{}\\
-The \kwd{ActiveSpaceOptimizer} keyword specifies the CI solver program to be used in the \kwd{DMRGSCF} module. Currently, only one value, namely\\
+
+The \kwd{ActiveSpaceOptimizer} keyword specifies the CI solver program to be used in the \kwd{DMRGSCF} module.
+Currently, only one value, namely\\
 \noindent\kwd{ActiveSpaceOptimizer=QCMaquis}\\
-is supported, which, obviously, calls the \qcm{} program. In the future, this option will be used to enable interfaces to other DMRG programs such as \textsc{Block} or \textsc{CheMPS2}.
+is supported, which, obviously, calls the \qcm{} program.
+In the future, this option will be used to enable other DMRG programs such as \textsc{Block} or \textsc{CheMPS2}.
 
 \paragraph{The \kwd{Fiedler} keyword}\mbox{}\\
-The \kwd{Fiedler} keyword, i.e.,\\
-\noindent\kwd{Fiedler=on}\\
-enables a state-specific orbital ordering for the MPS optimization
-by exploiting concepts from graph theory. The ordering follows from the elements of the
-Fiedler vector \cite{fied73,fied75} which is the eigenvector corresponding to the second
-lowest eigenvalue of the so-called graph Laplacian.
+
+The \kwd{Fiedler} keyword, i.e., \noindent\kwd{Fiedler=on}, enables a state-specific orbital ordering for the MPS optimization by exploiting concepts from graph theory. 
+The ordering follows from the elements of the Fiedler vector \cite{fied73,fied75} which is the eigenvector corresponding to the second lowest eigenvalue of the so-called graph Laplacian.
 Further details concerning our implementation of the Fiedler-vector ordering within \qcm\ can be found in Ref.\cite{interface}.
@@
 \paragraph{The \kwd{CIDEAS} keyword}\mbox{}\\
-The \kwd{CIDEAS} keyword, i.e.,\\
-\noindent\kwd{CIDEAS=on}\\
-enables a more advanced algorithm to construct a suitable initial MPS (see the keyword \kwd{init\_state} in Table \ref{tab:standard_kw_auto} for other options) provided by the configuration interaction dynamically extended active space (CI-DEAS) approach \cite{Legeza2003,fiedler}.
-The CI-DEAS protocol can be interpreted as an orbital entanglement entropy guided configuration selection and the quality of this initial guess depends on the quality of the initial CAS vector.
+
+The \kwd{CIDEAS} keyword, i.e., \noindent\kwd{CIDEAS=on}, enables a more advanced algorithm to construct a suitable guess MPS (see the keyword \kwd{init\_state} in Table \ref{tab:standard_kw_auto} for other options) provided by the configuration interaction dynamically extended active space (CI-DEAS) approach \cite{Legeza2003,fiedler}.
+The CI-DEAS protocol can be interpreted as an orbital entanglement entropy guided configuration selection whose quality depends on that of the initial CAS vector.
 Further details concerning our implementation of the CI-DEAS approach within \qcm\ can be found in Ref.\cite{interface}.
+
 \begin{framed}
-\noindent \textbf{Warning!} The CI-DEAS functionality is currently restricted to calculations performed with C$_1$\ symmetry. Support for other point group symmetries will be available in due time.
-
-\noindent \textbf{Note}: The \kwd{CIDEAS} option requires to set the ``HF occupation" for each state in the \texttt{OOptimizationSettings}\ input section by means of the \kwd{SOCC}\ keyword (see Table \ref{table:opt-keyword}).
+  \noindent \textbf{Warning!} The CI-DEAS functionality is currently restricted to calculations performed with C$_1$ symmetry (\textit{i.e.}, without point group symmetry).
+  Support for other point group symmetries will be available in due time.
+  \noindent \textbf{Note}: The \kwd{CIDEAS} option requires to set the ``HF occupatio'' for each state in the \texttt{OOptimizationSettings}\ input section by means of the \kwd{SOCC}\ keyword (see Table \ref{table:opt-keyword}).
 \end{framed}
@@
 \paragraph{The \kwd{DMRGSettings} block}\mbox{}\\
+
 The \kwd{DMRGSettings} block contains DMRG-specific options, which will be passed on to the \qcm{} program.
 The mandatory keywords, which must be present in each calculation, are summarized in Table~\ref{tab:standard_kw}. 
-In addition to these keywords, any \qcm{} keyword (see Table~\ref{tab:advanced_kw} may be used in this section.
-% In addition, keyword \kwd{Compress} may be used to enable wavefunction compression, e.\,g.\ for a subsequent NEVPT2 calculation.
+In addition to these keywords, any \qcm{} keyword (see Table~\ref{tab:advanced_kw}) may be used in this section.
 
 \begin{table}[h]
- \caption{\label{tab:standard_kw}Mandatory and special \qcm{} keywords to be specified in the \kwd{DMRGSettings} block of the \kwd{DMRGSCF} input.}
+  \caption{Mandatory and special \qcm{} keywords to be specified in the \kwd{DMRGSettings} block of the \kwd{DMRGSCF} input.}
   \begin{tabular}{ll}
@@
      \midrule
-
- \multirow{3}{*}{\texttt{max\_bond\_dimension}} & \multirow{3}{11cm}{Maximum number of renormalized states
-(commonly referred to as $m$-value or maximum bond dimension)
-kept during each microiteration step of a sweep.}\\
-& \\
-& \\
-\multirow{4}{*}{\texttt{nsweeps}} & \multirow{4}{11cm}{Maximum number of DMRG sweeps. Please be aware that \texttt{nsweeps} sets
-the number of combined forward and backward sweeps. Thus, the actual number of sweeps is $2\ \times $ \texttt{nsweeps}.} \\
-& \\
-& \\
-& \\
-%   \cmidrule(rl){1-2}
-%  \multirow{6}{*}{\texttt{Compress} (optional)} & \multirow{5}{12cm}{Compress the MPS obtained from the wavefunction optimization to an MPS with a smaller maximum number of renormalized states (provided as a parameter to the option) and use the compressed MPS in a subsequent 4-RDM evaluation.
-As the 4-RDM evaluation scales as $m^{3}L^8$, this is a convenient way to save computational cost in DMRG-NEVPT2 calculations.}\\
-%   & \\
-%   & \\
-%   & \\
-%   & \\
-%   & \\
-%   & \textbf{default:} \texttt{Compress=0}, which disables the compression. \\
+     \multirow{3}{*}{\texttt{max\_bond\_dimension}} & 
+     \multirow{3}{11cm}{Maximum number of renormalized states (commonly referred to as $m$-value or maximum bond dimension) kept during each microiteration step of a sweep.} \\
+       & \\
+       & \\
+     \multirow{4}{*}{\texttt{nsweeps}} & 
+     \multirow{4}{11cm}{Maximum number of DMRG sweeps.
+Please be aware that \texttt{nsweeps} sets the number of combined forward and backward sweeps.
+Thus, the actual number of sweeps is $2\ \times $ \texttt{nsweeps}.} \\
+       & \\
+       & \\
+       & \\
      \bottomrule
   \end{tabular}
+  \label{tab:standard_kw}
 \end{table}
@@
 \begin{table}[h]
+  \caption{Inoperative \kwd{RASSCF} keywords in the \kwd{OOptimizationSettings} block.}
   \begin{center}
-\caption{Inoperative \kwd{RASSCF} keywords in the \kwd{OOptimizationSettings} block.}\label{table:disabled-kw}
     \begin{tabular}{l@{\hspace{1.25cm}}l}
@@
       \midrule
-\multirow{2}{*}{\texttt{RASScf}} & \multirow{3}{12cm}{Keywords are unavailable as excitation level restrictions between subspaces are unsupported in DMRG calculations.}\\
+      \multirow{2}{*}{\texttt{RASScf}} & 
+      \multirow{3}{12cm}{Keywords are unavailable as excitation level restrictions between subspaces are unsupported in DMRG calculations.}  \\
       \multirow{2}{*}{\texttt{GASScf}}  & \\
       & \\
-\multirow{4}{*}{\texttt{TIGHt}} & \multirow{4}{12cm}{This keyword is not available because the Jacobi-Davidson diagonalization is
-independent and can be controlled with the \texttt{ietl\_jcd\_tol} and \texttt{ietl\_jcd\_maxiter} parameters (see
-Section \ref{sec:qcmaquis-kw}) in the \kwd{DMRGSettings} section.}\\
+      \multirow{4}{*}{\texttt{TIGHt}} 
+      & \multirow{4}{12cm}{This keyword is not available because the accuracy of the Jacobi-Davidson diagonalization can be controlled with the \texttt{ietl\_jcd\_tol} and \texttt{ietl\_jcd\_maxiter} parameters (see Section~\ref{sec:qcmaquis-kw}) in the \kwd{DMRGSettings} section.}\\
       & \\
@@
   \end{center}
+  \label{table:disabled-kw}
 \end{table}
@@
 \paragraph{The \kwd{OOptimizationSettings} block}\mbox{}\\
+
 The \kwd{OOptimizationSettings} block contains general, non DMRG-specific options required for the wavefunction optimisation (such as number of the active electrons, active orbital specification etc.) and resemble the input for an analogous CASCI or CASSCF calculation with the \mol{} \kwd{RASSCF} module.
-Most of the \kwd{RASSCF} keywords are accepted, with the exception of the keywords in Table~\ref{table:disabled-kw}.
+Most of the \kwd{RASSCF} keywords are accepted, with the exception of these listed in Table~\ref{table:disabled-kw}.
 Please consult the \kwd{RASSCF} entry of the \mol{} manual for a detailed description of the input.
@@
 
-% \begin{longtable}{l@{\hspace{1.25cm}}l}
 \begin{table}[h]
@@
   \begin{tabular}{ll}
-\\
     \toprule
@@
     \midrule
-% \endfirsthead
-%
-% \multicolumn{2}{c}%
-% {{\bfseries \tablename\ \thetable{} -- continued from previous page}} \\
-% \toprule
-% keyword & description \\
-% \midrule
-% \endhead
-\multirow{4}{*}{\texttt{FCIDUMP}} & \multirow{4}{12cm}{Skip the optimization and write out the transformed active MO integrals to a
-\kwd{FCIDUMP} file in \kwd{\$WorkDir} (see Ref. \citenum{fcidump} for a detailed description of the format) which can be used in subsequent \qcm\ DMRG calculations.}\\
- & \\
- & \\
- & \\
-\multirow{6}{*}{\texttt{SOCCupy}} & \multirow{6}{12cm}{Initial electronic configuration for the calculated state(s).
-This keyword is equivalent to the \texttt{hf\_occ} card in the \qcm\ input (see Section
-\ref{sec:qcmaquis-kw}), but allows input for multiple states.
+    \multirow{4}{*}{\texttt{FCIDUMP}} & 
+    \multirow{4}{12cm}{Skip the optimization and write out the transformed active MO integrals to a
+					   \kwd{FCIDUMP} file in \kwd{\$WorkDir} (see Ref. \citenum{fcidump} for a detailed description of the format) which can be used in subsequent \qcm\ DMRG calculations.} \\
+      & \\
+      & \\
+      & \\
+    \multirow{6}{*}{\texttt{SOCCupy}} & 
+    \multirow{6}{12cm}{Initial electronic configuration for the calculated state(s). This keyword is equivalent to the
+ 					   \texttt{hf\_occ} card in the \qcm\ input (see Section~\ref{sec:qcmaquis-kw}), but allows input for multiple states.
 The occupation is inserted as a string (strings) of aliases of occupations of the active (RAS2) orbitals with the aliases $2 \,=\,$ full, $u \,=\,$ up, $d \,=\,$ down, $0 \,=\,$ empty.
-For several states, the occupation strings for each state are separated by newlines.}\\
+For several states, the occupation strings for each state are separated by newlines.} \\
       & \\
@@
     \bottomrule
-% \end{longtable}
   \end{tabular}
@@
   \noindent \textbf{Note}: The input format for DMRG/DMRGSCF calculations with the \kwd{RASSCF} module of \mol{} described in this section is \textbf{deprecated} and will be \textbf{unavailable} in future versions of \mol{}!
-
   \noindent Instead, please use the \kwd{DMRGSCF} module as described in Section~\ref{sec:dmrgscf}.
@@
 
-Instead of using the new \kwd{DMRGSCF} module, DMRG calculations with \qcm{} are also supported from within the \kwd{RASSCF} module directly.
+Instead of using the new \kwd{DMRGSCF} module, DMRG calculations with \qcm{} are also supported within the \kwd{RASSCF} module directly.
 Compared to the \kwd{DMRGSCF} input, the \kwd{RASSCF} input changes as follows:
-\begin{samepage}
+
 \begin{itemize}
@@
 \end{itemize}
+
 An input example is provided in Listing~\ref{lst:rasscf}.
-\end{samepage}
 
@@
 
-\FloatBarrier
-\newpage
 \subsubsection{Input Examples}
-\begin{lstlisting}[language=molcas,caption={(File \texttt{001.input} in the test directory) Input example for a DMRG-SCF calculation with the \kwd{DMRGSCF} module (\ch{N2}, 6 orbitals, with symmetry).},label=lst:dmrgscf]
+
+\begin{lstlisting}[language=molcas,caption={(File \texttt{001.input} in the test directory) 
+				   Input example for a DMRG-SCF calculation with the \kwd{DMRGSCF} module (N$_2$, 6 orbitals, with symmetry).},label=lst:dmrgscf]
 &GATEWAY
@@
 \vspace{-0.5em}
-\begin{lstlisting}[language=molcas,caption={Input example for a DMRG-SCF calculation with the \kwd{RASSCF} module (\ch{N2}, 6 orbitals, with symmetry)},label=lst:rasscf]
+\begin{lstlisting}[language=molcas,caption={Input example for a DMRG-SCF calculation with the \kwd{RASSCF} module (N$_2$, 6 orbitals, with symmetry)},label=lst:rasscf]
 &GATEWAY
@@
 
-\section{State interaction with DMRG (MPS-SI)}\label{sec:mps-si}
+\section{State interaction with DMRG (MPS-SI)}
+\label{sec:mps-si}
 
@@
 
-The state-interaction module for MPS wave functions (MPS-SI) introduced in
-Ref.~\cite{knec16b} is based on the \texttt{CASSI/RASSI} framework \cite{malm89,malm02} of \hostp.
-MPS-SI is a generalized state-interaction approach for
-both \textit{nonorthogonal} and \textit{orthonormal} spin-free MPS wave
-functions which enables the evaluation of arbitrary one- and
-two-particle transition matrix elements as well as, for example, matrix
-elements of the spin-orbit coupling operator.
-For instance, diagonalization of the spin-orbit Hamiltonian matrix
-yields spin-orbit coupled wave functions as linear combinations of the
-uncoupled, spin-pure MPS states. The latter can (but do not have to)
-be obtained as results from one or several DMRG-SCF orbital
-optimization calculations as described in Section \ref{sec:dmrgscf}.
-
-Following the work of Malmqvist \cite{malm86}, the central element of the MPS-SI
-algorithm is the transformation of the bra and ket MPS wave functions
-to a biorthonormal basis representation.
-It is important to note that the latter transformation is not needed
-if the MPS wave functions considered for state interaction
-share a common MO basis. In this particular case, the MPS-SI program directly
-proceeds with the calculation of the reduced (transition) one- and two-particle
-density matrices.
-We emphasize that our approach is applicable to the general case with MPS
-wave functions built from mutually nonorthogonal molecular orbital bases.
-It therefore provides the desired flexibility
-to find the best individual molecular orbital basis to represent
-wave functions of different spin and/or spatial symmetry.
-After solving a generalized eigenvalue
-equation of the form
-\begin{equation}\label{eq:eigproblem}
+The state-interaction module for MPS wave functions (MPS-SI) introduced in Ref.~\cite{knec16b} is based on the \texttt{CASSI/RASSI} framework \cite{malm89,malm02} of \hostp.
+MPS-SI is a generalized state-interaction approach for both \textit{nonorthogonal} and \textit{orthonormal} spin-free MPS wave
+functions which enables the evaluation of arbitrary one- and two-particle transition matrix elements as well as, for example, matrix elements of the spin-orbit coupling operator.
+For instance, diagonalization of the spin-orbit Hamiltonian matrix yields spin-orbit coupled wave functions as linear combinations of the uncoupled, spin-pure MPS states.
+The latter can (but do not have to) be obtained as results from one or several DMRG-SCF orbital optimization calculations as described in Section~\ref{sec:dmrgscf}.
+
+Following the work of Malmqvist \cite{malm86}, the central element of the MPS-SI algorithm is the transformation of the bra and ket MPS wave functions to a biorthonormal basis representation.
+It is important to note that the latter transformation is not needed if the MPS wave functions considered for state interaction
+share a common MO basis. In this particular case, the MPS-SI program directly proceeds with the calculation of the reduced (transition) one- and two-particle density matrices.
+We emphasize that our approach is applicable to the general case with MPS wave functions built from mutually nonorthogonal molecular orbital bases.
+It therefore provides the desired flexibility to find the best individual molecular orbital basis to represent wave functions of different spin and/or spatial symmetry.
+After solving a generalized eigenvalue equation of the form
+
+\begin{equation}
   {\textbf{Hc}} = E {\textbf{Sc}}\ ,
+  \label{eq:eigproblem}
 \end{equation}
-with the Hamiltonian matrix \textbf{H} expressed in the basis of the DMRG-SCF MPS
-wave functions and the wave function overlap matrix \textbf{S},
-a set of fully orthogonal and non-interacting states are obtained
-as linear combinations of the DMRG-SCF MPS wave functions with
-the expansion coefficients given by \textbf{c}.
-
-Moreover, it is possible to ``dress" the diagonal elements of the Hamiltonian in Eq.~\eqref{eq:eigproblem} for \kwd{MPS-SI} by adding
-a correlation-correction term obtained, for example, from a preceding NEVPT2
-calculation (see Section \ref{sec:nevpt2}), by either using the \texttt{HDIAG} keyword within the \kwd{RASSI}\ module or provide the \texttt{nevpt2.h5}\ wave function file as input option to the \kwd{\&MPSSI} module (see the description of the keyword \kwd{FILE} in the \hostp\ manual for the \kwd{MPSSI} module for details on the correct syntax).
-
-\subsection{Running MPS-SI}\label{sec:run-mpssi}
-
-An MPS-SI calculation requires reference DMRG-SCF or NEVPT2
-wave functions as obtained with the \kwd{DMRGSCF} (see Section \ref{sec:dmrgscf}) or \kwd{NEVPT2} modules (see Section \ref{sec:nevpt2}), respectively.
+%
+with the Hamiltonian matrix \textbf{H} expressed in the basis of the DMRG-SCF MPS wave functions and the wave function overlap matrix \textbf{S}, a set of fully orthogonal and non-interacting states are obtained as linear combinations of the DMRG-SCF MPS wave functions with the expansion coefficients given by \textbf{c}.
+
+Moreover, it is possible to ``dress" the diagonal elements of the Hamiltonian in Eq.~\eqref{eq:eigproblem} for \kwd{MPS-SI} by adding a correlation-correction term obtained, for example, from a preceding NEVPT2 calculation (see Section~\ref{sec:nevpt2}), by either using the \texttt{HDIAG} keyword within the \kwd{RASSI}\ module or providing the \texttt{nevpt2.h5}\ wave function file as input option to the \kwd{\&MPSSI} module (see the description of the keyword \kwd{FILE} in the \hostp\ manual for the \kwd{MPSSI} module for details on the correct syntax).
+
+\subsection{Running MPS-SI}
+\label{sec:run-mpssi}
+
+An MPS-SI calculation requires reference DMRG-SCF or NEVPT2 wave functions as obtained with the \kwd{DMRGSCF} (see Section~\ref{sec:dmrgscf}) or \kwd{NEVPT2} modules (see Section~\ref{sec:nevpt2}), respectively.
+
 \begin{framed}
- \noindent Note: Currently, the MPS-SI implementation supports all point groups implemented in \hostp\
- if there is only one set of input states (i.e., all states have the same spatial \underline{and} spin symmetry).
-By contrast, if there is more than one set of input states of different spin symmetry, the \kwd{MPS-SI} module
- supports only calculations in C$_1$\ symmetry. The latter restriction will be removed in a forthcoming release.
+ \noindent Note: Currently, the MPS-SI implementation supports all point groups implemented in \hostp\ if there is only one set of input states (i.e., all states have the same spatial \underline{and} spin symmetry).
+ By contrast, if there is more than one set of input states of different spin symmetry, the \kwd{MPS-SI} module supports only calculations in C$_1$\ symmetry.
+The latter restriction will be removed in a forthcoming release.
 \end{framed}
-Listing \ref{lst:mpssi} shows an input example for an MPS-SI calculation (a calculation of spin-orbit coupling matrix elements for the \ch{CH2} molecule) where
-the important keywords are highlighted in blue. A detailed explanation of the input is given below.
-
-\begin{lstlisting}[language=molcas,caption={(File \texttt{007.input} in the QCMaquis test directory) Input example for a one-shot MPS-SI spin-orbit calculation (Singlet and triplet state, \ch{CH2} molecule, 6 electrons in 6 orbitals, C1 symmetry).},label=lst:mpssi]
+
+Listing~\ref{lst:mpssi} shows an input example for an MPS-SI calculation (a calculation of spin-orbit coupling matrix elements for the CH$_2$ molecule) where the important keywords are highlighted in blue.
+A detailed explanation of the input is given here below.
+
+\begin{lstlisting}[language=molcas,caption={(File \texttt{007.input} in the QCMaquis test directory) Input example for a one-shot MPS-SI spin-orbit calculation (Singlet and triplet state, CH$_2$ molecule, 6 electrons in 6 orbitals, C1 symmetry).},label=lst:mpssi]
 &GATEWAY
@@
 
-\subsection{Input options for MPS-SI}\label{sec:input-mpssi}
+\subsection{Input options for MPS-SI}
+\label{sec:input-mpssi}
 
@@
 \end{table}
-\FloatBarrier
-
-\clearpage
-\newpage
 
@@
 
-\begin{lstlisting}[language=molcas,caption={(\texttt{009.input} in the QCMaquis test subdirectory) Input example for a one-shot NEVPT2 calculation (Singlet \ch{CH2}, 6 electrons in 6 orbitals, RICD)},label=lst:nevpt2]
+\begin{lstlisting}[language=molcas,caption={(\texttt{009.input} in the QCMaquis test subdirectory) Input example for a one-shot NEVPT2 calculation (Singlet CH$_2$, 6 electrons in 6 orbitals, RICD)},label=lst:nevpt2]
 &GATEWAY
@@
 \end{longtable}
+
 \FloatBarrier
+
 \subsection{Massively parallel distributed 4-RDM calculations}\label{sec:4rdm-distributed}
+
 The computational cost of the RDM evaluation grows as $N^8$ with the number of active orbitals, and becomes prohibitively expensive even for moderate active spaces.
@@
 Note that the DMRG-SCF/NEVPT2 calculation need not be performed on the same system as the 4-RDM evaluation.
+
 \subsubsection*{How to run NEVPT2 calculations with distributed 4-RDM evaluation}
+
 \subsubsection{Prerequisites}
+
 \begin{itemize}
@@
 \end{itemize}
+
 If your system administrator has not set up DRMAA and Python DRMAA, you might need to download and install these libraries yourself.
 After the installation, the environment variable \kwd{DRMAA\_LIBRARY\_PATH} must be set to the path to \kwd{libdrmaa.so} and, if Python does not find the DRMAA Python binding, also \kwd{PYTHONPATH} to the path of the Python DRMAA library.
+
 \subsubsection{Workflow}\label{sec:4rdm-workflow}
+
 \begin{enumerate}
@@
 \subsubsection{Troubleshooting}
+
 The \kwd{jobmanager.py} script is experimental, and also batch jobs in queuing systems are prone to crash, therefore we provide a mechanism to identify and restart the crashed batch jobs.
@@
 In this case several options are available:
+
 \begin{itemize}
@@
  should not yield an empty result -- otherwise the corresponding calculation should be rerun.
+
 \end{itemize}
@@
 \subsubsection{Transition 3-RDM distributed calculations}
+
 \kwd{jobmanager.py} also supports distributed calculations of t-3RDMs (required for QD-NEVPT2).
@@
 
-\clearpage
-\newpage
-
 \section{\qcm{} Standalone Version}
 
-Section \ref{sec:install-qcm-stand}\ describes in details the installation process for a standalone version of \qcm.
+As already anticipated, \qcm\ can be used also as a standalone code.
+The standalone version of \qcm\ is used especially in two cases: (i) for electronic-structure calculations relying on integrals that are calculated with programs other than \mol{} and (ii) for DMRG calculations beyond time-independent electronic-structure simulations.
+In the present section, we focus on the former application field, while the latter will be discussed in the following Sections. \\
+Section~\ref{sec:install-qcm-stand}\ describes in details the installation process for a standalone version of \qcm.
+
 \begin{framed}
@@
 
-% To install \qcm\ as external module of the quantum chemistry software \mol\ proceed to Section \ref{sec:install-qcm-mol}.
-
-% QCMaquis standalone
-
-\subsection{Installation}\label{sec:install-qcm-stand}
-
-\noindent In the following steps \ref{subsubsec:download-stand}-\ref{subsubsec:runtime-stand}\ we describe how to successfully build and install  the \qcm\ software suite in a standalone manner.
-The installation of \qcm\ has been tested for different operating systems and compiler/math libraries environments. Their list can be found in Section \ref{subsec:supportedOS}.
+\subsection{Installation}
+\label{sec:install-qcm-stand}
+
+\noindent We describe in the following how to build and install the standalone version of the \qcm\ software.
+The installation of \qcm\ has been tested for different operating systems and compiler/math libraries environments.
+Their list can be found in Section \ref{subsec:supportedOS}.
 While other combinations might work equally well they are \emph{not} officially supported.
@@
 The following libraries are required and need to be installed manually by the user (but are usually provided by the system package manager)
+
 \begin{itemize}
@@
 \end{itemize}
-Additionally, \qcm{} distribution contains a customised version of the ALPS 2.3.0 library, which will be downloaded and installed automatically.
-
-\subsubsection{Download \qcm}\label{subsubsec:download-stand}
-
-As mentioned in Section~\ref{sec:reg}, \qcm{} standalone version may be downloaded from \texttt{\url{https://scine.ethz.ch/download/qcmaquis}}.
+
+Additionally, \qcm{} distribution includes a customised version of the ALPS 2.3.0 library, which will be compiled automatically.
+
+\subsubsection{Download \qcm}
+\label{subsubsec:download-stand}
+
+As mentioned in Section~\ref{sec:reg}, the \qcm{} standalone version may be found in \texttt{\url{https://github.com/qcscine/qcmaquis}}.
 
@@
 
-Extract the downloaded archive \texttt{public.tar.bz2} into a new directory \texttt{\qcmsrc}.
-\noindent Create a build folder \texttt{\qcmbuild} -- note that this folder \emph{does not} necessarily have to be a subfolder
-of \qcmsrc\ -- and change to this new folder:
+Clone the GitHub repository into a new directory \texttt{\qcmsrc}.
+Create a build folder \texttt{\qcmbuild} -- note that this folder \emph{does not} necessarily have to be a subfolder of \qcmsrc\ -- and change to this new folder:
+
 \begin{verbatim}
@@
 
-\subsubsection{Configuration}\label{subsubsec:configure-stand}
-
-\qcm{} supports operating systems, C++ compilers and math libraries as listed in Section~\ref{subsec:supportedOS}. Below we will show the configuration steps for
-the most popular compiler suites GNU (Section \ref{sec:gnu-conf-stand}) and Intel (Section \ref{sec:intel-conf-stand}), respectively.
+\subsubsection{Configuration}
+\label{subsubsec:configure-stand}
+
+\qcm{} supports operating systems, C++ compilers and math libraries as listed in Section~\ref{subsec:supportedOS}. 
+Below we will show the configuration steps for the most popular compiler suites GNU (Section \ref{sec:gnu-conf-stand}) and Intel (Section \ref{sec:intel-conf-stand}), respectively.
 How to setup and use a shared-memory OMP installation of \qcm\ is described in Section \ref{sec:parallel-stand}.
 
-\paragraph{Configuration with the GNU compiler suite}\label{sec:gnu-conf-stand}$\;$\\
+\paragraph{Configuration with the GNU compiler suite}
+\label{sec:gnu-conf-stand}$\;$\\
 
 \noindent To configure \qcm\ with the GNU compiler suite type
+
 \begin{verbatim}
@@
 \end{verbatim}
-
-\noindent where we assumed that the Intel Math Kernel Library (MKL) is available (recommended option).
-If the MKL libraries are not available \qcm\ will search for other suitable math libraries installed on the operating system.
+%
+where we assumed that the Intel Math Kernel Library (MKL) is available (recommended option).
+If the MKL libraries are not available, \qcm\ will search for other suitable math libraries installed on the operating system.
 If none are found the configuration step will stop with an appropriate message.
@@
 
-\paragraph{Shared-memory OMP parallel configuration}\label{sec:parallel-stand}$\;$\\
+\paragraph{Shared-memory OMP parallel configuration}
+\label{sec:parallel-stand}$\;$\\
 
@@
 
-In order to exploit the shared-memory OMP parallelization of \qcm\ the user is strongly encouraged to set at runtime the environment
-variable
+In order to exploit the shared-memory OMP parallelization of \qcm\ the user is strongly encouraged to set at runtime the environment variable
+
 \begin{verbatim}
@@
 \end{verbatim}
-where \texttt{XX}\ specifies the number of shared-memory cores to be used. The default (\emph{depending on the operating system!!!}) is to use all available cores.
-
-\subsubsection{Building and installation}\label{subsubsec:build-stand}
+%
+where \texttt{XX}\ specifies the number of shared-memory cores to be used.
+The default (\emph{depending on the operating system!!!}) is to use all available cores.
+
+\subsubsection{Building and installation}
+\label{subsubsec:build-stand}
 
@@
 \end{verbatim}
-\noindent or
+%
+or
+
 \begin{verbatim}
@@
 \end{verbatim}
+%
 to compile \qcm\ (in parallel on 8 cores) and install all its components in the build folder \texttt{\qcmbuild}.
@@
 
-In the following we describe (i) compulsory keywords (Section \ref{sec:compul-qcm}), (ii)
-optional keywords (Section \ref{sec:optional-qcm}) for \qcm\ DMRG calculations as well as (iii)
-keywords for property calculations (Section \ref{sec:prop-qcm}).
+In the following we describe (i) compulsory keywords (Section \ref{sec:compul-qcm}), (ii) optional keywords (Section \ref{sec:optional-qcm}) for \qcm\ DMRG calculations as well as (iii) keywords for property calculations (Section \ref{sec:prop-qcm}).
 Most of the \qcm\ keywords have default settings that guarantee convergence in the general case and are inserted automatically
-by the host program (\hostp\ in this case). A reasonable choice of default values for optional keywords is given in our
-example \qcm\ input file in Section \ref{sec:input-qcm}.
+by the host program (\hostp\ in this case). 
+A reasonable choice of default values for optional keywords is given in our
+example \qcm\ input file in Section~\ref{sec:input-qcm}.
 
@@
   \noindent\textbf{Note:}\ \textsc{Maquis} has many features beyond quantum chemistry, e.g. related to solid state physics.
-Some keywords listed in the example file in Section \ref{sec:input-qcm} (e.\,g.\ \kwd{LATTICE}, \kwd{lattice\_library} etc.)\ are therefore explained only briefly in the following
-and are not to be changed from the default values (see Listing~\ref{lst:qcm}), if a quantum chemical calculation is desired.
+  Some keywords listed in the example file in Section \ref{sec:input-qcm} (e.\,g.\ \kwd{LATTICE}, \kwd{lattice\_library} etc.)\ are therefore explained only briefly in the following and are not to be changed from the default values (see Listing~\ref{lst:qcm}), if a quantum chemical calculation is desired.
 \end{framed}
 
-\subsubsection{Compulsory keywords}\label{sec:compul-qcm}
+\subsubsection{Compulsory keywords}
+\label{sec:compul-qcm}
 
 The keywords in Tables \ref{tab:standard_kw} and \ref{tab:standard_kw_auto} have to be set for every DMRG calculation since they may crucially affect the accuracy of the final result.
-Their choice depends for example on the molecule under consideration
-(i.\,e. whether you expect strong static electron correlation and/or dynamical correlation to play a major role),
-the nature of the reference orbitals (Hartree-Fock orbitals, natural orbitals of some kind, \ldots), the size of the
-active space, and many other aspects.
+Their choice depends for example on the molecule under consideration (i.\,e.
+whether you expect strong static electron correlation and/or dynamical correlation to play a major role), the nature of the reference orbitals (Hartree-Fock orbitals, natural orbitals of some kind, \ldots), the active space size, and many other aspects.
 
@@
 \FloatBarrier
+
 \begin{framed}
   \noindent\textbf{Note:}\ We strongly encourage any new user of \qcm\ to carefully read first Ref.\ \citenum{Keller_CHIMIA_Determining_2014}\ which gives a detailed introduction to
-DMRG calculations in quantum chemistry. Further quantum chemical DMRG sample calculations starting from different computational setups are discussed for
-example in Refs. \cite{interface,ru-dmrg,fde-dmrg,srdft-dmrg}.
+  DMRG calculations in quantum chemistry.
+Further quantum chemical DMRG sample calculations starting from different computational setups are discussed for example in Refs.
+\cite{interface,ru-dmrg,fde-dmrg,srdft-dmrg}.
 \end{framed}
-\subsubsection{Optional keywords}\label{sec:optional-qcm}
+
+\subsubsection{Optional keywords}
+\label{sec:optional-qcm}
 
@@
 
-\subsubsection{Keywords for expectation value calculations}\label{sec:prop-qcm}
+\subsubsection{Keywords for expectation value calculations}
+\label{sec:prop-qcm}
 
@@
 
-\subsection{\qcm\ examples library}
+\section{QCMaquis vibrational DMRG module}
+\label{sec:vDMRG}
+
+The vibrational DMRG (vDMRG) module enables calculating ground- and excited-state energies of the Born--Oppenheimer vibrational Hamiltonian:
+
+\begin{equation}
+  \underbrace{\left[ \mathcal{T}(Q_1, \ldots, Q_L) + \mathcal{V}(Q_1, \ldots, Q_L) \right]}_{\mathcal{H}_\text{vib}(Q_1, \ldots, Q_L)} \Psi_n(Q_1, \ldots, Q_L) = E_n \Psi_n(Q_1, \ldots, Q_L)
+  \label{eq:VibHam}
+\end{equation}
+
+We designed two vDMRG formulations, one based on the so-called canonical quantization format,\cite{Hirata2014_NormalOrdered} and one based on the $n$-mode second quantization.\cite{Christiansen2004_nModeQuantization}
+These two formulations differ in the way in which the Hamiltonian $\mathcal{H}_\text{vib}$ is encoded based on the second-quantization formalism.
+In the following, we will describe how this encoding is obtained.
+
+\subsection{Canonical quantization vDMRG}
+\label{sec:canonical}
+
+The canonical quantization relies on the harmonic oscillator theory.\cite{Hirata2014_NormalOrdered}
+The vibrational wave function is expanded, in practice, as follows:
+
+\begin{equation}
+  \Psi(Q_1, \ldots, Q_L) = \sum_{n_1}^{N_\text{max}} \cdots \sum_{n_L=1}^{N_\text{max}}
+	C_{n_1 \cdots n_L} | n_1 \cdots n_L \rangle \, .
+  \label{eq:CanonicalQuantizationCI}
+\end{equation}
+
+$| n_1 \cdots n_L \rangle$ are occupation number vectors that are constructed using the harmonic oscillator eigenfunction as the one-particle basis set, \textit{i.e.}
+
+\begin{equation}
+  | n_1 \cdots n_L \rangle \equiv \prod_{i=1}^L \chi_{n_i}(Q_i)
+  \label{eq:HarmonicOscillatorBasis}
+\end{equation}
+%
+$\chi_{n_i}(Q_i)$ being the $n_i$-th eigenfunction of the monodimensional harmonic oscillator Hamiltonian associated with the $i$-th mode.
+In order to encode the Hamiltonian in a format analog to Eq.~\ref{eq:CanonicalQuantizationCI}, we express it as a power series in terms of the vibrational coordinates $Q_i$.
+By including up to fourth-order force constants, the expansion reads:
+
+\begin{equation}
+  \mathcal{V}(Q_1, \ldots, Q_L) = \sum_{i,j=1}^L k_{ij} Q_i Q_j + \sum_{i,j,k=1}^L k_{ijk} Q_i Q_j Q_k
+    + \sum_{i,j,k,l=1}^L k_{ijkl} Q_i Q_j Q_k Q_l
+  \label{eq:TaylorExpansion}
+\end{equation}
+%
+where we neglected the first-order force constants since we assume that the expansion is performed around a stationary point of the PES.
+The position operators $Q_i$ are, then, encoded in second quantization based on the well-known canonical quantization rules.
+
+\begin{equation}
+  Q_i = \frac{1}{2} \left( b_i^\dagger + b_i \right) \, .
+  \label{eq:CanonicalQuantizationPosition}
+\end{equation}
+
+Note that Eq.~\ref{eq:CanonicalQuantizationPosition} holds only if $Q_i$ are the dimensionless Cartesian-based normal coordinates.
+Moreover, the kinetic energy operator can be expressed in second quantization based on the counterpart of Eq.~(\ref{eq:CanonicalQuantizationPosition}) for the conjugate momentum operator
+
+\begin{equation}
+  P_i = \frac{\mathrm{i}}{2} \left( b_i^\dagger - b_i \right) \, .
+  \label{eq:CanonicalQuantizationMomentum}
+\end{equation}
+
+The resulting form of the vibrational Hamiltonian is then converted in the MPO format by \qcm\ based on the algorithm described in Ref.~\citenum{Baiardi2017_vDMRG}. \\
+
+\begin{lstlisting}[language=qcmaquis,
+				   caption={Input example for a canonicalq quantization vDMRG-FCI calculation},
+				   label=lst:canonicalVibInput]
+  // == Generic input section ==
+  nsweeps                         = 6
+  max_bond_dimension              = 100
+  model_library                   = coded
+  lattice_library                 = coded
+  optimization                    = twosite 
+  integral_file                   = FCIDUMP_Vibrational
+  integral_cutoff                 = 1.0E-10
+  init_state                      = 'default'
+  L                               = 12
+  // == vDMRG-specific input ==
+  symmetry                        = none
+  LATTICE                         = 'watson lattice'
+  MODEL                           = 'watson'
+  Nmax                            = 10
+\end{lstlisting}
+
+We report in Listing~\ref{lst:canonicalVibInput} an example input for a vDMRG calculation relying on the canonical-quantization based formalism.
+The first part of the input has the same structure as that of an electronic-structure DMRG calculation.
+Note that the lattice dimension $L$ is, in this case, equal to the overall number of modes of the molecule.
+Unlike for the electronic-structure case, the DMRG symmetry must be set to \texttt{none}, and the \texttt{MODEL} keyword should be set to \texttt{watson}.
+As the expansion reported in Eq.~\ref{eq:CanonicalQuantizationCI} is virtually unbounded, it must be truncated, in practice, by setting a threshold $N_\text{max}$ for the maximum excitation degree of a mode.
+Such a threshold must be specified in input, as shown in Listing~\ref{lst:canonicalVibInput}.
+
+The integral file \texttt{FCIDUMP\_Vibrational} contain the force constants that define the PES through Eq.~\ref{eq:TaylorExpansion}.
+\qcm\ supports up to sixth-order force constants, that must be provided as input in the \texttt{FCIDUMP\_Vibrational} file as illustrated in Listing~\ref{lst:CanonicalFCIDUMP}
+\vspace{.5cm}
+\begin{lstlisting}[language=qcmaquis,caption={Example of an vDMRG integral file for the potential
+											  expressed in the canonical quantization format.},
+				   label=lst:CanonicalFCIDUMP, mathescape=true, rulecolor=\color{black}]
+  $\smash{k_{ij}}$     i    j   0   0   0   0
+  $\smash{\vdots}$
+  $\smash{h_{ij}}$    -i   -j   0   0   0   0
+  $\smash{\vdots}$
+  $\smash{k_{ijk}}$     i    j   k   0   0   0
+  $\smash{\vdots}$
+  $\smash{k_{ijkh}}$     i    j   k   h   0   0
+  $\smash{\vdots}$
+  $\smash{k_{ijkhm}}$     i    j   k   h   m   0
+  $\smash{\vdots}$
+  $\smash{k_{ijkhmn}}$     i    j   k   h   m   n
+\end{lstlisting}
+\vspace{.5cm}
+The $h$ and $k$ constants define the PES as follows:
+
+\begin{equation}
+  \begin{aligned}
+    V(\tilde{Q}_1, \ldots, \tilde{Q}_L) 
+      &= \sum_{i=1}^L k_{ii} \tilde{Q}_i^2 + \sum_{i=1}^L h_{ii} \tilde{P}_i^2 
+      + \sum_{ijk=1}^L k_{ijk} \tilde{Q}_i \tilde{Q}_j \tilde{Q}_k
+      + \sum_{ijkl=1}^L k_{ijkl} \tilde{Q}_i \tilde{Q}_j \tilde{Q}_k \tilde{Q}_l \\
+      &+ \sum_{ijklm=1}^L k_{ijkl} \tilde{Q}_i \tilde{Q}_j \tilde{Q}_k \tilde{Q}_l \tilde{Q}_m
+      + \sum_{ijklmn=1}^L k_{ijkl} \tilde{Q}_i \tilde{Q}_j \tilde{Q}_k \tilde{Q}_l \tilde{Q}_m \tilde{Q}_n \, ,
+  \end{aligned}
+  \label{eq:PES_Representation}
+\end{equation}
+%
+where we introduced the scaled dimensionless normal coordinates $\tilde{Q}_i$ that are defined as:
+
+\begin{equation}
+  \tilde{Q}_i = \sqrt{2} Q_i = \left( b_i^\dagger + b_i \right)
+  \label{eq:ScaledNM}
+\end{equation}
+%
+$Q_i$ being the $i$-th dimensionless normal coordinate operator.
+The corresponding scaled dimensionless momentum operator $\tilde{P}_i$ is defined as:
+
+\begin{equation}
+  \tilde{P}_i = \frac{\sqrt{2}}{\mathrm{i}} P_i = \left( b_i^\dagger - b_i \right)
+  \label{eq:ScaledNM_Momentum}
+\end{equation}
+
+\subsection{$n$-mode vDMRG}
+\label{sec:nMode}
+
+The canonical quantization-based vDMRG theory requires the PES to be expressed as a Taylor series in terms of the Cartesian-based normal coordinates.
+Such a format is, however, ill-suited to describe
+\qcm\ can apply the vDMRG method to potentials that are encoded as generic many-body expansions \cite{Bowman1997_ManyBody}
+
+\begin{equation}
+  \mathcal{V}(\bm{Q}) = \sum_{i=1}^L V^{[i]}(Q_i) + \sum_{i<j}^L V^{[ij]}(Q_i, Q_j)
+    + \sum_{ijk}^L V^{[ijk]}(Q_i, Q_j, Q_k)
+  \label{eq:PES_ManyBody}
+\end{equation}
+%
+and supports un to three-body coupling terms.
+To encode Eq.~(\ref{eq:PES_ManyBody}) as a matrix product operator, \qcm\ relies on the so-called $n$-mode second quantization \cite{Christiansen2004_nModeQuantization} to encode the vibrational Hamiltonian.
+Let $\{ \chi_{1}^{(i)}(Q_i), \ldots, \chi_{N_i}^{(i)}(Q_i) \}$ be the $N_i$-dimensional modal basis describing the $i$-th mode.
+The full Vibrational Configuration Interaction (VCI) space is spanned by the basis constructed by forming all possible products of single particle bases, \textit{i.e.} by the functions $\Psi^{i_1,\ldots,i_L}(Q_1,\ldots,Q_L)$ defined as:
+
+\begin{equation}
+  \Psi^{i_1,\ldots,i_L}(Q_1,\ldots,Q_L) = \prod_{j=1}^L \chi_{i_j}^{(j)}(Q_j)
+  \label{eq:ProductBasis}
+\end{equation}
+
+It is possible to encode each basis function of Eq.~(\ref{eq:ProductBasis}) as an occupation number vector format, analogously to the electronic structure case, and define the corresponding creation and annihilation operators.
+Following Ref.~\citenum{Christiansen2004_nModeQuantization}, we define a pair of creation and annihilation operators per mode and per modals ($a_{i_j}^\dagger$ and $a_{i_j}$, respectively).
+The second-quantized form of Eq.~\ref{eq:PES_ManyBody} then reads, for a two-body potential (the extension to three-body coupling terms is trivial):
+
+\begin{equation}
+ \begin{aligned}
+  \mathcal{V}(\bm{Q}) &= \sum_{i=1}^L \sum_{m_i, n_i}^{N_i}
+    \langle \chi_{m_i}^{(i)}(Q_i) | V^{[i]}(Q_i) | \chi_{n_i}^{(i)}(Q_i) \rangle \\
+  &+ \sum_{i<j}^L \sum_{m_i, n_i}^{N_i} \sum_{m_j, n_j}^{N_j}
+     \langle \chi_{m_i}^{(i)}(Q_i) \chi_{m_j}^{(j)}(Q_j) | V^{[ij]}(Q_i, Q_j) | 
+             \chi_{n_i}^{(i)}(Q_i) \chi_{n_j}^{(j)}(Q_j) \rangle \,
+ \end{aligned}
+ \label{eq:PES_SecondQuantization}
+\end{equation}
+%
+\qcm\ constructs Eq.~(\ref{eq:PES_SecondQuantization}) starting from the integrals of the potential operator in the modal basis, which are provided as input following the formatting given in Listing~\ref{lst:vibdump}
+
+\begin{lstlisting}[language=qcmaquis,caption={Example of an vDMRG integral file for the potential
+											  expressed in the $n$-mode format.},
+				   label=lst:vibdump, mathescape=true]
+  1 0   1 0   $\langle \chi_{0}^{(1)}(Q_1) | V^{[1]}(Q_1) | \chi_{0}^{(1)}(Q_1) \rangle$
+  1 1   1 0   $\langle \chi_{1}^{(1)}(Q_1) | V^{[1]}(Q_1) | \chi_{0}^{(1)}(Q_1) \rangle$
+  1 0   1 1   $\langle \chi_{0}^{(1)}(Q_1) | V^{[1]}(Q_1) | \chi_{1}^{(1)}(Q_1) \rangle$
+  $\vdots$
+  2 0   2 0   $\langle \chi_{0}^{(2)}(Q_2) | V^{[2]}(Q_2) | \chi_{0}^{(2)}(Q_2) \rangle$
+  2 1   2 0   $\langle \chi_{1}^{(2)}(Q_2) | V^{[2]}(Q_2) | \chi_{0}^{(2)}(Q_2) \rangle$
+  2 0   2 1   $\langle \chi_{0}^{(2)}(Q_2) | V^{[2]}(Q_2) | \chi_{1}^{(2)}(Q_2) \rangle$
+  $\vdots$
+  3 0   3-0   $\langle \chi_{0}^{(3)}(Q_3) | V^{[3]}(Q_3) | \chi_{0}^{(3)}(Q_3) \rangle$
+  3 1   3-0   $\langle \chi_{1}^{(3)}(Q_3) | V^{[3]}(Q_3) | \chi_{0}^{(3)}(Q_3) \rangle$
+  3 0   3-1   $\langle \chi_{0}^{(3)}(Q_3) | V^{[3]}(Q_3) | \chi_{1}^{(3)}(Q_3) \rangle$ 
+  $\vdots$
+  1 0   1-0   2 0   2 0  $\langle \chi_{0}^{(1)}(Q_1) \chi_{0}^{(2)}(Q_2) | V^{[1,2]}(Q_1, Q_2) | \chi_{0}^{(1)}(Q_1) \chi_{0}^{(2)}(Q_2) \rangle$
+  1 0   1-0   2 0   2 1  $\langle \chi_{0}^{(1)}(Q_1) \chi_{0}^{(2)}(Q_2) | V^{[1,2]}(Q_1, Q_2) | \chi_{0}^{(1)}(Q_1) \chi_{1}^{(2)}(Q_2) \rangle$
+  $\vdots$
+  1 0   1 0   2 0   2 0   3 0   3 0   $\langle \chi_{0}^{(1)}(Q_1) \chi_{0}^{(2)}(Q_2) \chi_{0}^{(3)}(Q_3) | V^{[1,2,3]}(Q_1, Q_2, Q_3) | \chi_{0}^{(1)}(Q_1) \chi_{0}^{(2)}(Q_2) \chi_{0}^{(3)}(Q_3) \rangle$
+\end{lstlisting}
+
+Note that modes are numbered starting from 1, while modals are numbered starting from 0.
+Moreover, Hermitianity is not assumed, so that both the $\langle \chi_{1}^{(1)}(Q_1) | V^{[1]}(Q_1) | \chi_{0}^{(1)}(Q_1) \rangle$ and the $\langle \chi_{0}^{(1)}(Q_1) | V^{[1]}(Q_1) | \chi_{1}^{(1)}(Q_1) \rangle$ entries must be given as input.
+The corresponding input file is given in Listing~\ref{lst:nModeVibInput}
+
+\begin{lstlisting}[language=qcmaquis,
+				   caption={Input example for a $n$-mode vDMRG-FCI calculation},
+				   label=lst:nModeVibInput]
+  // == Generic input section ==
+  nsweeps                         = 6
+  max_bond_dimension              = 1000
+  model_library                   = coded
+  lattice_library                 = coded
+  optimization                    = twosite 
+  integral_file                   = FCIDUMP_Vibrational
+  integral_cutoff                 = 1.0E-10
+  init_state                      = 'default'
+
+  // == vDMRG-specific input ==
+  // The symmetry group is nu1 for all n-mode vDMRG calculations
+  symmetry                        = nu1
+  // This is the only lattice associated with the n-mode vDMRG method
+  LATTICE                         = "nmode lattice" 
+  // n-mode vDMRG model
+  MODEL                           = nMode
+  // String with the overall number of modals
+  nmode_num_modes                 = 20
+  // Maximum order of many-body coupling (1, 2, or 3)
+  nmode_max_coupling              = 2
+  // Number of basis functions per mode
+  nmode_num_basis                 = "10 10"
+\end{lstlisting}
+
+Note that the size of the vector given as input for the \texttt{nmode\_num\_basis} basis must be equal to the integer given as \texttt{nmode\_num\_modes} parameter.
+
+\subsection{DMRG for excitonic Hamiltonians}
+\label{sec:excitonic}
+
+\qcm\ can be applied also to simulate vibronic processes, \textit{i.e.} phenomena that are tuned simultaneously by electronic and vibrational effects.
+Specifically, \qcm\ implements two classes of Hamiltonians: the Holstein-Hubbard Hamiltonian and \textit{ab initio} vibronic Hamiltonians based on the linear vibronic coupling (LVC) model.
+
+The Holstein-Hubbard Hamiltonian is defined by the following parameters:
+
+\begin{itemize}
+  \item The number $N_\text{mol}$ of monomers forming the aggregate (input parameter \texttt{n\_excitons}).
+  \item The number of vibrations $N_\text{vib}$ of vibrational degrees of freedom that are included per aggregate (input parameter \texttt{vibronic\_nmodes}).
+  \item The harmonic frequencies $\omega_i$ associated with the $N_\text{vib}$ vibrations (defined in the \texttt{integral\_file}, as described below).
+  \item The $N_\text{vib}$ parameters $g_j$ of the LVC Hamiltonian (defined in the \texttt{integral\_file}, as described below).
+  \item The Coulomb coupling $J$ (keyword \texttt{J\_coupling}).
+\end{itemize}
+
+The Holstein-Hubbard Hamiltonian $\mathcal{H}_\text{Hol}$ reads, in practice:
+
+\begin{equation}
+ \begin{aligned}
+  \mathcal{H}_\text{Hol} =& | 0_1 0_2 \cdots 0_{N_\text{mol}} \rangle \langle 0_1 0_2 \cdots 0_{N_\text{mol}} |
+	\left[ \sum_{i=1}^{N_\text{mol}} \sum_{j=1}^{N_\text{vib}} 
+			\left( -\frac{1}{2} \mathcal{T}(q_{j}^{(i)}) + \frac{1}{2} \omega_j \left( q_j^{(i)} \right)^2 \right)
+	\right] \\
+	+& \sum_{i=1}^{N_\text{mol}} | 0_1 \cdots 1_i \cdots 0_{N_\text{mol}} \rangle 
+								 \langle 0_1 \cdots 1_i \cdots 0_{N_\text{mol}} | 
+	   \sum_{j=1}^{N_\text{vib}}
+	   \left( -\frac{1}{2} \mathcal{T}(q_{j}^{(i)}) + \frac{1}{2} \omega_j \left( q_j^{(i)} \right)^2 
+			  + g_j q_{j}^{(i)} \right) \\
+    +& J \sum_{i=1}^{N_\text{mol}-1} | 0_1 \cdots 1_i \cdots 0_{N_\text{mol}} \rangle
+	   								 \langle 0_1 \cdots 1_{i+1} \cdots 0_{N_\text{mol}} |
+ \end{aligned}
+ \label{eq:HolsteinHubbard}
+\end{equation}
+
+In Eq.~(\ref{eq:HolsteinHubbard}), $| 0_1 0_2 \cdots 0_{N_\text{mol}} \rangle$ represents the electronic configuration, where all the molecules are in their ground-state electronic state, while $| 0_1 \cdots 1_i \cdots 0_{N_\text{mol}} \rangle$ represents the configuration in which the $i$-th molecule of the aggregate has been excited.
+
+In order to apply DMRG to the Hamiltonian defined by Eq.~(\ref{eq:HolsteinHubbard}), the degrees of freedom must be mapped onto an appropriate one-dimensional lattice.
+Following Ref.~\citenum{Baiardi2019_TDDMMRG}, we construct a DMRG lattice that includes $N_\text{mol}$ electronic sites, each one describing a different molecule of the aggregate, and $N_\text{vib}$ vibrational sites per molecule.
+The overall number of sites is, therefore, $N_\text{mol} \times (N_\text{vib}+1)$.
+Currently, \qcm\ supports two different choices for sorting these sites into the DMRG lattice, that can be specified \textit{via} the \texttt{vibronic\_sorting} keyword:
+
+\begin{itemize}
+  \item The first sorting is obtained as follows: the first site is the electronic site associated with the first molecule of the aggregate.
+The following $N_\text{vib}$ sites are associated with the vibrations of the first molecule.
+The next site is then mapped to the electronic site associated with the second molecule, and the following $N_\text{vib}$ sites are used to represent the corresponding vibrational degrees of freedom.
+This procedure is repeated until all the molecules of the aggregate have been included.
+This mapping is obtained by setting the keyword \texttt{vibronic\_sorting} to \texttt{intertwined}.
+  \item Alternatively, it is possible to map the electronic degrees of freedom onto the first $N_\text{mol}$ sites of the lattice, and use the remaining ones for the vibrational parts.
+  Such a mapping is obtained by setting the \texttt{vibronic\_sorting} keyword to \texttt{firstele}.
+\end{itemize}
+
+\begin{lstlisting}[language=qcmaquis,
+				   caption={Input example for an excitonic DMRG calculation},
+				   label=lst:excitonicInput]
+  // == Generic input section ==
+  nsweeps                         = 6
+  max_bond_dimension              = 20
+  model_library                   = coded
+  lattice_library                 = coded
+  optimization                    = singlesite
+  integral_file                   = FCIDUMP_Excitonic
+  init_state                      = 'default'
+  // The symmetry group is u1 for all vibronic DMRG calculations
+  symmetry                        = u1
+  // This is lattice is used for both vibronic and excitonic calculations.
+  LATTICE                         = "vibronic lattice" 
+  // Vibronic model
+  MODEL                           = vibronic
+  J_coupling                      = -500
+  // This could have been also 'firstele'
+  vibronic_sorting                = 'intertwined'
+  // Number of modes and of molecules
+  vibronic_nmodes                 = 10
+  n_excitons                      = 6
+\end{lstlisting}
+
+We report in Listing~\ref{lst:excitonicInput} an examples of an input file for running a DMRG calculation on an excitonic Hamiltonian composed by 6 molecules, each one with 10 vibrational modes, coupled by a Coulomb coupling of -500.
+
+\begin{lstlisting}[language=qcmaquis,caption={Excitonic integral file format.},
+				   label=lst:excitonicDump]
+	 103.00      1     1
+	-103.00     -1    -1
+	 105.50      2     2
+	-105.50     -2    -2
+	 270.00      3     3
+	-270.00     -3    -3
+	 276.00      4     4
+	-276.00     -4    -4
+	 375.50      5     5
+	-375.50     -5    -5
+	 662.50      6     6
+	-662.50     -6    -6
+	 685.50      7     7
+	-685.50     -7    -7
+	 734.50      8     8
+	-734.50     -8    -8
+	 785.50      9     9
+	-785.50     -9    -9
+	 814.50     10    10
+	-814.50    -10   -10
+	 129.30      1     0
+	 138.36      2     0
+	 105.26      3     0
+	 150.16      4     0
+	 192.93      5     0
+	 187.38      6     0
+	 884.27      7     0
+	 425.76      8     0
+	 639.67      9     0
+	 454.68     10     0
+\end{lstlisting}
+
+The parameters defining the excitonic Hamiltonian must be included in the integral file, as shown in Listing~\ref{lst:vibdump}.
+The first 20 lines of the file define the Harmonic component of (see Eq.~(\ref{eq:HolsteinHubbard})).
+The terms are encoded as for the canonical quantization-based Hamiltonian of vDMRG.
+The last 10 lines of the file define the $g_j$ parameter of Eq.~(\ref{eq:HolsteinHubbard}).
+Also in this case, we follow the same encoding as for the canonical quantization vDMRG.
+
+\subsection{DMRG for vibronic Hamiltonians}
+\label{sec:vibronic}
+
+\qcm\ can apply DMRG also to \textit{ab-initio} vibronic Hamiltonians beyond the excitonic approximation to study the excited-state dynamics of isolated molecular systems.
+The specific form of the Hamiltonian that \qcm\ supports is the following:
+
+\begin{equation}
+  \mathcal{H} =
+    \begin{bmatrix}
+	   \mathcal{H}_1(Q_1,\ldots,Q_{N_\text{vib}})
+	     & \mathcal{V}_{12}(Q_1,\ldots,Q_{N_\text{vib}}) 
+         & \cdots 
+		 & \mathcal{V}_{1 N_\text{ele}}(Q_1,\ldots,Q_{N_\text{vib}}) \\
+	   \mathcal{V}_{21}(Q_1,\ldots,Q_{N_\text{vib}})
+	     & \mathcal{H}_2(Q_1,\ldots,Q_{N_\text{vib}})
+         &  
+		 & \\
+	  \vdots
+	     & 
+         & \ddots 
+		 & \\
+	  \mathcal{V}_{N_\text{ele} 1}(Q_1,\ldots,Q_{N_\text{vib}})
+	     &
+         &  
+		 & \mathcal{H}_{N_\text{ele}}(Q_1,\ldots,Q_{N_\text{vib}}) \\
+   \end{bmatrix}
+  \label{eq:VibronicHamiltonian}
+\end{equation}
+%
+for a system with $N_\text{ele}$ electronic states and $N_\text{vib}$ vibrational degrees of freedom.
+$\mathcal{H}_{i}(Q_1,\ldots,Q_{N_\text{vib}})$ represents the vibrational Hamiltonian associated with the $i$-th electronic state, and is expressed as:
+
+\begin{equation}
+  \mathcal{H}_{i}(Q_1,\ldots,Q_{N_\text{vib}}) = E_i^{(eq)}
+	+ \sum_{j=1}^{N_\text{vib}} \omega_j^{(i),2} \left( Q_j^2  + P_j^2 \right) \, ,
+  \label{eq:DiagonalVibronic}
+\end{equation}
+%
+where $E_i^{(eq)}$ is the electronic energy of the equilibrium position, and $Q_j$ and $P_j$ are the $j$-th dimensionless  normal coordinate and the corresponding conjugate momentum, respectively, with harmonic frequency $\omega_j$.
+$\mathcal{V}_{ij}(Q_1,\ldots,Q_{N_\text{vib}})$ is the non-adiabatic coupling between the $i$-th and the $j$-th electronic state, and is expressed as:
+
+\begin{equation}
+  \mathcal{V}_{ij}(Q_1,\ldots,Q_{N_\text{vib}}) =
+   \sum_{k=1}^{N_\text{vib}} g_k^{(i,j)} Q_k
+	+ \sum_{k,l=1}^{N_\text{vib}} h_{k,l}^{(i,j)} Q_k Q_l \, ,
+  \label{eq:OffDiagonalVibronic}
+\end{equation}
+%
+where $g_k^{(i,j)}$ and $h_{k,l}^{(i,j)}$ are the first- and second-order coupling non-adiabatic coupling terms, respectively.
+
+\begin{lstlisting}[language=qcmaquis,
+				   caption={Input example for a vibronic DMRG calculation},
+				   label=lst:vibronicInput]
+  nsweeps                         = 6
+  max_bond_dimension              = 20
+  model_library                   = coded
+  lattice_library                 = coded
+  optimization                    = singlesite
+  integral_file                   = FCIDUMP_Vibronic
+  init_state                      = 'default'
+  symmetry                        = u1
+  LATTICE                         = "vibronic lattice"
+  MODEL                           = vibronic
+  // Note that here the [n_excitons] parameter is replaced 
+  // by the [vibronic_nstates] one
+  vibronic_nmodes                 = 4
+  vibronic_nstates                = 2
+\end{lstlisting}
+
+A sample input file for a DMRG calculation on a vibronic Hamiltonian is given in Listing~\ref{lst:vibronicInput}.
+The key difference between the input for the excitonic calculation, reported in Listing~\ref{lst:excitonicInput}, is that the \texttt{n\_excitons} parameter is replaced by the \texttt{vibronic\_nstates} one, which corresponds to the $N_\text{ele}$ parameter introduced above.
+The integral file \texttt{FCIDUMP\_Vibronic} should contain the definition of the parameters entering the vibronic Hamiltonian, \textit{i.e.}, $\omega_j^{(i)}$, $g_k^{(i,j)}$, and $h_{k,l}^{(i,j)}$.
+We show in Listing~\ref{lst:vibronicDump} the format of the integral file for the case of the 4-mode vibronic Hamiltonian of pyrazine that describes the S$_1$ and S$_2$ electronic states, taken from Ref.~\citenum{Raab1999_Pyrazine}.
+
+\begin{lstlisting}[language=qcmaquis,caption={Vibronic integral file format.},
+				   label=lst:vibronicDump]
+EL_ST 0 0
+-4114.23                0      0
+  325.84799             1      1     
+ -325.84799            -1     -1
+  559.34550             2      2     
+ -559.34550            -2     -2
+  674.68277             3      3     
+ -674.68277            -3     -3     
+  518.21124             4      4     
+ -518.21124            -4     -4     
+EL_ST 1 1
+ 4114.23                0      0
+  325.84799             1      1     
+ -325.84799            -1     -1
+  559.34550             2      2     
+ -559.34550            -2     -2
+  674.68277             3      3     
+ -674.68277            -3     -3     
+  518.21124             4      4     
+ -518.21124            -4     -4     
+EL_ST 0 0
+ -791.22989             1      0
+ -405.69687             2      0
+ 1171.11703             3      0
+EL_ST 1 1
+ 1092.881251            1      0
+-1379.877165            2      0
+  302.457910            3      0
+EL_ST 0 0
+   0.16131088           1      1
+   8.71078783           1      2
+ -16.45371035           1      3
+   8.71078783           2      1
+ -65.33090875           2      2
+  38.23067993           2      3
+ -16.45371035           3      1
+  38.23067993           3      2
+  -9.35603137           3      3
+ -93.47965832           4      4
+EL_ST 1 1 
+ -73.96104114           1      1
+ -24.03532198           1      2
+ -15.24387871           1      3
+ -24.03532198           2      1
+  39.35985614           2      2
+   9.27537593           2      3
+ -15.24387871           3      1
+   9.27537593           3      2
+   1.77441974           3      3
+ -93.47965832           4      4
+EL_ST 0 1
+ 1677.63321200          4      0
+  -80.65544294          4      1
+  -44.44114904          4      2
+   10.24324125          4      3
+   50.65161814          1      4
+  -44.44114904          2      4
+   10.24324125          3      4
+EL_ST 1 0
+ 1677.63321200          4      0
+  -80.65544294          4      1
+  -44.44114904          4      2
+   10.24324125          4      3
+   50.65161814          1      4
+  -44.44114904          2      4
+   10.24324125          3      4
+\end{lstlisting}
+
+Note that the constants entering the vibronic Hamiltonian are represented as in the excitonic case.
+The header \texttt{EL\_ST X Y}, where \texttt{X} and \texttt{Y} are integer smaller than $N_\text{ele}$, indicates the electronic states to which the vibrational constants are associated with.
+For instance, the header \texttt{EL\_ST 0 0} indicates that the constants in lines 2-10 are associated with the 0-0 block of Eq.~\ref{eq:VibronicHamiltonian}.
+Therefore, they represent, in practice, the vibrational Hamiltonian for the S$_0$ state.
+Note that the coefficients associated with the \texttt{0   0} pair of indices (line 3) is the constant energy shift $E_i^{(eq)}$ of Eq.~(\ref{eq:DiagonalVibronic}) (also referred to as $\Delta$ in Ref.~(\citenum{Raab1999_Pyrazine})).
+Each block of Listing~\ref{lst:vibronicDump} corresponds to a class of terms entering Eq.~(1) of Ref.~(\citenum{Raab1999_Pyrazine}).
+Specifically:
+
+\begin{enumerate}
+  \item Lines 3-10 are the $\omega_i$ coefficients for the S$_1$ state.
+  \item Lines 13-20 are the $\omega_i$ coefficients for the S$_2$ state.
+  \item Lines 22-24 are the $a_i$ coefficients.
+  \item Lines 26-28 are the $b_i$ coefficients.
+  \item Lines 30-39 are the $a_{ij}$ coefficients.
+  \item Lines 41-50 are the $b_{ij}$ coefficients.
+  \item Lines 52-58 and 60-66 collect the $c_i$ and $c_{ij}$ coefficients.
+\end{enumerate}
+
+Note that the vibronic Hamiltonian is Hermitian and, therefore, the $\mathcal{V}_{ij}$ and $\mathcal{V}_{ji}$ operators form a hermitian conjugate pair.
+However, this hermiticity is not exploited in the integral file, where both the $\mathcal{V}_{01}$ and $\mathcal{V}_{10}$ operators are explicitly given.
+
+\subsection{Pratical considerations}
+\label{sec:Practical-vDMRG}
+
+% Informations on how to compile the vibrational module of QCMaquis
+The vDMRG library is not activated if the default CMake settings are used when compiling \qcm.
+In order to run vDMRG simulation, the \texttt{BUILD\_VIBRATIONAL} CMake flag should be set to \texttt{ON}.
+This is, however, not sufficient.
+In fact, the Hamiltonians supported by \qcm\ are classified based on the symmetry group to which they belong.
+These groups must be included, at compile time, in the \texttt{BUILD\_SYMMETRIES} CMake keyword.
+By default, \texttt{BUILD\_SYMMETRIES} includes \texttt{TwoU1PG} (\textit{i.e.}, the symmetry of the non spin-adapted non-relativistic Hamiltonian, with the possibility of including point group symmetry) and \texttt{SU2U1PG} (the spin-adapted counterpart).
+The \texttt{NONE} point group must be added in order to compile the canonical quantization version of vDMRG, and the \texttt{NU1} one must be added to compile the $n$-mode variant.
+
+Similarly, in order to enable the support for vibronic and excitonic Hamiltonians, two parameters must be changed when the \texttt{Makefile} is generated.
+First, the \texttt{BUILD\_VIBRONIC} CMake flag should be set to \texttt{ON}.
+Second, the \texttt{U1} group should be added to the list of compiled symmetries, under the \texttt{BUILD\_SYMMETRIES} keyword.
+
+In conclusion, we note that both excitonic and vibronic Hamiltonians are mostly applied to quantum dynamics simulations of non-equilibrium processes.
+We will discuss in the following how to run quantum dynamics simulations with \qcm.
+
+\section{QCMaquis pre-Born--Oppenheimer module}
+\label{sec:preBO}
+
+The \qcm\ pre-Born--Oppenheimer (pre-BO) module provides a tool to perform DMRG-CASCI calculations for \textit{multicomponent} Hamiltonians consisting of multiple spin-$\frac12$ particle types.
+
+QCMaquis follows the formatting of the integral file shown in Listing~\ref{lst:prebodump}.
+The first line, with only a single float, contains a constant shift in energy, e.g., point-charge repulsion of classical nuclei.
+The following lines contain the terms in the Hamiltonian where pairs of numbers $m-n$ describe a single second-quantization operator.
+The first number denotes the particle type (starting with 0) and the second number the orbital index associated with the given particle type (also starting with 0 for all types).
+We assume physics notation and normal-ordering, i.e., the following would be a valid term
+
+\begin{equation}
+  \left(\int \phi_{1-1}(1)\phi_{0-1}(2)g(1,2)\phi_{1-0}(1)\phi_{0-1}(2)\right)
+  a^\dagger_{1-1} a^\dagger_{0-1} a_{0-1} a_{1-0}\,.
+\end{equation}
+
+The second and third lines contain single-paricle type \textit{hopping} terms.
+The fourth line stands for a single-particle type interaction, and the fifth
+line for an interaction between different particle types. 
+Note here that the aforementioned convention implies that operators belonging to
+the same particle types in a given string of operators must be either at
+positions 1 and 4 or 2 and 3.
+
+\begin{lstlisting}[language=qcmaquis,caption={Pre-BO integral file
+format.},label=lst:prebodump]
+1.145901278714102                       
+   0-0         0-0                                      -2.451508889071943
+   1-0         1-0                                      2.30233235666604
+   0-2         0-0         0-1         0-1              -0.08690785254423253
+   1-1         0-1         0-1         1-0              4.887021698358933e-05
+\end{lstlisting}
+
+Listing~\ref{lst:qcm} shows an input example for a pre-BO DMRG-FCI calculation. 
+It assumes that a pre-BO integral file with has been generated. Additionally,
+the 1-particle RDM and the mutual information can be measured and printed to a
+file.
+
+\begin{lstlisting}[language=qcmaquis,
+				   caption={Input example for a pre-BO DMRG-FCI calculation},
+				   label=lst:preboinput]
+  nsweeps                         = 6
+  max_bond_dimension              = 1000
+
+  // Pre-BO specific input
+  // String with the number of parciles for each type
+  PreBO_ParticleTypeVector        = "4 1"
+  // Spin-1/2 fermion (1) or spin-0 boson (0)
+  // Attention: the boson-related functionality has not been fully tested, yet.
+  PreBO_FermionOrBosonVector      = "1 1"
+  // Number of orbitals for each type.
+  PreBO_OrbitalVector             = "3 3"
+  // Number of alpha-beta particles for each type
+  // Note: for spin-0 bosons it would be a single number.
+  PreBO_InitialStateVector        = "2 2 1 0"
+
+optimization                    = twosite 
+symmetry                        = nu1
+
+model_library                   = coded
+lattice_library                 = coded
+LATTICE                         = "preBO lattice"
+
+// Your model
+MODEL                           = PreBO
+init_state                      = 'default'
+
+// Hamiltonian file
+integral_file                   = FCIDUMP.E_4_H1_1
+integral_cutoff                 = 1.0E-10
+
+MEASURE[1rdm]          = 1
+MEASURE[mutinf]        = 1
+\end{lstlisting}
+
+%\subsection{Practical considerations}
+
+\section{\qcm\ examples library}
 \label{sec:examples}
@@
  For this example calculation we provide the input files for both the single-site (\texttt{Benzene\_6e6o.TD.SS.inp}) and the two-site (\texttt{Benzene\_6e6o.TD.TS.inp}) propagators.
+ \item \texttt{vDMRG}: example of a vDMRG calculation based on the $n$-mode second-quantization (\texttt{FAD\_2Mode\_TS.inp}).
+ The integral file is generated for a two-mode potential describing the formic acid dimer based on a discrete variable representation primitive basis set.
+ For each mode, 10 modal basis functions are included. \\
+ The same folder also contains an example calculation for the canonical second-quantization, based on the single-site (\texttt{C2H4.vDMRG.m50.SS.input}) and on the two-site (\texttt{C2H4.vDMRG.m50.TS.input}) optimization algorithm.
+ The PES is taken from Ref.~\citenum{Berkelback2021_vHBCI} and represents the potential energy surface of ethylene as a sixth-order force field.
+ The calculation uses the single-site optimizer and is initiated from the ONV associated with the ground-state wave function. \\
+ Note that the calculations should be run with the executable \texttt{applications/dmrg/dmrg}, which is automatically generated by each \qcm\ compilation configuration. \\
 \end{itemize}
@@
 For all input files, we provide also the corresponding output file that can be used as a reference to ensure that the \qcm\ executables have been compiled correctly.
-
-\subsection{QCMaquis pre-Born--Oppenheimer module}\label{sec:preBO}
-
-The \qcm\ pre-Born--Oppenheimer (pre-BO) module provides a tool to perform DMRG-CASCI calculations for \textit{multicomponent} Hamiltonians consisting of multiple spin-$\frac12$ particle types.
-
-QCMaquis follows the formatting of the integral file shown in Listing~\ref{lst:prebodump}.
-The first line, with only a single float, contains a constant shift in energy, e.g., point-charge repulsion of classical nuclei.
-The following lines contain the terms in the Hamiltonian where pairs of numbers $m-n$ describe a single second-quantization operator.
-The first number denotes the particle type (starting with 0) and the second number the orbital index associated with the given particle type (also starting with 0 for all types).
-We assume physics notation and normal-ordering, i.e., the following would be a valid term
-
-\begin{equation}
-  \left(\int \phi_{1-1}(1)\phi_{0-1}(2)g(1,2)\phi_{1-0}(1)\phi_{0-1}(2)\right)
-  a^\dagger_{1-1} a^\dagger_{0-1} a_{0-1} a_{1-0}\,.
-\end{equation}
-
-The second and third lines contain single-paricle type \textit{hopping} terms.
-The fourth line stands for a single-particle type interaction, and the fifth
-line for an interaction between different particle types. 
-Note here that the aforementioned convention implies that operators belonging to
-the same particle types in a given string of operators must be either at
-positions 1 and 4 or 2 and 3.
-
-\begin{lstlisting}[language=qcmaquis,caption={Pre-BO integral file
-format.},label=lst:prebodump]
-1.145901278714102                       
-   0-0         0-0                                      -2.451508889071943
-   1-0         1-0                                      2.30233235666604
-   0-2         0-0         0-1         0-1              -0.08690785254423253
-   1-1         0-1         0-1         1-0              4.887021698358933e-05
-\end{lstlisting}
-
-Listing~\ref{lst:qcm} shows an input example for a pre-BO DMRG-FCI calculation. 
-It assumes that a pre-BO integral file with has been generated. Additionally,
-the 1-particle RDM and the mutual information can be measured and printed to a
-file.
-
-\begin{lstlisting}[language=qcmaquis,caption={Input example for a pre-BO DMRG-FCI
-calculation},label=lst:preboinput]
-nsweeps                         = 6
-max_bond_dimension              = 1000
-
-// Pre-BO specific input
-// String with the number of parciles for each type
-PreBO_ParticleTypeVector        = "4 1"
-// Spin-1/2 fermion (1) or spin-0 boson (0)
-// Attention: the boson-related functionality has not been fully tested, yet.
-PreBO_FermionOrBosonVector      = "1 1"
-// Number of orbitals for each type.
-PreBO_OrbitalVector             = "3 3"
-// Number of alpha-beta particles for each type
-// Note: for spin-0 bosons it would be a single number.
-PreBO_InitialStateVector        = "2 2 1 0"
-
-optimization                    = twosite 
-symmetry                        = nu1
-
-model_library                   = coded
-lattice_library                 = coded
-LATTICE                         = "preBO lattice"
-
-// Your model
-MODEL                           = PreBO
-init_state                      = 'default'
-
-// Hamiltonian file
-integral_file                   = FCIDUMP.E_4_H1_1
-integral_cutoff                 = 1.0E-10
-
-MEASURE[1rdm]          = 1
-MEASURE[mutinf]        = 1
-\end{lstlisting}
-
-%\clearpage
-\clearpage
-\newpage
 

### 2023-04-20 dmrg/framework/dmrg/models/generate_mpo/1D_mpo_maker.hpp
Version 3.1.3

@@ -1,29 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2015 Laboratory of Physical Chemistry, ETH Zurich
- *               2015-2015 by Sebastian Keller <sebkelle@phys.ethz.ch>
- *               2021 by Alberto Baiardi <abaiardi@ethz.ch>
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #ifndef ONED_MPO_MAKER_H
 #define ONED_MPO_MAKER_H
@@ -244,7 +224,7 @@ namespace generate_mpo
      */
     template<class Matrix, class SymmGroup>
     MPO<Matrix, SymmGroup>
-    make_destroy_mpo(const Lattice& lat, Model<Matrix, SymmGroup>& model, int iOrb, IonizedOrbital alphaOrBeta=IonizedOrbital::Up)
[…]

### 2023-04-20 dmrg/framework/dmrg/models/generate_mpo/utils.hpp

@@ -1,28 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2011-2011 by Bela Bauer <bauerb@phys.ethz.ch>
- * 
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- * 
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #ifndef GENERATE_MPO_UTILS_H
 #define GENERATE_MPO_UTILS_H
@@ -172,126 +153,6 @@ namespace generate_mpo
             if (count(out_taken.begin(), out_taken.end(), k) == 0 &&
                 count(in_taken.begin(), in_taken.end(), k) == 0)
                 return k;// +-------------+
-//  NMODE LATTICE
-// +-------------+
-//TODO ALB Maybe move the setup of the orbital order here. When doing it, remember to change also get_prop_
-
-class NModeLattice : public lattice_impl
-{
-public:
-    // Types definition
[…]

### 2023-04-20 dmrg/framework/dmrg/models/measurements/overlap.h

@@ -1,29 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2011-2011 by Bela Bauer <bauerb@phys.ethz.ch>
- *               2011-2013    Michele Dolfi <dolfim@phys.ethz.ch>
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #ifndef MEASUREMENTS_OVERLAP_H
 #define MEASUREMENTS_OVERLAP_H

### 2023-04-20 dmrg/framework/dmrg/models/measurements/tagged_nrankrdm_2u1.h

@@ -1,0 +1,242 @@
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
+
+#ifndef MEASUREMENTS_TAGGED_NRANKRDM_TWOU1_H
+#define MEASUREMENTS_TAGGED_NRANKRDM_TWOU1_H
+
+namespace measurements {
+
+template <class Matrix, class SymmGroup, class = void>
+class TaggedNRankRDM : public measurement<Matrix, SymmGroup> {
+  // Types declaration
+  typedef measurement<Matrix, SymmGroup> base;
+  typedef typename Model<Matrix, SymmGroup>::term_descriptor term_descriptor;
+  typedef Lattice::pos_t pos_t;
+  typedef std::vector<pos_t> positions_type;
+  typedef typename base::op_t op_t;
+  typedef typename OPTable<Matrix, SymmGroup>::tag_type tag_type;
+  typedef typename base::value_type value_type;
+  typedef std::vector<tag_type> tag_vec;
+  typedef std::vector<tag_vec> bond_term;
+  typedef std::pair<std::vector<tag_vec>, value_type> scaled_bond_term;
+
+public:
+  /** @brief Class constructor */
+  TaggedNRankRDM(std::string const& name_, const Lattice & lat, std::shared_ptr<TagHandler<Matrix, SymmGroup> > tag_handler_,
+                 tag_vec const & identities_, tag_vec const & fillings_, std::vector<scaled_bond_term> const& ops_,
+                 bool half_only_, positions_type const& positions_ = positions_type(), std::string const& ckp_ = std::string(""))
+    : base(name_), lattice(lat), tag_handler(tag_handler_), positions_first(positions_), identities(identities_),
+      fillings(fillings_), operator_terms(ops_), half_only(half_only_), bra_ckp(ckp_)
+  {
+    pos_t extent = lattice.size();
+    // the default setting is only required for "measure_correlation"
+    if (positions_first.size() == 0 && operator_terms[0].first.size() == 2)
+    {
+        positions_first.resize(extent);
+        std::iota(positions_first.begin(), positions_first.end(), 0);
+    }
+    //this->cast_to_real = is_hermitian_meas(ops[0]);
+    this->cast_to_real = false;
+  }
+
+  /** @brief Calculates the RDM elements */
[…]

### 2023-04-20 dmrg/framework/dmrg/models/vibrational/none/model.hpp

@@ -1,49 +1,30 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2021 Institute for Theoretical Physics, ETH Zurich
- *               2021- by Alberto Baiardi <alberto.baiardi@phys.chem.ethz.ch>
- * 
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- * 
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #ifndef MODELS_VIBRATIONAL_NONE_H
 #define MODELS_VIBRATIONAL_NONE_H
 
-#ifdef DMRG_VIBRATIONAL
+// #ifdef DMRG_VIBRATIONAL
 
+#include <set>
 #include <sstream>
 #include "dmrg/models/model.h"
 #include "dmrg/models/measurements.h"
 #include "dmrg/utils/BaseParameters.h"
 #include "dmrg/models/model_helper.hpp"
 #include "dmrg/models/vibrational/VibrationalIntegralParser.hpp"
[…]

### 2023-04-20 dmrg/framework/dmrg/models/vibrational/u1/VibronicModel.hpp

@@ -1,28 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2021 Institute for Theoretical Physics, ETH Zurich
- *               2021- by Alberto Baiardi <abaiardi@ethz.ch>
- * 
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- * 
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #ifndef MODELS_VIBRONIC_U1_H
 #define MODELS_VIBRONIC_U1_H
@@ -31,7 +12,6 @@
 
 #include "dmrg/models/measurements/local_at.h"
 #include "dmrg/models/model_helper.hpp"
-#include "dmrg/models/vibrational/VibrationalModelTraitClass.hpp"
 #include "dmrg/models/vibrational/VibrationalHelperClass.hpp"
 #include "dmrg/models/vibrational/VibronicIntegralParser.hpp"
 
@@ -67,7 +47,7 @@ public:
     {
         // Variable definition
         int nMax = parameters_["Nmax"];
[…]

### 2023-04-20 dmrg/framework/dmrg/mp_tensors/mpotensor_detail.h

@@ -1,28 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2013-2013 by Sebastian Keller <sebkelle@phys.ethz.ch>
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #ifndef MPOTENSOR_DETAIL_H
 #define MPOTENSOR_DETAIL_H

### 2023-04-20 dmrg/framework/dmrg/mp_tensors/mps.hpp

@@ -1,32 +1,12 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2011-2011 by Bela Bauer <bauerb@phys.ethz.ch>
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #include "dmrg/mp_tensors/mps.h"
 #include "contractions.h"
-
 #include "dmrg/utils/archive.h"
 
 #include <limits>
@@ -48,27 +28,20 @@ std::string MPS<Matrix, SymmGroup>::description() const
 
 template<class Matrix, class SymmGroup>
 MPS<Matrix, SymmGroup>::MPS()
-: canonized_i(std::numeric_limits<size_t>::max())
-{ }
+    : canonized_i(std::numeric_limits<size_t>::max()) { }
 
 template<class Matrix, class SymmGroup>
-MPS<Matrix, SymmGroup>::MPS(size_t L)
-: data_(L)
[…]

### 2023-04-20 dmrg/framework/dmrg/mp_tensors/mps_initializers_helper.h

@@ -1,49 +1,33 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2021 Institute for Theoretical Physics, ETH Zurich
- *               2021 by Alberto Baiardi <alberto.baiardi@sns.it>
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #ifndef MPS_INITIALIZER_HELPER_H
 #define MPS_INITIALIZER_HELPER_H
 
 #include "dmrg/block_matrix/indexing.h"
 #include "dmrg/block_matrix/symmetry.h"
+#include "dmrg/utils/BaseParameters.h"
+
+#include <boost/tuple/tuple.hpp>
 
 /**
  * @brief Helper class for the MPS initialization.
- * 
+ *
  * This class wraps all the methods that are useful when initializing an MPS
  * from a given set of input data.
- * 
+ *
  * For now, we include only a method, [GenerateIndexFromString], that converts
  * an input of integer (provided with the `init_basis_state` input parameter)
[…]

### 2023-04-20 dmrg/framework/dmrg/mp_tensors/ts_reshape.h

@@ -1,38 +1,15 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *                    Laboratory for Physical Chemistry, ETH Zurich
- *               2011-2011 by Sebastian Keller <sebkelle@phys.ethz.ch>
- *                            Michele Dolfi <dolfim@phys.ethz.ch>
- * 
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- * 
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #include <vector>
 #include <utility>
 #include <alps/numeric/real.hpp>
 
-#include "dmrg/utils/random.hpp"
 #include "dmrg/mp_tensors/mpstensor.h"
-
 #include "dmrg/mp_tensors/reshapes.h"
 #include "dmrg/block_matrix/block_matrix_algorithms.h"
 

### 2023-04-20 dmrg/framework/dmrg/mp_tensors/twositetensor.hpp

@@ -1,29 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2011-2011 by Sebastian Keller <sebkelle@phys.ethz.ch>
- *                            Michele Dolfi <dolfim@phys.ethz.ch>
- * 
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- * 
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #include "dmrg/mp_tensors/twositetensor.h"
 #include "dmrg/mp_tensors/mpstensor.h"
@@ -31,7 +11,6 @@
 #include "dmrg/mp_tensors/reshapes.h"
 #include "dmrg/block_matrix/block_matrix_algorithms.h"
 
-#include "dmrg/utils/random.hpp"
 #include <alps/numeric/real.hpp>
 
 #include <vector>
@@ -43,18 +22,12 @@
[…]

### 2023-04-20 dmrg/framework/dmrg/utils/checks.h

@@ -1,28 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2015 Institute for Theoretical Physics, ETH Zurich
- *               2015-2015 by Sebastian Keller <sebkelle@phys.ethz.ch>
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #ifndef CHECKS_H_
 #define CHECKS_H_

### 2023-04-20 dmrg/framework/dmrg/utils/parallel/scheduler.hpp

@@ -1,29 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2011-2013 by Alexandr Kosenkov <alex.kosenkov@gmail.com>
- *                         by Michele Dolfi <dolfim@phys.ethz.ch>
- * 
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- * 
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #ifndef PARALLEL_SCHEDULER_HPP
 #define PARALLEL_SCHEDULER_HPP

### 2023-04-20 dmrg/framework/dmrg/utils/parallel/scheduler/inplace.hpp

@@ -1,29 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2011-2013 by Alexandr Kosenkov <alex.kosenkov@gmail.com>
- *                         by Michele Dolfi <dolfim@phys.ethz.ch>
- * 
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- * 
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #ifndef PARALLEL_SCHEDULER_INPLACE_HPP
 #define PARALLEL_SCHEDULER_INPLACE_HPP

### 2023-04-20 dmrg/framework/dmrg/utils/time_stopper.h

@@ -1,28 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2013-2013 by Michele Dolfi <dolfim@phys.ethz.ch>
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 /// Adapted from alps/ngs/scheduler/stop_callback.hpp
 

### 2023-04-20 dmrg/tests/Fixtures/LiHFixture.h

@@ -1,0 +1,115 @@
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
+
+#ifndef TEST_LiH_FIXTURE_H
+#define TEST_LiH_FIXTURE_H
+
+#include "maquis_dmrg.h"
+#include "dmrg/block_matrix/symmetry.h"
+
+/**
+ * @brief Fixture class for LiH.
+ * The integrals are generated for LiH with d=1 Angstrom, singlet, cc-pVDZ basis set, CAS(2,4).
+ * Note that the core orbitals of Li are excluded.
+ * The data were generated with MOLCAS by Leon Freitag.
+ */
+struct LiHFixture
+{
+// Types definition
+  using RealIntegralMapType = typename maquis::integral_map<double>;
+
+  /** @brief Constructor for the fixture class */
+  LiHFixture() {
+    // Real-valued integrals
+    integralsLiH = RealIntegralMapType {
+                                        { { 1, 1, 1, 1 },   0.597263715971     },
+                                        { { 1, 1, 2, 1 },   0.106899493032E-01 },
+                                        { { 2, 1, 2, 1 },   0.215106203079E-02 },
+                                        { { 2, 2, 2, 1 },   0.310109624236E-02 },
+                                        { { 1, 1, 2, 2 },   0.205585394307     },
+                                        { { 2, 2, 2, 2 },   0.266391740477     },
+                                        { { 1, 1, 3, 1 },  -0.294917730728E-08 },
+                                        { { 2, 1, 3, 1 },   0.185509604436E-07 },
+                                        { { 2, 2, 3, 1 },  -0.437276233593E-07 },
+                                        { { 3, 1, 3, 1 },   0.983873746561E-01 },
+                                        { { 3, 2, 3, 1 },  -0.550389961495E-02 },
+                                        { { 3, 3, 3, 1 },  -0.148187533022E-07 },
+                                        { { 1, 1, 3, 2 },   0.656187599279E-07 },
+                                        { { 2, 1, 3, 2 },  -0.686302758417E-08 },
+                                        { { 2, 2, 3, 2 },  -0.185120456521E-07 },
+                                        { { 3, 2, 3, 2 },   0.329840207584E-02 },
+                                        { { 3, 3, 3, 2 },   0.705823689976E-07 },
[…]

### 2023-04-20 dmrg/tests/VibrationalDMRG/test_BilinearlyCoupled.cpp

@@ -1,0 +1,47 @@
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
+
+#define BOOST_TEST_MAIN
+
+#ifdef DMRG_VIBRATIONAL
+
+#include <boost/test/included/unit_test.hpp>
+#include "Fixtures/WatsonFixture.h"
+#include "maquis_dmrg.h"
+
+/**
+ * @brief Watson-based calculation on a bilinearly-coupled Harmonic Hamiltonian
+ * Note that the reference energy is taken from the work
+ * "Calculating vibrational spectra with sum of product basis functions without
+ * storing full-dimensional vectors or matrices"
+ * by the group of Tucker Carrington
+ */
+BOOST_FIXTURE_TEST_CASE(Test_vDMRG_Calculation_BilinearlyCoupled, WatsonFixture)
+{
+#ifdef HAVE_TrivialGroup
+  using InterfaceType = maquis::DMRGInterface<double>;
+  // Adds the final input parameters
+  parametersBilinearly.set("init_type", "basis_state_generic");
+  parametersBilinearly.set("init_basis_state", "0,0,0,0,0,0");
+  parametersBilinearly.set("optimization", "singlesite");
+  parametersBilinearly.set("alpha_initial", 1.0E-8);
+  parametersBilinearly.set("alpha_initial", 1.0E-15);
+  parametersBilinearly.set("alpha_initial", 0.);
+  parametersBilinearly.set("nsweeps", 20);
+  parametersBilinearly.set("ngrowsweeps", 2);
+  parametersBilinearly.set("nmainsweeps", 2);
+  parametersBilinearly.set("max_bond_dimension", 20);
+  parametersBilinearly.set("MODEL", "watson");
+  // Creates the interface
+  InterfaceType interface(parametersBilinearly);
+  interface.optimize();
+  // The reference value can be calculated based on the Harmonic approximation
+  BOOST_CHECK_CLOSE(interface.energy(), 3.8164041, 1.0E-5);
+#endif
+}
+
+#endif // DMRG_VIBRATIONAL

### 2023-04-20 dmrg/tests/prebo/SymbolicJordanWignerTest.cpp

@@ -1,28 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2021 Institute for Theoretical Physics, ETH Zurich
- *               2021 by Alberto Baiardi <alberto.baiardi@phys.chem.ethz.ch>
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #define BOOST_TEST_MODULE SymbolicJordanWigner
 

### 2023-04-20 dmrg/tests/prebo/symbolicoperator_test.cpp

@@ -1,28 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2021 Institute for Theoretical Physics, ETH Zurich
- *               2021 by Alberto Baiardi <alberto.baiardi@phys.chem.ethz.ch>
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #define BOOST_TEST_MODULE SymbolicOperator
 

### 2023-04-20 dmrg/tests/test_realtime.cpp

@@ -1,28 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2021 Institute for Theoretical Physics, ETH Zurich
- *               2021-2022 by Alberto Baiardi <abaiardi@ethz.ch>
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #define BOOST_TEST_MAIN
 
@@ -84,27 +65,27 @@ BOOST_FIXTURE_TEST_CASE( TestRealTime, BenzeneFixture )
  * @brief Tests that the energy is conserved along a relativistic TD-DMRG propagation.
  * The data are obtained for N2+ and the 3-21G basis set.
  */
-BOOST_FIXTURE_TEST_CASE( TestRealTimeRelativistic, TestTimeEvolverFixture )
+BOOST_FIXTURE_TEST_CASE(TestRealTimeRelativistic, TestTimeEvolverFixture)
 {
-    // Generic settings
[…]

### 2023-04-20 dmrg/doc/qcmaquis_manual_standalone.tex

\documentclass[bibliography=totoc,12pt,a4paper]{scrartcl}

% -- Loads packages --
\usepackage[nott]{kpfonts}
\usepackage{array}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{braket}
\usepackage[bf,singlelinecheck=off]{caption}
\usepackage{colortbl}
\usepackage[T1]{fontenc}
\usepackage{framed}
\usepackage{graphicx}
\usepackage[colorlinks=true]{hyperref} % web links
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{longtable}
\usepackage{multirow}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{placeins}
\usepackage{varwidth}
\usepackage{xcolor}

% Definition of colours for hyperref
\definecolor{linkcolour}{rgb}{0,0.2,0.6}
\definecolor{citecolour}{rgb}{0,0.8,0.2}
\hypersetup{colorlinks,breaklinks,urlcolor=linkcolour,linkcolor=linkcolour,citecolor=citecolour}
\setcounter{secnumdepth}{4}

% -- New commands --
\newcommand{\relv}{release version 3.1.3}
\newcommand{\qcm}{\textsc{QCMaquis}}
\newcommand{\scine}{SCINE}
\newcommand{\hostp}{\mol}
\newcommand{\molbuild}{\texttt{my-Molcas-build}}
\newcommand{\molsrc}{\texttt{my-Molcas-src}}
\newcommand{\qcmsrc}{\texttt{my-QCMaquis-src}}
\newcommand{\qcmbuild}{\texttt{my-QCMaquis-build}}
\newcommand{\kwd}[1]{\texttt{#1}}

\newcommand{\tableoptionskip}{0.3em}
\newcommand{\myemph}[1]{\textbf{#1}}

\lstdefinelanguage{qcmaquis}
{
%
    keywords=[1]{L,nelec,spin,nsweeps,max_bond_dimension,integral_file,
				 init_type,LATTICE,lattice_library,MODEL,model_library,irrep,symmetry},
	morekeywords=[2]{conv_thresh,integral_cutoff,truncation_initial,truncation_final,
					 MEASURE,chkpfile,resultfile,storagedir},
	keywordstyle=[1]\color{green!20!red},
	keywordstyle=[2]\color{green!20!blue},
	commentstyle=\color{gray},
	sensitive=false,
	morecomment=[l]{//},
	morecomment=[s]{/*}{*/},
}

\lstset{
	numbers=left,
	numberstyle=\tiny,
	numberblanklines=true,
	basicstyle=\ttfamily\small,
	frame=single,
	showlines=true,
	flexiblecolumns=true,
	breaklines=true,
	postbreak={\llap{\textcolor{red}{$\lhook\joinrel\longrightarrow$\kern2.1em}}},
	escapechar=|
}

\title{User guide to the standalone version of the \scine-\qcm\ software}
\author{Alberto Baiardi, Robin Feldmann, Leon Freitag, Sebastian Keller, \\ Stefan Knecht, Yingjin Ma, Christopher Stein and Markus Reiher}
\publishers{ETH Z\"urich, Laboratorium f\"ur Physikalische Chemie, Vladimir-Prelog-Weg 2,\\ CH-8093 Z\"urich}
\date{\today}

\begin{document}

\bibliographystyle{achemso}

\pagenumbering{roman}
\maketitle
\thispagestyle{empty}
\begin{center}
\end{center}
\vspace{-1.85cm}
\centerline{\large{\relv}}

\vspace{2cm}

\noindent We kindly request that, for reproducibility reasons, any use of the \qcm\ software suite for density matrix renormalization group (DMRG) calculations that results in published material should cite the following paper:

\begin{framed}
\noindent Keller,~S.; Dolfi,~M.; Troyer,~M.; Reiher,~M. \emph{J. Chem. Phys.}
  \textbf{2015}, \emph{143}, 244118, \href{https://doi.org/10.1063/1.4939000}{doi:10.1063/1.4939000}.
\end{framed}

The reference works for the vibrational DMRG method are the following:

\begin{framed}
  \noindent Baiardi,~A.; Stein,~C.J.; Barone,~V.; Reiher,~M. \emph{J. Chem. Theory Comput.}
  \textbf{2017}, \emph{13}, 3764, \href{https://doi.org/10.1021/acs.jctc.7b00329}{doi:10.1021/acs.jctc.7b00329}. \\
  \noindent Baiardi,~A.; Stein,~C.J.; Barone,~V.; Reiher,~M. \emph{J. Chem. Phys.}
  \textbf{2019}, \emph{150}, 094113, \href{https://doi.org/10.1063/1.5068747jctc.7b00329}{doi:10.1063/1.5068747jctc.7b00329}. \\
  \noindent Glaser,~N.; Baiardi~A.; Reiher,~M. ``Tensor Network States for Vibrational Spectroscopy'', chapter 3 in: Joel M. Bowman (ed.), Vibrational Dynamics of Molecules, pp. 80-144 (\textbf{2022}), World Scientific, \href{https://arxiv.org/abs/2109.08961}{arXiv.org}. \\
\end{framed}

The reference papers for the Nuclear-Electron All-Particle DMRG method (see Section~\ref{sec:preBO}) are the following:

\begin{framed}
  \noindent Muolo,~A.; Baiardi,~A.; Feldmann,~R.; Reiher,~M. \emph{J. Chem. Phys.} \textbf{2020}, \emph{152}, 204103, \href{https://doi.org/10.1063/5.0007166}{doi:10.1063/5.0007166}. \\
  \noindent Feldmann,~R.; Muolo,~A.; Baiardi,~A.; Reiher,~M. \emph{J. Chem. Theory Comput.} \textbf{2022}, \emph{18}, 234, \href{https://doi.org/10.1021/acs.jctc.1c00913}{doi:10.1021/acs.jctc.1c00913}. 
\end{framed}

The reference paper for DMRG[FEAST] is the following:

\begin{framed}
  \noindent Baiardi,~A.; Kelemen,~A.~K.; Reiher,~M. \emph{J. Chem. Theory Comput.} \textbf{2022}, \emph{18}, 415, \href{https://doi.org/10.1021/acs.jctc.1c00984}{doi:10.1021/acs.jctc.1c00984}.
\end{framed}

\qcm\ builds upon the \href{http://alps.comp-phys.org/static/mps\_doc/index.html}{ALPS MPS code}, which has been developed at ETH Zurich by Michele Dolfi and Bela Bauer in the group of Matthias Troyer with contributions from Sebastian Keller and Alexandr Kosenkov and at the University of Geneva by Timoth{\'e}e Ewart and Adrian Kantian in the group of Thierry Giamarchi.
ALPS is a general open-source framework for the description of strongly correlated many-particle systems.
For further information on the ALPS project, please visit \href{alps.comp-phys.org}{alps.comp-phys.org} and refer to the following reference papers:

\begin{framed}
  \noindent M. Dolfi, B. Bauer, S. Keller, A. Kosenkov, T. Ewart, A. Kantian, T. Giamarchi, M. Troyer, \textit{Comp. Phys. Commun.} \textbf{2014}, 12, 3430. \href{https://doi.org/10.1016/j.cpc.2014.08.019}{doi:10.1016/j.cpc.2014.08.019} \\
  B. Bauer, \textit{et al.} (ALPS Collaboration), The ALPS project release 2.0: open source software for strongly correlated systems, \textit{J. Stat. Mech.} \textbf{2011} P05001. \href{http://dx.doi.org/10.1088/1742-5468/2011/05/P05001}{http://dx.doi.org/10.1088/1742-5468/2011/05/P05001}.
\end{framed}

\clearpage
\tableofcontents
\clearpage

\pagenumbering{arabic}

\section{Introduction to the \qcm\ Software Suite}

\qcm\ can be used as a standalone code.
The standalone version of \qcm\ is used especially in two cases:

\begin{enumerate}
  \item for electronic-structure calculations relying on integrals that are calculated with programs other than \texttt{OpenMolcas}
  \item for DMRG calculations beyond time-independent electronic-structure simulations.
\end{enumerate}

Section~\ref{sec:install-qcm-stand}\ describes in details the installation process for a standalone version of \qcm.
We will then focus on the former application field, while the latter will be discussed in the remaining Section of this user guide.
Note that, for a description of the \qcm{} software as integrated within \textsc{OpenMolcas}, refer to the \texttt{qcmaquis\_manual.pdf} user guide (referred to in the following simply as ``\textsc{OpenMolcas}-\qcm{} guide'')

\section{Installation}
\label{sec:install-qcm-stand}

\noindent We describe in the following how to build and install the standalone version of the \qcm\ software.
The installation of \qcm\ has been tested for different operating systems and compiler/math libraries environments.
Their list can be found in the other user guide.
While other combinations might work equally well they are \emph{not} officially supported.

\subsection{Prerequisite}
\label{subsec:requirements}

The following libraries are required and need to be installed manually by the user, although they are usually provided by the system package manager.

\begin{itemize}
 \item GSL
 \item HDF5
 \item SZIP
 \item Boost
\end{itemize}

Additionally, \qcm{} distribution includes a customised version of the ALPS 2.3.0 library, which is compiled automatically.

\subsection{Download \qcm}
\label{subsec:download-stand}

The \qcm{} standalone version may be found in the \texttt{\url{https://github.com/qcscine/qcmaquis}} repository.

\subsection{Setting up a build folder}

Clone the GitHub repository into a new directory \texttt{\qcmsrc}.
Create a build folder \texttt{\qcmbuild} -- note that this folder \emph{does not} necessarily have to be a subfolder of \qcmsrc\ -- and change to this new folder:

\begin{verbatim}
  mkdir /path/to/my-QCMaquis-build && cd /path/to/my-QCMaquis-build
\end{verbatim}

\subsection{Configuration}
\label{subsec:configure-stand}

For simplicity, we will describe the configuration steps for the most popular compiler suites GNU.
How to setup and use a shared-memory OMP installation of \qcm\ is described in Section~\ref{subsubsec:parallel-stand}.

\subsubsection{Configuration with the GNU compiler suite}
\label{subsubsec:gnu-conf-stand}

\noindent To configure \qcm\ with the GNU compiler suite type

\begin{verbatim}
  CC=gcc CXX=g++ cmake /path/to/my-QCMaquis-src
\end{verbatim}
%
where we assumed that the Intel Math Kernel Library (MKL) is available (recommended option).
If the MKL libraries are not available, \qcm\ will search for other suitable math libraries installed on the operating system. If none are found the configuration step will stop with an appropriate message.

\begin{framed}
  \noindent Note: Intel compiler is currently not supported. Running a binary compiled with Intel compilers may result in segmentation faults. We are currently investigating the issue.
\end{framed}

\subsubsection{Shared-memory OMP parallel configuration}
\label{subsubsec:parallel-stand}

By default \qcm\ is built as shared-memory OMP parallelized version which should work with either compiler suite, GNU or Intel.

\vspace{2ex}

In order to exploit the shared-memory OMP parallelization of \qcm\ the user is strongly encouraged to set at runtime the environment variable

\begin{verbatim}
  export OMP_NUM_THREADS=XX
\end{verbatim}
%
where \texttt{XX}\ specifies the number of shared-memory cores to be used.
The default (\emph{depending on the operating system!}) is to use all available cores.

\subsubsection{Building and installation}
\label{subsubsec:build-stand}

After a successful configuration, type
\begin{verbatim}
  make
\end{verbatim}
%
or

\begin{verbatim}
  make -j8
\end{verbatim}
%
to compile \qcm\ (in parallel on 8 cores) and install all its components in the build folder \texttt{\qcmbuild}.

\section{Keywords and Options}
\label{sec:qcmaquis-kw}

In the following we describe:

\begin{enumerate} 
  \item compulsory keywords (Section \ref{subsec:compul-qcm})
  \item optional keywords (Section \ref{subsec:optional-qcm}) for \qcm\ DMRG calculations 
  \item keywords for property calculations (Section \ref{subsec:prop-qcm})
\end{enumerate}

Most of the \qcm\ keywords have default settings that guarantee convergence in the general case.
A reasonable choice of default values for optional keywords is given in our example \qcm\ input file in Section~\ref{sec:input-qcm}.

\subsection{Compulsory keywords}
\label{subsec:compul-qcm}

The keywords in Table~\ref{tab:standard_kw_auto} have to be set for every DMRG calculation because they may crucially affect the accuracy of the final result.
Their choice depends, e.g., on the molecule under consideration, the specific Hamiltonian to which DMRG is applied to (electronic, vibrational, vibronic $\ldots$), and on the active space size.

\begin{framed}
  \noindent\textbf{Note:}\ We strongly encourage any new user of \qcm\ to carefully read first Ref.\ \citenum{Keller_CHIMIA_Determining_2014}\ which gives a detailed introduction to
  DMRG calculations in quantum chemistry.
  Further quantum chemical DMRG sample calculations starting from different computational setups are discussed, for example, in Refs.~\cite{interface,ru-dmrg,fde-dmrg,srdft-dmrg}.
\end{framed}

\begin{table}
  \caption{Compulsory \qcm{} keywords to be set by the user.}
  \label{tab:standard_kw_auto}
  \begin{tabular}{cV{11.5cm}}
    \toprule
        Keyword     &             Description    \\
    \midrule
	\rowcolor{gray!20}
	\texttt{nsweeps} & Overall number of optimization sweeps \\
    \texttt{nelec}   & Total number of electrons 
					   (used only for electronic-structure calculations). \\
	\rowcolor{gray!20}
    \texttt{irrep}   & Irreducible representation of the point group symmetry of the target state. 
					   Counting starts with 0 which has to be the totally symmetric
				       representation. (\textbf{Note:} this keyword is compulsory only for
					   electronic-structure calculations) \\
    \texttt{spin}   & Total spin ($2\times S$) of the target state, for example:
					  \texttt{spin=0} (singlet), \texttt{spin=1} (doublet),
					  \texttt{spin=2} (triplet) 
					  (used only for electronic-structure calculations) \\
    \rowcolor{gray!20}
    \texttt{L}      & Length of the DMRG lattice. For electronic-structure calculations, this 
                      corresponds to the number of orbitals in the active space. \\
    \texttt{integral\_file} 
                    & Path and filename of the file 
					  (for example \texttt{integral\_file = /path/to/file/FCIDUMP}) where the
					  definition of the Hamiltonian is stored. See below for specific examples for 
					  different classes of Hamiltonians. \\
    \rowcolor{gray!20}
    \texttt{symmetry} & Defines the total symmetry group.
					    For electronic-structure calculations, the default is the combined spin- (SU2) and point symmetry group (PG) \texttt{su2u1pg}, where the \texttt{pg}-suffix should be omitted for better performance if the molecule is C1-symmetric.
					    For test purposes, it is possible to switch off spin-adaptation, again with or without point group symmetry: \texttt{2u1(pg)} .
						In the latter case the keywords \texttt{spin} and \texttt{nelec} have no meaning.
						Instead \texttt{u1\_total\_charge1} and \texttt{u1\_total\_charge2} corresponding to the number of up and down electrons have to be specified.
 						Other symmetry groups must be used for other Hamiltonians, see below. \\
    \texttt{init\_state} 
					 & Algorithm to construct the initial guess for the MPS.
					   Possible options are \texttt{default} (corresponding to an MPS with entries
					   initialized randomly), \texttt{const} (corresponding to an MPS where all
					   the entries are initialized to the same values) and \texttt{hf}.
				 	   Usage of \texttt{hf} generates an MPS consisting of only the determinant
					   defined on the \texttt{hf\_occ} keyword. \\
	\rowcolor{gray!20}
	\texttt{LATTICE} & Definition of the mapping from the single-particle basis onto the DMRG 
					   lattice (must be set to \texttt{orbitals} for electronic-structure calculations, see below for the definition of this keyword for other 
					   Hamiltonians). \\
	\texttt{MODEL}   & Type of the Hamiltonian to which DMRG is applied. \\
	\rowcolor{gray!20}
    \texttt{lattice\_library} 
					 & Must be set to \texttt{"coded"}. \\
	\texttt{model\_library}
					 & Must be set to \texttt{"coded"}. \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Optional keywords}
\label{subsec:optional-qcm}

\begin{table}[htbp!]
  \caption{Optional keywords for \qcm\ calculations.}
  \label{tab:advanced_kw}
  \begin{tabular}{cV{11.5cm}}
	\toprule
    Keyword & Description \\
	\midrule
	\rowcolor{gray!20}
    \texttt{orbital\_order}
			& Sorting of the orbitals along the one dimensional lattice.
			  The order has to be entered as a string of comma separated orbital numbers.
		      By default, the orbitals are sorted in increasing value of the index used 
		      in the \texttt{integral\_file}. \\
    \texttt{max\_bond\_dimensions}
			& Maximum value of the bond dimension $m$ of the MPS. \\
	\rowcolor{gray!20}
    \texttt{conv\_thresh}
			& Energy convergence threshold (in the same units used in \texttt{integral\_file}).
			  If the lowest energy from the previous sweep differs from the lowest energy of the current sweep by less than \texttt{conv\_thresh}, the DMRG calculation stops.
			  For example, \texttt{conv\_thresh = 1e-6}. \\
    \texttt{ietl\_jcd\_tol} 
		    & Convergence threshold for the Jacobi-Davidson diagonalization that is carried out
			  for each site. Example: \texttt{ietl\_jcd\_tol = 1e-6}. \\
	\rowcolor{gray!20}
    \texttt{ietl\_jcd\_maxiter}
			& Maximum number of iterations in the Jacobi-Davidson diagonalization.\\
    \texttt{truncation\_initial}
			& If during the \texttt{ngrowsweeps}, the sum of the squared value of the discarded 
			  singular values for $m$ retained states is lower than the value defined here, more block states will be discarded until the sum increases 
		      to \texttt{truncation\_initial}. \\
	\rowcolor{gray!20}
    \texttt{truncation\_final}
			& If during the \texttt{nmainsweeps}, the sum of the discarded singular values for $m$ retained states is lower than the value defined here, more block states will be discarded until the discarded sum increases to \texttt{truncation\_final}. \\
	\texttt{measure\_each} 
			& Tells the program to compute the expectation values every $2\ \times$ \texttt{measure\_each} sweeps. \\
	\rowcolor{gray!20}
	\texttt{chkp\_each}
			& Tells the program to update the checkpoint file every $2\ \times$ \texttt{chkp\_each} sweeps. \\
    \texttt{hf\_occ}
			& Occupation of the starting orbitals when \texttt{init\_state} = \texttt{hf}.
			  Comma-separated string of occupation aliases, defined as: 4~=~full, 3~=~up,
			  2~=~down, 1~=~empty. \\
	\rowcolor{gray!20}
	\texttt{nmainsweeps} 
			& Number of sweeps in which \texttt{truncation\_final} is used in the singular value decomposition. \\
	\texttt{ngrowsweeps}
			& Number of sweeps in which \texttt{truncation\_initial} is used in the singular value decomposition. \\
	\rowcolor{gray!20}
	\texttt{storagedir} 
			& Scratch directory for temporary files (this tames the memory usage of the code). \\
	\texttt{chkpfile} 
			& Path and name of the folder in which the MPS is stored.\\
	\rowcolor{gray!20}
	\texttt{resultfile} 
			& Path and filename of the file storing the results (energy, expectation values...). \\
	\bottomrule
  \end{tabular}
\end{table}

The keywords summarized in Table~\ref{tab:advanced_kw} may be exploited by the more experienced user but can be safely ignored in conventional DMRG calculation.
They may affect the convergence and accuracy of the final result, though.

\subsection{Keywords for excited-state calculations}
\label{subsec:ExcitedState}

By default, \qcm{} calculates excited states with the methods described in Ref.~\citenum{Keller_JChemPhys_efficient_2015}.
Specifically, the first excited state is obtained with a DMRG optimization that is constrained onto the space orthogonal to the ground state.
The second excited state is obtained with an optimization orthogonal to the ground and first excited state, and so on.
Excited states are, therefore, calculated sequentially.
Table~\ref{tab:es_kw} describes the keywords to be activated to run excited-state DMRG calculations.

\begin{table}[htbp!]
  \centering
  \caption{Keywords for excited-state \qcm\ calculations.}
  \label{tab:eskw}
  \begin{tabular}{cV{11.5cm}}
	\toprule
    Keyword & Description \\
	\midrule
	\rowcolor{gray!20}
	\texttt{n\_ortho\_states}
		    & Number of states the current wave function is to be orthogonalized against to 
			  must be specified here. \\
	\texttt{ortho\_states}
			& Path(s) and filename(s) of the MPS checkpoint file(s) that store the lower 
			  lying states to which the current MPS shall be orthogonal to. \\
	\bottomrule
  \end{tabular}
\end{table}

\noindent \textbf{Note}: two other methods for calculating excited states, DMRG[IPI] and DMRG[FEAST], will be presented below.
\qcm\ currently supports them only for vibrational-structure calculations.

\subsection{Keywords for expectation value calculations}
\label{subsec:prop-qcm}

\qcm\ can (in principle) compute expectation values for any one- or two-particle operator that can be formulated in second quantization.
Table~\ref{tab:expval} comprises a list of the available property keywords available in \qcm.

The one-particle reduced density matrix as well as the one-particle spin-density matrix are implicitly computed from the expectation values of some of the operators contained in \texttt{MEASURE[ChemEntropy]}.

If a property is available only in 2U1 symmetry, the SU2U1 checkpoint may be transformed into 2U1 group with the \texttt{mps\_transform(\_pg)} tool (see Section~\ref{sec:tools-qcm})

\begin{table}
 \footnotesize
 \label{tab:expval}
 \caption{Expectation value calculations available in the release version of \qcm.}
  \begin{tabular}{cV{8.5cm}}
    \toprule
    Keyword & Description \\
    \midrule
    \texttt{MEASURE[ChemEntropy]} 
			& All expectation values over the operators required to calculate the mutual
			  information (as specified in Ref.~\cite{entanglement}) will be computed.
			  Available only for 2U1(PG) symmetry. \\
	\texttt{MEASURE[1rdm]} 
			& Computes the one-particle reduced density matrix, without the additional
			  correlators contained in the \texttt{ChemEntropy} measurement. \\
	\texttt{MEASURE[2rdm]}
			& Computes the two-particle reduced density matrix \\
	\texttt{MEASURE[3rdm]}
			& Computes the three-particle reduced density matrix. 
			  A two-index subblock may be calculated with \texttt{MEASURE[3rdm]="p1:p2@\textit{j,i}"}.
			  Available only in 2U1(pg) symmetry. \\
	\texttt{MEASURE[4rdm]}
			& Computes the four-particle reduced density matrix, respectively.
			  \texttt{MEASURE[4rdm]="p4:p3:p1:p2@\textit{l,k,i,j}"} allows to compute only a 4-index sub-block of the 4-RDM with the indices \texttt{\textit{i,j,k,l}}.
              Available only in 2U1(pg) symmetry. \\
	\texttt{MEASURE[trans2rdm]="\textit{chk}"}
			& Computes the transition two--particle reduced density matrix between the current 
			  state and the state specified in the \texttt{\textit{chk}} checkpoint file.
			  Available only for 2U1(pg) symmetry. \\
	\texttt{MEASURE[trans3rdm]="\textit{chk}"}
			& Computes the transition three-particle reduced density matrix between the 
			  current state and the state specified in the \texttt{\textit{chk}} checkpoint file. Two-index subblocks can be calculated with \texttt{MEASURE[trans3rdm]="\textit{chk};p1:p2@\textit{j,i}"}. 
			  Available only for 2U1(pg) symmetry. \\
	\texttt{MEASURE\_LOCAL[\textit{name}]="\textit{op}"}
			& Computes $\langle \psi | op_i | \psi \rangle, \, i = 1 \ldots L$. \texttt{Nup}, \texttt{Ndown} and \texttt{Nup*Ndown} are meaningful choices for \textit{op}. Available for \texttt{2U1(OG) only.}  (\emph{Note}: \texttt{\textit{name}}\ defines the name under which the expectation values are stored on the \texttt{resultfile}.) \\
	\texttt{MEASURE\_HALF}\texttt{\_CORRELATIONS[\textit{name}]="\textit{op$_1$:op$_2$}"}
			& Computes $\langle \psi | op_{1i} op_{2j}| \psi \rangle, \, i = 1 \ldots L, j=i+1 \ldots L$. \texttt{Nup}, \texttt{Ndown}, \texttt{Nup*Ndown}, \texttt{cdag\_up}, \texttt{cdag\_down}, \texttt{c\_up}, \texttt{c\_down}, \texttt{cdag\_up*Ndown}, \texttt{c\_up*Ndown}, \texttt{cdag\_down*Nup}, \texttt{c\_down*Nup}, \texttt{cdag\_up*cdag\_down}, \texttt{c\_up*c\_down}, \texttt{cdag\_up*c\_down}, and \texttt{cdag\_down*c\_up}, as \textit{op$_1$} and \textit{op$_2$} are recognized by the program. Available for \texttt{2U1(PG)} only. (\emph{Note}: \texttt{\textit{name}}\ defines the name under which the expectation values are stored on the \texttt{resultfile}.) \\
	\bottomrule
  \end{tabular}
\end{table}

\clearpage

\subsection{QCMaquis input example}
\label{sec:input-qcm}

\begin{lstlisting}[language=qcmaquis,
				   caption={Input example for a DMRG-CASCI(8,8) standalone \qcm{} calculation},
				   label=lst:qcm]
  // Lattice size (= number of orbitals) and number of electrons
  L                    = 8
  nelec                = 8
  // Target symmetry (spin and spatial) of the wave function.
  spin                 = 0
  // Number of optimization sweeps
  nsweeps              = 8
  // Maximum allowed value for the bond dimension
  max_bond_dimension   = 256
  // MPS initialization
  init_type            = 'default'
  // Wave function symmetry (su2u1pg: spin-adapted DMRG + point group)
  symmetry             = 'su2u1pg'
  // File with the one- and two-body integrals
  integral_file        = "FCIDUMP"
  // Definition of the lattice
  LATTICE              = "orbitals"
  // Definition of the Hamiltonian
  MODEL                = "quantum_chemistry"
  // Internal QCMaquis parameters (not to be changed!)
  lattice_library      = "coded"
  model_library        = "coded"
  // Wavefunction irreducible representation (0 == totally symmetric)
  irrep                = 0
  // HDF5 file where the MPS and the results will be stored
  chkpfile             = 'chkp.h5'
  resultfile           = 'result.h5'
  // Convergence threshold for the sweep optimization
  conv_thresh          = 1.0e-6
  // Threshold for the SVD truncation for the initial and sweeps
  truncation_initial   = 1.0e-16
  truncation_final     = 1.0e-7
  // Threshold below which integrals inside the integral_file are neglected
  integral_cutoff      = 1.0e-10
  // Path to the scratch folder (reduces RAM usage)
  storagedir           = '/scratch/$USER/boundaries'
  // List of measurements to be performed at the end of the optimization
  MEASURE[ChemEntropy] = 1
\end{lstlisting}

Listing~\ref{lst:qcm} provides an input example for a ground-state electronic-structure DMRG-CASCI calculation based on an active space comprising 8 electrons and 8 orbitals.
We assume that an integral file \texttt{FCIDUMP} has been generated with an external software (e.g., \textsl{OpenMolcas}) and placed in the same directory with the input file.
We indicate the mandatory keywords in red, and the optional ones in blue.

\section{\qcm{} vibrational DMRG module}
\label{sec:vDMRG}

\qcm{} implements the vibrational DMRG (vDMRG) method introduced in Ref.~\citenum{Baiardi2017_vDMRG}.
vDMRG applies the DMRG algorithm to solve the Schr\"odinger equation associated with the Born--Oppenheimer vibrational Hamiltonian.
For a system with $L$ vibrational degrees of freedom, the latter reads:

\begin{equation}
  \underbrace{\left[ \mathcal{T}(Q_1, \ldots, Q_L) + \mathcal{V}(Q_1, \ldots, Q_L) \right]}_{\mathcal{H}_\text{vib}(Q_1, \ldots, Q_L)} \Psi_n(Q_1, \ldots, Q_L) = E_n \Psi_n(Q_1, \ldots, Q_L)
  \label{eq:VibHam}
\end{equation}
%
where $Q_i$ are the dimensionless Cartesian-based normal coordinates.
The time-independent Schr\"odinger equation associated with Eq.~(\ref{eq:VibHam}) can be solved with methods based on the configuration interaction approach, similarly to their electronic-structure counterpart.
This leads to the vibrational configuration interaction method that, however, becomes impractical for $L$ values larger than 10/20.
The Hamiltonian defined in Eq.~(\ref{eq:VibHam}) must be encoded in a second-quantized form in order to make it possible to apply DMRG on it.
Such a form can be obtained based on the harmonic oscillator-based canonical quantization formalism\cite{Hirata2014_NormalOrdered} described in the next section.
%Such a form can be obtained either based on the harmonic oscillator-based canonical quantization formalism\cite{Hirata2014_NormalOrdered} or on the $n$-mode second quantization introduced by Ove Christiansen.\cite{Christiansen2004_nModeQuantization}
%These two options lead to two different vDMRG formulations, which we will describe in the following sections.

\subsection{Canonical quantization vDMRG}
\label{sec:canonical}

The canonical quantization relies on the harmonic oscillator theory.\cite{Hirata2014_NormalOrdered}
The vibrational wave function is expanded, in practice, as follows:

\begin{equation}
  \Psi(Q_1, \ldots, Q_L) = \sum_{n_1}^{N_\text{max}} \cdots \sum_{n_L=1}^{N_\text{max}}
	C_{n_1 \cdots n_L} | n_1 \cdots n_L \rangle \, ,
  \label{eq:CanonicalQuantizationCI}
\end{equation}
%
where $| n_1 \cdots n_L \rangle$ are occupation number vectors that are constructed based on the well-known harmonic oscillator second quantization.
This means that the $| n_1 \cdots n_L \rangle$ state corresponds, in real-space representation, to:

\begin{equation}
  | n_1 \cdots n_L \rangle \equiv \prod_{i=1}^L \chi_{n_i}(Q_i)
  \label{eq:HarmonicOscillatorBasis}
\end{equation}
%
$\chi_{n_i}(Q_i)$ being the $n_i$-th eigenfunction of the monodimensional harmonic oscillator Hamiltonian associated with the $i$-th mode.
To express the Hamiltonian in terms of such ONVs, we approximate the potential as a power series in terms of the vibrational coordinates $Q_i$ around a given reference geometry $\bm{Q}_\text{eq}$.
For a fourth-order power series, the expansion reads:

\begin{equation}
  \mathcal{V}(Q_1, \ldots, Q_L) 
	= \sum_{i,j=1}^L k_{ij} Q_i Q_j + \sum_{i,j,k=1}^L k_{ijk} Q_i Q_j Q_k
    + \sum_{i,j,k,l=1}^L k_{ijkl} Q_i Q_j Q_k Q_l
  \label{eq:TaylorExpansion}
\end{equation}
%
where we neglected the first-order force constants since we assume that the expansion is performed around a stationary point of the PES.
The $k_{ij}$, $k_{ijk}$, $k_{ijkl}$ parameters are usually known as force-constants.
The position operators $Q_i$ are, then, expressed in terms of the harmonic oscillator ladder operators as follows:

\begin{equation}
  Q_i = \frac{1}{2} \left( b_i^\dagger + b_i \right) \, .
  \label{eq:CanonicalQuantizationPosition}
\end{equation}

Eq.~\ref{eq:CanonicalQuantizationPosition} can be used to express the potential (cfr. Eq.~(\ref{eq:TaylorExpansion})) in terms of solely the $b_i^\dagger$ and $b_i$ operators.
Th kinetic energy operator can be expressed in a similar format based on the counterpart of Eq.~(\ref{eq:CanonicalQuantizationPosition}) for the conjugate momentum operator $P_i$, which is defined as:

\begin{equation}
  P_i = \frac{\mathrm{i}}{2} \left( b_i^\dagger - b_i \right) \, .
  \label{eq:CanonicalQuantizationMomentum}
\end{equation}

The resulting form of the vibrational Hamiltonian is then converted in the MPO format by \qcm\ based on the algorithm described in Ref.~\citenum{Baiardi2017_vDMRG}. \\

\begin{lstlisting}[language=qcmaquis,
				   caption={Input example for a canonical quantization-based vDMRG-FCI calculation},
				   label=lst:canonicalVibInput]
  // == Generic input section ==
  nsweeps                         = 6
  max_bond_dimension              = 100
  model_library                   = coded
  lattice_library                 = coded
  optimization                    = twosite 
  integral_file                   = FCIDUMP_Vibrational
  integral_cutoff                 = 1.0E-10
  init_type                       = 'default'
  L                               = 12
  // == vDMRG-specific input ==
  symmetry                        = none
  LATTICE                         = 'watson lattice'
  MODEL                           = 'watson'
  Nmax                            = 10
\end{lstlisting}

We report in Listing~\ref{lst:canonicalVibInput} an example input for a vDMRG calculation relying on the canonical-quantization based formalism.
The first part of the input has the same structure as that of an electronic-structure DMRG calculation.
Note that the lattice dimension $L$ is, in this case, equal to the overall number of modes.
Unlike for the electronic-structure case, the DMRG symmetry must be set to \texttt{none}, since the canonical quantization-based vibrational Hamiltonan does not possess any symmetry.
Moreover, the \texttt{MODEL} keyword should be set to \texttt{watson} in order to indicate that the DMRG calculation relies on the Watson-based vibrational Hamiltonian.
As the expansion reported in Eq.~\ref{eq:CanonicalQuantizationCI} is virtually unbounded, it must be truncated, in practice, by setting a threshold $N_\text{max}$ for the maximum excitation degree of a mode.
Such a threshold must be specified in input, as shown in Listing~\ref{lst:canonicalVibInput}.

The integral file \texttt{FCIDUMP\_Vibrational} contain the force constants that define the PES through Eq.~\ref{eq:TaylorExpansion}.
\qcm\ supports up to sixth-order force constants by default, which must be provided as input in the \texttt{FCIDUMP\_Vibrational} file as illustrated in Listing~\ref{lst:CanonicalFCIDUMP}.
If higher-order force constants are to be included, the \texttt{DMRG\_ORDERNONE} CMake flag should be set to the corresponding integer $n$ denoting the highest force constant order and the integral files need to provide $n$ integer indices per entry accordingly.
\vspace{.5cm}
\begin{lstlisting}[language=qcmaquis,caption={Example of an vDMRG integral file for the potential
											  expressed in the canonical quantization format.},
				   label=lst:CanonicalFCIDUMP, mathescape=true, rulecolor=\color{black}]
  $\smash{k_{ij}}$     i    j   0   0   0   0
  $\smash{\vdots}$
  $\smash{h_{ij}}$    -i   -j   0   0   0   0
  $\smash{\vdots}$
  $\smash{k_{ijk}}$     i    j   k   0   0   0
  $\smash{\vdots}$
  $\smash{k_{ijkh}}$     i    j   k   h   0   0
  $\smash{\vdots}$
  $\smash{k_{ijkhm}}$     i    j   k   h   m   0
  $\smash{\vdots}$
  $\smash{k_{ijkhmn}}$     i    j   k   h   m   n
\end{lstlisting}
\vspace{.5cm}
The $h$ and $k$ constants define the PES as follows:

\begin{equation}
  \begin{aligned}
    V(\tilde{Q}_1, \ldots, \tilde{Q}_L) 
      &= \sum_{i=1}^L k_{ii} \tilde{Q}_i^2 + \sum_{i=1}^L h_{ii} \tilde{P}_i^2 
      + \sum_{ijk=1}^L k_{ijk} \tilde{Q}_i \tilde{Q}_j \tilde{Q}_k
      + \sum_{ijkl=1}^L k_{ijkl} \tilde{Q}_i \tilde{Q}_j \tilde{Q}_k \tilde{Q}_l \\
      &+ \sum_{ijklm=1}^L k_{ijkl} \tilde{Q}_i \tilde{Q}_j \tilde{Q}_k \tilde{Q}_l \tilde{Q}_m
      + \sum_{ijklmn=1}^L k_{ijkl} \tilde{Q}_i \tilde{Q}_j \tilde{Q}_k \tilde{Q}_l \tilde{Q}_m \tilde{Q}_n \, ,
  \end{aligned}
  \label{eq:PES_Representation}
\end{equation}
%
where we introduced the scaled dimensionless normal coordinates $\tilde{Q}_i$ that are defined as:

\begin{equation}
  \tilde{Q}_i = \sqrt{2} Q_i = \left( b_i^\dagger + b_i \right)
  \label{eq:ScaledNM}
\end{equation}
%
$Q_i$ being the $i$-th dimensionless normal coordinate operator.
The corresponding scaled dimensionless momentum operator $\tilde{P}_i$ is defined as:

\begin{equation}
  \tilde{P}_i = \frac{\sqrt{2}}{\mathrm{i}} P_i = \left( b_i^\dagger - b_i \right)
  \label{eq:ScaledNM_Momentum}
\end{equation}

\subsection{DMRG for excitonic Hamiltonians}
\label{sec:excitonic}

\qcm\ can be applied also to simulate vibronic processes, \textit{i.e.} phenomena that are tuned simultaneously by electronic and vibrational effects.
Specifically, \qcm\ implements two classes of Hamiltonians: the Holstein-Hubbard Hamiltonian and \textit{ab initio} vibronic Hamiltonians based on the linear vibronic coupling (LVC) model.
The Holstein-Hubbard Hamiltonian describes excitation energy processes in molecular aggregates of multiple chromophores, and is defined by the following parameters:

\begin{itemize}
  \item The number $N_\text{mol}$ of monomers forming the aggregate (defined by the input parameter \texttt{n\_excitons}).
  \item The number of vibrational degrees of freedom $N_\text{vib}$ included per monomer (defined by the keyword \texttt{vibronic\_nmodes}).
  \item The harmonic frequencies $\omega_i$ associated with the $N_\text{vib}$ vibrations (defined in the \texttt{integral\_file} file, as described below).
  \item The $N_\text{vib}$ shift parameters $g_j$ of the LVC Hamiltonian (defined in the \texttt{integral\_file} file, as described below).
  \item The Coulomb coupling $J$ between chromophores (defined by the keyword \texttt{J\_coupling}).
\end{itemize}

The Holstein-Hubbard Hamiltonian $\mathcal{H}_\text{Hol}$ reads, in practice:

\begin{equation}
 \begin{aligned}
  \mathcal{H}_\text{Hol} =& | 0_1 0_2 \cdots 0_{N_\text{mol}} \rangle \langle 0_1 0_2 \cdots 0_{N_\text{mol}} |
	\left[ \sum_{i=1}^{N_\text{mol}} \sum_{j=1}^{N_\text{vib}} 
		   \left( \mathcal{T}(q_{j}^{(i)}) + \frac{1}{2} \omega_j \left( q_j^{(i)} \right)^2 \right)
	\right] \\
	+& \sum_{i=1}^{N_\text{mol}} | 0_1 \cdots 1_i \cdots 0_{N_\text{mol}} \rangle 
								 \langle 0_1 \cdots 1_i \cdots 0_{N_\text{mol}} | 
	   \sum_{j=1}^{N_\text{vib}} \left[ 
	   \left( \mathcal{T}(q_{j}^{(i)}) 
			+ \frac{1}{2} \omega_j \left( q_j^{(i)} \right)^2 + g_j q_{j}^{(i)} \right) 
	   \right] \\
    +& J \sum_{i=1}^{N_\text{mol}-1} | 0_1 \cdots 1_i \cdots 0_{N_\text{mol}} \rangle
	   								 \langle 0_1 \cdots 1_{i+1} \cdots 0_{N_\text{mol}} |
 \end{aligned}
 \label{eq:HolsteinHubbard}
\end{equation}

In Eq.~(\ref{eq:HolsteinHubbard}), $| 0_1 0_2 \cdots 0_{N_\text{mol}} \rangle$ represents the electronic configuration in which all the chromphores of the aggregate are in their ground-state electronic state.
The vibrational Hamiltonian associated with this ground-state wave function (\textit{i.e.}) the term between square brackets in the first line of Eq.~(\ref{eq:HolsteinHubbard})) is a $N_\text{vib}$-dimensional harmonic oscillator with normal coordinates $q_{j}^{(i)}$ and frequencies $\omega_j$.
Conversely, $| 0_1 \cdots 1_i \cdots 0_{N_\text{mol}} \rangle$ represents the configuration in which the $i$-th molecule of the aggregate has been excited to the first electronic excited state.
The corresponding vibrational Hamiltonian (\textit{i.e.} the term between square brackets in the second line of Eq.~(\ref{eq:HolsteinHubbard})) is a $N_\text{vib}$-dimensional harmonic oscillator with the same modes and frequencies as for the electronic ground state, but with a linear term ($g_j q_{j}^{(i)}$) associated with a shift in the equilibrium geometry of the chromophore.
The last term of Eq.~(\ref{eq:HolsteinHubbard}) describes the coupling between excited states that are localized on neighboring chromophores.
In practice, this term allows excitation to be transferred from a chromophore to the neighboring one.

In order to apply DMRG to the Holstein-Hubbard Hamiltonian, the degrees of freedom must be mapped onto an appropriate one-dimensional lattice.
Following Ref.~\citenum{Baiardi2019_TDDMMRG}, we construct the DMRG lattice by including $N_\text{mol}$ electronic sites, each one describing a different chromophore of the aggregate.
The physical basis associated with these sites contains two basis functions, one corresponding to the electronic ground-state and one associated with the excited-state.
Moreover, we include $N_\text{vib}$ vibrational sites for each chromophore.
The overall number of vibrational sites is, therefore, $N_\text{mol} N_\text{vib}$, while the overall lattice size (including also the electronic degrees of freedom) is $N_\text{mol} \times (N_\text{vib}+1)$.
Currently, \qcm\ supports two different strategies for mapping these sites onto the DMRG lattice, that can be specified \textit{via} the \texttt{vibronic\_sorting} keyword:

\begin{itemize}
  \item The first sorting is obtained as follows: the first site is the electronic site associated with the first molecule of the aggregate. The following $N_\text{vib}$ sites are associated with the vibrations of the first molecule. The next site is then mapped to the electronic site associated with the second molecule, and the following $N_\text{vib}$ sites are used to represent the corresponding vibrational degrees of freedom. This procedure is repeated until all the molecules of the aggregate have been included. This mapping is obtained by setting the keyword \texttt{vibronic\_sorting} to \texttt{intertwined}, and is maximally efficient when the inter-chromophore coupling is low.
  \item Alternatively, it is possible to map the electronic degrees of freedom onto the first $N_\text{mol}$ sites of the lattice, and use the remaining ones for the vibrational parts.
  Such a mapping is obtained by setting the \texttt{vibronic\_sorting} keyword to \texttt{firstele}.
\end{itemize}

\begin{lstlisting}[language=qcmaquis,
				   caption={Input example for an excitonic DMRG calculation},
				   label=lst:excitonicInput]
  // == Generic input section ==
  nsweeps                         = 6
  max_bond_dimension              = 20
  model_library                   = coded
  lattice_library                 = coded
  optimization                    = singlesite
  integral_file                   = FCIDUMP_Excitonic
  init_type                      = `default'
  // The symmetry group is u1 for all vibronic DMRG calculations
  symmetry                        = u1
  // This is lattice is used for both vibronic and excitonic calculations.
  LATTICE                         = `vibronic lattice'
  // Vibronic model
  MODEL                           = vibronic
  J_coupling                      = -500
  // This could have been also 'firstele'
  vibronic_sorting                = `intertwined'
  // Number of modes and of molecules
  vibronic_nmodes                 = 10
  n_excitons                      = 6
\end{lstlisting}

We report in Listing~\ref{lst:excitonicInput} an examples of an input file for running a DMRG calculation on an excitonic Hamiltonian composed by 6 molecules, each one with 10 vibrational modes, coupled by a Coulomb coupling of -500 (note that the units in which the $J$ parameter is reported must be the same used in the \texttt{integral\_file}).
Note that the \texttt{symmetry} keyword is now set to \texttt{u1}.
In fact, as we discuss in Ref.~\citenum{Baiardi2019_TDDMMRG}, the Holstein-Hubbard Hamiltonian conserves the overall excitation degree or, equivalently, the overall number of excitons.
This means, in practice, that if only one chromophore is excited, the non-equilibrium dynamics governed by the Holstein--Hubbard Hamiltonian will not populate electronic configurations where two (or more) chromophores are simultaneously excited.

\begin{lstlisting}[language=qcmaquis,caption={Excitonic integral file format.},
				   label=lst:excitonicDump]
	 103.00      1     1
	-103.00     -1    -1
	 105.50      2     2
	-105.50     -2    -2
	 270.00      3     3
	-270.00     -3    -3
	 276.00      4     4
	-276.00     -4    -4
	 375.50      5     5
	-375.50     -5    -5
	 662.50      6     6
	-662.50     -6    -6
	 685.50      7     7
	-685.50     -7    -7
	 734.50      8     8
	-734.50     -8    -8
	 785.50      9     9
	-785.50     -9    -9
	 814.50     10    10
	-814.50    -10   -10
	 129.30      1     0
	 138.36      2     0
	 105.26      3     0
	 150.16      4     0
	 192.93      5     0
	 187.38      6     0
	 884.27      7     0
	 425.76      8     0
	 639.67      9     0
	 454.68     10     0
\end{lstlisting}

The parameters defining the excitonic Hamiltonian must be included in the \texttt{integral\_file}.
We report in Listing~\ref{lst:excitonicDump} one example of such a file.
The first 20 lines of the file define the Harmonic component of (see Eq.~(\ref{eq:HolsteinHubbard})).
The last 10 lines of the file define the $g_j$ parameter of Eq.~(\ref{eq:HolsteinHubbard}).
For both classes of terms, we follow the same encoding as for the canonical quantization vDMRG.
This means that the coefficients should be reported in terms of the reduced coordinates $\tilde{Q}_i$ and the corresponding conjugate momenta $\tilde{P}_i$.

\subsection{DMRG for vibronic Hamiltonians}
\label{sec:vibronic}

\qcm\ supports also \textit{ab-initio} vibronic Hamiltonians for studying the excited-state dynamics of isolated molecular systems.
A generic vibronic Hamiltonian reads:

\begin{equation}
  \mathcal{H} =
    \begin{bmatrix}
	   \mathcal{H}_1(Q_1,\ldots,Q_{N_\text{vib}})
	     & \mathcal{V}_{12}(Q_1,\ldots,Q_{N_\text{vib}}) 
         & \cdots 
		 & \mathcal{V}_{1 N_\text{ele}}(Q_1,\ldots,Q_{N_\text{vib}}) \\
	   \mathcal{V}_{21}(Q_1,\ldots,Q_{N_\text{vib}})
	     & \mathcal{H}_2(Q_1,\ldots,Q_{N_\text{vib}})
         &  
		 & \\
	  \vdots
	     & 
         & \ddots 
		 & \\
	  \mathcal{V}_{N_\text{ele} 1}(Q_1,\ldots,Q_{N_\text{vib}})
	     &
         &  
		 & \mathcal{H}_{N_\text{ele}}(Q_1,\ldots,Q_{N_\text{vib}}) \\
   \end{bmatrix}
  \label{eq:VibronicHamiltonian}
\end{equation}
%
for a molecule with $N_\text{ele}$ electronic states and $N_\text{vib}$ vibrational degrees of freedom.
$\mathcal{H}_{i}(Q_1,\ldots,Q_{N_\text{vib}})$ represents the vibrational Hamiltonian associated with the $i$-th electronic state, and is expressed as:

\begin{equation}
  \mathcal{H}_{i}(Q_1,\ldots,Q_{N_\text{vib}}) = E_i^{(eq)}
	+ \frac{1}{2} \sum_{j=1}^{N_\text{vib}} \omega_j^{(i),2} \left( Q_j^2  + P_j^2 \right) 
	+ \sum_{j=1}^{N_\text{vib}} g_{j}^{(i)} Q_j \, ,
  \label{eq:DiagonalVibronic}
\end{equation}
%
where $E_i^{(eq)}$ is the electronic energy of the equilibrium position, and $Q_j$ and $P_j$ are the $j$-th dimensionless  normal coordinate and the corresponding conjugate momentum, respectively, with harmonic frequency $\omega_j$ and linear shift coefficient $g_{j}^{(i)}$.
$\mathcal{V}_{ij}(Q_1,\ldots,Q_{N_\text{vib}})$ is instead the non-adiabatic coupling between the $i$-th and the $j$-th electronic state.
Similarly to the PES, it can be expressed as a Taylor series in terms of $Q_j$ and $P_j$ as follows:

\begin{equation}
  \mathcal{V}_{ij}(Q_1,\ldots,Q_{N_\text{vib}}) =
   \sum_{k=1}^{N_\text{vib}} g_k^{(i,j)} Q_k
	+ \sum_{k,l=1}^{N_\text{vib}} h_{k,l}^{(i,j)} Q_k Q_l \, ,
  \label{eq:OffDiagonalVibronic}
\end{equation}
%
where $g_k^{(i,j)}$ and $h_{k,l}^{(i,j)}$ are the first- and second-order coupling non-adiabatic coupling terms, respectively.

\begin{lstlisting}[language=qcmaquis,
				   caption={Input example for a vibronic DMRG calculation},
				   label=lst:vibronicInput]
  nsweeps                         = 6
  max_bond_dimension              = 20
  model_library                   = coded
  lattice_library                 = coded
  optimization                    = singlesite
  integral_file                   = FCIDUMP_Vibronic
  init_type                      = 'default'
  symmetry                        = u1
  LATTICE                         = "vibronic lattice"
  MODEL                           = vibronic
  // Note that here the [n_excitons] parameter is replaced 
  // by the [vibronic_nstates] one
  vibronic_nmodes                 = 4
  vibronic_nstates                = 2
\end{lstlisting}

A sample input file for a DMRG calculation on a vibronic Hamiltonian is given in Listing~\ref{lst:vibronicInput}.
The key difference between this input and the one reported in Listing~\ref{lst:excitonicInput} for excitonic Hamiltonians is that the keyword \texttt{n\_excitons} is replaced by \texttt{vibronic\_nstates}, which corresponds to the $N_\text{ele}$ parameter in Eq.~(\ref{eq:VibronicHamiltonian}).
The integral file \texttt{FCIDUMP\_Vibronic} should contain the definition of the parameters entering the vibronic Hamiltonian, \textit{i.e.}, $\omega_j^{(i)}$, $g_k^{(i)}$, $g_k^{(i,j)}$, and $h_{k,l}^{(i,j)}$.
We show in Listing~\ref{lst:vibronicDump} the format of the integral file for the case of the 4-mode vibronic Hamiltonian of pyrazine that describes the S$_1$ and S$_2$ electronic states, taken from Ref.~\citenum{Raab1999_Pyrazine}.

\begin{lstlisting}[language=qcmaquis,caption={Vibronic integral file format.},
				   label=lst:vibronicDump]
EL_ST 0 0
-4114.23                0      0
  325.84799             1      1     
 -325.84799            -1     -1
  559.34550             2      2     
 -559.34550            -2     -2
  674.68277             3      3     
 -674.68277            -3     -3     
  518.21124             4      4     
 -518.21124            -4     -4     
EL_ST 1 1
 4114.23                0      0
  325.84799             1      1     
 -325.84799            -1     -1
  559.34550             2      2     
 -559.34550            -2     -2
  674.68277             3      3     
 -674.68277            -3     -3     
  518.21124             4      4     
 -518.21124            -4     -4     
EL_ST 0 0
 -791.22989             1      0
 -405.69687             2      0
 1171.11703             3      0
EL_ST 1 1
 1092.881251            1      0
-1379.877165            2      0
  302.457910            3      0
EL_ST 0 0
   0.16131088           1      1
   8.71078783           1      2
 -16.45371035           1      3
   8.71078783           2      1
 -65.33090875           2      2
  38.23067993           2      3
 -16.45371035           3      1
  38.23067993           3      2
  -9.35603137           3      3
 -93.47965832           4      4
EL_ST 1 1 
 -73.96104114           1      1
 -24.03532198           1      2
 -15.24387871           1      3
 -24.03532198           2      1
  39.35985614           2      2
   9.27537593           2      3
 -15.24387871           3      1
   9.27537593           3      2
   1.77441974           3      3
 -93.47965832           4      4
EL_ST 0 1
 1677.63321200          4      0
  -80.65544294          4      1
  -44.44114904          4      2
   10.24324125          4      3
   50.65161814          1      4
  -44.44114904          2      4
   10.24324125          3      4
EL_ST 1 0
 1677.63321200          4      0
  -80.65544294          4      1
  -44.44114904          4      2
   10.24324125          4      3
   50.65161814          1      4
  -44.44114904          2      4
   10.24324125          3      4
\end{lstlisting}

The file is divided in different sections by the header \texttt{EL\_ST X Y}, where \texttt{X} and \texttt{Y} are integer smaller than $N_\text{ele}$.
When, while parsing the integral file, a ``\texttt{EL\_ST X Y}'' header is found, then all parameters until the next header are used to define the vibrational Hamitonian for the state \texttt{X} (if \texttt{X}=\texttt{Y}), or for the non-adiabatic coupling between the electronic states \texttt{X} and \texttt{Y} otherwise.
For instance, the header \texttt{EL\_ST 0 0} in Listing~\ref{lst:vibronicDump} indicates that the constants reported in lines 2-10 are associated with the 0-0 block of Eq.~\ref{eq:VibronicHamiltonian}.
They represent, therefore, the vibrational Hamiltonian for the S$_0$ state.
Note that the definition of the vibrational Hamiltonian for a given pair of \texttt{X} and \texttt{Y} values do not have to appear in the same blocks, but can instead be split in different blocks.
For instance, the PES of the \texttt{0} state is defined by the coefficients on lines 2-10, but also on lines 21-24 and on lines 30-39.
Note also that the coefficients associated with the \texttt{0   0} pair of indices (line 3) is the constant energy shift $E_i^{(eq)}$ of Eq.~(\ref{eq:DiagonalVibronic}) (also referred to as $\Delta$ in Ref.~(\citenum{Raab1999_Pyrazine})).
Each block of Listing~\ref{lst:vibronicDump} corresponds to a class of terms entering Eq.~(1) of Ref.~(\citenum{Raab1999_Pyrazine}).
Specifically:

\begin{enumerate}
  \item Lines 3-10 are the $\omega_i$ coefficients for the S$_1$ state.
  \item Lines 13-20 are the $\omega_i$ coefficients for the S$_2$ state.
  \item Lines 22-24 are the $a_i$ coefficients.
  \item Lines 26-28 are the $b_i$ coefficients.
  \item Lines 30-39 are the $a_{ij}$ coefficients.
  \item Lines 41-50 are the $b_{ij}$ coefficients.
  \item Lines 52-58 and 60-66 collect the $c_i$ and $c_{ij}$ coefficients.
\end{enumerate}

Note that the vibronic Hamiltonian is Hermitian and, therefore, the $\mathcal{V}_{ij}$ and $\mathcal{V}_{ji}$ operators form a hermitian conjugate pair.
However, this hermiticity is not exploited in the integral file, where both the $\mathcal{V}_{01}$ and $\mathcal{V}_{10}$ operators are explicitly given.
Finally, as for the excitonic case, also for vibronic Hamiltonians the \texttt{integral\_file} should contain the coefficients expressed in terms of the reduced normal coordinate $\tilde{Q}_i$ and the corresponding conjugate momenta $\tilde{P}_i$.

\subsection{Pratical considerations on vibrational and vibronic DMRG}
\label{sec:Practical-vDMRG}

% Informations on how to compile the vibrational module of QCMaquis
By default, \qcm\ does not compile the module for running vibrational and vibronic DMRG calculations.
In order to enable running vDMRG simulations, the \texttt{BUILD\_VIBRATIONAL} CMake flag should be set to \texttt{ON}.
Moreover, the symmetry group describing the vibrational Hamiltonian in canonical quantization must be included, at compile time, in the \texttt{BUILD\_SYMMETRIES} CMake keyword.
By default, \texttt{BUILD\_SYMMETRIES} includes only \texttt{TwoU1PG} and \texttt{SU2U1PG}.
The \texttt{NONE} point group must be added in order to compile the canonical quantization version of vDMRG.
In fact, the canonical quantization-based vibrational Hamiltonian does not display any symmetry.

Similarly, in order to enable the support for vibronic and excitonic Hamiltonians, two parameters must be changed when the \texttt{Makefile} is generated.
First, the \texttt{BUILD\_VIBRONIC} CMake flag should be set to \texttt{ON}.
Second, the \texttt{U1} group should be added to the list of compiled symmetries, under the \texttt{BUILD\_SYMMETRIES} keyword.

In conclusion, we note that, although in principle one can run time-independent DMRG calculations for optimizing the ground-state of excitonic and vibronic Hamiltonians, these Hamiltonians are mostly applied to study non-equilibrium processes.
They are, therefore, particularly appealing when combined with the DMRG module for quantum-dynamics simulations.

\section{QCMaquis pre-Born--Oppenheimer module}
\label{sec:preBO}

The \qcm\ pre-Born--Oppenheimer (pre-BO) module provides a tool to perform DMRG-CASCI calculations for \textit{multicomponent} Hamiltonians consisting of multiple spin-$\frac12$ particle types.

QCMaquis follows the formatting of the integral file shown in Listing~\ref{lst:prebodump}.
The first line, with only a single float, contains a constant shift in energy, e.g., point-charge repulsion of classical nuclei.
The following lines contain the terms in the Hamiltonian where pairs of numbers $m-n$ describe a single second-quantization operator.
The first number denotes the particle type (starting with 0) and the second number the orbital index associated with the given particle type (also starting with 0 for all types).
We assume physics notation and normal-ordering, i.e., the following would be a valid term

\begin{equation}
  \left(\int \phi_{1-1}(1)\phi_{0-1}(2)g(1,2)\phi_{1-0}(1)\phi_{0-1}(2)\right)
  a^\dagger_{1-1} a^\dagger_{0-1} a_{0-1} a_{1-0}\,.
\end{equation}

The second and third lines contain single-paricle type \textit{hopping} terms.
The fourth line stands for a single-particle type interaction, and the fifth
line for an interaction between different particle types. 
Note here that the aforementioned convention implies that operators belonging to
the same particle types in a given string of operators must be either at
positions 1 and 4 or 2 and 3.

\begin{lstlisting}[language=qcmaquis,caption={Pre-BO integral file
format.},label=lst:prebodump]
1.145901278714102                       
   0-0         0-0                                      -2.451508889071943
   1-0         1-0                                      2.30233235666604
   0-2         0-0         0-1         0-1              -0.08690785254423253
   1-1         0-1         0-1         1-0              4.887021698358933e-05
\end{lstlisting}

Listing~\ref{lst:qcm} shows an input example for a pre-BO DMRG-FCI calculation. 
It assumes that a pre-BO integral file with has been generated. Additionally,
the 1-particle RDM and the mutual information can be measured and printed to a
file.

\begin{lstlisting}[language=qcmaquis,
				   caption={Input example for a pre-BO DMRG-FCI calculation},
				   label=lst:preboinput]
  nsweeps                         = 6
  max_bond_dimension              = 1000

  // Pre-BO specific input
  // String with the number of parciles for each type
  PreBO_ParticleTypeVector        = "4 1"
  // Spin-1/2 fermion (1) or spin-0 boson (0)
  // Attention: the boson-related functionality has not been fully tested, yet.
  PreBO_FermionOrBosonVector      = "1 1"
  // Number of orbitals for each type.
  PreBO_OrbitalVector             = "3 3"
  // Number of alpha-beta particles for each type
  // Note: for spin-0 bosons it would be a single number.
  PreBO_InitialStateVector        = "2 2 1 0"

optimization                    = twosite 
symmetry                        = nu1

model_library                   = coded
lattice_library                 = coded
LATTICE                         = "preBO lattice"

// Your model
MODEL                           = PreBO
init_type                      = 'default'

// Hamiltonian file
integral_file                   = FCIDUMP.E_4_H1_1
integral_cutoff                 = 1.0E-10

MEASURE[1rdm]          = 1
MEASURE[mutinf]        = 1
\end{lstlisting}

\section{Excited state targeting with DMRG}

Being based on the variational principle, the DMRG is inherently tailored to ground-state wave functions.
However, it can be extended to target excited states as well.
To date, \qcm\ supports two excited-state DMRG methods:

\begin{enumerate}
  \item \texttt{DMRG[ortho]}~\cite{Keller_JChemPhys_efficient_2015}, which relies on a constrained optimization search.
  \item \texttt{DMRG[IPI]}~\cite{Baiardi2022_DMRG-FEAST}, which extends the inverse power iteration to MPS wave functions.
  \item \texttt{DMRG[FEAST]}~\cite{Baiardi2022_DMRG-FEAST}, which applied the FEAST algorithm originally introduced by Polizzi\cite{Polizzi2009_FEAST} to MPS wave functions.
\end{enumerate}

\noindent In the following, we will review these three algorithms and describe how to run them with \qcm.

\subsection{\texttt{DMRG[ortho]}}

\noindent Suppose one wants to target the lowest-energy excited state of a given Hamiltonian.
If the exact wave function $\ket{\Psi_\text{GS}}$ is available, then this state is the lowest-eigenvalue of the following projected Hamiltonian $\mathcal{H}_p^{(1)}$:

\begin{equation}
  \mathcal{H}_p^{(1)} = \left( \mathcal{I} - \ket{\Psi_\text{GS}} \bra{\Psi_\text{GS}} \right)
  				  		\mathcal{H}
				  		\left( \mathcal{I} - \ket{\Psi_\text{GS}} \bra{\Psi_\text{GS}} \right) \, .
  \label{eq:ProjectedHamiltonian}
\end{equation}

\noindent The MPS representation of the first excited state $\ket{\Psi_\text{ES}^{(1)}}$ can, therefore, be optimized by applying DMRG to the operator defined in Eq.~(\ref{eq:ProjectedHamiltonian}).
The second excited state can then be optimized by applying DMRG to the following Hamiltonian, where both the ground state and the first excited-state are projected out:

\begin{equation}
  \mathcal{H}_p^{(1)} = \left( \mathcal{I} - \ket{\Psi_\text{GS}} \bra{\Psi_\text{GS}} \right)
						\left( \mathcal{I} - \ket{\Psi_\text{ES}^{(1)}} \bra{\Psi_\text{ES}^{(1)}} \right)
  				  		\mathcal{H}
						\left( \mathcal{I} - \ket{\Psi_\text{ES}^{(1)}} \bra{\Psi_\text{ES}^{(1)}} \right)
				  		\left( \mathcal{I} - \ket{\Psi_\text{GS}} \bra{\Psi_\text{GS}} \right) \, .
  \label{eq:ProjectedHamiltonian2}
\end{equation}

\noindent This procedure can be repeated to target excited states with an increasingly hight energy.
A \texttt{DMRG[ortho]} calculation is activated in \qcm\ by two keywords:

\begin{itemize}
  \item \texttt{n\_ortho\_states}: indicates the number of states for which the projector is constructed, as in Eq.~(\ref{eq:ProjectedHamiltonian2}).
  \item \texttt{ortho\_states}: comma separated list of checkpoint files containing the MPS for constructing the projection operators, as in Eq.~(\ref{eq:ProjectedHamiltonian2}).
\end{itemize}
%
Therefore, for targeting, for example, the two lowest-energy excited states of a given Hamiltonian, three calculations are needed:

\begin{enumerate}
  \item conventional ground-state DMRG optimization, where the \texttt{chkpfile} keyword is set to the name of the checkpoint file that will store the ground-state MPS (let us suppose that this name is \texttt{GS.checkpoint.h5}).
  \item constrained DMRG calculation where \texttt{n\_ortho\_states} is set to 1 and where \texttt{ortho\_states} is set to \texttt{GS.checkpoint.h5}.
  Since we will optimize also the second excited state, we specify again the \texttt{chkpfile} keyword to store the MPS associated with the first excited state (let us suppose that this second file is called \texttt{``ES.checkpoint.h5''}).
  \item second constrained DMRG calculation where \texttt{n\_ortho\_states} is set to 2 and where \texttt{ortho\_states} is set to \texttt{``GS.checkpoint.h5,ES.checkpoint.h5''}
\end{enumerate}

\noindent \texttt{DMRG[ortho]} is very efficient for low-lying excited states, but becomes extremely unpractical for optimizing highly-excited states for two main reasons.
First, targeting a given excited state with \texttt{DMRG[ortho]} requires optimizing all the lower-lying states with DMRG -- a computational task that becomes unfeasible for high-energy states.
Second, due to so-called root-flipping effect, the \texttt{DMRG[ortho]} optimization may be trapped into local minima in high-density regions of the Hamiltonian spectrum.
These two limitations are lifted by \texttt{DMRG[IP]} and \texttt{DMRG[FEAST]}.

\subsection{\texttt{DMRG[IPI]}}
\label{subsec:DMRG-IPI}

The inverse power iteration (IPI) method is an iterative scheme for calculating the lowest-energy eigenstate of the Hamiltonian operator.
IPI relies on the idea that, starting from a random normalized guess wave function $\ket{\Psi_0}$, the series of wave functions $\ket{\Psi_n}$ obtained recursively as:

\begin{equation}
  \ket{\tilde{\Psi}_n} = \mathcal{H}^{-1} \ket{\Psi_{n-1}} \qquad \qquad
  \ket{\Psi_n} = \frac{\ket{\tilde{\Psi}_n}}{\left\| \ket{\tilde{\Psi}_n} \right\|} \, ,
  \label{eq:IPI_Series}
\end{equation}
%
converges to the ground state wave function for $n \rightarrow +\infty$.
Similarly, if we introduce a shift parameter $\omega$, the series obtained as:

\begin{equation}
  \ket{\tilde{\Psi}_n} = \left( \mathcal{H} - \omega \mathcal{I} \right)^{-1} \ket{\Psi_{n-1}} \qquad \qquad
  \ket{\Psi_n} = \frac{\ket{\tilde{\Psi}_n}}{\left\| \ket{\tilde{\Psi}_n} \right\|} \, ,
  \label{eq:IPI_Series_Shifted}
\end{equation}
%
converges to the eigenfunction with the closest energy to $\omega$.
DMRG[IPI] applies IPI to wave functions $\ket{\Psi_n}$ encoded as MPSs.
The key challenge associated to generalizing IPI to DMRG[IPI] is that calculating the $n$-th wave function $\ket{\Psi_n}$ from the $(n-1)$-th one $\ket{\Psi_{n-1}}$ requires solving the following linear system:

\begin{equation}
  \left( \mathcal{H} - \omega \mathcal{I} \right) \ket{\Psi_n} = \ket{\Psi_{n-1}}
  \label{eq:LinearSystem}
\end{equation}

Eq.~(\ref{eq:LinearSystem}) can be solved for MPS wave functions by recasting it as a minimum problem, and by solving it with a sweep-based algorithm.\cite{Rakhuba2016_vDMRG,Baiardi2022_DMRG-FEAST}
For each microiteration of the sweep-based algorithm, a site-centered linear system must be solved, similarly to the sweep-based energy minimization, where a site-centered eigenvalue problem is solved.
The $n \rightarrow +\infty$ is realized, in practice, by monitoring the convergence of the wave function $\ket{\Psi_n}$ and of the corresponding energy $E[\ket{\Psi_n}]$. \\

A DMRG[IPI] calculation is run in \qcm\ with the \texttt{dmrgIPI} executable, located (starting from the build directory) in the \texttt{applications/dmrg} folder.
The relevant parameter (and the corresponding keywords) for the simulation are divided in two classes.
The first ones, associated with keywords with a name that starts with \texttt{linsystem}, are the parameters of the solution of the local site-centered linear systems:

\begin{itemize}
  \item \texttt{linsystem\_precond}: if \texttt{yes}, use a diagonal preconditioner within the solution of the local linear system. If \texttt{no}, does not apply the preconditionditioner.
  \item \texttt{linsystem\_init}: initial guess for the iterative solution of the local linear system.
  Can be either \texttt{mps}, if the initial guess should be constructed from the corresponding MPS tensor at the current iteration, or \texttt{zero}, if an MPS with zero entries should be used as initial guess.
  The default value, which should be used in conventional calculations, is \texttt{mps}.
  \item \texttt{linsystem\_krylov\_dim}: maximum dimension of the Krylov space for the iterative solution of the linear system.
  \item \texttt{linsystem\_max\_it}: maximum number of macroiterations for the iterative solution of the linear system. For each macroiteration, a new Krylov vector space, of maximum size equal to \texttt{linsystem\_krylov\_dim}.
  \item \texttt{linsystem\_solver}: iterative algorithm for solving the linear system (can be either \texttt{GMRES} of \texttt{MINRES}).
  \item \texttt{linsystem\_tol}: convergence threshold on the wave function for the iterative solver of the linear system.
\end{itemize}

The second set of parameters and keywords are instead associated with the IPI algorithm itself, and are:

\begin{itemize}
  \item \texttt{ipi\_sweep\_overlap\_threshold}: the series of Eq.~(\ref{eq:IPI_Series_Shifted}) is stopped as soon as the norm of the overlap between the MPSs $\langle \Psi_n | \Psi_{n-1} \rangle$ falls below this threshold.
  \item \texttt{ipi\_sweep\_energy\_threshold}: the series of Eq.~(\ref{eq:IPI_Series_Shifted}) is stopped as soon as the difference between the energy calculated at two subsequent iterations falls below this threshold.
  \item \texttt{ipi\_sweeps\_per\_system}: maximum number of sweeps to be used to solve a given linear system.
  \item \texttt{ipi\_shift}: $\omega$ parameter, expressed in the same units used for the Hamiltonian.
  \item \texttt{ipi\_iterations}: maximum number of times the linear system of Eq.~(\ref{eq:IPI_Series_Shifted}) is solved (i.e., upper value for the $n$ parameter).
\end{itemize}

\subsection{\texttt{DMRG[FEAST]}}
\label{subsec:FEAST}

DMRG[FEAST] is the third strategy supported by \qcm\ for calculating excited states with DMRG.
DMRG[FEAST]\cite{Baiardi2022_DMRG-FEAST} generalizes the FEAST algorithm\cite{Polizzi2009_FEAST} to wave functions encoded as MPSs.
Instead of targeting a single eigenpair, as in DMRG[IP], FEAST targets all the eigenfunctions $\left\{ \ket{\Psi_1}, \ldots, \ket{\Psi_n} \right\}$ with energy lying in a predetermined interval $\left[ E_\text{min}, E_\text{max} \right]$.
It does so by approximating the spectral projector onto the space spanned by these eigenfunctions, \textit{i.e.}

\begin{equation}
  \mathcal{P} = \sum_{i=1}^n \ket{\Psi_i} \bra{\Psi_i} \, ,
  \label{eq:SpectralProjector}
\end{equation}
%
by applying the Cauchy residual theorem

\begin{equation}
  \mathcal{P}  = \frac{1}{2\pi\mathrm{i}} \oint_{\mathcal{C}} 
	\left( z \mathcal{I} - \mathcal{H} \right)^{-1} \mathrm{d}z \, ,
  \label{eq:ComplexIntegral}
\end{equation}
%
where $\mathcal{C}$ is a circle in the complex plane that encloses the $\left[ E_\text{min}, E_\text{max} \right]$ interval on the real axis. 
Eq.~(\ref{eq:ComplexIntegral}) is then approximated with a the Gauss-Hermite quadrature.
This yields the following form for $\mathcal{P}$:

\begin{equation}
  \mathcal{P} \approx \frac{1}{2\pi\mathrm{i}} \sum_{n_q=1}^{N_\text{quad}}
	w_k \left( z_k \mathcal{I} - \mathcal{H} \right)^{-1} \mathrm{d}z \, ,
  \label{eq:ComplexIntegralApproximate}
\end{equation}
%
where $w_k$ and $z_k$ are the quadrature weights and nodes, respectively.
Applying $\mathcal{P}$ from Eq.~(\ref{eq:ComplexIntegralApproximate}) onto $n$ randomly generated wave functions will, therefore, yield a basis for the subspace spanned by the $n$ eigenfunctions $\left\{ \ket{\Psi_1}, \ldots, \ket{\Psi_n} \right\}$.
It is, therefore, sufficient to diagonalize $\mathcal{H}$ in this subspace to retrieve the eigenfunctions $\left\{ \ket{\Psi_1}, \ldots, \ket{\Psi_n} \right\}$.
Applying the approximate projector of Eq.~(\ref{eq:ComplexIntegralApproximate}) onto a wave function requires solving $N_\text{quad} \times n$ linear system.
In FEAST, these systems are solved with conventional Krylov-based iterative methods, such as the GMRES one.
Instead, DMRG[FEAST] applies the sweep-based algorithm already introduced for DMRG[IPI] to iteratively solve the linear systems in order to ensure that the solution can be represented as an MPS with a given bond dimension $m$. \\

\qcm\ implements the DMRG[FEAST] algorithm as described in Ref.~\citenum{Baiardi2022_DMRG-FEAST}.
A simulation can be run with the \texttt{dmrgFEAST} executable, which is stored -- relatively to the build directory -- in the \texttt{applications/FEAST} directory.
In order to compile this executable, the CMake variable \texttt{BUILD\_DMRG\_FEAST} should be set to \texttt{ON} (by default, its value is set to \texttt{OFF}).
The keywords that can be used to tune a DMRG[FEAST] calculation are the following:

\begin{itemize}
  \item \texttt{feast\_num\_states}: number of initial guesses.
  It must be $\geq$ than the number of states with energy inside the $\left[ E_\text{min}, E_\text{max} \right]$ interval in order to ensure convergence of the FEAST algorithm.
  \item \texttt{feast\_max\_iter}: maximum number of outer FEAST iterations.
  \item \texttt{feast\_emin}: lower bound of the FEAST energy interval.
  \item \texttt{feast\_emax}: upper bound of the FEAST energy interval.
  \item \texttt{feast\_num\_points}: number of quadrature points for approximating the numerical integration (the default value, which is also recommended in Ref.~\citenum{Polizzi2009_FEAST}, is 8).
  %\item \texttt{feast\_integral\_type}: it is possible to show~\cite{Polizzi2009_FEAST} that the complex contour integral can be applied only to half of the circle \texttt{C} by exploiting the Hermiticity of the operator.
  %The \texttt{feast\_integral\_type} parameter should be set to \texttt{half} for applying this simplification, and to \texttt{full} for performing a conventional complex integration.
  \item \texttt{feast\_truncation\_type}: if more than a single FEAST macroiteration is performed, the eigenfunctions of the FEAST Hamiltonian at the $(N-1)$-th iteration are used as a starting guess for the $N$-th iteration.
  However, these eigenfunctions are expressed as a linear combination of $n$ MPSs, each one expressed as a sum of $N_q$ solutions of the linear systems introduced above.
  Since the sum of two MPSs with bond dimension $m$ is an MPSs with bond dimension $2m$, this series of sums would increase drastically the bond dimension of the guesses.
  To avoid that, the guess MPSs are compressed back to a bond dimension $m$ before starting a new FEAST iteration.
  If the \texttt{feast\_truncation\_type} keyword is set to \texttt{end}, this truncation is applied only once, after the $N_q \times n$ sum are performed.
  If, instead, this keyword is set to \texttt{each}, the truncation is performed after each sum.
  This second option is faster, but yields a less accurate initial guess.
  The default value for \texttt{feast\_truncation\_type} is \texttt{end}.
  %\item \texttt{feast\_init\_type}: analog of \texttt{init\_type} for FEAST-based simulation -- this keyword sets the initial guess for DMRG[FEAST].
  %\item \texttt{feast\_init\_onv}: comma-separated list of integers indicating the ONV used as the initial guess (this keyword is valid only if \texttt{feast\_init\_type} is set to \texttt{basis\_state\_generic}).
  \item \texttt{feast\_overlap\_convergence\_threshold}: first threshold used to assess the convergence of DMRG[FEAST].
  The FEAST macroiterations are stopped if the sum of changes in the FEAST eigenfunctions exceeds the value set by this keyword.
  Note that the sum includes only the eigenstates whose energy lies in the interval defined by the keywords \texttt{feast\_emin} and \texttt{feast\_emax}.
  \item \texttt{feast\_energy\_convergence\_threshold}: second threshold used to assess the convergence of DMRG[FEAST].
  The FEAST macroiterations are stopped if the sum of variation of the FEAST energies exceeds the value set by this keyword.
  Note that the sum includes only the eigenstates whose energy lies in the interval defined by the keywords \texttt{feast\_emin} and \texttt{feast\_emax}.
  \item \texttt{feast\_verbose}: by default set to \texttt{no}, can be set to \texttt{yes} to increase the output verbosity for at DMRG[FEAST] calculation.
  \item \texttt{feast\_calculate\_standard\_deviation}: it may happen that the number of guesses exceeds the number of eigenpairs with energy included in the [\texttt{feast\_emin}, \texttt{feast\_emax}] interval.
  In this case, only some of the FEAST eigenfunctions correspond to ``true'' eigenfunctions of the full Hamiltonian.
  These ``true'' states can be identified based on two criteria. 
  Of corse, if the energy of the eigenstate does not lie in the [\texttt{feast\_emin}, \texttt{feast\_emax}] interval, the corresponding state does not correspond to an eigenpair of the Hamiltonian.
  However, if eigenpair has energy in the target interval, it may still not correspond to a ``true'' eigenstate.
  To exclude this situation, the energy standard deviation can be calculated for each FEAST eigenstate by setting the \texttt{feast\_calculate\_standard\_deviation} keyword to \texttt{yes}.
  ``True'' eigenfunction of the full Hamiltonian must correspond, for a sufficiently high bond dimension, to a vanishing variance.
  In practice, the acceptable states are identified based on the \texttt{feast\_standard\_deviation\_threshold} keyword described above. 
  \item \texttt{feast\_standard\_deviation\_threshold}: states with variance lower than the threshold set by this keyword are not considered as ``true'' states.
  Moreover, if \texttt{feast\_max\_iter} > 1, the corresponding eigenfunction is not taken as a guess for the following FEAST iteration.
  %, and a random MPSs is used instead as guess.
\end{itemize}

Note that the keywords starting with ``\texttt{linsystem\_}'' introduced in the previous section can be used also in DMRG[FEAST] to tune the iterative solution of the FEAST linear systems.

\section{\qcm\ examples library}
\label{sec:examples}

We provide in the folder \texttt{example} a set of example input files to run the different kinds of DMRG-based calculations supported by \qcm.
For all input files, we provide the corresponding output file that can be used as a reference to ensure that the \qcm\ executables have been compiled correctly (in addition to the test suite).
The examples are organized in different folder, one per feature provided by \qcm, as listed below:

\begin{itemize}
 \item \texttt{iTD-DMRG}: sample input for an imaginary-time DMRG calculation on H$_2$.
 The corresponding input file name is \texttt{H2\_2e4o.iTD.SS.inp}.
 The integral file is generated for the H$_2$ molecule and based on the 6-31G basis set.
 The calculation should be run with the executable \texttt{applications/evolve/dmrg\_evolve}, which is generated by setting the \texttt{BUILD\_DMRG\_EVOLVE} CMake flag to \texttt{ON}.
 We also provide in the same folder also \texttt{H2\_2e4o.TI.SS.inp}, \textit{i.e.} the input for running a conventional \texttt{DMRG} calculation, which must be run with the \texttt{applications/dmrg/dmrg} executable and returns the same energy as the iTD-DMRG simulation.
 \item \texttt{TD-DMRG}: example calculation of a real-time DMRG calculation.
 The integral file is associated with the $\pi$ orbital space (CAS(6,6)) of benzene calculated with the 6-31G* basis set.
 The calculations should be run with the \texttt{applications/evolve/dmrg\_evolve} executable.
 Note that the initial state for the propagation is a random MPS because the \texttt{init\_state} keyword is set to \texttt{default}.
 Therefore, the calculation does not correspond to any physically meaningful non-equilibrium process. 
 However, this calculation can be used to assess that the \texttt{dmrg\_evolve} executable has been compiled correctly by verifying that the energy is conserved along the propagation.
 For this example calculation we provide the input files for both the single-site (\texttt{Benzene\_6e6o.TD.SS.inp}) and the two-site (\texttt{Benzene\_6e6o.TD.TS.inp}) propagators.
 \item \texttt{vDMRG}: 
 %example of a $n$-mode vDMRG calculation.
 %The corresponding input name is \texttt{FAD\_2Mode\_TS.inp}.
 %The integral file is generated for a two-mode potential describing the formic acid dimer based on a discrete %variable representation primitive basis set.
 %For each mode, 10 modal basis functions are included. \\
 %The same folder also contains an
 Example calculation for the canonical quantization-based vDMRG, based on the single-site and on the two-site (the corresponding input files are named \texttt{C2H4.vDMRG.m50.SS.input} and \texttt{C2H4.vDMRG.m50.TS.input}, respectively) optimization algorithm.
 The PES is taken from Ref.~\citenum{Berkelback2021_vHBCI} and represents the potential energy surface of ethylene as a sixth-order force field.
 The calculation is initiated from the ONV associated with the ground-state wave function.
 The calculations should be run with the \texttt{applications/dmrg/dmrg} executable, which is automatically generated by each \qcm\ compilation configuration.
 \item \texttt{vDMRG\_IPI}: example of a vDMRG[IPI] calculation.
 %based on the canonical quantization vDMRG variant.
 The test system is ethylene, with the same PES described for the folder \texttt{vDMRG}.
 The calculation uses the single-site linear solver, and runs 5 iterations of the inverse power iteration algorithm.
 The simulation should be run with the \texttt{applications/dmrg/dmrgIPI} executable, which is generated automatically with the standard compilation parameters of \qcm.
 \item \texttt{vDMRG-FEAST}: example of a vDMRG[FEAST] calculation to optimize the vibrational ground state of ethylene.
 The PES is the same as for the folder \texttt{vDMRG}.
 The calculation uses a single-site linear solver, and repeat multiple FEAST iterations, although the energy is converged already with the first one.
 The simulation can be run with the \texttt{applications/FEAST/dmrgFEAST} executable, which is generated by setting the \texttt{BUILD\_DMRG\_FEAST} CMake varialbe to \texttt{ON}.                
\end{itemize}

\clearpage
\newpage

\bibliography{qcmaquis_manual}

\end{document}

### 2023-05-08 dmrg/doc/qcmaquis_manual.tex
Version 3.1.3

@@
 \usepackage{booktabs}
+\usepackage{braket}
 \usepackage[bf,singlelinecheck=off]{caption}
@@
 %
-    keywords=[1]{L,nelec,spin,irrep,nsweeps,max_bond_dimension,integral_file,chkpfile,resultfile,init_state,LATTICE,lattice_library,MODEL,model_library},
+    keywords=[1]{L,nelec,spin,irrep,nsweeps,max_bond_dimension,integral_file,chkpfile,resultfile,init_type,LATTICE,lattice_library,MODEL,model_library},
 	morekeywords=[2]{conv_thresh,integral_cutoff,truncation_initial,truncation_final,symmetry,MEASURE},
@@
 
-\section{\qcm{} Standalone Version}
-
-As already anticipated, \qcm\ can be used also as a standalone code.
-The standalone version of \qcm\ is used especially in two cases: (i) for electronic-structure calculations relying on integrals that are calculated with programs other than \mol{} and (ii) for DMRG calculations beyond time-independent electronic-structure simulations.
-In the present section, we focus on the former application field, while the latter will be discussed in the following Sections. \\
-Section~\ref{sec:install-qcm-stand}\ describes in details the installation process for a standalone version of \qcm.
-
-\begin{framed}
- \noindent Note: If you have installed \qcm{} with the \qcm{}-\mol{} interface, you may skip this Section.
-\end{framed}
-
-\subsection{Installation}
-\label{sec:install-qcm-stand}
-
-\noindent We describe in the following how to build and install the standalone version of the \qcm\ software.
-The installation of \qcm\ has been tested for different operating systems and compiler/math libraries environments.
-Their list can be found in Section \ref{subsec:supportedOS}.
-While other combinations might work equally well they are \emph{not} officially supported.
-
-The following libraries are required and need to be installed manually by the user (but are usually provided by the system package manager)
-
-\begin{itemize}
- \item GSL
- \item HDF5
- \item SZIP
- \item Boost
-\end{itemize}
-
-Additionally, \qcm{} distribution includes a customised version of the ALPS 2.3.0 library, which will be compiled automatically.
-
-\subsubsection{Download \qcm}
-\label{subsubsec:download-stand}
-
-As mentioned in Section~\ref{sec:reg}, the \qcm{} standalone version may be found in \texttt{\url{https://github.com/qcscine/qcmaquis}}.
-
-\subsubsection{Setting up a build folder}
-
-Clone the GitHub repository into a new directory \texttt{\qcmsrc}.
-Create a build folder \texttt{\qcmbuild} -- note that this folder \emph{does not} necessarily have to be a subfolder of \qcmsrc\ -- and change to this new folder:
-
-\begin{verbatim}
-  mkdir /path/to/my-QCMaquis-build && cd /path/to/my-QCMaquis-build
-\end{verbatim}
-
-\subsubsection{Configuration}
-\label{subsubsec:configure-stand}
-
-\qcm{} supports operating systems, C++ compilers and math libraries as listed in Section~\ref{subsec:supportedOS}. 
-Below we will show the configuration steps for the most popular compiler suites GNU (Section \ref{sec:gnu-conf-stand}) and Intel (Section \ref{sec:intel-conf-stand}), respectively.
-How to setup and use a shared-memory OMP installation of \qcm\ is described in Section \ref{sec:parallel-stand}.
-
-\paragraph{Configuration with the GNU compiler suite}
-\label{sec:gnu-conf-stand}$\;$\\
-
-\noindent To configure \qcm\ with the GNU compiler suite type
-
-\begin{verbatim}
-  CC=gcc CXX=g++ cmake /path/to/my-QCMaquis-src
-\end{verbatim}
-%
-where we assumed that the Intel Math Kernel Library (MKL) is available (recommended option).
-If the MKL libraries are not available, \qcm\ will search for other suitable math libraries installed on the operating system.
-If none are found the configuration step will stop with an appropriate message.
-
-% \paragraph{Configuration with the Intel compiler suite}\label{sec:intel-conf-stand}$\;$\\
-%
-% \noindent To configure \qcm\ with the Intel compiler suite including MKL type
-% \begin{verbatim}
-% FC=ifort CC=icc CXX=icpc cmake -DQCM_standalone=ON /path/to/my-QCMaquis-src
-% \end{verbatim}
-
-\begin{framed}
- \noindent Note: Intel compiler is currently not supported. Running a binary compiled with Intel compilers may result in segmentation faults. We are currently investigating the issue.
-\end{framed}
-
-\paragraph{Shared-memory OMP parallel configuration}
-\label{sec:parallel-stand}$\;$\\
-
-By default \qcm\ is built as shared-memory OMP parallelized version which should work with either compiler suite, GNU or Intel.
-
-\vspace{2ex}
-
-In order to exploit the shared-memory OMP parallelization of \qcm\ the user is strongly encouraged to set at runtime the environment variable
-
-\begin{verbatim}
-  export OMP_NUM_THREADS=XX
-\end{verbatim}
-%
-where \texttt{XX}\ specifies the number of shared-memory cores to be used.
-The default (\emph{depending on the operating system!!!}) is to use all available cores.
-
-\subsubsection{Building and installation}
-\label{subsubsec:build-stand}
-
-After a successful configuration, type
-\begin{verbatim}
-  make
-\end{verbatim}
-%
-or
-
-\begin{verbatim}
-  make -j8
-\end{verbatim}
-%
-to compile \qcm\ (in parallel on 8 cores) and install all its components in the build folder \texttt{\qcmbuild}.
-
-\subsection{Keywords and Options}\label{sec:qcmaquis-kw}
-
-In the following we describe (i) compulsory keywords (Section \ref{sec:compul-qcm}), (ii) optional keywords (Section \ref{sec:optional-qcm}) for \qcm\ DMRG calculations as well as (iii) keywords for property calculations (Section \ref{sec:prop-qcm}).
-Most of the \qcm\ keywords have default settings that guarantee convergence in the general case and are inserted automatically
-by the host program (\hostp\ in this case). 
-A reasonable choice of default values for optional keywords is given in our
-example \qcm\ input file in Section~\ref{sec:input-qcm}.
-
-\begin{framed}
-  \noindent\textbf{Note:}\ \textsc{Maquis} has many features beyond quantum chemistry, e.g. related to solid state physics.
-  Some keywords listed in the example file in Section \ref{sec:input-qcm} (e.\,g.\ \kwd{LATTICE}, \kwd{lattice\_library} etc.)\ are therefore explained only briefly in the following and are not to be changed from the default values (see Listing~\ref{lst:qcm}), if a quantum chemical calculation is desired.
-\end{framed}
-
-\subsubsection{Compulsory keywords}
-\label{sec:compul-qcm}
-
-The keywords in Tables \ref{tab:standard_kw} and \ref{tab:standard_kw_auto} have to be set for every DMRG calculation since they may crucially affect the accuracy of the final result.
-Their choice depends for example on the molecule under consideration (i.\,e.
-whether you expect strong static electron correlation and/or dynamical correlation to play a major role), the nature of the reference orbitals (Hartree-Fock orbitals, natural orbitals of some kind, \ldots), the active space size, and many other aspects.
-
-% \noindent Some of the compulsory keywords listed in Table \ref{standard_kw}\ are indeed automatically set if you run a \qcm\ DMRG calculation through the \qcm\ host program driver which is the case for \mol-\qcm\ DMRG
-% calculations. In this case skip the upper part of Table \ref{standard_kw}\ and proceed immediately to the lower part
-% marked by ``{\color{red}{Keywords \emph{NOT} set by the host program \mol}}".
-
-\begin{table}
-  \caption{\label{tab:standard_kw_auto}Compulsory \qcm{} keywords automatically set by the \qcm{}-\mol{} interface.}
-  \begin{tabular}{p{3.6cm}l}
-  \toprule
-keyword & description \\
-  \midrule
-
-\texttt{nelec} & Total number of electrons. \\[\tableoptionskip]
-\multirow{3}{*}{\texttt{irrep}} & \multirow{3}{12cm}{Irreducible representation of the point group symmetry of the target state.
-\textbf{Note:} Counting starts with 0 which has to be the totally symmetric representation.} \\
-& \\
-& \\[\tableoptionskip]
-\multirow{2}{*}{\texttt{spin}} & \multirow{2}{12cm}{Total spin ($2\times S$) of the target state, for example:\\
-\texttt{spin=0} (singlet), \texttt{spin=1} (doublet), \texttt{spin=2} (triplet), \ldots}\\
-& \\[\tableoptionskip]
-\texttt{L} & Length of lattice = number of orbitals in the active space. \\[\tableoptionskip]
-\multirow{2}{*}{\texttt{integral\_file}} & \multirow{2}{12cm}{Path and filename of the integral file, for example\\
-\texttt{integral\_file = /path/to/file/FCIDUMP}} \\
-& \\[\tableoptionskip]
-\texttt{chkpfile} & Path and name of the folder in which the MPS is stored. \\[\tableoptionskip]
-\texttt{resultfile} & Path and filename of the result file. \\[\tableoptionskip]
-\multirow{3}{*}{\texttt{n\_ortho\_states}} &\multirow{3}{12cm}{If an excited state calculation is desired, the number of states the current wave function is to be orthogonalized against shall be specified here.} \\
-& \\
-& \\[\tableoptionskip]
-\multirow{3}{*}{\texttt{ortho\_states}} & \multirow{3}{12cm}{Path(s) and filename(s) of the MPS checkpoint file(s) that store the lower lying states to which the current MPS shall be orthogonal to.} \\
-& \\
-& \\[\tableoptionskip]
-\multirow{9}{*}{\texttt{init\_state}} & \multirow{9}{12cm}{Possible options are \texttt{default}, \texttt{thin} and \texttt{hf}.
-The \texttt{default} and \texttt{thin} initializations fill the initial MPS with random numbers, the difference being
-that a singular value decomposition reduces the bond dimension to \texttt{init\_bond\_dimension} in the case of
-\texttt{thin}. Usage of \texttt{hf} generates an MPS consisting of only the determinant defined on the \texttt{hf\_occ}
-card. Note that the CI-DEAS procedure \cite{Legeza2003,fiedler,interface} (as invoked by \texttt{dmrginit.py}, see Section
-\ref{sec:pytools-qcm}) behaves like a restart from the newly generated CI-DEAS MPS.\\
-% {\textbf{default in \mol: \texttt{init\_state = "default"}}}
-}\\
-& \\
-& \\
-& \\
-& \\
-& \\
-& \\
-& \\
-& \\[\tableoptionskip]
-\multirow{7}{*}{\begin{minipage}{5cm}
-                  \texttt{%
-                  LATTICE\\
-                  lattice\_library\\
-                  MODEL\\
-                  model\_library\\}
-                \end{minipage}}
- & \multirow{3}{11cm}{These options set the Hamiltonian and the lattice type in \qcm{} calculations. For a quantum chemical calculation, these must be set to (see also Listing~\ref{lst:qcm})}\\
- & \\
- & \\
-  & \texttt{LATTICE = "orbitals"}\\
-  & \texttt{lattice\_library = "coded"}\\
-  & \texttt{MODEL = "quantum\_chemistry"}\\
-  & \texttt{model\_library = "coded"}\\
-\bottomrule
-\end{tabular}
-\end{table}
-\FloatBarrier
-
-\begin{framed}
-  \noindent\textbf{Note:}\ We strongly encourage any new user of \qcm\ to carefully read first Ref.\ \citenum{Keller_CHIMIA_Determining_2014}\ which gives a detailed introduction to
-  DMRG calculations in quantum chemistry.
-Further quantum chemical DMRG sample calculations starting from different computational setups are discussed for example in Refs.
-\cite{interface,ru-dmrg,fde-dmrg,srdft-dmrg}.
-\end{framed}
-
-\subsubsection{Optional keywords}
-\label{sec:optional-qcm}
-
-The keywords summarized in Table \ref{tab:advanced_kw} may be exploited by the more experienced user but can be safely
-ignored by those who just want to get started. They may affect the convergence and accuracy of the final result, though.
-For the inexperienced user however, we advise to not change these settings and accept the default values provided by \qcm.
-
-% \vspace{3ex}
-
-\begin{longtable}{ll}
-\caption{Optional keywords for \qcm\ calculations.}\label{tab:advanced_kw}
-\\
-  \toprule
-keyword & description \\
-  \midrule
-\endfirsthead
-
-\multicolumn{2}{c}%
-{{\bfseries \tablename\ \thetable{} -- continued from previous page}} \\
-\toprule
-keyword & description \\
-\midrule
-\endhead
-
-\multirow{6}{*}{\texttt{orbital\_order}} & \multirow{6}{12cm}{Manual ordering of the orbitals along the one dimensional lattice.
-The order has to be entered as a string of comma separated orbital numbers.
-We recommend the Fiedler ordering based on the mutual information \cite{interface,fiedler} which can be obtained by means of the python script \texttt{fiedler.py} (see Section \ref{sec:pytools-qcm}).\\
-	\textbf{default}: \texttt{orbital\_order = "1,2,3,4,5,6,\ldots"}} \\
-& \\
-& \\
-& \\
-& \\
-& \\[\tableoptionskip]
-\multirow{4}{*}{\texttt{sweep\_bond\_dimensions}} & \multirow{4}{12cm}{Sets \texttt{max\_bond\_dimension}\ for each sweep separately.\\
-	\emph{Note}: Replaces \texttt{max\_bond\_dimension} which does \emph{NOT} need to be specified in this case.\\
-	Example (\texttt{nsweeps=3}): \texttt{sweep\_bond\_dimensions}=``300,400,500"} \\
-& \\
-& \\
-& \\[\tableoptionskip]
-\multirow{2}{*}{\texttt{init\_bond\_dimension}} & \multirow{2}{12cm}{Adjusts the maximal bond dimension of the MPS produced by the
-CI-DEAS procedure \cite{Legeza2003,fiedler,interface}.} \\
-& \\[\tableoptionskip]
-\multirow{7}{*}{\texttt{conv\_thresh}} &\multirow{7}{12cm}{Sets the energy convergence threshold (in Hartree).
-If the lowest energy from the previous sweep differs from the lowest energy of the current sweep by less than \texttt{conv\_thresh}, the DMRG calculation stops.\\
-		\emph{Note}: Requires to set also \texttt{truncation\_final} and \texttt{ietl\_jcd\_tol}.\\
-	{Numerical format}: \texttt{xe-y} with \texttt{x}\ and \texttt{y} being integers. \\
-	Example: \texttt{conv\_thresh = 1e-6}.}\\
-& \\
-& \\
-& \\
-& \\
-& \\
-& \\[\tableoptionskip]
-\multirow{3}{*}{\texttt{ietl\_jcd\_tol}} & \multirow{3}{12cm}{Convergence threshold for the Jacobi-Davidson diagonalization.\\
-	{Numerical format}: \texttt{xe-y} with \texttt{x}\ and \texttt{y} being integers. \\
-	Example: \texttt{ietl\_jcd\_tol = 1e-6}.} \\
-& \\
-& \\[\tableoptionskip]
-\multirow{2}{*}{\texttt{ietl\_jcd\_maxiter}} & \multirow{2}{12cm}{Maximum number of iterations in the Jacobi-Davidson diagonalization.} \\
-& \\ \pagebreak
-\multirow{6}{*}{\texttt{truncation\_initial}} & \multirow{6}{12cm}{If during the \texttt{ngrowsweeps}, the sum of the discarded singular values for $m$ retained states is lower than the value defined here, more block states will be discarded until the discarded sum increases to \texttt{truncation\_initial}.\\
-		{Numerical format}: \texttt{xe-y} with \texttt{x}\ and \texttt{y} being integers. \\
-		Example: \texttt{truncation\_initial = 1e-6}.}\\
-& \\
-& \\
-& \\
-& \\
-& \\[\tableoptionskip]
-\multirow{6}{*}{\texttt{truncation\_final}} & \multirow{6}{12cm}{If during the \texttt{nmainsweeps}, the sum of the discarded singular values for $m$ retained states is lower than the value defined here, more block states will be discarded until the discarded sum increases to \texttt{truncation\_final}.\\
-	{Numerical format}: \texttt{xe-y} with \texttt{x}\ and \texttt{y} being integers. \\
-	Example: \texttt{truncation\_final = 1e-6}.}\\
-& \\
-& \\
-& \\
-& \\
-& \\[\tableoptionskip]
-\multirow{2}{*}{\texttt{measure\_each}} &\multirow{2}{12cm}{Tells the program to compute the expectation values every\\$2\ \times$
-	\texttt{measure\_each} sweeps.} \\
-& \\[\tableoptionskip]
-\multirow{10}{*}{\texttt{symmetry}} & \multirow{10}{11cm}{Defines the total symmetry group. Default is the combined spin- (SU2) and point
-symmetry group (PG) \texttt{su2u1pg}, where the \texttt{pg}-suffix should be omitted for better performance if the molecule is C1-symmetric.
-For test purposes, it is possible to switch off spin-adaptation, again with or without point group symmetry: \texttt{2u1(pg)} .
-In the latter case the keywords \texttt{spin} and \texttt{nelec} (see Section \ref{sec:compul-qcm}) have no meaning.
-Instead \texttt{u1\_total\_charge1} and \texttt{u1\_total\_charge2} corresponding to the number of up and down electrons have to be specified.} \\
-& \\
-& \\
-& \\
-& \\
-& \\
-& \\
-& \\
-& \\
-& \\[\tableoptionskip]
-\multirow{2}{*}{\texttt{chkp\_each}} & \multirow{2}{11cm}{Tells the program to update the checkpoint file every\\ $2\ \times$
-\texttt{chkp\_each} sweeps.} \\
-& \\[\tableoptionskip]
-\multirow{9}{*}{\texttt{hf\_occ}} & \multirow{9}{11cm}{Occupation of the starting orbitals (e.g. Hartree-Fock occupation) to be
-entered as a comma separated string of occupation aliases. The aliases are defined as follows: 4~=~full, 3~=~up,
-2~=~down, 1~=~empty. This information has to be entered in case of \texttt{hf} as \texttt{init\_state} and for the
-CI-DEAS procedure \cite{Legeza2003,fiedler,interface} as invoked by \texttt{dmrginit.py}.\\
-Example: \texttt{hf\_occ = "4,4,4,2,2,1"}\ for a CAS(8,6) triplet state setup where the unpaired electrons are ``located"
-in orbital \# 4 and \# 5.}\\
-& \\
-& \\
-& \\
-& \\
-& \\
-& \\
-& \\
-& \\[\tableoptionskip]
-\multirow{2}{*}{\texttt{nmainsweeps}} & \multirow{2}{11cm}{Number of sweeps in which \texttt{truncation\_final} is used in the singular value decomposition.} \\
-& \\[\tableoptionskip]
-\multirow{2}{*}{\texttt{ngrowsweeps}} &\multirow{2}{11cm}{Number of sweeps in which \texttt{truncation\_initial} is used in the singular value decomposition.}\\
-& \\[\tableoptionskip]
-\texttt{storagedir} &Scratch directory for temporary files.\\
-\bottomrule
-\end{longtable}
-
-\subsubsection{Keywords for expectation value calculations}
-\label{sec:prop-qcm}
-
-\qcm\ can (in principle) compute expectation values for any one- or two-particle operator that can be formulated in
-second quantization. Table \ref{tab:expval}\ comprises a list of the available property keywords available in \qcm.
-% For further updates/other properties please contact \href{mailto:dmrg@phys.chem.ethz.ch}{dmrg@phys.chem.ethz.ch}.
-
-The one-particle reduced density matrix as well as the one-particle spin-density matrix are implicitly computed from the
-expectation values of some of the operators contained in \texttt{MEASURE[ChemEntropy]}.
-
-If a property is available only in 2U1 symmetry, the SU2U1 checkpoint may be transformed into 2U1 group with the \texttt{mps\_transform(\_pg)} tool (see Section~\ref{sec:tools-qcm})
-
-\begin{longtable}{ll}
- \caption{\label{tab:expval}Expectation value calculations available in the release version of \qcm.}
-\\
-  \toprule
-keyword & description \\
-  \midrule
-\endfirsthead
-
-\multicolumn{2}{c}%
-{{\bfseries \tablename\ \thetable{} -- continued from previous page}} \\
-\toprule
-keyword & description \\
-\midrule
-\endhead
-% \begin{tabular}{p{5.6cm}l}
-
-\multirow{3}{*}{\texttt{MEASURE[ChemEntropy]}} & \multirow{3}{11cm}{All expectation values over the operators required to
-calculate the mutual information (as specified in Ref.~\cite{entanglement}) will be computed. Available only for SU2 symmetry.} \\
- & \\
- & \\[\tableoptionskip]
-\multirow{3}{*}{\texttt{MEASURE[1rdm]}} & \multirow{3}{11cm}{Computes the one-particle reduced density matrix, without the additional correlators
-contained in the \texttt{ChemEntropy} measurement.}\\
- & \\
- & \\[\tableoptionskip]
-\multirow{10}{*}{\begin{minipage}{5cm}
-                 \texttt{MEASURE[2rdm]}\\
-                 \texttt{MEASURE[3rdm]}\\
-                 \texttt{MEASURE[4rdm]}\\
-                \end{minipage}
-                }& \multirow{10}{11cm}{Computes the two-particle, three-particle and four-particle reduced density matrix, respectively.
-In case of 4-RDM, \texttt{MEASURE[4rdm]="p4:p3:p1:p2@\textit{l,k,i,j}"} allows to compute only a 4-index sub-block of the 4-RDM with the indices \texttt{\textit{i,j,k,l}}.
-This functionality is used in the distributed RDM calculations described in Section~\ref{sec:4rdm-distributed}.
-Analogously, for 3-RDM a two-index subblock may be calculated with \texttt{MEASURE[3rdm]="p1:p2@\textit{j,i}"}.\\ 3-RDM and 4-RDM calculations are available only in 2U1(pg) symmetry.}\\
-& \\
-& \\
-&\\
-&\\
-&\\
-&\\
-&\\
-&\\
-&\\[\tableoptionskip]
-\multirow{6}{*}{\begin{minipage}{5.6cm}
-%                  \texttt{MEASURE[trans1rdm]="\textit{chk}"}\\ %% trans1rdm_aa, _bb, _ab, _ba ??
-                 \texttt{MEASURE[trans2rdm]="\textit{chk}"}\\
-                 \texttt{MEASURE[trans3rdm]="\textit{chk}"}\\
-                \end{minipage}} & \multirow{6}{11cm}{Computes the transition two- or three-particle reduced density matrix with the current state and the state specified in the \texttt{\textit{chk}} checkpoint file.
-For t-3RDM calculations, two-index subblocks can be calculated with \texttt{MEASURE[trans3rdm]="\textit{chk};p1:p2@\textit{j,i}"}.
-Available only for 2U1(pg) symmetry.}\\
- & \\
- & \\
- & \\
- & \\
- & \\ \pagebreak
-\multirow{4}{*}{\texttt{MEASURE\_LOCAL[\textit{name}]="\textit{op}"}} &
-            \multirow{4}{11cm}{Computes $\langle \psi | op_i | \psi \rangle, \, i = 1 \ldots L$. \texttt{Nup}, \texttt{Ndown} and \texttt{Nup*Ndown}
-            are meaningful choices for \textit{op}. Available for \texttt{2u1(pg) only.}\\
-            \emph{Note}: \texttt{\textit{name}}\ defines the name under which the expectation values are stored on the \texttt{resultfile}.}\\
- & \\
- & \\
- & \\[\tableoptionskip]
-\multirow{8}{5.6cm}{\texttt{MEASURE\_HALF}\texttt{\_CORRELATIONS} \texttt{[\textit{name}]="\textit{op$_1$:op$_2$}"}} &
-            \multirow{8}{11cm}{Computes $\langle \psi | op_{1i} op_{2j}| \psi \rangle, \, i = 1 \ldots L, j=i+1 \ldots L$.
-            \texttt{Nup}, \texttt{Ndown}, \texttt{Nup*Ndown}, \texttt{cdag\_up}, \texttt{cdag\_down}, \texttt{c\_up}, \texttt{c\_down},
-            \texttt{cdag\_up*Ndown},
-            \texttt{c\_up*Ndown},
-            \texttt{cdag\_down*Nup},
-            \texttt{c\_down*Nup},
-            \texttt{cdag\_up*cdag\_down},
-            \texttt{c\_up*c\_down},
-            \texttt{cdag\_up*c\_down},
-            and \texttt{cdag\_down*c\_up},
-            as \textit{op$_1$} and \textit{op$_2$} are recognized by the program. Available for \texttt{2u1(pg)} only.\\
-            \emph{Note}: \texttt{\textit{name}}\ defines the name under which the expectation values are stored on the \texttt{resultfile}.}\\
-%\multirow{2}{*}{= "\textit{op$_1$:op$_2$}"}} & \\
- & \\
- & \\
- & \\
- & \\
- & \\
- & \\
- & \\
-\bottomrule
-\end{longtable}
-\FloatBarrier
-
-\subsection{QCMaquis input example}
-\label{sec:input-qcm}
-
-Listing~\ref{lst:qcm} provides an input example for a DMRG-CASCI calculation.
-It assumes that an integral file \texttt{FCIDUMP} has been generated and placed in the same directory with the input file.
-The mandatory keywords are indicated in red, and the optional keywords in blue.
-The \texttt{FCIDUMP} for this particular example has been generated by the \mol-\qcm{} interface with the \texttt{001.input} input file from the \mol-\qcm{} test directory.
-
-\begin{lstlisting}[language=qcmaquis,caption={Input example for a DMRG-CASCI(8,8) standalone \qcm{} calculation},label=lst:qcm]
-  // Lattice size (= number of orbitals)
-  L     = 8
-  // Number of electrons
-  nelec = 8
-  // Target symmetry (spin and spatial) of the wave function:
-  // 2*spin = 0 (singlet)
-  spin  = 0
-  // Wavefunction irreducible representation (0 == totally symmetric)
-  irrep = 0
-  // Parameters to control the actual DMRG calculation
-  nsweeps              = 8
-  max_bond_dimension   = 256
-  conv_thresh          = 1e-6
-  // MPS initialization
-  init_state = 'default'
-  // Wave function symmetry (su2u1pg for spin-adapted DMRG + point group)
-  symmetry             = 'su2u1pg'
-  // Threshold below which integrals inside the integral_file are neglected
-  integral_cutoff      = 1e-40
-  // Threshold for the SVD truncation for the initial sweeps
-  truncation_initial   = 1e-16
-  // Threshold for the SVD truncation for the final sweeps
-  truncation_final     = 1e-7
-  // HDF5 file where the MPS will be stored
-  chkpfile             = 'chkp.h5'
-  // HDF5 file where the results of the DMRG calculations will be stored
-  resultfile           = 'result.h5'
-  // File with the one- and two-body integrals
-  integral_file        = "FCIDUMP"
-  // Path to the scratch folder
-  storagedir           = '/scratch/$USER/boundaries'
-  // Lattice-related parameters (not to be changed for qc calculations)
-  LATTICE              = "orbitals"
-  lattice_library      = "coded"
-  MODEL                = "quantum_chemistry"
-  model_library        = "coded"
-  // List of measurements to be performed at the end of the optimization
-  MEASURE[ChemEntropy] = 1
-\end{lstlisting}
-
-\section{QCMaquis vibrational DMRG module}
-\label{sec:vDMRG}
-
-The vibrational DMRG (vDMRG) module enables calculating ground- and excited-state energies of the Born--Oppenheimer vibrational Hamiltonian:
-
-\begin{equation}
-  \underbrace{\left[ \mathcal{T}(Q_1, \ldots, Q_L) + \mathcal{V}(Q_1, \ldots, Q_L) \right]}_{\mathcal{H}_\text{vib}(Q_1, \ldots, Q_L)} \Psi_n(Q_1, \ldots, Q_L) = E_n \Psi_n(Q_1, \ldots, Q_L)
-  \label{eq:VibHam}
-\end{equation}
-
-We designed two vDMRG formulations, one based on the so-called canonical quantization format,\cite{Hirata2014_NormalOrdered} and one based on the $n$-mode second quantization.\cite{Christiansen2004_nModeQuantization}
-These two formulations differ in the way in which the Hamiltonian $\mathcal{H}_\text{vib}$ is encoded based on the second-quantization formalism.
-In the following, we will describe how this encoding is obtained.
-
-\subsection{Canonical quantization vDMRG}
-\label{sec:canonical}
-
-The canonical quantization relies on the harmonic oscillator theory.\cite{Hirata2014_NormalOrdered}
-The vibrational wave function is expanded, in practice, as follows:
-
-\begin{equation}
-  \Psi(Q_1, \ldots, Q_L) = \sum_{n_1}^{N_\text{max}} \cdots \sum_{n_L=1}^{N_\text{max}}
-	C_{n_1 \cdots n_L} | n_1 \cdots n_L \rangle \, .
-  \label{eq:CanonicalQuantizationCI}
-\end{equation}
-
-$| n_1 \cdots n_L \rangle$ are occupation number vectors that are constructed using the harmonic oscillator eigenfunction as the one-particle basis set, \textit{i.e.}
-
-\begin{equation}
-  | n_1 \cdots n_L \rangle \equiv \prod_{i=1}^L \chi_{n_i}(Q_i)
-  \label{eq:HarmonicOscillatorBasis}
-\end{equation}
-%
-$\chi_{n_i}(Q_i)$ being the $n_i$-th eigenfunction of the monodimensional harmonic oscillator Hamiltonian associated with the $i$-th mode.
-In order to encode the Hamiltonian in a format analog to Eq.~\ref{eq:CanonicalQuantizationCI}, we express it as a power series in terms of the vibrational coordinates $Q_i$.
-By including up to fourth-order force constants, the expansion reads:
-
-\begin{equation}
-  \mathcal{V}(Q_1, \ldots, Q_L) = \sum_{i,j=1}^L k_{ij} Q_i Q_j + \sum_{i,j,k=1}^L k_{ijk} Q_i Q_j Q_k
-    + \sum_{i,j,k,l=1}^L k_{ijkl} Q_i Q_j Q_k Q_l
-  \label{eq:TaylorExpansion}
-\end{equation}
-%
-where we neglected the first-order force constants since we assume that the expansion is performed around a stationary point of the PES.
-The position operators $Q_i$ are, then, encoded in second quantization based on the well-known canonical quantization rules.
-
-\begin{equation}
-  Q_i = \frac{1}{2} \left( b_i^\dagger + b_i \right) \, .
-  \label{eq:CanonicalQuantizationPosition}
-\end{equation}
-
-Note that Eq.~\ref{eq:CanonicalQuantizationPosition} holds only if $Q_i$ are the dimensionless Cartesian-based normal coordinates.
-Moreover, the kinetic energy operator can be expressed in second quantization based on the counterpart of Eq.~(\ref{eq:CanonicalQuantizationPosition}) for the conjugate momentum operator
-
-\begin{equation}
-  P_i = \frac{\mathrm{i}}{2} \left( b_i^\dagger - b_i \right) \, .
-  \label{eq:CanonicalQuantizationMomentum}
-\end{equation}
-
-The resulting form of the vibrational Hamiltonian is then converted in the MPO format by \qcm\ based on the algorithm described in Ref.~\citenum{Baiardi2017_vDMRG}. \\
-
-\begin{lstlisting}[language=qcmaquis,
-				   caption={Input example for a canonicalq quantization vDMRG-FCI calculation},
-				   label=lst:canonicalVibInput]
-  // == Generic input section ==
-  nsweeps                         = 6
-  max_bond_dimension              = 100
-  model_library                   = coded
-  lattice_library                 = coded
-  optimization                    = twosite 
-  integral_file                   = FCIDUMP_Vibrational
-  integral_cutoff                 = 1.0E-10
-  init_state                      = 'default'
-  L                               = 12
-  // == vDMRG-specific input ==
-  symmetry                        = none
-  LATTICE                         = 'watson lattice'
-  MODEL                           = 'watson'
-  Nmax                            = 10
-\end{lstlisting}
-
-We report in Listing~\ref{lst:canonicalVibInput} an example input for a vDMRG calculation relying on the canonical-quantization based formalism.
-The first part of the input has the same structure as that of an electronic-structure DMRG calculation.
-Note that the lattice dimension $L$ is, in this case, equal to the overall number of modes of the molecule.
-Unlike for the electronic-structure case, the DMRG symmetry must be set to \texttt{none}, and the \texttt{MODEL} keyword should be set to \texttt{watson}.
-As the expansion reported in Eq.~\ref{eq:CanonicalQuantizationCI} is virtually unbounded, it must be truncated, in practice, by setting a threshold $N_\text{max}$ for the maximum excitation degree of a mode.
-Such a threshold must be specified in input, as shown in Listing~\ref{lst:canonicalVibInput}.
-
-The integral file \texttt{FCIDUMP\_Vibrational} contain the force constants that define the PES through Eq.~\ref{eq:TaylorExpansion}.
-\qcm\ supports up to sixth-order force constants, that must be provided as input in the \texttt{FCIDUMP\_Vibrational} file as illustrated in Listing~\ref{lst:CanonicalFCIDUMP}
-\vspace{.5cm}
-\begin{lstlisting}[language=qcmaquis,caption={Example of an vDMRG integral file for the potential
-											  expressed in the canonical quantization format.},
-				   label=lst:CanonicalFCIDUMP, mathescape=true, rulecolor=\color{black}]
-  $\smash{k_{ij}}$     i    j   0   0   0   0
-  $\smash{\vdots}$
-  $\smash{h_{ij}}$    -i   -j   0   0   0   0
-  $\smash{\vdots}$
-  $\smash{k_{ijk}}$     i    j   k   0   0   0
-  $\smash{\vdots}$
-  $\smash{k_{ijkh}}$     i    j   k   h   0   0
-  $\smash{\vdots}$
-  $\smash{k_{ijkhm}}$     i    j   k   h   m   0
-  $\smash{\vdots}$
-  $\smash{k_{ijkhmn}}$     i    j   k   h   m   n
-\end{lstlisting}
-\vspace{.5cm}
-The $h$ and $k$ constants define the PES as follows:
-
-\begin{equation}
-  \begin{aligned}
-    V(\tilde{Q}_1, \ldots, \tilde{Q}_L) 
-      &= \sum_{i=1}^L k_{ii} \tilde{Q}_i^2 + \sum_{i=1}^L h_{ii} \tilde{P}_i^2 
-      + \sum_{ijk=1}^L k_{ijk} \tilde{Q}_i \tilde{Q}_j \tilde{Q}_k
-      + \sum_{ijkl=1}^L k_{ijkl} \tilde{Q}_i \tilde{Q}_j \tilde{Q}_k \tilde{Q}_l \\
-      &+ \sum_{ijklm=1}^L k_{ijkl} \tilde{Q}_i \tilde{Q}_j \tilde{Q}_k \tilde{Q}_l \tilde{Q}_m
-      + \sum_{ijklmn=1}^L k_{ijkl} \tilde{Q}_i \tilde{Q}_j \tilde{Q}_k \tilde{Q}_l \tilde{Q}_m \tilde{Q}_n \, ,
-  \end{aligned}
-  \label{eq:PES_Representation}
-\end{equation}
-%
-where we introduced the scaled dimensionless normal coordinates $\tilde{Q}_i$ that are defined as:
-
-\begin{equation}
-  \tilde{Q}_i = \sqrt{2} Q_i = \left( b_i^\dagger + b_i \right)
-  \label{eq:ScaledNM}
-\end{equation}
-%
-$Q_i$ being the $i$-th dimensionless normal coordinate operator.
-The corresponding scaled dimensionless momentum operator $\tilde{P}_i$ is defined as:
-
-\begin{equation}
-  \tilde{P}_i = \frac{\sqrt{2}}{\mathrm{i}} P_i = \left( b_i^\dagger - b_i \right)
-  \label{eq:ScaledNM_Momentum}
-\end{equation}
-
-\subsection{$n$-mode vDMRG}
-\label{sec:nMode}
-
-The canonical quantization-based vDMRG theory requires the PES to be expressed as a Taylor series in terms of the Cartesian-based normal coordinates.
-Such a format is, however, ill-suited to describe
-\qcm\ can apply the vDMRG method to potentials that are encoded as generic many-body expansions \cite{Bowman1997_ManyBody}
-
-\begin{equation}
-  \mathcal{V}(\bm{Q}) = \sum_{i=1}^L V^{[i]}(Q_i) + \sum_{i<j}^L V^{[ij]}(Q_i, Q_j)
-    + \sum_{ijk}^L V^{[ijk]}(Q_i, Q_j, Q_k)
-  \label{eq:PES_ManyBody}
-\end{equation}
-%
-and supports un to three-body coupling terms.
-To encode Eq.~(\ref{eq:PES_ManyBody}) as a matrix product operator, \qcm\ relies on the so-called $n$-mode second quantization \cite{Christiansen2004_nModeQuantization} to encode the vibrational Hamiltonian.
-Let $\{ \chi_{1}^{(i)}(Q_i), \ldots, \chi_{N_i}^{(i)}(Q_i) \}$ be the $N_i$-dimensional modal basis describing the $i$-th mode.
-The full Vibrational Configuration Interaction (VCI) space is spanned by the basis constructed by forming all possible products of single particle bases, \textit{i.e.} by the functions $\Psi^{i_1,\ldots,i_L}(Q_1,\ldots,Q_L)$ defined as:
-
-\begin{equation}
-  \Psi^{i_1,\ldots,i_L}(Q_1,\ldots,Q_L) = \prod_{j=1}^L \chi_{i_j}^{(j)}(Q_j)
-  \label{eq:ProductBasis}
-\end{equation}
-
-It is possible to encode each basis function of Eq.~(\ref{eq:ProductBasis}) as an occupation number vector format, analogously to the electronic structure case, and define the corresponding creation and annihilation operators.
-Following Ref.~\citenum{Christiansen2004_nModeQuantization}, we define a pair of creation and annihilation operators per mode and per modals ($a_{i_j}^\dagger$ and $a_{i_j}$, respectively).
-The second-quantized form of Eq.~\ref{eq:PES_ManyBody} then reads, for a two-body potential (the extension to three-body coupling terms is trivial):
-
-\begin{equation}
- \begin{aligned}
-  \mathcal{V}(\bm{Q}) &= \sum_{i=1}^L \sum_{m_i, n_i}^{N_i}
-    \langle \chi_{m_i}^{(i)}(Q_i) | V^{[i]}(Q_i) | \chi_{n_i}^{(i)}(Q_i) \rangle \\
-  &+ \sum_{i<j}^L \sum_{m_i, n_i}^{N_i} \sum_{m_j, n_j}^{N_j}
-     \langle \chi_{m_i}^{(i)}(Q_i) \chi_{m_j}^{(j)}(Q_j) | V^{[ij]}(Q_i, Q_j) | 
-             \chi_{n_i}^{(i)}(Q_i) \chi_{n_j}^{(j)}(Q_j) \rangle \,
- \end{aligned}
- \label{eq:PES_SecondQuantization}
-\end{equation}
-%
-\qcm\ constructs Eq.~(\ref{eq:PES_SecondQuantization}) starting from the integrals of the potential operator in the modal basis, which are provided as input following the formatting given in Listing~\ref{lst:vibdump}
-
-\begin{lstlisting}[language=qcmaquis,caption={Example of an vDMRG integral file for the potential
-											  expressed in the $n$-mode format.},
-				   label=lst:vibdump, mathescape=true]
-  1 0   1 0   $\langle \chi_{0}^{(1)}(Q_1) | V^{[1]}(Q_1) | \chi_{0}^{(1)}(Q_1) \rangle$
-  1 1   1 0   $\langle \chi_{1}^{(1)}(Q_1) | V^{[1]}(Q_1) | \chi_{0}^{(1)}(Q_1) \rangle$
-  1 0   1 1   $\langle \chi_{0}^{(1)}(Q_1) | V^{[1]}(Q_1) | \chi_{1}^{(1)}(Q_1) \rangle$
-  $\vdots$
-  2 0   2 0   $\langle \chi_{0}^{(2)}(Q_2) | V^{[2]}(Q_2) | \chi_{0}^{(2)}(Q_2) \rangle$
-  2 1   2 0   $\langle \chi_{1}^{(2)}(Q_2) | V^{[2]}(Q_2) | \chi_{0}^{(2)}(Q_2) \rangle$
-  2 0   2 1   $\langle \chi_{0}^{(2)}(Q_2) | V^{[2]}(Q_2) | \chi_{1}^{(2)}(Q_2) \rangle$
-  $\vdots$
-  3 0   3-0   $\langle \chi_{0}^{(3)}(Q_3) | V^{[3]}(Q_3) | \chi_{0}^{(3)}(Q_3) \rangle$
-  3 1   3-0   $\langle \chi_{1}^{(3)}(Q_3) | V^{[3]}(Q_3) | \chi_{0}^{(3)}(Q_3) \rangle$
-  3 0   3-1   $\langle \chi_{0}^{(3)}(Q_3) | V^{[3]}(Q_3) | \chi_{1}^{(3)}(Q_3) \rangle$ 
-  $\vdots$
-  1 0   1-0   2 0   2 0  $\langle \chi_{0}^{(1)}(Q_1) \chi_{0}^{(2)}(Q_2) | V^{[1,2]}(Q_1, Q_2) | \chi_{0}^{(1)}(Q_1) \chi_{0}^{(2)}(Q_2) \rangle$
-  1 0   1-0   2 0   2 1  $\langle \chi_{0}^{(1)}(Q_1) \chi_{0}^{(2)}(Q_2) | V^{[1,2]}(Q_1, Q_2) | \chi_{0}^{(1)}(Q_1) \chi_{1}^{(2)}(Q_2) \rangle$
-  $\vdots$
-  1 0   1 0   2 0   2 0   3 0   3 0   $\langle \chi_{0}^{(1)}(Q_1) \chi_{0}^{(2)}(Q_2) \chi_{0}^{(3)}(Q_3) | V^{[1,2,3]}(Q_1, Q_2, Q_3) | \chi_{0}^{(1)}(Q_1) \chi_{0}^{(2)}(Q_2) \chi_{0}^{(3)}(Q_3) \rangle$
-\end{lstlisting}
-
-Note that modes are numbered starting from 1, while modals are numbered starting from 0.
-Moreover, Hermitianity is not assumed, so that both the $\langle \chi_{1}^{(1)}(Q_1) | V^{[1]}(Q_1) | \chi_{0}^{(1)}(Q_1) \rangle$ and the $\langle \chi_{0}^{(1)}(Q_1) | V^{[1]}(Q_1) | \chi_{1}^{(1)}(Q_1) \rangle$ entries must be given as input.
-The corresponding input file is given in Listing~\ref{lst:nModeVibInput}
-
-\begin{lstlisting}[language=qcmaquis,
-				   caption={Input example for a $n$-mode vDMRG-FCI calculation},
-				   label=lst:nModeVibInput]
-  // == Generic input section ==
-  nsweeps                         = 6
-  max_bond_dimension              = 1000
-  model_library                   = coded
-  lattice_library                 = coded
-  optimization                    = twosite 
-  integral_file                   = FCIDUMP_Vibrational
-  integral_cutoff                 = 1.0E-10
-  init_state                      = 'default'
-
-  // == vDMRG-specific input ==
-  // The symmetry group is nu1 for all n-mode vDMRG calculations
-  symmetry                        = nu1
-  // This is the only lattice associated with the n-mode vDMRG method
-  LATTICE                         = "nmode lattice" 
-  // n-mode vDMRG model
-  MODEL                           = nMode
-  // String with the overall number of modals
-  nmode_num_modes                 = 20
-  // Maximum order of many-body coupling (1, 2, or 3)
-  nmode_max_coupling              = 2
-  // Number of basis functions per mode
-  nmode_num_basis                 = "10 10"
-\end{lstlisting}
-
-Note that the size of the vector given as input for the \texttt{nmode\_num\_basis} basis must be equal to the integer given as \texttt{nmode\_num\_modes} parameter.
-
-\subsection{DMRG for excitonic Hamiltonians}
-\label{sec:excitonic}
-
-\qcm\ can be applied also to simulate vibronic processes, \textit{i.e.} phenomena that are tuned simultaneously by electronic and vibrational effects.
-Specifically, \qcm\ implements two classes of Hamiltonians: the Holstein-Hubbard Hamiltonian and \textit{ab initio} vibronic Hamiltonians based on the linear vibronic coupling (LVC) model.
-
-The Holstein-Hubbard Hamiltonian is defined by the following parameters:
-
-\begin{itemize}
-  \item The number $N_\text{mol}$ of monomers forming the aggregate (input parameter \texttt{n\_excitons}).
-  \item The number of vibrations $N_\text{vib}$ of vibrational degrees of freedom that are included per aggregate (input parameter \texttt{vibronic\_nmodes}).
-  \item The harmonic frequencies $\omega_i$ associated with the $N_\text{vib}$ vibrations (defined in the \texttt{integral\_file}, as described below).
-  \item The $N_\text{vib}$ parameters $g_j$ of the LVC Hamiltonian (defined in the \texttt{integral\_file}, as described below).
-  \item The Coulomb coupling $J$ (keyword \texttt{J\_coupling}).
-\end{itemize}
-
-The Holstein-Hubbard Hamiltonian $\mathcal{H}_\text{Hol}$ reads, in practice:
-
-\begin{equation}
- \begin{aligned}
-  \mathcal{H}_\text{Hol} =& | 0_1 0_2 \cdots 0_{N_\text{mol}} \rangle \langle 0_1 0_2 \cdots 0_{N_\text{mol}} |
-	\left[ \sum_{i=1}^{N_\text{mol}} \sum_{j=1}^{N_\text{vib}} 
-			\left( -\frac{1}{2} \mathcal{T}(q_{j}^{(i)}) + \frac{1}{2} \omega_j \left( q_j^{(i)} \right)^2 \right)
-	\right] \\
-	+& \sum_{i=1}^{N_\text{mol}} | 0_1 \cdots 1_i \cdots 0_{N_\text{mol}} \rangle 
-								 \langle 0_1 \cdots 1_i \cdots 0_{N_\text{mol}} | 
-	   \sum_{j=1}^{N_\text{vib}}
-	   \left( -\frac{1}{2} \mathcal{T}(q_{j}^{(i)}) + \frac{1}{2} \omega_j \left( q_j^{(i)} \right)^2 
-			  + g_j q_{j}^{(i)} \right) \\
-    +& J \sum_{i=1}^{N_\text{mol}-1} | 0_1 \cdots 1_i \cdots 0_{N_\text{mol}} \rangle
-	   								 \langle 0_1 \cdots 1_{i+1} \cdots 0_{N_\text{mol}} |
- \end{aligned}
- \label{eq:HolsteinHubbard}
-\end{equation}
-
-In Eq.~(\ref{eq:HolsteinHubbard}), $| 0_1 0_2 \cdots 0_{N_\text{mol}} \rangle$ represents the electronic configuration, where all the molecules are in their ground-state electronic state, while $| 0_1 \cdots 1_i \cdots 0_{N_\text{mol}} \rangle$ represents the configuration in which the $i$-th molecule of the aggregate has been excited.
-
-In order to apply DMRG to the Hamiltonian defined by Eq.~(\ref{eq:HolsteinHubbard}), the degrees of freedom must be mapped onto an appropriate one-dimensional lattice.
-Following Ref.~\citenum{Baiardi2019_TDDMMRG}, we construct a DMRG lattice that includes $N_\text{mol}$ electronic sites, each one describing a different molecule of the aggregate, and $N_\text{vib}$ vibrational sites per molecule.
-The overall number of sites is, therefore, $N_\text{mol} \times (N_\text{vib}+1)$.
-Currently, \qcm\ supports two different choices for sorting these sites into the DMRG lattice, that can be specified \textit{via} the \texttt{vibronic\_sorting} keyword:
-
-\begin{itemize}
-  \item The first sorting is obtained as follows: the first site is the electronic site associated with the first molecule of the aggregate.
-The following $N_\text{vib}$ sites are associated with the vibrations of the first molecule.
-The next site is then mapped to the electronic site associated with the second molecule, and the following $N_\text{vib}$ sites are used to represent the corresponding vibrational degrees of freedom.
-This procedure is repeated until all the molecules of the aggregate have been included.
-This mapping is obtained by setting the keyword \texttt{vibronic\_sorting} to \texttt{intertwined}.
-  \item Alternatively, it is possible to map the electronic degrees of freedom onto the first $N_\text{mol}$ sites of the lattice, and use the remaining ones for the vibrational parts.
-  Such a mapping is obtained by setting the \texttt{vibronic\_sorting} keyword to \texttt{firstele}.
-\end{itemize}
-
-\begin{lstlisting}[language=qcmaquis,
-				   caption={Input example for an excitonic DMRG calculation},
-				   label=lst:excitonicInput]
-  // == Generic input section ==
-  nsweeps                         = 6
-  max_bond_dimension              = 20
-  model_library                   = coded
-  lattice_library                 = coded
-  optimization                    = singlesite
-  integral_file                   = FCIDUMP_Excitonic
-  init_state                      = 'default'
-  // The symmetry group is u1 for all vibronic DMRG calculations
-  symmetry                        = u1
-  // This is lattice is used for both vibronic and excitonic calculations.
-  LATTICE                         = "vibronic lattice" 
-  // Vibronic model
-  MODEL                           = vibronic
-  J_coupling                      = -500
-  // This could have been also 'firstele'
-  vibronic_sorting                = 'intertwined'
-  // Number of modes and of molecules
-  vibronic_nmodes                 = 10
-  n_excitons                      = 6
-\end{lstlisting}
-
-We report in Listing~\ref{lst:excitonicInput} an examples of an input file for running a DMRG calculation on an excitonic Hamiltonian composed by 6 molecules, each one with 10 vibrational modes, coupled by a Coulomb coupling of -500.
-
-\begin{lstlisting}[language=qcmaquis,caption={Excitonic integral file format.},
-				   label=lst:excitonicDump]
-	 103.00      1     1
-	-103.00     -1    -1
-	 105.50      2     2
-	-105.50     -2    -2
-	 270.00      3     3
-	-270.00     -3    -3
-	 276.00      4     4
-	-276.00     -4    -4
-	 375.50      5     5
-	-375.50     -5    -5
-	 662.50      6     6
-	-662.50     -6    -6
-	 685.50      7     7
-	-685.50     -7    -7
-	 734.50      8     8
-	-734.50     -8    -8
-	 785.50      9     9
-	-785.50     -9    -9
-	 814.50     10    10
-	-814.50    -10   -10
-	 129.30      1     0
-	 138.36      2     0
-	 105.26      3     0
-	 150.16      4     0
-	 192.93      5     0
-	 187.38      6     0
-	 884.27      7     0
-	 425.76      8     0
-	 639.67      9     0
-	 454.68     10     0
-\end{lstlisting}
-
-The parameters defining the excitonic Hamiltonian must be included in the integral file, as shown in Listing~\ref{lst:vibdump}.
-The first 20 lines of the file define the Harmonic component of (see Eq.~(\ref{eq:HolsteinHubbard})).
-The terms are encoded as for the canonical quantization-based Hamiltonian of vDMRG.
-The last 10 lines of the file define the $g_j$ parameter of Eq.~(\ref{eq:HolsteinHubbard}).
-Also in this case, we follow the same encoding as for the canonical quantization vDMRG.
-
-\subsection{DMRG for vibronic Hamiltonians}
-\label{sec:vibronic}
-
-\qcm\ can apply DMRG also to \textit{ab-initio} vibronic Hamiltonians beyond the excitonic approximation to study the excited-state dynamics of isolated molecular systems.
-The specific form of the Hamiltonian that \qcm\ supports is the following:
-
-\begin{equation}
-  \mathcal{H} =
-    \begin{bmatrix}
-	   \mathcal{H}_1(Q_1,\ldots,Q_{N_\text{vib}})
-	     & \mathcal{V}_{12}(Q_1,\ldots,Q_{N_\text{vib}}) 
-         & \cdots 
-		 & \mathcal{V}_{1 N_\text{ele}}(Q_1,\ldots,Q_{N_\text{vib}}) \\
-	   \mathcal{V}_{21}(Q_1,\ldots,Q_{N_\text{vib}})
-	     & \mathcal{H}_2(Q_1,\ldots,Q_{N_\text{vib}})
-         &  
-		 & \\
-	  \vdots
-	     & 
-         & \ddots 
-		 & \\
-	  \mathcal{V}_{N_\text{ele} 1}(Q_1,\ldots,Q_{N_\text{vib}})
-	     &
-         &  
-		 & \mathcal{H}_{N_\text{ele}}(Q_1,\ldots,Q_{N_\text{vib}}) \\
-   \end{bmatrix}
-  \label{eq:VibronicHamiltonian}
-\end{equation}
-%
-for a system with $N_\text{ele}$ electronic states and $N_\text{vib}$ vibrational degrees of freedom.
-$\mathcal{H}_{i}(Q_1,\ldots,Q_{N_\text{vib}})$ represents the vibrational Hamiltonian associated with the $i$-th electronic state, and is expressed as:
-
-\begin{equation}
-  \mathcal{H}_{i}(Q_1,\ldots,Q_{N_\text{vib}}) = E_i^{(eq)}
-	+ \sum_{j=1}^{N_\text{vib}} \omega_j^{(i),2} \left( Q_j^2  + P_j^2 \right) \, ,
-  \label{eq:DiagonalVibronic}
-\end{equation}
-%
-where $E_i^{(eq)}$ is the electronic energy of the equilibrium position, and $Q_j$ and $P_j$ are the $j$-th dimensionless  normal coordinate and the corresponding conjugate momentum, respectively, with harmonic frequency $\omega_j$.
-$\mathcal{V}_{ij}(Q_1,\ldots,Q_{N_\text{vib}})$ is the non-adiabatic coupling between the $i$-th and the $j$-th electronic state, and is expressed as:
-
-\begin{equation}
-  \mathcal{V}_{ij}(Q_1,\ldots,Q_{N_\text{vib}}) =
-   \sum_{k=1}^{N_\text{vib}} g_k^{(i,j)} Q_k
-	+ \sum_{k,l=1}^{N_\text{vib}} h_{k,l}^{(i,j)} Q_k Q_l \, ,
-  \label{eq:OffDiagonalVibronic}
-\end{equation}
-%
-where $g_k^{(i,j)}$ and $h_{k,l}^{(i,j)}$ are the first- and second-order coupling non-adiabatic coupling terms, respectively.
-
-\begin{lstlisting}[language=qcmaquis,
-				   caption={Input example for a vibronic DMRG calculation},
-				   label=lst:vibronicInput]
-  nsweeps                         = 6
-  max_bond_dimension              = 20
-  model_library                   = coded
-  lattice_library                 = coded
-  optimization                    = singlesite
-  integral_file                   = FCIDUMP_Vibronic
-  init_state                      = 'default'
-  symmetry                        = u1
-  LATTICE                         = "vibronic lattice"
-  MODEL                           = vibronic
-  // Note that here the [n_excitons] parameter is replaced 
-  // by the [vibronic_nstates] one
-  vibronic_nmodes                 = 4
-  vibronic_nstates                = 2
-\end{lstlisting}
-
-A sample input file for a DMRG calculation on a vibronic Hamiltonian is given in Listing~\ref{lst:vibronicInput}.
-The key difference between the input for the excitonic calculation, reported in Listing~\ref{lst:excitonicInput}, is that the \texttt{n\_excitons} parameter is replaced by the \texttt{vibronic\_nstates} one, which corresponds to the $N_\text{ele}$ parameter introduced above.
-The integral file \texttt{FCIDUMP\_Vibronic} should contain the definition of the parameters entering the vibronic Hamiltonian, \textit{i.e.}, $\omega_j^{(i)}$, $g_k^{(i,j)}$, and $h_{k,l}^{(i,j)}$.
-We show in Listing~\ref{lst:vibronicDump} the format of the integral file for the case of the 4-mode vibronic Hamiltonian of pyrazine that describes the S$_1$ and S$_2$ electronic states, taken from Ref.~\citenum{Raab1999_Pyrazine}.
-
-\begin{lstlisting}[language=qcmaquis,caption={Vibronic integral file format.},
-				   label=lst:vibronicDump]
-EL_ST 0 0
--4114.23                0      0
-  325.84799             1      1     
- -325.84799            -1     -1
-  559.34550             2      2     
- -559.34550            -2     -2
-  674.68277             3      3     
- -674.68277            -3     -3     
-  518.21124             4      4     
- -518.21124            -4     -4     
-EL_ST 1 1
- 4114.23                0      0
-  325.84799             1      1     
- -325.84799            -1     -1
-  559.34550             2      2     
- -559.34550            -2     -2
-  674.68277             3      3     
- -674.68277            -3     -3     
-  518.21124             4      4     
- -518.21124            -4     -4     
-EL_ST 0 0
- -791.22989             1      0
- -405.69687             2      0
- 1171.11703             3      0
-EL_ST 1 1
- 1092.881251            1      0
--1379.877165            2      0
-  302.457910            3      0
-EL_ST 0 0
-   0.16131088           1      1
-   8.71078783           1      2
- -16.45371035           1      3
-   8.71078783           2      1
- -65.33090875           2      2
-  38.23067993           2      3
- -16.45371035           3      1
-  38.23067993           3      2
-  -9.35603137           3      3
- -93.47965832           4      4
-EL_ST 1 1 
- -73.96104114           1      1
- -24.03532198           1      2
- -15.24387871           1      3
- -24.03532198           2      1
-  39.35985614           2      2
-   9.27537593           2      3
- -15.24387871           3      1
-   9.27537593           3      2
-   1.77441974           3      3
- -93.47965832           4      4
-EL_ST 0 1
- 1677.63321200          4      0
-  -80.65544294          4      1
-  -44.44114904          4      2
-   10.24324125          4      3
-   50.65161814          1      4
-  -44.44114904          2      4
-   10.24324125          3      4
-EL_ST 1 0
- 1677.63321200          4      0
-  -80.65544294          4      1
-  -44.44114904          4      2
-   10.24324125          4      3
-   50.65161814          1      4
-  -44.44114904          2      4
-   10.24324125          3      4
-\end{lstlisting}
-
-Note that the constants entering the vibronic Hamiltonian are represented as in the excitonic case.
-The header \texttt{EL\_ST X Y}, where \texttt{X} and \texttt{Y} are integer smaller than $N_\text{ele}$, indicates the electronic states to which the vibrational constants are associated with.
-For instance, the header \texttt{EL\_ST 0 0} indicates that the constants in lines 2-10 are associated with the 0-0 block of Eq.~\ref{eq:VibronicHamiltonian}.
-Therefore, they represent, in practice, the vibrational Hamiltonian for the S$_0$ state.
-Note that the coefficients associated with the \texttt{0   0} pair of indices (line 3) is the constant energy shift $E_i^{(eq)}$ of Eq.~(\ref{eq:DiagonalVibronic}) (also referred to as $\Delta$ in Ref.~(\citenum{Raab1999_Pyrazine})).
-Each block of Listing~\ref{lst:vibronicDump} corresponds to a class of terms entering Eq.~(1) of Ref.~(\citenum{Raab1999_Pyrazine}).
-Specifically:
-
-\begin{enumerate}
-  \item Lines 3-10 are the $\omega_i$ coefficients for the S$_1$ state.
-  \item Lines 13-20 are the $\omega_i$ coefficients for the S$_2$ state.
-  \item Lines 22-24 are the $a_i$ coefficients.
-  \item Lines 26-28 are the $b_i$ coefficients.
-  \item Lines 30-39 are the $a_{ij}$ coefficients.
-  \item Lines 41-50 are the $b_{ij}$ coefficients.
-  \item Lines 52-58 and 60-66 collect the $c_i$ and $c_{ij}$ coefficients.
-\end{enumerate}
-
-Note that the vibronic Hamiltonian is Hermitian and, therefore, the $\mathcal{V}_{ij}$ and $\mathcal{V}_{ji}$ operators form a hermitian conjugate pair.
-However, this hermiticity is not exploited in the integral file, where both the $\mathcal{V}_{01}$ and $\mathcal{V}_{10}$ operators are explicitly given.
-
-\subsection{Pratical considerations}
-\label{sec:Practical-vDMRG}
-
-% Informations on how to compile the vibrational module of QCMaquis
-The vDMRG library is not activated if the default CMake settings are used when compiling \qcm.
-In order to run vDMRG simulation, the \texttt{BUILD\_VIBRATIONAL} CMake flag should be set to \texttt{ON}.
-This is, however, not sufficient.
-In fact, the Hamiltonians supported by \qcm\ are classified based on the symmetry group to which they belong.
-These groups must be included, at compile time, in the \texttt{BUILD\_SYMMETRIES} CMake keyword.
-By default, \texttt{BUILD\_SYMMETRIES} includes \texttt{TwoU1PG} (\textit{i.e.}, the symmetry of the non spin-adapted non-relativistic Hamiltonian, with the possibility of including point group symmetry) and \texttt{SU2U1PG} (the spin-adapted counterpart).
-The \texttt{NONE} point group must be added in order to compile the canonical quantization version of vDMRG, and the \texttt{NU1} one must be added to compile the $n$-mode variant.
-
-Similarly, in order to enable the support for vibronic and excitonic Hamiltonians, two parameters must be changed when the \texttt{Makefile} is generated.
-First, the \texttt{BUILD\_VIBRONIC} CMake flag should be set to \texttt{ON}.
-Second, the \texttt{U1} group should be added to the list of compiled symmetries, under the \texttt{BUILD\_SYMMETRIES} keyword.
-
-In conclusion, we note that both excitonic and vibronic Hamiltonians are mostly applied to quantum dynamics simulations of non-equilibrium processes.
-We will discuss in the following how to run quantum dynamics simulations with \qcm.
-
-\section{QCMaquis pre-Born--Oppenheimer module}
-\label{sec:preBO}
-
-The \qcm\ pre-Born--Oppenheimer (pre-BO) module provides a tool to perform DMRG-CASCI calculations for \textit{multicomponent} Hamiltonians consisting of multiple spin-$\frac12$ particle types.
-
-QCMaquis follows the formatting of the integral file shown in Listing~\ref{lst:prebodump}.
-The first line, with only a single float, contains a constant shift in energy, e.g., point-charge repulsion of classical nuclei.
-The following lines contain the terms in the Hamiltonian where pairs of numbers $m-n$ describe a single second-quantization operator.
-The first number denotes the particle type (starting with 0) and the second number the orbital index associated with the given particle type (also starting with 0 for all types).
-We assume physics notation and normal-ordering, i.e., the following would be a valid term
-
-\begin{equation}
-  \left(\int \phi_{1-1}(1)\phi_{0-1}(2)g(1,2)\phi_{1-0}(1)\phi_{0-1}(2)\right)
-  a^\dagger_{1-1} a^\dagger_{0-1} a_{0-1} a_{1-0}\,.
-\end{equation}
-
-The second and third lines contain single-paricle type \textit{hopping} terms.
-The fourth line stands for a single-particle type interaction, and the fifth
-line for an interaction between different particle types. 
-Note here that the aforementioned convention implies that operators belonging to
-the same particle types in a given string of operators must be either at
-positions 1 and 4 or 2 and 3.
-
-\begin{lstlisting}[language=qcmaquis,caption={Pre-BO integral file
-format.},label=lst:prebodump]
-1.145901278714102                       
-   0-0         0-0                                      -2.451508889071943
-   1-0         1-0                                      2.30233235666604
-   0-2         0-0         0-1         0-1              -0.08690785254423253
-   1-1         0-1         0-1         1-0              4.887021698358933e-05
-\end{lstlisting}
-
-Listing~\ref{lst:qcm} shows an input example for a pre-BO DMRG-FCI calculation. 
-It assumes that a pre-BO integral file with has been generated. Additionally,
-the 1-particle RDM and the mutual information can be measured and printed to a
-file.
-
-\begin{lstlisting}[language=qcmaquis,
-				   caption={Input example for a pre-BO DMRG-FCI calculation},
-				   label=lst:preboinput]
-  nsweeps                         = 6
-  max_bond_dimension              = 1000
-
-  // Pre-BO specific input
-  // String with the number of parciles for each type
-  PreBO_ParticleTypeVector        = "4 1"
-  // Spin-1/2 fermion (1) or spin-0 boson (0)
-  // Attention: the boson-related functionality has not been fully tested, yet.
-  PreBO_FermionOrBosonVector      = "1 1"
-  // Number of orbitals for each type.
-  PreBO_OrbitalVector             = "3 3"
-  // Number of alpha-beta particles for each type
-  // Note: for spin-0 bosons it would be a single number.
-  PreBO_InitialStateVector        = "2 2 1 0"
-
-optimization                    = twosite 
-symmetry                        = nu1
-
-model_library                   = coded
-lattice_library                 = coded
-LATTICE                         = "preBO lattice"
-
-// Your model
-MODEL                           = PreBO
-init_state                      = 'default'
-
-// Hamiltonian file
-integral_file                   = FCIDUMP.E_4_H1_1
-integral_cutoff                 = 1.0E-10
-
-MEASURE[1rdm]          = 1
-MEASURE[mutinf]        = 1
-\end{lstlisting}
-
-%\subsection{Practical considerations}
-
-\section{\qcm\ examples library}
-\label{sec:examples}
-
-We provide in the folder \texttt{example} a set of example input files to run the different kinds of DMRG-based calculations that are supported by \qcm{}.
-These example calculations are organized in different folder, one per feature provided by \qcm, as listed below:
-
-\begin{itemize}
- \item \texttt{iTD-DMRG}: example calculation (\texttt{H2\_2e4o.iTD.SS.inp}) of an imaginary-time DMRG calculation.
- The integral file is generated for the H$_2$ molecule and based on the 6-31G basis set.
- Note that the calculation should be run with the executable \texttt{applications/evolve/dmrg\_evolve}, which is generated by setting the \texttt{BUILD\_DMRG\_EVOLVE} CMake flag to \texttt{ON}.
- We provide also the input for running a conventional \texttt{DMRG} calculation (\texttt{H2\_2e4o.TI.SS.inp}), which must be run with the \texttt{applications/dmrg/dmrg} executable and should return the same energy as the iTD-DMRG simulation.
- \item \texttt{TD-DMRG}: example calculation of a real-time DMRG calculation.
-The integral file is associated with the $\pi$ orbital space (CAS(6,6)) of benzene calculated with the 6-31G* basis set.
-As already noted above, the calculations should be run with the \texttt{applications/evolve/dmrg\_evolve} executable.
- Note that the initial state for the propagation is a random MPS (the \texttt{init\_state} keyword is set to \texttt{default}).
- Therefore, the calculation does not correspond to any physically meaningful non-equilibrium process. 
- However, this calculation can be used to assess that the \texttt{dmrg\_evolve} executable has been compiled correctly by verifying that the energy is conserved along the propagation.
- For this example calculation we provide the input files for both the single-site (\texttt{Benzene\_6e6o.TD.SS.inp}) and the two-site (\texttt{Benzene\_6e6o.TD.TS.inp}) propagators.
- \item \texttt{vDMRG}: example of a vDMRG calculation based on the $n$-mode second-quantization (\texttt{FAD\_2Mode\_TS.inp}).
- The integral file is generated for a two-mode potential describing the formic acid dimer based on a discrete variable representation primitive basis set.
- For each mode, 10 modal basis functions are included. \\
- The same folder also contains an example calculation for the canonical second-quantization, based on the single-site (\texttt{C2H4.vDMRG.m50.SS.input}) and on the two-site (\texttt{C2H4.vDMRG.m50.TS.input}) optimization algorithm.
- The PES is taken from Ref.~\citenum{Berkelback2021_vHBCI} and represents the potential energy surface of ethylene as a sixth-order force field.
- The calculation uses the single-site optimizer and is initiated from the ONV associated with the ground-state wave function. \\
- Note that the calculations should be run with the executable \texttt{applications/dmrg/dmrg}, which is automatically generated by each \qcm\ compilation configuration. \\
-\end{itemize}
-
-For all input files, we provide also the corresponding output file that can be used as a reference to ensure that the \qcm\ executables have been compiled correctly.
-
 \section{Additional Considerations}

### 2023-05-08 dmrg/applications/tools/mps2ci.cpp

@@ -1,28 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2014-2014 by Sebastian Keller <sebkelle@phys.ethz.ch>
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #include <cmath>
 #include <iterator>
@@ -47,7 +28,6 @@ using std::endl;
 #include "dmrg/models/chem/util.h"
 
 #include "ci_encode.hpp"
-#include "sampling.hpp"
 
 typedef TwoU1PG grp;
 

### 2023-05-08 dmrg/framework/dmrg/models/OperatorHandlers/OpTable.h

@@ -1,0 +1,34 @@
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
+
+#ifndef OPTABLE_H
+#define OPTABLE_H
+
+#include <vector>
+#include <utility>
+#include <stdexcept>
+#include "dmrg/models/tag_detail.h"
+
+template <class Matrix, class SymmGroup>
+class OPTable : public std::vector<typename operator_selector<Matrix, SymmGroup>::type>
+{
+public:
+    typedef tag_detail::tag_type tag_type;
+    typedef typename operator_selector<Matrix, SymmGroup>::type op_t;
+
+private:
+    typedef typename Matrix::value_type mvalue_type;
+
+public:
+    tag_type register_op(op_t const & op_);
+    std::pair<tag_type, mvalue_type> checked_register(const op_t& sample);
+    bool hasRegistered(const op_t& sample);
+};
+
+#include "OpTable.hpp"
+
+#endif

### 2023-05-08 dmrg/framework/dmrg/models/chem/rel/chem_helper.h

@@ -1,29 +1,9 @@
-/*****************************************************************************
- *
- * QCMaquis DMRG Project
- *
- * Copyright (C) 2013 Laboratory for Physical Chemistry, ETH Zurich
- *               2012-2013 by Sebastian Keller <sebkelle@phys.ethz.ch>
- *
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #ifndef REL_QC_CHEM_DETAIL_H
 #define REL_QC_CHEM_DETAIL_H
@@ -41,13 +21,14 @@ namespace detail {
         typedef ::term_descriptor<value_type> term_descriptor;
         typedef typename TagHandler<M, S>::tag_type tag_type;
         typedef Lattice::pos_t pos_t;
+        using InputType = std::complex<double>;
 
         RelChemHelper(BaseParameters & parms, Lattice const & lat_,
                    std::vector<tag_type> const & ident_, std::vector<tag_type> const & fill_,
                    std::shared_ptr<TagHandler<M, S> > tag_handler_)
[…]

### 2023-05-08 dmrg/framework/dmrg/models/generate_mpo/corr_maker.hpp

@@ -1,42 +1,21 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2011-2011 by Bela Bauer <bauerb@phys.ethz.ch>
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #ifndef GENERATE_MPO_CORR_MAKER_H
 #define GENERATE_MPO_CORR_MAKER_H
 
 #include "dmrg/models/generate_mpo/utils.hpp"
-
 #include "dmrg/block_matrix/block_matrix.h"
 #include "dmrg/block_matrix/block_matrix_algorithms.h"
 #include "dmrg/block_matrix/symmetry.h"
-
 #include "dmrg/mp_tensors/mpo.h"
-
-#include "dmrg/models/lattice.h"
-#include "dmrg/models/op_handler.h"
+#include "dmrg/models/lattice/lattice.h"
[…]

### 2023-05-08 dmrg/framework/dmrg/models/generate_mpo/tagged_mpo_maker_optim.hpp

@@ -1,800 +1,739 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2013-2013 by Sebastian Keller <sebkelle@phys.ethz.ch>
- *                            Michele Dolfi <dolfim@phys.ethz.ch>
- *               2021 by Alberto Baiardi <abaiardi@ethz.ch>
- *               2021 by Robin Feldmann <robinfe@ethz.ch>
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #ifndef GENERATE_MPO_TAGGED_MPO_MAKER_H
 #define GENERATE_MPO_TAGGED_MPO_MAKER_H
 
 #include "dmrg/models/generate_mpo/utils.hpp"
-
 #include "dmrg/block_matrix/block_matrix.h"
[…]

### 2023-05-08 dmrg/framework/dmrg/models/lattice.h

@@ -1,142 +1,0 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2011-2013 by Bela Bauer <bauerb@phys.ethz.ch>
- *                            Michele Dolfi <dolfim@phys.ethz.ch>
- *               2020-2021 by Robin Feldmann <robinfe@phys.chem.ethz.ch>
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
-
-#ifndef LATTICE_H
-#define LATTICE_H
-
-#include "dmrg/utils/BaseParameters.h"
-
-#include <vector>
-#include <string>
-#include <boost/shared_ptr.hpp>
-#include <boost/any.hpp>
-#include <boost/filesystem.hpp>
-
-/// lattice common base
-class lattice_impl {
-public:
-    typedef int pos_t;
-    typedef int part_type;
-
-    virtual ~lattice_impl() {}
-
-    virtual std::vector<pos_t> forward(pos_t) const = 0;
-    virtual std::vector<pos_t> all(pos_t) const = 0;
-
-    // non-virtual!
-    template<class T> T get_prop(std::string property) const
[…]

### 2023-05-08 dmrg/framework/dmrg/models/lattice/ChainLattice.hpp

@@ -1,0 +1,119 @@
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
+
+#ifndef MAQUIS_DMRG_lattice_hpp
+#define MAQUIS_DMRG_lattice_hpp
+
+#include <numeric>
+#include <sstream>
+#include <set>
+#include <vector>
+#include <boost/lexical_cast.hpp>
+#include <boost/lambda/lambda.hpp>
+#include "dmrg/models/lattice/lattice.h"
+#include "dmrg/utils/BaseParameters.h"
+
+class ChainLattice : public lattice_impl
+{
+public:
+  // Types definition
+  using pos_t = typename lattice_impl::pos_t;
+
+  /** @brief Constructor from a parameter container */
+  ChainLattice (BaseParameters & parms, bool pbc_=false)
+    : L(parms["L"]) , pbc(pbc_) { }
+
+  /** @brief Constructor from parameters */
+  ChainLattice (int L_, bool pbc_=false)
+    : L(L_), pbc(pbc_) { }
+
+  /** @brief Get next position in the lattice */
+  std::vector<pos_t> forward(pos_t i) const
+  {
+    std::vector<pos_t> ret;
+    if (i < L-1)
+      ret.push_back(i+1);
+    if (pbc && i == L-1)
+      ret.push_back(0);
+    return ret;
+  }
+
+  /** @brief Getter for the neighbours of a given site */
+  std::vector<pos_t> all(pos_t i) const
+  {
+    std::vector<pos_t> ret;
+    if (i < L-1)
+      ret.push_back(i+1);
+    if (i > 0)
+      ret.push_back(i-1);
+    if (pbc && i == L-1)
+      ret.push_back(0);
+    if (pbc && i == 0)
+      ret.push_back(L-1);
+    return ret;
+  }
+
+  /** @brief Getter for a generic property of the lattice */
+  boost::any get_prop_(std::string const & property, std::vector<pos_t> const & pos) const
+  {
+    if (property == "label" && pos.size() == 1)
+      return boost::any( site_label(pos[0]) );
+    else if (property == "label" && pos.size() == 2)
+      return boost::any( bond_label(pos[0], pos[1]) );
+    else if (property == "type" && pos.size() == 1)
+      return boost::any( 0 );
+    else if (property == "type" && pos.size() == 2)
+      return boost::any( 0 );
[…]

### 2023-05-08 dmrg/framework/dmrg/models/lattice/SquareLattice.hpp

@@ -1,0 +1,109 @@
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
+
+#ifndef MAQUIS_DMRG_SQUARE_LATTICE
+#define MAQUIS_DMRG_SQUARE_LATTICE
+
+#include "dmrg/models/lattice/lattice.h"
+#include <sstream>
+#include <vector>
+#include <set>
+#include <boost/lexical_cast.hpp>
+#include <boost/lambda/lambda.hpp>
+#include <numeric>
+#include "dmrg/utils/BaseParameters.h"
+
+class SquareLattice : public lattice_impl
+{
+public:
+  SquareLattice(BaseParameters & parms) 
+    : L_(parms["L"]) , W_(parms["W"]) , a(parms["a"]) { }
+
+  /*
+   0 4  8 12
+   1 5  9 13
+   2 6 10 14
+   3 7 11 15
+   */
+  std::vector<int> forward(int p) const
+  {
+    std::vector<int> ret;
+    if (p+1 < L_*W_ && (p+1) % W_ != 0)
+      ret.push_back(p+1);
+    if (p+W_ < L_*W_)
+      ret.push_back(p+W_);
+    return ret;
+  }
+
+  std::vector<int> all(int p) const
+  {
+    std::vector<int> ret = forward(p);
+    if (p >= 1 && p % W_ != 0)
+      ret.push_back(p-1);
+    if (p >= W_)
+      ret.push_back(p-W_);
+    return ret;
+  }
+
+  /** @brief Getter for the lattice size */
+  int size() const { return L_*W_; }
+  
+  /** @brief Getter for the number of types of sites */
+  int getMaxType() const override { return 1; }
+
+  boost::any get_prop_(std::string const & property, std::vector<pos_t> const & pos) const
+  {
+    if (property == "label" && pos.size() == 1)
+      return boost::any( site_label(pos[0]) );
+    else if (property == "label" && pos.size() == 2)
+      return boost::any( bond_label(pos[0], pos[1]) );
+    else if (property == "type" && pos.size() == 1)
+      return boost::any( 0 );
+    else if (property == "type" && pos.size() == 2)
+      return boost::any( 0 );
+    else if (property == "x" && pos.size() == 1)
+      return boost::any( x(pos[0]) );
[…]

### 2023-05-08 dmrg/framework/dmrg/models/lattice/WatsonLattice.hpp

@@ -1,0 +1,121 @@
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
+
+#ifndef MAQUIS_DMRG_WATSON_LATTICE
+#define MAQUIS_DMRG_WATSON_LATTICE
+
+#ifdef DMRG_VIBRATIONAL
+
+#include <sstream>
+#include <vector>
+#include <set>
+#include <boost/lexical_cast.hpp>
+#include <boost/lambda/lambda.hpp>
+#include <numeric>
+#include "dmrg/utils/BaseParameters.h"
+#include "dmrg/models/lattice/lattice.h"
+
+/**
+ * @brief Watson lattice class.
+ * 
+ * Lattice representing the canonical quantization-based vibrational Hamiltonian.
+ * Unlike the NU1 case, each site is mapped to a mode.
+ */
+
+class WatsonLattice : public lattice_impl
+{
+public:
+  // Types definition
+  using post_t = lattice_impl::pos_t;
+    
+  /**
+   * @brief Class constructor for the lattice
+   * @param parameters parameter container
+   */
+  WatsonLattice(BaseParameters& parameters) : L(parameters["L"]), numTypes(parameters["L"]) { }
+    
+  /** @brief Returns the next position in the lattice */
+  std::vector<pos_t> forward(pos_t i) const {
+    std::vector<pos_t> ret;
+    if (i < L-1)
+      ret.push_back(i+1);
+    return ret;
+  }
+    
+  /** @brief Returns the neighbors of a given site */
+  std::vector<pos_t> all(pos_t i) const {
+    std::vector<pos_t> ret;
+    if (i < L-1)
+      ret.push_back(i+1);
+    if (i > 0)
+      ret.push_back(i-1);
+    return ret;
+  }
+
+  /**
+   * @brief Getter for the property
+   * 
+   * Note that, in additional to the usual properties of a lattice, we code
+   * the additional property "sublatticePos" which states where the sublattice
+   * associated with a given mode is starting
+   * 
+   * @param property string identifier for the property
+   * @param pos vector of positions 
+   * @return boost::any requested property
+   */
+  boost::any get_prop_(std::string const & property, std::vector<pos_t> const & pos) const
+  {
+    if (property == "label" && pos.size() == 1)
+      return boost::any(site_label(pos[0]));
[…]

### 2023-05-08 dmrg/framework/dmrg/models/lattice/lattice_factory.cpp

@@ -1,0 +1,40 @@
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
+
+#include "dmrg/models/lattice/lattice.h"
+#include "dmrg/models/factories/factory_lattice.hpp"
+
+#ifdef ENABLE_ALPS_MODELS
+#include "dmrg/models/alps/lattice.hpp"
+#endif
+
+#ifdef ENABLE_LL_MODELS
+#include "dmrg/models/ll/ll_models.h"
+#endif
+
+/// lattice factory
+std::shared_ptr<lattice_impl>
+lattice_factory(BaseParameters & parms)
+{
+    typedef std::shared_ptr<lattice_impl> impl_ptr;
+
+    if (parms["lattice_library"] == "coded") {
+        return coded_lattice_factory(parms);
+    } else if (parms["lattice_library"] == "alps") {
+#ifdef ENABLE_ALPS_MODELS
+        return impl_ptr( new alps_lattice(parms) );
+#else
+        throw std::runtime_error("This code was compiled without alps lattice.");
+#endif
+#ifdef ENABLE_LL_MODELS
+    } else if (parms["lattice_library"] == "ll") {
+        return ll_lattice_factory(parms);
+#endif
+    } else {
+        throw std::runtime_error("Don't know this lattice_library!");
+    }
+}

### 2023-05-08 dmrg/framework/dmrg/models/meas_prepare.hpp

@@ -1,34 +1,14 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2011-2011 by Bela Bauer <bauerb@phys.ethz.ch>
- *               2011-2013    Michele Dolfi <dolfim@phys.ethz.ch>
- * 
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- * 
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #ifndef MAQUIS_DMRG_MODEL_MEAS_PREPARE_HPP
 #define MAQUIS_DMRG_MODEL_MEAS_PREPARE_HPP
 
-#include "dmrg/models/lattice.h"
+#include "dmrg/models/lattice/lattice.h"
 #include "dmrg/models/generate_mpo.hpp"
 
 namespace meas_prepare {

### 2023-05-08 dmrg/framework/dmrg/models/measurements/measurements_details.h

@@ -1,32 +1,9 @@
-/*****************************************************************************
-*
-* ALPS MPS DMRG Project
-*
-* Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
-*               2011-2011 by Bela Bauer <bauerb@phys.ethz.ch>
-*               2011-2013    Michele Dolfi <dolfim@phys.ethz.ch>
-*               2014-2014    Sebastian Keller <sebkelle@phys.ethz.ch>
-*               2016         Stefan Knecht <stknecht@ethz.ch>
-*               2020         Leon Freitag <lefreita@ethz.ch>
-*
-* This software is part of the ALPS Applications, published under the ALPS
-* Application License; you can use, redistribute it and/or modify it under
-* the terms of the license, either version 1 or (at your option) any later
-* version.
-*
-* You should have received a copy of the ALPS Application License along with
-* the ALPS Applications; see the file LICENSE.txt. If not, the license is also
-* available from http://alps.comp-phys.org/.
-*
-* THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
-* IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
-* FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
-* SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
-* FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
-* ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
-* DEALINGS IN THE SOFTWARE.
-*
-*****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 #ifndef MEASUREMENTS_DETAILS_H
 #define MEASUREMENTS_DETAILS_H
 

### 2023-05-08 dmrg/framework/dmrg/models/model.h

@@ -1,28 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2011-2011 by Michele Dolfi <dolfim@phys.ethz.ch>
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #ifndef MAQUIS_DMRG_MODELS_MODELS_H
 #define MAQUIS_DMRG_MODELS_MODELS_H
@@ -30,9 +11,8 @@
 #include "dmrg/utils/BaseParameters.h"
 #include "dmrg/models/term_descriptor.h"
 #include "dmrg/models/measurement.h"
-
-#include "dmrg/models/lattice.h"
-#include "dmrg/models/op_handler.h"
+#include "dmrg/models/lattice/lattice.h"
+#include "dmrg/models/OperatorHandlers/TagHandler.h"
[…]

### 2023-05-08 dmrg/framework/dmrg/models/op_handler.hpp

@@ -1,329 +1,0 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2013-2013 by Sebastian Keller <sebkelle@phys.ethz.ch>
- *
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
-
-#ifndef MAQUIS_DMRG_MODELS_OP_HANDLER_HPP
-#define MAQUIS_DMRG_MODELS_OP_HANDLER_HPP
-
-// **************************************************************************
-// OPTable implementation
-// **************************************************************************
-
-template <class Matrix, class SymmGroup>
-typename OPTable<Matrix, SymmGroup>::tag_type
-OPTable<Matrix, SymmGroup>::register_op(op_t const & op_)
-{
-    tag_type ret = this->size();
-    this->push_back(op_);
-    return ret;
-}
-
-template <class Matrix, class SymmGroup>
-std::pair<typename OPTable<Matrix, SymmGroup>::tag_type, typename OPTable<Matrix, SymmGroup>::mvalue_type>
-OPTable<Matrix, SymmGroup>::checked_register(op_t const& sample)
-{
-    std::pair<bool, mvalue_type> cmp_result;
[…]

### 2023-05-08 dmrg/framework/dmrg/models/vibrational/VibrationalModelTraitClass.hpp

@@ -1,0 +1,37 @@
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
+
+#ifndef VIBRATIONAL_MODEL_TRAIT_CLASS_H
+#define VIBRATIONAL_MODEL_TRAIT_CLASS_H
+
+#include "dmrg/block_matrix/symmetry.h"
+
+/** @brief Trait class containing data required for vibrational models */
+template<class SymmGroup>
+class VibrationalModelTraitClass {};
+
+/**
+ * @brief Class specialization for the TrivialGroup
+ * The trait class includes the maximum number of coupling terms that are supported.
+ */
+template<>
+class VibrationalModelTraitClass<TrivialGroup> {
+public:
+  static constexpr int maximumNumberOfCouplings = 6;
+};
+
+/**
+ * @brief Class specialization for the TrivialGroup
+ * The trait class includes the maximum number of coupling terms that are supported.
+ */
+template<>
+class VibrationalModelTraitClass<U1> {
+public:
+  static constexpr int maximumNumberOfCouplings = 2;
+};
+
+#endif // VIBRATIONAL_MODEL_TRAIT_CLASS_H

### 2023-05-08 dmrg/framework/dmrg/models/vibrational/u1/VibronicModel.hpp

@@ -1,0 +1,252 @@
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
+
+#ifndef MODELS_VIBRONIC_U1_H
+#define MODELS_VIBRONIC_U1_H
+
+#ifdef DMRG_VIBRONIC
+
+#include "dmrg/models/measurements/local_at.h"
+#include "dmrg/models/model_helper.hpp"
+#include "dmrg/models/vibrational/VibrationalHelperClass.hpp"
+#include "dmrg/models/vibrational/VibronicIntegralParser.hpp"
+
+/**
+ * @brief Class representing an ab-initio vibronic Hamiltonian.
+ * 
+ * The Hamiltonian is defined by providing as input the parameters
+ * entering the definition of the vibronic Hamiltonian.
+ * So far, coupling terms only up to the second-order can be given
+ * as input (therefore, only the LVC and the QVC models are supported)
+ */
+
+template<class Matrix>
+class VibronicModel : public model_impl<Matrix, U1>
+{
+public:
+    // Types definition
+    using base = model_impl<Matrix, U1>;
+    using table_type = typename base::table_type;
+    using table_ptr = typename base::table_ptr;
+    using tag_type = typename base::tag_type;
+    using op_t = typename base::op_t;
+    using measurements_type = typename base::measurements_type;
+    using value_type = typename Matrix::value_type;
+    using pos_t = typename Lattice::pos_t;
+    using positions_type = typename std::vector<pos_t>;
+    using operators_type = typename std::vector<tag_type>;
+
+    /** @brief Class constructor */
+    VibronicModel(const Lattice& lattice_, BaseParameters& parameters_)
+      : lat(lattice_), parameters(parameters_), tag_handler(new table_type()), L_(parameters_["L"]),
+        n_ele_states_(parameters_["vibronic_nstates"]), n_vib_states_(parameters_["vibronic_nmodes"])
+    {
+        // Variable definition
+        int nMax = parameters_["Nmax"];
+        int maxCoupling = chem::getIndexDim(chem::Hamiltonian::Vibronic) / 2; // Two indices are required as the electronic state is also included
+        op_t ident_vib_op, ident_ele_op, create_ele_op, destroy_ele_op, count_ele_op;
+        op_t position_vib_op, momentum_vib_op;
[…]

### 2023-05-08 dmrg/framework/dmrg/mp_tensors/contractions/abelian/special.hpp

@@ -1,29 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2011-2011 by Bela Bauer <bauerb@phys.ethz.ch>
- *               2014-2014 by Michele Dolfi <dolfim@phys.ethz.ch>
- * 
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- * 
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #ifndef CONTRACTIONS_SPECIAL_HPP
 #define CONTRACTIONS_SPECIAL_HPP

### 2023-05-08 dmrg/framework/dmrg/mp_tensors/multigrid.h

@@ -1,29 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2011-2012 by Bela Bauer <bauerb@phys.ethz.ch>
- *                            Michele Dolfi <dolfim@phys.ethz.ch>
- * 
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- * 
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #ifndef MULTIGRID_H
 #define MULTIGRID_H

### 2023-05-08 dmrg/framework/dmrg/mp_tensors/ts_ops.h

@@ -1,30 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2015 Institute for Theoretical Physics, ETH Zurich
- *                    Laboratory for Physical Chemistry, ETH Zurich
- *               2013-2015 by Sebastian Keller <sebkelle@phys.ethz.ch>
- *
- * 
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- * 
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #ifndef MAQUIS_DMRG_TS_OPS_H
 #define MAQUIS_DMRG_TS_OPS_H
@@ -35,7 +14,7 @@
 #include "dmrg/block_matrix/multi_index.h"
 #include "dmrg/block_matrix/site_operator.h"
 #include "dmrg/block_matrix/site_operator_algorithms.h"
-
+#include "dmrg/models/OperatorHandlers/KronHandler.h"
 
 namespace ts_ops_detail
 {
@@ -60,126 +39,116 @@ namespace ts_ops_detail
[…]

### 2023-05-08 dmrg/framework/dmrg/sim/sim.h

@@ -1,29 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2011-2013 by Bela Bauer <bauerb@phys.ethz.ch>
- *                            Michele Dolfi <dolfim@phys.ethz.ch>
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #ifndef APP_SIM_H
 #define APP_SIM_H
@@ -31,39 +11,31 @@
 #include <cmath>
 #include <iterator>
 #include <iostream>
-
 #include <boost/filesystem.hpp>
 #include <boost/optional.hpp>
-
 #include "utils/data_collector.hpp"
-
 #include "dmrg/utils/DmrgParameters.h"
-
 #include "dmrg/mp_tensors/mps.h"
 #include "dmrg/mp_tensors/mps_initializers.h"
 #include "dmrg/mp_tensors/mpo.h"
 #include "dmrg/models/generate_mpo.hpp"
-
[…]

### 2023-05-08 dmrg/framework/dmrg/utils/storage.h

@@ -1,29 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2011-2011 by Bela Bauer <bauerb@phys.ethz.ch>
- *               2011-2012 by Michele Dolfi <dolfim@phys.ethz.ch>
- * 
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- * 
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT 
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE 
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE, 
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER 
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #ifndef STORAGE_H
 #define STORAGE_H
@@ -50,316 +30,435 @@ namespace storage {
 }
 #endif
 
-template<class Matrix, class SymmGroup> class Boundary;
-template<class Matrix, class SymmGroup> class MPSTensor;
-template<class Matrix, class SymmGroup> class block_matrix;
-template<class Matrix, class SymmGroup> class SiteOperator;
+// Forward declaration of all the serializable objects.
+template<class Matrix, class SymmGroup>
+class Boundary;
 
[…]

### 2023-05-08 dmrg/tests/Fixtures/VibronicFixture.h

@@ -1,0 +1,253 @@
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
+
+#ifndef TEST_VIBRONIC_FIXTURE_H
+#define TEST_VIBRONIC_FIXTURE_H
+
+#include "dmrg/utils/DmrgParameters.h"
+#include "maquis_dmrg.h"
+
+/** @brief Fixture class for the test of the Vibronic DMRG */
+struct VibronicFixture
+{
+    /** @brief Constructor for the fixture class */
+    VibronicFixture() {
+        // Single-excited state excitonic model
+        parametersExcitonicAggregate.set("L", 66);
+        parametersExcitonicAggregate.set("symmetry", "u1");
+        parametersExcitonicAggregate.set("LATTICE", "vibronic lattice");
+        parametersExcitonicAggregate.set("MODEL", "excitonic");
+        parametersExcitonicAggregate.set("Nmax", 8);
+        parametersExcitonicAggregate.set("vibronic_nstates", 1);
+        parametersExcitonicAggregate.set("vibronic_nmodes", 10);
+        parametersExcitonicAggregate.set("n_excitons", 6);
+        parametersExcitonicAggregate.set("J_coupling", -500);
+        parametersExcitonicAggregate.set("J_interaction", "nn");
+        parametersExcitonicAggregate.set("integral_file", "integral_file_Excitonic");
+        //
+        parametersExcitonicAggregateTwoSites.set("L", 22);
+        parametersExcitonicAggregateTwoSites.set("symmetry", "u1");
+        parametersExcitonicAggregateTwoSites.set("LATTICE", "vibronic lattice");
+        parametersExcitonicAggregateTwoSites.set("MODEL", "excitonic");
+        parametersExcitonicAggregateTwoSites.set("Nmax", 8);
+        parametersExcitonicAggregateTwoSites.set("vibronic_nstates", 1);
+        parametersExcitonicAggregateTwoSites.set("vibronic_nmodes", 10);
+        parametersExcitonicAggregateTwoSites.set("n_excitons", 2);
+        parametersExcitonicAggregateTwoSites.set("J_coupling", -500);
+        parametersExcitonicAggregateTwoSites.set("J_interaction", "nn");
[…]

### 2023-05-08 dmrg/tests/LinearSystem/LinearSystemSolverElectronic.cpp

@@ -1,0 +1,63 @@
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
+
+#define BOOST_TEST_MODULE LinearSolverElectronic
+
+#include "dmrg/LinearSystem/LinSystemTraitsClass.h"
+#include "dmrg/mp_tensors/mps.h"
+#include "dmrg/mp_tensors/mpo.h"
+#include "dmrg/models/model.h"
+#include "dmrg/models/lattice/lattice.h"
+#include "dmrg/models/generate_mpo.hpp"
+#include "dmrg/sim/matrix_types.h"
+#include "dmrg/SweepBasedAlgorithms/SweepBasedLinearSystem.h"
+#include "Fixtures/BenzeneFixture.h"
+#include "utils/fpcomparison.h"
+#include <boost/test/included/unit_test.hpp>
+
+/**
+ * @brief Validates the linear solver for electronic problems.
+ * The test-case is, in this case, Benzene.
+ */
+BOOST_FIXTURE_TEST_CASE(Test_LinearSolver_Electronic, BenzeneFixture) {
+#ifdef HAVE_TwoU1PG
+  using SimulatorType = SweepBasedLinearSystem<cmatrix, TwoU1PG, storage::disk, SweepOptimizationType::TwoSite>;
+  parametersBenzene.set("max_bond_dimension", 100);
+  parametersBenzene.set("init_type", "hf");
+  parametersBenzene.set("hf_occ", "4,4,4,1,1,1");
+  parametersBenzene.set("nsweeps", 10);
+  parametersBenzene.set("truncation_initial", 1.0E-30);
+  parametersBenzene.set("truncation_final", 1.0E-30);
+  parametersBenzene.set("linsystem_init", "last");
+  parametersBenzene.set("linsystem_max_it", 1);
+  parametersBenzene.set("linsystem_tol", 1.0E-10);
+  parametersBenzene.set("linsystem_krylov_dim", 10);
+  parametersBenzene.set("linsystem_exact_error", "yes");
+  // Defines variables required by the simulator
+  auto benzeneLattice = Lattice(parametersBenzene);
+  auto benzeneModel = Model<cmatrix, TwoU1PG>(benzeneLattice, parametersBenzene);
+  auto benzeneMPO = make_mpo(benzeneLattice, benzeneModel);
[…]

### 2023-05-08 dmrg/tests/SweepOptimizationTools/SweepBasedLinearSystemVibrational.cpp

@@ -1,0 +1,60 @@
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
+
+#define BOOST_TEST_MODULE SweepBasedLinearSystemVibrational
+
+#include <iostream>
+#include <boost/test/included/unit_test.hpp>
+#include "dmrg/SweepBasedAlgorithms/SweepBasedLinearSystem.h"
+#include "dmrg/models/generate_mpo.hpp"
+#include "dmrg/models/lattice/lattice.h"
+#include "dmrg/models/model.h"
+#include "dmrg/sim/matrix_types.h"
+#include "Fixtures/WatsonFixture.h"
+
+#ifdef DMRG_VIBRATIONAL
+
+/** @brief Checks that the linear system solver via interface works for vibrational problems. */
+BOOST_FIXTURE_TEST_CASE(Test_SweepBasedLinearSystemSS_Vibrational_Bilinearlty, WatsonFixture)
+{
+#ifdef HAVE_TrivialGroup
+  parametersBilinearly.set("nsweeps", 5);
+  parametersBilinearly.set("max_bond_dimension", 100);
+  parametersBilinearly.set("truncation_initial", 1.0E-20);
+  parametersBilinearly.set("truncation_main", 1.0E-15);
+  auto bilinearlyLattice = Lattice(parametersBilinearly);
+  auto bilinearlyModel = Model<matrix, TrivialGroup>(bilinearlyLattice, parametersBilinearly);
+  parametersBilinearly.set("init_type", "const");
+  auto bilinearlyMPS = MPS<matrix, TrivialGroup>(bilinearlyLattice.size(),
+                                                 *(bilinearlyModel.initializer(bilinearlyLattice, parametersBilinearly)));
+  bilinearlyMPS.normalize_right();
+  // Calculates the energy via the interface
+  parametersBilinearly.set("optimization", "singlesite");
+  parametersBilinearly.set("symmetry", "none");
+  maquis::DMRGInterface<double> interfaceBilinearly(parametersBilinearly);
+  interfaceBilinearly.optimize();
+  double energyFromInterface = interfaceBilinearly.energy();
+  // Parameters that are specific for the solution of the linear system.
[…]

### 2023-05-08 dmrg/tests/mpo/test_1D_mpo_electronic.cpp

@@ -1,0 +1,51 @@
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
+
+#define BOOST_TEST_MAIN
+
+#include <boost/test/included/unit_test.hpp>
+#include <boost/mpl/list.hpp>
+#include "dmrg/models/generate_mpo/1D_mpo_maker.hpp"
+#include "dmrg/mp_tensors/mpo_times_mps.hpp"
+#include "dmrg/mp_tensors/mps.h"
+#include "dmrg/models/model.h"
+#include "dmrg/models/lattice/lattice.h"
+#include "dmrg/sim/matrix_types.h"
+#include "Fixtures/H2Fixture.h"
+
+typedef boost::mpl::list<
+#ifdef HAVE_TwoU1PG
+TwoU1PG
+#endif
+> symmetries;
+
+/**
+ * @brief Checks that applying the destructor we get the same MPS of the cation.
+ */
+BOOST_FIXTURE_TEST_CASE_TEMPLATE( Test_MPO_Times_MPS_ExpVal, S, symmetries, H2Fixture ) 
+{
+    // Generates the HF MPS
+    parametersH2.set("init_type", "hf");
+    parametersH2.set("hf_occ", "4,1");
+    auto lattice = Lattice(parametersH2);
+    auto model = Model<matrix, S>(lattice, parametersH2);
+    auto mpsHF = MPS<matrix, S>(lattice.size(), *(model.initializer(lattice, parametersH2)));
+    // Creates the destructor operator
+    auto traitClass = MPOTimesMPSTraitClass<matrix, S>(mpsHF, model, lattice, 
+                                                       model.total_quantum_numbers(parametersH2),
+                                                       parametersH2["max_bond_dimension"]);
+    auto ionizedMPS = traitClass.ionizeMPS(0, generate_mpo::IonizedOrbital::Up);
+    // Creates the ionized MPS from the mps_intializer
+    parametersH2.set("hf_occ", "2,1");
+    parametersH2.set("u1_total_charge1", 0);
+    parametersH2.set("u1_total_charge2", 1);
+    auto modelCation = Model<matrix, S>(lattice, parametersH2);
[…]

### 2023-05-08 dmrg/tests/prebo_test1.cpp

@@ -1,29 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2011-2013 by Michele Dolfi <dolfim@phys.ethz.ch>
- *               2021 by Robin Feldmann <robin.feldmann@phys.chem.ethz.ch>
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #define BOOST_TEST_MAIN
 
@@ -205,7 +185,7 @@ BOOST_AUTO_TEST_CASE( PreBO_Test1 )
     optimizer.push_back("twosite");
 
     // Attention: const guess in test, so that the results are deterministic.
-    p.set("init_state", "const");
+    p.set("init_type", "const");
     // Measure RDMs
     p.set("MEASURE[1rdm]","1");
     p.set("MEASURE[mutinf]","1");
[…]

### 2023-05-08 dmrg/tests/test2.cpp

@@ -1,29 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2014 Institute for Theoretical Physics, ETH Zurich
- *               2011-2013 by Michele Dolfi <dolfim@phys.ethz.ch>
- *               2019 by Leon Freitag <lefreita@ethz.ch>
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #define BOOST_TEST_MAIN
 
@@ -34,428 +14,244 @@
 #include <iostream>
 #include "maquis_dmrg.h"
 #include "test_detail.h"
+#include "Fixtures/LiHFixture.h"
 
 // Test 2: LiH with d=1 Angstrom, singlet, cc-pVDZ basis set, CAS(2,4), integrals generated by MOLCAS
 // Ground and excited state calculations
-BOOST_AUTO_TEST_CASE( Test2 )
[…]

### 2023-05-08 dmrg/tests/test_ExcitonicCoupling.cpp

@@ -1,0 +1,70 @@
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
+
+#define BOOST_TEST_MAIN
+
+#ifdef DMRG_VIBRONIC
+
+#include <boost/test/included/unit_test.hpp>
+#include "Fixtures/VibronicFixture.h"
+#include "maquis_dmrg.h"
+
+/** @brief Watson-based calculation on an excitonic Hamiltonian. */
+BOOST_FIXTURE_TEST_CASE(Test_Vibronic_Excitonic_SingleSite, VibronicFixture)
+{
+#ifdef HAVE_U1
+    using InterfaceType = maquis::DMRGInterface<double>;
+    // Adds the final input parameters
+    parametersExcitonicAggregate.set("init_type", "default");
+    //parametersExcitonicAggregate.set("init_basis_state", "1,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0");
+    parametersExcitonicAggregate.set("optimization", "singlesite");
+    parametersExcitonicAggregate.set("nsweeps", 10);
+    parametersExcitonicAggregate.set("max_bond_dimension", 20);
+    parametersExcitonicAggregate.set("integral_file", "integral_file_Excitonic_Harmonic");
+    parametersExcitonicAggregate.set("ngrowsweeps", 2);
+    parametersExcitonicAggregate.set("nmainsweeps", 2);
+    parametersExcitonicAggregate.set("alpha_initial", 1.0E-8);
+    parametersExcitonicAggregate.set("alpha_main", 1.0E-15);
+    parametersExcitonicAggregate.set("alpha_final", 0.);
+    // Note that here we are imposing that J=0
+    parametersExcitonicAggregate.set("J_coupling", 0.);
+    // Creates the interface
+    InterfaceType interface(parametersExcitonicAggregate);
+    interface.optimize();
+    // The reference value can be calculated based on the Harmonic approximation.
+    auto refEnergy = 4812.5;
[…]

### 2023-05-08 dmrg/tests/test_mps_mpo_ops/test_mpo_times_mps_vibrational.cpp

@@ -1,0 +1,116 @@
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
+
+#define BOOST_TEST_MAIN
+
+#include <boost/test/included/unit_test.hpp>
+#include <boost/mpl/list.hpp>
+#include "dmrg/models/generate_mpo.hpp"
+#include "dmrg/mp_tensors/mps.h"
+#include "dmrg/mp_tensors/mpo_times_mps.hpp"
+#include "dmrg/models/model.h"
+#include "dmrg/models/lattice/lattice.h"
+#include "dmrg/sim/matrix_types.h"
+#include "Fixtures/WatsonFixture.h"
+
+#ifdef DMRG_VIBRATIONAL
+
+/**
+ * @brief Checks that [mpo_times_mps] gives results that are coherent with expval.
+ * Note that here we use 1) the harmonic MPS and 2) the harmonic Hamiltonian.
+ */
+BOOST_FIXTURE_TEST_CASE(Test_MPO_Times_MPS_None, WatsonFixture)
+{
+#ifdef HAVE_TrivialGroup
+  // Generates the HF MPS
+  parametersEthyleneWatsonHarmonic.set("init_type", "basis_state_generic");
+  parametersEthyleneWatsonHarmonic.set("init_basis_state", "0,0,0,0,0,0,0,0,0,0,0,0");
+  parametersEthyleneWatsonHarmonic.set("max_bond_dimension", 200);
+  parametersEthyleneWatsonHarmonic.set("Nmax", 6);
+  auto lattice = Lattice(parametersEthyleneWatsonHarmonic);
+  auto watsonModel = Model<matrix, TrivialGroup>(lattice, parametersEthyleneWatsonHarmonic);
+  auto watsonHarmonicMPO = make_mpo(lattice, watsonModel);
+  auto mps = MPS<matrix, TrivialGroup>(lattice.size(), *(watsonModel.initializer(lattice, parametersEthyleneWatsonHarmonic)));
+  // Calculates the MPS-MPO contraction
+  auto traitClass = MPOTimesMPSTraitClass<matrix, TrivialGroup>(mps, watsonModel, lattice, watsonModel.total_quantum_numbers(parametersEthyleneWatsonHarmonic),
[…]

### 2023-05-08 dmrg/tests/test_mps_mpo_ops/test_mps_mpo_ops_vibronic.cpp

@@ -1,0 +1,106 @@
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
+
+#define BOOST_TEST_MAIN
+
+#include <iostream>
+#include <boost/test/included/unit_test.hpp>
+#include "dmrg/models/lattice/lattice.h"
+#include "dmrg/models/model.h"
+#include "dmrg/mp_tensors/mps.h"
+#include "dmrg/mp_tensors/mps_rotate.h"
+#include "Fixtures/VibronicFixture.h"
+#include "dmrg/sim/matrix_types.h"
+
+#ifdef DMRG_VIBRONIC
+
+/** @brief Checks that the Hamonic energy is calculated correctly via [expval] */
+BOOST_FIXTURE_TEST_CASE(Test_ExpVal_U1_HarmonicEnergy_Vibronic, VibronicFixture)
+{
+#ifdef HAVE_U1
+    parametersVibronicPyrazineRedDim.set("init_type", "basis_state_generic");
+    parametersVibronicPyrazineRedDim.set("init_basis_state", "1,0,0,0,0,0");
+    auto vibronicLattice = Lattice(parametersVibronicPyrazineRedDim);
+    auto vibronicModel = Model<matrix, U1>(vibronicLattice, parametersVibronicPyrazineRedDim);
+    auto vibronicHarmonicMPO = make_mpo(vibronicLattice, vibronicModel);
+    auto mps = MPS<matrix, U1>(vibronicLattice.size(), *(vibronicModel.initializer(vibronicLattice, parametersVibronicPyrazineRedDim)));
+    BOOST_CHECK_CLOSE(norm(mps), 1., 1.0E-15);
+    auto energy = expval(mps, vibronicHarmonicMPO)/norm(mps);
+    auto referenceHarmonicEnergy = -2036.1425;
+    BOOST_CHECK_CLOSE(referenceHarmonicEnergy, energy, 1e-7);
+#endif // HAVE_U1
+}
+
+#ifdef HAVE_U1
+
+/** @brief Checks that the Hamonic excitation energy is calculated correctly via [expval] */
+BOOST_FIXTURE_TEST_CASE(Test_ExpVal_U1_HarmonicEnergyDifference_Vibronic, VibronicFixture)
+{
+    // Creates the vibrational ground state
[…]

### 2023-05-08 dmrg/tests/test_mps_mpo_ops/test_mps_overlap_vibrational.cpp

@@ -1,0 +1,68 @@
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
+
+#define BOOST_TEST_MODULE MPSOverlapVibrational
+
+#include <boost/test/included/unit_test.hpp>
+#include <boost/mpl/list.hpp>
+#include "dmrg/models/lattice/lattice.h"
+#include "dmrg/models/model.h"
+#include "dmrg/mp_tensors/mps.h"
+#include "dmrg/sim/matrix_types.h"
+#include "Fixtures/WatsonFixture.h"
+#include "dmrg/models/generate_mpo.hpp"
+#include "dmrg/mp_tensors/mps_mpo_ops.h"
+
+#ifdef DMRG_VIBRATIONAL
+
+
+/** @brief Checks that overlap between two ONV is zero for the TrivialGroup case */
+BOOST_FIXTURE_TEST_CASE( Test_MPS_Overlap_Vibrational, WatsonFixture )
+{
+#ifdef HAVE_TrivialGroup
+    // ONV1 
+    parametersEthyleneWatson.set("init_type", "basis_state_generic");
+    parametersEthyleneWatson.set("init_basis_state", "0,0,0,0,0,0,0,0,0,0,0,0");
+    auto lattice = Lattice(parametersEthyleneWatson);
+    auto model = Model<matrix, TrivialGroup>(lattice, parametersEthyleneWatson);
+    auto mpsHF1 = MPS<matrix, TrivialGroup>(lattice.size(), *(model.initializer(lattice, parametersEthyleneWatson)));
+    // ONV2
+    parametersEthyleneWatson.set("init_type", "basis_state_generic");
+    parametersEthyleneWatson.set("init_basis_state", "0,0,0,0,0,0,0,0,1,0,0,0");
+    lattice = Lattice(parametersEthyleneWatson);
+    model = Model<matrix, TrivialGroup>(lattice, parametersEthyleneWatson);
+    auto mpsHF2 = MPS<matrix, TrivialGroup>(lattice.size(), *(model.initializer(lattice, parametersEthyleneWatson)));
+    // Overlap calculationb
+    double overlapHF1 = overlap(mpsHF1, mpsHF2);
+    double overlapHF2 = overlap(mpsHF2, mpsHF1);
[…]

### 2023-05-08 dmrg/tests/test_realtime.cpp

@@ -1,28 +1,9 @@
-/*****************************************************************************
- *
- * ALPS MPS DMRG Project
- *
- * Copyright (C) 2021 Institute for Theoretical Physics, ETH Zurich
- *               2021 by Alberto Baiardi <abaiardi@ethz.ch>
- *
- * This software is part of the ALPS Applications, published under the ALPS
- * Application License; you can use, redistribute it and/or modify it under
- * the terms of the license, either version 1 or (at your option) any later
- * version.
- *
- * You should have received a copy of the ALPS Application License along with
- * the ALPS Applications; see the file LICENSE.txt. If not, the license is also
- * available from http://alps.comp-phys.org/.
- *
- * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
- * IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
- * FITNESS FOR A PARTICULAR PURPOSE, TITLE AND NON-INFRINGEMENT. IN NO EVENT
- * SHALL THE COPYRIGHT HOLDERS OR ANYONE DISTRIBUTING THE SOFTWARE BE LIABLE
- * FOR ANY DAMAGES OR OTHER LIABILITY, WHETHER IN CONTRACT, TORT OR OTHERWISE,
- * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
- * DEALINGS IN THE SOFTWARE.
- *
- *****************************************************************************/
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
 
 #define BOOST_TEST_MAIN
 
@@ -34,6 +15,7 @@
 #include "Fixtures/BenzeneFixture.h"
 #include "Fixtures/TimeEvolversFixture.h"
 #include "Fixtures/PreBOTimeEvolversFixture.h"
+#include "Fixtures/VibronicFixture.h"
 
 /**
  * @brief Tests that the energy is conserved along a TD-DMRG propagation.
@@ -83,27 +65,27 @@ BOOST_FIXTURE_TEST_CASE( TestRealTime, BenzeneFixture )
[…]

### 2023-05-08 dmrg/tests/vibronic/ExcitonicModel.cpp

@@ -1,0 +1,29 @@
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
+
+#define BOOST_TEST_MODULE MODEL_EXCITONIC_U1
+
+#include <boost/test/included/unit_test.hpp>
+#include <boost/mpl/assert.hpp>
+#include "dmrg/models/lattice/lattice.h"
+#include "dmrg/models/vibrational/u1/ExcitonicModel.hpp"
+#include "Fixtures/VibronicFixture.h"
+#include "maquis_dmrg.h"
+#include "dmrg/sim/matrix_types.h"
+
+/** Test for the integral parser for the trivial vibronic Hamiltonian */
+BOOST_FIXTURE_TEST_CASE(Test_Integral_Parser_Excitonic, VibronicFixture)
+{
+#ifdef HAVE_U1
+    auto lattice = Lattice(parametersExcitonicAggregate);
+    auto integrals = Vibrational::detail::parseIntegralExcitonic<double>(parametersExcitonicAggregate, lattice);
+    // Checks sizes
+    BOOST_CHECK_EQUAL(integrals.first.size(), 30);
+    BOOST_CHECK_EQUAL(integrals.second.size(), 30);
+    BOOST_CHECK_EQUAL(integrals.first[0].size(), 2);
+#endif // HAVE_U1
+}

### 2023-05-08 dmrg/tests/vibronic/VibronicModel.cpp

@@ -1,0 +1,85 @@
+/**
+ * @file
+ * @copyright This code is licensed under the 3-clause BSD license.
+ *            Copyright ETH Zurich, Laboratory of Physical Chemistry, Reiher Group.
+ *            See LICENSE.txt for details.
+ */
+
+#define BOOST_TEST_MODULE MODEL_VIBRONIC_U1
+
+#include <boost/test/included/unit_test.hpp>
+#include <boost/mpl/assert.hpp>
+#include "dmrg/models/lattice/lattice.h"
+#include "dmrg/models/vibrational/u1/VibronicModel.hpp"
+#include "Fixtures/VibronicFixture.h"
+#include "maquis_dmrg.h"
+#include "dmrg/sim/matrix_types.h"
+
+/** Test for the integral parser for the trivial vibronic Hamiltonian */
+BOOST_FIXTURE_TEST_CASE(Test_Integral_Parser_Vibronic, VibronicFixture)
+{
+#ifdef HAVE_U1
+    auto lattice = Lattice(parametersFakeVibronic);
+    auto integrals = Vibrational::detail::parseIntegralVibronic<double>(parametersFakeVibronic, lattice);
+    // Checks sizes
+    BOOST_CHECK_EQUAL(integrals.first.size(), 6);
+    BOOST_CHECK_EQUAL(integrals.second.size(), 6);
+    BOOST_CHECK_EQUAL(integrals.first[0].size(), 4);
+#endif // HAVE_U1
+}
+
+#ifdef HAVE_U1
+
+/** Tests the [create_terms] method for the fake vibronic model */
+BOOST_FIXTURE_TEST_CASE(Test_VibronicModel_Create_Terms, VibronicFixture)
+{
+    auto lattice = Lattice(parametersFakeVibronic);
+    auto fakeVibronicModel = VibronicModel<tmatrix<double>>(lattice, parametersFakeVibronic);
+    auto sizeBefore = fakeVibronicModel.hamiltonian_terms().size();
+    BOOST_CHECK_EQUAL(sizeBefore, 0);
+    fakeVibronicModel.create_terms();
+    auto sizeAfter = fakeVibronicModel.hamiltonian_terms().size();
+    BOOST_CHECK_EQUAL(sizeAfter, 6);
+}
+
+/** Checks consistency for the physical dimensions for the "trivial" vibronic model */
[…]
